### 2016-04-21 dissertation-export.tex
Copy core LaTeX files from Rebecca manuscript

ABSTRACT

\begin{longtable}[]{@{}ll@{}}
\toprule
Title of Document: & HOW ELECTRICAL ENGINEERING STUDENTS DESIGN COMPUTER
PROGRAMS\tabularnewline
&\tabularnewline
& Brian Adam Danielak, Doctor of Philosophy, 2014\tabularnewline
&\tabularnewline
Directed By: & Associate Professor Andrew Elby, Department of Teaching,
Learning, Policy, and Leadership\tabularnewline
\bottomrule
\end{longtable}

When professional programmers begin designing programs, we know they
often spend time away from a computer, using tools such as pens, paper,
and whiteboards as they discuss and plan their designs (Petre, van der
Hoek, \& Baker, 2010). But, we're only beginning to analyze and
understand the complexity of what happens during such early-stage design
work. And, our accounts are almost exclusively about what professionals
do. For all we've begun to understand about what happens in early-stage
software design, we rarely apply the same research questions and methods
to students' early-stage design work. This dissertation tries to redress
that imbalance. I present two case studies --- derived from my 10 study
participants --- of electrical engineering (EE) students designing
computer programs in a second-semester computer programming course.

In study 1, I show how analyzing a student's code snapshot history
\emph{and} conducting clinical interviews tells us far more about her
design trajectory than either method could alone. From that combined
data I argue students' overall software designs can be consequentially
shaped by factors --- such as students' stances toward trusting their
code or believing a current problem is a new instance of an old one ---
that existing code snapshot research is poorly equipped to explain.
Rather, explanations that add non-conceptual constructs including
affective state and epistemological stance can offer a more complete and
satisfactory account of students' design activities.

In study 2, I argue computer science and engineering education should
move beyond conceptual-knowledge and concept deficit explanations of
students' difficulties (and capabilities) in programming. I show that in
doing design students do, say, write, and gesture things that:

\begin{itemize}
\item
  Are outside the phenomenological scope of most (mis)conceptions
  accounts of programming
\item
  Would be explained differently under frameworks that emphasize
  manifold epistemological resources. Some student difficulties can be
  recast as epistemological blocks in activity rather than conceptual
  knowledge deficits. Similarly, some students' productive capacities
  can be understood as epistemologically-related stances toward an
  activity, rather than evidencing particular knowledge of specific
  computational concepts.
\item
  Would suggest different instructional interventions if teachers
  attended to the stabilizing aspects --- such as epistemological
  dynamics --- that help these episodes of activity cohere for students.
\end{itemize}

\begin{longtable}[]{@{}l@{}}
\toprule
HOW ELECTRICAL ENGINEERING STUDENTS DESIGN COMPUTER
PROGRAMS\tabularnewline
\bottomrule
\end{longtable}

By

\begin{longtable}[]{@{}l@{}}
\toprule
Brian Adam Danielak\tabularnewline
\bottomrule
\end{longtable}

Dissertation submitted to the Faculty of the Graduate School of the

University of Maryland, College Park, in partial fulfillment

of the requirements for the degree of

\begin{longtable}[]{@{}l@{}}
\toprule
Doctor of Philosophy\tabularnewline
2014\tabularnewline
\bottomrule
\end{longtable}

Advisory Committee:

\begin{longtable}[]{@{}l@{}}
\toprule
Associate Professor Andrew Elby, Chair\tabularnewline
Ayush Gupta\tabularnewline
Edward F. Redish

James G. Greeno\tabularnewline
William E. J. Doane\tabularnewline
Benjamin B. Bederson\tabularnewline
\bottomrule
\end{longtable}

\begin{longtable}[]{@{}l@{}}
\toprule
© Copyright by\tabularnewline
Brian Adam Danielak\tabularnewline
2014\tabularnewline
\bottomrule
\end{longtable}

\section{Dedication}\label{dedication}

To my grandparents. Anna and Josef Danielak endured and survived so that
I could be here. Sabina Schwab loved me and made funny faces at me when
no one else was looking. Henry Schwab continues to teach me.

I also dedicate this to my brother, Jason. My name may be on the cover
of this thing, but without his support it wouldn't exist.

I love you all.

\section{\texorpdfstring{\\
Acknowledgements}{ Acknowledgements}}\label{acknowledgements}

Once, in my high school Economics class, my teacher Mr. Mooney\footnote{That's
  not a pseudonym.} told us to take out a sheet of paper. We were to
spend 5 minutes writing down a list of all the people who were
responsible for getting us to school that day. I started off slowly.
Mom, for waking me up when my alarm clock didn't. Me, for remembering to
put gas in the car.\footnote{Yes, I counted myself.} Then I thought
about all the people who had maintained the car over the years. I
thought about the people who had designed the car, when it was just
modeling clay and blueprints, before a single piece of metal was bent
into a fender.\footnote{Or, for that matter, \emph{re-bent} when mom hit
  a lightpost in a parking lot.} Soon I was writing down the people who
made the antifreeze that kept the engine cool, Charles Goodyear for
developing vulcanization techniques that make modern tires work, and the
guy at Jiffy Lube who told me---and I think I'm quoting here---''normal
transmission fluid should be red, like fruit punch; your transmission
fluid is brown, like Thanksgiving gravy with metal shavings in it.''

I was surprised---having filled the front of the sheet and most of the
back---when Mr. Mooney told us to stop. He said the point of the
exercise was to show that as a society we are interdependent. That
interdependence is key to understanding the very basics of economic
theory: scarcity, pricing, trade, comparative advantage. But, I was
still stuck on how many people were on my list. Most of them were people
I'd never even met. It wasn't that I didn't care about economics; it was
that for the first time I realized in a big way how many people were
responsible for where I was.

I can't exhaustively thank everyone who helped me make this. Moreover,
some people, having read its contents, might wish to dissociate
themselves entirely from it. To those who made this dissertation
possible, I hope my thanks below will suffice. To those I leave out,
please don't let my imperfections reflect on the job you did. Lastly, to
those who wanted no public record of having been associated with this,
I'm sorry.

I would first like to thank my advisors, Andy Elby and Ayush Gupta.
Nearly six years ago I made the decision to leave my Ph.D. program, my
school, and my city behind to join Andy's program. It wasn't long until
I was working with, joking with, writing with, and being advised by Andy
and Ayush. From Andy, I learned to be clever, argue well, and always try
to find the good, redeeming features of things. From Ayush, I learned
the importance of bringing empathy and compassion to my work and my
life. Any graduate student would be lucky to have just one of them as an
advisor. I struck gold with both.

My thanks go also to my committee. Ben Bederson graciously welcomed me
into Maryland's human-computer interaction lab, all the while offering
grounded and focused feedback on my work. Joe Redish has been an upbeat
supporter of---and at times loyal opposition to---my research. My
thoughts and insights have only grown stronger with his contributions.
Jim Greeno has seen my intellectual development from the beginning. It
is not an exaggeration to say parts of this dissertation were forged in
the crucible of DC rush hour as I tried to navigate traffic and answer
his questions about my research. I thank him for entrusting me with his
vehicular safety. I thank him as well because at a dinner in June of
2011 he restored my faith that some people get into academia to try and
do good. Finally, Wil Doane helped get me into this crazy mess in the
first place. In addition to being a mentor in computer science, Wil has
been an unflinching champion of my ideas for as long as we've known each
other. He first showed me what passion and innovation can look like in a
computer science classroom; I hope this dissertation is a natural
extension of those ideals.\footnote{And if it's not, I hope you'll
  forgive me.}

I'd like to thank my compatriots in the Physics Education Research Group
(PERG) at Maryland. Jessica Watkins, Chandra Turpen, Vashti Sawtelle,
and Julia (Svoboda) Gouvea were model post-docs and amazing listeners. I
look up to all of them, which at least Julia has confirmed makes her
uncomfortable. Tiffany Sikorski has always been impressed with my work,
which means so much to me because I think she's a very hard person to
impress. Lama Jaber was a constant source of warmth, compassion, and
joy. I was very sad to see her leave for Boston, but I'm so honored to
have her as a friend. Ben Dreyfus offered a sympathetic ear and ongoing
encouragement when I got stuck thinking about students thinking about
programming. Luke Conlin was my stalwart conference roommate,
intellectual companion, cheerleader from afar, and fellow pillow fort
architect. Gina Quan taught me how to hug; her non-hug-related work
teaching girls to program Arduino robots has me endlessly excited. And,
thanks to PERG's impressive alumni network, Rosemary Russ helped me
think through some tough data analysis once I got to Madison.

Eric Kuo and Mike Hull were my teammates in making sense of engineering
students making sense of mathematics. Without Eric and Mike taping,
analyzing, and dissecting classroom video my work couldn't exist. Eric
in particular endured a roundtrip car ride from Boston to Bar Harbor in
which I panicked about everything I didn't understand about our
research.\footnote{I still panic about it. I'm just older now.}

Jen Richards, Colleen Gillespie, and Jason Yip have been with me in
Science Ed since the beginning. It was a painful change when we stopped
taking the same courses, changed advisors, and changed geographies. But,
we gave writing feedback to each other, supplied references to each
other, and kept each other in good humor when things got hard. I'd like
to single out Colleen for being the first one of us to prove this whole
dissertation thing was possible. And, I'd like to single out Jen for
watching over me and watching out for me. When Jen noticed I was
struggling, she talked to me about it --- a simple act that got me into
engineering education research and led to all of this. In so many ways,
I wouldn't be here without her.

It may buck tradition, but I'd also like to thank my undergraduate
mentors who have been with me for almost a decade. Robert Daly first
convinced me to come to the University at Buffalo. He later, to my
delight, became my mentor in English and agreed to supervise my
undergraduate thesis. It turned out the ideas in that thesis---while
poorly argued and barely cogent---would form the basis of my
intellectual life for the past six years.\footnote{I know. I was
  surprised too.} Josephine Capuana backed me from the beginning. As my
advisor she supported me no matter what I wanted to do and how many
majors I cycled through to do it. She is also the reason I met my second
undergraduate mentor, Kenneth Takeuchi. Dr. Takeuchi has backed me, time
and time again, since I was a freshman in college. He and Amy Marschilok
taught me about mentoring, inspired me to become an academic, and
continue to serve as models for what it means to care about students.

I'd like to thank the friends who have been there to prop me up and help
me think. The first group is so big it gets its own paragraph. Robert
Davis and Anne Davis welcomed me into their home and their family where
everyone seemed to me to be strangely excited by what I was working on.
Will Davis and Jamie Davis gave C help when I needed it. Russ Davis
inspires me with his dedication to serving the country. Nora Deram and
Mat Deram, in addition to being improbably lovely people, are both
helping to build the future of mathematics. Travis Pond took me to the
hallowed ground of the Apollo 1 launchpad so I could witness the
engineering marvel of putting a rocket into space. And Nan Pond has been
my friend, bandmate, fellow drum major, R convert, transatlantic pen
pal, and support system for nearly 15 years. Here's to uncountably
infinitely many more.

Al Stein took me to lunch and reminded me that I've been boldly taking
chances with my writing since I was in tenth grade. Vanessa Svihla has
been my colleague, co-author, and co-conspirator in embracing the
importance of design in our work. Jae Vick, Cheryl Medley, Danielle
Champney, and Christie Veitch constantly affirmed I could do this,
despite my doubts. Heidi Thalman and Adam Lloyd reminded me I am
awesome, despite my stern objections to the contrary. And, a ragtag
group of Physics Education and Science Education Researchers have been
there for me electronically: Warren Christensen, Eleanor Close, Hunter
Close, Dedra Demaree, Matty Lau, Sissi Li, Michael Loverude, Sandy
Martinuk, Sam McKagan, Rachel Scherr, and Michael Wittmann. I'd
especially like to thank Rachel Scherr, Hunter Close, and Michael
Wittmann for being so welcoming to me at ICLS 2010.

The people dearest to me get the last word. Mom, thank you for having
me. Thank you for raising me. Thank you for loving me. Thank you for
doing all of that when it wasn't easy. Also, thank you for that time I
called you at 4 am when I was sick.

Dad, thank you for loving me. Thank you for supporting me. Thank you for
that time I climbed the tree in our driveway and didn't know how to get
back down, so you helped me. And thank you for teaching me how to ride
my bike. You might not know it, but almost everything I think about when
I think about teaching traces back to that.

Jason, I already dedicated this to you. Stop looking for more credit.
It's unbecoming.

And, guys: this is it. This is my Science Book.

\section{\texorpdfstring{\\
Table of Contents}{ Table of Contents}}\label{table-of-contents}

Dedication ii

Acknowledgements iii

Table of Contents x

List of Tables 1

List of Figures 2

1 Introduction 1

2 Study 1 -- Retrospectively analyzing the design of a student's project
12

2.1 Abstract 12

2.2 Introduction 13

2.2.1 Studying the software design practices of experts 13

2.2.2 The missing attention to students' design thinking 18

2.2.3 The current state of research on students' code snapshots 22

2.3 Context and Methods 26

2.3.1 Context of the study 26

2.3.2 Methods 28

2.4 Code Snapshot Data and Analysis 31

2.4.1 Rebecca's file-scanning solution is computationally complex 31

2.4.2 Rebecca's file-scanning solution ignores an assignment directive
33

2.4.3 Rebecca repeats the same chunk of code seven times 36

2.5 Interview Data and Analysis 39

2.5.1 Rebecca employed fscanf loops because she was deliberately reusing
from an Basic Programming Assignment. 40

2.5.2 Rebecca repeated code because she wanted to re-use functionality
she could trust 44

2.6 Discussion 48

2.6.1 Rebecca's key design decisions are made early, persist through to
her final submission, and carry consequences 48

2.6.2 Her design decisions may be influenced by framing and overzealous
transfer 50

2.7 Conclusion 53

3 Study 2 -- What directs and sustains students' in-the-moment
programming activity? 55

3.1 Introduction 55

3.2 Literature Review 57

3.2.1 Computing culture should support a diversity of ways of knowing 58

3.2.2 Misconceptions research in computing education tends to ignore
students' productive knowledge 63

3.2.3 Not all cognitive programming bugs imply a problem with the
student 73

3.2.4 Examples motivate the need for contextual-sensitivity in modeling
programming cognition 76

3.2.5 Manifold models of cognition explain context-dependence and the
growth of expertise 84

3.3 Methods and Theoretical Commitments 88

3.3.1 Student population, course background, and selection 88

3.3.2 Studying design as a complex phenomenon of disciplinary practice
90

3.3.3 Deploying methods to capture the complexity of early-stage design
work 92

3.3.4 Developing a revelatory case of students' early-stage design work
95

3.4 Lionel's approach to design and to programming 97

3.4.1 While modifying a bike Lionel decomposed tasks, explained
solutions to himself, and created intermediate design representations 97

3.4.2 Lionel's approach to programming resonates strongly with the
design stance he held when working on his bike 105

3.4.3 Gestural, inscriptional, and verbal in-interview evidence reveals
Lionel's resources for structuring a program 109

3.4.4 We can trace continuities in Lionel across different kinds of
design activities 126

3.5 Rebecca's approach to programming 127

3.5.1 Rebecca's struggles on the iTunes project make sense in light of
some prior programming experiences she had 129

3.5.2 Rebecca has productive capacities for making progress on software
design work 150

3.6 Conclusion 176

3.6.1 Students' early-stage design activity reveals patterns outside the
explanatory scope of (mis)conceptions accounts 176

3.6.2 Rebecca had epistemological resources to support expert-like
practices, but she framed those practices differently 177

3.6.3 Students displayed a diversity of approaches to programming in the
moment 181

3.6.4 Dynamic epistemological models can offer a lens for reforming
assessment and instruction. 185

4 Conclusion 188

4.1 We should think carefully about what students' programming design
knowledge means for assessment 188

4.2 What We Might Change About Classroom Practice 191

4.2.1 Instructors could look beyond content to understand student
difficulties 191

4.2.2 Instructors could use code history to inform interventions 192

4.2.3 Instructors could establish a norm of asking why a design choice
makes sense 194

4.3 Final Remarks 195

5 Appendix 1 -- Transcript conventions 197

6 Appendix 2 -- Visual conventions for gestures 198

7 Appendix 3 -- Transcript of Rebecca's pseudo-code episode
\emph{without} gesture codes 199

8 Appendix 4 -- Conceptual knowledge frameworks in computing don't tell
us much about how Lionel structured an in-interview program 202

8.1 Analyzing Lionel's Solution From a Conceptual Knowledge View 202

9 Appendix 5 -- Neverly-Asked Questions (NAQs) 205

9.1.1 Neverly Asked Questions (NAQs) about my conceptual analysis of
Lionel's code 205

9.1.2 Neverly-asked questions about Lionel's verbal pseudo-code
description 207

10 References 213

\section{List of Tables}\label{list-of-tables}

Table 1 -- An overview of my research participants and the types of data
I collected 9

Table 2 -- Five different interpretations of semantically equivalent
programming statements in the same language 78

Table 3 -- Three syntactically-similar statements with very different
semantics 80

Table 4 -- Using room lighting configurations to think about
epistemological framing 87

Table 5 -- Comparing Lionel's words and actions across different stages
of ``programming'' over the same putative section of his program. I have
bolded some language-specific syntax. 122

Table 6 -- Comparing Lionel's design activity across contexts 126

Table 7 - Rebecca describes an array of structures using gestures
(Interview 4 of 5, April 6, 2012) 153

Table 8 -- Rebecca says gets lost on the array of pointers pointing to
track names (Interview 4 of 5, April 6, 2012) 156

Table 9 -- Rebecca starts talking out pseudo-code and asks for the pen
to write it 167

Table 10 -- Rebecca's verbal/gestural overview of looping through the
input track titles (Interview 4 of 5, April 6, 2012) 169

Table 11 -- Rebecca concludes her visual and gestural explanation for
scanning in track titles (Interview 4 of 5, April 6, 2012) 174

Table 12 -- Comparing Lionel's and Rebecca's views toward pseudo-code
178

\section{\texorpdfstring{\\
List of Figures}{ List of Figures}}\label{list-of-figures}

Figure 2‑1 -- A side-by-side comparison of two blocks of Rebecca's
flight-scanning code 37

Figure 3‑1 -- An comic-based overview of my methods for capturing
students' design practices 95

Figure 3‑2 -- An overview of Lionel's representational cascade and the
features and affordances of each representational layer. 111

Figure 3‑3 -- Lionel's final source code for the range-finding prompt
113

Figure 3‑4 -- Lionel makes a cycloid/helix gesture as he says ``parse
through this array'' 117

Figure 3‑5 -- Rebecca's compilation activity over time for Project 2.
Each compile a student initiates creates a commit, so long as there has
been a change to the underlying code. The height of each bar maps to the
number of commits recorded that day. The red line charts
\emph{cumulative} commits over time. The red line is steepest in periods
of frequent activity and shallowest in periods with little or no compile
activity. 144

Figure 3‑6 -- The instructor required students to use this data
arrangement for storing information in the music server. The scheme uses
an array of pointers to represent the track names of an album. 151

Figure 3‑7 -- I write down what I understand as Rebecca explains the
overall data structure of albums to me 155

Figure 3‑8 -- The candidate pseudo-code Rebecca dictated to me for
declaring an array of pointers 166

Figure 3‑9 -- Rebecca's pseudo-code for scanning in and storing track
names 167

Figure 8‑1 -- Highlighting the conceptual features of Lionel's code
given Elliott Tew's (2010) outline of CS 1 concepts 204

\section{}\label{section}

\section{Introduction}\label{introduction}

Seymour Papert's (1980) book \emph{Mindstorms} has a curious section. Or
at least I thought so. The book collects years of Papert's thoughts,
observations, and experiments with teaching children at school to
program. The famous\footnote{Bearing in mind that ``fame,'' in the
  subfield that is computing education, is a relative thing.} Turtle and
LOGO --- the language spoken and understood by it --- let students
express geometrical ideas in the form of procedures and debug procedures
using the spatial affordances of geometry. My school did not have LOGO,
so I came to see the beauty of programming much, much later in life.
But, for many children in Papert's studies, LOGO became a venue for
expression and creativity.

The curious section in \emph{Mindstorms} is one in which Papert
describes ``structured programming,'' a technique for breaking down
large procedures into ``mind-sized bites.'' In short, it's possible to
write lots of programs ahierarchically as a list of instructions:

FORWARD 50;

RIGHT 120;

FORWARD 50;

RIGHT 120;

FORWARD 50;

RIGHT 120;

This little program would make a LOGO turtle draw an equilateral
triangle. By telling the turtle how much to go forward (in distance),
how much to turn (in degrees), and in which direction to turn (in this
case right), one can make a turtle trace out this equilateral triangle
as a sequence of steps.

But, this program doesn't have much structure. And, as the program grows
--- say, to accommodate having the turtle draw a simple house ---
managing the program becomes more complicated. It's possible, in
principle, to have dozens or hundreds of simple little statements and it
can become more and more difficult to figure out which part of the
program drew which part of the picture.

So, I'll propose a modification:

TO MAKEATRIANGLE:

FORWARD 50;

RIGHT 120;

FORWARD 50;

RIGHT 120;

FORWARD 50;

RIGHT 120;

What I've done is define a \emph{procedure} in LOGO. In a way, the
procedure bundles together the list of discrete steps and offers a sense
of \emph{structure} to the program. The procedure is now also something
I can compose to make more complicated things. Now, for example, I can
draw a bunch of triangles that all share a vertex and make a kind of
pinwheel:

MAKEATRIANGLE;

RIGHT 120;

MAKEATRIANGLE;

RIGHT 120;

MAKEATRIANGLE;

RIGHT 120;

If I apply the same kind of thinking to the pinwheel that I did to the
triangle, I might do this:

TO MAKEAPINWHEEL:

MAKEATRIANGLE;

RIGHT 120;

MAKEATRIANGLE;

RIGHT 120;

MAKEATRIANGLE;

RIGHT 120;

Now imagine two students, both of whom are trying to get the turtle to
draw a pinwheel.

\begin{longtable}[]{@{}ll@{}}
\toprule
Alan's Code & Ada's Code\tabularnewline
\midrule
\endhead
\includegraphics[width=2.00000in,height=5.04861in]{media/image1.emf} &
\includegraphics[width=2.40278in,height=4.62500in]{media/image2.emf}\tabularnewline
\bottomrule
\end{longtable}

The picture Alan's turtle makes should be identical to the picture Ada's
turtle makes.\footnote{Unless I screwed up, which is possible. If so,
  that's my fault; don't hold Alan and Ada accountable.} So, if all we
care about is the final art, Alan and Ada are making the same thing.
They're indistinguishable as students from the perspective of what the
turtle draws on paper.

But, clearly Alan's code and Ada's code aren't the same. Ada's program
evinces what Papert (1980) would call \emph{structured programming}:
pinwheels are composed of triangles, while triangles are composed of
straight line segments.\footnote{Not to get pedantic on you, but yes,
  Ada could have written a program that was perfectly syntactically
  valid --- and produced the same picture --- while choosing different
  and possibly bogus procedure names. By renaming functions, for
  example, Ada could have forced me to say that in her program
  ``unicorns are composed of Cheetos, and Cheetos are composed of
  dinosaur fingers.'' And, while \emph{my restatement} would make no
  sense, neither LOGO nor the turtle would have a problem interpreting
  Ada's procedure to make what Ada would call a pinwheel. The dinosaur
  might have a problem, I guess.} Alan's code doesn't compose things.
Every line in Alan's code is a direct instruction to the turtle and
there is no sense of hierarchy. Instead, there's just a serial flow of
instructions. To think more about the differences between Alan and Ada,
let's go back to Papert.

Papert (1980) discusses structured programming in the case of two
children. One child, Robert, embraced structured programming. Robert
extolled the style: ``see, all my procedures are mind-sized bites''
(Papert, 1980). And, indeed, programmers today are given similar advice
in books on professional programming practice (Martin, 2009). But, Keith
was a child who resisted. His programs remained ahierarchical;
``featureless'' in Papert's words. Debugging was harder for Keith, too,
because it was harder to locate problems in a program devoid of
structure.

If Keith was rather stubbornly sticking to his ahierarchical style,
should a teacher have proactively \emph{instructed} Keith to change?
Papert says no. Rather, as Keith encountered troublesome errors
debugging his code, he would meet with familiar advice:

When a child in this predicament asks what to do, it is usually
sufficient to say: ``You know what to do!'' And often the child will
say, sometimes triumphantly, some-times sheepishly: ``I guess I should
turn it into subprocedures?'' The ``right way'' was not imposed on
Keith; the computer gave him enough flexibility and power so that his
exploration could be genuine and his own. (Papert, 1980)

But, largely, that's where Papert leaves things. As Papert would have
it, most children will inevitably run into increasing frustration when
they debug. When children hit those problems, it's up to the teacher to
patiently remind the student that, in a sense, ``we've covered this.''
After repeatedly hitting this wall most students will see the light ---
eventually --- and come to save themselves from frustration by adopting
structured programming.

\emph{But what if they don't?}

I knew from decades of research in science education that students do
not always adopt the practices we want them to. As Hammer (1994)
essentially put it, \emph{not all novices go on to develop expertise,}
and many cognitive formulations of the novice-expert continuum didn't
(and still don't) really explain why. Was structured programming
something more than a practice of keeping code tidy and easing
debugging? \emph{Yes}, I believed so. Structured programming wasn't the
computer science equivalent of ``always labeling your units'' in
science\footnote{Don't misunderstand me; labeling units is important.
  The fact that early versions of LOGO didn't drives me \emph{nuts}.
  But, if you asked a bunch of sophisticated chemists what are the chief
  intellectual practices of their discipline, I don't think ``labeling
  units'' will top the list. If you ask Abelson \& Sussman (1996) what
  the chief intellectual practice of computer programming is, they'll
  tell you it's managing the complexity of large software systems.}; it
reflected a larger view on how we think about procedure and a
concomitant willingness to compose big complex ideas out of small
atomically-understandable ones (Abelson \& Sussman, 1996; Papert, 1980).

Back to Alan and Ada. With all this talk about structured programming,
we've mostly been focusing on \emph{code}: how it's structured
differently and what affordances and drawbacks result from the design.
We still don't have a sense of how Alan and Ada actually built their
programs. We can't know, just from these snippets, whether Ada's code
started with a compositional structure. Perhaps Ada started just the way
Alan did, so there was some intermediate state of Ada's code before she
refactored it when it was just as ahierarchical and featureless as
Alan's code. Or, perhaps Alan started by \emph{trying} to structure his
program, but he ran into some errors. Maybe Alan's literal code is the
result of a design retreat late in the game. It's not that he doesn't
\emph{appreciate} structured programming; he just couldn't get it to
work and, feeling pressured to produce a proper pinwheel, resorted to
taking out the structure and leaving just concrete instructions for the
Turtle.

I had questions.

Would all students come to structured programming after banging their
heads enough against a buggy featureless program? \emph{Maybe}. Would
they do it on a timescale we could see? \emph{I wasn't sure}. Was it
possible students could get through one, or even multiple semesters of
programming without ever embracing structured programming for
themselves? \emph{Quite possibly}. Outside of Papert's schools, was it
possible for research to not only track but explain why some students
used structured programming and other students didn't? Were there Alans
and Adas out there in real life whose complex stories could help us
understand more about how students learn structured programming?

This dissertation is my stab at beginning to answer those questions. The
focus of my study is a programming course for electrical engineering
students. I chose that course and population for two reasons. First and
foremost, I had preliminary evidence suggesting students were emerging
from the course with wildly different mindsets about its relationship to
engineering and, by extension, design. One student, Larry, saw
structured programming as a natural extension of the way he tried to
make sense of the physical world. Another student dismissed the class
entirely as ``not even a real engineering course.'' Second, my
connections to the school of engineering allowed me relatively easy
access to the class.

Over two semesters I observed lectures and interviewed students. In
interviews, I was able to use video recordings, screen capture software,
and an electronic pen to capture what students said, gestured, typed,
and wrote when they programmed. I later augmented the study by tracking
students' code histories. I co-developed a simple automated system to
create version-controlled repositories of the multi-week projects
students worked on for the course. Each time a student compiled, our
system captured a snapshot of the entire codebase (called a ``commit''
in Git, the versioning system we used). With each snapshot we also
captured the input students passed to the compiler and any compile-time
messages (including errors and warnings) they would have seen.

Table 1 shows a summary of all the participants in my studies and the
kind of data I collected.\footnote{The ``Inscriptions'' column is
  currently blank because I've been having trouble accessing that
  portion of my dataset. I hope to resolve my technical difficulties so
  I can provide inscription counts in later revisions of this
  dissertation.} The solicitation process began at the beginning of each
semester. During the second class lecture I read a 3-minute IRB-approved
speech explaining that I was studying how students understand
programming code. Students who expressed interest were invited to
participate in interviews, and for each interview they completed they
were paid \$15. Interviews were scheduled opportunistically based on
when students were available. In total, the data corpus comprises more
than 20 hours of clinical interviews and more than 2,500 distinct code
snapshots. Of the 10 students who participated, I chose to focus on
Rebecca (who appears in both study 1 and study 2) and Lionel (who
appears in study 2).

\protect\hypertarget{ux5fToc252445955}{}{}Table -- An overview of my
research participants and the types of data I collected

\begin{longtable}[]{@{}llllll@{}}
\toprule
\textbf{Semester} & Participant & Interviews\footnote{All interviews
  were videorecorded with the exception of Toby's. Toby requested to be
  audiorecorded only.} & Screencaptures & Inscriptions\footnote{At
  present, technical difficulties prevent me from reporting an exact
  count of inscriptions collected.} & Code Snapshots\tabularnewline
\midrule
\endhead
\textbf{Fall 2011} & Lionel & 1 & 1 session & & ---\tabularnewline
& CJ & 1 & --- & & ---\tabularnewline
& Donna & 2 & 2 sessions & & ---\tabularnewline
& Sam & 1 & --- & & ---\tabularnewline
& Toby & 2 & --- & & ---\tabularnewline
& Will & 1 & 1 session & & ---\tabularnewline
\textbf{Spring 2012} & Isaac & 3 & 1 session & & 434\footnote{Due to an
  unidentified error in Isaac's code snapshot history, we lack data
  between February 21 and April 12. we know in aggregate what changed in
  his code between those two dates; we just can't resolve that aggregate
  change down into individual changes for each time Isaac compiled
  between those dates.}\tabularnewline
& Dana & 3 & 2 sessions & & 878\tabularnewline
& Natalie & 4 & 1 session & & 262\tabularnewline
& Rebecca & 5 & 3 sessions & & 959\tabularnewline
\bottomrule
\end{longtable}

This dissertation is structured into two studies that analyze data from
the corpus. Study 1 asks what we can learn if a student submits a
program like Alan's. In other words, when one student's final code
submission is structured in peculiar ways (or not at all), can research
help us retrospectively recover the story of how that design came to be?
Study 2 in effect asks what directs and sustains Alan and Ada's
in-the-moment programming activity. I analyze rich data --- including
talk, gestures, inscriptions, and screen captures --- of two real
students. The analysis tries to understand what constitutes the
approaches they take when they code. Along the way, study 2 offers an
illustration of why strictly ``conceptual'' accounts of students may not
be enough to explain certain classes of phenomena in learning to
program.

\section{Study 1 -- Retrospectively analyzing the design of a student's
project}\label{study-1-retrospectively-analyzing-the-design-of-a-students-project}

\subsection{Abstract}\label{abstract}

This paper focuses on a historically understudied area in computing
education: attending to students' \emph{design thinking} in
university-level introductory programming courses. We detail the case of
Rebecca, a first-year Electrical Engineering student taking a required
2\textsuperscript{nd}-semester programming course in C. Our analysis
focuses on two related aspects of Rebecca's code for a multi-week
project:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The origin, nature, and evolution of unusual structural and behavioral
  features of Rebecca's code
\item
  The subtle, yet complex reasons that led Rebecca to make particular
  design choices in her code
\end{enumerate}

Our data comes from ethnographic observation of Rebecca's class,
fine-grained compile-time snapshots of Rebecca's codebase, and
semistructured interviews with Rebecca. We first present an analysis of
the compile-time snapshots, detailing Rebecca's unusual use of
file-scanning loops and her seven-fold repetition of a particular code
chunk (once for each day of the week). We then augment that analysis
with data from semi-structured interviews with Rebecca, which reveal
that affect (Eynde \& Hannula, 2006; Hannula, Evans, Philippou, \& Zan,
2004) and framing (Hammer, Elby, Scherr, \& Redish, 2005; van de Sande
\& Greeno, 2012) offer substantial explanatory power for understanding
why Rebecca made particular design choices.

\subsection{Introduction}\label{introduction-1}

\subsubsection{Studying the software design practices of
experts}\label{studying-the-software-design-practices-of-experts}

In 2010, the journal \emph{Design Studies} devoted an entire issue to
studies of how professional software engineers design complex systems.
The journal featured five different research perspectives on the same
dataset: videos of three different professional software engineering
teams trying to design a traffic simulator. The existence of the journal
issue, and the investigations contained therein, were motivated by what
the issue's editors saw as a growing need to know more about how
software design gets done in practice. In particular, the editors
contended, fields of design studies, interaction analysis, and
human-computer interaction don't know enough about how software
engineers use representations and collaborative exchanges to organize
the beginning phases of a design:

{[}d{]}uring formative design, software engineers spend a great deal of
time engaging in creative, exploratory design thinking using pen and
paper or a whiteboard---\emph{whether alone or in a small group}.
However, not enough is known about how software designers work in such
settings. What do designers actually do during early software design?
How do they communicate? What sorts of drawings do they create? What
kinds of strategies do they apply in exploring the vast space of
possible designs? (Petre et al., 2010, p. 533; my emphasis added)

Research in that issue takes on a variety of challenges in the study of
expert software design practice. One challenge, for example, is that of
understanding how engineers process, prioritize, and cope with design
requirements. Ball, Onarheim, and Christensen (2010) observed that
engineers deploy mixed strategies, developing solutions breadth-first
for easy problems and depth-first for more complex problems. They also
found that the more complex a requirement became, the more likely
engineers were to create speculative simulations (through talk and
representations) about how a system might work to solve that problem
(Ball et al., 2010).

Looking at design sessions longitudinally, Baker and van der Hoek (2010)
explored the shape and trajectory of how ideas generated in the design
process develop and relate to one another. Those researchers found that
roughly a third of the ideas discussed in a typical design session
``were reiterations or rephrasings of previously stated ideas'' (Baker
\& van der Hoek, 2010, p. 604). While it is perhaps frustrating that
ideas would be repeated so much, the authors argue such repetition can
be viewed as a kind of continual revisitation to make sure a proposed
design coheres:

Rather than representing a failure on the part of the designers, this
repetition seems to be a necessary character of successful design
sessions. Each time an idea is resurrected it is placed them {[}sic{]}
in a new context, and compared to different aspects of the system. In
this way, a concept of compatible, elegant design ideas is slowly
converged upon. (Baker \& van der Hoek, 2010, p. 607)

Perhaps most germane to the work presented herein, Jackson (2010)
explores the role of \emph{structure} in software design. Jackson's
meaning in using the word ``structure'' is broad. It can best be
described as the arrangement of and relationships between the elements
of a system. But, the broadness is deliberate, because Jackson's overall
argument is that the work of software often involves a coordination of
different structures, some of which drive the organization of others. In
the case of Jackson's (2010) study, designers worked with various kinds
of structured sets to build software that could simulate traffic
patterns in a city section. The work of design then, in part, becomes
that of defining and coordinating structural relationships at various
levels of abstraction:

\begin{itemize}
\item
  The \emph{system-to-be-modeled}, in this case a system of roads and
  intersections on which one can simulate traffic. Schematically, this
  system can be structured on the Cartesian plane of a map. Its
  representation (a diagrammatic map-like network of streets) works to
  organize how the system-to-be-modeled looks in the real world.
\item
  The \emph{actors} present in the system-to-be-modeled. In this case,
  that means abstracting from the map to all of the \emph{actors} that
  simulator software would need to send data to and get data from:
  including traffic signal units, traffic signal controllers, drivers,
  and vehicle sensors.
\item
  The \emph{meta-entities} that must exist to execute the actual
  simulation. Crucially, these abstractions may not be the same as the
  actors \emph{in} the system.\footnote{For example, suppose we were
    building an AI for chess. The ``entities in the system'' are the
    pieces (rooks, queens, knights, etc.) and the actual board. But,
    those pieces alone aren't enough to create a chess AI. One must also
    create entities that make decisions, move the pieces, maintain the
    state of the board, determine the legality of moves, etc. Thus,
    bishops may be ``entities in the system'' that know how to move
    (only along diagonals, up to 8 spaces). But, bishops need another
    meta-system entity to tell them \emph{when} and \emph{where} to
    move. That distinction is important because it means simulating
    chess involves much more than building a board and pieces, even
    though the board and pieces may constitute the only outward-facing
    aspects of a chess AI.} In this case, additional computational
  entities must be added to carry out a traffic simulation, such as an
  arrivals model and a simulation clock (Jackson, 2010, p. 559)
\item
  The \emph{programming code} that embodies the data and behaviors of
  the entities and meta-entities. While to some degree aesthetic, the
  symbolic structure of code is in many ways driven by decisions about
  structural relationships in other levels of abstraction.
\end{itemize}

How engineers develop and interact with even the first three kinds of
structural representations can profoundly influence the nature and
quality of the code they produce. As Jackson (2010) observes (in notable
contrast to (Baker \& van der Hoek, 2010)), early negotiations of
structure led to design decisions that were \emph{not} revisited later:

The traffic simulation problem was of a kind unfamiliar to all three
design teams. All three quite rightly tried immediately to assimilate
the problem to something they already knew. Two teams took the view that
the problem was an instance of the Model-View-Controller (MVC) software
pattern; the third identified the problem as `like a drawing program'.
Unfortunately, these hasty classifications were inadequate; but in every
case they were accepted uncritically and never explicitly questioned.
(Jackson, 2010, p. 564)

Ultimately, careful study of how software engineers design shows us the
enormous complexity involved in producing effective and thorough code.
And, much of that complexity is both afforded and constrained by talk,
representational infrastructure, and knowledge of larger-scale solution
patterns. It's striking, then, that this research perspective on how
programing design work gets done---for example, the detailed
ethnomethodological analysis by Rooksby and colleagues on how engineers
use a whiteboard (Rooksby \& Ikeya, 2012; Rooksby, 2010)---seems largely
underrepresented in studies of how students learn to program. Instead,
the latest generation of research on student learning in programming has
focused much more extensively on questions like \emph{how can we assess
and mitigate students' difficulty in programming?} than it has on
questions like \emph{how do students learn and display evidence of
design thinking in programming?}

\subsubsection{The missing attention to students' design
thinking}\label{the-missing-attention-to-students-design-thinking}

It's difficult to point directly to the absence of a research direction
regarding how students design software. Rather, one can instead show
that most computing education research focuses elsewhere. In 2004,
Valentine conducted a meta-analysis of 20 years of conference papers at
SIGCSE\footnote{SIGCSE, as used here, is the abbreviation for the annual
  meeting of the Association for Computing Machinery (ACM) Special
  Interest Group for Computer Science Education. SIGCSE is one of the
  premier venues for computer science education.}. His analysis focused
on papers focused on first-year university courses, which as of 2004
comprised about 1/3 of the total papers presented at SIGCSE. Valentine's
resulting taxonomy reveals how---in a conference about computer science
education---there is at best a minimal focus on how students engage in
design thinking. In the decade from 1994-2003:

\begin{itemize}
\item
  About 24\% of conference papers were ``Marco Polo'' studies, which
  Valentine (2004) describes as teacher- or administrator-centered
  accounts of describing a new method, course, curriculum, or approach
  and documenting how well it works.
\item
  Another 42\% of papers were classified as either ``Nifty'' or
  ``Tools,'' and centered on either innovative assignments to give to
  students or software tools to augment learning, instruction, and
  assessment.
\item
  Only 22\% of studies were classified as ``Experimental,'' a broad
  category which includes everything from quasi-experimental comparisons
  of teaching methods to ethnographic interviews with students solving
  protocol problems.
\end{itemize}

Joy et al. (2009) offer a similar, more recent view of the state of
computing education research. Their 2009 survey included both conference
papers and journals, from which they constructed a taxonomy of over
3,500 papers in two overlapping fields: ``education in computer
science'' and ``computers for education'' (Joy et al., 2009, p. 112).
While Joy et al.'s (2009) classification scheme isn't identical to that
of Valentine's (2004), the most ready parallel to Valentine's
``Experimental'' studies are what Joy et al. call \emph{theoretical
pedagogy}:

The focus of the paper is principally educational, and reports results
grounded in education theory (i.e. explicitly references, discusses and
applies pure education theory). This category is used for ``Learning
Psychology'' where the ``learning'' is the predominant focus of theory,
such as Bruner and Vygotsky. (Joy et al., 2009, p. 114)

As a percentage of the pool of 3,500 papers, research classified as
``theoretical pedagogy'' totaled less than 5\%.\footnote{Joy et al.
  (2009) don't report raw quantitative numbers, so I estimated this
  percentage by measuring their data graphic on p. 117.} Admittedly, the
authors don't disaggregate their final results by ``education in
computer science'' vs. ``computers for education.'' Still, it seems
reasonable to assume that if even half the publications were from
``computers for education'' and none of that half focused on theoretical
pedagogy, that still means less than 10\% of publications \emph{in}
education in computer science foreground learning as a primary
phenomenon.

If one is looking for the kinds of studies that would involve deep,
fine-grained analysis of learning via students in action, the pool gets
even thinner. In a survey of 79 papers from the International Computing
Education Research Conference (ICER)---published between 2003 and
2009---Malmi et al. (2010) found that of the 79\% of papers that
included any mention of a theoretical framework, 39\% were ``surveys.''
(p. 8). ``Experimental'' papers account for another 15\% of papers that
explicitly identified a framework. The total number of papers that
employed case study methodology, ethnography, phenomenology, \emph{or}
phenomenology was 10, or approximately 14\% (Malmi et al., 2010, p. 8).
But, since Malmi et al. (2010) do not identify those case
study/ethnography/phenomenology/phenomenograpy papers by name, it's
difficult to assess just how closely those papers hew to their nominal
frameworks.

Indeed, one of the most useful collections of research on the psychology
of how people design software isn't recent; it's almost thirty years
old. In the 1980s, Elliot Soloway, James Spohrer, and others were trying
to establish what expert programming behavior looked like (Adelson \&
Soloway, 1985; Soloway, 1986), and what kept novice programmers from
developing expertise (Bonar \& Soloway, 1983, 1985; Mayer, 1981; Pea,
Soloway, \& Spohrer, 1987; Soloway \& Spohrer, 1989). One of the most
fundamental arguments to emerge from Soloway and Spohrer's work was a
refutation of the claim that novices make programming errors because
they don't understand the elements of a programming language:

Our empirical study leads us to argue that (1) yes, a few bug types
account for a large percentage of program bugs, and (2) no,
misconceptions about language constructs do not seem to be as widespread
or as troublesome as is generally believed. Rather, many bugs arise as a
result of plan composition problems---difficulties in putting the
``pieces'' of a program together (see sidebar)---and not as a result of
construct-based problems, which are misconceptions about language
constructs. (Spohrer \& Soloway, 1986, p. 401)

Ultimately, Spohrer and Soloway (1986) found strong support for two
kinds of problems that transcend any particular programming language:

Interpretation Problem: When novices read a programming assignment, they
do not always infer the correct interpretation from the specifications.

Composition Problem: Novices may not detect negative interactions
between sections of code that are locally correct, but globally
incorrect. For example, the code to perform the output may be correct,
but in the wrong place in the program. (Spohrer \& Soloway, 1986, p.
191)

Despite Spohrer and Soloway's (1986) finding that program composition
accounts explain more than do language construct accounts of student
error, research since has persisted in documenting students'
``misconceptions'' about programming language constructs (Fleury, 1991,
1993, 2000; Herman, Kaczmarczyk, Loui, \& Zilles, 2008; Kaczmarczyk,
Petrick, East, \& Herman, 2010). Moreover, in a 2006 paper analyzing the
programming behavior of hundreds of novices, Jadud argues that a focus
on syntax-level analyses can help improve research and practice:
``{[}b{]}y identifying and understanding the behaviour of novices
learning to program, we hope to build up to later making sound cognitive
and constructivist inquiries and recommendations'' (Jadud, 2006, p. 80).

In the next section, I argue that a growing syntax- and language-focused
trend in computing education research is using bigger and richer
datasets to push us farther away from the kind of fine-grained studies
that help us understand design thinking in software creation. First, I
explain how code-snapshotting systems help us capture changes to
students' code over time. Then, I outline why---in their current
usage---such systems fail to help us consider larger design thinking
issues that might be at play for students. I conclude my introduction
with a claim that we can use the same code-snapshot repositories we've
been mining for syntactical analyses to look for deeper explanations of
why students struggle with software design.

\subsubsection{The current state of research on students' code
snapshots}\label{the-current-state-of-research-on-students-code-snapshots}

A growing trend in computer science education research is the collection
and analysis of code snapshot data---records of the state of and changes
to students' code as they develop it (Jadud, 2006; Rodrigo, Tabanao,
Lahoz, \& Jadud, 2009; Spacco, Pugh, Ayewah, \& Hovemeyer, 2006). Though
specific implementations differ, the general strategy in such projects
is that a student event (typically compiling code or saving a file)
triggers a procedure that creates a record containing the entire content
of all of a student's relevant files, as well as associated metadata
(time of save/compilation, for example, and any compiler errors that may
have been generated). Mining data from such snapshotting systems has led
to large-scale documentation of common student errors (Spacco, Pugh, et
al., 2006), the development of compile-time detectors to catch common
student errors (Spacco, Strecker, Hovemeyer, \& Pugh, 2005), and the
proposal of formative assessment models to predict student success
(Jadud, 2006; Tabanao, Rodrigo, \& Jadud, 2011). Building on existing
momentum, some researchers are actively pushing for a continued scale-up
of how we collect code snapshot data. One current proposal even calls
for creating an international database by collecting snapshot data from
thousands of introductory programming students worldwide (Kölling \&
Utting, 2012).

What's common to these threads of research is how they mine the data. In
most applications of code snapshot\footnote{What I call ``code
  snapshot'' research may be more formally called ``online protocol''
  research (Jadud, 2006; Rodrigo, Tabanao, Lahoz, \& Jadud, 2009;
  Tabanao, Rodrigo, \& Jadud, 2011). ``Online'' here isn't mean to mean
  the sense of globally-connected computers, but rather that student
  materials are collected in a minimally-intrusive way \emph{while
  students work}.} research, the aim is to average across events and
sessions to arrive at a characteristic measure of a student's behavior.
Jadud (2006), for example, develops the idea of a student's ``Error
Quotient,'' a measure of how effectively a student addresses
compile-time errors in their code.\footnote{A useful way of thinking
  about Jadud's (2006) error quotient is that the computation believes
  in second chances. It doesn't penalize students for making errors; it
  penalizes students if they don't fix a given error before compiling
  the same piece of code \emph{again.}} Rodrigo and colleagues extend
the idea of error quotient (Rodrigo et al., 2009) and introduce similar
measures such as students' ``compilation profiles'' (Rodrigo et al.,
2009), ``frustration profiles'' (Rodrigo \& Baker, 2009), and ``error
profiles'' (Tabanao et al., 2011). Because these measures can be
collected in real-time as students code, they offer the potential to
identify at-risk students (Tabanao et al., 2011) and respond with early
interventions (Jadud, 2006).

But, for all the data these methods collect their focus is primarily
what Jadud (2006) calls students' ``syntactic'' struggles: the challenge
of articulating well-formed statements a compiler can properly parse.
This focus on syntax-level struggles leads to both a methodological and
theoretical trade-off: we can see in detail how students struggle with
wording, but we see less of how they struggle with the meaning and
intent of their code. Syntactical analyses necessarily ignore the
specific \emph{content} of students' code because they abstract
meta-information about the event: error type, error location, frequency
of error.

As an analogy, suppose that instead of studying computer science
students we were studying screenwriting students. Further suppose we
have a system to track data whenever students save their screenplays and
scripts. For each save, we get a copy of the entire script at that time.
We can also run an automated analysis to check whether they've properly
formatted slug lines\footnote{In a screenplay, a slug line gives
  information about where and when a scene takes place. For example:
  ``INT. FORTRESS OF SOLITUDE -- DAY.''}, put character names in
all-capital letters, properly numbered scenes, etc. We might even know
where students are when they write (coffee shop, library, home, etc.).
If we mine that data, we could learn something about students'
\emph{screenwriting behavior}, but only at the level of how they
struggle with screenwriting syntax. Using only error-based data it's
much harder to answer questions such as:

\begin{itemize}
\item
  How does this student construct a scene or handle dialog in their
  script?
\item
  Does the student seem to have a grasp of pacing, story beats, and
  efficient exposition?
\item
  How does their writing manage and develop character arcs?
\end{itemize}

We also still can't answer fundamental questions about the context of
students' writing processes:

\begin{itemize}
\item
  Do they use particular techniques to ``break story'' and decompose a
  narrative into its key beats (index cards? Whiteboard?)
\item
  Do they participate in a writing group?
\item
  How do they respond to people giving them notes on revising a script?
\end{itemize}

To sum up, because snapshot systems are designed to collect the totality
of a codebase at frequent intervals, they offer a rich record of data
that captures the results---both major and minor---of students' design
decisions. But, computing education research that \emph{uses}
code-snapshotting has focused much more on detecting, classifying, and
predicting student errors than it has on showing how students progress
in programming and design expertise. Nevertheless, snapshot-based
research shows tremendous promise. Given that :

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Snapshotting students' code represents a cutting-edge way to resolve
  the way code---as a design artifact---evolves over time, \emph{but}
\item
  Code snapshots, as they're currently used, explore neither the
  totality of a student's design nor the rich context of that design's
  production, \emph{and}
\item
  There is a lack of parity between studies of how professionals design
  software and studies of how students do so,
\end{enumerate}

It seems sensible to ask: \emph{can code snapshots be used---possibly in
synthesis with ethnographically-oriented methods---to start studying how
students' design thinking plays a role in their introductory programming
work?} We believe the answer is yes.

In what follows, we present work that proceeds from an empirical
challenge: how can we develop accounts of students' programming activity
that explain the form and evolution of their code on a design project?
We offer an account of one student---Rebecca---and her experiences and
code from a second-semester course on programming concepts for
engineers. Using data from both code snapshots and clinical interviews,
we explicate both the challenges of studying students' software design
processes and the potential for such study to inform accounts of
teaching and learning.

\subsection{Context and Methods}\label{context-and-methods}

\subsubsection{Context of the study}\label{context-of-the-study}

The data presented in this case are taken from an ongoing IRB-approved
study of undergraduate electrical engineering majors undertaken at
\emph{Flagship State}, a large, public research institution on the east
coast. For two semesters, I have followed a total of 10 students taking
``Intermediate Programming Concepts for Engineers.'' It is the second of
a required two-semester course sequence in programming.\footnote{Hereafter,
  I use ``Intermediate Programming'' to refer to the second course and
  ``Basic Programming'' to refer to the first.} Students can (and some
do) place out of Basic Programming via AP Computer Science credit, but
all Electrical and Computer Engineering students must take Intermediate
Programming.

Intermediate Programming has two 75-minute lectures per week and a
weekly discussion section led by an undergraduate Teaching Assistant
(TA). Typical enrollment is between 60 and 80 students per semester.
Like Basic Programming, Intermediate Programming is taught using the C
programming language, and it incorporates multi-week projects in C as
part of its assessment structure. Students must work individually on
four projects over the course of the semester, which together comprise
45\% of their final course grade.

Grading projects involves running students' compiled code against
automated tests that determine whether a student program's output
matches the instructor's canonical output. If a student's program
completely matches the canonical output, the student receives at least a
90\% grade on a project. The remaining 10\% are discretionarily
allocated ``style'' points, awarded for things like proper formatting,
code commenting, and functional decomposition (Field Notes).

This study centers on ``Flights Database,'' the second of four projects
assigned to Intermediate Programming students during the spring 2012
semester. Students were asked to build a text menu-based program that
would let users query information about airports and plan non-stop and
one-stop flights between airports. For this project, the instructor gave
students three separate text files as source material. The
\textbf{airports} file contained names of airports and their
three-letter abbreviation codes; the \textbf{routes} file contained
3-tuples of two airport codes and the route number of a flight flying
between them; the \textbf{flights} file contained a list of specific
flight information (including arrival and departure times) by route
number. Crucially, in order to be able to respond to user queries
students would need to build a program that could coordinate information
across all three files to return an answer.

My analysis details the work of Rebecca, a female first-year electrical
engineering major. I focus specifically on Rebecca's code for finding
``one-stop'' flights, which the instructor defined as ``all pairs of
ﬂights that route the user between the departure and arrival airports
with exactly 1 stop (i.e., a one-connection ﬂight)'' (Flights Database
handout, 2012). The one-stop problem is particularly challenging. To
solve it successfully students' code must accept a user's choice of
airports and day, then stitch together routes that involve two separate
flights in a way that passes stringent constraints for acceptable
layover times.

\subsubsection{Methods}\label{methods}

This study proceeds from an empirical challenge: how can we develop
accounts of students' programming activity that explain the form and
evolution of their code on a design project? The focused form of that
challenge for this study is ``how can we understand the unconventional
design choices embedded in Rebecca's one-stop flight code?'' To answer
that question, my study draws from three data streams: ethnographic
observation, clinical interviewing, and code snapshot analysis.

For two semesters, I ethnographically embedded myself in the same
instructor's section of Intermediate Programming. My aim throughout was
to see what students see in terms of course material, assignment
directives, and instruction. In fall 2011 I observed approximately 50\%
of the course lectures. I also independently completed all class
homeworks and three out of the four course projects to more fully
understand the course's assessments. In spring 2012 I continued
attending lectures, though less frequently, and began attending select
TA-led discussion sections. During both lectures and discussion sections
I took field notes while recording ambient audio using a LiveScribe
Pulse pen.

Rebecca was one of four students (three female, one male) willing and
able to participate in a series of 1-hour outside-of-class clinical
interviews during the spring 2012 semester.\footnote{In total, roughly
  30 students responded to my initial in-class solicitation to be
  contacted by email about my study. Of the students I emailed,
  approximately 8 students responded to my emails to schedule interview
  times. Of those 8, only four students were able to successfully find
  interview time slots that fit our respective schedules.} I interviewed
Rebecca five times, and in typical interviews I split time between
asking about her experiences in the course and giving her time in the
interview to work on her project code. During each interview, I
simultaneously used:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  A Kodak Zi8 camera for video-recording our interactions
\item
  A LiveScribe Pulse pen to capture Rebecca's on-paper penstrokes
\item
  A MacBook Pro (early 2011) with screen-recording software to capture
  everything on-screen while Rebecca programmed
\end{enumerate}

The final component of data gathering is modeled after Jadud's (2006)
system for capturing students' code. My colleagues and I developed
software, built around the open-source version control system called
Git, that effectively creates an entire copy of a student's code---what
we call ``snapshots''---every time students invoke the compiler on their
code. Our software then sends those snapshots to a secure,
researcher-accessible server in real-time as they're created.
Consequently, I could plan each interview with Rebecca around
up-to-the-minute knowledge of her work---in some cases work she had
completed just hours before the interview---and tailor my interview
questions to emerging patterns in her code. In total, Rebecca's work
resulted in 958 compilation snapshots over the course of the semester.

In what follows, I recount snapshot and interview data to explore the
form and history of certain features of Rebecca's one-stop flight code.
First, I simulate an analysis of Rebecca's code in the methodological
vein of Jadud (2006) and others (Rodrigo, Tabanao, et al., 2009; Tabanao
et al., 2011), analyzing only Rebecca's code snapshots. In this first
analysis, I focus on the form and evolution of particular design
features of Rebecca's code. Then, I present a second, complementary
analysis using contextual data from my clinical interviews with Rebecca.
In my second analysis I go beyond the snapshots to highlight why, in
Rebecca's own words, she made those particular design choices.

\subsection{Code Snapshot Data and
Analysis}\label{code-snapshot-data-and-analysis}

In this section, I simulate what data on Rebecca would look like using
an existing method of analysis described in Jadud (2006). Namely, I work
at a level of remove from Rebecca herself. I base my inferences only on
her actual code, limiting my interpretations about what she may have
\emph{intended} to do or \emph{designed} her code to do only to what's
supported in the snapshot-by-snapshot record of her code. Through this
``snapshot-only'' analysis, I show that we can see in great detail how
Rebecca's code embodies particular design features that diverge from how
an expert might have handled this design task. I restrict my analysis
here to just two such features of Rebecca's code:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Rebecca's use of repeated file-scanning loops to create a depth-first
  search algorithm
\item
  Rebecca's 7-fold duplication of code to handle checking for flights on
  each day of the week
\end{enumerate}

\subsubsection{Rebecca's file-scanning solution is computationally
complex}\label{rebeccas-file-scanning-solution-is-computationally-complex}

After declaring variables and opening the three provided text files
(flights, routes, and airports), Rebecca's one-stop flight code enters a
series of conditionally-nested `fscanf()` commands. The purpose of
`fscanf()` is to scan through the characters of a file, most often one
line at a time. The programmer can specify patterns of text to look for,
e.g., ``in each line, look for a word, followed by a space, followed by
two digits.'' `fscanf()` also gives the programmer the flexibility to
store such matched patterns. ``Once you find a word followed by two
digits, store the word in the following location in memory.'' What's
interesting isn't \emph{that} Rebecca used `fscanf()` to read through
files like the list of airports. Any suitable solution would need to
read through those files. Rather, what's interesting is \emph{how} she
uses `fscanf()` in her design.

Rebecca's file-scanning logic never persistently stores the contents of
the files it reads in. Rather, her program reads through files one line
at a time, and it essentially can't process or act on airport/flight
information not in the line currently being scanned. Instead, it's been
designed to copy single patterns temporarily, then rewind the file back
to the top and s\protect\hypertarget{GoBack}{}{}tart reading in
one-line-at-a-time again.

What's consequential about Rebecca's design choice? Computationally, her
code has to repeatedly open multiple files and scan them one line at a
time in order to coordinate information. So, the task of finding a
one-stop flight between two cities becomes a series of repeated,
one-line-at-a-time scans of external files:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Scan the \textbf{airports} file one line at a time to check the
  correctness of the user's input (line 20)
\item
  Scan the file of pairwise airport \textbf{routes} one line at a time
  (line 24)
\item
  To find possible connection cities, scan the \textbf{routes} file one
  line at a time again (line 38)
\item
  If a route matches, scan the \textbf{flights} file one line at a time
  to verify whether the time/day constraints are acceptable (one of the
  following lines depending on the chosen day: 52, 79, 106, 134, 162,
  190, 218).
\end{enumerate}

Rebecca's code is both visually and computationally complex. The
multiply-nested blocks can make it difficult for a human reader to
follow the code flow, which may have made it challenging for Rebecca to
debug her own work. Moreover, nested for- and while-loops increase the
complexity space of a program---what computer scientists call the big-O
characterization of her algorithm. For every nesting of a scan loop
(there are 4 here) Rebecca increases by 1 the degree of a polynomial
that represents the execution time of her program. So, from a
performance perspective, Rebecca's design suffers a trade-off in that
with each invocation of a scanning loop, we see a geometric increase in
the time complexity of her program.\footnote{} In fairness to Rebecca,
the point I make here is not about her but about our mode of analysis.
From the perspective of a typical software engineer, Rebecca made a
decision involving a design trade-off. Still, inferring only from the
code we cannot know whether her decision was deliberate. We cannot know
from code whether she knew or understood the kind of complexity and
performance trade-off she made, nor can we know how she felt about the
consequences of the decision. We know only that her final submitted code
contained the results of her decision to use nested scanning loops.

\subsubsection{Rebecca's file-scanning solution ignores an assignment
directive
}\label{rebeccas-file-scanning-solution-ignores-an-assignment-directive}

A second feature or Rebecca's `fscanf()` design is that she ignores
instructions in the course assignment, which produces a solution that
uses abstraction in a non-obvious way. One design solution to the
problem of creating a relational database---and one seemingly dictated
by the assignment---is to create three separate arrays in the computer's
memory:

To parse the 3 airline ﬂight database ﬁles, you will need to declare
arrays that will receive all the data. For the purposes of determining
array sizes, you may assume there will never be more than 100 airports
in the ``airports.txt'' ﬁle, 500 route IDs in the ``routes.txt'' ﬁle,
and 3000 ﬂights in the ``ﬂights.txt'' ﬁle. (Flights Database class
assignment, 2012)

Presumably, from the instructor's directive, one ``will need'' to have
an array of airports (mapping 3-letter code to full airport name), an
array of routes (mapping a pair of airports to a unique routing number),
and an array of flights (mapping unique flight numbers to a collection
of information about that flight).

Rebecca creates no arrays. Instead, her code attempts to accomplish the
same task that a memory-persistent data structure would, only without
the persistence. A consequence of Rebecca's approach is that she has no
easy way to refer to arbitrary airports, routes, or flights in her code,
since her program has no mechanism to store such information
persistently. A second consequence is that since she avoids persistent
data structures, the complex work her program does to read through each
line of each file, in some cases multiple times (above) is repeated
every single time a user initiates a query.

Given Rebecca's particular design pattern, we asked the question of
whether she may have tried creating arrays before ultimately settling on
her scanning-loop solution. The answer, as far as we can tell, is no. We
analyzed the history of both Rebecca's main() method and her one-stop
flight code module. Our search revealed that no snapshots exist in which
Rebecca created arrays---either through dynamically allocating them
(through heap memory), or, as the assignment recommended, creating
overprovisioned fixed-size arrays on the stack. In other words, at the
limit of resolution of our data collection, and within the scope of the
code Rebecca typed, she never tried an array solution.\footnote{If
  Rebecca had tried an array solution and compiled---whether error-free
  or not---our automated snapshot collection system would have captured
  it.}

Curiously, we have evidence \emph{outside} of Rebecca's code that
suggests she knew, and even perhaps had seen, an array-based design
solution. In a file called ``notes.txt'' contained in her project
directory, we see the following lines:

think about using: sscanf, array of pointers

his header file!!!

-max line lenght: 2000

-max string lenght: 100

-defined true and false

-max airports: 100

-max routes: 500

-max flights: 3000

-min connect time: 60.0

-max connect time: 120

-daily maxk: 254 ???

-char airports{[}max airports{]}{[}4{]}

-char aiport\_cities{[}max airports{]}{[}max string lenght{]}

-he has 3D array for routes.... char routes{[}max
routes{]}{[}2{]}{[}4{]}

(notes.txt file, created March 19, 2012)

The context of the file is not entirely apparent, because we did not
observe lecture on March 19, the day the notes.txt file entered
Rebecca's snapshot history. Also, whether ``his'' refers to the
instructor or perhaps another classmate is unclear. ---What seems clear,
however, is that Rebecca was responding to items she had seen in someone
else's header file. Consequently, putting together the notes.txt file
with Rebecca's final code submission reveals Rebecca was exposed to a
design solution involving arrays, but never implemented it in her code.
Thus, a lingering and consequential question remains unanswered: why did
Rebecca adopt a solution that defied the directions of the assignment,
especially when she'd seen part of a potential design solution that did
use arrays?

We return to this question in a later section, but first we turn our
attention to another unusual feature of Rebecca's work: seven-fold
repetition of code.

\subsubsection{Rebecca repeats the same chunk of code seven
times}\label{rebecca-repeats-the-same-chunk-of-code-seven-times}

A second key feature of Rebecca's code is the almost identical
repetition of a single 23-line code chunk seven times (lines 50--240).
Because users can run queries by choosing a day to fly (and some flights
only run on certain days), students' code must be able to handle each of
the seven possible days for when a user would want to fly. In principle,
Rebecca's code achieves just that.\footnote{I say ``in principle''
  because Rebecca's code would not compile on my machine. So, in
  practice, her design contains compile-time errors (and possibly
  run-time errors). Nevertheless, her code provides ample evidence that
  she was attempting conditional logic to handle each possible day.} In
practice, her code creates seven different conditional branches---one
branch for each day of the week---where the code within each branch is
duplicated.

Figure 2-1 represents a side-by-side delta-comparison of two such
day-specific branches of code. Lines 158--184 of Rebecca's original code
are on the left; lines 214--241 are on
\includegraphics[width=5.99722in,height=3.89722in]{media/image3.png}the
right, and in the figure lines have been renumbered (from 1) to ease
comparison. In this delta view, lines that differ are highlighted in
pale red, and characters that differ are shown in bold red.

The two code blocks demonstrate just how much code is duplicated for
handling user input based on days of the week. Between these two chunks
there are only three differences (lines 1, 3, and 11): all of the
references to day are changed from 5 (on the left) to 7 (on the
right).\footnote{In the text-input files students were given, days of
  the week were represented as integers (rather than the perhaps more
  familiar ``Tuesday,'' ``Wednesday,'' etc.).} Moreover, the changes
from block to block are patternistic and predictable: the first line of
the block is a non-functioning comment, the third line of each block
just checks whether the rest of the block should run, while the eleventh
line of each block compares an array entry to the day of interest.
Everything else is duplicate boilerplate that is essentially repeated 7
times; once for each day of the week. I say ``essentially repeated''
because, as we'll now explore, there are minute differences between some
of the code chunk's seven incarnations.

Repeating code as Rebecca has done can be problematic because each
repetition multiplies the number of places she has to examine and modify
if she wants to introduce a systematic change. If, for example, Rebecca
wanted to change the internal names she gives to scanned-in variables,
she has to make that change in seven different blocks of code: once for
each of the seven days of the week she's hard-coded. And, since any
given change may inadvertently introduce an error, increasing the number
of places she repeats code also makes the code that much more vulnerable
to inconsistently-applied changes.

Indeed, a repeated, inconsistently-applied scan pattern change seems to
be exactly what occurred in Rebecca's code history. Between 10:04pm and
10:37pm on March 26, Rebecca introduced a large set of changes to the
one-stop flight module. Among those changes Rebecca added the
`d\_letter` file-scanning-parameter to what would become line 218, but
not to what would become line 190. We can reasonably infer Rebecca added
this parameter as a way of capturing the ``am'' or ``pm'' specifier
given the input file's format. Moreover, we can verify through Git that
once introduced, Rebecca's omission of the parameter was never modified
or corrected. The problem percolated through to her final submitted
code.

\subsection{Interview Data and
Analysis}\label{interview-data-and-analysis}

In the previous section, I described two unusual features of Rebecca's
code for searching one-stop flights:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Her use of multiply nested loops that scan through source information
  files \emph{without} storing the information in those files
  persistently in long-term memory

  The code for handling a user's chosen day, which was essentially the
  same block of code copied and pasted 7 times
\end{enumerate}

In this section, I offer explanations of Rebecca's design choices by
interpreting data from over five hours of clinical interviews I
conducted with her. I draw from those interviews to explain how design
decisions that might seem unusual to an expert in fact grew rather
unproblematically (for Rebecca) as ways of deliberately transferring
prior knowledge and designs (which explains feature 1) or coping with a
constraint to produce a reliable solution she could trust (which
explains feature 2).

\subsubsection{Rebecca employed fscanf loops because she was
deliberately reusing from an Basic Programming
Assignment.}\label{rebecca-employed-fscanf-loops-because-she-was-deliberately-reusing-from-an-basic-programming-assignment.}

Rebecca's choices become easier to understand when you consider what she
said in the interview. When Rebecca initially saw the flights database
project, her first reaction was ``this is just a lot like our fantasy
football project that we did last year'' (Interview, March 16,
2012).\footnote{Transcript conventions are shown in Appendix 1.} But,
the previous fantasy football project was easier because all of the
information she needed was in one file. Flights database, by comparison,
fractured necessary information across multiple input files.

Since Rebecca said the fantasy football project was like an easier
version of flights database, I asked whether she tried making this
project more like the easier one. Her response was an emphatic ``Oh
yeah! Definitely!'' (Interview, March 16, 2012). She said ``as soon as
we got this project I was like, ah! fantasy football! I'm just gonna go
and see how much code I can rework from that and like, use /mmhmm/ in
this project'' (Interview, March 16, 2012). Specifically, Rebecca went
back to her fantasy football code ``and I was looking at how I scanned
in the information from the files, cuz, we haven't really done anything
like that this semester. Uh, scanning in from files before---''
(Interview, March 16, 2012). Since the fantasy football draft project
was the last in which she'd needed to do file-scanning, she ``just, uh,
went back to check how I did that /mmhmm/ and then, if I could I copied,
but because a lot of the variables were different, uh, like these were
more var---less, less variables, and more strings than last year /mmhmm/
uh, I just retyped it out. I just looked at how it was similar''
(Interview, March 16, 2012).

My first opportunity to discuss Rebecca's one-stop flight work was on
March 16, 2012, in what would be her third of five interviews that
semester. This interview was conducted very early into the time window
for the Flight Database project, before Rebecca had done the bulk of her
coding. We were discussing her prospective design plans. As Rebecca
began explaining how the logic for a one-stop flight search was supposed
to work, she described what she saw as one of the central difficulties
of the project: the relevant information for answering user queries was
spread across multiple files (Interview, March 16, 2012).

As Rebecca explained, something as simple as finding a flight from, say,
JFK to BWI ``involves scanning through multiple files, because it's not
like one file that has everything conveniently like, there'' (Interview,
March 16, 2012). When I asked what would make things easier if,
hypothetically, all the information she needed were in one file, Rebecca
responded by appealing to a previous assignment from last semester. In
Rebecca's first-semester programming course (Basic Programming for
Engineers), one of several multi-week projects had students create a
system for users to conduct a fantasy football draft. The project
involved, among other things, topics related to basic file management in
the C language, including how to read in a file from disk (Interview,
March 16, 2012). Rebecca explained:

Rebecca: Um, like, cuz when we first got this project, uh, I actually
was thinking ``oh, well this is just a lot like our fantasy football
project we did last year.'' /Huh/ We uh, had to scan in, uh, someone had
to enter like ``I wanna pick a quarterback,'' so then you had to scan in
and go and look for all the quarterbacks in the file and say ``OK, this
is the quarterback'' and everything. But, in that project we only had
the one file that had everyone listed: quarterbacks, runningbacks, wide
receiver, in one file. And all the information you needed there /Mmhmm/
So, you could just get it all and compare it all at once with one scanf
/mmm/ whereas this you have to, take, uh, you scan in the flights, ah,
the flights file. So, then you find the flight number. You have to save
the ID from that flight number, use that ID to scan into the routes file
/mmhmm/ and then save the routes information and then print it out with
the flights information. (Interview, March 16, 2012)

Rebecca's comments suggest she saw a coupling between the arrangement of
the input information and the structure (and complexity) of the
computational logic needed to process it. When one fantasy football file
contained all of the relevant information (player, position, team, etc.)
it could be read in and processed one line at a time. When information
was fractured across files (pairs of airport codes in one file, full
spell-outs of airport names in another, for example), Rebecca felt she'd
need to use information shared across files (such as a route ID) to
coordinate a scan across one file with a scan across another (and
possibly an additional scan across a third file) before she would have
all the necessary information and computations to return a result.

I was interested in the connections Rebecca saw across projects, so I
pressed on. When I asked whether she thought about trying to make this
project like her fantasy football project, her answer was an emphatic
``Oh yeah, definitely!'' As she elaborated:

Rebecca: That was like, as soon as we got this project I was like, ah!
fantasy football! I'm just gonna go and see how much code I can rework
from that and like, use /mmhmm/ in this project. And, my whole main
file, like all those NULL checks and everything, I mean they're really
simple to write, but I just copied `em and put `em there, cuz, we had
the same thing. /Mmhmm/ Um, just changed, like, the names of the files.

As she explained, ``reading in files'' was a topic covered extensively
in Basic Programming---the first course of the sequence---but they
hadn't talked much about it in this semester's course.

Rebecca: Because {[}reading in from files{]} was a big {[}Basic
Programming{]} topic we hadn't talked about it much. /Yup/ So, I just,
uh, went back to check how I did that /mmhmm/ and then, if I could I
copied, but because a lot of the variables were different, uh, like
these were more var---less, less variables, and more strings than last
year /mmhmm/ uh, I just retyped it out. I just looked at how it was
similar. (Interview, March 16, 2012)

I started doing the scanning, and then I realized like, ``I do not have
all the information I need in this one file,'' so that's when I realized
I would have to make the dummy variable to compare to the other file,
so, which, there was no other file in the, um, fantasy football one. So
that's when I realized, I was like OK, so, there's gonna be other stuff
that I'm gonna have to do, and /mmhmm/ I mean, I can still kinda compare
cuz it's still scanning files /yup/, but the idea of the dummy variable
and all that, that would not have come from fantasy football, so /sure/.

In sum, then, Rebecca's repeated, nested scan loops were a structure she
deliberately borrowed from a previous semester's project. By her
telling, what seemed obvious was that ``scanning in from files'' was a
topic she'd already covered, which meant she'd already developed a
workable solution for how to solve that problem. Thus, she saw the
problem of how to coordinate airport information from different files as
a new instance of the old problem of reading information in from one
file. Her flight database work, accordingly, tried opportunistically
adapting a previously working solution to fit the current circumstances.

\subsubsection{Rebecca repeated code because she wanted to re-use
functionality she could
trust}\label{rebecca-repeated-code-because-she-wanted-to-re-use-functionality-she-could-trust}

By our interview on April 6, Rebecca had already completed and submitted
her code for the flights database project. When I looked at the final
form of her code for finding one-stop flights I noted an unusual pattern
described in section 2.4.3 above: she had a code chunk repeated almost
character-for-character 7 times. In the interview, this section is what
Rebecca referred to as ``my obnoxiously long part of my code''
(Interview, April 6, 2012):

Rebecca: So, the way I did it was really long and probably, there was
probably like a much easier way, but I just did a giant if---if
statements \{swings cursor from line 50 to line 63\} If they wanted to
fly on Monday /OK/ I went through and checked to see if the route ID was
the same \{wiggles cursor across line 54\}, and if it did, I went
through to che---uh, I made a check\_days function \{wiggles cursor
across line 60\} uh, I ended up commenting that out cuz I didn't end up
/mmhmm/ finishing it. But, uh, my check\_days function worked, it just
didn't work completely with the code /OK/ (Interview, April 6, 2012)

As I scrolled the screen to look at each of the repeated blocks of code,
Rebecca elaborated:

Rebecca: And this is why my code, I feel like, is not uh, concise
enough, or, I don't really, I forget the word they use, but uh
/\{inaudible\}/ it's very long because I couldn't figure out if I should
do a while loop or whatever /Uh-huh/ But, so I was just like, I know
this way should work if I get everything else right, that uh, just go
through, if input's 1, if input's 2 and just do the same thing in each
of `em just /Mmmhmm/ check for, ``oh, if days is 2, if days is 1''
instead of, like---Cuz I probably could have done, like, maybe a giant
while loop, um, to try and, and if, while, inputs something, uh, then
you check to see whatever i is. But, I could, I didn't---couldn't figure
out how that would work, so I just did the same thing six times.

Interviewer: So, in, in each one of these it's like, looks, and I'm not
sure about this, but it looks like the way you wrote it---so this
\{highlights line 79\} is pretty much the same in all of them, right?
/Yes/

Interviewer: So's this one \{highlights line 81\} /Yes/ this one
\{highlights line 83\} Here's where it's different \{highlights line
85\}

Rebecca: Yes, because it just checks if it's a 2 instead of a 1.

Interviewer: OK. Um. /And then everything else is still the same/
Layover's still the same. OK.

Rebecca: Yeah. So that's why it's prob---it's not, uh, the neatest code
or whatever, because it's the same thing six times. (Interview, April 6,
2012)

Given Rebecca's assertions that her code wasn't neat, I asked what, if
anything she might change if she hypothetically had another week to work
on the project.

Rebecca: Um, first I'd try and get it to make sure it worked completely
/Ahh, OK, yeah/ this way, \{laughs\}, uh, and then, if I had the week
after that whatever, I'd probably go through and see if I could figure
out a way to make it concise-r because he likes uh, neat, as, like, code
that's, uh, easy for the user to see /uh-huh/ I guess. Uh, I forget
what, I keep forgetting what the word he used was at the beginning of
the year, but uh, just very concise and, uh, this is {[}a{]} very
expanded \{laughs\} way of coding, but, it made sense to me at the time
and I was just like ``I just want something that makes sense right
now.'' /Right/ So, that I can actually work with and have an idea.

Interviewer: Um, OK. So, so it would take you some extra thinking to
figure out /Mmmhmm/ how to break this down into /Yes/ smaller stuff
/smaller code/ Do you feel like you've had a lot of practice doing that,
or like?

Rebecca: Uh, a little. Like, but, a lot of times in {[}Basic
Programming{]} they didn't really mention too much about being concise.
They were just like ``if you can do it, do it'' \{laughs\} /OK/ So I
usually stuck to what made sense to me /Right/ uh, to turn the projects
in. (Interview, April 6, 2012)

In summary, Rebecca's ``expanded way of coding'' was a way of expressing
ideas in code that, in her own words, ``made sense'' to her. Moreover,
her Basic Programming course seemed, to her, to set expectations that
functionality comes first; ``neatness'' second. If she hypothetically
had more time to work on the project, her first priority would be to get
her existing code working. Consequently, Rebecca's repetition of code
can be understood as a kind of pragmatic solution to a difficult
problem: choosing which computational techniques were best for
accomplishing a complex goal. Moreover, her approach was shaped by the
fact that her Basic Programming course historically valued a philosophy
of ``if you can do it, do it'' (Interview, April 6, 2012). Ultimately,
those factors seem to be what led Rebecca to choose repeating code that
made sense to her over the difficult-to-envision alternative of a
``giant while loop.''

\subsection{Discussion}\label{discussion}

When we consider evidence from both code snapshot histories and clinical
interviews with Rebecca, we can draw several plausible conclusions about
Rebecca's patterns of software development on this project. Moreover,
those conclusions can spur larger considerations about theorizing
student learning and considering alternative instructional strategies
that may be more responsive to student needs. Below, I synthesize and
discuss patterns in Rebecca's development.

\subsubsection{Rebecca's key design decisions are made early, persist
through to her final submission, and carry
consequences}\label{rebeccas-key-design-decisions-are-made-early-persist-through-to-her-final-submission-and-carry-consequences}

On both her non-stop flight code and her one-stop flight code, key
structural features of how Rebecca's code works never meaningfully
change after Rebecca introduces them. In particular, Rebecca's use of
fscanf() patterns (as opposed to storing the data persistently in
memory) occurs extremely early in the work history of the project. The
fscanf() patterns initially appear in Rebecca's flight-checking code on
March 19, the third of what would be 286 chronological snapshots we have
comprising Rebecca's work on the project.\footnote{The project was
  distributed to students on March 5 and was due March 28, but the bulk
  of Rebecca's work on the project (284 of the 286 snapshots) occurs
  beginning March 19.
  https://github.com/TLPLEngineeringEdResearch/Rebecca/compare/e0a7caab4eb2d8401a0e50e8c6c6d8b514e9054e...0e8a5c606ef30f3b404024fdfad17d8d50e32456}
We note several consequences that arise \emph{because} these design
choices were made early and persisted to the final code.

First, Rebecca's \emph{early} choice not to persistently store data
likely forced her into the complicated scanning loop logic she developed
\emph{later} to solve the one-stop flight problem. A bit of explanation
may help here. Because the database students were building has multiple
uses, including uses that find flights for users, to be successful
Rebecca had to write functionality that found both non-stop and one-stop
flights for users. But, Rebecca initially designed each part of her
code---airport listings, flight listing, non-stop flight search, and
one-stop flight search---with its own internal fscanf() loops, as
opposed to having the four of them each refer out to a separate module
that could handle reading in data. The result is that none of her
individual modules share file-scanning code, and repeat data look-ups
necessarily have to be handled by creating another nested scan loop. A
design where code with a common function is pasted repeatedly into
separate modules contrasts strongly with one in which separate modules
all refer out to a single external piece of code.\footnote{As an
  analogy, imagine a supermarket where every unique type of item on the
  shelf has a cash register specific to that item next to it. To buy an
  item shoppers have to check-out individual items as they move through
  the store. It's a pain for shoppers because shoppers have to swipe
  their cards at dozens of separate item-specific registers across the
  store just to pay for a single grocery run. It's a pain for
  store-owners because now there are literally hundreds of individual
  registers to manage, maintain, and upgrade.}

We can understand Rebecca's development as involving a kind of design
inertia. Because she decided from the start to scan information using
fscanf() separately in each module, her later code was also beholden to
that commitment. The end result is that increasing the problem
complexity a small amount (the challenge of handling one-stop flights
versus non-stop flights) forces her to write much more code that is
itself more complex in its flow of control. And, when we consider her
seven-fold repeated code for checking flight availability on given days,
there are similar consequences. The decision---whether intentional or
not---to create seven different conditional paths forces Rebecca to
write and maintain much more code over time as the project evolves. As
we suggested in section 2.4.3, having a large number of locations to
keep in sync is likely what made Rebecca vulnerable to one of the errors
she ultimately introduced into her code.

\subsubsection{Her design decisions may be influenced by framing and
overzealous
transfer}\label{her-design-decisions-may-be-influenced-by-framing-and-overzealous-transfer}

Rebecca said in our interviews that she explicitly tried to reuse
solutions from the previous semester's fantasy football project. Our
code snapshots help us understand the extent to which that's true.
Rebecca's first compile for her flights database work begins with this
commented line of code:

//Rebecca Wells's Fantasy Football Team Maker: Project 1

Thus, our interview with Rebecca corroborates the snapshot evidence that
Rebecca directly copied code from a prior project. Moreover, Rebecca's
behavior of copying her old fantasy football code suggests several
implications.

First, to extend our point in section 2.6.1, Rebecca's ``design
inertia'' actually traces as far back as the decisions she made during
the first project of her first semester of programming. That is, by
copying code (and particularly file-scanning logic) from an old project,
she was incorporating core functionality that she designed when she
first learned to program. Admittedly, code reuse isn't in and of itself
a problem in software development. Parson and Saunders (2004), for
example, have written on cognitive heuristics that keep professional
software engineers from reusing code, even when reusing and extending
existing software artifacts is the best course of action on a software
project. So, the concern isn't that Rebecca reused code, but rather the
matter of what cognitive dynamics were at play that directed her choice
to reuse that code. In that regard, the constructs of framing and
transfer may help us better understand Rebecca's activity.

Framing, as it has been applied in contexts such as physics education
(Elby \& Hammer, 2010; Hammer et al., 2005; Scherr \& Hammer, 2009) and
mathematics education (van de Sande \& Greeno, 2012) concerns how
participants understand the social and intellectual activities in which
they're engaged. Of particular relevance to studying Rebecca is the
notion of \emph{epistemological framing}, which van de Sande and Greeno
(2012) summarize as

participants' understanding of kinds of knowledge that are relevant for
use in their activity and the kinds of knowledge, understanding, and
information they need to construct to succeed in their activity (e.g.,
what kind of information would count as a solution to the problem they
are working on). (van de Sande \& Greeno, 2012, p. 2)

Rebecca's decision to copy code from fantasy football implicitly
reflects her orientation toward what kinds of knowledge (the course
topic of scanning information from files) are relevant to solving the
problem. More broadly, note that in Rebecca's interview she explains
that her primary objective is to get a solution that works, which stands
in contrast to having a design priority like having a solution that is
elegant, or one that transparently manages complexity. That commitment
again reflects a manifestation of epistemological framing as ``what
would count as a solution,'' where for Rebecca what counts is a solution
that works.

Additionally, Rebecca saw the flight database project as a new instance
of a prior problem: fantasy football. Consequently, she consciously
adapted past solution patterns, because the flights database project
looked like it contained problems she had already solved in previous
code. Rebecca's deliberate reuse of old code can be readily understood
as an example of what Schwartz, Chase, and Bransford (2012) call
``overzealous transfer'':

Of particular concern are situations where students transfer skills,
knowledge, and routines that are effective for the task at hand but may
nevertheless be sub-optimal in the long run because they block
additional learning. We will call this \emph{overzealous transfer}
(OZT)---people transfer solutions that appear to be positive because
they are working well enough, but they are nevertheless negative with
respect to learning what is new. (Schwartz et al., 2012, p. 206)

In short, when students overzealously transfer prior knowledge as
Rebecca did, ``they may believe they are doing the right thing, and
without appropriate feedback they cannot know otherwise'' (Schwartz et
al., 2012, p. 206).

In Rebecca's case, the constructs of framing and overzealous transfer
together let us describe why she would have repurposed a solution that
she felt was adequate, even to the exclusion of the topics being taught
in class and the explicit directions in the project brief. By framing
the flights database problem as a new instance of an old problem,
Rebecca treated it as she did the old problem. But, she arguably
transferred \emph{too} \emph{much} of the old code's structure; so much
that she had to introduce even more complexity into her code just to
make the transferred parts work properly under the new constraints. And,
because her framing of the task seemed to privilege a philosophy of ``if
you can do it, do it'' (Interview, April 6, 2012), her primary goals
were to get her program to work, by whatever means she could understand
and trust.

\subsection{Conclusion }\label{conclusion}

In closing, even with a complete snapshot history and over five hours of
clinical interviews, it's still an enormous challenge to thoroughly
understand one student's design trajectory on a project. The challenge
is both methodological and theoretical. Through interviews we know what
Rebecca was thinking retrospectively, but we still work at a remove from
understanding what she was thinking in the precise moments she wrote her
code. Conversely, our snapshots provide a detailed record of the state
of her code every time she compiled. But, our snapshots are limited by
the kinds of materials they can track and the frequency with which
they're taken. We don't know, for instance, what Rebecca might have been
talking to others about while sitting in a dorm lounge working on her
project. And, because Rebecca wouldn't always compile frequently, some
snapshots detail large changes to the code made over long periods of
time---hours or days---which can limit our ability to precisely
determine what happened, and when.

From a theoretical perspective, we find our attempts to understand her
design decisions involve appeals to constructs of a finer granularity
and dynamism than are typically considered when studying first-year
programming courses. Understanding why fscanf() loops dominate Rebecca's
flight search code, for example, pushes us to consider the effects of
how Rebecca framed the task and the subsequent overzealous transfer that
resulted. And, understanding why code gets repeated seven times pushes
us to consider how emotion couples with sense-making in students' design
decisions. In Rebecca's case, a time-pressured preference for code that
``makes sense right now'' translated to a design of explicitly repeating
procedures instead of elegantly abstracting them.

\section{\texorpdfstring{\protect\hypertarget{ux5fToc247188552}{}{\protect\hypertarget{ux5fToc252445913}{}{}}Study
2 -- What directs and sustains students' in-the-moment programming
activity?}{Study 2 -- What directs and sustains students' in-the-moment programming activity?}}\label{study-2-what-directs-and-sustains-students-in-the-moment-programming-activity}

\subsection{Introduction}\label{introduction-2}

Although imaginative and carefully designed, the Harvard course taught
that there is only one right way to approach the computer, a way that
emphasizes control through structure and planning. There are many
virtues to this computational approach (it certainly makes sense when
dividing the labor on a large programming project), but Lisa and Robin
have intellectual styles at war with it. Lisa says she has ``turned
herself into a different kind of person'' in order to perform, and Robin
says she has learned to ``fake it.'' Although both women are able to get
good grades in their programming course, they represent casualties of
this war. Both deny who they are in order to succeed. (Turkle \& Papert,
1991, p. 165)

Over 20 years ago Sherry Turkle and Seymour Papert called for
epistemological pluralism in computing education. Their argument --- in
my view ---had multiple parts:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Programming a computer involved a shift in how one thought about the
  nature knowledge from a propositional \emph{what is} to an imperative
  \emph{how to} (Abelson \& Sussman, 1996; Papert, 1980).
\item
  Historically, there was a tendency to treat computers and the act of
  programming them as an extension of the same Western logico-deductive
  knowledge tradition that fueled science.
\item
  Recent scholarship, often at the intersection of post-structuralism
  and social studies of science, had been troubling the idea that
  knowledge-work in science proceeded solely by way of logical
  deductions (Keller, 1983; Latour, 1987; Traweek, 1988)
\item
  Data from professionals suggested programming, too, didn't proceed
  solely by way of the structured-planning approach emergent from
  Western logico-deductive knowledge traditions. There were many ways of
  knowing and constructing knowledge in programming.
\item
  A viewpoint of \emph{epistemological pluralism} --- embracing multiple
  kinds and ways of knowing --- was informing the field of science
  studies but not, to a large degree, how education researchers thought
  about or taught computing.
\item
  The lack of support for diverse ways of knowing in computing
  classrooms demonstrably hurt students.
\end{enumerate}

To Turkle and Papert (1991), the result of this culturally-rooted
epistemological tension was a war. More specifically, it was a war where
students of computing were the casualties, and the aggressor was a
distributed failure to recognize, embrace, and support diverse ways of
knowing. While I think their strong-form characterization of a ``war''
no longer applies, their consideration of epistemological issues is
still relevant two decades later. An understanding that computing does
in fact involve diverse ways of knowing should continue to inform
research and discussions on learning

This paper tries to model what directs and sustains students'
in-the-moment activity when they program. Its focus is on early-stage
program design; analyzing how what students say, do, write, and gesture
\emph{even before they type code} can help us improve theories of
cognition and activity in learning programming. Using a ``revelatory
case study'' (Yin, 2009) of two students in an introductory programming
course, I argue the following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Students' early-stage design activity reveals patterns outside the
  explanatory scope of misconception-based accounts of cognition.
\item
  For a subset of phenomena, we can recast students' productive
  capacities and their difficulties in terms of epistemological stances.
\item
  Such recastings are analytically powerful. Beyond Rebecca and Lionel,
  they could potentially explain the diversity of practices I saw other
  students take up.
\item
  As evidenced by work in science and math education, dynamic
  epistemological models can offer a lens for reforming assessment and
  instruction.
\end{enumerate}

\subsection{Literature Review}\label{literature-review}

I begin by exploring the historical legacy of Turkle and Papert's (1991)
remarks on epistemological pluralism in computing. I'll then describe
what I see as a well-intentioned obstacle: the research-based
preoccupation with students' misconceptions in computing education.
These two stra.

\subsubsection{Computing culture should support a diversity of ways of
knowing}\label{computing-culture-should-support-a-diversity-of-ways-of-knowing}

Turkle \& Papert (1991) describe a tension between individuals and a
kind of cultural collective. In this tension, individual students and
their personally-identified ways of knowing were in conflict with the
larger culture of computing and its manifested ways of knowing. Lisa,
for instance, was a Harvard student who identified as a poet:

Lisa experiences language as transparent, she knows where all the
elements are at every point in the development of her ideas. She wants
her relationship to computer language to be similarly transparent. When
she builds large programs she prefers to write her own smaller
``building block'' procedures even though she could use prepackaged ones
from a program library; she resents the latter's opacity. (Turkle \&
Papert, 1991, p. 164)

Turkle and Papert connect Lisa's programming practices to Lisa's
personal view of knowledge. She writes building block procedures because
of a continuity between herself as knower-of-her-own-poetry and that of
knower-of-her-own-programs. Across both contexts, Lisa \emph{resents
opacity} because it is an obstacle to her knowing. But, that resentment
slows her progress on projects in the class.

In part because of her commitments to transparency and the difficulty
they created, Lisa changed her approach to programming. She ultimately
had to, in her words, ``be a different kind of person with the machine''
(Turkle \& Papert, 1991, p. 164). Turkle and Papert explain:

She had been told that the ``right way'' to do things was to control a
program through planning and black­boxing, the technique that lets you
exploit opacity to plan something large without knowing in advance how
the details will be managed. Lisa recognized the value of these
techniques --- for someone else. She struggled against using them as the
starting points for her learning. Lisa ended up abandoning the fight,
doing things ``their way,'' and accepting the inevitable alienation from
her work. (Turkle \& Papert, 1991, p. 164)

The phrase ``inevitable alienation'' is particularly damning. One might
accuse Turkle and Papert (1991) of hyperbole, but research from other
disciplines suggests they're not overstating the case. Identity
alienation for epistemological reasons --- or some variation of them ---
is playing out in other subjects as well.

Connections and tensions between identities, learning, and personal ways
of knowing are not unique to computing education. Wortham (2006), for
example, detailed instances of social identification and academic
identity as \emph{jointly emerging} in a combined high school English
and History class. Nasir and colleagues explored how different learning
environments afford access to identity, domain practices, and
self-expression (Nasir \& Cooks, 2009; Nasir \& Hand, 2008). To single
out one study, Nasir \& Hand (2008) followed players ``from the
{[}basketball{]} court to the classroom.'' They found the court afforded
players a sense of role on the team, chances to be expressive through
play, and ``access to the domain {[}of basketball{]} as a whole'' (Nasir
\& Hand, 2006, p. 147). Comparatively, the mathematics classroom offered
little (if any) sense that students were on a team, a much narrower role
for them as having either right or wrong answers, and no strong sense of
self-expression or creativity through activity. These markedly different
contexts led not only to differences in kinds of activity but, Nasir and
Hand (2006) argue, divergent \emph{practice-linked identities} for the
players. In a sense, players had a different identity on the court than
they did in the mathematics classroom because who they were was strongly
coupled to a sense of their relationship with a practice.

Boaler's work on learning and identity in mathematics classrooms (Boaler
\& Greeno, 2000; Boaler, 1998, 2000, 2002) offers resonant findings. In
particular, Boaler (2000) strongly echoes the student voices in Turkle
and Papert (1991). Where Turkle and Papert's aggressor is the dominant
computing culture, Boaler describes a beast with similar effects in the
form of ``school mathematics'' --- the \emph{monotonous},
\emph{meaningless} (Boaler, 2002, pp. 383--384) imposition of routinized
mathematical procedures in ways that seem far removed from the real
world. Of school mathematics, Boaler writes:

School mathematics, for many of them, was \emph{of another world} and to
fully engage in that world, students needed to suspend their knowledge
of the real world, suppress their desire to interact with others, and
strive to reproduce standard procedures that held little meaning for
them. (Boaler, 2002, p. 392 emphasis in original)

Schoenfeld (1988, 1991) had already argued school mathematics could and
did depart substantially from the core of mathematical thinking. Boaler
established that such school mathematics could alienate learners
(Boaler, 1998, 2000) and influence their feelings about pursuing
mathematics after secondary school (Boaler \& Greeno, 2000).

Most recently, Danielak, Gupta, and Elby (in press) extended those
findings in engineering education. In their 3 ½ year case study of an
electrical engineering undergraduate named ``Michael,'' those authors
argued Michael's practice of sense-making in engineering coursework was
strongly coupled to his identity. Michael's passion for deeply
understanding concepts, much like Lisa's (Turkle \& Papert, 1991) desire
to understand every element of her program, did not mesh neatly with the
larger culture of his courses. But, because Michael's practice of
sense-making entwined with his identity, forces that pushed against
sense-making alienated him. That tension was particularly pointed when
Michael described his father's resistance to Michael's way of thinking:

{[}My dad said{]} ``you're just an undergraduate. Nobody expects
undergraduates to understand how anything works. That's why you go to
graduate school.'' I was like ``look, you know. I'm gonna be an unhappy
person if I have to....I have life goals other than to just get a good
grade on the exam. Other things are important to me.'' (Danielak et al.,
in press)

And, much like Turkle and Papert's (1991) Lisa, Michael learned to
curtail parts of his identity to get by in class:

I think the reason {[}pursuing deep learning{]} actually hasn't affected
my GPA is because I view learning as a hobby. So, as with any hobby, you
shouldn't let it interfere your GPA. But it \emph{is} one of my hobbies,
and I \emph{do} enjoy learning, I just---up to the point where I get my
grades done \{raises eyebrow\}. (Danielak et al., in press)

Ultimately, research suggests the identity-epistemology tensions Turkle
and Papert (1991) described in computing resonates with those found in
related disciplines. And, what such accounts --- particularly those of
Boaler and others (Boaler \& Greeno, 2000; Boaler, 1998, 2000, 2002;
Danielak et al., in press) --- have in common is a culturally-sanctioned
``right way'' (Turkle \& Papert, 1991, p. 164) of thinking or knowing
that alienates some students. Avowedly such ways of knowing have use and
value. Turkle \& Papert themselves acknowledge ``there are many
virtues'' to black-boxing and top-down design in programming (Turkle \&
Papert, 1991, p. 165). But,

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  When a preponderance of results from science studies suggests a
  plurality of ways of knowing among practicing scientists and
  programmers, and
\item
  When research from education shows that the inflexible imposition of
  one way of knowing above all else is alienating students, then
\item
  It's worth reconsidering whether a narrow sense of what counts as
  knowing might still be hindering improvements in computing education.
\end{enumerate}

In the next section, I turn to misconceptions research in computing
education. In keeping with observations 1--3, I argue research
approaches that inflexibly privilege canonical knowledge do so at the
expense of other productive knowledge and ways of knowing that students
have.

\subsubsection{Misconceptions research in computing education tends to
ignore students' productive
knowledge}\label{misconceptions-research-in-computing-education-tends-to-ignore-students-productive-knowledge}

In the past three decades, educational research has had a marked focus
on students' misconceptions in programming. It's a focus with a sensible
origin. Students get things wrong in programming --- often
systematically so --- and in ways that seem resistant to instruction.
The cause of those errors is theorized to be something cognitive,
whether it's a ``bug'' (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
``misconception'' (Bayman \& Mayer, 1983; Bonar \& Soloway, 1985;
Clancy, 2004; Gal-Ezer \& Zur, 2004; Herman et al., 2008; Kaczmarczyk et
al., 2010), a ``belief'' (Fleury, 1993) or a ``student-constructed
rule'' (Fleury, 1991, 2000). Instruction should try to identify,
address, and correct these misconceptions (Clancy, 2004) because they
can represent barriers to learning.

How that line of thinking and research becomes problematic is two-fold.
First, when taken in total the alleged brokenness of student knowledge
begins eclipsing all else in studying the cognition of learning to
program. In other words, most cognitively-focused educational research
in computer science treats students as having varied degrees of
deficiency with respect to canonical knowledge. Below is an unordered,
partial sampling of topics about which researchers have documented
students' misconceptions. Note across the list the variation in both the
grain sizes of students' misconceptions and the programming languages in
which they manifest:

\begin{itemize}
\item
  Objects in object-oriented programming (Holland, Griffiths, \&
  Woodman, 1997)
\item
  Algorithms and data structures (Danielsiek, Paul, \& Vahrenhold, 2012;
  Paul \& Vahrenhold, 2013)
\item
  Programming statements in BASIC (Bayman \& Mayer, 1983)
\item
  Programming in Java (Fleury, 2000)
\item
  Programming in Pascal (Fleury, 1993)
\item
  Parameter-passing (Fleury, 1991)
\item
  Arrays in Java (Kaczmarczyk et al., 2010)
\item
  Objects in Java (Kaczmarczyk et al., 2010)
\item
  Algorithms and computational complexity (Trakhtenbrot, 2013)
\item
  Boolean logic (Herman et al., 2008)
\item
  The efficiency of algorithms (Gal-Ezer \& Zur, 2004)
\item
  The Build-Heap algorithm (Seppälä, Malmi, \& Korhonen, 2006)
\item
  Hashtables (Patitsas, Craig, \& Easterbrook, 2013)
\item
  The correctness of programs (Kolikant \& Mussai, 2008)
\end{itemize}

Clancy (2004) provides a comprehensive overview of this line of
research, though in the past decade it has only grown. Indeed, roughly
half the articles above were published in the ten years since Clancy's
overview.

Identifying and removing barriers to student learning seems like a good
thing. So, it should follow that cataloging student misconceptions and
developing remedies for them should also be a good thing. But, the
logical implication isn't that clean. In some cases students display
productive, useful knowledge that's either ignored or outright
criticized by researchers. Aligning students toward canonical knowledge
makes sense, but doing so at the expense of---or in direct contradiction
to---useful ways of knowing seems undesirable at best. Next, I expand on
two examples from misconceptions research in programming. Specifically,
I show how and why I think a misconceptions focus in programming casts
aside students' useful intuitions and understandings.

The first example comes from Kaczmarczyk et al. (2010). Part of that
study involved giving students snippets of Java code and asking students
to diagram (or pseudo-code) how the information would be stored in
memory. Below I have reproduced the code for Problem 2 (Kaczmarczyk et
al., 2010, p. 110):

Cheese{[}{]} cheeses = new Cheese{[}4{]};

Meat{[}{]} meats = new Meat{[}2{]};

Turkey turkey;

Ham ham;

RoastBeef roastBeef;

boolean lettuce = true;

boolean tomato = true;

SauceType sauceType = new SauceType();

int numMeat;

int numCheese;

In diagramming this information, a student in the study makes a mistake:

Student3 makes incorrect assumptions about connections between variables
to the extent that the student makes a mistake concerning the types of
the variables. As a result, the student places Objects of different
types in an array whose type matches none of them: ``And so because
there's two arrays, cheese and meats, uh, all those turkey and ham and
roast beef are gonna be sorted into the meats array.'' (Kaczmarczyk et
al., 2010, p. 110)

The researchers are correct in the sense that \textbf{turkey} and
\textbf{ham} and \textbf{roastBeef} will \emph{not} be sorted into the
\textbf{meats} array. First, there is no code here that places
\textbf{turkey} and \textbf{ham} in the array; there is only code that
declares them as variables. Moreover, as written, an attempt to place
\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} into the array
would fail. Because of type restrictions in Java, only objects of class
Meat (or objects that inherit from class Meat) can go in the array.
\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} are, perhaps
confusingly, references to object instances of classes \textbf{Turkey}
and \textbf{Ham} and \textbf{RoastBeef}, so in the current snippet they
cannot enter an array of type \textbf{Meat} because (1) they don't yet
exist as objects and (2) even if they did exist, their types don't match
the array's type. The authors call this misconception \emph{semantics to
semantics}, which occurs ``when the student inappropriately
assume{[}s{]} details about the relationship and operation of code
samples, although such information was neither given nor implied''
(Kaczmarczyk et al., 2010, p. 110).

Again, the researchers are right that the student is failing to describe
the code in a way consistent with canon.\footnote{In this case, canon is
  the specifications and operations of the Java language and its
  compilers.} But, in their non-canonical thinking Student3 evidences
potentially productive insights about design. Precisely \emph{because}
there is no code stating that \textbf{turkey} and \textbf{roastBeef} and
ham are sorted into the array, the student is \emph{inferring} that to
be true. And, while that behavior is not what's happening, it
\emph{would} be sensible to design a program where specific instances of
classes \textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} could go
into an array of type \textbf{Meat}. To do so, a designer could define
\textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} as inheriting
from \textbf{Meat}.

Student3 has an idea about a relationship between entities where that
relationship is not specified in the code. Kaczmarczyk et al.
(Kaczmarczyk et al., 2010) focus only on the downside of the idea: the
student fails to display a proper understanding of how arrays work in
Java. Moreover, the student might be prone to similar mistakes of
inferring information that does not actually exist in code. But, there
is also an upside of this idea. Because Student3 is thinking about
real-world propositions like turkey and ham being kinds of meat, they
might be prepared to appreciate and discuss an object-oriented way to
put turkey in a \textbf{Meat} array. But, that possibility is
speculative conjecture. We can't know for certain whether Student3 could
be tipped into a productive object-oriented design activity around the
meats example because that question was not a focus of the research.

My second example of research that criticizes students' non-canonical
understandings comes from Bonar and Soloway's (1983) study of Pascal
programmers, part of which is discussed in Pea (1986). A student in
Bonar \& Soloway's study was asked to ``Write a program which reads in
ten integers and prints the average of those integers'' (Bonar \&
Soloway, 1983, p. 12). In pseudo-code, she wrote:

Repeat

(1) Read a number (Num)

(1a) Count := Count + 1

(2) Add the number to Sum

(2a) Sum := Sum + Num

(3) until Count :=10

(4) Average := Sum div Num

(5) writeln (`average = `,Average)

The interviewer then asked whether (1a) and (2a) were ``the same kinds
of statements.'' That interchange is reproduced here:

Interviewer: Steps 1a and 2a: are those the same kinds of statements?

Subject: How's that, are they the same \emph{kind}. Ahhh, ummm, not
exactly, because with this {[}1a{]} you are adding - you initialize it
at zero and you're adding one to it {[}points to the right side of 1a{]}
which is just a constant kind of thing.

Interviewer: Yes

Subject: {[}points to 2a{]} Sum, initialized, to, uhh Sum to Sum plus
Num, ahh - thats {[}points to left side of 2a{]} storing two values in
one, two variables {[}points to Sum and Num on the right side of 2a{]}.
That's {[}now points to 1a{]} a counter, that's what keeps the whole
loop under control. Whereas, this thing {[}points to 2a{]} was probably
the most interesting thing\ldots{}about Pascal when I hit it. That you
could have the same, you sorta have the same thing here {[}points to
1a{]}, it was interesting that you cold have, you could save space by
having the Sum re-storing information on the left with two different
things there {[}points to right side of 2a{]}, so I didn't need to have
two. No, they're different to me.

Interviewer: So - in summary, how do you think of 1a?

Subject: I think of this {[}point to 1a{]} as just a constant, something
that keeps the loop under control. And this {[}points to 2a{]} has
something to do with something that you are gonna, that stores more
kinds of information that you are going to take out of the loop with
you. (Bonar \& Soloway, 1983, p. 12)

Pea's (1986) interpretation? ``Here, again, we see the student believing
that the programming language knows more about her intentions than it
possibly can'' (p. 32).

As before, this student has an idea about relationships in code. Pea
(1986) see the downside of her idea: believing PASCAL can understand
shades of programmer intent when, in fact, it cannot. And again, that
downside is real. It could cause trouble for this programmer later on if
she expects PASCAL to interpret her intent and it cannot.

In defense of the student, the question --- as asked --- is vague. Are
those statements the same \emph{to whom and in what way?} Pea (1986)
treats the data as though she meant ``the same to PASCAL.'' Indeed,
maybe she did, in which case his interpretation has traction. But,
another interpretation is that she meant to herself, or to someone else
reading the code. Those statements might not be the same \emph{to her}
because she treats (1a) as having a function of controlling iteration
while (2a)'s job is to combine two numbers into a new sum.

These two purposes, which for the sake of description I'll call
\emph{keeping control} and \emph{totaling up} are, in a sense,
different. The PASCAL compiler (and runtime) does not differentiate
them, but humans can. And, humans may well \emph{want} to differentiate
them. diSessa (1986) describes exactly this kind of differentiation as a
consequence of separating the structural understanding of a programming
language from a functional understanding of a language. As an example,
he discusses the structure/function difference with respect to
variables:

The structural aspects of a variable in a computer language are given
primarily by the rules for setting their values and for getting access
to their values. These rules apply in all contexts. In contrast, a
variable's functions might vary. Sometimes they might be described as
``a flag'' or more generally, as ``a communications device.'' At other
times a variable might function as ``a counter,'' ``data,'' or
``input.'' (diSessa, 1986, p. 202)

The student in Bonar \& Soloway's (1983) study did not show evidence of
understanding the structural similarities between (1a) and (2a) in her
pseudo-code. And, those authors as well as Pea (1986) justly insist that
similarity is important for students to understand. From a conceptual
standpoint, seeing the structural similarity constitutes a part of
``knowing'' PASCAL. But, even if knowing PASCAL were not the goal,
seeing the similarity helps one to take the perspective of a computing
agent that has no means for discerning programmer intent. Such
perspective-taking may help students avoid mistakes that arise from
over-assuming what a computer ``understands.''

The student did show evidence of understanding a functional difference
between (1a) and (2a), but Pea (1986) does not remark on that kind of
understanding at all.\footnote{Also glossed over is, to me, another
  important difference: a programmer might not know in advance which
  numbers are being passed in to the sum statement. So, in advance the
  programmer can say nothing about how the value of \textbf{Sum} will
  change as the loop iterates. In contrast, the programmer knows exactly
  how the value of Count will change with each loop iteration.} Again, I
claim this oversight is part of a subtle but observable trend in
programming misconceptions literature. While, or perhaps because
research has been so preoccupied warring with students' problematic
knowledge, it has sometimes failed to recover the productive knowledge
(or resources for building it) students have. In this example, the
student already has a grasp that syntactically similar statements could
serve different conceptual purposes. Hypothetically, the student might
use that information in design by calling up the ``◻=︎ + 1'' syntactical
template when the situation seems to demand \emph{keeping control},
while calling up ``◻= ︎ + number'' when \emph{totaling up} is the goal.
And, the idea that structurally identical symbol templates can serve
different functional and conceptual purposes fits precisely in line with
Sherin's (2001) theory of symbolic forms. For example, the symbolic
forms \emph{parts-of-a-whole} and \emph{base+change} have different
conceptual schemata. Parts-of-a-whole refers to the contributions of
component entities while base±change describes a kind of accumulation.
Specifically, the terms in base±change ``play different roles'' (Sherin,
2001, p. 534) But, the two distinct conceptual schemata share what I
would argue is the same concrete symbol template of how to write an
expression: ◻=︎ + ◻.

The problem, for learning to program, comes in needing to fluidly
interpret and write code in languages that may demand incommensurable,
or at least distinct, conceptual schemata. As I show later in Table 4,
three current programming languages make remarkably different use of the
plus sign (+) as an operator. Crucially, some of the entailing ways to
make sense of how + works in those languages don't exist in Sherin's
(2001) catalog of conceptual schemata. In other words, I would argue
there are conceptual ways a programmer may need to think about
interpreting or writing a ◻=︎ +◻ symbol template that Sherin doesn't
enumerate.\footnote{One of the most obvious is the movement from seeing
  ``◻=︎ +◻'' as a statement of equality to seeing it as the storage of a
  sum to a variable.} To follow that implication, the canonical body of
knowledge about which programmers must reason is itself fractured,
because different languages design their operations around different
symbolic and conceptual metaphors.

To return to misconceptions, what drives research on students'
misconceptions is largely a need to get students to program computers
and reason about computation in ways that are canonically correct. And,
those are assuredly worthwhile goals. But, as I've argued, we can
already identify cases where a narrow misconceptions focus is silent
about or dismissive of students' useful intuitions. We can also identify
the further problems whereby unilateral emphasis on one language's canon
opposes the vocabulary of symbolic forms (and diverse conceptual
schemata) expert programmers ultimately need (Sherin, 2001) Taken in
total, that silence, dismissiveness, and narrow view of refining
knowledge perpetuates a deficit-focused discourse about student's
knowledge in computing.\footnote{Worse still, in my view, is that there
  is a pattern of researchers failing to discuss or acknowledge
  students' productive knowledge \emph{even when their own data supports
  it}.} Such a perspective also fails to address what aspects of
students reasoning might get broken by attempts to ``fix'' such
``misconceptions''.

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge this way. I begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.

\subsubsection{Not all cognitive programming bugs imply a problem with
the
student}\label{not-all-cognitive-programming-bugs-imply-a-problem-with-the-student}

What's curious about Pea's (1986) comments on Bonar \& Soloway (1983) is
that Pea drew different conclusions from the data than Bonar and Soloway
did. Pea (1986) emphasized that when the student thought two
semantically-equivalent assignment statements in PASCAL were different,
it was a problem of egocentrism: ``students assume that there is
\emph{more} of their meaning for what they want to accomplish in the
program than is actually present in the code they have written'' (p.
30). In other words, Pea treated the data as a fairly clear example of a
class of bugs. Specifically, he saw a bug class in which students simply
assumed the interpreter or runtime could infer the code author's shades
of intent.

Bonar and Soloway (1983), by contrast, were less quick to make
inferences either about the nature of the bug or about the intervention
it entailed. Rather than jump to definitive conclusions, they were
circumspect:

It is not clear exactly how to react to the bugs we have uncovered in
novice understanding of programming. In some cases it may be appropriate
to design new languages or constructs. Often, better instruction would
take care of the problem. The intent of our studies is to better
understand the source of the mismatches and misconceptions that cause
novice bugs. Only once a bug is uncovered and understood are we ready to
create a remedy for that bug. (Bonar \& Soloway, 1983, p. 12)

That the authors would even consider developing new constructs or
languages is a noteworthy distinction. Rather than assume in toto that
students with ``bugs'' had wrong knowledge, the authors instead suppose
cognitive bugs have plausible origins worth designing around. Moreover,
they treat students' divergence from canon as \emph{an opportunity for
research to learn from students}. Precisely because students saw
functional differences (cf., diSessa, 1986) in semantically-equivalent
PASCAL statements, Bonar and Soloway (1983) reflect that perhaps
programming languages should be more expressive:

We find it quite interesting that novices seem to understand the role or
strategy of statements more clearly than the standard semantics. Such
roles discussed here include ``counter variable,'' ``running total
variable,'' ``running total loop,'' and ``first, then rest loop''. (See
Soloway et al {[}1982b{]} for a detailed discussion of novice looping
strategies.) Much work in programming languages is concerned with
allowing a programmer to more accurately express his or her intentions
in the program. Perhaps we can learn something from novices here - our
programming systems should support recording the roles the programmer
intends for various statements and variables. (Bonar \& Soloway, 1983,
p. 12)

Again, what's noteworthy here is that rather than treating students'
non-canonical views as a burden for instruction, Bonar and Soloway
instead see them as an opportunity for programming language designers to
make languages better.

That insight---that instruction and design can meet novices where they
are---carries through to their final remarks about studying and
analyzing novice programming knowledge:

The experience and understanding of a novice are available for analysis.
In particular, our results suggest that the knowledge people bring from
natural language has a key effect on their early programming efforts.
Our work suggests that we need serious study of the knowledge novices
bring to a computing system. For most computerized tasks there is some
model that a novice will use in his or her first attempts. We need to
understand when it is appropriate to appeal to this model, and, when
necessary, how to move a novice to some more appropriate model. (Bonar
\& Soloway, 1983, p. 13)

I assume Bonar and Soloway structured their final line deliberately. If
so, their phrasing has three consequences:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Appealing to novice's existing models gets precedence. That is,
  understanding how to leverage novices' existing knowledge comes first
  in research.
\item
  Changing the models novices have comes next, and ``when necessary.''
\item
  They speak of ``how to move a novice to some more appropriate model,''
  which does not necessarily entail the removal or destruction of
  novice's existing models.
\end{enumerate}

Taken together, these points convey a sense of how Bonar and Soloway
view learning and instruction in computing. Instruction explicitly
includes appeals to prior models and knowledge students might already
have. Learning, meanwhile, involves the movement \emph{when necessary}
to more appropriate models of computation. As I explain in the next
section, such a view exactly aligns with a particular branch of
constructivism, where cognition is viewed as the complex activation of
manifold resources for thinking and knowing.

\subsubsection{Examples motivate the need for contextual-sensitivity in
modeling programming
cognition}\label{examples-motivate-the-need-for-contextual-sensitivity-in-modeling-programming-cognition}

Let's begin with two motivating examples. My aim with these examples is
to show how a practicing programmer might employ specific, distinct
conceptual models to reason locally about a piece of code. First,
consider variants of the PASCAL statements from Bonar and Soloway
(1983), where an assignment statement worked to increment a value and
store the result back to that value. In Table 2 below, I imagine five
different ways of writing a programming statement. Each of the five ways
is semantically equivalent to the others (or near enough for explanatory
purposes). And, in each example I add a layer of specificity to the
syntax. I also propose a corresponding interpretation of how I might
apply interpret and think about a given statement.\footnote{I assume for
  simplicity's sake that all variables and functions in my code samples
  evaluate to floating-point numbers that can be combined under
  addition. I also assume, crucially, that all statements are in the
  same hypothetical language and that the variable names connote nothing
  to the computational interpreter.}

\protect\hypertarget{ux5fToc252445956}{}{}Table -- Five different
interpretations of semantically equivalent programming statements in the
same language

\begin{longtable}[]{@{}lll@{}}
\toprule
\textbf{Statement } & Programming Syntax & How I might think about
it\tabularnewline
\midrule
\endhead
\textbf{1} & x = x + d & An incrementer or accumulator\tabularnewline
\textbf{2} & x = x + update & Updating the value of x\tabularnewline
\textbf{3} & x\_position = x\_position + update & Updating the
x-position with uniform velocity\tabularnewline
\textbf{4} & x\_position = x\_position + update(t) & Acceleration on the
x-position\tabularnewline
\textbf{5} & x\_position = x\_position + update(\ldots{}) & Generalized
x-position update (update could be any function of any number of
parameters: random noise, low-level hardware functions getting mouse or
keyboard input, etc.).\tabularnewline
\bottomrule
\end{longtable}

I stress that the third column is about \emph{how I might think about}
each programming statement\emph{.} By no means am I making normative
claims about how one \emph{ought} to think about it, or whether the
statement actually does what its names might suggest it does. Rather,
``How I might think about it'' reflects the kind of local meaning or
interpretation I might attach to such a statement when I work with it,
given my understanding of its role and context.

Consider specifically statements 1 and 4. Statement 1 could very well be
an incrementer in some kind of iterative code. If I had to debug code
that employs statement 1, what I might do is exploit my knowledge of
what d is (does it hold a constant value? Is it 1?) and try inserting
intermediate print statements into the iterative code. Doing so, I can
inspect textually how values change with each iteration. In statement
4's case, it could very well be that the position-update code animates
images on a screen. If there's a problem with that code, one of the
easiest ways I might notice is that the acceleration seems off in the
graphics. Consequently, my debugging might call upon knowledge I have
from kinematics. I might try inserting code that draws a dot at the
object's on-screen position with each iteration, leaving a trail I can
visually inspect. I could then look at the path of the object's
trajectory and the spacing patterns between successive dots as a
first-pass test of whether my code achieves the motion I want.

To be clear, it's not just that graphics might improve my efficiency in
debugging a statement. Rather, applying an interpretive frame that
treats an assignment statement as a kinematic position update lets me
\emph{use conceptual knowledge from physics} to diagnose and fix
problems in my code. If I view the assignment statement as saying
something about the motion of an object (cf., Hammer, 1994, p. 165), a
field of knowledge and concomitant techniques from physics becomes
available to me to think with. But, if I focus only on the
semantic-level equivalence of statements 1 through 5, there would be no
obvious reason for me to access what I know about physics in order to
reason about the code.

My second example concerns statements that are syntactically-similar,
rather than semantically-similar. I proposed that the statements in
Table 2 all came from the same language. But, another phenomenon comes
into play when different languages use the same symbology for
conceptually different operations. Table 3 shows examples of what look
like semantically-equivalent operations, but in fact are not.

\protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
syntactically-similar statements with very different semantics

\begin{longtable}[]{@{}llll@{}}
\toprule
\textbf{Statement } & Programming Syntax & Language & How I think about
it\tabularnewline
\midrule
\endhead
\textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
\textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
the string w\tabularnewline
\textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
layer of points onto a plot p\tabularnewline
\bottomrule
\end{longtable}

The catch here is that the plus operator takes on different roles in
different languages because of how those languages define its use.
Statement 1 increments a number in C; Statement 2 appends the letters
``ly'' to a string; Statement 3 adds a layer of points to a statistical
graphics plot. These different kinds of operations become even more
apparent and consequential when, for example, such statements are
repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
  operator can be written as a directional arrow. The symbol
  \textless{}- (``less-than, hyphen'') indicates the value on the right
  side of the symbol is being assigned to the variable on the left side
  of the symbol. As used, the symbol itself is semantically equivalent
  to having written an equals sign (``=``).} I'm using multiple
reassignment statements to compose a statistical graphics plot. The
layered creation of a plot invites a very different kind of conceptual
interpretation than, say, repeatedly accumulating numbers into a running
sum:

p \textless{}- InitializeGgplot\_1w()

p \textless{}- p + GrandMeanLine(owp)

p \textless{}- p + GrandMeanPoint(owp)

p \textless{}- p + ScaleX\_1w(owp)

p \textless{}- p + ScaleY\_1w(owp)

p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)

One obvious reason for thinking about this code with a different
interpretive frame is that numeric addition is commutative; composing a
plot is not necessarily commutative.\footnote{To convince yourself that
  plot composition isn't commutative, imagine a scale function that
  squares up the aspect ratio of a plot and another scale function that
  sets the aspect ratio to 1.5:1. Applying the square function last
  produces a square plot; applying it first produces a rectangular plot.
  String concatenation is also not necessarily commutative. ``cat'' +
  ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
  ``dogcat.''} So, despite the syntactic similarity, reassignments that
compose a plot using reassignment (as above) does not obey the same
rules as reassignments for a running total. But, even within this code
block, statements that look alike perform operations of a different
nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
visual layer to a plot, others modify features of the plot (e.g, ``+
Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
the numeric range of the data).

Given these motivating examples, it seems sensible to think there's
utility in a programmer having different conceptual metaphors available
to think about and work with code. Example 1 shows that choosing to
apply knowledge from physics to a piece of code can change the cognitive
nature of debugging. There, debugging a position-update statement
becomes, in part, reasoning kinematically about the properties of motion
trails.\footnote{For comparison, consider the argument that ringing an
  aircraft speedometer with physical markers changes the nature of
  cognition when pilots work to land a plane (Hutchins, 1995b).} Example
2 shows that across languages, programmers might have to deploy
different conceptual metaphors to reason about statements in a
locally-consistent way. Knowing that plots in ggplot2 can be composed
layer-by-layer with reassignment is crucial if you're trying to write or
understand code that creates statistical graphics. But, I would argue
that thinking about ◻ = ◻ + ◻ as ``compose new layer onto plot'' can and
does appeal to different kinds of knowledge when compared to thinking
about ◻ = ◻ + ◻ as ``include this addend in the sum,'' which itself can
and does appeal to different kinds of concepts when compared to thinking
about ◻ = ◻ + ◻ as ``increment the counter.''

Stepping back, I can build the following argument

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Programming can be helped by applying conceptual models to code,
  particularly when relevant domain-knowledge structures can
  advantageously transform a problem (example 1)
\item
  But, conceptual models don't work all the time for all statements.
  Because languages are designed differently, the same syntax can
  actually correspond to very different operations in code (example 2).
  And, that's true both within and across languages.
\item
  Consequently, it makes less sense to treat conceptual models as right
  or wrong, and more sense to treat them as differentially advantageous
  for thinking about what a piece of code does. (Thinking a ``+''
  implies numerical addition isn't \emph{globally wrong} in JavaScript,
  but it won't explain why 1 + ``1'' yields ``11'' as a result.)
\item
  It seems plausible that successful programmers, when reasoning about
  or writing code, are able to dynamically access or deploy conceptual
  models that are advantageous given the context (language, syntax,
  surrounding code). Certainly such a supposition is in line with work
  suggesting students have resources for thinking conceptually and
  dynamically about how mathematics models real-world situations (Izsák,
  2004; Sherin, 2001)
\item
  To model how programmers think with conceptual models, a suitable
  framework should be able to account for the dynamic, context-sensitive
  deployment of conceptual knowledge.
\item
  To model how programmers develop expertise, a suitable framework
  should be able to describe higher-order phenomena. Such phenomena
  include explaining how programmers come to have conceptual models or
  generate new ones, why they decide to deploy them, and how programmers
  consider which conceptual model (i.e., which way to think about code)
  is appropriate.
\end{enumerate}

Taken together, these assertions propose criteria for how we might
strive to model cognition in programming. Our modeling frameworks should
be context-dependent, dynamic, and capable of explaining where
conceptual models come from. They should also be able to account for
phenomena that are not themselves conceptual, including what directs the
use of certain kinds of conceptual knowledge. In the learning sciences,
such models already exist and have proven useful and productive for
thinking about thinking.

\subsubsection{Manifold models of cognition explain context-dependence
and the growth of
expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}

In 1993, a pair of articles in the learning sciences staked a strong
claim for viewing knowledge as a network of pieces, isolated enough to
be locally triggered but trainable enough to fire in larger concerted
patterns (diSessa, 1993; Smith, diSessa, \& Roschelle, 1993). Informed
in part by agent-based accounts of cognition (Minsky, 1986) and complex
systems models (diSessa, 2002), the central tenets of an ``in-pieces''
approach hold that knowledge is emergent from interacting primitives,
rather than unitary and monolithic. An example from Smith, diSessa, and
Roschelle helps illustrate the point.

The authors show that we might think of a rubber band as a different
conceptual entity depending on context. In several different
situations---wrapped around a newspaper, pulled taut as a string, spun
to store energy in a toy plane propeller---we intuitively think about
the rubber band's physical behavior differently: one as a negligible
part of the newspaper's point mass, another as a transverse pendulum
(likely) obeying Hooke's Law, and the third as a torsional spring. Those
differences in intuitive thinking reflect the contextual dependency of
what we know about the physical world:

In each of the rubberband examples, various pieces of intuitive physical
knowledge describe the mechanism at work: the rubber band binds the
newspaper, grips the jar lid, and acts a source of springiness for the
bobbing object. Although a mapping cannot be made from the rubberband to
scientific entities, it is quite easy to map these qualitatively
distinct physical processes to scientific entities and laws. For
example, instances of binding almost always map to a practically rigid
body. Likewise, gripping maps to friction forces, and springiness maps
to Hooke's law. This suggests that applicability can depend directly on
our intuitive knowledge---knowledge that exists prior to any formal
scientific training (Smith et al., 1993, p. 144).

The in-pieces approach to modeling cognition has been used, among other
things, to explain how experts reason about fractions and decimals
(Smith et al., 1993), how students reason about forces in physics
(diSessa \& Sherin, 1998; diSessa, 1993; Hammer, 1996; Sherin, 2001),
how students construct and evaluate algebraic representations of
physical situations (Izsák, 2004), and how knowledge transfers across
contexts (Hammer et al., 2005; Wagner, 2006). Because its starting
assumption is that knowledge is fragmented, knowledge-in-pieces can
account for wide variations of how people---particularly novices---use
knowledge on a moment-to-moment basis. In other words, because it
assumes knowledge is local, it can still explain the kinds of
globally-inconsistent ways people might reason about physical situations
(diSessa, 1993). As a framework, an in-pieces approach ultimately argues
that models of concept replacement and good/bad criteria for knowledge
should be supplanted by a learning model of alignment/refinement of
prior knowledge and the consideration of knowledge as
productive/unproductive.

Hammer and colleagues have worked to extend the in-pieces approach to
explain how students' epistemological activity---how they orient toward
knowledge and knowing in a context (Hammer et al., 2005; Hammer \& Elby,
2002, 2003). Specifically, those authors use two core theoretical
constructs to explain students' stances toward knowledge and knowing:

\begin{itemize}
\item
  \emph{Epistemological Resources} are the epistemological equivalent of
  diSessa's phenomenological primitives (p-prims). Resources, the
  authors propose, are the atomic units involved in how people cognize
  about the source of knowledge, the nature of knowledge, and
  epistemological activities (Hammer \& Elby, 2003; Louca, Elby, Hammer,
  \& Kagey, 2004).
\item
  \emph{Epistemological Frames} are the emergent result of subsets of
  resources acting in concert. Drawing from both Goffman's (1974)
  sociological notion of frame as structures of expectations and
  subsequent work on framing in discourse (Tannen, 1993),
  epistemological frames are a participant's local answer to the
  question ``what is it {[}specifically, what knowledge activity{]}
  that's going on here'' (Goffman, 1974, p. 8).
\end{itemize}

Resources can frames can interact in activity settings to produce
larger-scale patterns called ``epistemological coherences'' (Rosenberg,
Hammer, \& Phelan, 2006) where evidence from data suggests that a
network of discrete cognitive units can nonetheless give rise to stable
cognition.

A useful conceptual metaphor for resources and framing is to think about
a lecture hall with different sets of lights: a spotlight in the back to
highlight a lecturer, chalkboard lights, house lights, and a projector.
Each light has an individual brightness, but the lights are only
controllable at the per-bulb level. In that sense, they are atomic. But,
light patterns interact with one another to create a field of lighting
for the room. Thus, lights are a bit like resources, and different light
configurations can correspond to different, sociologically-stable uses
of the room.

\protect\hypertarget{ux5fToc252445958}{}{}Table -- Using room lighting
configurations to think about epistemological framing

\begin{longtable}[]{@{}lllll@{}}
\toprule
\textbf{What is it that's going on?} & Houselights & Spotlight &
Chalkboard & Projector\tabularnewline
\midrule
\endhead
\textbf{On-stage monologue} & Low & ON & Low & OFF\tabularnewline
\textbf{Presenting slides} & Low & ON & Low & ON\tabularnewline
\textbf{Working through equations} & Low & ON & High &
OFF\tabularnewline
\textbf{Students discussing with each other} & High & OFF & Low &
OFF\tabularnewline
\textbf{Watching a movie} & Low & OFF & Low & ON\tabularnewline
\textbf{Cleaning the room after a movie} & High & OFF & High &
OFF\tabularnewline
\bottomrule
\end{longtable}

To convince yourself of the sociological stability of these
configurations, imagine you were watching a movie when suddenly the
house and chalkboard lights came up to full intensity. It would jar you
out of the experience. You might wonder whether something was wrong: is
there an emergency? Should you evacuate?

The metaphor isn't perfect. Ontologically, for example, resource and
framing theory propose these constructs exist not in the world but in
the minds of individuals. But, the metaphor is quite useful for
understanding how individual elements---in this case lights---can work
in concert to create and sustain a stable frame. And, crucially for a
cognitive system, the metaphor lets us account for variation in
activity. The same lecture hall can become a window onto one person's
thoughts (monologue), a site for instruction (working through equations
at the board), a shared space for collaboration (student discussion), or
a place in need of repair (cleaning up) simply by varying the
intensities of lighting banks in particular ways.

An example helps ground this in-pieces approach to epistemology. In
Russ, Coffey, Hammer, and Hutchison (2008), the authors describe the
situation where an elementary student reasons about why an empty juice
box collapses when you suck on the straw. One student gives what the
authors deem to be an excellent mechanistic account of why the juice box
collapses. But, as the teacher seems to steer the discussion toward
vocabulary---in this case, ``pressure''---the student clearly pulls back
from her mechanistic reasoning, seems much more diffident, and claims
that pressure is hard to explain. That example highlights the disconnect
between doing science as knowing vocabulary and doing science as
reasoning mechanistically. Moreover, it strikingly highlights that a
student who by all accounts produced an excellent explanation of how
pressure works was left nonetheless with the impression that pressure
was hard to explain.\footnote{In fairness to the teacher, I'm trying to
  focus on the result of the interaction and far less on the intent the
  teacher might have had. The student still left with a sense that
  science might be about vocabulary, even if that's not the view the
  teacher would espouse or was trying to enact in the moment.}

\subsection{Methods and Theoretical
Commitments}\label{methods-and-theoretical-commitments}

\subsubsection{Student population, course background, and selection
}\label{student-population-course-background-and-selection}

This paper focuses on Electrical Engineering (EE) students from Flagship
State, a large public research institution in the mid-Atlantic. The
students I studied were taking Intermediate Programming, the second
semester of an introductory programming course taught in C. The course
was exclusively for EE majors and taught by EE faculty, and its
enrollees were typically first-year or second-year EE majors. Its model
was two 75-minute lectures and a 1-hour teaching assistant-led
discussion section for students each week. Course topics included:

\begin{itemize}
\item
  Strings
\item
  Pointers
\item
  Dynamic Memory Allocation
\item
  Testing
\item
  Debugging
\item
  Hash tables
\item
  Trees
\item
  Linked Lists
\item
  Abstract Data Types
\item
  Functional Decomposition
\end{itemize}

Students in the course had varying degrees of experience with
programming. Partly, that variation is because students with Advanced
Placement (AP) Computer Science credit could place out of Basic
Programming, the first semester C course. So, some students in
Intermediate Programming came from technical magnet schools where they
may have already had one or more years of programming in multiple
languages, while other students may have been first-time programmers
with only one semester of programming experience: Basic Programming.

I studied the same course, Intermediate Programming, during the fall
2011 semester and the spring 2012 semester. During fall 2011 I
ethnographically observed over 50\% of the course lectures, and during
spring 2012 I sat in on several discussion sections. At the beginning of
each semester I solicited student participants for semi-structured
clinical interviews. Of the students who responded to the solicitation,
I scheduled interviews opportunistically with students given my
resources as the sole researcher on the project. In total, I worked with
a cohort of 6 students in fall 2011 and 4 students in spring 2012.

My specific interest was in how students design computer programs. By
that, I mean I wanted to know not just how students program, but whether
and how they structured programs, how they did (or did not) try to
incorporate modularity into their programs, and how the final form of a
program reflected what they had learned about how to manage complexity
through software. In the sections that follow I outline the palette of
methods I used to pursue those questions.\footnote{The language in
  sections 3.3.2 and 3.3.3 is taken with slight modification, from
  ``Studying students' early-stage software design practices,'' a paper
  I authored, along with William Doane, and submitted to the 2014
  International Conference of the Learning Sciences (ICLS).}

\subsubsection{Studying design as a complex phenomenon of disciplinary
practice}\label{studying-design-as-a-complex-phenomenon-of-disciplinary-practice}

Practicing professionals make complex use of talk, gesture, and
representational artifacts in physics (Gupta, Hammer, \& Redish, 2010;
Kaiser, 2005; Ochs, Gonzales, \& Jacoby, 1996); chemistry (Stieff,
2007); field biology (Hall, Stevens, \& Torralba, 2002); civil
engineering (Stevens \& Hall, 1998); structural engineering (Gainsburg,
2006); mechanical and electrical engineering (Bucciarelli, 1994;
Henderson, 1999); and architectural design (Hall et al., 2002; Hall,
1999). Given that:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Science and engineering education research has made progress by
  looking for continuities in how novice learners develop disciplinary
  practices (Gupta et al., 2010; Hall \& Stevens, 1995; Smith et al.,
  1993; Stevens \& Hall, 1998), and
\item
  Emerging research on software engineering reveals that early-stage
  software design involves complex inscriptional, discursively, and
  epistemic practices,
\end{enumerate}

it seems striking that there is no contemporary body of research,
comparable to studies of expert practice, that looks at students'
software design practices. In other words, we have every reason to
believe that expertise in software design involves complex practices,
but little (if any) research on what productive capacities students have
for them. Finding those productive design capacities requires a shift
from questions such as \emph{how can we assess and mitigate students'
difficulty in programming?} toward questions such as \emph{how do
students learn and display evidence of design thinking in programming?}

Rephrasing research questions asked of professional software engineers
(Petre et al., 2010, p. 533) and instead treating \emph{students} as
designers, I ask:

\begin{itemize}
\item
  What do students actually do during early stage software design work?
\item
  What does students' exploratory design thinking look like?
\item
  How do students communicate?
\item
  What sorts of drawings do students create when they design software?
\item
  What kinds of strategies do students apply in exploring the vast space
  of possible software designs?
\end{itemize}

\subsubsection{Deploying methods to capture the complexity of
early-stage design
work}\label{deploying-methods-to-capture-the-complexity-of-early-stage-design-work}

The methods below form the core of my developing program to study
students as software designers. None of the methods below are new; all
have been used in prior educational research. What \emph{is} new, I
believe, is the opportunity to combine them all under the umbrella of
understanding what happens when students design software. For each of
the four students in my spring 2012 cohort I captured multiple streams
of data.

\begin{itemize}
\item
  I collected their \textbf{code history}. By this, I mean I preserved
  frequent snapshots in time of what students' code looks like. Research
  has already shown code snapshotting to be a useful method for
  understanding large-scale patterns of student error (Jadud, 2006;
  Rodrigo et al., 2009; Spacco, Hovemeyer, et al., 2006). And, the
  resolution of snapshots is extremely fine: Spacco et al. are able to
  capture the contents of a file each and every time a student saves it
  to disk. But, that research takes an aggregate view: it identifies
  large-scale error patterns at the expense of detailed naturalistic
  understandings of why students make those errors. Moreover, it's
  primarily used to identify what \emph{mistakes} students make, which
  is distinctly different from a research orientation that considers the
  negative \emph{and} positive consequences of students' design choices.
  A currently untapped advantage of collecting code history data, then,
  is that while we historically use it to aggregate \emph{across
  programmers} it nevertheless also gives us fine-grained individual or
  team-based histories of how designs evolve.
\item
  I conducted \textbf{clinical interviews} with them. Clinical
  interviews have proven historically useful in understanding the
  substance of students' knowledge and the nature of conceptual change
  (diSessa \& Sherin, 1998; Duckworth, 2006; Ginsburg \& Opper, 1988;
  Ginsburg, 1997). Crucially for Computer Science Education (CSEd)
  research, clinical interviews can tell us about students'
  epistemologies---how they view knowledge and knowing in a discipline
  (Hofer \& Pintrich, 1997)---which can affect how they approach and
  adopt that discipline's practices (Hammer, 1989, 1994; Lising \& Elby,
  2005). Moreover, because my interviews were videotaped and analyzed
  from perspectives of interaction analysis (Goodwin, 2000; Jordan \&
  Henderson, 1995), they offer rich evidence of the substance of
  students' design practices
\item
  I analyzed their in-interview \textbf{inscriptions} when they designed
  --- what they wrote, how they wrote it, and how those writings got
  used. Evidence from both science studies (Hall et al., 2002;
  Henderson, 1999; Hutchins, 1995a; Kaiser, 2005; Latour, 1990; Ochs et
  al., 1996) and educational research (Hall \& Stevens, 1995; Lehrer,
  Schauble, Carpenter, \& Penner, 2000; Stevens \& Hall, 1998)
  highlights the centrality of inscriptions to disciplinary practice in
  science and mathematics. Ethnomethdological data from studies of
  professional software engineers supports the same result:
  inscriptional practice is central to how professional engineers design
  software (Rooksby \& Ikeya, 2012; Rooksby, 2010). And, since part of
  the inscriptional environment when designing software is the computer
  itself, I captured and analyzed what happened on-screen as students
  design programs.
\item
  I paid close attention to students' in-interview \textbf{gestures}.
  Perspectives of gestural analysis hold that gestures can not only
  support or extend thinking, they can also communicate entirely
  different information than what's being said (Goldin-Meadow, 2003).
  Moreover, perspectives from cognitive anthropology and embodied
  cognition studies argue that bodily motion \emph{is itself} cognition
  (Hall \& Nemirovsky, 2012; Hutchins, 1995a; Nemirovsky, Rasmussen,
  Sweeney, \& Wawro, 2012). For example, when students describe a part
  of code by moving their hand across an imaginary row of items and
  tapping each item, their bodies convey information we can interpret
  about how they understand iteration.
\end{itemize}

Figure 3-1 presents a visual overview of some of these methods and modes
of analysis. In particular, the first (top-left) panel depicts how we
deploy these data collection methods during a clinical interview:

\begin{itemize}
\item
  a voice recorder captures speech (often a redundancy in case another
  recording device fails)
\item
  a LiveScribe Pulse pen digitizes written inscriptions, allowing us to
  play back what was written in time
\item
  a videocamera records data for knowledge analysis, interaction
  analysis, and gestural analysis,
\item
  an in-interview computer tracks code history and its screen-capturing
  software records real-time activity.
\end{itemize}

\includegraphics[width=6.25556in,height=4.20000in]{media/image4.jpeg}

\protect\hypertarget{ux5fToc252445967}{}{}Figure ‑ -- An comic-based
overview of my methods for capturing students' design practices

\subsubsection{Developing a revelatory case of students' early-stage
design
work}\label{developing-a-revelatory-case-of-students-early-stage-design-work}

In the empirical work that follows I explore what productive knowledge
and resources students have for structuring programs. I begin with
Lionel, a male student in Intermediate Programming, fall 2011. First, I
propose the idea that Lionel connects ``designerly'' (Archer, 1979)
stances across contexts. I analyze his retrospective account of
modifying a bike and compare it to his retrospective accounts of how he
programs in Intermediate Programming. Then I detail Lionel's multi-stage
cascading approach to structuring a program solution in an interview.
The argument throughout is that from bike design to code design Lionel
evinces knowledge and stances that productively help him solve problems.

Next I transition to Rebecca, first explaining the epistemological
difficulty she felt when coding, then showing how she circumvented that
block by structuring a program piece using gesture and talk. I argue
that what Rebecca and Lionel have in common are resources for handling
design obstacles as they emerge. Where the two students differ is in
their in-the-moment judgments about what sorts of knowledge and
activities are appropriate to work through a design problem.

I argue that taken together, Rebecca and Lionel constitute what Yin
(2009) calls a ``revelatory case,'' by which I mean my investigation
arises from ``an opportunity to observe and analyze a phenomenon
previously inaccessible to social science inquiry.'' In my case, I argue
the phenomenon I'm observing is what engineering students do in the
early stages of design work on a program. Its inaccessibility is
demonstrated by the almost complete paucity of studies that draw on
real-time (e.g., video records) of how students begin work on a complex
program. Contrary to being ``representative'' or ``typical,'' my
revelatory case does not aim to generalize. Rather, its value is in
phenomenological richness. As Yin (2009) writes, and I think others
including Erickson (1986) would agree, such case studies are worth doing
``because the descriptive information alone will be revelatory.''

\subsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188553}{}{\protect\hypertarget{ux5fToc252445926}{}{}}Lionel's
approach to design and to
programming}{Lionel's approach to design and to programming}}\label{lionels-approach-to-design-and-to-programming}

\subsubsection{While modifying a bike Lionel decomposed tasks, explained
solutions to himself, and created intermediate design
representations}\label{while-modifying-a-bike-lionel-decomposed-tasks-explained-solutions-to-himself-and-created-intermediate-design-representations}

Prior to entering my study, Lionel had already earned a Bachelor's
degree in Business Information Technology at a different university. We
began the interview by discussing why he re-matriculated to earn an
electrical engineering degree.

Interviewer: How did you, um end up coming out of your business program
and your business experience and deciding that electrical engineering
was what you wanted?

Lionel: Um, well, since I was Business Information Technology, basically
all my job was doing was writing SQL code and, you know, short
statements like that. I also worked at a help desk for---I guess I
worked at Department of Homeland Security. And, pretty much besides the
SQL statements I worked with this one program and whenever somebody had
a problem with it they'd come to me. And basically that---that's what my
whole degree was---was going down that track and then that's really not
what I wanted to do: just sitting at a desk, you know people coming to
me being like ``oh, you gotta do this for me.'' So, I was like, alright.
Or, I have to fix a problem for them, pretty much. And that's really
what I didn't want to do. I wanted to do something more hands-on where I
could, you know, I'd be given a project where I'd do something on my
own, create something on my own. And, um, also my Dad and my uncle were
both electrical engineers. And my brother became a computer scientist.
Cu---so I've always been around, you know, that type of environment,
where my dad's always building something and, you know, I always see it
and I'm like, you know, that's very interesting. I'd like to do
something like that too. So it's really just uh, a mixture of bad
experience with my previous job and being around company that likes to
do that sort of stuff. (Interview 1 of 1, October 17, 2011)

Lionel apprenticed building things with his father first, taking on
small, highly-directed roles in projects.

Lionel: Be-before I started my degree, you know, I really didn't know
how to do any of this stuff. So, I would just kinda help my dad out a
little bit. He'd be like ``Oh, here,'' you know, ``to, to do this you
have to do exactly this.'' You know, ``solder this here,'' or, you know,
``drill this,'' blah blah blah. He'd tell me exactly what to do. So, I
just kind of followed along. (Interview 1 of 1, October 17, 2011)

Presumably in those projects Lionel was at a periphery and direction
came from his father. More recently, though, Lionel had struck out on
projects of his own. As Lionel said, ``now that uh, my dad kinda showed
me how to do certain typical types of things, I've tried to become
\textbar{}\textbar{}more adventurous\textbar{}\textbar{} \textbar{}\{air
quotes\}\textbar{} /Sure/ and do things on my own'' (Interview 1 of 1,
October 17, 2011).

One of Lionel's recent adventurous projects was adding a motor to a
beach cruiser bicycle. The gist of the project was simple: using a
low-priced motor he found online, Lionel realized he could mount the
motor to an inexpensive bicycle and make a motorized bike. With one
exception, Lionel was solely responsible for the project from start to
finish.\footnote{Lionel told me that the only part his father helped him
  with was showing him how to use a Dremel power tool to smooth, sand,
  and cut edges (Interview 1 of 1, October 17, 2011).} He bought the
parts, he mounted the motor, and he maintained it as a working vehicle.
But, in practice the project was not straightforward. In the next
subsections I show at length how Lionel addressed the emergent
challenges of modifying\footnote{Often called ``modding'', for short.}
his bike with a motor. My analysis focuses on three features of Lionel's
story that I argue carry over to his programming approach:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  When faced with ambiguity, Lionel worked to decompose large tasks into
  smaller subtasks he could understand.
\item
  Throughout the task, Lionel was metacognitive about the state of his
  design. His constant refrain was ``\emph{how} am I gonna make this
  work?''
\item
  Before committing work and materials to a solution, Lionel created
  intermediate designs whose suitability he could evaluate.
\end{enumerate}

I argue in later sections that these three features of Lionel's
approach---decomposing subtasks, explaining procedures to himself, and
explicitly planning prior to committing time and resources to a final
product---have strong resonances with the approach Lionel took to while
programming.

\paragraph{Lionel had to decompose Step 1, which simply said ``Install
the
Engine''}\label{lionel-had-to-decompose-step-1-which-simply-said-install-the-engine}

Lionel bought an inexpensive motor online with the plan of attaching it
to a cheap bike. As he discovered, though, frugality came with a hidden
cost: poor instructions. Attaching the motor to the bike was not at all
a straightforward process.

Lionel: it was funny. When I, when I got the motor /Mmmhmm/ Um, oh,
it---like I said it was like cheaply made and whatnot and it came with
instructions, but um, no exaggeration at all, step 1 said ``install the
engine''. And that was it. \{Laughs\}

Interviewer: Huh

Lionel: And then step 2 was, um, I guess it was somethin like ``put the
gas tank on.'' It really didn't tell me anything. It was kinda like
``OK, those are the obvious steps, but *how* do I do that? So I kinda
had to figure stuff out on my own. /So/ yup

Interviewer: Did it---how did it end up going? Was it a lot of trial and
error? Like, did they even have pictures of how it was supposed to look
when it /There, there/ was installed?

Lionel: There are maybe two or three pictures, but they were black and
white pictures printed out on like a, I guess like a crappy printer. So
they were kinda hard to see. Um, so I guess my first step was to go look
online at better quality pictures. Unfortunately nobody really---I was
hopin to find you know, like a YouTube video: this is how I did it.

Interviewer: Mmmhmm.

Lionel: I couldn't really find somethin like that, so I just looked at
pictures. (Interview 1 of 1, October 17, 2011).

Lionel recognized that having an attached motor is an obvious goal
state, but that goal state was poorly-specified by ``crappy pictures''
that were hard to interpret. Moreover, the states in between start and
goal were undocumented and not obvious. Lionel's subsequent search for
materials that showed a step-by-step progression came up empty. Faced
with a poorly-specified goal-state and no intermediate guidance,
Lionel's response was to force himself to think carefully about how he
would proceed.

\paragraph{In making modifications Lionel frequently asked himself ``how
am I gonna make this
work?''}\label{in-making-modifications-lionel-frequently-asked-himself-how-am-i-gonna-make-this-work}

Faced with the challenge of interpolating between ``crappy pictures,''
Lionel thought hard about what to do.

Lionel: I spent, I guess the, I guess the first five hours I just spent
kinda staring at it like, ``hmm, *what* am I gonna do?'' Like, ``*how*
am I gonna make this work?'' (Interview 1 of 1, October 17, 2011)

Ultimately, he was struggling with more than just poor instructions.
Even if he did figure out how the motor was supposed to be mounted, he
still faced another problem: the motor was designed for a mountain bike.

Lionel: I had to make a few modifications on my own. Because I guess
it's built for uh, a mountain bike. And the tubes on mountain bikes are,
y'know like *that* thin \{makes circle with right thumb and index
finger\}, but on my beach cruiser there's one tube \{widens thumb and
forefinger\} that's pretty thick where I had to mount the engine.

Lionel: And since the engine mount was built for a motorized bike it was
really small. So I kinda had to like devise my own plan on *how* to,
y'know, add a bigger mount.

Interviewer: Mmmhmm

Lionel: So, that was, that was my first problem. That was, y'know the
first step was to put the motor on I'm thinkin, well, how am I gonna put
the motor on if, y'know the, the specs don't even align. So, I guess
spent y'know five hours just thinkin', like, ``what am I gonna do?''

Lionel: Finally I came up with a plan that seems to be working. So. And
then, I just kinda went from there. I'd put it together, look at the
picture and say ``OK, this piece looks like that piece, and y'know I
guess it looks like it attaches that way, so I'm gonna try that.'' And
just put it together and, y'know if it fit, it fit. If it didn't, then
I'd, y'know I'd be like ``OK, maybe that was wrong. Lemme, lemme work on
somethin' else on it.'' So. (Interview 1 of 1, October 17, 2011)

Asking himself \emph{how} to do something helped bring sub-problems into
focus. Within a larger mod task (viz., adding a motor to a bike), Lionel
in effect entered a nested task: modding a mountain bike engine mount to
fit a beach cruiser. Modding the mount itself was entirely undocumented,
so Lionel's solution in part was to look for affordances in the physical
structure of parts to see what he could attach. Crucially, the
environment afforded Lionel rapid, iterative feedback. It wasn't hard to
tell when a design choice worked because ``if it fit, it fit.''

\paragraph{Lionel used intermediate representations to consider
alternative
designs}\label{lionel-used-intermediate-representations-to-consider-alternative-designs}

Lionel grew increasingly frustrated as he saw how difficult it would be
to mount the motor. Crucially, part of his tactic around that
frustration was to resolve to keep trying and thinking of possible
solutions.

Lionel: So, I guess I had to give myself, like a few hours just to cool
down and then actually think ``OK, well maybe, you know I've already
bought it, so, I'm gonna continue on with it. So I had to think, you
know, OK, what's my next step. *How* could I make this work? You know,
I---I thought of a few different ideas. You know, I'd think of one idea
and think ``OK, would this work? It might, but let me think of another
idea.'' So I'd come up with another idea and say, ``would this work?''
You know, so once I got a few ideas I would just kind of, in my head
picture it. Say, ``OK, this is how it would mount, or this is how it
would mount. Which one would be safer? Or will one of these not even
work?''

Interviewer: Right.

Lionel: So, I guess after, you know, awhile of deciding which one was
better, you know, which one would be easier and safe at the same time
/mmhmm/ so then, that's when I made my decision, and just kinda, went
for it to see if it would work. And luckily it did \{laughs\}.
(Interview 1 of 1, October 17, 2011)

Before committing to a design, Lionel forced himself to evaluate
alternatives first. Naively, he could have gone with the first idea that
came to him. Instead, even if he thought an idea might work he pushed
himself to consider others first. Then, with a collection of possible
designs he evaluated what he saw as their suitability in his head. He
applied criteria as he evaluated, trying to balance the ease with which
a mounting scheme might work with a concern for the safety of the
user---in this case, himself. Again, at least in Lionel's telling, those
steps happened \emph{before} committing to a final implementation. In
other words, Lionel generated and evaluated multiple configurations with
an eye toward both the builder of and user of those configurations
before settling on a path to take.

In sum, I note three key features of Lionel's approach to the design
challenges of modifying his bike. First, he repeatedly asked himself
``how am I gonna do this,'' forcing himself to think in terms of a plan
or procedure to fill in the gaps between the instructional pictures.
Second, he proposed ideas, tried to evaluate their feasibility, and
continued pressing for alternate solutions (``would this work? It might,
but let me think of another solution''). Third, so much of his planning
took place before he ever mounted the motor. By his account, he fiddled
with pieces, envisioned multiple ways the engine could mount, and
evaluated their feasibility before ever taking concrete steps to mount
the motor.

\subsubsection{Lionel's approach to programming resonates strongly with
the design stance he held when working on his
bike}\label{lionels-approach-to-programming-resonates-strongly-with-the-design-stance-he-held-when-working-on-his-bike}

\paragraph{Lionel thinks about steps at a
chalkboard}\label{lionel-thinks-about-steps-at-a-chalkboard}

Later in the interview I asked Lionel to compare his experiences in
Basic Programming with those he was having in Intermediate Programming.
His response revealed a great deal about how he starts projects in
Intermediate Programming: he starts at a chalkboard.

Lionel: Well, {[}Basic Programming{]} I already learned the basics of
the class so I know the basics of where to start my code. And then,
obviously the projects in this class {[}Intermediate Programming{]} are
harder than last class, but I have the basics of, ``OK, well, I know how
a program runs. So how can I translate that into my program, for, for
{[}Intermediate Programming{]}?'' And, yeah, so I guess I'll sit down,
like, yeah, and I'll have like a chalkboard and I'll write things out,
and say OK well first, you know, \{sweeps hands{]} not even think about
the code. Just, in \{air quotes\} English words, so you know, my
project's gonna do this. You know, first I need to make this
calculation, then I'm gonna print this. And then I'm gonna make this
calculation. Et cetera. And I'll sit down and write that out, in the
order that it needs to be done. And then from there I'll go back and
think, ``OK, what's the code, to, you know, make this calculation or
print that?''

Interviewer: So, when you start a project, you don't start at a
computer, it sounds like you kinda start at your chalkboard.

Lionel: Right. Exactly. Yeah, cuz I've tried to start a project at a
computer, and you know I'll write a---I'll start writing a few lines of
code, but then I just really zone in on what I'm doing and kinda lose
sight of the whole picture.

Lionel: So then I'll write this code for, you know, this one little
section of the project, and then after I've done that I'll look back and
say ``Oh, yeah, I need to write the whole project,'' but then this
little code doesn't actually fit into my project, so it's kinda, you
know I guess I get tunnel vision when I start at a computer. (Interview
1 of 1, October 17, 2011)

For Lionel, the chalkboard is a pivotal object in his design activity.
It holds the top-down ``English words'' description of his program. In
so doing, it serves as the earliest durable representation of his
program's intended hierarchy and relationships. That function is
important, because without it Lionel is apt to ``lose sight of the whole
picture'' of what his code is supposed to do. Moreover, as the next
section reveals, the board continues to help Lionel orient his activity
as he digs into the specifics of how his program will work.

\paragraph{Lionel copies pseudo-code into the computer even though it
won't
compile}\label{lionel-copies-pseudo-code-into-the-computer-even-though-it-wont-compile}

As Lionel described the role of the chalkboard, I initially thought
there was a distinct separation between the chalkboard and the computer
as activity objects. The chalkboard, I assumed, was where Lionel
sketched out design plans while the computer was where he wrote code.
When I probed Lionel based on my hunch, I discovered Lionel's activity
wasn't cleanly separable into design on the board / code on the
computer.

Interviewer: Did you, when you start thinking about the code, do you end
up usually filling it in on your board, or will you actually go straight
over to the computer and do it?

Lionel: Um, on my board, I'll do, I guess like
\textbar{}\textbar{}pseudo\textbar{}\textbar{} \textbar{}\{air
quotes\}\textbar{} code /mmhmm/ uh, you know, first on the board it
would only say ``this function calls this function, this function calls
this function.''

Interviewer: Mmmhmm

Lionel: So then on the board I'd write down, ``alright, well this
function'' um, I dunno, for instance I'll, I'll say you know, I'm workin
with a function that checks to make sure that it's on the---that I'm
moving *on* the board between spaces 1 and 24.\footnote{At that point in
  the semester, Lionel's class was working through a Backgammon project,
  so the spaces and moves he's referring to pertain to programming a
  basic Backgammon game.} So then, you know I'll write down like,
pseudocode, write ``OK, well I'm gonna start. I'm gonna have a for-loop
that will go from 1 to 24, and then here I'll, you know I'll have an
if() statement that says if it's between you know, 1 and 24 it's valid,
if not it's invalid.'' So, that'd be my pseudocode, and then when I
actually---I'd, I'd try to write that out for the entire program, or as
much as I could.

Interviewer: Filling out the tree, you mean.

Lionel: Right, fillin out the tree. And then I'd go to my computer, so
that at least I already had a baseline of what to work with. And then,
on my computer I'd, you know I'd write out the pseudocode, I'd ac---I'd
literally write out the pseudocode, even though obviously it wouldn't
compile and actually work.

Interviewer: Mmmhmm

Lionel: So, then, from the pseudocode, as, as I read down on my computer
what the pseudocode said, I'd actually try to write the actual code that
would work. (Interview 1 of 1, October 17, 2011)

Lionel blurred my initial chalkboard / computer distinction in two ways.
First, his overall ``English words'' plan of functions calling functions
evolved on the chalkboard to include computational idioms, including
if-statements and for-loops. But, then, according to Lionel, he would
copy that pseudo-code as-was into the computer, knowing full well it
wouldn't compile. These observations---that Lionel writes computational
pseudo-code on the board, then copies that same code into the
computer---may seem trivial, but I contend they aren't.

Because of the choices Lionel makes, neither chalkboard nor computer has
a distinct, privileged kind of syntax. Rather, ``English words'' and
pseudo-code can peacefully co-exist on the chalkboard, and non-working
pseudo-code gets deliberately typed into the computer before it's
expanded out into ``actual code.'' Put another way, Lionel is capable of
distinguishing between ``English,'' ``pseudo-code,'' and ``actual
code,'' but all three forms of expression are a part of his design
process. It's perfectly acceptable for him to have something other than
``actual code'' exist in intermediate representations of his design.

\subsubsection{Gestural, inscriptional, and verbal in-interview evidence
reveals Lionel's resources for structuring a
program}\label{gestural-inscriptional-and-verbal-in-interview-evidence-reveals-lionels-resources-for-structuring-a-program}

To probe Lionel's retrospective account of how he programs, I gave him
an in-interview task to work on. In this section, I argue Lionel's
development process on that task moved rather linearly through what I
call a ``representational cascade.''\footnote{As I'll explain later, I
  think it would be a mistake to assume that all development activity
  proceeds linearly and top-down in the manner Lionel's does. Not only
  do I think Lionel's clean linearity is rare, I also think it's not
  normative. So, my point here isn't to stress that Lionel moved
  linearly, monotonically through the cascade. Rather, my point is to
  show that he moved between different representational levels \emph{at
  all}.} My schematic for that cascade --- along with the features and
affordances in each layer of the cascade --- is depicted in Figure 3.
What I show here is that actually, the final code Lionel produces is in
fact one representation of a procedure that actually existed in
different forms and modalities before it became the final code in Figure
3.

In the analysis that follows, I explore two of those modalities---verbal
description and written pseudo-code---and explain why each is crucial to
understanding the final produced code. Ultimately, I argue there is a
need to explain \emph{how} Lionel creates and moves fluidly between
these markedly different cascade layers. Specifically, I attempt to
answer the question of \emph{Why does he see the upper layers (verbal
code, hand-written pseudocode, and handwritten source code) as
legitimate activity?} My candidate answer is that Lionel's approach to
programming is in part stabilized by epistemological resources that
allow him to treat the upper layers of the cascade as valid, productive
knowledge expressions in programming. Such an answer is strongly in line
with the findings from sections 3.4.1, where we see Lionel deliberately
making use of other intermediate design representations.

\includegraphics[width=5.74444in,height=4.31111in]{media/image5.jpeg}

\protect\hypertarget{ux5fToc252445968}{}{}Figure ‑ -- An overview of
Lionel's representational cascade and the features and affordances of
each representational layer.

\paragraph{The range-finding prompt and Lionel's final source
code}\label{the-range-finding-prompt-and-lionels-final-source-code}

About 35 minutes into our interview, I gave Lionel the programming task
reproduced below:

Suppose you want to write a program that finds the range of a set of
numbers. So, the program would take 10 numbers as input, then it would
compute the range as the difference between the largest number and the
smallest number.

How would you write such a program? (Interview prompt, October 10, 2011)

The prompt was deliberately vague about certain constraints. It did not
prescribe \emph{how} the program should get input (Should the programmer
expect the user might type it in, or feed in a formatted file?). It also
did not prescribe how the program should provide the result (should the
programmer print the result on screen? Store it to a file? Return the
result to be used by another function?) The reasons for vagueness were
in part because I wanted to see what assumptions students might make
during their design process. Some solutions, I reasoned, might decouple
the procedures of getting input, computing range, and returning output.
Others might simply hard-code a procedure for getting keyboard-based
input. I wanted to prescribe no specific approach up front if I didn't
have to. I have reconstructed his original source code from
screen-capture footage of his in-interview programming:

\includegraphics[width=5.74444in,height=7.11111in]{media/image6.emf}

\protect\hypertarget{ux5fToc252445969}{}{}Figure ‑ -- Lionel's final
source code for the range-finding prompt

Let's explore some top-level features of Lionel's code. First, Lionel
chose to have a user enter text directly using the keyboard (lines
9--12), and his program displays the range result as printed text (lines
28--29). His strategy for finding the maximum and minimum is to
implement a procedure that iterates through the array of numbers and
compares the current number to the globally stored max and
min.\footnote{I think of this as a ``king of the court'' procedure. The
  current value stays on as an extreme if and only if it beats the
  current ``king'' for that extreme. Otherwise, the challenger is
  replaced and the contest resumes with a new challenger until there are
  no more challengers.} If the current number (array{[}i{]}) ``beats''
the global min (by being lower than min), its value replaces that of the
global min. Similarly, if the current number exceeds the global max, its
value replaces that of max. Before the procedure starts, Lionel assigns
the first number of the array as the value for both extrema.

\paragraph{Lionel's verbal description of the
program}\label{lionels-verbal-description-of-the-program}

Lionel chose to start the task on pencil and paper. After scribbling a
line about scanning input into an array (which I discuss in a later
section), Lionel stopped and began explaining out loud to me.

Lionel: so I'm thinkin, my general concept is I'm going to parse through
this array, check each number /Mmmhmm/ and I'll have two variables, min
and max, and, first I'll set both min and max to the first variable in
array, er, yeah first number in array /Mmmhmm/ and, I'll parse through
the rest of the array, checking to see if,

Lionel: Umm, you know if---I'll s---I'll start with max, for example.
/Mmmhmm/ And I'll check to see if the number in that section of the
array is bigger than max. /uh-huh/ And if it is, I'll set max to that
new number. /OK/ And then I'll do the same thing for minimum. And then
at the very end, I'll have my minimum and max and then I'll just
subtract `em.

Interviewer: OK

Lionel: So that's my main concept. (Interview, October 17, 2011)

This verbal description is in some ways bound by the least stringent
rules of expression. If we're hard-nosed about syntax, Lionel's speech
stops and starts (line 5 into line 6), which leans on the unspoken
understanding that ideas in talk are under negotiation. Stopping
mid-sentence and leaving that sentence unfinished is a valid and
acceptable move in conversational space. Moreover, there is a strong
mapping between Lionel's temporal language and the intended
flow-of-control for the program. ``First I'll set,'' ``then I'll do,''
``and then at the very end'' all describe when instructions will be run.

Finally, Lionel's instructions begin at the high level, and in some
cases implementation details are completely left out. For example,
Lionel does not specify precisely how he'll ``parse through the array.''
At this stage of the cascade, it seems sufficient for him to explain
\emph{that} he plans to parse through, but now how. What Lionel
\emph{does} do is gesture while describing ``parsing''. This gesturing,
I argue, creates a concrete representation of iteration and
incrementation --- the gestural marker of a loop.

\includegraphics[width=5.74444in,height=3.23333in]{media/image7.png}

\protect\hypertarget{ux5fToc252445970}{}{}Figure ‑ -- Lionel makes a
cycloid/helix gesture as he says ``parse through this array''

Contrast how Lionel specifies ``parsing'' through the array with how he
describes ``checking'' its contents. First, Lionel explains that he'll
``check each number'' (line 2), decides that he has underspecified how
``checking'' happens (line 5--6), and finally makes ``checking''
relatively more explicit (lines 7--8). Once checking for the maximum has
been made explicit (lines 7--8), Lionel says ``and then I'll do the same
thing for minimum'' (line 9). Line 9 thus becomes the verbal equivalent
of a repeated, parameterized procedure. It is, in a sense, the
talk-space seed of a function. Conversationally, both participants
understand that Lionel saying, ``then I'll do the same thing for
minimum'' (line 9) refers to a procedure defined elsewhere in the
conversation, not a new set of instructions.

\paragraph{Lionel's pseudo-code}\label{lionels-pseudo-code}

Immediately following the verbal description above, Lionel begins
creating what he calls his ``pseudo-code,'' narrating it as he writes.
There are four key features in Lionel's talk (below) that I discuss in
my analysis. Below is an overview of each feature and what importance I
think it has:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Lionel iterates over his initial verbal description by fleshing out
  some implementation specifics --- structures that further specify
  \emph{how} something in a program is to be done.\footnote{It's a
    limitation of my writing style that in this document I use
    ``iteration'' both to describe aspects of Lionel's design process
    and as a formal term of art in computer science. In both cases,
    though, the conceptual meaning is the same: the incremental,
    rule-based set of transitions to the state of an object.} This
  iteration is important because it suggests that his subgoals --- and
  criteria for satisficing them --- change from one micro-coherence to
  the next.
\item
  Reconciling what Lionel \emph{narrates} with what Lionel \emph{writes}
  reveals peculiar mismatches. The mismatches suggests a productive
  capacity: that Lionel can \emph{background} certain details within a
  micro-coherence by omitting them from the written code while still
  maintaining an understanding of how his program will work. Put another
  way: the inscriptional object in a microcoherence captures what Lionel
  thinks is important within that microcoherence, but the inscription
  does not necessarily represent the totality of his conceptual
  understanding.
\item
  At one point in his code, Lionel writes ``same for min,'' which we can
  reasonably infer means he intends to repeat a procedure for
  calculating the minimum. That line encapsulates the precursors of
  functional abstraction: Lionel not only recognizes procedures that
  share the same structure, he can background the specifics (point 2) of
  code he repeats.
\end{enumerate}

Lionel: So it'll be scanf, um, and I'll have my for-loop, and I'll just
have my int-i, and I'll go from zero uh, zero to, four actually cuz
it'll be zero, one, two, three, four, yeah, zero to four. /Mmmhmm/

Lionel: Umm, and I'll have, oh, up top here, to set the, right---right
after the scanf /Mmmhmm/, I'll just say min and max equal the
first---mm, hiccup, sorry

Interviewer: It's OK.

Lionel: \{laughs\} the first number in the array. So I'll say min and
max equal array of zero. And that's just setting them so that I have
something to compare to, I guess. /uh-huh/ I think that will work. Yeah.

Lionel: So, so then, as I parse through my array of five variables, I'll
say ``if min,'' er no, I'll say ``if array of i, you know of, of that,
if, uh, the number in the array that I'm looking at is greater than max,
then max equals that number, array of i'' /Mmmhmm/ and then I'll do the
same thing, same for min. And then at the very bottom, I'll have, uh,
range equals max minus min.

Lionel: So that's my pseudo-code, just so that *I* understand what's
goin' on in my head.

\includegraphics[width=5.75000in,height=3.00208in]{media/image8.emf}

First, Lionel's talk iterates over his initial verbal description by
fleshing out \emph{implementation details} --- programming structures
that further specify \emph{how} something is to be done. Compare here
how Lionel instantiates ideas of iteration in Table 1 below.

\protect\hypertarget{ux5fToc252445959}{}{}Table -- Comparing Lionel's
words and actions across different stages of ``programming'' over the
same putative section of his program. I have bolded some
language-specific syntax.

\begin{longtable}[]{@{}lll@{}}
\toprule
\textbf{Design Coherence} & What he says & What he does\tabularnewline
\midrule
\endhead
\textbf{Verbal Description} & ``I'm going to parse through this array,
check each number'' &
\includegraphics[width=2.22222in,height=1.48889in]{media/image9.jpeg}\tabularnewline
\textbf{Narrated pseudocode} & ``I'll have my \textbf{for-loop}, and
I'll just have my \textbf{int-i}, and I'll go from zero uh, zero to,
four actually cuz it'll be zero, one, two, three, four, yeah, zero to
four. &
\includegraphics[width=2.23333in,height=0.35556in]{media/image8.emf}\tabularnewline
\bottomrule
\end{longtable}

In his narrated pseudo-code, he introduces the C-specific syntax of
``for-loop'' and ``int-i,'' which handle iteration and indexing,
respectively. Moreover, he counts in order to specify the index bounds
of the loop (zero to four).

Second, it's actually somewhat complicated to address the question ``how
do Lionel's inscriptions align with what he says?'' He \emph{says}
``I'll have a for-loop'', but the visual evidence shows that he hasn't
bounded his for-loop body in curly-braces.\footnote{As a language, C
  typically uses curly braces to bound instructions that span multiple
  lines. So, a typical for-loop would be written like this, where the
  loop preamble is bounded by parenthesis and the loop body is bounded
  by curly braces (bolded):

  for(int i=0; i \textless{} maximum\_value; i++) \textbf{\{}

  // Do something

  \textbf{\}}} If he had done so, his for-loop preamble would end with
an opening curly brace; it doesn't (Table 1 graphic above). Also, for
matching, his for-loop should have concluded with a closing curly brace;
it doesn't. Nevertheless, his verbal evidence --- particularly him
saying ``int-i'' --- suggests that he understands details that variables
need to be declared. Together, these mismatches suggest that features
(viz., denoting iteration and setting its bounds) are important enough
to be included in this representation. Other specifics, including proper
declaration of variables, are things Lionel understands but does not
insist on writing.\footnote{One might argue that Lionel is simply being
  sloppy and forgetting to include details, even though he said them.
  The point I'd make here is that even if he is being sloppy, I would
  think it unfair to say, for example, that Lionel has conceptual issues
  with declaring variables. His talk (and many later iterations of his
  design) suggests he understands that declarations are necessary for
  working with variables in C. Moreover, even if he is being sloppy on
  the particular point of variable declarations, the overall character
  of this pseudo-code is still starkly different from that of the full,
  working source code solution he'll ultimately produce. So, even if we
  propose that Lionel has some unintended tolerance --- such as
  sometimes overlooking missing variable declarations --- for assessing
  whether his pseudo-code satisfices, there is still strong evidence
  that his criteria for evaluating his work in pseudo-code stage are
  markedly different from the criteria he'll apply to later iterations
  of design.

  no one would confuse the pseudo-code skeleton here for Lionel's
  fleshed-out, fully-working code later. So, even if Lionel \emph{meant}
  to put them in and forgot, he still judged this pseudocode as
  satisficing his subgoal} Together, and combined with later evidence of
Lionel's successful working program, these observations suggest that
ignoring the details is actually productive for Lionel.

Third, Lionel's talk reveals productive capacities for dealing with
design work. For example, he describes the guts of his program through
brief perspective-taking (lines 14 -- 18 above). As a reminder, Lionel's
procedure is what I refer to as a ``king of the court'' approach: if the
number currently-being-inspected exceeds the reigning champ, the champ
is dethroned and replaced by the number currently-being-inspected. But,
the way Lionel articulates that process actually uses multiple senses of
``I,'' where it's at times actually not clear whether ``I'' means ``I
the programmer'' or ``I the program.''\footnote{To keep your head from
  exploding, I remind you that the lower-case version of ``i'' in
  Lionel's talk actually refers to the subscript for indexing an array.}
For example, the I in ``I'll say `if the array of i'' could very well be
the programmer, where by ``I'll say'' he means ``I the programmer will
write a conditional test.'' But, that phrase is immediately followed by
``if, uh, the number in the array that I'm looking at is greater than
max'', where I believe we may be seeing Lionel taking the perspective of
the program itself.\footnote{My point here is that technically, Lionel
  can never inspect an array element himself; he can only write code
  that does that. You might disagree, in which case we get to have a
  really fun discussion over beer about how we think programming is a
  case of ``distributing cognition'' (Hall, Wright, \& Wieckert, 2007).}
If anything, the seemingly untroubled, interchangeable use of ``I'' here
might suggest that to do the work of programming one can occupy a hybrid
space of both being the program and being the programmer at the same
time.

Lionel concludes this procedure with the invocation ``same for min.''
Picking up the thread of hybridity, this phrase could mean, ``I the
programmer will write a similar procedure for min'' or ``I the program
will perform similar operations to determine min.'' That Lionel leaves
his pseudocode there (``so that's my pseudocode'', line 20) suggests
that whatever perspective it might be written from, this designed
artifact satisfices his pseudo-code production subgoal. And, that
satisfaction is remarkable because if ``same for min'' is an appropriate
place-holder, it suggests Lionel has ways of seeing that procedure as
\emph{similar enough} in structure to the procedure already laid out as
to not require further specification. This is an important point: if a
precursor to functional abstraction is recognizing repeated code, Lionel
has actually recognized repeated code \emph{before ever writing it}. As
we'll see later, he ends up writing this procedure out explicitly, which
means that whatever answer we might come up with as to why Lionel
doesn't abstract this to a function \emph{can't} be that he doesn't see
the repeated computation; to the contrary, repeating core parts of the
max computation was built into his design from the start.

\subsubsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188557}{}{\protect\hypertarget{ux5fToc252445930}{}{}}We
can trace continuities in Lionel across different kinds of design
activities}{We can trace continuities in Lionel across different kinds of design activities}}\label{we-can-trace-continuities-in-lionel-across-different-kinds-of-design-activities}

We can trace patterns in Lionel's design activity by asking questions
about the intermediate design artifacts he creates. Specifically, we can
pay attention to \emph{what} representations he creates, \emph{where}
they live, and \emph{how} he uses them. Table 3 outlines those
relationships for the three contexts discussed: Lionel's bike modding
project, Lionel's retrospective account of his programming practices,
and Lionel's in-interview work on the range-finding task. Table 3 helps
support three key points:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  There is a specific consistency to Lionel's design activity across
  contexts. All activities---given the limits of Lionel's
  self-reports---\emph{involved the generation and use of intermediate
  design representations}. Those representations may at times have lived
  in Lionel's head, on paper, or in electronic format.
\item
  A thorough account of Lionel's in-interview programming activity
  cannot and should not ignore those intermediate representations and
  how he used them.
\item
  Because intermediate representations are a feature of all three design
  contexts, it seems likely that what stabilizes their creation and use
  is tied to something deeper or more global than just conceptual
  knowledge in a domain such as computer programming.
\end{enumerate}

\protect\hypertarget{ux5fToc252445960}{}{}Table -- Comparing Lionel's
design activity across contexts

\begin{longtable}[]{@{}llll@{}}
\toprule
\textbf{Context / Question} & Modding the bike & Retrospective
programming account & In-interview task\tabularnewline
\midrule
\endhead
\textbf{\emph{What} intermediate representations did he create?} &
Possible configurations for mounting the engine & Relationships of
functions to each other; pseudo-code for how functions would work & A
verbal description; narrated pseudo-code; hand-written pseudo-code;
hand-written source code\tabularnewline
\textbf{\emph{Where} do the intermediate representations live?} & In
Lionel's head & On a chalkboard and, later, in the text editor & In
Lionel's talk and gestures; on paper; in the text editor\tabularnewline
\textbf{\emph{How} did he use those intermediate representations?} & To
compare mounting configurations for ease and safety & To ``see the whole
picture'' of a project & To ``understand what's goin on in my head''
when structuring his code\tabularnewline
\bottomrule
\end{longtable}

\subsection{Rebecca's approach to
programming}\label{rebeccas-approach-to-programming}

In this section, I analyze data from a different student --- Rebecca. I
offer this data as an example of what developing design expertise might
look like. If Lionel's approach to programming uses high-level work
(e.g., his pseudo-code) to cascade down the specifics of syntax, it
seems reasonable to ask whether other students use pseudo-code in the
same way. And, if they do not, it seems worth asking how uses of
pseudo-code differ and what that difference might tell us about
students' approaches to programming.

In Rebecca's case, the data I'll present begins with her stuck on the
third of four projects in the course: the ``iTunes'' project.\footnote{In
  this project, students were given a series of text files ---
  representing the user's music collection --- that contained formatted
  albums and song titles. The task was to create a program to read in
  the music information, create a database representation of it, and
  allow the user to transact with that database. Typical transactions
  might be asking for a listing of all the albums in a user's library or
  creating a playlist by allowing the user to choose specific songs.} My
analysis will expand on the following points:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  To understand Rebecca's struggle on the iTunes project and her
  approach to programming during that struggle, we need to understand
  some of her prior experiences in programming. In our fourth interview
  Rebecca recalls a pivotal project experience from her Basic
  Programming course. At the cost of hours of work, half her codebase,
  and a good grade, Rebecca discovered that her sense of how her program
  should work did not match what was expressible in C. I analyze
  Rebecca's recollection because it sits at the intersection of
  self-efficacy and disciplinary practice. Rebecca lost confidence and
  felt ``demoralized'' because her attempts at expressing a procedure
  did not match the constructs and underlying workings of the C
  language. Moreover, the fallout from that experience persisted into
  Intermediate Programming. By our fourth interview she had considered
  dropping Intermediate Programming entirely because of her low grades.
  ``My *biggest* problem'' in programming, she said, ``is I think I have
  the logic, my logic just doesn't transfer to code. Or I don't know
  *how* to transfer it correctly'' (Interview 4 of 5, April 6, 2012).
\item
  Rebecca has the capacity to make sense of program designs. As she
  tries to understand why the instructor would prescribe an array of
  pointers to store track titles in the music server, Rebecca makes
  progress by reasoning about what she knows about pointers and arrays.
  Like Lionel, Rebecca can use gestures, talk, and inscriptions (viz.,
  pseudocode) to express how she might intend for her program to work.
  Rebecca also has a grasp of the conceptual issues --- from a
  computational and programming perspective --- relevant to the part of
  her design she's struggling with. But, Because Rebecca's approach to
  programming so strongly orients toward producing valid, working
  syntax, it may be limiting Rebecca from seeing other ways in which she
  can make progress on her design. In other words, Rebecca \emph{can} do
  things like write pseudo-code to plan, but she doesn't tend toward
  those behaviors when she's stuck. I'll offer an explanation based on
  epistemological dynamics to explain why that's the case.
\end{enumerate}

In what follows, I first present point 1 to give readers a sense of how
Rebecca's prior experiences may inform her approach to programming. I
then outline the particular part of the iTunes project where she feels
``stuck,'' discussing point 2. Before I jump into that analysis, I
\emph{strongly} urge readers who are not familiar with pointers or
dynamic memory allocation to refresh themselves on the subject.

\subsubsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188558}{}{\protect\hypertarget{ux5fToc252445932}{}{}}Rebecca's
struggles on the iTunes project make sense in light of some prior
programming experiences she
had}{Rebecca's struggles on the iTunes project make sense in light of some prior programming experiences she had}}\label{rebeccas-struggles-on-the-itunes-project-make-sense-in-light-of-some-prior-programming-experiences-she-had}

At the time of this study Rebecca was a first-year electrical
engineering major who had gone to high school in rural Maryland. Her
university courses in Basic and Intermediate Programming were the first
experiences she had doing any programming at all, let alone in the C
language. Though she felt academically at ease in high school, a tough
project in her first-semester Basic Programming course shook her
confidence in programming. In the data presentation and analysis that
follows, there are four sub-points I'd like to make about Rebecca's
development in programming.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Prior to the semester I worked with her, Rebecca had difficult
  experience trying to understand arrays on a Basic Programming course
  project.
\item
  That difficult experience had two pieces to it that I believe relate
  to personal epistemology:

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \item
    Rebecca came to feel that her idea for a solution wasn't something
    expressible in C.
  \item
    When she got help from a friend who was ``very smart in
    programming,'' she felt like she typed the code he told her to into
    her project without really coming to understand it.
  \end{enumerate}
\item
  Reflecting on it in our first interview, Rebecca pinpoints the
  experience of that array project as demoralizing and having a lasting
  effect on her confidence in programming.
\item
  Epistemological issues continue to factor into Rebecca's confidence in
  programming. By our fourth interview, Rebecca was getting low grades
  in Intermediate Programming. She had strongly considered dropping the
  course. Trying to articulate her ``biggest problem,'' Rebecca
  explained that she felt like she understood the theory behind the code
  examples in class but had difficulty transferring her logic into code.
\end{enumerate}

\paragraph{Rebecca had a difficult experience on a Basic Programming
arrays
project}\label{rebecca-had-a-difficult-experience-on-a-basic-programming-arrays-project}

In our first interview, I asked Rebecca about how her experiences in
Basic Programming compared to those she was having in Intermediate
Programming. She explained that in that first semester, she ``kinda got
lost on one part'' --- arrays --- ``that was key to the rest of the
year.''

Rebecca: {[}I{]}n {[}Beginning Programming{]} we had three projects, and
the first one, uh, we did---I did discuss it with other students, and we
all worked together kind of. But that one I felt much more like I had a
grasp on everything I was doing. Whereas the second project, I would
stare at the code and be like ``ohmygosh, I don't even understand what,
like, it's asking me to do.'' Like, or how to make it do what it needed
to do, which was to like store in numbers in arrays, change them, and
add `em together and stuff. (Interview 1 of 5, February 10, 2012)

I asked Rebecca if she felt she had a better grasp by the time she
turned in the project.

Rebecca: No. Cuz, there was a friend on my floor, who---very smart at
programming, he's had like years of experience. And he gave me a lot of
pointers and tips doing it, so I felt like I was just using him a lot
instead of actually learning it, which was a downfall for me.

Interviewer: Huh. So, how, I mean. I feel like sometimes you could---you
might ask somebody for help, and then they show you something and then
you might be like oh, you've learned it now /yes/, but it sounds like
you're saying /no/

Rebecca: Just, I don't, like, I don't blame him at all because he helped
me, but, it was kind of more of a, he showed me how to do it, I was like
\{mimes exaggerated typing, in an almost singsong voice\} *OK, sure,
I'll type that code*, but I don't really, I never really, I don't know I
never made the connection. (Interview 1 of 5, February 10, 2012).

\paragraph{Rebecca's ``terrible'' array experience has epistemological
components}\label{rebeccas-terrible-array-experience-has-epistemological-components}

In our second interview a week later when we discussed debugging,
Rebecca returned to that experience in her Basic Programming class:

Rebecca: My array project, my---the second project we did last semester,
that I had a big error. I had to delete like half the code I worked on.
Like, just cuz, like I had said last week, I didn't do very well on
arrays. So.

Interviewer: Oh. Was it cuz of one thing, or was it like a repeated---

Rebecca: It was kind like a general, like, what I was trying to do
didn't work with C.

Interviewer: Do you remember what it was you were trying to do?

Rebecca: I don't know exactly, but uh, I just remember, like, whatever I
was tryin to do, cuz I had my friend help me, he's like ``you can't do
this this way,'' like, it just, C doesn't recognize whatever I was
trying to do.

Interviewer: Huh. So that was like, it sounds like it's almost kind of a
case where your, your plain English of what should happen /yeah/ can't
actually translate---it's not even that you're not sure how it's---

Rebecca: It's like, it was impossible.

Interviewer: It was never---

Rebecca: It was never allowed.

Interviewer: There is no word for that /yeah/ in this language.

Rebecca: I \{laughs\}, that was back then, so. That was always fun.

(Interview 2 of 5, February 17, 2012)

Epistemologically, two patterns stand out in Rebecca's recollections.
First, as Rebecca remembers it, one of her core problems was that her
visions for how her program should accomplish things were ``impossible''
and ``never allowed'' in C. Second, in order to get past that obstacle,
she ultimately typed code a friend told her to without feeling like she
understood it. As this analysis unfolds, I think these two points give
us a crucial starting point to understand Rebecca's practice. The
experience underscored a gulf between ideas Rebecca wanted to invoke and
the syntax she needed to realize them in C. And, her way of dealing with
that gulf was to make an end-run around it by taking and using code she
didn't fully understand.\footnote{I say this without any intended
  judgment (or pejorative connotations) about what Rebecca did. Given
  the assessment structure of the course, where 90\% of a student's
  grade came entirely from whether the program worked to spec, she may
  have felt that getting a good grade by using code she didn't
  understand was preferable to getting a bad grade with code that didn't
  work.}

It's worth noting that whether her friend's code worked to
spec\footnote{Here, I'm using ``worked to spec'' as a shorthand for
  ``worked in such a way that it passed all of the input/output tests
  the instructor would run to determine 90\% of a student's grade.''} is
immaterial to the two epistemological issues I raise. First, if the code
she copied worked, Rebecca would get a good grade, but she wouldn't
understand how her project functioned. And, if it didn't work, she would
get a bad grade \emph{and} she wouldn't understand how her project
functioned. Ultimately, no matter how the code performed, Rebecca would
likely be in a tough spot if she ever had to maintain, revisit, or
refactor the code she copied. Second, no matter how her copied-code
project functioned, it represented \emph{someone else's} articulation of
how a procedure should go. That is, the copied code was an artifact of
someone else bridging the gulf between high-level idea and articulation
in C. Rebecca still found difficulty traversing that gulf.

\paragraph{Rebecca's array project experience had a lasting effect on
her confidence in
programming}\label{rebeccas-array-project-experience-had-a-lasting-effect-on-her-confidence-in-programming}

As that interview drew to a close, I wrapped it up with what had become
standard protocol: asking whether there was anything else she wanted to
talk about.

Rebecca: I think I'm good, I---just

Interviewer: Anything else on your mind?

Rebecca: Sorry I'm not the greatest programmer \{laughs\}

Interviewer: No, hey hey hey. That's \{shakes head\}, first of all, I
mean, um, how come you're apologizing?

Rebecca: I dunno. It's a study, and I'm probably not the best test
subject. (Interview 2 of 5, February 17, 2012)

Rebecca's spontaneous apology spawned a discussion about why she feels
diffident in programming. For her, that diffidence traced its way back
Basic Programming. Feeling lost and stuck on that array project was more
than just a tough experience; it was a turning point:

Rebecca: When I got lost I think is when I started losing that
confidence thing. And then I never really got it back, so.

Interviewer: When was it you started to get lost?

Rebecca: Halfway through last semester, like with the arrays in the
second project. So\ldots{}. I guess it was like, demoralizing, kind of,
and, like, like I don't want to sound cocky or arrogant or anything, but
in high school I did very well in academics. And, so, I never really,
like anything that I had to work for, I worked for for a short period of
time, got it, and like I'd understand it, and if I didn't, like it was
really easy for someone usually to explain it to me. And, I felt like I
didn't really get that after I lost everything on that second project, I
was like, I wasn't able to build myself back up out of that, which
probably stems to why, not as confident in the class. (Interview 2 of 5,
February 17, 2012)

As I interpret how this array project experience may be playing a part
in Rebecca's later approaches to programming, I stress that my analysis
is about Rebecca's perceptions and recollections.\footnote{Due to
  deliberate study limitations I don't have access to the actual code
  from that project. To be as conservative as possible in our data
  collection and respect the privacy of participants, we asked
  participants to grant us access only to folders that would contain
  their Intermediate Programming projects. Outside code/data, including
  code from prior semesters, was excluded from the scope of our code
  snapshot data collection.} It's possible that her ideas for how her
project should work wouldn't have been implementable in any language,
let alone in C. And, while I think that possibility is unlikely, I have
to concede that I only know what Rebecca recalled; I wasn't present for
the experiences she's describing. Nevertheless, the question for my
research is not ``can we know what really happened in that experience?''
Rather, my question is, ``how might Rebecca's \emph{image} of that
experience --- with its specific epistemological facets --- be playing a
part in her approach to programming now?''

Part of the answer to that question may be in her quote above. She talks
about losing her confidence and ``never really {[}getting{]} it back.''
Later, she says ``I felt like I didn't really get that after I lost
everything on that second project.'' Because of the peculiarities of
speech, I had a difficult time parsing Rebecca's statement about losing
everything on that second project. It wasn't clear to me whether there
was a pause between ``get that'' and after, but the meaning of ``that,''
and in turn the meaning of the sentences, depends on the antecedent of
``that.'' So what she said could have two consequentially different
meanings:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  ``I feel like I didn't really get \textbf{that}. After I lost
  everything I lost everything on that second project, I was like, I
  wasn't able to build myself back up out of that.'' In this version,
  ``that'' may mean the topic---arrays---or the project as a whole. I
  might then interpret Rebecca's speech as saying this project was a
  turning point because of the particular topic. Her being ``unable to
  build myself back up out of that'' reflects a kind of debt that might
  have piled up when later assessments depended cumulatively on that
  early topic. Her failure to grasp arrays was thus cumulatively
  punishing because arrays formed a core part of many projects that
  followed.
\item
  ``I feel like I didn't really get \textbf{that} after I lost
  everything on the second project. I was like, I wasn't able to build
  myself back up out of that.'' In this version, ``that'' may mean her
  ready insight into topics or her facility in understanding when
  friends explain them to her. The distinction I draw in this second
  interpretation is ``that'' refers to a kind of continuity in her
  identity---someone for whom school always came easy. The array project
  disturbed that continuity because it was an instance in which things
  no longer came easy to her. And, the experience may have been so
  troubling that she was never able to re-establish being able to easily
  grasp things in programming. The result, then, is that she couldn't
  build herself---in the sense of her continuous identity as one who
  quickly grasps school topics in general and programming topics in
  particular---back up after the array project.
\end{enumerate}

Distinguishing between these possible meanings of Rebecca's speech might
seem pedantic. But, I think the bifurcation is important for theorizing
about learning and instruction in computing. If Rebecca's difficulty was
about the topic of arrays, a plausible narrative is that the difficulty
of that concept---and the assessments that continued to depend on
it---caused a run of bad grades, and the bad grades shook Rebecca's
confidence. But, if Rebecca's difficulty comes from troubling her sense
that she has a ready understanding of programming, the attack on her
confidence is more complex. It's not just that she doesn't get arrays,
it's that she feels she doesn't readily understand other new code she's
seeing in the course---even code that doesn't strongly rely on arrays.
In this view, the confidence Rebecca speaks of isn't just a product of
getting good grades; it's part of a feedback loop in which grades,
confidence, and a feeling of ready insight into programming are all
components.

That potential coupling of grades, confidence, and feelings about ready
insight into code matters because it further pushes the discussion
toward epistemological considerations. It is why, for example, I was
paying such careful attention to Rebecca's recollection of copying down
code her classmate told her. Instead of helping her confidence in
programming by potentially boosting her grade, the copying may have
diminished her confidence because she couldn't understand why someone
else's code worked.

\paragraph{Epistemological issues continued to interact with Rebecca's
confidence in Intermediate
Programming}\label{epistemological-issues-continued-to-interact-with-rebeccas-confidence-in-intermediate-programming}

Rebecca started our fourth interview with a warning to me that she might
``rant'' about programming.

Interviewer: What would you rant about? Go ahead.

Rebecca: Ohh god. Um, I am not doing well \{laughs\} in the class right
now.

Interviewer: OK

Rebecca: And it's not good, so. Yeah, like, we just had our last project
due, which I did a lot better on than the first project

Interviewer: Uh-huh

Rebecca: Which is making me really happy. Um, but, everything about
programming just upsets me, because there's the two guys, like I had
mentioned before who, uh, help me

Interviewer: Mmmhmm

Rebecca: Uh, one of them, he---like I worked every day from the day we
got this project on this project, and my project still didn't turn out
perfect. He wrote his twice in one day because he lost the whole thing.
And, his was perfect. And, I dunno---it just irks me *so* bad that that,
it can be that much of a disparity between people and programming.
(Interview 4 of 5, April 6, 2012)

In her telling, Rebecca wasn't the only person upset by the relative
ease with which this student programmed.

Rebecca: we didn't say it to him, but some of us were like, it's, we
just found it like, so, like aggravating. Like, cuz there's me and
another girl who, on my floor who, she's uh, never had programming
experience either---

Interviewer: Uh-huh

Rebecca: And we're just like it gets so aggravating that it comes easy,
like to him, and that he does it well, but, that's just---he ju---he is
very good at school. He's very smart. So. (Interview 4 of 5, April 6,
2012)

Note here that Rebecca is connecting the idea that programming came easy
to this student to the notion that he was ``very good at school.'' In
this moment, the statement seems to suggest Rebecca has a \emph{fixed
mindset} \{Dweck and colleagues\} about programming expertise. While it
would be naïve to say Rebecca believed --- globally --- that programming
came easily to those who were ``good at school,'' it's nonetheless true
she saw a discrepancy between how easy programming was for her and her
friend and how easy it was for this ``very smart'' classmate.

As we kept talking about this discrepancy, Rebecca offered further
evidence that struggling with programming troubled her academic
identity. In high school, she ``didn't have to study a lot'' because ``a
lot of things came relatively easy'' (Interview 4 of 5, April 6, 2012).
But, her experiences in Intermediate Programming pushed her into a new
position. As she said,

now I'm on the other end and I'm realizing how aggravating it
is\ldots{}. And, I don'---I guess it's just a shock, cuz I do not like
not being good at things \{laughs\}. So...'' (Interview 4 of 5, April 6,
2012).

Again, Rebecca's comments stress the frustration of feeling like
programming comes easily to some people but not to her.\footnote{Or, at
  least, at that moment, in that semester, it did not seem to be coming
  easily to her.} And, they are charged with emotion. She speaks of
being ``on the other end,'' connoting an idea of a structure whose poles
have people who ``get'' programming on one end and those who don't on
the other. It's ``aggravating'' to be on that end, and a ``shock'' to be
there for the first time.

I asked Rebecca to reflect on whether she could ever remember seeing
classmates in the position she was now, where something seemed to come
easily to Rebecca but not to them.

Rebecca: Um. I mean, there was a---there's a friend of mine, who, from
my high school, who, I, I mean just always got math. And, uh, she was
struggling in math a lot.

Interviewer: Mmmhmm

Rebecca: But her attitude was, uh, ``I don't need math. I'm gonna be an
English major.''

Interviewer: OK

Rebecca: So, she didn't even care about it

Interviewer: Oh, OK

Rebecca: Whereas, like, I was like, I'm an electrical engineer major, I
actually need this. So. Like that's the only time I think I've ever felt
like someone's had that feeling that I probably did.

Interviewer: Right.

Rebecca: And, I mean I'm sure there are people that I---uh, that I
didn't know, or didn't realize they had that feeling. But, it's just,
interesting to be on the other end. \{6 second pause\}

Rebecca: Umm. Just. Ugh. I've like, debated withdrawing from the class.
And, like, I've talked to three different advisors or whatever, and
they're like ``don't do it. It's not---you can still bring your---you
can bring your grade up. And even if you don't you have freshman
forgiveness,\footnote{In the interview, Rebecca explained that Freshman
  Forgiveness was a policy extended to students during their first 24
  credit-hours at the university. If a student gets a grade she's
  unhappy with, she can retake that course and her second final grade
  will replace the first.} so.''

Rebecca, unlike her friend, couldn't afford \emph{not} to care about
Intermediate Programming; it was a required course for her electrical
engineering major. But, the combination of feelings she had---consisting
partly of a feeling that programming did not come easily to her---pushed
her to think about withdrawing from the course anyway. And, Rebecca was
not in the habit of withdrawing from classes. Intermediate Programming
was the first and only class where she was experiencing this level of
difficulty (Interview 4 of 5, April 6, 2012).

It would be tempting to think Rebecca struggled in Intermediate
Programming because she didn't work hard enough. But, by her own
assertions she \emph{did} put a considerable amount of effort on
projects. On Project 2, for instance, she said ``I worked every day from
the day we got this project on this project, and my project still didn't
turn out perfect'' (Interview 4 of 5, April 6, 2012). Figure 5 below
shows Rebecca's activity over time from March 5, the day Project 2 was
handed out, to March 28, the day it was due:

\includegraphics[width=5.73333in,height=3.98889in]{media/image10.emf}

\protect\hypertarget{ux5fToc252445971}{}{}Figure ‑ -- Rebecca's
compilation activity over time for Project 2. Each compile a student
initiates creates a commit, so long as there has been a change to the
underlying code. The height of each bar maps to the number of commits
recorded that day. The red line charts \emph{cumulative} commits over
time. The red line is steepest in periods of frequent activity and
shallowest in periods with little or no compile activity.

From the data we can tell Rebecca was compiling Project 2 code almost
every day between March 19 and March 26. Moreover, roughly two thirds of
compilations for the entire project happened in the last 3 days leading
up to the deadline. So, at the very least our data shows she was
actively compiling her code for 8 out the last 10 days.

Rebecca registered no commits between March 12 and March 19. There were
also other days for which no commits were recorded. One possible cause
for those zero values is our data collection system was faulty. A second
possible explanation is that Rebecca was not working. But, a third
explanation is that Rebecca was working but not compiling. This third
explanation actually seems most plausible. It fits the pattern of
Rebecca's work on project 1, where she wrote dozens of lines of code
before ever compiling (Interview 2 of 5, February 17, 2012). It also
accounts for Rebecca's lack of compilations despite her claim that she
worked on Project 2 every day.

So, we can be fairly certain Rebecca was spending time on
projects.\footnote{The specific nature of \emph{what} Rebecca's work
  looked like on Project 2 is the subject of the first study of this
  dissertation, so I don't expand upon it substantially in this study.}
Another possibility for why Rebecca was struggling might be that she
wasn't studying. To address that conjecture, we have to step back to
when I asked Rebecca whether other classes --- classes other than
Intermediate Programming --- had ever pushed her close to withdrawing
from them. She said no.

Rebecca: the thing is, I *don't* know how to like, study for
programming. Cuz like, I look at code and stuff, and I try, and I work
every day on it, but like, even when---I dunno, it just, does not come
naturally.

Interviewer: So, what is it like when you study it? What do you usually
try to do?

Rebecca: Uhh, a lot of times, like, I'll read through the notes we took
in class, like, cuz, what we're doing now---we have a project, uh, we're
doing linked lists, and like, malloc'ing data, which is, uh, *you*
storing it in, as you go. And, like, I get the theory and everything
behind it, and, like it makes sense what it does, but, I just like try
and look at his code examples and stuff, and try and replicate it, but I
don't know what's happening right now, but it's not doing it correctly.
I just, like, my *biggest* problem, I realized, in programming, is I
think I have the logic, my logic just doesn't transfer to code. Or I
don't know *how* to transfer it correctly.

Interviewer: So, um. when you're spending time studying, um, do you
spend most of it on your computer trying to actually program or---

Rebecca: A lot of it is on my computer and a lot of it is also trying to
just look at his code and notes in class. Uh, that's probably about
split, 50/50. Uh, because, sometimes I'll just end up staring at my
computer screen, like, ``what do I even type in to try?'' So, I try and
go look at his examples, so.

(Interview 4 of 5, April 6, 2012)

There are two striking features of this exchange.\footnote{There are
  actually three striking features. The one I left out above is that
  Rebecca describes malloc'ing as \emph{you} storing in data. I think a
  great deal of Rebecca's difficulty with malloc specifically hinges on
  what she means by ``you'' storing it in.} The first is that Rebecca
felt she didn't know how to study for programming. The second striking
feature of this exchange is how Rebecca elaborated on what it might mean
for programming to come naturally to her.

Rebecca feels programming doesn't come naturally to her because she
routinely finds herself unable to transfer her ``logic'' to code. As a
follow-up to Rebecca's answer, I asked what ``logic'' meant to her in
this case.

Rebecca: Like, like how to get the---like, if I need to get a certain
data, like I have a process of how I want to get it, I ju---and I'm
like, ``Oh, you would just do this,'' but can the *computer* do that?
Can I---and it's my---me being able to tell the computer to do that is
where I get lost. (Interview 4 of 5, April 6, 2012)

The tension Rebecca is describing is one I see as strongly
epistemological. First, it's epistemological in the sense that it
reflects a struggle to articulate knowledge of \emph{how to do
something}. Papert (1980) and Abelson \& Sussman (1996) would describe
such a struggle as a pertaining to \emph{procedural epistemology}: the
struggle concerns a particular kind of knowledge work in which the goal
is not to describe what is, but rather how to. The second sense in which
it's an epistemological struggle is closer to how Hammer, Elby, and
colleagues use the word epistemological: what does students' activity
reveal about how they orient toward knowledge and knowing in a
discipline (Hammer et al., 2005; Hammer \& Elby, 2002, 2003; Scherr \&
Hammer, 2009). In this second sense, we can observe that Rebecca defines
her difficulty in terms of the creation of proper code. We can also pose
the question ``if Rebecca struggles to transfer logic to code, what
directs her activity as she \emph{tries} to transfer logic to code?''

Later in the interview, she compared the iTunes project --- on which she
was stuck --- to Project 2.\footnote{Project 2 did not require students
  to use malloc() to dynamically allocate memory. Project 3 did.} In so
doing, she offered another telling idea about the relationship between
her logic and ``syntax'':

Rebecca: Like, cuz, I guess I feel like I did a lot better on the last
project because it was so similar to stuff that I'm used to /Mmmhmm/ um,
whereas all this is brand new and I still don't have any of the syntax
down yet. /Mmmhmm/ Like, if-statements and while-loops, they make sense
to me, whereas I'm still like, trying to grasp at this stuff.

By ``this stuff,'' the context of the conversation suggests Rebecca was
referring to dynamic memory allocation and malloc, which we had been
discussing.

It seems sensible to think if-statements and while-loops might ``make
sense'' to Rebecca because their syntactical form aligns strongly with
their computational operation.\footnote{To borrow terminology from
  Sherin (2001) and rebut Saussure (1986) at the same time, the symbol
  template for an if-statement is strongly connected to its conceptual
  schema because we so often use everyday language in a manner
  concordant with an if-statement's branched control flow. In this
  instance, the relationship between signifier and signified is the
  \emph{opposite} of arbitrary. The similarity between the syntax for an
  if-statement and an understanding of what that statement does is by
  design, because C itself was a designed language.} By that, I mean in
principle the designers of C could have chosen \emph{any} symbol to
represent branched control flow, but they chose a word --- if --- that
invites connections to its meaning in everyday speech. If-statements in
C split control flow based on a Boolean expression's value; while-loops
allow continued iteration until a Boolean expression turns up false.
Both structures have everyday analogues with similar conceptual
properties. A parent might say, ``if you're staying out past 9, let me
know,'' or, ``while we're gone, can you make sure Sadie doesn't eat
anything from the garbage?'' If a child will be home by 8, there is no
need to let that parent know. Similarly, it's reasonable to expect that
once the parents return, the responsibility of keeping Sadie out of the
garbage no longer falls solely to the child.\footnote{Some parents might
  object that family members should \emph{always} be on the lookout for
  Sadie rooting through the trash, that these parents were just
  emphasizing the point. And, reasonable children might in general try
  to heed that. But, what makes computers computers is their lack of
  qualms about abandoning any and all responsibility to keep Sadie from
  the trash once the parents return.} The point of this observation is
that Rebecca may feel certain kinds of programming structures ``make
sense'' not simply because they were covered in introductory
programming, but because something about their syntactical form offers
an affordance for thinking about how they work.

\subsubsection{Rebecca has productive capacities for making progress on
software design
work}\label{rebecca-has-productive-capacities-for-making-progress-on-software-design-work}

In our fourth of five interviews, Rebecca was having trouble making
progress on Project 3, which the instructor described as an ``online
music server.''\footnote{Students in my study colloquially referred to
  this project as the ``iTunes project,'' so ``the iTunes project'' and
  ``Project 3'' all refer to this project.} ``Like, I am so lost, like I
don't even know where to start,'' she said. We discussed why she was
stuck. Then, with my help, we began to work through what Rebecca
identified as a core cause of her confusion. In brief, the assignment
contained a diagram showing the data scheme students were required to
use to store data about the music (Figure 8 below). Specifically, the
scheme required an array of pointers --- each entry of which pointed to
an array of characters --- as a mechanism for storing data about song
titles for the music server. Rebecca's work to resolve her confusion
offers evidence for her productive capacities in designing programs. In
this section, I argue:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Rebecca made sense of the array-of-pointers design\footnote{By
    ``design scheme,'' here, I mean the instructor-provided diagram
    indicating how students were to structure their data for Project 3.}
  using talk and gesture. She was able to explain both why an array of
  pointers \emph{should} exist in C as well as suggest candidate syntax
  for what might create such an array.
\item
  When I proposed we pretend her candidate syntax worked, Rebecca
  articulated through gesture, talk, and written pseudo-code how part of
  her design would incorporate that syntax.
\item
  After the interview, Rebecca resumed work on the project and
  incorporated a version of the design she developed in the interview
  into her code.
\end{enumerate}

\paragraph{Rebecca made sense of the array-of-pointers design using talk
and
gesture}\label{rebecca-made-sense-of-the-array-of-pointers-design-using-talk-and-gesture}

The figure below shows the instructor-provided diagram for how students
should structure their data.

\includegraphics[width=5.36667in,height=4.10000in]{media/image11.emf}

\protect\hypertarget{ux5fToc252445972}{}{}Figure ‑ -- The instructor
required students to use this data arrangement for storing information
in the music server. The scheme uses an array of pointers to represent
the track names of an album.

Rebecca would have seen this diagram before because the project
assignment was given out 11 days prior to our interview. But, the data I
present is the first time I had any access to how she was thinking about
it.

At the time of the interview Rebecca felt stuck on handling issues with
\emph{music albums} --- the top-left and top-right structures depicted
on the diagram. The assignment required that ``all of the data
structures described {[}in this diagram{]}'', including music albums and
all of the data they referenced, ``must be allocated dynamically''
(Instructor-provided Project 3 description). So, Rebecca and I started
discussing what parts of that scheme she understood. Table 3 below
presents what she said and what she did while beginning to explain her
understanding.\footnote{}

\protect\hypertarget{ux5fToc252445961}{}{}Table - Rebecca describes an
array of structures using gestures (Interview 4 of 5, April 6, 2012)

\begin{longtable}[]{@{}ll@{}}
\toprule
What she says & What she does\tabularnewline
\midrule
\endhead
Like, the---the thing that makes sense to me that I, uh, got right now,
is that, to malloc the array, to \{bounds a wide space with her hands\}
make the array a certain length, do that. &
\includegraphics[width=2.03333in,height=2.72222in]{media/image12.png}\tabularnewline
And then I know that each of the \{chops out three invisible boxes in
the air\} I guess nu---characters---I---uh, I'll call `em characters. &
\includegraphics[width=1.37778in,height=3.66667in]{media/image13.emf}\tabularnewline
Like, each of the spots in the array holds a structure and each of the
structures holds three things. & (gestures obscured by the computer
)\tabularnewline
\bottomrule
\end{longtable}

Rebecca's gestures create a kind of virtual object that looks much like
the music albums array in the diagram. As she chops out spaces for each
of the ``characters,'' those spaces are in series and fit --- more or
less --- within the larger area she bounded with her when she said ``to
malloc the array.'' The spatial subdivision is key because it offers
supporting evidence that in some way, Rebecca might have been thinking
about the album structures themselves as being \emph{contained} in or by
the array. And, that spatial containment metaphor resurfaced later when
we discussed the array of pointers, which I'll get to in a moment.

As Rebecca explained her understanding to me, I took to writing down
what I understood her to be saying on paper:

Interviewer: OK. So you have, so here's what you just said. You said
\{writes\} ``malloc the array'' um, ``each element has a structure,''
and then ``each structure has'' you said---

Rebecca: Has three parts

Interviewer: OK. Has three parts. OK.

Rebecca: Yeah. So, and the first part I can do `cuz it's an integer
\{laughs\}. The first part is just the number of tracks on the album.
So---

Interviewer: Oh, oh. So, of the three parts {[}{[}I'm sorry

Rebecca: Yes{]}{]}

Interviewer: I {[}{[}\{untillegible\}

Rebecca: Oh yeah, sorry{]}{]}

Interviewer: So you're saying, um, and then---and then this is number of
tracks, which is an integer, right? /yeah/ OK (Interview 4 of 5, April
6, 2012)

Figure 9 below shows what I wrote as we worked:

\includegraphics[width=5.62222in,height=1.48889in]{media/image14.emf}

\protect\hypertarget{ux5fToc252445973}{}{}Figure ‑ -- I write down what
I understand as Rebecca explains the overall data structure of albums to
me

It wasn't the first time --- either in my series of interviews with
Rebecca or in that interview in particular --- that I tried to write
down what she said. But, I highlight it here because it became part of
my activity as we worked together in this episode. Rebecca would
articulate something, often through talk and gesture, and I would try to
write down and offer back to her what I understood her to be explaining.

As we continued, we landed on trying to understand the ``tracks''
portion of the album data structure. Storing tracks was where Rebecca
felt she was getting ``lost'':

\protect\hypertarget{ux5fToc252445962}{}{}Table -- Rebecca says gets
lost on the array of pointers pointing to track names (Interview 4 of 5,
April 6, 2012)

\begin{longtable}[]{@{}ll@{}}
\toprule
What she says & What she does\tabularnewline
\midrule
\endhead
Rebecca: And, the next one is where I get lost, on the tracks part.
Because it's a pointer \{2 sec pause\} to \{hands make a vertical
cylinder in front of her; note that the array is represented on the
assignment as a narrow, vertical rectangle\}--- &
\includegraphics[width=3.40000in,height=2.62222in]{media/image15.emf}\tabularnewline
it looks like another array that's
\textbar{}\textbar{}pointing\textbar{}\textbar{} \textbar{}\{left hand
crosses from her left to right, at chest height, index finger extended
in the direction of motion; this is the same direction (from Rebecca's
perspective) in which the pointer array on paper points to each track
name\}\textbar{} to the names. &
\includegraphics[width=1.98889in,height=3.38889in]{media/image16.emf}\tabularnewline
\bottomrule
\end{longtable}

Rebecca's gestures for the array of pointers again spatially mimic the
depiction of the array on the page: a vertical cylinder. The reason I
note the orientation of her array gestures is that for practical
purposes arrays as represented by bits in C have no orientation in space
that the programmer could know, much less care about. An array cannot be
parallel to the ground, nor can it be oriented toward the
ceiling.\footnote{The computational hardware representing the array does
  have an orientation in space, but that's clearly not what is at issue
  here. The whole point of the array is to be an abstraction away from
  the soldered transistors, capacitive plates, flipped switches, vacuum
  tubes, tinker toys, or other physical means that store the state of
  the array.} But, the assignment \emph{depicted} the album array as
horizontal, and so too did Rebecca's gestures for it. The assignment
\emph{depicted} the pointer array as vertical, and so too did Rebecca's
gestures for it. In other words, the observation that Rebecca's gestures
align with the assignments pictures might suggest that she is doing more
than thinking about arrays in the abstract. Rather, some elements of the
representational features of the page have made their way into her
activity and, arguably, her cognition.\footnote{One could argue that
  gesture is possibly just undirected flailing, and that the correlation
  between Rebecca's gestures and the page is weak at best --- the sort
  of thing you'd expect to happen some percentage of the time anyway
  under the assumption that the pictures on the page have nothing to do
  with how she's thinking about it. I concede that it's \emph{possible}
  the correlation isn't structurally meaningful. But, I think the
  correlation is meaningful given how many other times her gestures
  align with canonical written representations of computational
  structures and processes.}

Rebecca continued on to try thinking through why an array of pointers
was part of the design.

Rebecca: But, I've always been confused as to why you---I guess, I
always---I was always like, when we learned pointers, I was like ``why
do you need pointers when you could just call it the name? Why do you
need two names for it?'' But, I think, what---at least what I'm seeing
here maybe is like, this is just another array /Mmmhmm/ because you
can't put more than one character in each element of the array, so that
is just an array of pointers that point to strings. And the strings are
the names.

Rebecca: So, that would make sense. So you would---the tracks would
point to the array, and so you could do the pointer of that \{1.5 sec
pause\} The pointer to the array, an array---the array of like 1 would
be a pointer to track 1's name.

Interviewer: Mmmhmm

Rebecca: So. OK, so---see that part makes sense now. I just---no idea
how I would ever access that. Or, store it, I guess, is the better word.
(Interview 4 of 5, April 6, 2012)

What ``confused'' Rebecca has an understandable origin. Most of the work
she would have been doing until that point would have used named
variables. For example:

int age;

float weight;

char *name;

If age, weight, and name are already-named pieces of data, why would
something ever need to refer to them indirectly? It would be as if we
had to call me ``The second-born son of the second-born daughter of
Henry'' for some reason. I already have a name. Why do I need another
name that uses the names and relationships of my forebears to refer to
me when historically my given name has perfectly fine for all
referential purposes?

Rebecca's way of finding sense in pointers was to observe that you can't
put names --- each of which is defined as an array of characters ---
into an array because ``you can't put more than one character in an
array.'' To verify I was understanding her idea correctly, I tried to
restate it:

Interviewer: OK, so you were saying the reason you---when you're tryin
to convince yourself like---you originally you were thinking ``why don't
you just put all the names {[}{[}here

Rebecca: Yeah{]}{]}

Interviewer: And you were {[}{[}saying the reason is

Rebecca: saying, no{]}{]}

Interviewer: because==

Rebecca: ==That makes sense now. Yeah, you can't put more than one
character in an element. And, it's a whole name of a song, so you would
need to point---that element would just \{gesture obscured by my
laptop\} be a pointer to a name. OK. (Interview 4 of 5, April 6, 2012)

The latching and overlapping turns of talk above offer strong evidence
that Rebecca's explanation aligned with my interpretation. And again,
notions of \emph{containment} come into play. Rebecca says, ``you can't
\emph{put} more than one character \emph{in} an element'' in part
because you're talking about ``a \emph{whole name} of a song'' (my
emphasis added here). If Rebecca has a capacity for thinking about the
canonical concept of type-restrictions on arrays, it's manifesting here
as a part-physical metaphor, evoked through both talk (the quote above)
and gesture (Table 4), in the service of making sense of a design
decision.

As she continued, Rebecca discerned that because the input data
contained the number of tracks on each album, she could use that number
to define how long the pointer array needed to be. But, something was
still troubling her.

Rebecca: So. \{4 sec pause\} That would \{6 sec pause\}

Interviewer: What're you thinkin?

Rebecca: Just that, um, I'm just tryna think, cuz like that makes sense,
but I don't know how to get to that. Cuz, how do you make a pointer
point to a random array that you just made? Uh, I guess \{4 sec pause\}

Rebecca: Uh, p---the pointer syntax I need---I would need to go back and
look at them because now you have an array, each array---my issue---cuz
like now that makes sense. Like, it's just how do I get to that point of
making that work?

Interviewer: {[}{[}So

Rebecca: And this is what{]}{]} I mean by my logic /OK/ I get the logic
and the theory behind it but I don't know how to actually put it into C.

Interviewer: OK, so in other words, what makes
\textbar{}\textbar{}sense\textbar{}\textbar{} \textbar{}\{air
quotes\}\textbar{} to you know /Mmmhmm/ that I guess didn't before is
that *that* \{gestures to the instructor's structural diagram on
screen\} is a---that's an understandable arrangement /yes/ for the data

Rebecca: Yes.

Interviewer: But then, how do I {[}{[}make it

Rebecca: make it{]}{]} Yeah \{nods\}. (Interview 4 of 5, April 6, 2012)

When Rebecca said ``I don't know how to get to that,'' I take ``how to
get to that'' to mean \emph{how to write the code for that procedure}.
We had just agreed that design-wise, she could use a number provided in
the input data to define the length of the pointer of array. But, she
was stuck because of her difficulty translating the ``logic'' we
developed into code.

From the outside, what Rebecca and I had done together was something I
think crucially important to the design of a program. We worked to
understand how someone else chose to structure their data.\footnote{That
  the ``someone else'' was in this case the instructor is important for
  larger power dynamic considerations.} We tried to decide for ourselves
whether and how those structures made sense. Along the way we created
ephemeral objects of that sense-making in the form of talk and gesture
as well as the durable artifacts of my inscriptions. But, none of that
work resulted in C code.

For Rebecca, the lack of certainty about \emph{how to make} our ideas in
C code was a marked concern. And, again, that concern makes sense. We
had just convinced ourselves that we could understand \emph{why} an
array-of-pointers. We even figured out \emph{how} we could use the input
data to provide the length of the arrays we would need to make. But we
had not laid out how to declare the array \emph{and} preserve a
reference to it (``how do you make a pointer point to a random array
that you just made?). Without that piece --- a piece Rebecca was unsure
of and felt she would need to look up --- our solution clearly wouldn't
work.

Absent that piece, Rebecca tried thinking of possible ways to fill in
the missing syntax.

Rebecca: Cuz, I'm like, like I'm trying to think right now of things
that I would try /Mmmhmm/ but I don't know what I would try first
because, all I'm thinking right now is that I know tracks is going to be
a pointer. /Mmmhmm/ So you make that a pointer.

Interviewer: OK.

Rebecca: But I'm not sure where it would point to, because you don't
have that array made yet. So I \{3 sec pause\} so you'd have to, I guess
you could make the array, and then make it point to the array, but \{3
sec pause\} Cuz you have to make an array of pointers.

Interviewer: Mmmhmm

Rebecca: And I don't know how to do that. Because I've made arrays of
characters and integers and stuff before /OK/ but never pointers, so.
(Interview 4 of 5, April 6, 2012)

Rebecca was seeing a temporal problem: how could she reference an array
that didn't exist yet?\footnote{As a crude analogy, imagine trying to
  write your own will. If you think forward in time, you might later
  have children to whom you'd like to bequeath your assets. But how can
  you enumerate those children in the will now, before you've had them?
  You need your will to refer to something that does not exist at the
  time you write the will. Extrapolating forward and rather ludicrously,
  we could imagine you are always capable of producing children. So, up
  until the day of your death, there is always the seeming risk that
  your will might need to refer to a beneficiary who does not yet exist.
  Recognizing this subtle problem as Rebecca did is an act I consider to
  be sense-making, even though she as-yet still did not have a solution.}

Rebecca started exploring ways to solve that problem of getting tracks
to point to the array of pointers.

Rebecca: But I'm not sure where it (tracks) would point to, because you
don't have that array made yet. So I \{3 sec pause\} so you'd have to, I
guess you could make the array, and then make it point to the array, but
\{3 sec pause\} Cuz you have to make an array of pointers.

Interviewer: Mmmhmm

Rebecca: And I don't know how to do that. Because I've made arrays of
characters and integers and stuff before /OK/ but never pointers, so.
(Interview 4 of 5, April 6, 2012)

Rebecca landed on a potential solution: make the array first, then make
tracks point to the array. But, with that solution came another problem:
how do you make an array of pointers?

\paragraph{Rebecca could propose candidate syntax and build pseudocode
around
it.}\label{rebecca-could-propose-candidate-syntax-and-build-pseudocode-around-it.}

Rebecca came up with an idea to create an array of pointers, but she
wasn't sure if it would work.

Rebecca: if I were to d---when I declared the array, if I were to---I
dunno if you even can, like, whenever we---we do ints, and then it would
be like star-p /Mmmhmm/ and then star-p would be the pointer /ok/ I
don't know if you can do like int star-p
\textbar{}\textbar{}bracket-bracket\textbar{}\textbar{}
\textbar{}\{makes square brackets with hands\}\textbar{} and that would
be an array or not.

Interviewer: {[}{[}OK

Rebecca: But I don't know{]}{]} if you can do that.

Notice what's happening here. Rebecca is taking patterns that she knows
work:

int age; // creates an integer variable called age

int ages{[}35{]}; // creates an array called ages,

// 35 integers long

int *p; // creates a pointer-to-an-integer;

// said pointer is called p

and considering a pattern that might work:

int *p{[}35{]}; // Rebecca is not exactly sure what this will do

Notice further that in this exchange, Rebecca wondered ``if you can do
that.'' Determining the referent of \emph{that} was, again,
consequential. One possibility was Rebecca meant ``int *p{[}35{]}''
might not be the proper syntax for creating an array of 35
pointers-to-ints. If so, Rebecca was struggling with her aforementioned
problem of not being able to translate her logic to code. Another
possibility was Rebecca meant ``I don't know if there exists \emph{any}
syntax in C to create an array of pointers.'' So, I followed up on this
latter interpretation. We both agreed it \emph{should} be possible to
create an array of pointers, it was indeed just a question of whether
that syntax accomplished it (Interview 4 of 5, April 6, 2012). So, that
left the former interpretation: it looked like Rebecca was unsure how to
``transfer her logic to code.''

We both agreed creating an array of pointers should be possible. So, I
asked Rebecca to suppose her candidate syntax worked. She readily
agreed.

Interviewer: OK. So then, um, what if we just pretended for a minute,
that that, like==

Rebecca: ==That that works==

Interviewer: ==That that worked==

Rebecca: ==OK==

Interviewer: ==OK. So then, um. \{begins writing\} So then you might
write like to make an array of \{stops writing\} actually what would we
call this? This is==

Rebecca: ==array of pointers, I guess==

Interviewer: ==OK. \{under breath\}
\textbar{}\textbar{}pointers\textbar{}\textbar{} \textbar{}\{writes
``pointers''\}\textbar{}. It'd be, well. So. The---the one we had was
like \textbar{}\textbar{}int, star p, um, it would be some
number\textbar{}\textbar{} \textbar{}\{writing\}\textbar{}, um, and I
guess that'd be it in order to just declare /Yes/ that right? OK.

In this exchange note every turn boundary is latched. Once I bid for
pretending, Rebecca accedes. When I'm unsure what to call the array,
Rebecca finishes my sentence. Rebecca rapidly approves the final product
before I even finish my last sentence. Figure 10 below is the syntax I
wrote in the exchange.

\includegraphics[width=5.12500in,height=1.09792in]{media/image17.emf}

If that syntax created the array-of-pointers, the next challenge for
storing track names was how to access elements in array-of-pointers.
Figure 11 presents the pseudo-code Rebecca ultimately wrote to scan in
and store track names.

\includegraphics[width=5.36667in,height=1.30000in]{media/image18.emf}

\protect\hypertarget{ux5fToc252445974}{}{}Figure ‑ -- Rebecca's
pseudo-code for scanning in and storing track names

The next section will show the rich talk and gestures that accompanied
Rebecca's generation of this pseudo-code. I argue activity in the form
of gesture and talk constituted and sustained Rebecca's in-the-moment
approach to programming.

\paragraph{Rebecca's talk and gestures constituted and sustained her
in-the-moment approach to
programming}\label{rebeccas-talk-and-gestures-constituted-and-sustained-her-in-the-moment-approach-to-programming}

Throughout this section I detail the speech and gestures that accompany
Rebecca's pseudocode production. Along both modalities I saw evidence of
a marked shift in Rebecca's orientation. It was a shift from being
hunched, silent, thinking over the keyboard to being animated, open, and
gestural toward me. It was also a shift from seeming diffident,
uncertain, and hedging in speech to seeming confident, assured, and able
to dispatch my questions. Taken in their totality across the episode,
these subtleties constitute evidence of an approach that strongly
differs from those Rebecca had when discussing her struggles in the
course (section 3.5.1).

First, Rebecca suggested she could write the pseudo-code as we talked.

\protect\hypertarget{ux5fToc252445963}{}{}Table -- Rebecca starts
talking out pseudo-code and asks for the pen to write it

\begin{longtable}[]{@{}ll@{}}
\toprule
What she said & What she did\tabularnewline
\midrule
\endhead
Uh, well, what I was think---like if I wanted to like,
\textbar{}\textbar{}save it\textbar{}\textbar{} \textbar{}\{pinches
right hand, thumb and index finger a few inches apart\}\textbar{} or
whatever, /Yeah/ &
\includegraphics[width=2.68889in,height=3.23333in]{media/image19.png}\tabularnewline
I could make a while loop, \textbar{}\textbar{}scan in
the\textbar{}\textbar{} \textbar{}\{drags hand across the
table\}\textbar{} ---scan in the data from the album, /OK/ \{interviewer
begins writing\} uh, which is, {[}{[}I can write it if you want &
\includegraphics[width=2.66667in,height=3.23333in]{media/image20.png}\tabularnewline
\bottomrule
\end{longtable}

Rebecca then took the pen and started describing her thinking while she
wrote.

\protect\hypertarget{ux5fToc252445964}{}{}Table -- Rebecca's
verbal/gestural overview of looping through the input track titles
(Interview 4 of 5, April 6, 2012)

\begin{longtable}[]{@{}lll@{}}
\toprule
Panel \# & What she said & What she did\tabularnewline
\midrule
\endhead
1 & And then my thinking at least, is you should be able to, um, say
that

``star p of i'' /mmhmm/ equals, uh, the title, and then you just do i++,
so then it'll \textbar{}\textbar{}move to the next
one\textbar{}\textbar{} \textbar{}\{makes looping gesture with left
hand\}\textbar{} /OK/ &
\includegraphics[width=2.03333in,height=2.33333in]{media/image21.png}\tabularnewline
2 & and you just keep \textbar{}\textbar{}saving each of the
pointers\textbar{}\textbar{} \textbar{}\{left hand makes horizontal
chops in the air, like rungs down a ladder\}\textbar{} &
\includegraphics[width=3.40000in,height=1.92222in]{media/image22.emf}\tabularnewline
3 & in the array \textbar{}\textbar{}to a title\textbar{}\textbar{}
\textbar{}\{right hand makes pinching motion, left-hand points to
it\}\textbar{} &
\includegraphics[width=2.18889in,height=2.24444in]{media/image23.png}\tabularnewline
4 & \textbar{}\textbar{}And\textbar{}\textbar{} \textbar{}\{right hand
makes a large loop counterclockwise (from Rebecca's
perspective)\}\textbar{} &
\includegraphics[width=3.13333in,height=1.54444in]{media/image24.emf}\tabularnewline
5 & \textbar{}\textbar{}you just increment by 1\textbar{}\textbar{}
\textbar{}\{right hand makes a cycloid to her right, in a plane parallel
to the back wall, like counting dots on a line\}\textbar{} until you
reach the end of file. &
\includegraphics[width=1.06667in,height=3.62222in]{media/image25.emf}\tabularnewline
\bottomrule
\end{longtable}

``Like, that would make sense to me,'' Rebecca said. I asked whether the
while loop would get a fresh line of text when it runs again.

\begin{longtable}[]{@{}lll@{}}
\toprule
Panel \# & What she said & What she did\tabularnewline
\midrule
\endhead
1 & Yeah, because{]}{]} uh, \textbar{}\textbar{}what the while loop
does\textbar{}\textbar{} \textbar{}\{pinches right thumb and
forefinger\}\textbar{} &
\includegraphics[width=1.74444in,height=1.95556in]{media/image26.png}\tabularnewline
2 & \textbar{}\textbar{}is it reads in the line\textbar{}\textbar{}
\textbar{}\{right hand, palm down, chest height, slides out to the
right\}\textbar{} &
\includegraphics[width=1.76667in,height=2.83333in]{media/image27.emf}\tabularnewline
3 & and then \textbar{}\textbar{}once it reaches\textbar{}\textbar{}
\textbar{}\{right hand rises, lowers in a chop\}\textbar{} &
\includegraphics[width=1.78889in,height=2.11111in]{media/image28.png}\tabularnewline
4 & the end of \textbar{}\textbar{}character it'll\textbar{}\textbar{}
\textbar{}\{right hand rises, swoops down in a crescent, index finger
extended, looping back up\}\textbar{} &
\includegraphics[width=2.74444in,height=1.36667in]{media/image29.emf}\tabularnewline
5 & go back down---it'll \textbar{}\textbar{}do the while
loop\textbar{}\textbar{} \textbar{}\{right hand moves back down toward
table\}\textbar{} \textbar{}\textbar{}and then\textbar{}\textbar{}
\textbar{}\{right hand moves back up above shoulder-height,
palm-down\}\textbar{} &
\includegraphics[width=2.63333in,height=1.55556in]{media/image30.emf}\tabularnewline
6 & \textbar{}\textbar{}go back\textbar{}\textbar{} \textbar{}\{right
hand creates a loop down and back up\}\textbar{} down to the next
thing---line, &
\includegraphics[width=3.38889in,height=1.31111in]{media/image31.emf}\tabularnewline
7 & it'll \textbar{}\textbar{}read in the line\textbar{}\textbar{}
\textbar{}\{right hand scans rightward at shoulder height\}\textbar{} &
\includegraphics[width=1.61111in,height=2.47778in]{media/image32.emf}\tabularnewline
8 & \textbar{}\textbar{}until it\textbar{}\textbar{} \textbar{}\{pinches
right thumb and forefinger, swipes right hand back to the left, palm
down\}\textbar{} reaches the end of file &
\includegraphics[width=3.38889in,height=1.45556in]{media/image33.emf}\tabularnewline
\bottomrule
\end{longtable}

In panel 4, notice that Rebecca's gesture of swooping back actually
\emph{precedes} her saying ``go back down.'' In other words, her body
articulates the idea of cycling back before she actually speaks it. A
similar argument could be made for panel 8, where Rebecca's gesture for
a process hitting the bottom (hand swooping and hitting an inflection
point) precedes her saying ``reaches the end of file''

I tried to sum up my understanding of Rebecca's description:

Interviewer: So as you step through this loop /mmhmm/ \{points to i++\}
i keeps going up by one==

Rebecca: ==Yes. (Interview 4 of 5, April 6, 2012)

As I thought about that, Rebecca continued explaining and gesturing:

\protect\hypertarget{ux5fToc252445965}{}{}Table -- Rebecca concludes her
visual and gestural explanation for scanning in track titles (Interview
4 of 5, April 6, 2012)

\begin{longtable}[]{@{}lll@{}}
\toprule
Panel \# & What she said & What she did\tabularnewline
\midrule
\endhead
1 & And the \textbar{}\textbar{}lines keep going
down\textbar{}\textbar{} \textbar{}\{left hand horizontally chops the
air, creating ``ladder rungs''\}\textbar{} & (Gesture partially obscured
by computer )\tabularnewline
2 & so, \textbar{}\textbar{}that way,\textbar{}\textbar{}
\textbar{}\{left hand raises up to her head, palm down and parallel to
the table, and it is replaced by right hand, index finger extended and
pointing to her left\}\textbar{} &
\includegraphics[width=2.10000in,height=2.60000in]{media/image34.emf}\tabularnewline
3 & \textbar{}\textbar{}the first line\textbar{}\textbar{}
\textbar{}\{right hand scans across to her right\}\textbar{} &
\includegraphics[width=1.57778in,height=3.50000in]{media/image35.emf}\tabularnewline
4 & is going to \textbar{}\textbar{}be\textbar{}\textbar{}
\textbar{}\{right hand rotates to become pinched thumb and forefinger,
palm out\}\textbar{} , \textbar{}\textbar{}uh the
p---\textbar{}\textbar{} \textbar{}\{right-hand wiggles\}\textbar{} &
\includegraphics[width=3.25556in,height=1.62222in]{media/image36.emf}\tabularnewline
5 & element zero /OK/ uh \textbar{}\textbar{}the
second\textbar{}\textbar{} \textbar{}\{hand swoops slightly up, then
curves down and locks in in a position below the prior one\}\textbar{}
one'll be element one &
\includegraphics[width=3.25556in,height=1.55556in]{media/image37.emf}\tabularnewline
\bottomrule
\end{longtable}

\subsection{Conclusion}\label{conclusion-1}

In pulling together concluding ideas, I revisit the points I established
at the end of section 3.1.

\subsubsection{Students' early-stage design activity reveals patterns
outside the explanatory scope of (mis)conceptions
accounts}\label{students-early-stage-design-activity-reveals-patterns-outside-the-explanatory-scope-of-misconceptions-accounts}

Lionel's strategy of working at a whiteboard, keeping himself at a
top-level of planning, and copying pseudo-code into a computer isn't
explained by appealing to ``concepts'' in computer science, which in
many research accounts are simply mappings of content in computer
science curricula. I demonstrate this gap in Appendix 4, where the
``conceptual features'' of Lionel's code reveal little (if anything)
about the resources involved in Lionel's design process.

Rebecca, in a similar fashion, has lots of productive knowledge for
thinking about arrays (section 3.5.2), but at the time of the interview
none of her code reflected that knowledge. Moreover, Rebecca was able to
sense-make (Danielak et al., in press) about an array-of-pointers
design, where her sense-making was again a complex activity that would
be poorly accounted for in CSEd frameworks that appeal to
misconceptions. In Rebecca's case, the situation is particularly
paradoxical if we consider conceptual knowledge about a topic to naively
be something students have or don't have.

Suppose Rebecca had the requisite conceptual knowledge for thinking
about an array of pointers. It's a puzzle, then, to explain solely via
conceptual knowledge why she was so frustrated with programming and felt
stuck when she began Interview 4. Why hadn't the knowledge she
possessed---relevant to solving the problem at hand---manifested already
some way? Why was she stuck?

The alternative is to assume that as of Interview 4 she didn't have the
conceptual knowledge for thinking about the problem at hand. But, that
doesn't make sense either. If Rebecca didn't know how to think about an
array of pointers, why was she able to do it so well during the
interview? Both possibilities---that either Rebecca had or did not have
conceptual knowledge about an array of pointers---lead to fairly
non-sensical conclusions under naïve have/don't have assumptions of
conceptual knowledge. Yet, again, the majority of conceptual knowledge
research in CSEd today is silent on the issue of degree when it comes to
conceptual knowledge: students either have a mental model that matches
canonical function or they don't.

\subsubsection{Rebecca had epistemological resources to support
expert-like practices, but she framed those practices
differently}\label{rebecca-had-epistemological-resources-to-support-expert-like-practices-but-she-framed-those-practices-differently}

Consider these summative statements made by Lionel and Rebecca on how
they use pseudo-code in their work.

\protect\hypertarget{ux5fToc252445966}{}{}Table -- Comparing Lionel's
and Rebecca's views toward pseudo-code

\begin{longtable}[]{@{}ll@{}}
\toprule
Lionel & Rebecca\tabularnewline
\midrule
\endhead
``on my computer I'd, you know I'd write out the pseudocode, I'd
ac---I'd literally write out the pseudocode, even though obviously it
wouldn't compile and actually work.'' (Interview 1 of 1, October 17,
2011) & ``So, I can't just type in ``if the white piece reaches in,''
I---the, the C language, I guess, putting it in those, their terms,
their terminology into programming language.'' (Interview 2 of 5,
February 17, 2012)\tabularnewline
\bottomrule
\end{longtable}

As illustrative examples, they capture a fundamental difference in
practice between Lionel and Rebecca. Lionel worked in pseudo-code on his
computer, even going so far as to copy it into his source-code files
despite the fact that it ``obviously wouldn't compile and actually
work.'' Rebecca also wrote pseudo-code, and she even wrote it into her
source code files. But, her local sense of the activity of writing
pseudo-code was different. It came to represent a discontinuity with the
final code she was trying to write because there was nonetheless a gap
between writing in English and writing C in ``their terms.''

These diverging stances take the same practice---in this case,
pseudo-coding---and situate it within a different kind of
epistemological coherence. For Lionel, the coherence associated with
pseudo-coding is productive and optimistic. Copying non-working
pseudo-code into the computer is, in a colloquial sense, all part of the
plan. It's a legitimate step toward creating a final working program
and, crucially, it reflects his own understanding of what's supposed to
happen in his code. For Rebecca, by contrast, writing pseudo-code is a
fallback. Pseudo-code is what she writes ``if I know what I want to put
underneath of it,'' by which she means if she's not sure how to flesh
out a loop or other control structure. Rather than being part of a
planful coherence, the practice of pseudo-coding for Rebecca gets
triggered as part of a stopgap coherence; a contingent measure at times
associated with a kind of diffidence.

Crucially, Rebecca has intellectual resources for pseudo-coding.
Moreover, her practice of writing pseudo-code to flesh out a loop is the
plausible beginning of ``method stubbing,'' the process by which one
might defer implementing the guts of a function and instead create a
simple stopgap. For example, the function below will always print 28
Fahrenheit, which is fairly useless as far as temperature-getting
functions go:

getOutsideTemperature \textless{}- function(sensor) \{

print(``28F'')

\}

But, the advantage of having \emph{something} in the function body, no
matter how trivial, is enormous. Now, because
\textbf{getOutsideTemperature} is defined and takes arguments, other
functions can safely call it. It also offers a diagnostic output---by
printing ``\textbf{28F}'' out---that we can rely on if we ever need to
make sure the function was called. At the time of creating
\textbf{getOutsideTemperature}, a programmer may have no idea how the
temperature will be obtained but will nonetheless need to be able to
write code that depends on it getting the temperature. Stubbing is, in
such cases, a highly productive thing for a programmer to be able to do.
But, stubbing is as much a strict practice as it is an epistemological
move, because it represents a decision that an entity-to-be-known can
be, for the moment, underspecified so it can be incorporated into a
system before its behavior is defined.

Rebecca is already capable of simple stubbing. It's part of an
epistemological coherence that's triggered when she doesn't know how to
flesh out a control structure. So, from a constructivist standpoint
(Smith et al., 1993), she has knowledge of something that can be further
refined into software engineering knowledge. Indeed, much of the
surrounding data from Rebecca suggests that she would have been served
well if an instructional intervention had taught her some simple ways to
stub out functions so she could call them even if she hadn't defined
their bodies yet. To my knowledge, she was never exposed to stubbing
formally, which is unfortunate considering how much it might have led to
improving her confidence in programming.

Lastly, consider Rebecca's episode in reasoning about an array of
pointers (section 3.5.2). As I show, Rebecca clearly had resources for
reasoning about the sensibility of an array-of-pointers design and for
predicting array-of-pointers syntax by extrapolating from patterns she
had seen before. And, when I explicitly suggested that we suppose a
candidate syntax works, Rebecca was able to fluidly articulate the
procedure she would build around that syntax (section 3.5.2.3). Rebecca
knew things, but her knowledge was deployed in such a highly contextual
way as to be sensitive to what kind of knowledge-activity she thought
she was supposed to be doing. It's thus sensible to model that set of
phenomena from an epistemological standpoint, rather than a strictly
conceptual one. Rebecca had certain kinds of knowledge, but part of my
in-interview intervention was to establish a frame where \emph{supposing
syntax worked} was a valid part of programming. Once we mutually
negotiated that frame there was a strong microcoherence of Rebecca
stably explaining her code. But, crucially, my intervention wasn't about
introducing conceptual content but rather establishing a kind of
knowledge game---supposing---that one could play as part of programming
design.

\subsubsection{Students displayed a diversity of approaches to
programming in the
moment}\label{students-displayed-a-diversity-of-approaches-to-programming-in-the-moment}

The diversity of programming approaches I saw students take is far
greater than just the space spanned by Lionel and Rebecca. Within and
across my data, I continually observed students saying and doing things
that not only distinguished them from each other, but constituted
phenomena I haven't seen described in prior CSEd research. I saw
students employ clever tests for debugging, create their own makeshift
debugger, and discuss at length whether certain programming constructs
represented natural ways of thinking. For illustrative purposes, I'll
discuss three such examples. While I don't analyze them at length, I
include them here as proof of principle of the kinds of phenomena we're
not currently capturing that nonetheless have a strong effect on
students' programming approaches. Moreover, these ``in the wild''
(Hutchins, 1995a) practices are further evidence of the kinds of
knowledge that could be further refined into student expertise (Smith et
al., 1993). So, not only do our research accounts miss these kinds
phenomena, by missing them they preclude research and practice from
building off them.

\paragraph{Isaac used a thoughtful debugging strategy that code
snapshots alone could never
capture}\label{isaac-used-a-thoughtful-debugging-strategy-that-code-snapshots-alone-could-never-capture}

When Isaac was debugging an error in his checkers game code, he made use
of a clever test to discover he had reversed array index subscripts. He
had become uncertain about which of the two subscripts corresponded to
what he thought of as the x-coordinate on the board and which controlled
the y-coordinate. So, he took a piece whose x-y position he knew and
instructed his program to print it as a ``7'' on the board. Then, he
told it to increment the first index of that known piece by 1, leave the
second index alone and print the resulting piece on the board as ``8''.
Finally, he told his program to increment the second index of the known
piece by 1, leave the first piece alone, and print the resulting piece
on the board as ``9''. By inspecting the visual output, Isaac could tell
which subscript controlled which position because he knew where 8 and 9
were relative to the unchanged 7. When he reflected on his work in the
interview, he explained that what was tricky was realizing that the
coordinate system for the checkers board didn't work the way standard
coordinate systems in math worked; in the checkers program the indices
were reversed.

Because Isaac had been set up as a code-snapshot participant, we have a
snapshot record of his work for that portion of the interview. But, the
only snapshot-visible changes are the reversal of the indices and the
accompanying test cases. The narrative surrounding Isaac's particular
debugging text, including his reflection that coordinate systems in
programming are ``trickier'' than those in math, is entirely invisible
to the snapshot record. If I hadn't interviewed him I never would have
captured it.

\paragraph{Dana created her own debugging
environment}\label{dana-created-her-own-debugging-environment}

Dana was working on her checkers program and needed to debug some errant
behavior. In class and in discussion students had been exposed to the
GNU Debugger (GDB), so in the interview I asked Dana if she'd like to
use it. She said she didn't feel comfortable using GDB. Instead, using
my MacBook, Dana decided to create three separate instances of a
terminal. In one instance, she had her checkers program source code
open. In the second, she had the source code for a set of test inputs
she was developing to try to pinpoint the error. The third window was
dedicated to compiling and running the code. Since the program was
designed to print an ASCII (text) representation of the board at each
turn, the dedicated program window was Dana's visual output for which
pieces were where at any state in the game.

The important finding from this episode is how Dana deliberately
structured her environment to support her activity. At the lowest level,
she could have tried to do everything in a single terminal window. But,
if she did that she'd have to constantly switch contexts. Because
students in the course used a terminal version of emacs, she had no easy
way to simultaneously view her code while compiling it. GDB would have
been in many respects the right tool for that problem: it would let her
set breakpoints, step through iterations of code, and inspect variable
values as she did so. But Dana chose not to use GDB.

Instead, Dana exploited the fact that Terminal.app on Mac OS X 10.7 can
have multiple instances of itself open at the same time on the same
screen. With three terminals running, she could persistently have her
source code visible and executing at the same time. With one glance
pattern she could move from the board's visual output to the test to the
program source and back again. Collectively, the field of displays
enabled a kind of programming that was substantially different---in
terms of its representational affordances---from using GDB or a single
terminal instance. Moreover, a snapshot-history-only research approach
would have never had access to how or why Dana (re)configured her
development environment.

\paragraph{Toby said recursion was the ``hardest programming way to
think''}\label{toby-said-recursion-was-the-hardest-programming-way-to-think}

When I gave Toby a code sample task in an interview, he immediately
commented on the fact that it used recursion. Recursion was, as he said,
``the hardest programming way to think.'' At one point his exact
phrasing, quoting the film \emph{Ice Age}, was that ``recursion is bad
juju.'' The task in question was adapted from an example in Abelson and
Sussman (1996) that approximates the square root of a number using an
elegant recursive approach.

Toby's first explanation for why recursion was bad was that it's a
completely unnatural way to think. He cited examples from cooking by
saying, in effect, ``Nobody ever stirs by saying stir once, and if it's
not stirred, stir once. They say stir 100 times.'' He cited simple
operations---including incrementing a number---where using recursion to
push and pop a stack seemed needlessly complex. One of his most damning
indictments of recursion was that his class introduced it as a solution
to generate numbers in the Fibbonacci sequence. Of course recursion
works for examples that are mathematically recursive, he reasoned, but
outside of those obviously contrived examples it was bad joo joo when
compared to understandable, sensible iteration.

The fuller account of Toby's resistance to recursion---and slight change
of heart after an instructional moment---is beyond the scope of what I
can present here. Suffice it to say that Toby's feelings toward
recursion---rooted in a sense of it being an unnatural way to
think---strongly directed his approach to programming. The approach was
so strong, in fact, that during our first interview he wasn't fully able
to produce working code examples of why recursion was absurd despite his
vehement conviction that it was. In other words, he maintained that
recursion was bad and inefficient even though every time he tried to
demonstrate its inferiority he made mistakes in his code and his
examples didn't work.

\subsubsection{Dynamic epistemological models can offer a lens for
reforming assessment and
instruction.}\label{dynamic-epistemological-models-can-offer-a-lens-for-reforming-assessment-and-instruction.}

Historically, a recognition of in-pieces cognitive dynamics in science
education (diSessa, 1993; Hammer, 1994) has led to more direct research
on how those frameworks inform instruction (Hammer \& Elby, 2003;
Hammer, 1996; Louca et al., 2004). What I describe here is research
still at the formative end of generating models to explain cognition.
And, because the focus of the research was explicitly on learning and
not instruction, I advise prudence in trying to draw certain kinds of
specific recommendations from the work I present.

Those restrictions being said, we can think carefully about what this
work means for instruction. For practical purposes, let's refine
terminology so we can talk more specifically and precisely about
components of teaching. Below are my working definitions for teaching
components:

\begin{itemize}
\item
  How should instructors change the way they act in the classroom
  (\emph{instruction})?
\item
  How should this research inform the set of intended learning outcomes
  instructors prepare (\emph{curriculum})?
\item
  How should we change the way we measure or otherwise go about
  ascertaining what students know (\emph{assessment})?
\end{itemize}

The clearest implications of my work are in instruction and assessment.

From an instruction perspective, it may help teachers to know that
students can enact different epistemological stances as part of their
programming practice. As stated, that finding is in principle not new
(cf., Gaspar \& Langevin, 2007). What is new, I think, is the
characterization that these stances evidence epistemological resources
that can be tapped for learning (cf., Hammer \& Elby, 2003). That
sense-making about program design choices or computational concepts is a
worthwhile activity, for example, is a message instructors can send to
students that resonates with larger findings in engineering education
research (Danielak et al., in press). Moreover, instructors can be more
sensitive to what kinds of knowledge-activity students think they're
doing, particularly when making epistemological moves like
\emph{supposing} might help students work through design problems.

My findings also speak to assessment. In the course I studied, there
were few (if any) assessment instruments that would have revealed the
kinds of knowledge I document Lionel and Rebecca having in this study.
At no point in my observational data of the course were students asked
to explain a design choice or reason about competing solutions to handle
a problem. At no point were they encouraged to deeply sense-make about
the conceptual content of the course, particularly the traditionally
tricky topics of pointers and pointer arithmetic. At no point in my data
did the instructor model the kind of sense-making Rebecca did in her
interviews, a fact which she noted at one point as distinguishing the
course from Basic Programming, its first-semester counterpart. Because
the course never explicitly asked students to sense-make, reason about
design choices, or explain designs, it never used information about what
knowledge students had about those things to inform its instruction. It
couldn't. Moreover, because the course never explicitly asked students
to sense-make, reason about design choices, or explain designs, I was
left to conclude such things were not prioritized learning outcomes of
the course. To be clear, all courses establish a kind of focus by
deciding what won't be covered, and I don't fault the instructor at all
for running a course where design knowledge wasn't an enacted learning
outcome. But, I'm also left to wonder: why wasn't it? Given that even
beginning students can think about software design, and given that
engineering as a discipline fundamentally involves design, shouldn't it
have been?

\section{Conclusion}\label{conclusion-2}

If there is one overarching finding from this dissertation, it's that
students clearly have resources for thinking about designing programs
and a diversity of approaches to programming in the moment. Below, I
explain the kind of diversity I saw in programming approaches. Then, I
conclude with a discussion of what my research might mean for
assessment.

\subsection{We should think carefully about what students' programming
design knowledge means for
assessment}\label{we-should-think-carefully-about-what-students-programming-design-knowledge-means-for-assessment}

My research helps us document and model the kinds of knowledge students
have. In so doing, it points out kinds of information that the course's
assessments were apt to miss:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  How students frame or otherwise approach the task of programming in
  the moment
\item
  The role of different kinds of prior experience in stabilizing (or
  potentially destabilizing) certain kinds of frames
\item
  In-the-moment practices---including talking out solutions, sketching
  out debugging strategies, and writing out pseudo-code---that display
  students' competence and sense-making
\item
  The code history that traces how designs evolve, including how
  students start projects and what parts of their designs become
  dead-ends
\end{enumerate}

Together, those four points are relevant for formative assessment (Black
\& Wiliam, 1998) in introductory programming. (1) and (2) point us
toward what students think they're doing when they're programming.
Analogous work from science education tells us that students' sense of
the kind of knowledge activity they're enacting matters for learning and
assessment (Russ et al., 2008). For example, knowing early on that
students are blindly copying code or randomly trying syntax gives
instructors a chance to intervene. But, my research suggests
intervention can't just be about stopping a bad behavior: knowing that
copy-paste behavior is happening is different from knowing \emph{why}
it's happening.

Formative assessment has to be about diagnosing causes, not just
identifying symptoms. Otherwise, we run the risk of ignoring or even
harming students' productive knowledge. Take the example of copy-paste
behavior, as documented by Gaspar and Langevin (Gaspar \& Langevin,
2007). One reason we may see copy-paste behavior, as Study 1
demonstrates, is that students might see situations as new instances of
already-solved problems. In professional practice, seeing old solved
problems in new situations can be productive. Such insights can, for
example, direct engineers to use a pre-built library of functions
instead of building their own. But, another related reason for
copy-paste may be efficiency. For Rebecca, copying and pasting code was
very fast; it required only a few quick keystrokes. In the short run,
Rebecca's choice to copy the code was a faster, less demanding, more
trustworthy route to go on than was abstracting the code to a function.
When we consider students trying to see common problems in new scenarios
and solve such problems efficiently, ``copy-paste'' becomes a symptom
rather than a root cause. Epistemological frameworks, then, can inform
assessments by offering explanations that aim at the root cause of
certain behaviors.

Points 3 and 4 complement knowledge analysis by drawing focus to
artifacts, practices, and history. As data from Lionel shows (Study 2),
a crucial part of his design process involves artifacts and activities
that were only distally knowable to the assessments he got in class.
Lionel's instructor had no direct access to:

\begin{itemize}
\item
  Lionel's whiteboard
\item
  the hours Lionel may have spent working out designs in chalk
\item
  how Lionel might have talked out design features
\item
  how Lionel's initial ``pseudo-code'' evolved into his final design.
\end{itemize}

In Rebecca's case, commented-out code in her final project submission
were perhaps the only clues at all that she tried abstracting some
flight day-checking procedures into functions (Study 1). Without
Rebecca's history, the instructor would have had no reliable way of
knowing:

\begin{itemize}
\item
  Rebecca began her flights database project by copying code from an
  earlier project
\item
  Rebecca's original solution for day-checking evolved from a seven-fold
  conditional structure, one for each day
\item
  Rebecca may have tried abstracting day-checking procedures, even
  creating a function called check\_days.
\end{itemize}

To sum up: traditional assessments can tell us what students finally
produce but not how they produce it, when they produce it, why they
produce it, or what the production process was. In the summative
assessment structure of the course, final student products were the only
products submitted to the instructor. My research shows that what's
happening in the interstices---before typing, between compiles, away
from the computer---can be captured and, in principle, analyzed and
acted upon by instructors.

\subsection{What We Might Change About Classroom
Practice}\label{what-we-might-change-about-classroom-practice}

What follows is my speculation about how we might specifically change
classroom practices in light of my research findings. I move from
recommendations I think are most strongly supported by my data to
recommendations that align with my findings but are more expansive, and
thus less strongly supported. As my recommendations broaden, I try to
offer not only an instructional recommendation but a concomitant
question for research.

\subsubsection{Instructors could look beyond content to understand
student
difficulties}\label{instructors-could-look-beyond-content-to-understand-student-difficulties}

I think first and foremost, instructors have to have a willingness to
recognize that the difficulties they think they're seeing in students
are difficulties they may be viewing through a lens of content. For
example, after seeing a student struggle an instructor might say ``the
student doesn't know assignment statements.'' But, those difficulties
almost certainly have a deeper explanation. I say ``almost certainly,''
where what I mean is that there are a number of patterns we've
identified as common novice errors, but most research stops before
asking why that's a common novice error. With rare exceptions (Fleury,
1991, 2000), the computing education community doesn't encourage asking
the question \emph{why might a student be doing this?} and \emph{what
might this tell me about the way a student is making sense of this?} By
contrast, my research suggests those orientations are ripe, low-hanging
fruit for instructors to take on. I think one of the first things as
instructor could ask is, \emph{why would a student be thinking that this
is the appropriate thing to be doing?} And that holds true whether the
thing in question is writing the statement this way, or interpreting the
code this way, or enacting this kind of programming activity this way.

Moreover, we know from research on metacognition (Schoenfeld, 1987,
1992) that thinking beyond content opens the palette of kinds of
interventions an instructor can make. Schoenfeld (1987), for example,
came to encourage metacognition in his classroom by doggedly asking
student groups what they were doing, why they were doing it, and how
they hoped it would lead them to a solution. Students ultimately came to
internalize such strategies and spent far less time floundering along
solution paths that weren't ultimately productive. Might the same idea
be true in introductory programming courses? To find out, we could begin
researching in earnest how often and in what ways instructors model
metacognition for their students. Currently, I would argue, we don't
know whether and how instructors do so.

\subsubsection{Instructors could use code history to inform
interventions}\label{instructors-could-use-code-history-to-inform-interventions}

When, for example, a student comes to office hours with a problem on a
project, an instructor needs to quickly come to grips with the state of
the students' project, the logic of their design, where they're stuck,
why they're stuck, and what might best help them. That's no small task.
But, given my analysis I have strong reason to believe that having code
history available to instructor could change both the nature of coming
to terms with a student's project and the conversation with a student
that results. If an instructor can see the evolution of a student's
code, at the very least they could see where the student started, how
the code was growing, and where the student was working most recently.
Even if the instructor had never before seen the code (or its history)
until that moment in office hours, having both available changes the
kind of view the instructor can get of the code and the specifics of the
intervention that might result. In Rebecca's case, for example, it might
have been an opportunity to explore why her check\_days abstracted
function was failing on her flights database project.

Having code-snapshot capabilities also changes the kind of research we
can do. First and foremost, a result of this dissertation was to create
a freely-available, lightweight, open-source framework for capturing and
visualizing students' code histories.\footnote{https://github.com/briandk/gitvisualizations}
So, the most basic kind of study would involve deploying that framework
from the instructional side of a course and exploring what happens, as,
for example, Hurd (2013) has done. One could ask questions of how
instructors build code snapshotting into their course, how such
information could or did inform assessment, and how such information
could or did change the nature of instructional interactions with
students. Was it for the better? If so, how do we know? If not, how do
we know?

\subsubsection{Instructors could establish a norm of asking why a design
choice makes
sense}\label{instructors-could-establish-a-norm-of-asking-why-a-design-choice-makes-sense}

The most far-reaching implication of the research I present here is that
instructors should establish a norm in their classes where anyone, at
any point, for any piece of code, is allowed to ask \emph{why does this
make sense as a design choice? Why is that an obvious choice to make?
How does that work?} Consider a parallel example from mathematics
education. In Don Saari's calculus class at UC Irvine, he ``invokes the
principle of what he calls ``WGAD''---``Who gives a damn?'' (Bain, 2004,
pp. 38--39). Bain explains:

At the beginning of his courses, he tells his students that they are
free to ask him the question on any day during the course, at any moment
in class. He will stop and explain to his students why the material
under consideration at that moment---however abstruse and minuscule a
piece of the big picture it may be---is important, and how it relates to
the larger questions and issues of the course. (Bain, 2004, p. 39)

What I'm suggesting is even broader than that, because it's not just a
question students can ask of professors; it's a question \emph{anyone}
can ask of \emph{anyone.} I see it as the programming and design
extension of a sociomathematical norm (Yackel \& Cobb, 1996), and I
think it could lead to collaborative sense-making. That is, I'm asking
for an accepted cultural practice that is also itself a design practice,
whereby an instructor can help create a safe, stable space for a
community to be reflective and critical about design choices. I also
think it's an idea that leads to others, such as letting, if not
requiring, students review one another's code. When students are forced
to reckon with someone else's code and understand their design
decisions, they're also forced to justify their own decisions. And
establishing ``why that design choice?'' or ``how does that make
sense?'' as a norm provides reciprocal opportunities for students, not
just instructors, to improve the code of others.

The research questions that come out of such an idea would include:

\begin{itemize}
\item
  How can an instructor satisfactorily establish norms about design in a
  classroom?
\item
  What does it look like when students collaboratively sense-make about
  a program's design choices?
\item
  What does it look like when students are asked to reflect on their own
  design choices?
\item
  How might introducing a code review component into a course change the
  way students approach design? How might it improve students'
  conceptual understanding? How might it improve the quality of the code
  students produce?
\end{itemize}

\subsection{Final Remarks}\label{final-remarks}

How students design programs matters for learning and instruction in
engineering. It matters because finished code reflects what students
know about design, whether or not instructors capture such information.
It matters because students have resources for learning about and
engaging in design; whether or not curriculum, instruction, and
assessment choose to tap into those resources. It matters because design
should be an intellectual thread that runs through all engineering
courses. That thread shouldn't stop when we introduce students to
programming.

\section{Appendix 1 -- Transcript
conventions}\label{appendix-1-transcript-conventions}

\begin{itemize}
\item
  Turns are not numbered, but they are blank-line-delimited
\item
  Short interjected speech that does not interrupt a speaker's turn is
  bounded /by slashes/
\item
  Matching double brackets show the {[}{[}onset and termination{]}{]} of
  overlapping talk across turns.
\item
  *emphasized speech* is bounded by asterisks
\item
  Parenthetical clarifications by the analyst appear in parentheses. (He
  considered but rejected square brackets.)
\item
  matching double equals signs mark turn boundaries== ==with minimal or
  no audible silence (also known as latching turns)
\item
  Where gestures don't overlap speech they are in-lined by curly braces
  when they happen \{folds arms, having made his point\}. All gestures
  enacted \emph{by} a speaker appear within that speaker's turn unless
  otherwise noted \{smiles, self-satisfied at having made this important
  clarification\} \{audience scoffs\}.
\item
  When gestures happen \emph{during} speech, the speech is presented
  first and bounded by double pipes. Gestures that happen simultaneously
  with such speech are bounded by curly braces, nested within double
  pipes, and immediately follow the speech they overlap. For example:
\end{itemize}

Throwing chainsaws \textbar{}\textbar{}*up*\textbar{}\textbar{}
\textbar{}\{throws chainsaw\}\textbar{} is easy. Just be careful when
they come \textbar{}\textbar{}*down*\textbar{}\textbar{}
\textbar{}\{catches chainsaw for a punctuated finish\}\textbar{}.

\section{Appendix 2 -- Visual conventions for
gestures}\label{appendix-2-visual-conventions-for-gestures}

A challenge of this dissertation was trying to use still-frames to
convey motion and animation. Where I could, I annotated pictures with
arrows to show trajectories of movement:

\includegraphics[width=5.62222in,height=0.87778in]{media/image38.emf}

\begin{longtable}[]{@{}ll@{}}
\toprule
\textbf{Description} & \textbf{Example}\tabularnewline
\midrule
\endhead
\textbf{Blue} arrows project forward to show action that will occur but
hasn't yet in the frame you're looking at. The origin is where a hand
\emph{is}; the terminus is where it \emph{will be} shortly. &
\includegraphics[width=1.67778in,height=1.58889in]{media/image39.png}\tabularnewline
\textbf{Pink} arrows show action that already occurred before the frame
you're looking at. The origin is where a hand \emph{was}; the terminus
is where it \emph{is}. &
\includegraphics[width=1.65556in,height=2.37778in]{media/image20.png}\tabularnewline
\bottomrule
\end{longtable}

\section{\texorpdfstring{Appendix 3 -- Transcript of Rebecca's
pseudo-code episode \emph{without} gesture
codes}{Appendix 3 -- Transcript of Rebecca's pseudo-code episode without gesture codes}}\label{appendix-3-transcript-of-rebeccas-pseudo-code-episode-without-gesture-codes}

Interviewer: OK. So then, um, what if we just pretended for a minute,
that that, like==

Rebecca: ==That that works=

Interviewer: ==That that worked==

Rebecca: ==OK==

Interviewer: ==OK. So then, um. So then you might write like to make an
array of \{stops writing\} actually what would we call this? This is==

Rebecca: ==array of pointers, I guess==

Interviewer: ==OK. Pointers. It'd be, well. So. The---the one we had was
like int, star p, um, it would be some number, um, and I guess that'd be
it in order to just declare /Yes/ that right? OK. And then you're
saying, how would you like access an element of it?

Rebecca: Uh, well, what I was think---like if I wanted to like, save it
or whatever, /Yeah/ I could make a while loop, scan in the---scan in the
data from the album, /OK/ uh, which is, I can write it {[}if you want

Interviewer: Sure, go ahead{]}{]}

Rebecca: Um, so, it was==

Interviewer: {[}{[}Oh, sorry, just have to{]}

Rebecca: {[}Oh{]}

Interviewer: It's /Oh, OK/ one of those annoying ticks, but it works
best cause of the camera if, ah, /Ah, OK/ if that's pointing away from
your hand

Rebecca: {[}OK{]}

Interviewer: {[}That's fine{]}

Rebecca: Um, there was album, and then, so there was, uh, percent d, and
percent s, and then give those names, just, I'll just call it number,
and title. /Mmmhmm/ And so, at least, my---until it does not equal EOF,
and then my thinking at least, is you should be able to, um, say that

Rebecca: ``star p of i'' /mmhmm/ equals, uh, the title, and then you
just do i++, so then it'll move to the next one /OK/ and you just keep
saving each of the pointers in the array to a title /Mmmhmm/ And you
just increment by 1 until you reach the end of file.

Interviewer: OK==

Rebecca: ==Like *that* would make sense to me.

Interviewer: So then, like, what would go in p of one would be the first
title we read in==

Rebecca: ==Yes,

Interviewer: uh, would, when the while loop runs again, does it get a
fresh line {[}{[}line from that

Rebecca: Yeah, because{]}{]} uh, what the while loop does is it reads in
the line and then once it reaches the end of character it'll go back
down---it'll do the while loop and then go back down to the next
thing---line, it'll read in the line until it reaches the end of file

Interviewer: So as you step through this loop /mmhmm/ i keeps going up
by one==

Rebecca: ==Yes.

Interviewer: Uh

Rebecca: And the lines keep going down so, that way, the first line is
going to be, uh the p---element zero /OK/ uh the second one'll be
element one (Interview 4 of 5, April 6, 2012)

\section{Appendix 4 -- Conceptual knowledge frameworks in computing
don't tell us much about how Lionel structured an in-interview
program}\label{appendix-4-conceptual-knowledge-frameworks-in-computing-dont-tell-us-much-about-how-lionel-structured-an-in-interview-program}

In this section, I explore Lionel's work in solving an in-interview
programming task. First, I reproduce the prompt Lionel was given. Next,
I show the final source code of his solution to the problem, written in
C. In the analysis that follows, I move outward to explore the context
of that code's production.

My choice to start with the code first is deliberate. I'm trying to
mimic what the instructor of a typical programming course might see: the
final submitted form of a student's code. This approach is admittedly a
bit backwards, since the data I show after I present Lionel's submission
is all about what happened \emph{before} the code reached its final
form. But, I think this approach useful because it invites us to explore
an artifact first, then raise questions about the conditions of its
production. In that sense, my presentation has the anthropological feel
of understanding a found object, which I think is a faithful way of
looking at what many instructors (not to mention researchers, and of
course active software developers) face in their day-to-day work.

\subsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188555}{}{\protect\hypertarget{ux5fToc252445950}{}{}}Analyzing
Lionel's Solution From a Conceptual Knowledge
View}{Analyzing Lionel's Solution From a Conceptual Knowledge View}}\label{analyzing-lionels-solution-from-a-conceptual-knowledge-view}

To infer conceptual knowledge, I'll use Elliott Tew's (2010) conceptual
categories for procedural programming in first-semester computer science
(CS1) courses. My reasons for doing so are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Elliott Tew (2010) sought concepts that were language-agnostic. That
  is, the concepts Elliott Tew (2010) identified are not idiomatic to
  just one language. Rather, they are general ideas implemented in
  nearly every major procedural language taught to undergraduates (Java,
  Python, C-based languages, Scheme).
\item
  Elliott Tew's (2010, pp. 23--27) selection process was extensive. She
  began with curricular documents, moved to canonical texts, then
  ultimately conducted a thematic analysis on emerging categories to
  produce her final list of concepts.
\item
  To date, Elliott Tew's (2010) identification of CS1 concepts is the
  only work I know of that has been carried forward to create,
  administer, and validate concept inventories in CS1 (Elliott Tew \&
  Guzdial, 2010, 2011)
\end{enumerate}

Figure 2 below presents an overview of the conceptual features I
identified in Lionel's code. Even in a short, relatively simple program
Lionel exhibits four of the ten concept categories Elliott Tew (2010)
identifies. Moreover, he uses each of them ``correctly,'' so to speak,
in that not only does his program compile, it also properly finds the
range of the numbers it's given. It's also worth noting that Lionel's
code displays consistent, helpful use of whitespace (indentation) as
well as informative comments. While these attributes aren't conceptual,
per se, they are the sorts of features an instructor might consider in
assigning a grade to this program.

\includegraphics[width=5.60000in,height=5.67778in]{media/image40.jpeg}

\protect\hypertarget{ux5fToc252445975}{}{}Figure ‑ -- Highlighting the
conceptual features of Lionel's code given Elliott Tew's (2010) outline
of CS 1 concepts

\section{Appendix 5 -- Neverly-Asked Questions
(NAQs)}\label{appendix-5-neverly-asked-questions-naqs}

\subsubsection{Neverly Asked Questions (NAQs) about my conceptual
analysis of Lionel's
code}\label{neverly-asked-questions-naqs-about-my-conceptual-analysis-of-lionels-code}

\emph{Didn't Elliott Tew (2010) define those concepts as a way of
creating a concept inventory?}

Yes. In her work, I believe each concept was represented by at least one
multiple-choice question (MCQ) with carefully-chosen distractors. In
each MCQ a student would see a snippet of pseudo-code and be asked to
choose from a list of responses. The structure was explicitly modeled
after Force Concept Inventory (FCI) work in physics education.

\emph{So, if her intent (and research) was about making a concept
\textbf{inventory}, aren't you misusing her ideas of concepts?}

What makes you say that?

\emph{Well, first off, she used categories of concept to create concrete
instances of code, which she tests students on}\ldots{}

That's my understanding, yep.

\emph{But you used a concrete instance of code --- Lionel's code --- and
inferred the existence of concepts in it. So, that's not the same thing
as what Elliott Tew (2010) did.}

I agree I'm not doing what she did, but I don't think that observation
necessarily undermines the usefulness of what I \emph{am} doing. If the
concepts she identifies really are general concepts, then it would seem
silly to think we couldn't find concrete instances of code that
exemplify them. I think I might rephrase your objection: ``how can I
trust that you properly identified concepts in Lionel's code?'' My
answer to that is, again, yes, I have no coding scheme from Elliott Tew
to apply. But, the code Lionel wrote is there and my candidates for
assigning concepts are there: my results are open to inspection and
challenge.

\emph{OK. Fine. But what ``conceptual knowledge analysis'' is there to
speak of? You just made a slide and identified which parts of his code
might correspond to which concepts.}

What I did was assign concepts to canonically-correct --- that is,
syntactically valid parts of Lionel's code that also work properly to
make the program produce the right results --- sections of Lionel's
code. I think your objection is about my inferential basis: ``why does
the presence of those code chunks --- and my tagging certain chunks as
aligning with Elliott Tew's (2010) concepts --- imply that Lionel has
those concepts in his head?''

\emph{I accept your amendment. So, what basis do you have for saying
``this concept is in Lionel's mind?''}

I feel like I'm on the same playing field as the concept inventory
folks, though in some ways I'm setting a modified standard for validity.
Elliott Tew (2010) says this about validity:

validity is the evidence that assures us that questions about a
particular concept are indeed measuring that concept. For instance, a
question about arrays should require a student to have knowledge about
arrays, but should not require knowledge about another concept, such as
recursion. In addition, it is important that the question cannot be
answered correctly without knowledge of arrays. (Elliott Tew, 2010, p.
10)

I don't know that I share her assumptions about validity. But, what
specifically might be at issue here is ``does the presence of a for-loop
in code imply the student has knowledge of definite loops?'' My answer
is, ``if not that, then what?''

\emph{But couldn't I argue that the for-loop could be there because he
copied and pasted it from the internet? That he wouldn't have had to
know anything about definite loops to do that?}

Yes, you could. My rejoinder to that is in section 3.3.3.

\subsubsection{Neverly-asked questions about Lionel's verbal pseudo-code
description}\label{neverly-asked-questions-about-lionels-verbal-pseudo-code-description}

Why do you so pedantically analyze such a short, small snippet of
conversation?

Because by Lionel's own assertion what was said constitutes his ``main
concept'' (line 12) for the program. That is, by his own account he has
just expressed the entire top-level design for the program. But, he's
done it all all without writing a single additional line of code.

So?

If anything, this episode suggests that Lionel can and does have ways of
expressing a process or procedure that don't rely on writing code. More
importantly, I think, is that he views it as a legitimate activity to
express ideas at a level \emph{above} the particularities of syntax and
implementation.

What do you mean ``views it as a legitimate activity?''

I think here I'm appealing to my interpretation of Goffman's (1974)
\emph{frames}: a participant's understanding of ``what is it'' that's
going on in a social situation. Lionel takes up my bid for him to
explain what he's thinking, but he does so first by explaining his
``general concept'' for the program.

Right. But, I mean, he's just doing what you \textbf{asked} him to do. I
don't understand how any of this relates to Goffman or frames. That just
seems like a needlessly complicated way to explain he did what you
asked.

OK, let's step back for a second. Suppose you've just started going to
group therapy. One of the things group therapy environments often stress
is that you should try to talk about how \emph{you feel}, which can be a
hard transition for people. Group therapy participants may be used to
saying ``you always do {[}X{]}'' when in a fight with someone. And, if
prompted by a group leader to discuss how they feel about troublesome
instance, a participant may adopt that kind of language. ``My partner
always yells at me if I'm out late,'' for example.

Group therapy can encourage inward reflection in a way that reshapes how
participants talk about themselves and their feelings. So, after some
time in group therapy our hypothetical participant may change the way
they orient toward questions. If asked ``how are you feeling about the
fight you and your partner had?'' our hypothetical participant may now
say ``I feel like I don't have the freedom to see the people I want to
see.'' The focus in speech shifts from the \emph{other}---in this case a
partner---to the \emph{self}. Moreover, the substance of the speech is
about feelings, rather than behavior.

I think these reshapings of speech are neither trivial nor incidental.
If we had transcripts of our hypothetical person's early sessions in
group therapy and compared them to those of later sessions, I think we
would not be surprised to see a marked change. When asked the same
question---in this case ``how are you feeling about {[}X{]},'' the
structure and character of our participant's response changes after
therapy.

My point here is that an everyday sense of how participants' speech
patterns change in therapy primes our intuition for thinking more
broadly about social phenomena. A cynical view of our participant would
still have to acknowledge that, observably, the structure of their
response to the question ``how does that make you feel?'' changes after
therapy. I think a sociolinguistic stance would be more generous: our
participant is \emph{orienting} differently toward the question.
Furthermore, I take the following as evidence of such an orientation:

\begin{itemize}
\item
  A change in sentence focus from the other to the self
\item
  Introspection into the source of a feeling
\item
  Acknowledgment of personal responsibility for feeling those feelings
\end{itemize}

I think we as analysts can acknowledge that after therapy, our
participant frames the event of being asked, ``how does that make you
feel?'' differently.

OK. So, you're saying that if we look at the change in how someone
answers the question ``how does that make you feel?'' we can find
evidence of the superstructure---a frame---that directs their sense of
appropriate ways to answer that question?

Yes.

So how does this get back to your claims about Lionel?

I think it's tempting to take this position that, ``it's
inevitable/totally expected that he would just verbalize the top-level
description of his procedure! I mean, what else could he possibly do?''

I mean, yeah. Kind of.

Right, but I think actually that assumption isn't in line with a classic
frame analysis approach. Even if the assumption is right, frame analysis
changes the explanation from ``it's inevitable'' to ``hmm, I guess
almost everyone I've ever seen brings or enacts the same structure of
expectations.''

So, my argument is, if you ask someone to describe to you how they're
thinking about their program, they're gonna do that. What else
\textbf{would} they do?

Right. And I don't think we disagree that they'll describe their
thinking. I think where we disagree is that everyone will describe in
the same way. I'm saying \emph{if} they describe their program to you,
\emph{that} they're able to express it, what they choose to verbalize,
what they choose to leave out, how their gestures accompany the
explanations, and their ultimate criteria for having satisfied the task
of describing for themselves are all driven underneath (or from above?)
by structures of expectations. And even if \emph{one thousand} out of
one thousand people do the same kind of thing Lionel does, that doesn't
undermine the idea that people have and bring structures of expectations
to a task. What I think it does is show that one thousand out of one
thousand people share the same sets of expectations about what it is
that's going on when they're asked to verbalize their thinking on a
programming protocol.

So, what's the point of this whole section on Lionel's verbal
description, then?

I think the point is that when I ask Lionel to verbalize his thinking,
he outlines through talk what the ``main concept'' of his program is. I
focus on \emph{that} he decides to describe it out in words, \emph{how}
he describes it, \emph{what} he leaves out\emph{, how} he repairs, and
how all of that reflects structures of expectations about the activity
and what he's being asked to do. Ultimately, the argument is that
Lionel's choice to take me up on that bid and how he does so
reflects---to me, the analyst---that he views \emph{verbal description
of a high-level program} as a legitimate activity and part of his
approach to programming.

\section{References}\label{references}

Abelson, H., \& Sussman, G. J. (1996). \emph{Structure and
Interpretation of Computer Programs} (2nd ed.). Cambridge, Mass: MIT
Press.

Adelson, B., \& Soloway, E. (1985). The Role of Domain Experience in
Software Design. \emph{IEEE Transactions on Software Engineering},
\emph{SE-11}(11), 1351 -- 1360. doi:10.1109/TSE.1985.231883

Archer, L. B. (1979). Whatever became of design methodology.
\emph{Design Studies}, \emph{1}(1), 17--18.

Bain, K. (2004). \emph{What the best college teachers do}. Cambridge,
Mass: Harvard University Press.

Baker, A., \& van der Hoek, A. (2010). Ideas, subjects, and cycles as
lenses for understanding the software design process. \emph{Design
Studies}, \emph{31}(6), 590--613. doi:10.1016/j.destud.2010.09.008

Ball, L. J., Onarheim, B., \& Christensen, B. T. (2010). Design
requirements, epistemic uncertainty and~solution development strategies
in software design. \emph{Design Studies}, \emph{31}(6), 567--589.
doi:10.1016/j.destud.2010.09.003

Bayman, P., \& Mayer, R. E. (1983). A diagnosis of beginning
programmers' misconceptions of BASIC programming statements.
\emph{Communications of the ACM}, \emph{26}(9), 677--679.
doi:http://doi.acm.org/10.1145/358172.358408

Black, P., \& Wiliam, D. (1998). Inside the Black Box: Raising Standards
Through Classroom Assessment. \emph{Phi Delta Kappan}, \emph{80}(2),
139--44.

Boaler, J. (1998). Open and closed mathematics: Student experiences and
understandings. \emph{Journal for Research in Mathematics Education},
\emph{29}(1), 41--62.

Boaler, J. (2000). Mathematics from another world: Traditional
communities and the alienation of learners. \emph{The Journal of
Mathematical Behavior}, \emph{18}(4), 379--397.
doi:10.1016/S0732-3123(00)00026-2

Boaler, J. (2002). The development of disciplinary relationships:
Knowledge, practice and identity in mathematics classrooms. \emph{For
the Learning of Mathematics}, \emph{22}(1), 42--47.

Boaler, J., \& Greeno, J. G. (2000). Identity, agency, and knowing in
mathematical worlds. In J. Boaler (Ed.), \emph{Multiple perspectives on
mathematics teaching and learning} (pp. 171--200). Westport, CT: Ablex
Pub.

Bonar, J., \& Soloway, E. (1983). Uncovering principles of novice
programming. \emph{Proceedings of the 10th ACM SIGACT-SIGPLAN Symposium
on Principles of Programming Languages}, 10--13.
doi:10.1145/567067.567069

Bonar, J., \& Soloway, E. (1985). Preprogramming Knowledge: A Major
Source of Misconceptions in Novice Programmers. \emph{Human-Computer
Interaction}, \emph{1}(2), 133.

Bucciarelli, L. L. (1994). \emph{Designing engineers}. Cambridge, Mass:
MIT Press.

Clancy, M. (2004). Misconceptions and Attitudes that Interfere with
Learning to Program. In S. Fincher \& M. Petre (Eds.), \emph{Computer
Science Education Research} (pp. 85--100). London, UK: RoutledgeFalmer.

Danielak, B. A., Gupta, A., \& Elby, A. (in press). The Marginalized
Identities of Sense-Makers: Reframing Engineering Student Retention.
\emph{Journal of Engineering Education}.

Danielsiek, H., Paul, W., \& Vahrenhold, J. (2012). Detecting and
Understanding Students' Misconceptions Related to Algorithms and Data
Structures. In \emph{Proceedings of the 43rd ACM Technical Symposium on
Computer Science Education} (pp. 21--26). New York, NY, USA: ACM.
doi:10.1145/2157136.2157148

diSessa, A. A. (1986). Models of Computation. In D. A. Norman \& S. W.
Draper (Eds.), \emph{User centered system design: new perspectives on
human-computer interaction} (pp. 201--218). Hillsdale, N.J: L. Erlbaum
Associates.

diSessa, A. A. (1993). Toward an Epistemology of Physics.
\emph{Cognition and Instruction}, \emph{10}(2/3), 105--225.

diSessa, A. A. (2002). Why ``Conceptual Ecology'' is a good idea. In M.
Limón \& L. Mason (Eds.), \emph{Reconsidering conceptual change: issues
in theory and practice} (pp. 29--60). Dordrecht ; Boston: Kluwer
Academic Publishers.

diSessa, A. A., \& Sherin, B. L. (1998). What changes in conceptual
change? \emph{International Journal of Science Education},
\emph{20}(10), 1155--1191. doi:10.1080/0950069980201002

Duckworth, E. R. (2006). \emph{``The having of wonderful ideas'' and
other essays on teaching and learning} (3rd ed.). New York: Teachers
College Press.

Elby, A., \& Hammer, D. (2010). Epistemological resources and framing: A
cognitive framework for helping teachers interpret and respond to their
students' epistemologies. In L. D. Bendixen \& F. C. Feucht (Eds.),
\emph{Personal epistemology in the classroom: theory, research, and
implications for practice} (pp. 409--434). Cambridge, UK ; New York:
Cambridge University Press.

Elliott Tew, A. (2010). \emph{Assessing fundamental introductory
computing concept knowledge in a language independent manner} (Ph.D.).
Georgia Institute of Technology, Ann Arbor. Retrieved from ProQuest
Dissertations \& Theses Full Text. (873212789)

Elliott Tew, A., \& Guzdial, M. (2010). Developing a validated
assessment of fundamental CS1 concepts. \emph{Proceedings of the 41st
ACM Technical Symposium on Computer Science Education}, 97--101.
doi:10.1145/1734263.1734297

Elliott Tew, A., \& Guzdial, M. (2011). The FCS1: a language independent
assessment of CS1 knowledge. In \emph{Proceedings of the 42nd ACM
technical symposium on Computer science education} (pp. 111--116). New
York, NY, USA: ACM. doi:10.1145/1953163.1953200

Erickson, F. (1986). Qualitative methods in research on teaching. In M.
C. Wittrock (Ed.), \emph{Handbook of research on teaching} (3rd ed., pp.
119--161). New York : London: Macmillan ; Collier Macmillan.

Eynde, P., \& Hannula, M. (2006). The Case Study of Frank.
\emph{Educational Studies in Mathematics}, \emph{63}(2), 123--129.
doi:10.1007/s10649-006-9030-8

Fleury, A. E. (1991). Parameter passing: the rules the students
construct. In \emph{Proceedings of the twenty-second SIGCSE technical
symposium on Computer science education} (pp. 283--286). New York, NY,
USA: ACM. doi:10.1145/107004.107066

Fleury, A. E. (1993). Student Beliefs about Pascal Programming.
\emph{Journal of Educational Computing Research}, \emph{9}(3), 355--371.
doi:10.2190/VECR-P8T6-GB10-MXJ5

Fleury, A. E. (2000). Programming in Java: student-constructed rules.
\emph{SIGCSE Bull.}, \emph{32}(1), 197--201. doi:10.1145/331795.331854

Gainsburg, J. (2006). The mathematical modeling of structural engineers.
\emph{Mathematical Thinking \& Learning}, \emph{8}(1), 3--36.
doi:10.1207/s15327833mtl0801\_2

Gal-Ezer, J., \& Zur, E. (2004). The efficiency of
algorithms-\/-misconceptions. \emph{Computers \& Education},
\emph{42}(3), 215--226.

Gaspar, A., \& Langevin, S. (2007). Restoring ``coding with intention''
in introductory programming courses. \emph{Proceedings of the 8th ACM
SIGITE Conference on Information Technology Education}, 91--98.
doi:10.1145/1324302.1324323

Ginsburg, H. P. (1997). \emph{Entering the child's mind: the clinical
interview in psychological research and practice}. Cambridge ; New York:
Cambridge University Press.

Ginsburg, H. P., \& Opper, S. (1988). \emph{Piaget's Theory of
Intellectual Development} (3rd ed.). Englewood Cliffs, N.J:
Prentice-Hall. Retrieved from http://lccn.loc.gov/87017353

Goffman, E. (1974). \emph{Frame Analysis: An Essay on the Organization
of Experience}. New York: Harper \& Row.

Goldin-Meadow, S. (2003). \emph{Hearing gesture: how our hands help us
think}. Cambridge, Mass: Belknap Press of Harvard University Press.

Goodwin, C. (2000). Action and embodiment within situated human
interaction. \emph{Journal of Pragmatics}, \emph{32}(10), 1489--1522.
doi:10.1016/S0378-2166(99)00096-X

Gupta, A., Hammer, D., \& Redish, E. F. (2010). The Case for Dynamic
Models of Learners' Ontologies in Physics. \emph{Journal of the Learning
Sciences}, \emph{19}(3), 285. doi:10.1080/10508406.2010.491751

Hall, R. (1999). Following mathematical practices in design-oriented
work. In C. Hoyles, C. Morgan, \& G. Woodhouse (Eds.), \emph{Rethinking
the Mathematics Curriculum} (pp. 29--47). London: Falmer Press.

Hall, R., \& Nemirovsky, R. (2012). Introduction to the Special Issue:
Modalities of Body Engagement in Mathematical Activity and Learning.
\emph{Journal of the Learning Sciences}, \emph{21}(2), 207--215.
doi:10.1080/10508406.2011.611447

Hall, R., \& Stevens, R. (1995). Making space: A comparison of
mathematical work in school and professional design practices. In S. L.
Star (Ed.), \emph{The cultures of computing} (pp. 118--145). Oxford, UK:
Blackwell Publisher.

Hall, R., Stevens, R., \& Torralba, T. (2002). Disrupting
representational infrastructure in conversations across disciplines.
\emph{Mind, Culture \& Activity}, \emph{9}(3), 179--210.

Hall, R., Wright, K., \& Wieckert, K. (2007). Interactive and Historical
Processes of Distributing Statistical Concepts Through Work
Organization. \emph{Mind, Culture \& Activity}, \emph{14}(1/2),
103--127. doi:10.1080/10749030701307770

Hammer, D. (1989). Two approaches to learning physics. \emph{The Physics
Teacher}, \emph{27}(9), 664--670. doi:10.1119/1.2342910

Hammer, D. (1994). Epistemological beliefs in introductory physics.
\emph{Cognition and Instruction}, \emph{12}(2), 151--183.
doi:10.2307/3233679

Hammer, D. (1996). Misconceptions or P-Prims: How May Alternative
Perspectives of Cognitive Structure Influence Instructional Perceptions
and Intentions? \emph{Journal of the Learning Sciences}, \emph{5}(2),
97--127. doi:10.1207/s15327809jls0502\_1

Hammer, D., \& Elby, A. (2002). On the form of a personal epistemology.
In B. K. Hofer \& P. R. Pintrich (Eds.), \emph{Personal epistemology:
The psychology of beliefs about knowledge and knowing} (pp. 169--190).
Mahwah, N.J: L. Erlbaum Associates.

Hammer, D., \& Elby, A. (2003). Tapping epistemological resources for
learning physics. \emph{The Journal of the Learning Sciences},
\emph{12}(1), 53--90. doi:10.2307/1466634

Hammer, D., Elby, A., Scherr, R. E., \& Redish, E. F. (2005). Resources,
framing, and transfer. In J. P. Mestre (Ed.), \emph{Transfer of learning
from a modern multidisciplinary perspective}. Greenwich, CT: IAP.

Hannula, M., Evans, J., Philippou, G., \& Zan, R. (2004). Affect in
Mathematics Education--Exploring Theoretical Frameworks. Research Forum.
\emph{International Group for the Psychology of Mathematics Education},
30.

Henderson, K. (1999). \emph{On line and on paper: visual
representations, visual culture, and computer graphics in design
engineering}. Cambridge, Mass: MIT Press.

Herman, G. L., Kaczmarczyk, L., Loui, M. C., \& Zilles, C. (2008). Proof
by incomplete enumeration and other logical misconceptions.
\emph{Proceeding of the Fourth International Workshop on Computing
Education Research}, 59--70. doi:10.1145/1404520.1404527

Hofer, B. K., \& Pintrich, P. R. (1997). The development of
epistemological theories: Beliefs about knowledge and knowing and their
relation to learning. \emph{Review of Educational Research},
\emph{67}(1), 88--140. doi:10.3102/00346543067001088

Holland, S., Griffiths, R., \& Woodman, M. (1997). Avoiding object
misconceptions. In \emph{Proceedings of the twenty-eighth SIGCSE
technical symposium on Computer science education} (pp. 131--134). New
York, NY, USA: ACM. doi:10.1145/268084.268132

Hurd, A. (2013, March). \emph{Assessment in an Introduction to
Programming Course}. Presented at the 16th Annual Course Technology
Conference, San Diego, CA, USA. Retrieved from
http://www.slideshare.net/CengageLearning/andrew-hurd-assessment-in-an-intro-to-programming-course

Hutchins, E. (1995a). \emph{Cognition in the Wild}. Cambridge, Mass: MIT
Press.

Hutchins, E. (1995b). How a cockpit remembers its speeds.
\emph{Cognitive Science}, \emph{19}(3), 265--288.
doi:10.1016/0364-0213(95)90020-9

Izsák, A. (2004). Students' Coordination of Knowledge When Learning to
Model Physical Situations. \emph{Cognition \& Instruction},
\emph{22}(1), 81--128.

Jackson, M. (2010). Representing structure in a software system design.
\emph{Design Studies}, \emph{31}(6), 545--566.
doi:10.1016/j.destud.2010.09.002

Jadud, M. C. (2006). Methods and tools for exploring novice compilation
behaviour. In \emph{Proceedings of the 2006 international workshop on
Computing education research - ICER '06} (p. 73). Canterbury, United
Kingdom. doi:10.1145/1151588.1151600

Jordan, B., \& Henderson, A. (1995). Interaction analysis: Foundations
and practice. \emph{Journal of the Learning Sciences}, \emph{4}(1), 39.
doi:10.1207/s15327809jls0401\_2

Joy, M., Sinclair, J., Sun, S., Sitthiworachart, J., \& López-González,
J. (2009). Categorising computer science education research.
\emph{Education and Information Technologies}, \emph{14}(2), 105--126.
doi:10.1007/s10639-008-9078-4

Kaczmarczyk, L. C., Petrick, E. R., East, J. P., \& Herman, G. L.
(2010). Identifying student misconceptions of programming.
\emph{Proceedings of the 41st ACM Technical Symposium on Computer
Science Education}, 107--111. doi:10.1145/1734263.1734299

Kaiser, D. (2005). \emph{Drawing theories apart: the dispersion of
Feynman diagrams in postwar physics}. Chicago: University of Chicago
Press.

Keller, E. F. (1983). \emph{A feeling for the organism: the life and
work of Barbara McClintock}. San Francisco: W.H. Freeman.

Kolikant, Y. B.-D., \& Mussai, M. (2008). ``So my program doesn't run!''
Definition, origins, and practical expressions of students'
(mis)conceptions of correctness. \emph{Computer Science Education},
\emph{18}(2), 135.

Kölling, M., \& Utting, I. (2012). Building an open, large-scale
research data repository of initial programming student behaviour. In
\emph{Proceedings of the 43rd ACM technical symposium on Computer
Science Education} (pp. 323--324). New York, NY, USA: ACM.
doi:10.1145/2157136.2157234

Latour, B. (1987). \emph{Science in action: how to follow scientists and
engineers through society}. Cambridge, Mass: Harvard University Press.

Latour, B. (1990). Drawing things together. In M. Lynch \& S. Woolgar
(Eds.), \emph{Representation in Scientific Practice} (1st MIT Press ed.,
pp. 19--68). Cambridge, Mass: MIT Press.

Lehrer, R., Schauble, L., Carpenter, S., \& Penner, D. (2000). The
interrrelated development of inscriptions and conceptual understanding.
In P. Cobb, E. Yackel, \& K. McClain (Eds.), \emph{Symbolizing and
Communicating in Mathematics Classrooms: Perspectives on Discourse,
Tools, and Instructional Design} (pp. 325--360). Mahwah, N.J: Lawrence
Erlbaum Associates.

Lising, L., \& Elby, A. (2005). The impact of epistemology on learning:
A case study from introductory physics. \emph{American Journal of
Physics}, \emph{73}(4), 372. doi:10.1119/1.1848115

Louca, L., Elby, A., Hammer, D., \& Kagey, T. (2004). Epistemological
Resources: Applying a New Epistemological Framework to Science
Instruction. \emph{Educational Psychologist}, \emph{39}(1), 57--68.

Malmi, L., Sheard, J., Simon, Bednarik, R., Helminen, J., Korhonen, A.,
\ldots{} Taherkhani, A. (2010). Characterizing research in computing
education: a preliminary analysis of the literature. In
\emph{Proceedings of the Sixth international workshop on Computing
education research} (pp. 3--12). New York, NY, USA: ACM.
doi:10.1145/1839594.1839597

Martin, R. C. (2009). \emph{Clean code: a handbook of agile software
craftsmanship}. Upper Saddle River, NJ: Prentice Hall.

Mayer, R. E. (1981). The Psychology of How Novices Learn Computer
Programming. \emph{ACM Computing Surveys (CSUR)}, \emph{13}, 121--141.
doi:10.1145/356835.356841

Minsky, M. L. (1986). \emph{The Society of Mind}. New York: Simon and
Schuster.

Nasir, N. S., \& Cooks, J. (2009). Becoming a Hurdler: How Learning
Settings Afford Identities. \emph{Anthropology \& Education Quarterly},
\emph{40}(1), 41--61. doi:10.1111/j.1548-1492.2009.01027.x

Nasir, N. S., \& Hand, V. (2008). From the court to the classroom:
Opportunities for engagement, learning, and identity in basketball and
classroom mathematics. \emph{Journal of the Learning Sciences},
\emph{17}(2), 143--179. doi:10.1080/10508400801986108

Nasir, N. S., \& Hand, V. M. (2006). Exploring Sociocultural
Perspectives on Race, Culture, and Learning. \emph{Review of Educational
Research}, \emph{76}(4), 449--475.

Nemirovsky, R., Rasmussen, C., Sweeney, G., \& Wawro, M. (2012). When
the Classroom Floor Becomes the Complex Plane: Addition and
Multiplication as Ways of Bodily Navigation. \emph{Journal of the
Learning Sciences}, \emph{21}(2), 287--323.
doi:10.1080/10508406.2011.611445

Ochs, E., Gonzales, P., \& Jacoby, S. (1996). ``When I come down I'm in
the domain state'': grammar and graphic representation in the
interpretive activity of physicists. In \emph{Interaction and Grammar}
(pp. 328--369). Cambridge: Cambridge University Press.

Papert, S. (1980). \emph{Mindstorms: Children, Computers, and Powerful
Ideas}. New York: Basic Books. Retrieved from
http://lccn.loc.gov/79005200

Parsons, J., \& Saunders, C. (2004). Cognitive heuristics in software
engineering: Applying and extending anchoring and adjustment to artifact
reuse. \emph{IEEE Transactions on Software Engineering}, \emph{30}(12),
873 -- 888. doi:10.1109/TSE.2004.94

Patitsas, E., Craig, M., \& Easterbrook, S. (2013). On the Countably
Many Misconceptions About \#Hashtables (Abstract Only). In
\emph{Proceeding of the 44th ACM Technical Symposium on Computer Science
Education} (pp. 739--739). New York, NY, USA: ACM.
doi:10.1145/2445196.2445443

Paul, W., \& Vahrenhold, J. (2013). Hunting High and Low: Instruments to
Detect Misconceptions Related to Algorithms and Data Structures. In
\emph{Proceeding of the 44th ACM Technical Symposium on Computer Science
Education} (pp. 29--34). New York, NY, USA: ACM.
doi:10.1145/2445196.2445212

Pea, R. D. (1986). Language-independent conceptual`` bugs'' in novice
programming. \emph{Journal of Educational Computing Research},
\emph{2}(1), 25--36.

Pea, R. D., Soloway, E., \& Spohrer, J. C. (1987). The Buggy Path to the
Development of Programming Expertise. \emph{Focus on Learning Problems
in Mathematics}, \emph{9}(1), 5--30.

Petre, M., van der Hoek, A., \& Baker, A. (2010). Editorial.
\emph{Design Studies}, \emph{31}(6), 533--544.
doi:10.1016/j.destud.2010.09.001

Rodrigo, M. M. T., \& Baker, R. S. J. d. (2009). Coarse-grained
detection of student frustration in an introductory programming course.
In \emph{Proceedings of the fifth international workshop on Computing
education research workshop} (pp. 75--80). New York, NY, USA: ACM.
doi:10.1145/1584322.1584332

Rodrigo, M. M. T., Tabanao, E., Lahoz, M. B. ., \& Jadud, M. C. (2009).
Analyzing Online Protocols to Characterize Novice Java Programmers.
\emph{Philippine Journal of Science}, \emph{138}(2), 177--190.

Rooksby, J. (2010). ``Just try to do it at the whiteboard'':
Researcher-participant interaction and issues of generalisation. In
\emph{Proceedings of the Studying Professional Software Design (SPSD)
Conference}. San Diego, CA, USA.

Rooksby, J., \& Ikeya, N. (2012). Collaboration in Formative Design:
Working Together at a Whiteboard. \emph{IEEE Software}, \emph{29}(1), 56
--60. doi:10.1109/MS.2011.123

Rosenberg, S., Hammer, D., \& Phelan, J. (2006). Multiple
Epistemological Coherences in an Eighth-Grade Discussion of the Rock
Cycle. \emph{Journal of the Learning Sciences}, \emph{15}(2), 261--292.
doi:10.1207/s15327809jls1502\_4

Russ, R. S., Coffey, J. E., Hammer, D., \& Hutchison, P. (2008). Making
classroom assessment more accountable to scientific reasoning: A case
for attending to mechanistic thinking. \emph{Science Education},
\emph{93}(5), 875--891. doi:10.1002/sce.20320

Saussure, F. de. (1986). \emph{Course in general linguistics}. LaSalle,
Ill: Open Court.

Scherr, R. E., \& Hammer, D. (2009). Student Behavior and
Epistemological Framing: Examples from Collaborative Active-Learning
Activities in Physics. \emph{Cognition \& Instruction}, \emph{27}(2),
147--174. doi:10.1080/07370000902797379

Schoenfeld, A. H. (1987). What's all the fuss about metacognition? In
\emph{Cognitive Science and Mathematics Education} (pp. 189--215).
Lawrence Erlbaum Associates.

Schoenfeld, A. H. (1988). When good teaching leads to bad results: The
disasters of ``well-taught'' mathematics courses. \emph{Educational
Psychologist}, \emph{23}(2), 145--166. doi:10.1207/s15326985ep2302\_5

Schoenfeld, A. H. (1991). On mathematics as sense-making: An informal
attack on the unfortunate divorce of formal and informal mathematics. In
\emph{Informal reasoning and education} (pp. 311--343).

Schoenfeld, A. H. (1992). Learning to Think Mathematically: Problem
Solving, Metacognition, and Sense-Making in Mathematics. In D. Grouws
(Ed.), \emph{Handbook for research on mathematics teaching and learning}
(pp. 334--370). New York: MacMillan.

Schwartz, D. L., Chase, C. C., \& Bransford, J. D. (2012). Resisting
Overzealous Transfer: Coordinating Previously Successful Routines With
Needs for New Learning. \emph{Educational Psychologist}, \emph{47}(3),
204--214. doi:10.1080/00461520.2012.696317

Seppälä, O., Malmi, L., \& Korhonen, A. (2006). Observations on student
misconceptions---A case study of the Build -- Heap Algorithm.
\emph{Computer Science Education}, \emph{16}(3), 241--255.
doi:10.1080/08993400600913523

Sherin, B. L. (2001). How students understand physics equations.
\emph{Cognition and Instruction}, \emph{19}(4), 479--541.

Smith, J. P., diSessa, A. A., \& Roschelle, J. (1993). Misconceptions
Reconceived: A Constructivist Analysis of Knowledge in Transition.
\emph{The Journal of the Learning Sciences}, \emph{3}(2), 115--163.

Soloway, E. (1986). Learning to program = learning to construct
mechanisms and explanations. \emph{Communications of the ACM},
\emph{29}, 850--858. doi:10.1145/6592.6594

Soloway, E., Bonar, J., \& Ehrlich, K. (1983). Cognitive Strategies and
Looping Constructs: An Empirical Study. \emph{Commun. ACM},
\emph{26}(11), 853--860. doi:10.1145/182.358436

Soloway, E., \& Spohrer, J. C. (Eds.). (1989). \emph{Studying the Novice
Programmer}. Hillsdale, N.J: L. Erlbaum Associates.

Spacco, J., Hovemeyer, D., Pugh, W., Hollingsworth, J., Padua-Perez, N.,
\& Emad, F. (2006). Experiences with Marmoset: Designing and Using an
Advanced Submission and Testing System for Programming Courses. In
\emph{ITiCSE '06: Proceedings of the 11th annual conference on
Innovation and technology in computer science education}. ACM Press.

Spacco, J., Pugh, W., Ayewah, N., \& Hovemeyer, D. (2006). The Marmoset
project: an automated snapshot, submission, and testing system. In
\emph{Companion to the 21st ACM SIGPLAN symposium on Object-oriented
programming systems, languages, and applications} (pp. 669--670). New
York, NY, USA: ACM. doi:10.1145/1176617.1176665

Spacco, J., Strecker, J., Hovemeyer, D., \& Pugh, W. (2005). Software
Repository Mining with Marmoset: An Automated Programming Project
Snapshot and Testing System. In \emph{Proceedings of the Mining Software
Repositories Workshop (MSR 2005)}. St. Louis, Missouri, USA.

Spohrer, J. C., \& Soloway, E. (1986). Alternatives to construct-based
programming misconceptions. \emph{ACM SIGCHI Bulletin}, \emph{17},
183--191. doi:10.1145/22339.22369

Stevens, R., \& Hall, R. (1998). Disciplined perception: Learning to see
in technoscience. In \emph{Talking Mathematics in School: Studies of
Teaching and Learning} (pp. 107--150). Cambridge, U.K: Cambridge
University Press.

Stieff, M. (2007). Mental rotation and diagrammatic reasoning in
science. \emph{Learning and Instruction}, \emph{17}(2), 219--234.
doi:10.1016/j.learninstruc.2007.01.012

Tabanao, E. S., Rodrigo, M. M. T., \& Jadud, M. C. (2011). Predicting
at-risk novice Java programmers through the analysis of online
protocols. In \emph{Proceedings of the seventh international workshop on
Computing education research} (pp. 85--92). New York, NY, USA: ACM.
doi:10.1145/2016911.2016930

Tannen, D. (Ed.). (1993). \emph{Framing in Discourse}. New York: Oxford
University Press.

Trakhtenbrot, M. (2013). Students Misconceptions in Analysis of
Algorithmic and Computational Complexity of Problems. In
\emph{Proceedings of the 18th ACM Conference on Innovation and
Technology in Computer Science Education} (pp. 353--354). New York, NY,
USA: ACM. doi:10.1145/2462476.2465604

Traweek, S. (1988). \emph{Beamtimes and lifetimes: the world of high
energy physicists}. Cambridge, Mass: Harvard University Press.

Turkle, S., \& Papert, S. (1991). Epistemological Pluralism and the
Reevaluation of the Concrete. In I. Harel, S. Papert, \& Massachusetts
Institute of Technology (Eds.), \emph{Constructionism: research reports
and essays, 1985-1990}. Norwood, N.J: Ablex Pub. Corp.

Valentine, D. W. (2004). CS educational research: a meta-analysis of
SIGCSE technical symposium proceedings. \emph{SIGCSE Bull.},
\emph{36}(1), 255--259. doi:10.1145/1028174.971391

Van de Sande, C. C., \& Greeno, J. G. (2012). Achieving Alignment of
Perspectival Framings in Problem-Solving Discourse. \emph{Journal of the
Learning Sciences}, \emph{21}(1), 1--44.
doi:10.1080/10508406.2011.639000

VanLehn, K. (1990). \emph{Mind Bugs: The Origins of Procedural
Misconceptions}. Cambridge, Mass: MIT Press.

Wagner, J. F. (2006). Transfer in Pieces. \emph{Cognition \&
Instruction}, \emph{24}(1), 1--71. doi:10.1207/s1532690xci2401\_1

Wortham, S. (2006). \emph{Learning Identity: The Joint Emergence of
Social Identification and Academic Learning}. Cambridge: Cambridge
University Press.

Yackel, E., \& Cobb, P. (1996). Sociomathematical Norms, Argumentation,
and Autonomy in Mathematics. \emph{Journal for Research in Mathematics
Education}, \emph{27}(4), 458--477. doi:10.2307/749877

Yin, R. K. (2009). \emph{Case study research: design and methods} (4th
ed.). Los Angeles, Calif: Sage Publications.

### 2016-04-21 manuscript-body.tex
Copy core LaTeX files from Rebecca manuscript

\section{Introduction}\label{introduction}

\subsection{Acknowledgments}\label{acknowledgments}

Blinded for review. \cite{hammer_tapping_2003}

\cite{hammer_two_1989}

\clearpage

\bibliographystyle{acm}
\bibliography{bibliography}

### 2016-04-21 dissertation-export.tex
Start carving out unused dissertation material

@@
-ABSTRACT
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Title of Document: & HOW ELECTRICAL ENGINEERING STUDENTS DESIGN COMPUTER
-PROGRAMS\tabularnewline
-&\tabularnewline
-& Brian Adam Danielak, Doctor of Philosophy, 2014\tabularnewline
-&\tabularnewline
-Directed By: & Associate Professor Andrew Elby, Department of Teaching,
-Learning, Policy, and Leadership\tabularnewline
-\bottomrule
-\end{longtable}
-
-When professional programmers begin designing programs, we know they
-often spend time away from a computer, using tools such as pens, paper,
-and whiteboards as they discuss and plan their designs (Petre, van der
-Hoek, \& Baker, 2010). But, we're only beginning to analyze and
-understand the complexity of what happens during such early-stage design
-work. And, our accounts are almost exclusively about what professionals
-do. For all we've begun to understand about what happens in early-stage
-software design, we rarely apply the same research questions and methods
-to students' early-stage design work. This dissertation tries to redress
-that imbalance. I present two case studies --- derived from my 10 study
-participants --- of electrical engineering (EE) students designing
-computer programs in a second-semester computer programming course.
-
-In study 1, I show how analyzing a student's code snapshot history
-\emph{and} conducting clinical interviews tells us far more about her
-design trajectory than either method could alone. From that combined
-data I argue students' overall software designs can be consequentially
-shaped by factors --- such as students' stances toward trusting their
-code or believing a current problem is a new instance of an old one ---
-that existing code snapshot research is poorly equipped to explain.
-Rather, explanations that add non-conceptual constructs including
-affective state and epistemological stance can offer a more complete and
-satisfactory account of students' design activities.
-
-In study 2, I argue computer science and engineering education should
-move beyond conceptual-knowledge and concept deficit explanations of
-students' difficulties (and capabilities) in programming. I show that in
-doing design students do, say, write, and gesture things that:
-
-\begin{itemize}
-\item
-  Are outside the phenomenological scope of most (mis)conceptions
-  accounts of programming
-\item
-  Would be explained differently under frameworks that emphasize
-  manifold epistemological resources. Some student difficulties can be
-  recast as epistemological blocks in activity rather than conceptual
-  knowledge deficits. Similarly, some students' productive capacities
-  can be understood as epistemologically-related stances toward an
-  activity, rather than evidencing particular knowledge of specific
-  computational concepts.
-\item
-  Would suggest different instructional interventions if teachers
-  attended to the stabilizing aspects --- such as epistemological
-  dynamics --- that help these episodes of activity cohere for students.
-\end{itemize}
-
-\begin{longtable}[]{@{}l@{}}
-\toprule
-HOW ELECTRICAL ENGINEERING STUDENTS DESIGN COMPUTER
-PROGRAMS\tabularnewline
-\bottomrule
-\end{longtable}
-
-By
-
-\begin{longtable}[]{@{}l@{}}
-\toprule
-Brian Adam Danielak\tabularnewline
-\bottomrule
-\end{longtable}
-
-Dissertation submitted to the Faculty of the Graduate School of the
-
-University of Maryland, College Park, in partial fulfillment
-
-of the requirements for the degree of
-
-\begin{longtable}[]{@{}l@{}}
-\toprule
-Doctor of Philosophy\tabularnewline
-2014\tabularnewline
-\bottomrule
-\end{longtable}
-
-Advisory Committee:
-
-\begin{longtable}[]{@{}l@{}}
-\toprule
-Associate Professor Andrew Elby, Chair\tabularnewline
-Ayush Gupta\tabularnewline
-Edward F. Redish
-
-James G. Greeno\tabularnewline
-William E. J. Doane\tabularnewline
-Benjamin B. Bederson\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}l@{}}
-\toprule
-© Copyright by\tabularnewline
-Brian Adam Danielak\tabularnewline
-2014\tabularnewline
-\bottomrule
-\end{longtable}
-
-\section{Dedication}\label{dedication}
-
-To my grandparents. Anna and Josef Danielak endured and survived so that
-I could be here. Sabina Schwab loved me and made funny faces at me when
-no one else was looking. Henry Schwab continues to teach me.
-
-I also dedicate this to my brother, Jason. My name may be on the cover
-of this thing, but without his support it wouldn't exist.
-
-I love you all.
-
-\section{\texorpdfstring{\\
-Acknowledgements}{ Acknowledgements}}\label{acknowledgements}
-
-Once, in my high school Economics class, my teacher Mr. Mooney\footnote{That's
-  not a pseudonym.} told us to take out a sheet of paper. We were to
-spend 5 minutes writing down a list of all the people who were
-responsible for getting us to school that day. I started off slowly.
-Mom, for waking me up when my alarm clock didn't. Me, for remembering to
-put gas in the car.\footnote{Yes, I counted myself.} Then I thought
-about all the people who had maintained the car over the years. I
-thought about the people who had designed the car, when it was just
-modeling clay and blueprints, before a single piece of metal was bent
-into a fender.\footnote{Or, for that matter, \emph{re-bent} when mom hit
-  a lightpost in a parking lot.} Soon I was writing down the people who
-made the antifreeze that kept the engine cool, Charles Goodyear for
-developing vulcanization techniques that make modern tires work, and the
-guy at Jiffy Lube who told me---and I think I'm quoting here---''normal
-transmission fluid should be red, like fruit punch; your transmission
-fluid is brown, like Thanksgiving gravy with metal shavings in it.''
-
-I was surprised---having filled the front of the sheet and most of the
-back---when Mr. Mooney told us to stop. He said the point of the
-exercise was to show that as a society we are interdependent. That
-interdependence is key to understanding the very basics of economic
-theory: scarcity, pricing, trade, comparative advantage. But, I was
-still stuck on how many people were on my list. Most of them were people
-I'd never even met. It wasn't that I didn't care about economics; it was
-that for the first time I realized in a big way how many people were
-responsible for where I was.
-
-I can't exhaustively thank everyone who helped me make this. Moreover,
-some people, having read its contents, might wish to dissociate
-themselves entirely from it. To those who made this dissertation
-possible, I hope my thanks below will suffice. To those I leave out,
-please don't let my imperfections reflect on the job you did. Lastly, to
-those who wanted no public record of having been associated with this,
-I'm sorry.
-
-I would first like to thank my advisors, Andy Elby and Ayush Gupta.
-Nearly six years ago I made the decision to leave my Ph.D. program, my
-school, and my city behind to join Andy's program. It wasn't long until
-I was working with, joking with, writing with, and being advised by Andy
-and Ayush. From Andy, I learned to be clever, argue well, and always try
-to find the good, redeeming features of things. From Ayush, I learned
-the importance of bringing empathy and compassion to my work and my
-life. Any graduate student would be lucky to have just one of them as an
-advisor. I struck gold with both.
-
-My thanks go also to my committee. Ben Bederson graciously welcomed me
-into Maryland's human-computer interaction lab, all the while offering
-grounded and focused feedback on my work. Joe Redish has been an upbeat
-supporter of---and at times loyal opposition to---my research. My
-thoughts and insights have only grown stronger with his contributions.
-Jim Greeno has seen my intellectual development from the beginning. It
-is not an exaggeration to say parts of this dissertation were forged in
-the crucible of DC rush hour as I tried to navigate traffic and answer
-his questions about my research. I thank him for entrusting me with his
-vehicular safety. I thank him as well because at a dinner in June of
-2011 he restored my faith that some people get into academia to try and
-do good. Finally, Wil Doane helped get me into this crazy mess in the
-first place. In addition to being a mentor in computer science, Wil has
-been an unflinching champion of my ideas for as long as we've known each
-other. He first showed me what passion and innovation can look like in a
-computer science classroom; I hope this dissertation is a natural
-extension of those ideals.\footnote{And if it's not, I hope you'll
-  forgive me.}
-
-I'd like to thank my compatriots in the Physics Education Research Group
-(PERG) at Maryland. Jessica Watkins, Chandra Turpen, Vashti Sawtelle,
-and Julia (Svoboda) Gouvea were model post-docs and amazing listeners. I
-look up to all of them, which at least Julia has confirmed makes her
-uncomfortable. Tiffany Sikorski has always been impressed with my work,
-which means so much to me because I think she's a very hard person to
-impress. Lama Jaber was a constant source of warmth, compassion, and
-joy. I was very sad to see her leave for Boston, but I'm so honored to
-have her as a friend. Ben Dreyfus offered a sympathetic ear and ongoing
-encouragement when I got stuck thinking about students thinking about
-programming. Luke Conlin was my stalwart conference roommate,
-intellectual companion, cheerleader from afar, and fellow pillow fort
-architect. Gina Quan taught me how to hug; her non-hug-related work
-teaching girls to program Arduino robots has me endlessly excited. And,
-thanks to PERG's impressive alumni network, Rosemary Russ helped me
-think through some tough data analysis once I got to Madison.
-
-Eric Kuo and Mike Hull were my teammates in making sense of engineering
-students making sense of mathematics. Without Eric and Mike taping,
-analyzing, and dissecting classroom video my work couldn't exist. Eric
-in particular endured a roundtrip car ride from Boston to Bar Harbor in
-which I panicked about everything I didn't understand about our
-research.\footnote{I still panic about it. I'm just older now.}
-
-Jen Richards, Colleen Gillespie, and Jason Yip have been with me in
-Science Ed since the beginning. It was a painful change when we stopped
-taking the same courses, changed advisors, and changed geographies. But,
-we gave writing feedback to each other, supplied references to each
-other, and kept each other in good humor when things got hard. I'd like
-to single out Colleen for being the first one of us to prove this whole
-dissertation thing was possible. And, I'd like to single out Jen for
-watching over me and watching out for me. When Jen noticed I was
-struggling, she talked to me about it --- a simple act that got me into
-engineering education research and led to all of this. In so many ways,
-I wouldn't be here without her.
-
-It may buck tradition, but I'd also like to thank my undergraduate
-mentors who have been with me for almost a decade. Robert Daly first
-convinced me to come to the University at Buffalo. He later, to my
-delight, became my mentor in English and agreed to supervise my
-undergraduate thesis. It turned out the ideas in that thesis---while
-poorly argued and barely cogent---would form the basis of my
-intellectual life for the past six years.\footnote{I know. I was
-  surprised too.} Josephine Capuana backed me from the beginning. As my
-advisor she supported me no matter what I wanted to do and how many
-majors I cycled through to do it. She is also the reason I met my second
-undergraduate mentor, Kenneth Takeuchi. Dr. Takeuchi has backed me, time
-and time again, since I was a freshman in college. He and Amy Marschilok
-taught me about mentoring, inspired me to become an academic, and
-continue to serve as models for what it means to care about students.
-
-I'd like to thank the friends who have been there to prop me up and help
-me think. The first group is so big it gets its own paragraph. Robert
-Davis and Anne Davis welcomed me into their home and their family where
-everyone seemed to me to be strangely excited by what I was working on.
-Will Davis and Jamie Davis gave C help when I needed it. Russ Davis
-inspires me with his dedication to serving the country. Nora Deram and
-Mat Deram, in addition to being improbably lovely people, are both
-helping to build the future of mathematics. Travis Pond took me to the
-hallowed ground of the Apollo 1 launchpad so I could witness the
-engineering marvel of putting a rocket into space. And Nan Pond has been
-my friend, bandmate, fellow drum major, R convert, transatlantic pen
-pal, and support system for nearly 15 years. Here's to uncountably
-infinitely many more.
-
-Al Stein took me to lunch and reminded me that I've been boldly taking
-chances with my writing since I was in tenth grade. Vanessa Svihla has
-been my colleague, co-author, and co-conspirator in embracing the
-importance of design in our work. Jae Vick, Cheryl Medley, Danielle
-Champney, and Christie Veitch constantly affirmed I could do this,
-despite my doubts. Heidi Thalman and Adam Lloyd reminded me I am
-awesome, despite my stern objections to the contrary. And, a ragtag
-group of Physics Education and Science Education Researchers have been
-there for me electronically: Warren Christensen, Eleanor Close, Hunter
-Close, Dedra Demaree, Matty Lau, Sissi Li, Michael Loverude, Sandy
-Martinuk, Sam McKagan, Rachel Scherr, and Michael Wittmann. I'd
-especially like to thank Rachel Scherr, Hunter Close, and Michael
-Wittmann for being so welcoming to me at ICLS 2010.
-
-The people dearest to me get the last word. Mom, thank you for having
-me. Thank you for raising me. Thank you for loving me. Thank you for
-doing all of that when it wasn't easy. Also, thank you for that time I
-called you at 4 am when I was sick.
-
-Dad, thank you for loving me. Thank you for supporting me. Thank you for
-that time I climbed the tree in our driveway and didn't know how to get
-back down, so you helped me. And thank you for teaching me how to ride
-my bike. You might not know it, but almost everything I think about when
-I think about teaching traces back to that.
-
-Jason, I already dedicated this to you. Stop looking for more credit.
-It's unbecoming.
-
-And, guys: this is it. This is my Science Book.
-
-\section{\texorpdfstring{\\
-Table of Contents}{ Table of Contents}}\label{table-of-contents}
-
-Dedication ii
-
-Acknowledgements iii
-
-Table of Contents x
-
-List of Tables 1
-
-List of Figures 2
-
-1 Introduction 1
-
-2 Study 1 -- Retrospectively analyzing the design of a student's project
-12
-
-2.1 Abstract 12
-
-2.2 Introduction 13
-
-2.2.1 Studying the software design practices of experts 13
-
-2.2.2 The missing attention to students' design thinking 18
-
-2.2.3 The current state of research on students' code snapshots 22
-
-2.3 Context and Methods 26
-
-2.3.1 Context of the study 26
-
-2.3.2 Methods 28
-
-2.4 Code Snapshot Data and Analysis 31
-
-2.4.1 Rebecca's file-scanning solution is computationally complex 31
-
-2.4.2 Rebecca's file-scanning solution ignores an assignment directive
-33
-
-2.4.3 Rebecca repeats the same chunk of code seven times 36
-
-2.5 Interview Data and Analysis 39
-
-2.5.1 Rebecca employed fscanf loops because she was deliberately reusing
-from an Basic Programming Assignment. 40
-
-2.5.2 Rebecca repeated code because she wanted to re-use functionality
-she could trust 44
-
-2.6 Discussion 48
-
-2.6.1 Rebecca's key design decisions are made early, persist through to
-her final submission, and carry consequences 48
-
-2.6.2 Her design decisions may be influenced by framing and overzealous
-transfer 50
-
-2.7 Conclusion 53
-
-3 Study 2 -- What directs and sustains students' in-the-moment
-programming activity? 55
-
-3.1 Introduction 55
-
-3.2 Literature Review 57
-
-3.2.1 Computing culture should support a diversity of ways of knowing 58
-
-3.2.2 Misconceptions research in computing education tends to ignore
-students' productive knowledge 63
-
-3.2.3 Not all cognitive programming bugs imply a problem with the
-student 73
-
-3.2.4 Examples motivate the need for contextual-sensitivity in modeling
-programming cognition 76
-
-3.2.5 Manifold models of cognition explain context-dependence and the
-growth of expertise 84
-
-3.3 Methods and Theoretical Commitments 88
-
-3.3.1 Student population, course background, and selection 88
-
-3.3.2 Studying design as a complex phenomenon of disciplinary practice
-90
-
-3.3.3 Deploying methods to capture the complexity of early-stage design
-work 92
-
-3.3.4 Developing a revelatory case of students' early-stage design work
-95
-
-3.4 Lionel's approach to design and to programming 97
-
-3.4.1 While modifying a bike Lionel decomposed tasks, explained
-solutions to himself, and created intermediate design representations 97
-
-3.4.2 Lionel's approach to programming resonates strongly with the
-design stance he held when working on his bike 105
-
-3.4.3 Gestural, inscriptional, and verbal in-interview evidence reveals
-Lionel's resources for structuring a program 109
-
-3.4.4 We can trace continuities in Lionel across different kinds of
-design activities 126
-
-3.5 Rebecca's approach to programming 127
-
-3.5.1 Rebecca's struggles on the iTunes project make sense in light of
-some prior programming experiences she had 129
-
-3.5.2 Rebecca has productive capacities for making progress on software
-design work 150
-
-3.6 Conclusion 176
-
-3.6.1 Students' early-stage design activity reveals patterns outside the
-explanatory scope of (mis)conceptions accounts 176
-
-3.6.2 Rebecca had epistemological resources to support expert-like
-practices, but she framed those practices differently 177
-
-3.6.3 Students displayed a diversity of approaches to programming in the
-moment 181
-
-3.6.4 Dynamic epistemological models can offer a lens for reforming
-assessment and instruction. 185
-
-4 Conclusion 188
-
-4.1 We should think carefully about what students' programming design
-knowledge means for assessment 188
-
-4.2 What We Might Change About Classroom Practice 191
-
-4.2.1 Instructors could look beyond content to understand student
-difficulties 191
-
-4.2.2 Instructors could use code history to inform interventions 192
-
-4.2.3 Instructors could establish a norm of asking why a design choice
-makes sense 194
-
-4.3 Final Remarks 195
-
-5 Appendix 1 -- Transcript conventions 197
-
-6 Appendix 2 -- Visual conventions for gestures 198
-
-7 Appendix 3 -- Transcript of Rebecca's pseudo-code episode
-\emph{without} gesture codes 199
-
-8 Appendix 4 -- Conceptual knowledge frameworks in computing don't tell
-us much about how Lionel structured an in-interview program 202
-
-8.1 Analyzing Lionel's Solution From a Conceptual Knowledge View 202
-
-9 Appendix 5 -- Neverly-Asked Questions (NAQs) 205
-
-9.1.1 Neverly Asked Questions (NAQs) about my conceptual analysis of
-Lionel's code 205
-
-9.1.2 Neverly-asked questions about Lionel's verbal pseudo-code
-description 207
-
-10 References 213
-
-\section{List of Tables}\label{list-of-tables}
-
-Table 1 -- An overview of my research participants and the types of data
-I collected 9
-
-Table 2 -- Five different interpretations of semantically equivalent
-programming statements in the same language 78
-
-Table 3 -- Three syntactically-similar statements with very different
-semantics 80
-
-Table 4 -- Using room lighting configurations to think about
-epistemological framing 87
-
-Table 5 -- Comparing Lionel's words and actions across different stages
-of ``programming'' over the same putative section of his program. I have
-bolded some language-specific syntax. 122
-
-Table 6 -- Comparing Lionel's design activity across contexts 126
-
-Table 7 - Rebecca describes an array of structures using gestures
-(Interview 4 of 5, April 6, 2012) 153
-
-Table 8 -- Rebecca says gets lost on the array of pointers pointing to
-track names (Interview 4 of 5, April 6, 2012) 156
-
-Table 9 -- Rebecca starts talking out pseudo-code and asks for the pen
-to write it 167
-
-Table 10 -- Rebecca's verbal/gestural overview of looping through the
-input track titles (Interview 4 of 5, April 6, 2012) 169
-
-Table 11 -- Rebecca concludes her visual and gestural explanation for
-scanning in track titles (Interview 4 of 5, April 6, 2012) 174
-
-Table 12 -- Comparing Lionel's and Rebecca's views toward pseudo-code
-178
-
-\section{\texorpdfstring{\\
-List of Figures}{ List of Figures}}\label{list-of-figures}
-
-Figure 2‑1 -- A side-by-side comparison of two blocks of Rebecca's
-flight-scanning code 37
-
-Figure 3‑1 -- An comic-based overview of my methods for capturing
-students' design practices 95
-
-Figure 3‑2 -- An overview of Lionel's representational cascade and the
-features and affordances of each representational layer. 111
-
-Figure 3‑3 -- Lionel's final source code for the range-finding prompt
-113
-
-Figure 3‑4 -- Lionel makes a cycloid/helix gesture as he says ``parse
-through this array'' 117
-
-Figure 3‑5 -- Rebecca's compilation activity over time for Project 2.
-Each compile a student initiates creates a commit, so long as there has
-been a change to the underlying code. The height of each bar maps to the
-number of commits recorded that day. The red line charts
-\emph{cumulative} commits over time. The red line is steepest in periods
-of frequent activity and shallowest in periods with little or no compile
-activity. 144
-
-Figure 3‑6 -- The instructor required students to use this data
-arrangement for storing information in the music server. The scheme uses
-an array of pointers to represent the track names of an album. 151
-
-Figure 3‑7 -- I write down what I understand as Rebecca explains the
-overall data structure of albums to me 155
-
-Figure 3‑8 -- The candidate pseudo-code Rebecca dictated to me for
-declaring an array of pointers 166
-
-Figure 3‑9 -- Rebecca's pseudo-code for scanning in and storing track
-names 167
-
-Figure 8‑1 -- Highlighting the conceptual features of Lionel's code
-given Elliott Tew's (2010) outline of CS 1 concepts 204
-
-\section{}\label{section}
-
-\section{Introduction}\label{introduction}
-
-Seymour Papert's (1980) book \emph{Mindstorms} has a curious section. Or
-at least I thought so. The book collects years of Papert's thoughts,
-observations, and experiments with teaching children at school to
-program. The famous\footnote{Bearing in mind that ``fame,'' in the
-  subfield that is computing education, is a relative thing.} Turtle and
-LOGO --- the language spoken and understood by it --- let students
-express geometrical ideas in the form of procedures and debug procedures
-using the spatial affordances of geometry. My school did not have LOGO,
-so I came to see the beauty of programming much, much later in life.
-But, for many children in Papert's studies, LOGO became a venue for
-expression and creativity.
-
-The curious section in \emph{Mindstorms} is one in which Papert
-describes ``structured programming,'' a technique for breaking down
-large procedures into ``mind-sized bites.'' In short, it's possible to
-write lots of programs ahierarchically as a list of instructions:
-
-FORWARD 50;
-
-RIGHT 120;
-
-FORWARD 50;
-
-RIGHT 120;
-
-FORWARD 50;
-
-RIGHT 120;
-
-This little program would make a LOGO turtle draw an equilateral
-triangle. By telling the turtle how much to go forward (in distance),
-how much to turn (in degrees), and in which direction to turn (in this
-case right), one can make a turtle trace out this equilateral triangle
-as a sequence of steps.
-
-But, this program doesn't have much structure. And, as the program grows
---- say, to accommodate having the turtle draw a simple house ---
-managing the program becomes more complicated. It's possible, in
-principle, to have dozens or hundreds of simple little statements and it
-can become more and more difficult to figure out which part of the
-program drew which part of the picture.
-
-So, I'll propose a modification:
-
-TO MAKEATRIANGLE:
-
-FORWARD 50;
-
-RIGHT 120;
-
-FORWARD 50;
-
-RIGHT 120;
-
-FORWARD 50;
-
-RIGHT 120;
-
-What I've done is define a \emph{procedure} in LOGO. In a way, the
-procedure bundles together the list of discrete steps and offers a sense
-of \emph{structure} to the program. The procedure is now also something
-I can compose to make more complicated things. Now, for example, I can
-draw a bunch of triangles that all share a vertex and make a kind of
-pinwheel:
-
-MAKEATRIANGLE;
-
-RIGHT 120;
-
-MAKEATRIANGLE;
-
-RIGHT 120;
-
-MAKEATRIANGLE;
-
-RIGHT 120;
-
-If I apply the same kind of thinking to the pinwheel that I did to the
-triangle, I might do this:
-
-TO MAKEAPINWHEEL:
-
-MAKEATRIANGLE;
-
-RIGHT 120;
-
-MAKEATRIANGLE;
-
-RIGHT 120;
-
-MAKEATRIANGLE;
-
-RIGHT 120;
-
-Now imagine two students, both of whom are trying to get the turtle to
-draw a pinwheel.
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Alan's Code & Ada's Code\tabularnewline
-\midrule
-\endhead
-\includegraphics[width=2.00000in,height=5.04861in]{media/image1.emf} &
-\includegraphics[width=2.40278in,height=4.62500in]{media/image2.emf}\tabularnewline
-\bottomrule
-\end{longtable}
-
-The picture Alan's turtle makes should be identical to the picture Ada's
-turtle makes.\footnote{Unless I screwed up, which is possible. If so,
-  that's my fault; don't hold Alan and Ada accountable.} So, if all we
-care about is the final art, Alan and Ada are making the same thing.
-They're indistinguishable as students from the perspective of what the
-turtle draws on paper.
-
-But, clearly Alan's code and Ada's code aren't the same. Ada's program
-evinces what Papert (1980) would call \emph{structured programming}:
-pinwheels are composed of triangles, while triangles are composed of
-straight line segments.\footnote{Not to get pedantic on you, but yes,
-  Ada could have written a program that was perfectly syntactically
-  valid --- and produced the same picture --- while choosing different
-  and possibly bogus procedure names. By renaming functions, for
-  example, Ada could have forced me to say that in her program
-  ``unicorns are composed of Cheetos, and Cheetos are composed of
-  dinosaur fingers.'' And, while \emph{my restatement} would make no
-  sense, neither LOGO nor the turtle would have a problem interpreting
-  Ada's procedure to make what Ada would call a pinwheel. The dinosaur
-  might have a problem, I guess.} Alan's code doesn't compose things.
-Every line in Alan's code is a direct instruction to the turtle and
-there is no sense of hierarchy. Instead, there's just a serial flow of
-instructions. To think more about the differences between Alan and Ada,
-let's go back to Papert.
-
-Papert (1980) discusses structured programming in the case of two
-children. One child, Robert, embraced structured programming. Robert
-extolled the style: ``see, all my procedures are mind-sized bites''
-(Papert, 1980). And, indeed, programmers today are given similar advice
-in books on professional programming practice (Martin, 2009). But, Keith
-was a child who resisted. His programs remained ahierarchical;
-``featureless'' in Papert's words. Debugging was harder for Keith, too,
-because it was harder to locate problems in a program devoid of
-structure.
-
-If Keith was rather stubbornly sticking to his ahierarchical style,
-should a teacher have proactively \emph{instructed} Keith to change?
-Papert says no. Rather, as Keith encountered troublesome errors
-debugging his code, he would meet with familiar advice:
-
-When a child in this predicament asks what to do, it is usually
-sufficient to say: ``You know what to do!'' And often the child will
-say, sometimes triumphantly, some-times sheepishly: ``I guess I should
-turn it into subprocedures?'' The ``right way'' was not imposed on
-Keith; the computer gave him enough flexibility and power so that his
-exploration could be genuine and his own. (Papert, 1980)
-
-But, largely, that's where Papert leaves things. As Papert would have
-it, most children will inevitably run into increasing frustration when
-they debug. When children hit those problems, it's up to the teacher to
-patiently remind the student that, in a sense, ``we've covered this.''
-After repeatedly hitting this wall most students will see the light ---
-eventually --- and come to save themselves from frustration by adopting
-structured programming.
-
-\emph{But what if they don't?}
-
-I knew from decades of research in science education that students do
-not always adopt the practices we want them to. As Hammer (1994)
-essentially put it, \emph{not all novices go on to develop expertise,}
-and many cognitive formulations of the novice-expert continuum didn't
-(and still don't) really explain why. Was structured programming
-something more than a practice of keeping code tidy and easing
-debugging? \emph{Yes}, I believed so. Structured programming wasn't the
-computer science equivalent of ``always labeling your units'' in
-science\footnote{Don't misunderstand me; labeling units is important.
-  The fact that early versions of LOGO didn't drives me \emph{nuts}.
-  But, if you asked a bunch of sophisticated chemists what are the chief
-  intellectual practices of their discipline, I don't think ``labeling
-  units'' will top the list. If you ask Abelson \& Sussman (1996) what
-  the chief intellectual practice of computer programming is, they'll
-  tell you it's managing the complexity of large software systems.}; it
-reflected a larger view on how we think about procedure and a
-concomitant willingness to compose big complex ideas out of small
-atomically-understandable ones (Abelson \& Sussman, 1996; Papert, 1980).
-
-Back to Alan and Ada. With all this talk about structured programming,
-we've mostly been focusing on \emph{code}: how it's structured
-differently and what affordances and drawbacks result from the design.
-We still don't have a sense of how Alan and Ada actually built their
-programs. We can't know, just from these snippets, whether Ada's code
-started with a compositional structure. Perhaps Ada started just the way
-Alan did, so there was some intermediate state of Ada's code before she
-refactored it when it was just as ahierarchical and featureless as
-Alan's code. Or, perhaps Alan started by \emph{trying} to structure his
-program, but he ran into some errors. Maybe Alan's literal code is the
-result of a design retreat late in the game. It's not that he doesn't
-\emph{appreciate} structured programming; he just couldn't get it to
-work and, feeling pressured to produce a proper pinwheel, resorted to
-taking out the structure and leaving just concrete instructions for the
-Turtle.
-
-I had questions.
-
-Would all students come to structured programming after banging their
-heads enough against a buggy featureless program? \emph{Maybe}. Would
-they do it on a timescale we could see? \emph{I wasn't sure}. Was it
-possible students could get through one, or even multiple semesters of
-programming without ever embracing structured programming for
-themselves? \emph{Quite possibly}. Outside of Papert's schools, was it
-possible for research to not only track but explain why some students
-used structured programming and other students didn't? Were there Alans
-and Adas out there in real life whose complex stories could help us
-understand more about how students learn structured programming?
-
-This dissertation is my stab at beginning to answer those questions. The
-focus of my study is a programming course for electrical engineering
-students. I chose that course and population for two reasons. First and
-foremost, I had preliminary evidence suggesting students were emerging
-from the course with wildly different mindsets about its relationship to
-engineering and, by extension, design. One student, Larry, saw
-structured programming as a natural extension of the way he tried to
-make sense of the physical world. Another student dismissed the class
-entirely as ``not even a real engineering course.'' Second, my
-connections to the school of engineering allowed me relatively easy
-access to the class.
-
-Over two semesters I observed lectures and interviewed students. In
-interviews, I was able to use video recordings, screen capture software,
-and an electronic pen to capture what students said, gestured, typed,
-and wrote when they programmed. I later augmented the study by tracking
-students' code histories. I co-developed a simple automated system to
-create version-controlled repositories of the multi-week projects
-students worked on for the course. Each time a student compiled, our
-system captured a snapshot of the entire codebase (called a ``commit''
-in Git, the versioning system we used). With each snapshot we also
-captured the input students passed to the compiler and any compile-time
-messages (including errors and warnings) they would have seen.
-
-Table 1 shows a summary of all the participants in my studies and the
-kind of data I collected.\footnote{The ``Inscriptions'' column is
-  currently blank because I've been having trouble accessing that
-  portion of my dataset. I hope to resolve my technical difficulties so
-  I can provide inscription counts in later revisions of this
-  dissertation.} The solicitation process began at the beginning of each
-semester. During the second class lecture I read a 3-minute IRB-approved
-speech explaining that I was studying how students understand
-programming code. Students who expressed interest were invited to
-participate in interviews, and for each interview they completed they
-were paid \$15. Interviews were scheduled opportunistically based on
-when students were available. In total, the data corpus comprises more
-than 20 hours of clinical interviews and more than 2,500 distinct code
-snapshots. Of the 10 students who participated, I chose to focus on
-Rebecca (who appears in both study 1 and study 2) and Lionel (who
-appears in study 2).
-
-\protect\hypertarget{ux5fToc252445955}{}{}Table -- An overview of my
-research participants and the types of data I collected
-
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-\textbf{Semester} & Participant & Interviews\footnote{All interviews
-  were videorecorded with the exception of Toby's. Toby requested to be
-  audiorecorded only.} & Screencaptures & Inscriptions\footnote{At
-  present, technical difficulties prevent me from reporting an exact
-  count of inscriptions collected.} & Code Snapshots\tabularnewline
-\midrule
-\endhead
-\textbf{Fall 2011} & Lionel & 1 & 1 session & & ---\tabularnewline
-& CJ & 1 & --- & & ---\tabularnewline
-& Donna & 2 & 2 sessions & & ---\tabularnewline
-& Sam & 1 & --- & & ---\tabularnewline
-& Toby & 2 & --- & & ---\tabularnewline
-& Will & 1 & 1 session & & ---\tabularnewline
-\textbf{Spring 2012} & Isaac & 3 & 1 session & & 434\footnote{Due to an
-  unidentified error in Isaac's code snapshot history, we lack data
-  between February 21 and April 12. we know in aggregate what changed in
-  his code between those two dates; we just can't resolve that aggregate
-  change down into individual changes for each time Isaac compiled
-  between those dates.}\tabularnewline
-& Dana & 3 & 2 sessions & & 878\tabularnewline
-& Natalie & 4 & 1 session & & 262\tabularnewline
-& Rebecca & 5 & 3 sessions & & 959\tabularnewline
-\bottomrule
-\end{longtable}
-
-This dissertation is structured into two studies that analyze data from
-the corpus. Study 1 asks what we can learn if a student submits a
-program like Alan's. In other words, when one student's final code
-submission is structured in peculiar ways (or not at all), can research
-help us retrospectively recover the story of how that design came to be?
-Study 2 in effect asks what directs and sustains Alan and Ada's
-in-the-moment programming activity. I analyze rich data --- including
-talk, gestures, inscriptions, and screen captures --- of two real
-students. The analysis tries to understand what constitutes the
-approaches they take when they code. Along the way, study 2 offers an
-illustration of why strictly ``conceptual'' accounts of students may not
-be enough to explain certain classes of phenomena in learning to
-program.
-
-\section{Study 1 -- Retrospectively analyzing the design of a student's
-project}\label{study-1-retrospectively-analyzing-the-design-of-a-students-project}
-
-\subsection{Abstract}\label{abstract}
-
-This paper focuses on a historically understudied area in computing
-education: attending to students' \emph{design thinking} in
-university-level introductory programming courses. We detail the case of
-Rebecca, a first-year Electrical Engineering student taking a required
-2\textsuperscript{nd}-semester programming course in C. Our analysis
-focuses on two related aspects of Rebecca's code for a multi-week
-project:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  The origin, nature, and evolution of unusual structural and behavioral
-  features of Rebecca's code
-\item
-  The subtle, yet complex reasons that led Rebecca to make particular
-  design choices in her code
-\end{enumerate}
-
-Our data comes from ethnographic observation of Rebecca's class,
-fine-grained compile-time snapshots of Rebecca's codebase, and
-semistructured interviews with Rebecca. We first present an analysis of
-the compile-time snapshots, detailing Rebecca's unusual use of
-file-scanning loops and her seven-fold repetition of a particular code
-chunk (once for each day of the week). We then augment that analysis
-with data from semi-structured interviews with Rebecca, which reveal
-that affect (Eynde \& Hannula, 2006; Hannula, Evans, Philippou, \& Zan,
-2004) and framing (Hammer, Elby, Scherr, \& Redish, 2005; van de Sande
-\& Greeno, 2012) offer substantial explanatory power for understanding
-why Rebecca made particular design choices.
-
-\subsection{Introduction}\label{introduction-1}
-
-\subsubsection{Studying the software design practices of
-experts}\label{studying-the-software-design-practices-of-experts}
-
-In 2010, the journal \emph{Design Studies} devoted an entire issue to
-studies of how professional software engineers design complex systems.
-The journal featured five different research perspectives on the same
-dataset: videos of three different professional software engineering
-teams trying to design a traffic simulator. The existence of the journal
-issue, and the investigations contained therein, were motivated by what
-the issue's editors saw as a growing need to know more about how
-software design gets done in practice. In particular, the editors
-contended, fields of design studies, interaction analysis, and
-human-computer interaction don't know enough about how software
-engineers use representations and collaborative exchanges to organize
-the beginning phases of a design:
-
-{[}d{]}uring formative design, software engineers spend a great deal of
-time engaging in creative, exploratory design thinking using pen and
-paper or a whiteboard---\emph{whether alone or in a small group}.
-However, not enough is known about how software designers work in such
-settings. What do designers actually do during early software design?
-How do they communicate? What sorts of drawings do they create? What
-kinds of strategies do they apply in exploring the vast space of
-possible designs? (Petre et al., 2010, p. 533; my emphasis added)
-
-Research in that issue takes on a variety of challenges in the study of
-expert software design practice. One challenge, for example, is that of
-understanding how engineers process, prioritize, and cope with design
-requirements. Ball, Onarheim, and Christensen (2010) observed that
-engineers deploy mixed strategies, developing solutions breadth-first
-for easy problems and depth-first for more complex problems. They also
-found that the more complex a requirement became, the more likely
-engineers were to create speculative simulations (through talk and
-representations) about how a system might work to solve that problem
-(Ball et al., 2010).
-
-Looking at design sessions longitudinally, Baker and van der Hoek (2010)
-explored the shape and trajectory of how ideas generated in the design
-process develop and relate to one another. Those researchers found that
-roughly a third of the ideas discussed in a typical design session
-``were reiterations or rephrasings of previously stated ideas'' (Baker
-\& van der Hoek, 2010, p. 604). While it is perhaps frustrating that
-ideas would be repeated so much, the authors argue such repetition can
-be viewed as a kind of continual revisitation to make sure a proposed
-design coheres:
-
-Rather than representing a failure on the part of the designers, this
-repetition seems to be a necessary character of successful design
-sessions. Each time an idea is resurrected it is placed them {[}sic{]}
-in a new context, and compared to different aspects of the system. In
-this way, a concept of compatible, elegant design ideas is slowly
-converged upon. (Baker \& van der Hoek, 2010, p. 607)
-
-Perhaps most germane to the work presented herein, Jackson (2010)
-explores the role of \emph{structure} in software design. Jackson's
-meaning in using the word ``structure'' is broad. It can best be
-described as the arrangement of and relationships between the elements
-of a system. But, the broadness is deliberate, because Jackson's overall
-argument is that the work of software often involves a coordination of
-different structures, some of which drive the organization of others. In
-the case of Jackson's (2010) study, designers worked with various kinds
-of structured sets to build software that could simulate traffic
-patterns in a city section. The work of design then, in part, becomes
-that of defining and coordinating structural relationships at various
-levels of abstraction:
-
-\begin{itemize}
-\item
-  The \emph{system-to-be-modeled}, in this case a system of roads and
-  intersections on which one can simulate traffic. Schematically, this
-  system can be structured on the Cartesian plane of a map. Its
-  representation (a diagrammatic map-like network of streets) works to
-  organize how the system-to-be-modeled looks in the real world.
-\item
-  The \emph{actors} present in the system-to-be-modeled. In this case,
-  that means abstracting from the map to all of the \emph{actors} that
-  simulator software would need to send data to and get data from:
-  including traffic signal units, traffic signal controllers, drivers,
-  and vehicle sensors.
-\item
-  The \emph{meta-entities} that must exist to execute the actual
-  simulation. Crucially, these abstractions may not be the same as the
-  actors \emph{in} the system.\footnote{For example, suppose we were
-    building an AI for chess. The ``entities in the system'' are the
-    pieces (rooks, queens, knights, etc.) and the actual board. But,
-    those pieces alone aren't enough to create a chess AI. One must also
-    create entities that make decisions, move the pieces, maintain the
-    state of the board, determine the legality of moves, etc. Thus,
-    bishops may be ``entities in the system'' that know how to move
-    (only along diagonals, up to 8 spaces). But, bishops need another
-    meta-system entity to tell them \emph{when} and \emph{where} to
-    move. That distinction is important because it means simulating
-    chess involves much more than building a board and pieces, even
-    though the board and pieces may constitute the only outward-facing
-    aspects of a chess AI.} In this case, additional computational
-  entities must be added to carry out a traffic simulation, such as an
-  arrivals model and a simulation clock (Jackson, 2010, p. 559)
-\item
-  The \emph{programming code} that embodies the data and behaviors of
-  the entities and meta-entities. While to some degree aesthetic, the
-  symbolic structure of code is in many ways driven by decisions about
-  structural relationships in other levels of abstraction.
-\end{itemize}
-
-How engineers develop and interact with even the first three kinds of
-structural representations can profoundly influence the nature and
-quality of the code they produce. As Jackson (2010) observes (in notable
-contrast to (Baker \& van der Hoek, 2010)), early negotiations of
-structure led to design decisions that were \emph{not} revisited later:
-
-The traffic simulation problem was of a kind unfamiliar to all three
-design teams. All three quite rightly tried immediately to assimilate
-the problem to something they already knew. Two teams took the view that
-the problem was an instance of the Model-View-Controller (MVC) software
-pattern; the third identified the problem as `like a drawing program'.
-Unfortunately, these hasty classifications were inadequate; but in every
-case they were accepted uncritically and never explicitly questioned.
-(Jackson, 2010, p. 564)
-
-Ultimately, careful study of how software engineers design shows us the
-enormous complexity involved in producing effective and thorough code.
-And, much of that complexity is both afforded and constrained by talk,
-representational infrastructure, and knowledge of larger-scale solution
-patterns. It's striking, then, that this research perspective on how
-programing design work gets done---for example, the detailed
-ethnomethodological analysis by Rooksby and colleagues on how engineers
-use a whiteboard (Rooksby \& Ikeya, 2012; Rooksby, 2010)---seems largely
-underrepresented in studies of how students learn to program. Instead,
-the latest generation of research on student learning in programming has
-focused much more extensively on questions like \emph{how can we assess
-and mitigate students' difficulty in programming?} than it has on
-questions like \emph{how do students learn and display evidence of
-design thinking in programming?}
-
-\subsubsection{The missing attention to students' design
-thinking}\label{the-missing-attention-to-students-design-thinking}
-
-It's difficult to point directly to the absence of a research direction
-regarding how students design software. Rather, one can instead show
-that most computing education research focuses elsewhere. In 2004,
-Valentine conducted a meta-analysis of 20 years of conference papers at
-SIGCSE\footnote{SIGCSE, as used here, is the abbreviation for the annual
-  meeting of the Association for Computing Machinery (ACM) Special
-  Interest Group for Computer Science Education. SIGCSE is one of the
-  premier venues for computer science education.}. His analysis focused
-on papers focused on first-year university courses, which as of 2004
-comprised about 1/3 of the total papers presented at SIGCSE. Valentine's
-resulting taxonomy reveals how---in a conference about computer science
-education---there is at best a minimal focus on how students engage in
-design thinking. In the decade from 1994-2003:
-
-\begin{itemize}
-\item
-  About 24\% of conference papers were ``Marco Polo'' studies, which
-  Valentine (2004) describes as teacher- or administrator-centered
-  accounts of describing a new method, course, curriculum, or approach
-  and documenting how well it works.
-\item
-  Another 42\% of papers were classified as either ``Nifty'' or
-  ``Tools,'' and centered on either innovative assignments to give to
-  students or software tools to augment learning, instruction, and
-  assessment.
-\item
-  Only 22\% of studies were classified as ``Experimental,'' a broad
-  category which includes everything from quasi-experimental comparisons
-  of teaching methods to ethnographic interviews with students solving
-  protocol problems.
-\end{itemize}
-
-Joy et al. (2009) offer a similar, more recent view of the state of
-computing education research. Their 2009 survey included both conference
-papers and journals, from which they constructed a taxonomy of over
-3,500 papers in two overlapping fields: ``education in computer
-science'' and ``computers for education'' (Joy et al., 2009, p. 112).
-While Joy et al.'s (2009) classification scheme isn't identical to that
-of Valentine's (2004), the most ready parallel to Valentine's
-``Experimental'' studies are what Joy et al. call \emph{theoretical
-pedagogy}:
-
-The focus of the paper is principally educational, and reports results
-grounded in education theory (i.e. explicitly references, discusses and
-applies pure education theory). This category is used for ``Learning
-Psychology'' where the ``learning'' is the predominant focus of theory,
-such as Bruner and Vygotsky. (Joy et al., 2009, p. 114)
-
-As a percentage of the pool of 3,500 papers, research classified as
-``theoretical pedagogy'' totaled less than 5\%.\footnote{Joy et al.
-  (2009) don't report raw quantitative numbers, so I estimated this
-  percentage by measuring their data graphic on p. 117.} Admittedly, the
-authors don't disaggregate their final results by ``education in
-computer science'' vs. ``computers for education.'' Still, it seems
-reasonable to assume that if even half the publications were from
-``computers for education'' and none of that half focused on theoretical
-pedagogy, that still means less than 10\% of publications \emph{in}
-education in computer science foreground learning as a primary
-phenomenon.
-
-If one is looking for the kinds of studies that would involve deep,
-fine-grained analysis of learning via students in action, the pool gets
-even thinner. In a survey of 79 papers from the International Computing
-Education Research Conference (ICER)---published between 2003 and
-2009---Malmi et al. (2010) found that of the 79\% of papers that
-included any mention of a theoretical framework, 39\% were ``surveys.''
-(p. 8). ``Experimental'' papers account for another 15\% of papers that
-explicitly identified a framework. The total number of papers that
-employed case study methodology, ethnography, phenomenology, \emph{or}
-phenomenology was 10, or approximately 14\% (Malmi et al., 2010, p. 8).
-But, since Malmi et al. (2010) do not identify those case
-study/ethnography/phenomenology/phenomenograpy papers by name, it's
-difficult to assess just how closely those papers hew to their nominal
-frameworks.
-
-Indeed, one of the most useful collections of research on the psychology
-of how people design software isn't recent; it's almost thirty years
-old. In the 1980s, Elliot Soloway, James Spohrer, and others were trying
-to establish what expert programming behavior looked like (Adelson \&
-Soloway, 1985; Soloway, 1986), and what kept novice programmers from
-developing expertise (Bonar \& Soloway, 1983, 1985; Mayer, 1981; Pea,
-Soloway, \& Spohrer, 1987; Soloway \& Spohrer, 1989). One of the most
-fundamental arguments to emerge from Soloway and Spohrer's work was a
-refutation of the claim that novices make programming errors because
-they don't understand the elements of a programming language:
-
-Our empirical study leads us to argue that (1) yes, a few bug types
-account for a large percentage of program bugs, and (2) no,
-misconceptions about language constructs do not seem to be as widespread
-or as troublesome as is generally believed. Rather, many bugs arise as a
-result of plan composition problems---difficulties in putting the
-``pieces'' of a program together (see sidebar)---and not as a result of
-construct-based problems, which are misconceptions about language
-constructs. (Spohrer \& Soloway, 1986, p. 401)
-
-Ultimately, Spohrer and Soloway (1986) found strong support for two
-kinds of problems that transcend any particular programming language:
-
-Interpretation Problem: When novices read a programming assignment, they
-do not always infer the correct interpretation from the specifications.
-
-Composition Problem: Novices may not detect negative interactions
-between sections of code that are locally correct, but globally
-incorrect. For example, the code to perform the output may be correct,
-but in the wrong place in the program. (Spohrer \& Soloway, 1986, p.
-191)
-
-Despite Spohrer and Soloway's (1986) finding that program composition
-accounts explain more than do language construct accounts of student
-error, research since has persisted in documenting students'
-``misconceptions'' about programming language constructs (Fleury, 1991,
-1993, 2000; Herman, Kaczmarczyk, Loui, \& Zilles, 2008; Kaczmarczyk,
-Petrick, East, \& Herman, 2010). Moreover, in a 2006 paper analyzing the
-programming behavior of hundreds of novices, Jadud argues that a focus
-on syntax-level analyses can help improve research and practice:
-``{[}b{]}y identifying and understanding the behaviour of novices
-learning to program, we hope to build up to later making sound cognitive
-and constructivist inquiries and recommendations'' (Jadud, 2006, p. 80).
-
-In the next section, I argue that a growing syntax- and language-focused
-trend in computing education research is using bigger and richer
-datasets to push us farther away from the kind of fine-grained studies
-that help us understand design thinking in software creation. First, I
-explain how code-snapshotting systems help us capture changes to
-students' code over time. Then, I outline why---in their current
-usage---such systems fail to help us consider larger design thinking
-issues that might be at play for students. I conclude my introduction
-with a claim that we can use the same code-snapshot repositories we've
-been mining for syntactical analyses to look for deeper explanations of
-why students struggle with software design.
-
-\subsubsection{The current state of research on students' code
-snapshots}\label{the-current-state-of-research-on-students-code-snapshots}
-
-A growing trend in computer science education research is the collection
-and analysis of code snapshot data---records of the state of and changes
-to students' code as they develop it (Jadud, 2006; Rodrigo, Tabanao,
-Lahoz, \& Jadud, 2009; Spacco, Pugh, Ayewah, \& Hovemeyer, 2006). Though
-specific implementations differ, the general strategy in such projects
-is that a student event (typically compiling code or saving a file)
-triggers a procedure that creates a record containing the entire content
-of all of a student's relevant files, as well as associated metadata
-(time of save/compilation, for example, and any compiler errors that may
-have been generated). Mining data from such snapshotting systems has led
-to large-scale documentation of common student errors (Spacco, Pugh, et
-al., 2006), the development of compile-time detectors to catch common
-student errors (Spacco, Strecker, Hovemeyer, \& Pugh, 2005), and the
-proposal of formative assessment models to predict student success
-(Jadud, 2006; Tabanao, Rodrigo, \& Jadud, 2011). Building on existing
-momentum, some researchers are actively pushing for a continued scale-up
-of how we collect code snapshot data. One current proposal even calls
-for creating an international database by collecting snapshot data from
-thousands of introductory programming students worldwide (Kölling \&
-Utting, 2012).
-
-What's common to these threads of research is how they mine the data. In
-most applications of code snapshot\footnote{What I call ``code
-  snapshot'' research may be more formally called ``online protocol''
-  research (Jadud, 2006; Rodrigo, Tabanao, Lahoz, \& Jadud, 2009;
-  Tabanao, Rodrigo, \& Jadud, 2011). ``Online'' here isn't mean to mean
-  the sense of globally-connected computers, but rather that student
-  materials are collected in a minimally-intrusive way \emph{while
-  students work}.} research, the aim is to average across events and
-sessions to arrive at a characteristic measure of a student's behavior.
-Jadud (2006), for example, develops the idea of a student's ``Error
-Quotient,'' a measure of how effectively a student addresses
-compile-time errors in their code.\footnote{A useful way of thinking
-  about Jadud's (2006) error quotient is that the computation believes
-  in second chances. It doesn't penalize students for making errors; it
-  penalizes students if they don't fix a given error before compiling
-  the same piece of code \emph{again.}} Rodrigo and colleagues extend
-the idea of error quotient (Rodrigo et al., 2009) and introduce similar
-measures such as students' ``compilation profiles'' (Rodrigo et al.,
-2009), ``frustration profiles'' (Rodrigo \& Baker, 2009), and ``error
-profiles'' (Tabanao et al., 2011). Because these measures can be
-collected in real-time as students code, they offer the potential to
-identify at-risk students (Tabanao et al., 2011) and respond with early
-interventions (Jadud, 2006).
-
-But, for all the data these methods collect their focus is primarily
-what Jadud (2006) calls students' ``syntactic'' struggles: the challenge
-of articulating well-formed statements a compiler can properly parse.
-This focus on syntax-level struggles leads to both a methodological and
-theoretical trade-off: we can see in detail how students struggle with
-wording, but we see less of how they struggle with the meaning and
-intent of their code. Syntactical analyses necessarily ignore the
-specific \emph{content} of students' code because they abstract
-meta-information about the event: error type, error location, frequency
-of error.
-
-As an analogy, suppose that instead of studying computer science
-students we were studying screenwriting students. Further suppose we
-have a system to track data whenever students save their screenplays and
-scripts. For each save, we get a copy of the entire script at that time.
-We can also run an automated analysis to check whether they've properly
-formatted slug lines\footnote{In a screenplay, a slug line gives
-  information about where and when a scene takes place. For example:
-  ``INT. FORTRESS OF SOLITUDE -- DAY.''}, put character names in
-all-capital letters, properly numbered scenes, etc. We might even know
-where students are when they write (coffee shop, library, home, etc.).
-If we mine that data, we could learn something about students'
-\emph{screenwriting behavior}, but only at the level of how they
-struggle with screenwriting syntax. Using only error-based data it's
-much harder to answer questions such as:
-
-\begin{itemize}
-\item
-  How does this student construct a scene or handle dialog in their
-  script?
-\item
-  Does the student seem to have a grasp of pacing, story beats, and
-  efficient exposition?
-\item
-  How does their writing manage and develop character arcs?
-\end{itemize}
-
-We also still can't answer fundamental questions about the context of
-students' writing processes:
-
-\begin{itemize}
-\item
-  Do they use particular techniques to ``break story'' and decompose a
-  narrative into its key beats (index cards? Whiteboard?)
-\item
-  Do they participate in a writing group?
-\item
-  How do they respond to people giving them notes on revising a script?
-\end{itemize}
-
-To sum up, because snapshot systems are designed to collect the totality
-of a codebase at frequent intervals, they offer a rich record of data
-that captures the results---both major and minor---of students' design
-decisions. But, computing education research that \emph{uses}
-code-snapshotting has focused much more on detecting, classifying, and
-predicting student errors than it has on showing how students progress
-in programming and design expertise. Nevertheless, snapshot-based
-research shows tremendous promise. Given that :
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Snapshotting students' code represents a cutting-edge way to resolve
-  the way code---as a design artifact---evolves over time, \emph{but}
-\item
-  Code snapshots, as they're currently used, explore neither the
-  totality of a student's design nor the rich context of that design's
-  production, \emph{and}
-\item
-  There is a lack of parity between studies of how professionals design
-  software and studies of how students do so,
-\end{enumerate}
-
-It seems sensible to ask: \emph{can code snapshots be used---possibly in
-synthesis with ethnographically-oriented methods---to start studying how
-students' design thinking plays a role in their introductory programming
-work?} We believe the answer is yes.
-
-In what follows, we present work that proceeds from an empirical
-challenge: how can we develop accounts of students' programming activity
-that explain the form and evolution of their code on a design project?
-We offer an account of one student---Rebecca---and her experiences and
-code from a second-semester course on programming concepts for
-engineers. Using data from both code snapshots and clinical interviews,
-we explicate both the challenges of studying students' software design
-processes and the potential for such study to inform accounts of
-teaching and learning.
-
-\subsection{Context and Methods}\label{context-and-methods}
-
-\subsubsection{Context of the study}\label{context-of-the-study}
-
-The data presented in this case are taken from an ongoing IRB-approved
-study of undergraduate electrical engineering majors undertaken at
-\emph{Flagship State}, a large, public research institution on the east
-coast. For two semesters, I have followed a total of 10 students taking
-``Intermediate Programming Concepts for Engineers.'' It is the second of
-a required two-semester course sequence in programming.\footnote{Hereafter,
-  I use ``Intermediate Programming'' to refer to the second course and
-  ``Basic Programming'' to refer to the first.} Students can (and some
-do) place out of Basic Programming via AP Computer Science credit, but
-all Electrical and Computer Engineering students must take Intermediate
-Programming.
-
-Intermediate Programming has two 75-minute lectures per week and a
-weekly discussion section led by an undergraduate Teaching Assistant
-(TA). Typical enrollment is between 60 and 80 students per semester.
-Like Basic Programming, Intermediate Programming is taught using the C
-programming language, and it incorporates multi-week projects in C as
-part of its assessment structure. Students must work individually on
-four projects over the course of the semester, which together comprise
-45\% of their final course grade.
-
-Grading projects involves running students' compiled code against
-automated tests that determine whether a student program's output
-matches the instructor's canonical output. If a student's program
-completely matches the canonical output, the student receives at least a
-90\% grade on a project. The remaining 10\% are discretionarily
-allocated ``style'' points, awarded for things like proper formatting,
-code commenting, and functional decomposition (Field Notes).
-
-This study centers on ``Flights Database,'' the second of four projects
-assigned to Intermediate Programming students during the spring 2012
-semester. Students were asked to build a text menu-based program that
-would let users query information about airports and plan non-stop and
-one-stop flights between airports. For this project, the instructor gave
-students three separate text files as source material. The
-\textbf{airports} file contained names of airports and their
-three-letter abbreviation codes; the \textbf{routes} file contained
-3-tuples of two airport codes and the route number of a flight flying
-between them; the \textbf{flights} file contained a list of specific
-flight information (including arrival and departure times) by route
-number. Crucially, in order to be able to respond to user queries
-students would need to build a program that could coordinate information
-across all three files to return an answer.
-
-My analysis details the work of Rebecca, a female first-year electrical
-engineering major. I focus specifically on Rebecca's code for finding
-``one-stop'' flights, which the instructor defined as ``all pairs of
-ﬂights that route the user between the departure and arrival airports
-with exactly 1 stop (i.e., a one-connection ﬂight)'' (Flights Database
-handout, 2012). The one-stop problem is particularly challenging. To
-solve it successfully students' code must accept a user's choice of
-airports and day, then stitch together routes that involve two separate
-flights in a way that passes stringent constraints for acceptable
-layover times.
-
-\subsubsection{Methods}\label{methods}
-
-This study proceeds from an empirical challenge: how can we develop
-accounts of students' programming activity that explain the form and
-evolution of their code on a design project? The focused form of that
-challenge for this study is ``how can we understand the unconventional
-design choices embedded in Rebecca's one-stop flight code?'' To answer
-that question, my study draws from three data streams: ethnographic
-observation, clinical interviewing, and code snapshot analysis.
-
-For two semesters, I ethnographically embedded myself in the same
-instructor's section of Intermediate Programming. My aim throughout was
-to see what students see in terms of course material, assignment
-directives, and instruction. In fall 2011 I observed approximately 50\%
-of the course lectures. I also independently completed all class
-homeworks and three out of the four course projects to more fully
-understand the course's assessments. In spring 2012 I continued
-attending lectures, though less frequently, and began attending select
-TA-led discussion sections. During both lectures and discussion sections
-I took field notes while recording ambient audio using a LiveScribe
-Pulse pen.
-
-Rebecca was one of four students (three female, one male) willing and
-able to participate in a series of 1-hour outside-of-class clinical
-interviews during the spring 2012 semester.\footnote{In total, roughly
-  30 students responded to my initial in-class solicitation to be
-  contacted by email about my study. Of the students I emailed,
-  approximately 8 students responded to my emails to schedule interview
-  times. Of those 8, only four students were able to successfully find
-  interview time slots that fit our respective schedules.} I interviewed
-Rebecca five times, and in typical interviews I split time between
-asking about her experiences in the course and giving her time in the
-interview to work on her project code. During each interview, I
-simultaneously used:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  A Kodak Zi8 camera for video-recording our interactions
-\item
-  A LiveScribe Pulse pen to capture Rebecca's on-paper penstrokes
-\item
-  A MacBook Pro (early 2011) with screen-recording software to capture
-  everything on-screen while Rebecca programmed
-\end{enumerate}
-
-The final component of data gathering is modeled after Jadud's (2006)
-system for capturing students' code. My colleagues and I developed
-software, built around the open-source version control system called
-Git, that effectively creates an entire copy of a student's code---what
-we call ``snapshots''---every time students invoke the compiler on their
-code. Our software then sends those snapshots to a secure,
-researcher-accessible server in real-time as they're created.
-Consequently, I could plan each interview with Rebecca around
-up-to-the-minute knowledge of her work---in some cases work she had
-completed just hours before the interview---and tailor my interview
-questions to emerging patterns in her code. In total, Rebecca's work
-resulted in 958 compilation snapshots over the course of the semester.
-
-In what follows, I recount snapshot and interview data to explore the
-form and history of certain features of Rebecca's one-stop flight code.
-First, I simulate an analysis of Rebecca's code in the methodological
-vein of Jadud (2006) and others (Rodrigo, Tabanao, et al., 2009; Tabanao
-et al., 2011), analyzing only Rebecca's code snapshots. In this first
-analysis, I focus on the form and evolution of particular design
-features of Rebecca's code. Then, I present a second, complementary
-analysis using contextual data from my clinical interviews with Rebecca.
-In my second analysis I go beyond the snapshots to highlight why, in
-Rebecca's own words, she made those particular design choices.
-
-\subsection{Code Snapshot Data and
-Analysis}\label{code-snapshot-data-and-analysis}
-
-In this section, I simulate what data on Rebecca would look like using
-an existing method of analysis described in Jadud (2006). Namely, I work
-at a level of remove from Rebecca herself. I base my inferences only on
-her actual code, limiting my interpretations about what she may have
-\emph{intended} to do or \emph{designed} her code to do only to what's
-supported in the snapshot-by-snapshot record of her code. Through this
-``snapshot-only'' analysis, I show that we can see in great detail how
-Rebecca's code embodies particular design features that diverge from how
-an expert might have handled this design task. I restrict my analysis
-here to just two such features of Rebecca's code:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Rebecca's use of repeated file-scanning loops to create a depth-first
-  search algorithm
-\item
-  Rebecca's 7-fold duplication of code to handle checking for flights on
-  each day of the week
-\end{enumerate}
-
-\subsubsection{Rebecca's file-scanning solution is computationally
-complex}\label{rebeccas-file-scanning-solution-is-computationally-complex}
-
-After declaring variables and opening the three provided text files
-(flights, routes, and airports), Rebecca's one-stop flight code enters a
-series of conditionally-nested `fscanf()` commands. The purpose of
-`fscanf()` is to scan through the characters of a file, most often one
-line at a time. The programmer can specify patterns of text to look for,
-e.g., ``in each line, look for a word, followed by a space, followed by
-two digits.'' `fscanf()` also gives the programmer the flexibility to
-store such matched patterns. ``Once you find a word followed by two
-digits, store the word in the following location in memory.'' What's
-interesting isn't \emph{that} Rebecca used `fscanf()` to read through
-files like the list of airports. Any suitable solution would need to
-read through those files. Rather, what's interesting is \emph{how} she
-uses `fscanf()` in her design.
-
-Rebecca's file-scanning logic never persistently stores the contents of
-the files it reads in. Rather, her program reads through files one line
-at a time, and it essentially can't process or act on airport/flight
-information not in the line currently being scanned. Instead, it's been
-designed to copy single patterns temporarily, then rewind the file back
-to the top and s\protect\hypertarget{GoBack}{}{}tart reading in
-one-line-at-a-time again.
-
-What's consequential about Rebecca's design choice? Computationally, her
-code has to repeatedly open multiple files and scan them one line at a
-time in order to coordinate information. So, the task of finding a
-one-stop flight between two cities becomes a series of repeated,
-one-line-at-a-time scans of external files:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Scan the \textbf{airports} file one line at a time to check the
-  correctness of the user's input (line 20)
-\item
-  Scan the file of pairwise airport \textbf{routes} one line at a time
-  (line 24)
-\item
-  To find possible connection cities, scan the \textbf{routes} file one
-  line at a time again (line 38)
-\item
-  If a route matches, scan the \textbf{flights} file one line at a time
-  to verify whether the time/day constraints are acceptable (one of the
-  following lines depending on the chosen day: 52, 79, 106, 134, 162,
-  190, 218).
-\end{enumerate}
-
-Rebecca's code is both visually and computationally complex. The
-multiply-nested blocks can make it difficult for a human reader to
-follow the code flow, which may have made it challenging for Rebecca to
-debug her own work. Moreover, nested for- and while-loops increase the
-complexity space of a program---what computer scientists call the big-O
-characterization of her algorithm. For every nesting of a scan loop
-(there are 4 here) Rebecca increases by 1 the degree of a polynomial
-that represents the execution time of her program. So, from a
-performance perspective, Rebecca's design suffers a trade-off in that
-with each invocation of a scanning loop, we see a geometric increase in
-the time complexity of her program.\footnote{} In fairness to Rebecca,
-the point I make here is not about her but about our mode of analysis.
-From the perspective of a typical software engineer, Rebecca made a
-decision involving a design trade-off. Still, inferring only from the
-code we cannot know whether her decision was deliberate. We cannot know
-from code whether she knew or understood the kind of complexity and
-performance trade-off she made, nor can we know how she felt about the
-consequences of the decision. We know only that her final submitted code
-contained the results of her decision to use nested scanning loops.
-
-\subsubsection{Rebecca's file-scanning solution ignores an assignment
-directive
-}\label{rebeccas-file-scanning-solution-ignores-an-assignment-directive}
-
-A second feature or Rebecca's `fscanf()` design is that she ignores
-instructions in the course assignment, which produces a solution that
-uses abstraction in a non-obvious way. One design solution to the
-problem of creating a relational database---and one seemingly dictated
-by the assignment---is to create three separate arrays in the computer's
-memory:
-
-To parse the 3 airline ﬂight database ﬁles, you will need to declare
-arrays that will receive all the data. For the purposes of determining
-array sizes, you may assume there will never be more than 100 airports
-in the ``airports.txt'' ﬁle, 500 route IDs in the ``routes.txt'' ﬁle,
-and 3000 ﬂights in the ``ﬂights.txt'' ﬁle. (Flights Database class
-assignment, 2012)
-
-Presumably, from the instructor's directive, one ``will need'' to have
-an array of airports (mapping 3-letter code to full airport name), an
-array of routes (mapping a pair of airports to a unique routing number),
-and an array of flights (mapping unique flight numbers to a collection
-of information about that flight).
-
-Rebecca creates no arrays. Instead, her code attempts to accomplish the
-same task that a memory-persistent data structure would, only without
-the persistence. A consequence of Rebecca's approach is that she has no
-easy way to refer to arbitrary airports, routes, or flights in her code,
-since her program has no mechanism to store such information
-persistently. A second consequence is that since she avoids persistent
-data structures, the complex work her program does to read through each
-line of each file, in some cases multiple times (above) is repeated
-every single time a user initiates a query.
-
-Given Rebecca's particular design pattern, we asked the question of
-whether she may have tried creating arrays before ultimately settling on
-her scanning-loop solution. The answer, as far as we can tell, is no. We
-analyzed the history of both Rebecca's main() method and her one-stop
-flight code module. Our search revealed that no snapshots exist in which
-Rebecca created arrays---either through dynamically allocating them
-(through heap memory), or, as the assignment recommended, creating
-overprovisioned fixed-size arrays on the stack. In other words, at the
-limit of resolution of our data collection, and within the scope of the
-code Rebecca typed, she never tried an array solution.\footnote{If
-  Rebecca had tried an array solution and compiled---whether error-free
-  or not---our automated snapshot collection system would have captured
-  it.}
-
-Curiously, we have evidence \emph{outside} of Rebecca's code that
-suggests she knew, and even perhaps had seen, an array-based design
-solution. In a file called ``notes.txt'' contained in her project
-directory, we see the following lines:
-
-think about using: sscanf, array of pointers
-
-his header file!!!
-
--max line lenght: 2000
-
--max string lenght: 100
-
--defined true and false
-
--max airports: 100
-
--max routes: 500
-
--max flights: 3000
-
--min connect time: 60.0
-
--max connect time: 120
-
--daily maxk: 254 ???
-
--char airports{[}max airports{]}{[}4{]}
-
--char aiport\_cities{[}max airports{]}{[}max string lenght{]}
-
--he has 3D array for routes.... char routes{[}max
-routes{]}{[}2{]}{[}4{]}
-
-(notes.txt file, created March 19, 2012)
-
-The context of the file is not entirely apparent, because we did not
-observe lecture on March 19, the day the notes.txt file entered
-Rebecca's snapshot history. Also, whether ``his'' refers to the
-instructor or perhaps another classmate is unclear. ---What seems clear,
-however, is that Rebecca was responding to items she had seen in someone
-else's header file. Consequently, putting together the notes.txt file
-with Rebecca's final code submission reveals Rebecca was exposed to a
-design solution involving arrays, but never implemented it in her code.
-Thus, a lingering and consequential question remains unanswered: why did
-Rebecca adopt a solution that defied the directions of the assignment,
-especially when she'd seen part of a potential design solution that did
-use arrays?
-
-We return to this question in a later section, but first we turn our
-attention to another unusual feature of Rebecca's work: seven-fold
-repetition of code.
-
-\subsubsection{Rebecca repeats the same chunk of code seven
-times}\label{rebecca-repeats-the-same-chunk-of-code-seven-times}
-
-A second key feature of Rebecca's code is the almost identical
-repetition of a single 23-line code chunk seven times (lines 50--240).
-Because users can run queries by choosing a day to fly (and some flights
-only run on certain days), students' code must be able to handle each of
-the seven possible days for when a user would want to fly. In principle,
-Rebecca's code achieves just that.\footnote{I say ``in principle''
-  because Rebecca's code would not compile on my machine. So, in
-  practice, her design contains compile-time errors (and possibly
-  run-time errors). Nevertheless, her code provides ample evidence that
-  she was attempting conditional logic to handle each possible day.} In
-practice, her code creates seven different conditional branches---one
-branch for each day of the week---where the code within each branch is
-duplicated.
-
-Figure 2-1 represents a side-by-side delta-comparison of two such
-day-specific branches of code. Lines 158--184 of Rebecca's original code
-are on the left; lines 214--241 are on
-\includegraphics[width=5.99722in,height=3.89722in]{media/image3.png}the
-right, and in the figure lines have been renumbered (from 1) to ease
-comparison. In this delta view, lines that differ are highlighted in
-pale red, and characters that differ are shown in bold red.
-
-The two code blocks demonstrate just how much code is duplicated for
-handling user input based on days of the week. Between these two chunks
-there are only three differences (lines 1, 3, and 11): all of the
-references to day are changed from 5 (on the left) to 7 (on the
-right).\footnote{In the text-input files students were given, days of
-  the week were represented as integers (rather than the perhaps more
-  familiar ``Tuesday,'' ``Wednesday,'' etc.).} Moreover, the changes
-from block to block are patternistic and predictable: the first line of
-the block is a non-functioning comment, the third line of each block
-just checks whether the rest of the block should run, while the eleventh
-line of each block compares an array entry to the day of interest.
-Everything else is duplicate boilerplate that is essentially repeated 7
-times; once for each day of the week. I say ``essentially repeated''
-because, as we'll now explore, there are minute differences between some
-of the code chunk's seven incarnations.
-
-Repeating code as Rebecca has done can be problematic because each
-repetition multiplies the number of places she has to examine and modify
-if she wants to introduce a systematic change. If, for example, Rebecca
-wanted to change the internal names she gives to scanned-in variables,
-she has to make that change in seven different blocks of code: once for
-each of the seven days of the week she's hard-coded. And, since any
-given change may inadvertently introduce an error, increasing the number
-of places she repeats code also makes the code that much more vulnerable
-to inconsistently-applied changes.
-
-Indeed, a repeated, inconsistently-applied scan pattern change seems to
-be exactly what occurred in Rebecca's code history. Between 10:04pm and
-10:37pm on March 26, Rebecca introduced a large set of changes to the
-one-stop flight module. Among those changes Rebecca added the
-`d\_letter` file-scanning-parameter to what would become line 218, but
-not to what would become line 190. We can reasonably infer Rebecca added
-this parameter as a way of capturing the ``am'' or ``pm'' specifier
-given the input file's format. Moreover, we can verify through Git that
-once introduced, Rebecca's omission of the parameter was never modified
-or corrected. The problem percolated through to her final submitted
-code.
-
-\subsection{Interview Data and
-Analysis}\label{interview-data-and-analysis}
-
-In the previous section, I described two unusual features of Rebecca's
-code for searching one-stop flights:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Her use of multiply nested loops that scan through source information
-  files \emph{without} storing the information in those files
-  persistently in long-term memory
-
-  The code for handling a user's chosen day, which was essentially the
-  same block of code copied and pasted 7 times
-\end{enumerate}
-
-In this section, I offer explanations of Rebecca's design choices by
-interpreting data from over five hours of clinical interviews I
-conducted with her. I draw from those interviews to explain how design
-decisions that might seem unusual to an expert in fact grew rather
-unproblematically (for Rebecca) as ways of deliberately transferring
-prior knowledge and designs (which explains feature 1) or coping with a
-constraint to produce a reliable solution she could trust (which
-explains feature 2).
-
-\subsubsection{Rebecca employed fscanf loops because she was
-deliberately reusing from an Basic Programming
-Assignment.}\label{rebecca-employed-fscanf-loops-because-she-was-deliberately-reusing-from-an-basic-programming-assignment.}
-
-Rebecca's choices become easier to understand when you consider what she
-said in the interview. When Rebecca initially saw the flights database
-project, her first reaction was ``this is just a lot like our fantasy
-football project that we did last year'' (Interview, March 16,
-2012).\footnote{Transcript conventions are shown in Appendix 1.} But,
-the previous fantasy football project was easier because all of the
-information she needed was in one file. Flights database, by comparison,
-fractured necessary information across multiple input files.
-
-Since Rebecca said the fantasy football project was like an easier
-version of flights database, I asked whether she tried making this
-project more like the easier one. Her response was an emphatic ``Oh
-yeah! Definitely!'' (Interview, March 16, 2012). She said ``as soon as
-we got this project I was like, ah! fantasy football! I'm just gonna go
-and see how much code I can rework from that and like, use /mmhmm/ in
-this project'' (Interview, March 16, 2012). Specifically, Rebecca went
-back to her fantasy football code ``and I was looking at how I scanned
-in the information from the files, cuz, we haven't really done anything
-like that this semester. Uh, scanning in from files before---''
-(Interview, March 16, 2012). Since the fantasy football draft project
-was the last in which she'd needed to do file-scanning, she ``just, uh,
-went back to check how I did that /mmhmm/ and then, if I could I copied,
-but because a lot of the variables were different, uh, like these were
-more var---less, less variables, and more strings than last year /mmhmm/
-uh, I just retyped it out. I just looked at how it was similar''
-(Interview, March 16, 2012).
-
-My first opportunity to discuss Rebecca's one-stop flight work was on
-March 16, 2012, in what would be her third of five interviews that
-semester. This interview was conducted very early into the time window
-for the Flight Database project, before Rebecca had done the bulk of her
-coding. We were discussing her prospective design plans. As Rebecca
-began explaining how the logic for a one-stop flight search was supposed
-to work, she described what she saw as one of the central difficulties
-of the project: the relevant information for answering user queries was
-spread across multiple files (Interview, March 16, 2012).
-
-As Rebecca explained, something as simple as finding a flight from, say,
-JFK to BWI ``involves scanning through multiple files, because it's not
-like one file that has everything conveniently like, there'' (Interview,
-March 16, 2012). When I asked what would make things easier if,
-hypothetically, all the information she needed were in one file, Rebecca
-responded by appealing to a previous assignment from last semester. In
-Rebecca's first-semester programming course (Basic Programming for
-Engineers), one of several multi-week projects had students create a
-system for users to conduct a fantasy football draft. The project
-involved, among other things, topics related to basic file management in
-the C language, including how to read in a file from disk (Interview,
-March 16, 2012). Rebecca explained:
-
-Rebecca: Um, like, cuz when we first got this project, uh, I actually
-was thinking ``oh, well this is just a lot like our fantasy football
-project we did last year.'' /Huh/ We uh, had to scan in, uh, someone had
-to enter like ``I wanna pick a quarterback,'' so then you had to scan in
-and go and look for all the quarterbacks in the file and say ``OK, this
-is the quarterback'' and everything. But, in that project we only had
-the one file that had everyone listed: quarterbacks, runningbacks, wide
-receiver, in one file. And all the information you needed there /Mmhmm/
-So, you could just get it all and compare it all at once with one scanf
-/mmm/ whereas this you have to, take, uh, you scan in the flights, ah,
-the flights file. So, then you find the flight number. You have to save
-the ID from that flight number, use that ID to scan into the routes file
-/mmhmm/ and then save the routes information and then print it out with
-the flights information. (Interview, March 16, 2012)
-
-Rebecca's comments suggest she saw a coupling between the arrangement of
-the input information and the structure (and complexity) of the
-computational logic needed to process it. When one fantasy football file
-contained all of the relevant information (player, position, team, etc.)
-it could be read in and processed one line at a time. When information
-was fractured across files (pairs of airport codes in one file, full
-spell-outs of airport names in another, for example), Rebecca felt she'd
-need to use information shared across files (such as a route ID) to
-coordinate a scan across one file with a scan across another (and
-possibly an additional scan across a third file) before she would have
-all the necessary information and computations to return a result.
-
-I was interested in the connections Rebecca saw across projects, so I
-pressed on. When I asked whether she thought about trying to make this
-project like her fantasy football project, her answer was an emphatic
-``Oh yeah, definitely!'' As she elaborated:
-
-Rebecca: That was like, as soon as we got this project I was like, ah!
-fantasy football! I'm just gonna go and see how much code I can rework
-from that and like, use /mmhmm/ in this project. And, my whole main
-file, like all those NULL checks and everything, I mean they're really
-simple to write, but I just copied `em and put `em there, cuz, we had
-the same thing. /Mmhmm/ Um, just changed, like, the names of the files.
-
-As she explained, ``reading in files'' was a topic covered extensively
-in Basic Programming---the first course of the sequence---but they
-hadn't talked much about it in this semester's course.
-
-Rebecca: Because {[}reading in from files{]} was a big {[}Basic
-Programming{]} topic we hadn't talked about it much. /Yup/ So, I just,
-uh, went back to check how I did that /mmhmm/ and then, if I could I
-copied, but because a lot of the variables were different, uh, like
-these were more var---less, less variables, and more strings than last
-year /mmhmm/ uh, I just retyped it out. I just looked at how it was
-similar. (Interview, March 16, 2012)
-
-I started doing the scanning, and then I realized like, ``I do not have
-all the information I need in this one file,'' so that's when I realized
-I would have to make the dummy variable to compare to the other file,
-so, which, there was no other file in the, um, fantasy football one. So
-that's when I realized, I was like OK, so, there's gonna be other stuff
-that I'm gonna have to do, and /mmhmm/ I mean, I can still kinda compare
-cuz it's still scanning files /yup/, but the idea of the dummy variable
-and all that, that would not have come from fantasy football, so /sure/.
-
-In sum, then, Rebecca's repeated, nested scan loops were a structure she
-deliberately borrowed from a previous semester's project. By her
-telling, what seemed obvious was that ``scanning in from files'' was a
-topic she'd already covered, which meant she'd already developed a
-workable solution for how to solve that problem. Thus, she saw the
-problem of how to coordinate airport information from different files as
-a new instance of the old problem of reading information in from one
-file. Her flight database work, accordingly, tried opportunistically
-adapting a previously working solution to fit the current circumstances.
-
-\subsubsection{Rebecca repeated code because she wanted to re-use
-functionality she could
-trust}\label{rebecca-repeated-code-because-she-wanted-to-re-use-functionality-she-could-trust}
-
-By our interview on April 6, Rebecca had already completed and submitted
-her code for the flights database project. When I looked at the final
-form of her code for finding one-stop flights I noted an unusual pattern
-described in section 2.4.3 above: she had a code chunk repeated almost
-character-for-character 7 times. In the interview, this section is what
-Rebecca referred to as ``my obnoxiously long part of my code''
-(Interview, April 6, 2012):
-
-Rebecca: So, the way I did it was really long and probably, there was
-probably like a much easier way, but I just did a giant if---if
-statements \{swings cursor from line 50 to line 63\} If they wanted to
-fly on Monday /OK/ I went through and checked to see if the route ID was
-the same \{wiggles cursor across line 54\}, and if it did, I went
-through to che---uh, I made a check\_days function \{wiggles cursor
-across line 60\} uh, I ended up commenting that out cuz I didn't end up
-/mmhmm/ finishing it. But, uh, my check\_days function worked, it just
-didn't work completely with the code /OK/ (Interview, April 6, 2012)
-
-As I scrolled the screen to look at each of the repeated blocks of code,
-Rebecca elaborated:
-
-Rebecca: And this is why my code, I feel like, is not uh, concise
-enough, or, I don't really, I forget the word they use, but uh
-/\{inaudible\}/ it's very long because I couldn't figure out if I should
-do a while loop or whatever /Uh-huh/ But, so I was just like, I know
-this way should work if I get everything else right, that uh, just go
-through, if input's 1, if input's 2 and just do the same thing in each
-of `em just /Mmmhmm/ check for, ``oh, if days is 2, if days is 1''
-instead of, like---Cuz I probably could have done, like, maybe a giant
-while loop, um, to try and, and if, while, inputs something, uh, then
-you check to see whatever i is. But, I could, I didn't---couldn't figure
-out how that would work, so I just did the same thing six times.
-
-Interviewer: So, in, in each one of these it's like, looks, and I'm not
-sure about this, but it looks like the way you wrote it---so this
-\{highlights line 79\} is pretty much the same in all of them, right?
-/Yes/
-
-Interviewer: So's this one \{highlights line 81\} /Yes/ this one
-\{highlights line 83\} Here's where it's different \{highlights line
-85\}
-
-Rebecca: Yes, because it just checks if it's a 2 instead of a 1.
-
-Interviewer: OK. Um. /And then everything else is still the same/
-Layover's still the same. OK.
-
-Rebecca: Yeah. So that's why it's prob---it's not, uh, the neatest code
-or whatever, because it's the same thing six times. (Interview, April 6,
-2012)
-
-Given Rebecca's assertions that her code wasn't neat, I asked what, if
-anything she might change if she hypothetically had another week to work
-on the project.
-
-Rebecca: Um, first I'd try and get it to make sure it worked completely
-/Ahh, OK, yeah/ this way, \{laughs\}, uh, and then, if I had the week
-after that whatever, I'd probably go through and see if I could figure
-out a way to make it concise-r because he likes uh, neat, as, like, code
-that's, uh, easy for the user to see /uh-huh/ I guess. Uh, I forget
-what, I keep forgetting what the word he used was at the beginning of
-the year, but uh, just very concise and, uh, this is {[}a{]} very
-expanded \{laughs\} way of coding, but, it made sense to me at the time
-and I was just like ``I just want something that makes sense right
-now.'' /Right/ So, that I can actually work with and have an idea.
-
-Interviewer: Um, OK. So, so it would take you some extra thinking to
-figure out /Mmmhmm/ how to break this down into /Yes/ smaller stuff
-/smaller code/ Do you feel like you've had a lot of practice doing that,
-or like?
-
-Rebecca: Uh, a little. Like, but, a lot of times in {[}Basic
-Programming{]} they didn't really mention too much about being concise.
-They were just like ``if you can do it, do it'' \{laughs\} /OK/ So I
-usually stuck to what made sense to me /Right/ uh, to turn the projects
-in. (Interview, April 6, 2012)
-
-In summary, Rebecca's ``expanded way of coding'' was a way of expressing
-ideas in code that, in her own words, ``made sense'' to her. Moreover,
-her Basic Programming course seemed, to her, to set expectations that
-functionality comes first; ``neatness'' second. If she hypothetically
-had more time to work on the project, her first priority would be to get
-her existing code working. Consequently, Rebecca's repetition of code
-can be understood as a kind of pragmatic solution to a difficult
-problem: choosing which computational techniques were best for
-accomplishing a complex goal. Moreover, her approach was shaped by the
-fact that her Basic Programming course historically valued a philosophy
-of ``if you can do it, do it'' (Interview, April 6, 2012). Ultimately,
-those factors seem to be what led Rebecca to choose repeating code that
-made sense to her over the difficult-to-envision alternative of a
-``giant while loop.''
-
-\subsection{Discussion}\label{discussion}
-
-When we consider evidence from both code snapshot histories and clinical
-interviews with Rebecca, we can draw several plausible conclusions about
-Rebecca's patterns of software development on this project. Moreover,
-those conclusions can spur larger considerations about theorizing
-student learning and considering alternative instructional strategies
-that may be more responsive to student needs. Below, I synthesize and
-discuss patterns in Rebecca's development.
-
-\subsubsection{Rebecca's key design decisions are made early, persist
-through to her final submission, and carry
-consequences}\label{rebeccas-key-design-decisions-are-made-early-persist-through-to-her-final-submission-and-carry-consequences}
-
-On both her non-stop flight code and her one-stop flight code, key
-structural features of how Rebecca's code works never meaningfully
-change after Rebecca introduces them. In particular, Rebecca's use of
-fscanf() patterns (as opposed to storing the data persistently in
-memory) occurs extremely early in the work history of the project. The
-fscanf() patterns initially appear in Rebecca's flight-checking code on
-March 19, the third of what would be 286 chronological snapshots we have
-comprising Rebecca's work on the project.\footnote{The project was
-  distributed to students on March 5 and was due March 28, but the bulk
-  of Rebecca's work on the project (284 of the 286 snapshots) occurs
-  beginning March 19.
-  https://github.com/TLPLEngineeringEdResearch/Rebecca/compare/e0a7caab4eb2d8401a0e50e8c6c6d8b514e9054e...0e8a5c606ef30f3b404024fdfad17d8d50e32456}
-We note several consequences that arise \emph{because} these design
-choices were made early and persisted to the final code.
-
-First, Rebecca's \emph{early} choice not to persistently store data
-likely forced her into the complicated scanning loop logic she developed
-\emph{later} to solve the one-stop flight problem. A bit of explanation
-may help here. Because the database students were building has multiple
-uses, including uses that find flights for users, to be successful
-Rebecca had to write functionality that found both non-stop and one-stop
-flights for users. But, Rebecca initially designed each part of her
-code---airport listings, flight listing, non-stop flight search, and
-one-stop flight search---with its own internal fscanf() loops, as
-opposed to having the four of them each refer out to a separate module
-that could handle reading in data. The result is that none of her
-individual modules share file-scanning code, and repeat data look-ups
-necessarily have to be handled by creating another nested scan loop. A
-design where code with a common function is pasted repeatedly into
-separate modules contrasts strongly with one in which separate modules
-all refer out to a single external piece of code.\footnote{As an
-  analogy, imagine a supermarket where every unique type of item on the
-  shelf has a cash register specific to that item next to it. To buy an
-  item shoppers have to check-out individual items as they move through
-  the store. It's a pain for shoppers because shoppers have to swipe
-  their cards at dozens of separate item-specific registers across the
-  store just to pay for a single grocery run. It's a pain for
-  store-owners because now there are literally hundreds of individual
-  registers to manage, maintain, and upgrade.}
-
-We can understand Rebecca's development as involving a kind of design
-inertia. Because she decided from the start to scan information using
-fscanf() separately in each module, her later code was also beholden to
-that commitment. The end result is that increasing the problem
-complexity a small amount (the challenge of handling one-stop flights
-versus non-stop flights) forces her to write much more code that is
-itself more complex in its flow of control. And, when we consider her
-seven-fold repeated code for checking flight availability on given days,
-there are similar consequences. The decision---whether intentional or
-not---to create seven different conditional paths forces Rebecca to
-write and maintain much more code over time as the project evolves. As
-we suggested in section 2.4.3, having a large number of locations to
-keep in sync is likely what made Rebecca vulnerable to one of the errors
-she ultimately introduced into her code.
-
-\subsubsection{Her design decisions may be influenced by framing and
-overzealous
-transfer}\label{her-design-decisions-may-be-influenced-by-framing-and-overzealous-transfer}
-
-Rebecca said in our interviews that she explicitly tried to reuse
-solutions from the previous semester's fantasy football project. Our
-code snapshots help us understand the extent to which that's true.
-Rebecca's first compile for her flights database work begins with this
-commented line of code:
-
-//Rebecca Wells's Fantasy Football Team Maker: Project 1
-
-Thus, our interview with Rebecca corroborates the snapshot evidence that
-Rebecca directly copied code from a prior project. Moreover, Rebecca's
-behavior of copying her old fantasy football code suggests several
-implications.
-
-First, to extend our point in section 2.6.1, Rebecca's ``design
-inertia'' actually traces as far back as the decisions she made during
-the first project of her first semester of programming. That is, by
-copying code (and particularly file-scanning logic) from an old project,
-she was incorporating core functionality that she designed when she
-first learned to program. Admittedly, code reuse isn't in and of itself
-a problem in software development. Parson and Saunders (2004), for
-example, have written on cognitive heuristics that keep professional
-software engineers from reusing code, even when reusing and extending
-existing software artifacts is the best course of action on a software
-project. So, the concern isn't that Rebecca reused code, but rather the
-matter of what cognitive dynamics were at play that directed her choice
-to reuse that code. In that regard, the constructs of framing and
-transfer may help us better understand Rebecca's activity.
-
-Framing, as it has been applied in contexts such as physics education
-(Elby \& Hammer, 2010; Hammer et al., 2005; Scherr \& Hammer, 2009) and
-mathematics education (van de Sande \& Greeno, 2012) concerns how
-participants understand the social and intellectual activities in which
-they're engaged. Of particular relevance to studying Rebecca is the
-notion of \emph{epistemological framing}, which van de Sande and Greeno
-(2012) summarize as
-
-participants' understanding of kinds of knowledge that are relevant for
-use in their activity and the kinds of knowledge, understanding, and
-information they need to construct to succeed in their activity (e.g.,
-what kind of information would count as a solution to the problem they
-are working on). (van de Sande \& Greeno, 2012, p. 2)
-
-Rebecca's decision to copy code from fantasy football implicitly
-reflects her orientation toward what kinds of knowledge (the course
-topic of scanning information from files) are relevant to solving the
-problem. More broadly, note that in Rebecca's interview she explains
-that her primary objective is to get a solution that works, which stands
-in contrast to having a design priority like having a solution that is
-elegant, or one that transparently manages complexity. That commitment
-again reflects a manifestation of epistemological framing as ``what
-would count as a solution,'' where for Rebecca what counts is a solution
-that works.
-
-Additionally, Rebecca saw the flight database project as a new instance
-of a prior problem: fantasy football. Consequently, she consciously
-adapted past solution patterns, because the flights database project
-looked like it contained problems she had already solved in previous
-code. Rebecca's deliberate reuse of old code can be readily understood
-as an example of what Schwartz, Chase, and Bransford (2012) call
-``overzealous transfer'':
-
-Of particular concern are situations where students transfer skills,
-knowledge, and routines that are effective for the task at hand but may
-nevertheless be sub-optimal in the long run because they block
-additional learning. We will call this \emph{overzealous transfer}
-(OZT)---people transfer solutions that appear to be positive because
-they are working well enough, but they are nevertheless negative with
-respect to learning what is new. (Schwartz et al., 2012, p. 206)
-
-In short, when students overzealously transfer prior knowledge as
-Rebecca did, ``they may believe they are doing the right thing, and
-without appropriate feedback they cannot know otherwise'' (Schwartz et
-al., 2012, p. 206).
-
-In Rebecca's case, the constructs of framing and overzealous transfer
-together let us describe why she would have repurposed a solution that
-she felt was adequate, even to the exclusion of the topics being taught
-in class and the explicit directions in the project brief. By framing
-the flights database problem as a new instance of an old problem,
-Rebecca treated it as she did the old problem. But, she arguably
-transferred \emph{too} \emph{much} of the old code's structure; so much
-that she had to introduce even more complexity into her code just to
-make the transferred parts work properly under the new constraints. And,
-because her framing of the task seemed to privilege a philosophy of ``if
-you can do it, do it'' (Interview, April 6, 2012), her primary goals
-were to get her program to work, by whatever means she could understand
-and trust.
-
-\subsection{Conclusion }\label{conclusion}
-
-In closing, even with a complete snapshot history and over five hours of
-clinical interviews, it's still an enormous challenge to thoroughly
-understand one student's design trajectory on a project. The challenge
-is both methodological and theoretical. Through interviews we know what
-Rebecca was thinking retrospectively, but we still work at a remove from
-understanding what she was thinking in the precise moments she wrote her
-code. Conversely, our snapshots provide a detailed record of the state
-of her code every time she compiled. But, our snapshots are limited by
-the kinds of materials they can track and the frequency with which
-they're taken. We don't know, for instance, what Rebecca might have been
-talking to others about while sitting in a dorm lounge working on her
-project. And, because Rebecca wouldn't always compile frequently, some
-snapshots detail large changes to the code made over long periods of
-time---hours or days---which can limit our ability to precisely
-determine what happened, and when.
-
-From a theoretical perspective, we find our attempts to understand her
-design decisions involve appeals to constructs of a finer granularity
-and dynamism than are typically considered when studying first-year
-programming courses. Understanding why fscanf() loops dominate Rebecca's
-flight search code, for example, pushes us to consider the effects of
-how Rebecca framed the task and the subsequent overzealous transfer that
-resulted. And, understanding why code gets repeated seven times pushes
-us to consider how emotion couples with sense-making in students' design
-decisions. In Rebecca's case, a time-pressured preference for code that
-``makes sense right now'' translated to a design of explicitly repeating
-procedures instead of elegantly abstracting them.
-
-\section{\texorpdfstring{\protect\hypertarget{ux5fToc247188552}{}{\protect\hypertarget{ux5fToc252445913}{}{}}Study
-2 -- What directs and sustains students' in-the-moment programming
-activity?}{Study 2 -- What directs and sustains students' in-the-moment programming activity?}}\label{study-2-what-directs-and-sustains-students-in-the-moment-programming-activity}
-
 \subsection{Introduction}\label{introduction-2}
@@
 \toprule
-\textbf{Statement } & Programming Syntax & How I might think about
-it\tabularnewline
+\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
+Statement
+\strut\end{minipage} &
+\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
+Programming Syntax
+\strut\end{minipage} &
+\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
+How I Might Think About It
+\strut\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{1} & x = x + d & An incrementer or accumulator\tabularnewline
-\textbf{2} & x = x + update & Updating the value of x\tabularnewline
-\textbf{3} & x\_position = x\_position + update & Updating the
-x-position with uniform velocity\tabularnewline
-\textbf{4} & x\_position = x\_position + update(t) & Acceleration on the
-x-position\tabularnewline
-\textbf{5} & x\_position = x\_position + update(\ldots{}) & Generalized
-x-position update (update could be any function of any number of
-parameters: random noise, low-level hardware functions getting mouse or
-keyboard input, etc.).\tabularnewline
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+1
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+\texttt{x\ =\ x\ +d}
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+An incremented or accumulator
+\strut\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+2
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+\texttt{x\ =\ x\ +\ update}
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+Updating the value of \texttt{x}
+\strut\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+3
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+\texttt{x\_position\ =\ x\_position\ +\ update}
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+Updating the x-position with uniform velocity
+\strut\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+4
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+\texttt{x\_position\ =\ x\_position\ +\ update(t)}
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+Acceleration in the x-direction
+\strut\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+5
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+\texttt{x\_position\ =\ x\_position\ +\ update(...)\_}
+\strut\end{minipage} &
+\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
+Generalized x-position update, where the update function might be
+anything from stochastic noise to low-level hardware functions that get
+keyboard or mouse input
+\strut\end{minipage}\tabularnewline
 \bottomrule

### 2016-04-22 dissertation-export.tex
Copy parts of exported dissertation into manuscript body
But, I forgot to apply `--no-wrap` to the pandoc export, so now I have
to go fix that.

@@
 
-\begin{longtable}[]{@{}lll@{}}
-\toprule
-\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
-Statement
-\strut\end{minipage} &
-\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
-Programming Syntax
-\strut\end{minipage} &
-\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
-How I Might Think About It
-\strut\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-1
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-\texttt{x\ =\ x\ +d}
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-An incremented or accumulator
-\strut\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-2
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-\texttt{x\ =\ x\ +\ update}
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-Updating the value of \texttt{x}
-\strut\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-3
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-\texttt{x\_position\ =\ x\_position\ +\ update}
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-Updating the x-position with uniform velocity
-\strut\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-4
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-\texttt{x\_position\ =\ x\_position\ +\ update(t)}
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-Acceleration in the x-direction
-\strut\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-5
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-\texttt{x\_position\ =\ x\_position\ +\ update(...)\_}
-\strut\end{minipage} &
-\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
-Generalized x-position update, where the update function might be
-anything from stochastic noise to low-level hardware functions that get
-keyboard or mouse input
-\strut\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-I stress that the third column is about \emph{how I might think about}
-each programming statement\emph{.} By no means am I making normative
-claims about how one \emph{ought} to think about it, or whether the
-statement actually does what its names might suggest it does. Rather,
-``How I might think about it'' reflects the kind of local meaning or
-interpretation I might attach to such a statement when I work with it,
-given my understanding of its role and context.
-
-Consider specifically statements 1 and 4. Statement 1 could very well be
-an incrementer in some kind of iterative code. If I had to debug code
-that employs statement 1, what I might do is exploit my knowledge of
-what d is (does it hold a constant value? Is it 1?) and try inserting
-intermediate print statements into the iterative code. Doing so, I can
-inspect textually how values change with each iteration. In statement
-4's case, it could very well be that the position-update code animates
-images on a screen. If there's a problem with that code, one of the
-easiest ways I might notice is that the acceleration seems off in the
-graphics. Consequently, my debugging might call upon knowledge I have
-from kinematics. I might try inserting code that draws a dot at the
-object's on-screen position with each iteration, leaving a trail I can
-visually inspect. I could then look at the path of the object's
-trajectory and the spacing patterns between successive dots as a
-first-pass test of whether my code achieves the motion I want.
-
-To be clear, it's not just that graphics might improve my efficiency in
-debugging a statement. Rather, applying an interpretive frame that
-treats an assignment statement as a kinematic position update lets me
-\emph{use conceptual knowledge from physics} to diagnose and fix
-problems in my code. If I view the assignment statement as saying
-something about the motion of an object (cf., Hammer, 1994, p. 165), a
-field of knowledge and concomitant techniques from physics becomes
-available to me to think with. But, if I focus only on the
-semantic-level equivalence of statements 1 through 5, there would be no
-obvious reason for me to access what I know about physics in order to
-reason about the code.
-
-My second example concerns statements that are syntactically-similar,
-rather than semantically-similar. I proposed that the statements in
-Table 2 all came from the same language. But, another phenomenon comes
-into play when different languages use the same symbology for
-conceptually different operations. Table 3 shows examples of what look
-like semantically-equivalent operations, but in fact are not.
-
-\protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
-syntactically-similar statements with very different semantics
-
-\begin{longtable}[]{@{}llll@{}}
-\toprule
-\textbf{Statement } & Programming Syntax & Language & How I think about
-it\tabularnewline
-\midrule
-\endhead
-\textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
-\textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
-the string w\tabularnewline
-\textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
-layer of points onto a plot p\tabularnewline
-\bottomrule
-\end{longtable}
-
-The catch here is that the plus operator takes on different roles in
-different languages because of how those languages define its use.
-Statement 1 increments a number in C; Statement 2 appends the letters
-``ly'' to a string; Statement 3 adds a layer of points to a statistical
-graphics plot. These different kinds of operations become even more
-apparent and consequential when, for example, such statements are
-repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
-  operator can be written as a directional arrow. The symbol
-  \textless{}- (``less-than, hyphen'') indicates the value on the right
-  side of the symbol is being assigned to the variable on the left side
-  of the symbol. As used, the symbol itself is semantically equivalent
-  to having written an equals sign (``=``).} I'm using multiple
-reassignment statements to compose a statistical graphics plot. The
-layered creation of a plot invites a very different kind of conceptual
-interpretation than, say, repeatedly accumulating numbers into a running
-sum:
-
-p \textless{}- InitializeGgplot\_1w()
-
-p \textless{}- p + GrandMeanLine(owp)
-
-p \textless{}- p + GrandMeanPoint(owp)
-
-p \textless{}- p + ScaleX\_1w(owp)
-
-p \textless{}- p + ScaleY\_1w(owp)
-
-p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)
-
-One obvious reason for thinking about this code with a different
-interpretive frame is that numeric addition is commutative; composing a
-plot is not necessarily commutative.\footnote{To convince yourself that
-  plot composition isn't commutative, imagine a scale function that
-  squares up the aspect ratio of a plot and another scale function that
-  sets the aspect ratio to 1.5:1. Applying the square function last
-  produces a square plot; applying it first produces a rectangular plot.
-  String concatenation is also not necessarily commutative. ``cat'' +
-  ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
-  ``dogcat.''} So, despite the syntactic similarity, reassignments that
-compose a plot using reassignment (as above) does not obey the same
-rules as reassignments for a running total. But, even within this code
-block, statements that look alike perform operations of a different
-nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
-visual layer to a plot, others modify features of the plot (e.g, ``+
-Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
-the numeric range of the data).
-
-Given these motivating examples, it seems sensible to think there's
-utility in a programmer having different conceptual metaphors available
-to think about and work with code. Example 1 shows that choosing to
-apply knowledge from physics to a piece of code can change the cognitive
-nature of debugging. There, debugging a position-update statement
-becomes, in part, reasoning kinematically about the properties of motion
-trails.\footnote{For comparison, consider the argument that ringing an
-  aircraft speedometer with physical markers changes the nature of
-  cognition when pilots work to land a plane (Hutchins, 1995b).} Example
-2 shows that across languages, programmers might have to deploy
-different conceptual metaphors to reason about statements in a
-locally-consistent way. Knowing that plots in ggplot2 can be composed
-layer-by-layer with reassignment is crucial if you're trying to write or
-understand code that creates statistical graphics. But, I would argue
-that thinking about ◻ = ◻ + ◻ as ``compose new layer onto plot'' can and
-does appeal to different kinds of knowledge when compared to thinking
-about ◻ = ◻ + ◻ as ``include this addend in the sum,'' which itself can
-and does appeal to different kinds of concepts when compared to thinking
-about ◻ = ◻ + ◻ as ``increment the counter.''
-
-Stepping back, I can build the following argument
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Programming can be helped by applying conceptual models to code,
-  particularly when relevant domain-knowledge structures can
-  advantageously transform a problem (example 1)
-\item
-  But, conceptual models don't work all the time for all statements.
-  Because languages are designed differently, the same syntax can
-  actually correspond to very different operations in code (example 2).
-  And, that's true both within and across languages.
-\item
-  Consequently, it makes less sense to treat conceptual models as right
-  or wrong, and more sense to treat them as differentially advantageous
-  for thinking about what a piece of code does. (Thinking a ``+''
-  implies numerical addition isn't \emph{globally wrong} in JavaScript,
-  but it won't explain why 1 + ``1'' yields ``11'' as a result.)
-\item
-  It seems plausible that successful programmers, when reasoning about
-  or writing code, are able to dynamically access or deploy conceptual
-  models that are advantageous given the context (language, syntax,
-  surrounding code). Certainly such a supposition is in line with work
-  suggesting students have resources for thinking conceptually and
-  dynamically about how mathematics models real-world situations (Izsák,
-  2004; Sherin, 2001)
-\item
-  To model how programmers think with conceptual models, a suitable
-  framework should be able to account for the dynamic, context-sensitive
-  deployment of conceptual knowledge.
-\item
-  To model how programmers develop expertise, a suitable framework
-  should be able to describe higher-order phenomena. Such phenomena
-  include explaining how programmers come to have conceptual models or
-  generate new ones, why they decide to deploy them, and how programmers
-  consider which conceptual model (i.e., which way to think about code)
-  is appropriate.
-\end{enumerate}
-
-Taken together, these assertions propose criteria for how we might
-strive to model cognition in programming. Our modeling frameworks should
-be context-dependent, dynamic, and capable of explaining where
-conceptual models come from. They should also be able to account for
-phenomena that are not themselves conceptual, including what directs the
-use of certain kinds of conceptual knowledge. In the learning sciences,
-such models already exist and have proven useful and productive for
-thinking about thinking.
-
-\subsubsection{Manifold models of cognition explain context-dependence
-and the growth of
-expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}
-
-In 1993, a pair of articles in the learning sciences staked a strong
-claim for viewing knowledge as a network of pieces, isolated enough to
-be locally triggered but trainable enough to fire in larger concerted
-patterns (diSessa, 1993; Smith, diSessa, \& Roschelle, 1993). Informed
-in part by agent-based accounts of cognition (Minsky, 1986) and complex
-systems models (diSessa, 2002), the central tenets of an ``in-pieces''
-approach hold that knowledge is emergent from interacting primitives,
-rather than unitary and monolithic. An example from Smith, diSessa, and
-Roschelle helps illustrate the point.
-
-The authors show that we might think of a rubber band as a different
-conceptual entity depending on context. In several different
-situations---wrapped around a newspaper, pulled taut as a string, spun
-to store energy in a toy plane propeller---we intuitively think about
-the rubber band's physical behavior differently: one as a negligible
-part of the newspaper's point mass, another as a transverse pendulum
-(likely) obeying Hooke's Law, and the third as a torsional spring. Those
-differences in intuitive thinking reflect the contextual dependency of
-what we know about the physical world:
-
-In each of the rubberband examples, various pieces of intuitive physical
-knowledge describe the mechanism at work: the rubber band binds the
-newspaper, grips the jar lid, and acts a source of springiness for the
-bobbing object. Although a mapping cannot be made from the rubberband to
-scientific entities, it is quite easy to map these qualitatively
-distinct physical processes to scientific entities and laws. For
-example, instances of binding almost always map to a practically rigid
-body. Likewise, gripping maps to friction forces, and springiness maps
-to Hooke's law. This suggests that applicability can depend directly on
-our intuitive knowledge---knowledge that exists prior to any formal
-scientific training (Smith et al., 1993, p. 144).
-
-The in-pieces approach to modeling cognition has been used, among other
-things, to explain how experts reason about fractions and decimals
-(Smith et al., 1993), how students reason about forces in physics
-(diSessa \& Sherin, 1998; diSessa, 1993; Hammer, 1996; Sherin, 2001),
-how students construct and evaluate algebraic representations of
-physical situations (Izsák, 2004), and how knowledge transfers across
-contexts (Hammer et al., 2005; Wagner, 2006). Because its starting
-assumption is that knowledge is fragmented, knowledge-in-pieces can
-account for wide variations of how people---particularly novices---use
-knowledge on a moment-to-moment basis. In other words, because it
-assumes knowledge is local, it can still explain the kinds of
-globally-inconsistent ways people might reason about physical situations
-(diSessa, 1993). As a framework, an in-pieces approach ultimately argues
-that models of concept replacement and good/bad criteria for knowledge
-should be supplanted by a learning model of alignment/refinement of
-prior knowledge and the consideration of knowledge as
-productive/unproductive.
-
-Hammer and colleagues have worked to extend the in-pieces approach to
-explain how students' epistemological activity---how they orient toward
-knowledge and knowing in a context (Hammer et al., 2005; Hammer \& Elby,
-2002, 2003). Specifically, those authors use two core theoretical
-constructs to explain students' stances toward knowledge and knowing:
-
-\begin{itemize}
-\item
-  \emph{Epistemological Resources} are the epistemological equivalent of
-  diSessa's phenomenological primitives (p-prims). Resources, the
-  authors propose, are the atomic units involved in how people cognize
-  about the source of knowledge, the nature of knowledge, and
-  epistemological activities (Hammer \& Elby, 2003; Louca, Elby, Hammer,
-  \& Kagey, 2004).
-\item
-  \emph{Epistemological Frames} are the emergent result of subsets of
-  resources acting in concert. Drawing from both Goffman's (1974)
-  sociological notion of frame as structures of expectations and
-  subsequent work on framing in discourse (Tannen, 1993),
-  epistemological frames are a participant's local answer to the
-  question ``what is it {[}specifically, what knowledge activity{]}
-  that's going on here'' (Goffman, 1974, p. 8).
-\end{itemize}
-
-Resources can frames can interact in activity settings to produce
-larger-scale patterns called ``epistemological coherences'' (Rosenberg,
-Hammer, \& Phelan, 2006) where evidence from data suggests that a
-network of discrete cognitive units can nonetheless give rise to stable
-cognition.
-
-A useful conceptual metaphor for resources and framing is to think about
-a lecture hall with different sets of lights: a spotlight in the back to
-highlight a lecturer, chalkboard lights, house lights, and a projector.
-Each light has an individual brightness, but the lights are only
-controllable at the per-bulb level. In that sense, they are atomic. But,
-light patterns interact with one another to create a field of lighting
-for the room. Thus, lights are a bit like resources, and different light
-configurations can correspond to different, sociologically-stable uses
-of the room.
-
-\protect\hypertarget{ux5fToc252445958}{}{}Table -- Using room lighting
-configurations to think about epistemological framing
-
-\begin{longtable}[]{@{}lllll@{}}
-\toprule
-\textbf{What is it that's going on?} & Houselights & Spotlight &
-Chalkboard & Projector\tabularnewline
-\midrule
-\endhead
-\textbf{On-stage monologue} & Low & ON & Low & OFF\tabularnewline
-\textbf{Presenting slides} & Low & ON & Low & ON\tabularnewline
-\textbf{Working through equations} & Low & ON & High &
-OFF\tabularnewline
-\textbf{Students discussing with each other} & High & OFF & Low &
-OFF\tabularnewline
-\textbf{Watching a movie} & Low & OFF & Low & ON\tabularnewline
-\textbf{Cleaning the room after a movie} & High & OFF & High &
-OFF\tabularnewline
-\bottomrule
-\end{longtable}
-
-To convince yourself of the sociological stability of these
-configurations, imagine you were watching a movie when suddenly the
-house and chalkboard lights came up to full intensity. It would jar you
-out of the experience. You might wonder whether something was wrong: is
-there an emergency? Should you evacuate?
-
-The metaphor isn't perfect. Ontologically, for example, resource and
-framing theory propose these constructs exist not in the world but in
-the minds of individuals. But, the metaphor is quite useful for
-understanding how individual elements---in this case lights---can work
-in concert to create and sustain a stable frame. And, crucially for a
-cognitive system, the metaphor lets us account for variation in
-activity. The same lecture hall can become a window onto one person's
-thoughts (monologue), a site for instruction (working through equations
-at the board), a shared space for collaboration (student discussion), or
-a place in need of repair (cleaning up) simply by varying the
-intensities of lighting banks in particular ways.
-
-An example helps ground this in-pieces approach to epistemology. In
-Russ, Coffey, Hammer, and Hutchison (2008), the authors describe the
-situation where an elementary student reasons about why an empty juice
-box collapses when you suck on the straw. One student gives what the
-authors deem to be an excellent mechanistic account of why the juice box
-collapses. But, as the teacher seems to steer the discussion toward
-vocabulary---in this case, ``pressure''---the student clearly pulls back
-from her mechanistic reasoning, seems much more diffident, and claims
-that pressure is hard to explain. That example highlights the disconnect
-between doing science as knowing vocabulary and doing science as
-reasoning mechanistically. Moreover, it strikingly highlights that a
-student who by all accounts produced an excellent explanation of how
-pressure works was left nonetheless with the impression that pressure
-was hard to explain.\footnote{In fairness to the teacher, I'm trying to
-  focus on the result of the interaction and far less on the intent the
-  teacher might have had. The student still left with a sense that
-  science might be about vocabulary, even if that's not the view the
-  teacher would espouse or was trying to enact in the moment.}
-
-\subsection{Methods and Theoretical
-Commitments}\label{methods-and-theoretical-commitments}
-
-\subsubsection{Student population, course background, and selection
-}\label{student-population-course-background-and-selection}
-
-This paper focuses on Electrical Engineering (EE) students from Flagship
-State, a large public research institution in the mid-Atlantic. The
-students I studied were taking Intermediate Programming, the second
-semester of an introductory programming course taught in C. The course
-was exclusively for EE majors and taught by EE faculty, and its
-enrollees were typically first-year or second-year EE majors. Its model
-was two 75-minute lectures and a 1-hour teaching assistant-led
-discussion section for students each week. Course topics included:
-
-\begin{itemize}
-\item
-  Strings
-\item
-  Pointers
-\item
-  Dynamic Memory Allocation
-\item
-  Testing
-\item
-  Debugging
-\item
-  Hash tables
-\item
-  Trees
-\item
-  Linked Lists
-\item
-  Abstract Data Types
-\item
-  Functional Decomposition
-\end{itemize}
-
-Students in the course had varying degrees of experience with
-programming. Partly, that variation is because students with Advanced
-Placement (AP) Computer Science credit could place out of Basic
-Programming, the first semester C course. So, some students in
-Intermediate Programming came from technical magnet schools where they
-may have already had one or more years of programming in multiple
-languages, while other students may have been first-time programmers
-with only one semester of programming experience: Basic Programming.
-
-I studied the same course, Intermediate Programming, during the fall
-2011 semester and the spring 2012 semester. During fall 2011 I
-ethnographically observed over 50\% of the course lectures, and during
-spring 2012 I sat in on several discussion sections. At the beginning of
-each semester I solicited student participants for semi-structured
-clinical interviews. Of the students who responded to the solicitation,
-I scheduled interviews opportunistically with students given my
-resources as the sole researcher on the project. In total, I worked with
-a cohort of 6 students in fall 2011 and 4 students in spring 2012.
-
-My specific interest was in how students design computer programs. By
-that, I mean I wanted to know not just how students program, but whether
-and how they structured programs, how they did (or did not) try to
-incorporate modularity into their programs, and how the final form of a
-program reflected what they had learned about how to manage complexity
-through software. In the sections that follow I outline the palette of
-methods I used to pursue those questions.\footnote{The language in
-  sections 3.3.2 and 3.3.3 is taken with slight modification, from
-  ``Studying students' early-stage software design practices,'' a paper
-  I authored, along with William Doane, and submitted to the 2014
-  International Conference of the Learning Sciences (ICLS).}
-
-\subsubsection{Studying design as a complex phenomenon of disciplinary
-practice}\label{studying-design-as-a-complex-phenomenon-of-disciplinary-practice}
-
-Practicing professionals make complex use of talk, gesture, and
-representational artifacts in physics (Gupta, Hammer, \& Redish, 2010;
-Kaiser, 2005; Ochs, Gonzales, \& Jacoby, 1996); chemistry (Stieff,
-2007); field biology (Hall, Stevens, \& Torralba, 2002); civil
-engineering (Stevens \& Hall, 1998); structural engineering (Gainsburg,
-2006); mechanical and electrical engineering (Bucciarelli, 1994;
-Henderson, 1999); and architectural design (Hall et al., 2002; Hall,
-1999). Given that:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Science and engineering education research has made progress by
-  looking for continuities in how novice learners develop disciplinary
-  practices (Gupta et al., 2010; Hall \& Stevens, 1995; Smith et al.,
-  1993; Stevens \& Hall, 1998), and
-\item
-  Emerging research on software engineering reveals that early-stage
-  software design involves complex inscriptional, discursively, and
-  epistemic practices,
-\end{enumerate}
-
-it seems striking that there is no contemporary body of research,
-comparable to studies of expert practice, that looks at students'
-software design practices. In other words, we have every reason to
-believe that expertise in software design involves complex practices,
-but little (if any) research on what productive capacities students have
-for them. Finding those productive design capacities requires a shift
-from questions such as \emph{how can we assess and mitigate students'
-difficulty in programming?} toward questions such as \emph{how do
-students learn and display evidence of design thinking in programming?}
-
-Rephrasing research questions asked of professional software engineers
-(Petre et al., 2010, p. 533) and instead treating \emph{students} as
-designers, I ask:
-
-\begin{itemize}
-\item
-  What do students actually do during early stage software design work?
-\item
-  What does students' exploratory design thinking look like?
-\item
-  How do students communicate?
-\item
-  What sorts of drawings do students create when they design software?
-\item
-  What kinds of strategies do students apply in exploring the vast space
-  of possible software designs?
-\end{itemize}
-
-\subsubsection{Deploying methods to capture the complexity of
-early-stage design
-work}\label{deploying-methods-to-capture-the-complexity-of-early-stage-design-work}
-
-The methods below form the core of my developing program to study
-students as software designers. None of the methods below are new; all
-have been used in prior educational research. What \emph{is} new, I
-believe, is the opportunity to combine them all under the umbrella of
-understanding what happens when students design software. For each of
-the four students in my spring 2012 cohort I captured multiple streams
-of data.
-
-\begin{itemize}
-\item
-  I collected their \textbf{code history}. By this, I mean I preserved
-  frequent snapshots in time of what students' code looks like. Research
-  has already shown code snapshotting to be a useful method for
-  understanding large-scale patterns of student error (Jadud, 2006;
-  Rodrigo et al., 2009; Spacco, Hovemeyer, et al., 2006). And, the
-  resolution of snapshots is extremely fine: Spacco et al. are able to
-  capture the contents of a file each and every time a student saves it
-  to disk. But, that research takes an aggregate view: it identifies
-  large-scale error patterns at the expense of detailed naturalistic
-  understandings of why students make those errors. Moreover, it's
-  primarily used to identify what \emph{mistakes} students make, which
-  is distinctly different from a research orientation that considers the
-  negative \emph{and} positive consequences of students' design choices.
-  A currently untapped advantage of collecting code history data, then,
-  is that while we historically use it to aggregate \emph{across
-  programmers} it nevertheless also gives us fine-grained individual or
-  team-based histories of how designs evolve.
-\item
-  I conducted \textbf{clinical interviews} with them. Clinical
-  interviews have proven historically useful in understanding the
-  substance of students' knowledge and the nature of conceptual change
-  (diSessa \& Sherin, 1998; Duckworth, 2006; Ginsburg \& Opper, 1988;
-  Ginsburg, 1997). Crucially for Computer Science Education (CSEd)
-  research, clinical interviews can tell us about students'
-  epistemologies---how they view knowledge and knowing in a discipline
-  (Hofer \& Pintrich, 1997)---which can affect how they approach and
-  adopt that discipline's practices (Hammer, 1989, 1994; Lising \& Elby,
-  2005). Moreover, because my interviews were videotaped and analyzed
-  from perspectives of interaction analysis (Goodwin, 2000; Jordan \&
-  Henderson, 1995), they offer rich evidence of the substance of
-  students' design practices
-\item
-  I analyzed their in-interview \textbf{inscriptions} when they designed
-  --- what they wrote, how they wrote it, and how those writings got
-  used. Evidence from both science studies (Hall et al., 2002;
-  Henderson, 1999; Hutchins, 1995a; Kaiser, 2005; Latour, 1990; Ochs et
-  al., 1996) and educational research (Hall \& Stevens, 1995; Lehrer,
-  Schauble, Carpenter, \& Penner, 2000; Stevens \& Hall, 1998)
-  highlights the centrality of inscriptions to disciplinary practice in
-  science and mathematics. Ethnomethdological data from studies of
-  professional software engineers supports the same result:
-  inscriptional practice is central to how professional engineers design
-  software (Rooksby \& Ikeya, 2012; Rooksby, 2010). And, since part of
-  the inscriptional environment when designing software is the computer
-  itself, I captured and analyzed what happened on-screen as students
-  design programs.
-\item
-  I paid close attention to students' in-interview \textbf{gestures}.
-  Perspectives of gestural analysis hold that gestures can not only
-  support or extend thinking, they can also communicate entirely
-  different information than what's being said (Goldin-Meadow, 2003).
-  Moreover, perspectives from cognitive anthropology and embodied
-  cognition studies argue that bodily motion \emph{is itself} cognition
-  (Hall \& Nemirovsky, 2012; Hutchins, 1995a; Nemirovsky, Rasmussen,
-  Sweeney, \& Wawro, 2012). For example, when students describe a part
-  of code by moving their hand across an imaginary row of items and
-  tapping each item, their bodies convey information we can interpret
-  about how they understand iteration.
-\end{itemize}
-
-Figure 3-1 presents a visual overview of some of these methods and modes
-of analysis. In particular, the first (top-left) panel depicts how we
-deploy these data collection methods during a clinical interview:
-
-\begin{itemize}
-\item
-  a voice recorder captures speech (often a redundancy in case another
-  recording device fails)
-\item
-  a LiveScribe Pulse pen digitizes written inscriptions, allowing us to
-  play back what was written in time
-\item
-  a videocamera records data for knowledge analysis, interaction
-  analysis, and gestural analysis,
-\item
-  an in-interview computer tracks code history and its screen-capturing
-  software records real-time activity.
-\end{itemize}
-
-\includegraphics[width=6.25556in,height=4.20000in]{media/image4.jpeg}
-
-\protect\hypertarget{ux5fToc252445967}{}{}Figure ‑ -- An comic-based
-overview of my methods for capturing students' design practices
-
-\subsubsection{Developing a revelatory case of students' early-stage
-design
-work}\label{developing-a-revelatory-case-of-students-early-stage-design-work}
-
-In the empirical work that follows I explore what productive knowledge
-and resources students have for structuring programs. I begin with
-Lionel, a male student in Intermediate Programming, fall 2011. First, I
-propose the idea that Lionel connects ``designerly'' (Archer, 1979)
-stances across contexts. I analyze his retrospective account of
-modifying a bike and compare it to his retrospective accounts of how he
-programs in Intermediate Programming. Then I detail Lionel's multi-stage
-cascading approach to structuring a program solution in an interview.
-The argument throughout is that from bike design to code design Lionel
-evinces knowledge and stances that productively help him solve problems.
-
-Next I transition to Rebecca, first explaining the epistemological
-difficulty she felt when coding, then showing how she circumvented that
-block by structuring a program piece using gesture and talk. I argue
-that what Rebecca and Lionel have in common are resources for handling
-design obstacles as they emerge. Where the two students differ is in
-their in-the-moment judgments about what sorts of knowledge and
-activities are appropriate to work through a design problem.
-
-I argue that taken together, Rebecca and Lionel constitute what Yin
-(2009) calls a ``revelatory case,'' by which I mean my investigation
-arises from ``an opportunity to observe and analyze a phenomenon
-previously inaccessible to social science inquiry.'' In my case, I argue
-the phenomenon I'm observing is what engineering students do in the
-early stages of design work on a program. Its inaccessibility is
-demonstrated by the almost complete paucity of studies that draw on
-real-time (e.g., video records) of how students begin work on a complex
-program. Contrary to being ``representative'' or ``typical,'' my
-revelatory case does not aim to generalize. Rather, its value is in
-phenomenological richness. As Yin (2009) writes, and I think others
-including Erickson (1986) would agree, such case studies are worth doing
-``because the descriptive information alone will be revelatory.''
-
-\subsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188553}{}{\protect\hypertarget{ux5fToc252445926}{}{}}Lionel's
-approach to design and to
-programming}{Lionel's approach to design and to programming}}\label{lionels-approach-to-design-and-to-programming}
-
-\subsubsection{While modifying a bike Lionel decomposed tasks, explained
-solutions to himself, and created intermediate design
-representations}\label{while-modifying-a-bike-lionel-decomposed-tasks-explained-solutions-to-himself-and-created-intermediate-design-representations}
-
-Prior to entering my study, Lionel had already earned a Bachelor's
-degree in Business Information Technology at a different university. We
-began the interview by discussing why he re-matriculated to earn an
-electrical engineering degree.
-
-Interviewer: How did you, um end up coming out of your business program
-and your business experience and deciding that electrical engineering
-was what you wanted?
-
-Lionel: Um, well, since I was Business Information Technology, basically
-all my job was doing was writing SQL code and, you know, short
-statements like that. I also worked at a help desk for---I guess I
-worked at Department of Homeland Security. And, pretty much besides the
-SQL statements I worked with this one program and whenever somebody had
-a problem with it they'd come to me. And basically that---that's what my
-whole degree was---was going down that track and then that's really not
-what I wanted to do: just sitting at a desk, you know people coming to
-me being like ``oh, you gotta do this for me.'' So, I was like, alright.
-Or, I have to fix a problem for them, pretty much. And that's really
-what I didn't want to do. I wanted to do something more hands-on where I
-could, you know, I'd be given a project where I'd do something on my
-own, create something on my own. And, um, also my Dad and my uncle were
-both electrical engineers. And my brother became a computer scientist.
-Cu---so I've always been around, you know, that type of environment,
-where my dad's always building something and, you know, I always see it
-and I'm like, you know, that's very interesting. I'd like to do
-something like that too. So it's really just uh, a mixture of bad
-experience with my previous job and being around company that likes to
-do that sort of stuff. (Interview 1 of 1, October 17, 2011)
-
-Lionel apprenticed building things with his father first, taking on
-small, highly-directed roles in projects.
-
-Lionel: Be-before I started my degree, you know, I really didn't know
-how to do any of this stuff. So, I would just kinda help my dad out a
-little bit. He'd be like ``Oh, here,'' you know, ``to, to do this you
-have to do exactly this.'' You know, ``solder this here,'' or, you know,
-``drill this,'' blah blah blah. He'd tell me exactly what to do. So, I
-just kind of followed along. (Interview 1 of 1, October 17, 2011)
-
-Presumably in those projects Lionel was at a periphery and direction
-came from his father. More recently, though, Lionel had struck out on
-projects of his own. As Lionel said, ``now that uh, my dad kinda showed
-me how to do certain typical types of things, I've tried to become
-\textbar{}\textbar{}more adventurous\textbar{}\textbar{} \textbar{}\{air
-quotes\}\textbar{} /Sure/ and do things on my own'' (Interview 1 of 1,
-October 17, 2011).
-
-One of Lionel's recent adventurous projects was adding a motor to a
-beach cruiser bicycle. The gist of the project was simple: using a
-low-priced motor he found online, Lionel realized he could mount the
-motor to an inexpensive bicycle and make a motorized bike. With one
-exception, Lionel was solely responsible for the project from start to
-finish.\footnote{Lionel told me that the only part his father helped him
-  with was showing him how to use a Dremel power tool to smooth, sand,
-  and cut edges (Interview 1 of 1, October 17, 2011).} He bought the
-parts, he mounted the motor, and he maintained it as a working vehicle.
-But, in practice the project was not straightforward. In the next
-subsections I show at length how Lionel addressed the emergent
-challenges of modifying\footnote{Often called ``modding'', for short.}
-his bike with a motor. My analysis focuses on three features of Lionel's
-story that I argue carry over to his programming approach:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  When faced with ambiguity, Lionel worked to decompose large tasks into
-  smaller subtasks he could understand.
-\item
-  Throughout the task, Lionel was metacognitive about the state of his
-  design. His constant refrain was ``\emph{how} am I gonna make this
-  work?''
-\item
-  Before committing work and materials to a solution, Lionel created
-  intermediate designs whose suitability he could evaluate.
-\end{enumerate}
-
-I argue in later sections that these three features of Lionel's
-approach---decomposing subtasks, explaining procedures to himself, and
-explicitly planning prior to committing time and resources to a final
-product---have strong resonances with the approach Lionel took to while
-programming.
-
-\paragraph{Lionel had to decompose Step 1, which simply said ``Install
-the
-Engine''}\label{lionel-had-to-decompose-step-1-which-simply-said-install-the-engine}
-
-Lionel bought an inexpensive motor online with the plan of attaching it
-to a cheap bike. As he discovered, though, frugality came with a hidden
-cost: poor instructions. Attaching the motor to the bike was not at all
-a straightforward process.
-
-Lionel: it was funny. When I, when I got the motor /Mmmhmm/ Um, oh,
-it---like I said it was like cheaply made and whatnot and it came with
-instructions, but um, no exaggeration at all, step 1 said ``install the
-engine''. And that was it. \{Laughs\}
-
-Interviewer: Huh
-
-Lionel: And then step 2 was, um, I guess it was somethin like ``put the
-gas tank on.'' It really didn't tell me anything. It was kinda like
-``OK, those are the obvious steps, but *how* do I do that? So I kinda
-had to figure stuff out on my own. /So/ yup
-
-Interviewer: Did it---how did it end up going? Was it a lot of trial and
-error? Like, did they even have pictures of how it was supposed to look
-when it /There, there/ was installed?
-
-Lionel: There are maybe two or three pictures, but they were black and
-white pictures printed out on like a, I guess like a crappy printer. So
-they were kinda hard to see. Um, so I guess my first step was to go look
-online at better quality pictures. Unfortunately nobody really---I was
-hopin to find you know, like a YouTube video: this is how I did it.
-
-Interviewer: Mmmhmm.
-
-Lionel: I couldn't really find somethin like that, so I just looked at
-pictures. (Interview 1 of 1, October 17, 2011).
-
-Lionel recognized that having an attached motor is an obvious goal
-state, but that goal state was poorly-specified by ``crappy pictures''
-that were hard to interpret. Moreover, the states in between start and
-goal were undocumented and not obvious. Lionel's subsequent search for
-materials that showed a step-by-step progression came up empty. Faced
-with a poorly-specified goal-state and no intermediate guidance,
-Lionel's response was to force himself to think carefully about how he
-would proceed.
-
-\paragraph{In making modifications Lionel frequently asked himself ``how
-am I gonna make this
-work?''}\label{in-making-modifications-lionel-frequently-asked-himself-how-am-i-gonna-make-this-work}
-
-Faced with the challenge of interpolating between ``crappy pictures,''
-Lionel thought hard about what to do.
-
-Lionel: I spent, I guess the, I guess the first five hours I just spent
-kinda staring at it like, ``hmm, *what* am I gonna do?'' Like, ``*how*
-am I gonna make this work?'' (Interview 1 of 1, October 17, 2011)
-
-Ultimately, he was struggling with more than just poor instructions.
-Even if he did figure out how the motor was supposed to be mounted, he
-still faced another problem: the motor was designed for a mountain bike.
-
-Lionel: I had to make a few modifications on my own. Because I guess
-it's built for uh, a mountain bike. And the tubes on mountain bikes are,
-y'know like *that* thin \{makes circle with right thumb and index
-finger\}, but on my beach cruiser there's one tube \{widens thumb and
-forefinger\} that's pretty thick where I had to mount the engine.
-
-Lionel: And since the engine mount was built for a motorized bike it was
-really small. So I kinda had to like devise my own plan on *how* to,
-y'know, add a bigger mount.
-
-Interviewer: Mmmhmm
-
-Lionel: So, that was, that was my first problem. That was, y'know the
-first step was to put the motor on I'm thinkin, well, how am I gonna put
-the motor on if, y'know the, the specs don't even align. So, I guess
-spent y'know five hours just thinkin', like, ``what am I gonna do?''
-
-Lionel: Finally I came up with a plan that seems to be working. So. And
-then, I just kinda went from there. I'd put it together, look at the
-picture and say ``OK, this piece looks like that piece, and y'know I
-guess it looks like it attaches that way, so I'm gonna try that.'' And
-just put it together and, y'know if it fit, it fit. If it didn't, then
-I'd, y'know I'd be like ``OK, maybe that was wrong. Lemme, lemme work on
-somethin' else on it.'' So. (Interview 1 of 1, October 17, 2011)
-
-Asking himself \emph{how} to do something helped bring sub-problems into
-focus. Within a larger mod task (viz., adding a motor to a bike), Lionel
-in effect entered a nested task: modding a mountain bike engine mount to
-fit a beach cruiser. Modding the mount itself was entirely undocumented,
-so Lionel's solution in part was to look for affordances in the physical
-structure of parts to see what he could attach. Crucially, the
-environment afforded Lionel rapid, iterative feedback. It wasn't hard to
-tell when a design choice worked because ``if it fit, it fit.''
-
-\paragraph{Lionel used intermediate representations to consider
-alternative
-designs}\label{lionel-used-intermediate-representations-to-consider-alternative-designs}
-
-Lionel grew increasingly frustrated as he saw how difficult it would be
-to mount the motor. Crucially, part of his tactic around that
-frustration was to resolve to keep trying and thinking of possible
-solutions.
-
-Lionel: So, I guess I had to give myself, like a few hours just to cool
-down and then actually think ``OK, well maybe, you know I've already
-bought it, so, I'm gonna continue on with it. So I had to think, you
-know, OK, what's my next step. *How* could I make this work? You know,
-I---I thought of a few different ideas. You know, I'd think of one idea
-and think ``OK, would this work? It might, but let me think of another
-idea.'' So I'd come up with another idea and say, ``would this work?''
-You know, so once I got a few ideas I would just kind of, in my head
-picture it. Say, ``OK, this is how it would mount, or this is how it
-would mount. Which one would be safer? Or will one of these not even
-work?''
-
-Interviewer: Right.
-
-Lionel: So, I guess after, you know, awhile of deciding which one was
-better, you know, which one would be easier and safe at the same time
-/mmhmm/ so then, that's when I made my decision, and just kinda, went
-for it to see if it would work. And luckily it did \{laughs\}.
-(Interview 1 of 1, October 17, 2011)
-
-Before committing to a design, Lionel forced himself to evaluate
-alternatives first. Naively, he could have gone with the first idea that
-came to him. Instead, even if he thought an idea might work he pushed
-himself to consider others first. Then, with a collection of possible
-designs he evaluated what he saw as their suitability in his head. He
-applied criteria as he evaluated, trying to balance the ease with which
-a mounting scheme might work with a concern for the safety of the
-user---in this case, himself. Again, at least in Lionel's telling, those
-steps happened \emph{before} committing to a final implementation. In
-other words, Lionel generated and evaluated multiple configurations with
-an eye toward both the builder of and user of those configurations
-before settling on a path to take.
-
-In sum, I note three key features of Lionel's approach to the design
-challenges of modifying his bike. First, he repeatedly asked himself
-``how am I gonna do this,'' forcing himself to think in terms of a plan
-or procedure to fill in the gaps between the instructional pictures.
-Second, he proposed ideas, tried to evaluate their feasibility, and
-continued pressing for alternate solutions (``would this work? It might,
-but let me think of another solution''). Third, so much of his planning
-took place before he ever mounted the motor. By his account, he fiddled
-with pieces, envisioned multiple ways the engine could mount, and
-evaluated their feasibility before ever taking concrete steps to mount
-the motor.
-
-\subsubsection{Lionel's approach to programming resonates strongly with
-the design stance he held when working on his
-bike}\label{lionels-approach-to-programming-resonates-strongly-with-the-design-stance-he-held-when-working-on-his-bike}
-
-\paragraph{Lionel thinks about steps at a
-chalkboard}\label{lionel-thinks-about-steps-at-a-chalkboard}
-
-Later in the interview I asked Lionel to compare his experiences in
-Basic Programming with those he was having in Intermediate Programming.
-His response revealed a great deal about how he starts projects in
-Intermediate Programming: he starts at a chalkboard.
-
-Lionel: Well, {[}Basic Programming{]} I already learned the basics of
-the class so I know the basics of where to start my code. And then,
-obviously the projects in this class {[}Intermediate Programming{]} are
-harder than last class, but I have the basics of, ``OK, well, I know how
-a program runs. So how can I translate that into my program, for, for
-{[}Intermediate Programming{]}?'' And, yeah, so I guess I'll sit down,
-like, yeah, and I'll have like a chalkboard and I'll write things out,
-and say OK well first, you know, \{sweeps hands{]} not even think about
-the code. Just, in \{air quotes\} English words, so you know, my
-project's gonna do this. You know, first I need to make this
-calculation, then I'm gonna print this. And then I'm gonna make this
-calculation. Et cetera. And I'll sit down and write that out, in the
-order that it needs to be done. And then from there I'll go back and
-think, ``OK, what's the code, to, you know, make this calculation or
-print that?''
-
-Interviewer: So, when you start a project, you don't start at a
-computer, it sounds like you kinda start at your chalkboard.
-
-Lionel: Right. Exactly. Yeah, cuz I've tried to start a project at a
-computer, and you know I'll write a---I'll start writing a few lines of
-code, but then I just really zone in on what I'm doing and kinda lose
-sight of the whole picture.
-
-Lionel: So then I'll write this code for, you know, this one little
-section of the project, and then after I've done that I'll look back and
-say ``Oh, yeah, I need to write the whole project,'' but then this
-little code doesn't actually fit into my project, so it's kinda, you
-know I guess I get tunnel vision when I start at a computer. (Interview
-1 of 1, October 17, 2011)
-
-For Lionel, the chalkboard is a pivotal object in his design activity.
-It holds the top-down ``English words'' description of his program. In
-so doing, it serves as the earliest durable representation of his
-program's intended hierarchy and relationships. That function is
-important, because without it Lionel is apt to ``lose sight of the whole
-picture'' of what his code is supposed to do. Moreover, as the next
-section reveals, the board continues to help Lionel orient his activity
-as he digs into the specifics of how his program will work.
-
-\paragraph{Lionel copies pseudo-code into the computer even though it
-won't
-compile}\label{lionel-copies-pseudo-code-into-the-computer-even-though-it-wont-compile}
-
-As Lionel described the role of the chalkboard, I initially thought
-there was a distinct separation between the chalkboard and the computer
-as activity objects. The chalkboard, I assumed, was where Lionel
-sketched out design plans while the computer was where he wrote code.
-When I probed Lionel based on my hunch, I discovered Lionel's activity
-wasn't cleanly separable into design on the board / code on the
-computer.
-
-Interviewer: Did you, when you start thinking about the code, do you end
-up usually filling it in on your board, or will you actually go straight
-over to the computer and do it?
-
-Lionel: Um, on my board, I'll do, I guess like
-\textbar{}\textbar{}pseudo\textbar{}\textbar{} \textbar{}\{air
-quotes\}\textbar{} code /mmhmm/ uh, you know, first on the board it
-would only say ``this function calls this function, this function calls
-this function.''
-
-Interviewer: Mmmhmm
-
-Lionel: So then on the board I'd write down, ``alright, well this
-function'' um, I dunno, for instance I'll, I'll say you know, I'm workin
-with a function that checks to make sure that it's on the---that I'm
-moving *on* the board between spaces 1 and 24.\footnote{At that point in
-  the semester, Lionel's class was working through a Backgammon project,
-  so the spaces and moves he's referring to pertain to programming a
-  basic Backgammon game.} So then, you know I'll write down like,
-pseudocode, write ``OK, well I'm gonna start. I'm gonna have a for-loop
-that will go from 1 to 24, and then here I'll, you know I'll have an
-if() statement that says if it's between you know, 1 and 24 it's valid,
-if not it's invalid.'' So, that'd be my pseudocode, and then when I
-actually---I'd, I'd try to write that out for the entire program, or as
-much as I could.
-
-Interviewer: Filling out the tree, you mean.
-
-Lionel: Right, fillin out the tree. And then I'd go to my computer, so
-that at least I already had a baseline of what to work with. And then,
-on my computer I'd, you know I'd write out the pseudocode, I'd ac---I'd
-literally write out the pseudocode, even though obviously it wouldn't
-compile and actually work.
-
-Interviewer: Mmmhmm
-
-Lionel: So, then, from the pseudocode, as, as I read down on my computer
-what the pseudocode said, I'd actually try to write the actual code that
-would work. (Interview 1 of 1, October 17, 2011)
-
-Lionel blurred my initial chalkboard / computer distinction in two ways.
-First, his overall ``English words'' plan of functions calling functions
-evolved on the chalkboard to include computational idioms, including
-if-statements and for-loops. But, then, according to Lionel, he would
-copy that pseudo-code as-was into the computer, knowing full well it
-wouldn't compile. These observations---that Lionel writes computational
-pseudo-code on the board, then copies that same code into the
-computer---may seem trivial, but I contend they aren't.
-
-Because of the choices Lionel makes, neither chalkboard nor computer has
-a distinct, privileged kind of syntax. Rather, ``English words'' and
-pseudo-code can peacefully co-exist on the chalkboard, and non-working
-pseudo-code gets deliberately typed into the computer before it's
-expanded out into ``actual code.'' Put another way, Lionel is capable of
-distinguishing between ``English,'' ``pseudo-code,'' and ``actual
-code,'' but all three forms of expression are a part of his design
-process. It's perfectly acceptable for him to have something other than
-``actual code'' exist in intermediate representations of his design.
-
-\subsubsection{Gestural, inscriptional, and verbal in-interview evidence
-reveals Lionel's resources for structuring a
-program}\label{gestural-inscriptional-and-verbal-in-interview-evidence-reveals-lionels-resources-for-structuring-a-program}
-
-To probe Lionel's retrospective account of how he programs, I gave him
-an in-interview task to work on. In this section, I argue Lionel's
-development process on that task moved rather linearly through what I
-call a ``representational cascade.''\footnote{As I'll explain later, I
-  think it would be a mistake to assume that all development activity
-  proceeds linearly and top-down in the manner Lionel's does. Not only
-  do I think Lionel's clean linearity is rare, I also think it's not
-  normative. So, my point here isn't to stress that Lionel moved
-  linearly, monotonically through the cascade. Rather, my point is to
-  show that he moved between different representational levels \emph{at
-  all}.} My schematic for that cascade --- along with the features and
-affordances in each layer of the cascade --- is depicted in Figure 3.
-What I show here is that actually, the final code Lionel produces is in
-fact one representation of a procedure that actually existed in
-different forms and modalities before it became the final code in Figure
-3.
-
-In the analysis that follows, I explore two of those modalities---verbal
-description and written pseudo-code---and explain why each is crucial to
-understanding the final produced code. Ultimately, I argue there is a
-need to explain \emph{how} Lionel creates and moves fluidly between
-these markedly different cascade layers. Specifically, I attempt to
-answer the question of \emph{Why does he see the upper layers (verbal
-code, hand-written pseudocode, and handwritten source code) as
-legitimate activity?} My candidate answer is that Lionel's approach to
-programming is in part stabilized by epistemological resources that
-allow him to treat the upper layers of the cascade as valid, productive
-knowledge expressions in programming. Such an answer is strongly in line
-with the findings from sections 3.4.1, where we see Lionel deliberately
-making use of other intermediate design representations.
-
-\includegraphics[width=5.74444in,height=4.31111in]{media/image5.jpeg}
-
-\protect\hypertarget{ux5fToc252445968}{}{}Figure ‑ -- An overview of
-Lionel's representational cascade and the features and affordances of
-each representational layer.
-
-\paragraph{The range-finding prompt and Lionel's final source
-code}\label{the-range-finding-prompt-and-lionels-final-source-code}
-
-About 35 minutes into our interview, I gave Lionel the programming task
-reproduced below:
-
-Suppose you want to write a program that finds the range of a set of
-numbers. So, the program would take 10 numbers as input, then it would
-compute the range as the difference between the largest number and the
-smallest number.
-
-How would you write such a program? (Interview prompt, October 10, 2011)
-
-The prompt was deliberately vague about certain constraints. It did not
-prescribe \emph{how} the program should get input (Should the programmer
-expect the user might type it in, or feed in a formatted file?). It also
-did not prescribe how the program should provide the result (should the
-programmer print the result on screen? Store it to a file? Return the
-result to be used by another function?) The reasons for vagueness were
-in part because I wanted to see what assumptions students might make
-during their design process. Some solutions, I reasoned, might decouple
-the procedures of getting input, computing range, and returning output.
-Others might simply hard-code a procedure for getting keyboard-based
-input. I wanted to prescribe no specific approach up front if I didn't
-have to. I have reconstructed his original source code from
-screen-capture footage of his in-interview programming:
-
-\includegraphics[width=5.74444in,height=7.11111in]{media/image6.emf}
-
-\protect\hypertarget{ux5fToc252445969}{}{}Figure ‑ -- Lionel's final
-source code for the range-finding prompt
-
-Let's explore some top-level features of Lionel's code. First, Lionel
-chose to have a user enter text directly using the keyboard (lines
-9--12), and his program displays the range result as printed text (lines
-28--29). His strategy for finding the maximum and minimum is to
-implement a procedure that iterates through the array of numbers and
-compares the current number to the globally stored max and
-min.\footnote{I think of this as a ``king of the court'' procedure. The
-  current value stays on as an extreme if and only if it beats the
-  current ``king'' for that extreme. Otherwise, the challenger is
-  replaced and the contest resumes with a new challenger until there are
-  no more challengers.} If the current number (array{[}i{]}) ``beats''
-the global min (by being lower than min), its value replaces that of the
-global min. Similarly, if the current number exceeds the global max, its
-value replaces that of max. Before the procedure starts, Lionel assigns
-the first number of the array as the value for both extrema.
-
-\paragraph{Lionel's verbal description of the
-program}\label{lionels-verbal-description-of-the-program}
-
-Lionel chose to start the task on pencil and paper. After scribbling a
-line about scanning input into an array (which I discuss in a later
-section), Lionel stopped and began explaining out loud to me.
-
-Lionel: so I'm thinkin, my general concept is I'm going to parse through
-this array, check each number /Mmmhmm/ and I'll have two variables, min
-and max, and, first I'll set both min and max to the first variable in
-array, er, yeah first number in array /Mmmhmm/ and, I'll parse through
-the rest of the array, checking to see if,
-
-Lionel: Umm, you know if---I'll s---I'll start with max, for example.
-/Mmmhmm/ And I'll check to see if the number in that section of the
-array is bigger than max. /uh-huh/ And if it is, I'll set max to that
-new number. /OK/ And then I'll do the same thing for minimum. And then
-at the very end, I'll have my minimum and max and then I'll just
-subtract `em.
-
-Interviewer: OK
-
-Lionel: So that's my main concept. (Interview, October 17, 2011)
-
-This verbal description is in some ways bound by the least stringent
-rules of expression. If we're hard-nosed about syntax, Lionel's speech
-stops and starts (line 5 into line 6), which leans on the unspoken
-understanding that ideas in talk are under negotiation. Stopping
-mid-sentence and leaving that sentence unfinished is a valid and
-acceptable move in conversational space. Moreover, there is a strong
-mapping between Lionel's temporal language and the intended
-flow-of-control for the program. ``First I'll set,'' ``then I'll do,''
-``and then at the very end'' all describe when instructions will be run.
-
-Finally, Lionel's instructions begin at the high level, and in some
-cases implementation details are completely left out. For example,
-Lionel does not specify precisely how he'll ``parse through the array.''
-At this stage of the cascade, it seems sufficient for him to explain
-\emph{that} he plans to parse through, but now how. What Lionel
-\emph{does} do is gesture while describing ``parsing''. This gesturing,
-I argue, creates a concrete representation of iteration and
-incrementation --- the gestural marker of a loop.
-
-\includegraphics[width=5.74444in,height=3.23333in]{media/image7.png}
-
-\protect\hypertarget{ux5fToc252445970}{}{}Figure ‑ -- Lionel makes a
-cycloid/helix gesture as he says ``parse through this array''
-
-Contrast how Lionel specifies ``parsing'' through the array with how he
-describes ``checking'' its contents. First, Lionel explains that he'll
-``check each number'' (line 2), decides that he has underspecified how
-``checking'' happens (line 5--6), and finally makes ``checking''
-relatively more explicit (lines 7--8). Once checking for the maximum has
-been made explicit (lines 7--8), Lionel says ``and then I'll do the same
-thing for minimum'' (line 9). Line 9 thus becomes the verbal equivalent
-of a repeated, parameterized procedure. It is, in a sense, the
-talk-space seed of a function. Conversationally, both participants
-understand that Lionel saying, ``then I'll do the same thing for
-minimum'' (line 9) refers to a procedure defined elsewhere in the
-conversation, not a new set of instructions.
-
-\paragraph{Lionel's pseudo-code}\label{lionels-pseudo-code}
-
-Immediately following the verbal description above, Lionel begins
-creating what he calls his ``pseudo-code,'' narrating it as he writes.
-There are four key features in Lionel's talk (below) that I discuss in
-my analysis. Below is an overview of each feature and what importance I
-think it has:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Lionel iterates over his initial verbal description by fleshing out
-  some implementation specifics --- structures that further specify
-  \emph{how} something in a program is to be done.\footnote{It's a
-    limitation of my writing style that in this document I use
-    ``iteration'' both to describe aspects of Lionel's design process
-    and as a formal term of art in computer science. In both cases,
-    though, the conceptual meaning is the same: the incremental,
-    rule-based set of transitions to the state of an object.} This
-  iteration is important because it suggests that his subgoals --- and
-  criteria for satisficing them --- change from one micro-coherence to
-  the next.
-\item
-  Reconciling what Lionel \emph{narrates} with what Lionel \emph{writes}
-  reveals peculiar mismatches. The mismatches suggests a productive
-  capacity: that Lionel can \emph{background} certain details within a
-  micro-coherence by omitting them from the written code while still
-  maintaining an understanding of how his program will work. Put another
-  way: the inscriptional object in a microcoherence captures what Lionel
-  thinks is important within that microcoherence, but the inscription
-  does not necessarily represent the totality of his conceptual
-  understanding.
-\item
-  At one point in his code, Lionel writes ``same for min,'' which we can
-  reasonably infer means he intends to repeat a procedure for
-  calculating the minimum. That line encapsulates the precursors of
-  functional abstraction: Lionel not only recognizes procedures that
-  share the same structure, he can background the specifics (point 2) of
-  code he repeats.
-\end{enumerate}
-
-Lionel: So it'll be scanf, um, and I'll have my for-loop, and I'll just
-have my int-i, and I'll go from zero uh, zero to, four actually cuz
-it'll be zero, one, two, three, four, yeah, zero to four. /Mmmhmm/
-
-Lionel: Umm, and I'll have, oh, up top here, to set the, right---right
-after the scanf /Mmmhmm/, I'll just say min and max equal the
-first---mm, hiccup, sorry
-
-Interviewer: It's OK.
-
-Lionel: \{laughs\} the first number in the array. So I'll say min and
-max equal array of zero. And that's just setting them so that I have
-something to compare to, I guess. /uh-huh/ I think that will work. Yeah.
-
-Lionel: So, so then, as I parse through my array of five variables, I'll
-say ``if min,'' er no, I'll say ``if array of i, you know of, of that,
-if, uh, the number in the array that I'm looking at is greater than max,
-then max equals that number, array of i'' /Mmmhmm/ and then I'll do the
-same thing, same for min. And then at the very bottom, I'll have, uh,
-range equals max minus min.
-
-Lionel: So that's my pseudo-code, just so that *I* understand what's
-goin' on in my head.
-
-\includegraphics[width=5.75000in,height=3.00208in]{media/image8.emf}
-
-First, Lionel's talk iterates over his initial verbal description by
-fleshing out \emph{implementation details} --- programming structures
-that further specify \emph{how} something is to be done. Compare here
-how Lionel instantiates ideas of iteration in Table 1 below.
-
-\protect\hypertarget{ux5fToc252445959}{}{}Table -- Comparing Lionel's
-words and actions across different stages of ``programming'' over the
-same putative section of his program. I have bolded some
-language-specific syntax.
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule
-\textbf{Design Coherence} & What he says & What he does\tabularnewline
-\midrule
-\endhead
-\textbf{Verbal Description} & ``I'm going to parse through this array,
-check each number'' &
-\includegraphics[width=2.22222in,height=1.48889in]{media/image9.jpeg}\tabularnewline
-\textbf{Narrated pseudocode} & ``I'll have my \textbf{for-loop}, and
-I'll just have my \textbf{int-i}, and I'll go from zero uh, zero to,
-four actually cuz it'll be zero, one, two, three, four, yeah, zero to
-four. &
-\includegraphics[width=2.23333in,height=0.35556in]{media/image8.emf}\tabularnewline
-\bottomrule
-\end{longtable}
-
-In his narrated pseudo-code, he introduces the C-specific syntax of
-``for-loop'' and ``int-i,'' which handle iteration and indexing,
-respectively. Moreover, he counts in order to specify the index bounds
-of the loop (zero to four).
-
-Second, it's actually somewhat complicated to address the question ``how
-do Lionel's inscriptions align with what he says?'' He \emph{says}
-``I'll have a for-loop'', but the visual evidence shows that he hasn't
-bounded his for-loop body in curly-braces.\footnote{As a language, C
-  typically uses curly braces to bound instructions that span multiple
-  lines. So, a typical for-loop would be written like this, where the
-  loop preamble is bounded by parenthesis and the loop body is bounded
-  by curly braces (bolded):
-
-  for(int i=0; i \textless{} maximum\_value; i++) \textbf{\{}
-
-  // Do something
-
-  \textbf{\}}} If he had done so, his for-loop preamble would end with
-an opening curly brace; it doesn't (Table 1 graphic above). Also, for
-matching, his for-loop should have concluded with a closing curly brace;
-it doesn't. Nevertheless, his verbal evidence --- particularly him
-saying ``int-i'' --- suggests that he understands details that variables
-need to be declared. Together, these mismatches suggest that features
-(viz., denoting iteration and setting its bounds) are important enough
-to be included in this representation. Other specifics, including proper
-declaration of variables, are things Lionel understands but does not
-insist on writing.\footnote{One might argue that Lionel is simply being
-  sloppy and forgetting to include details, even though he said them.
-  The point I'd make here is that even if he is being sloppy, I would
-  think it unfair to say, for example, that Lionel has conceptual issues
-  with declaring variables. His talk (and many later iterations of his
-  design) suggests he understands that declarations are necessary for
-  working with variables in C. Moreover, even if he is being sloppy on
-  the particular point of variable declarations, the overall character
-  of this pseudo-code is still starkly different from that of the full,
-  working source code solution he'll ultimately produce. So, even if we
-  propose that Lionel has some unintended tolerance --- such as
-  sometimes overlooking missing variable declarations --- for assessing
-  whether his pseudo-code satisfices, there is still strong evidence
-  that his criteria for evaluating his work in pseudo-code stage are
-  markedly different from the criteria he'll apply to later iterations
-  of design.
-
-  no one would confuse the pseudo-code skeleton here for Lionel's
-  fleshed-out, fully-working code later. So, even if Lionel \emph{meant}
-  to put them in and forgot, he still judged this pseudocode as
-  satisficing his subgoal} Together, and combined with later evidence of
-Lionel's successful working program, these observations suggest that
-ignoring the details is actually productive for Lionel.
-
-Third, Lionel's talk reveals productive capacities for dealing with
-design work. For example, he describes the guts of his program through
-brief perspective-taking (lines 14 -- 18 above). As a reminder, Lionel's
-procedure is what I refer to as a ``king of the court'' approach: if the
-number currently-being-inspected exceeds the reigning champ, the champ
-is dethroned and replaced by the number currently-being-inspected. But,
-the way Lionel articulates that process actually uses multiple senses of
-``I,'' where it's at times actually not clear whether ``I'' means ``I
-the programmer'' or ``I the program.''\footnote{To keep your head from
-  exploding, I remind you that the lower-case version of ``i'' in
-  Lionel's talk actually refers to the subscript for indexing an array.}
-For example, the I in ``I'll say `if the array of i'' could very well be
-the programmer, where by ``I'll say'' he means ``I the programmer will
-write a conditional test.'' But, that phrase is immediately followed by
-``if, uh, the number in the array that I'm looking at is greater than
-max'', where I believe we may be seeing Lionel taking the perspective of
-the program itself.\footnote{My point here is that technically, Lionel
-  can never inspect an array element himself; he can only write code
-  that does that. You might disagree, in which case we get to have a
-  really fun discussion over beer about how we think programming is a
-  case of ``distributing cognition'' (Hall, Wright, \& Wieckert, 2007).}
-If anything, the seemingly untroubled, interchangeable use of ``I'' here
-might suggest that to do the work of programming one can occupy a hybrid
-space of both being the program and being the programmer at the same
-time.
-
-Lionel concludes this procedure with the invocation ``same for min.''
-Picking up the thread of hybridity, this phrase could mean, ``I the
-programmer will write a similar procedure for min'' or ``I the program
-will perform similar operations to determine min.'' That Lionel leaves
-his pseudocode there (``so that's my pseudocode'', line 20) suggests
-that whatever perspective it might be written from, this designed
-artifact satisfices his pseudo-code production subgoal. And, that
-satisfaction is remarkable because if ``same for min'' is an appropriate
-place-holder, it suggests Lionel has ways of seeing that procedure as
-\emph{similar enough} in structure to the procedure already laid out as
-to not require further specification. This is an important point: if a
-precursor to functional abstraction is recognizing repeated code, Lionel
-has actually recognized repeated code \emph{before ever writing it}. As
-we'll see later, he ends up writing this procedure out explicitly, which
-means that whatever answer we might come up with as to why Lionel
-doesn't abstract this to a function \emph{can't} be that he doesn't see
-the repeated computation; to the contrary, repeating core parts of the
-max computation was built into his design from the start.
-
-\subsubsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188557}{}{\protect\hypertarget{ux5fToc252445930}{}{}}We
-can trace continuities in Lionel across different kinds of design
-activities}{We can trace continuities in Lionel across different kinds of design activities}}\label{we-can-trace-continuities-in-lionel-across-different-kinds-of-design-activities}
-
-We can trace patterns in Lionel's design activity by asking questions
-about the intermediate design artifacts he creates. Specifically, we can
-pay attention to \emph{what} representations he creates, \emph{where}
-they live, and \emph{how} he uses them. Table 3 outlines those
-relationships for the three contexts discussed: Lionel's bike modding
-project, Lionel's retrospective account of his programming practices,
-and Lionel's in-interview work on the range-finding task. Table 3 helps
-support three key points:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  There is a specific consistency to Lionel's design activity across
-  contexts. All activities---given the limits of Lionel's
-  self-reports---\emph{involved the generation and use of intermediate
-  design representations}. Those representations may at times have lived
-  in Lionel's head, on paper, or in electronic format.
-\item
-  A thorough account of Lionel's in-interview programming activity
-  cannot and should not ignore those intermediate representations and
-  how he used them.
-\item
-  Because intermediate representations are a feature of all three design
-  contexts, it seems likely that what stabilizes their creation and use
-  is tied to something deeper or more global than just conceptual
-  knowledge in a domain such as computer programming.
-\end{enumerate}
-
-\protect\hypertarget{ux5fToc252445960}{}{}Table -- Comparing Lionel's
-design activity across contexts
-
-\begin{longtable}[]{@{}llll@{}}
-\toprule
-\textbf{Context / Question} & Modding the bike & Retrospective
-programming account & In-interview task\tabularnewline
-\midrule
-\endhead
-\textbf{\emph{What} intermediate representations did he create?} &
-Possible configurations for mounting the engine & Relationships of
-functions to each other; pseudo-code for how functions would work & A
-verbal description; narrated pseudo-code; hand-written pseudo-code;
-hand-written source code\tabularnewline
-\textbf{\emph{Where} do the intermediate representations live?} & In
-Lionel's head & On a chalkboard and, later, in the text editor & In
-Lionel's talk and gestures; on paper; in the text editor\tabularnewline
-\textbf{\emph{How} did he use those intermediate representations?} & To
-compare mounting configurations for ease and safety & To ``see the whole
-picture'' of a project & To ``understand what's goin on in my head''
-when structuring his code\tabularnewline
-\bottomrule
-\end{longtable}
-
-\subsection{Rebecca's approach to
-programming}\label{rebeccas-approach-to-programming}
-
-In this section, I analyze data from a different student --- Rebecca. I
-offer this data as an example of what developing design expertise might
-look like. If Lionel's approach to programming uses high-level work
-(e.g., his pseudo-code) to cascade down the specifics of syntax, it
-seems reasonable to ask whether other students use pseudo-code in the
-same way. And, if they do not, it seems worth asking how uses of
-pseudo-code differ and what that difference might tell us about
-students' approaches to programming.
-
-In Rebecca's case, the data I'll present begins with her stuck on the
-third of four projects in the course: the ``iTunes'' project.\footnote{In
-  this project, students were given a series of text files ---
-  representing the user's music collection --- that contained formatted
-  albums and song titles. The task was to create a program to read in
-  the music information, create a database representation of it, and
-  allow the user to transact with that database. Typical transactions
-  might be asking for a listing of all the albums in a user's library or
-  creating a playlist by allowing the user to choose specific songs.} My
-analysis will expand on the following points:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  To understand Rebecca's struggle on the iTunes project and her
-  approach to programming during that struggle, we need to understand
-  some of her prior experiences in programming. In our fourth interview
-  Rebecca recalls a pivotal project experience from her Basic
-  Programming course. At the cost of hours of work, half her codebase,
-  and a good grade, Rebecca discovered that her sense of how her program
-  should work did not match what was expressible in C. I analyze
-  Rebecca's recollection because it sits at the intersection of
-  self-efficacy and disciplinary practice. Rebecca lost confidence and
-  felt ``demoralized'' because her attempts at expressing a procedure
-  did not match the constructs and underlying workings of the C
-  language. Moreover, the fallout from that experience persisted into
-  Intermediate Programming. By our fourth interview she had considered
-  dropping Intermediate Programming entirely because of her low grades.
-  ``My *biggest* problem'' in programming, she said, ``is I think I have
-  the logic, my logic just doesn't transfer to code. Or I don't know
-  *how* to transfer it correctly'' (Interview 4 of 5, April 6, 2012).
-\item
-  Rebecca has the capacity to make sense of program designs. As she
-  tries to understand why the instructor would prescribe an array of
-  pointers to store track titles in the music server, Rebecca makes
-  progress by reasoning about what she knows about pointers and arrays.
-  Like Lionel, Rebecca can use gestures, talk, and inscriptions (viz.,
-  pseudocode) to express how she might intend for her program to work.
-  Rebecca also has a grasp of the conceptual issues --- from a
-  computational and programming perspective --- relevant to the part of
-  her design she's struggling with. But, Because Rebecca's approach to
-  programming so strongly orients toward producing valid, working
-  syntax, it may be limiting Rebecca from seeing other ways in which she
-  can make progress on her design. In other words, Rebecca \emph{can} do
-  things like write pseudo-code to plan, but she doesn't tend toward
-  those behaviors when she's stuck. I'll offer an explanation based on
-  epistemological dynamics to explain why that's the case.
-\end{enumerate}
-
-In what follows, I first present point 1 to give readers a sense of how
-Rebecca's prior experiences may inform her approach to programming. I
-then outline the particular part of the iTunes project where she feels
-``stuck,'' discussing point 2. Before I jump into that analysis, I
-\emph{strongly} urge readers who are not familiar with pointers or
-dynamic memory allocation to refresh themselves on the subject.
-
-\subsubsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188558}{}{\protect\hypertarget{ux5fToc252445932}{}{}}Rebecca's
-struggles on the iTunes project make sense in light of some prior
-programming experiences she
-had}{Rebecca's struggles on the iTunes project make sense in light of some prior programming experiences she had}}\label{rebeccas-struggles-on-the-itunes-project-make-sense-in-light-of-some-prior-programming-experiences-she-had}
-
-At the time of this study Rebecca was a first-year electrical
-engineering major who had gone to high school in rural Maryland. Her
-university courses in Basic and Intermediate Programming were the first
-experiences she had doing any programming at all, let alone in the C
-language. Though she felt academically at ease in high school, a tough
-project in her first-semester Basic Programming course shook her
-confidence in programming. In the data presentation and analysis that
-follows, there are four sub-points I'd like to make about Rebecca's
-development in programming.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Prior to the semester I worked with her, Rebecca had difficult
-  experience trying to understand arrays on a Basic Programming course
-  project.
-\item
-  That difficult experience had two pieces to it that I believe relate
-  to personal epistemology:
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \item
-    Rebecca came to feel that her idea for a solution wasn't something
-    expressible in C.
-  \item
-    When she got help from a friend who was ``very smart in
-    programming,'' she felt like she typed the code he told her to into
-    her project without really coming to understand it.
-  \end{enumerate}
-\item
-  Reflecting on it in our first interview, Rebecca pinpoints the
-  experience of that array project as demoralizing and having a lasting
-  effect on her confidence in programming.
-\item
-  Epistemological issues continue to factor into Rebecca's confidence in
-  programming. By our fourth interview, Rebecca was getting low grades
-  in Intermediate Programming. She had strongly considered dropping the
-  course. Trying to articulate her ``biggest problem,'' Rebecca
-  explained that she felt like she understood the theory behind the code
-  examples in class but had difficulty transferring her logic into code.
-\end{enumerate}
-
-\paragraph{Rebecca had a difficult experience on a Basic Programming
-arrays
-project}\label{rebecca-had-a-difficult-experience-on-a-basic-programming-arrays-project}
-
-In our first interview, I asked Rebecca about how her experiences in
-Basic Programming compared to those she was having in Intermediate
-Programming. She explained that in that first semester, she ``kinda got
-lost on one part'' --- arrays --- ``that was key to the rest of the
-year.''
-
-Rebecca: {[}I{]}n {[}Beginning Programming{]} we had three projects, and
-the first one, uh, we did---I did discuss it with other students, and we
-all worked together kind of. But that one I felt much more like I had a
-grasp on everything I was doing. Whereas the second project, I would
-stare at the code and be like ``ohmygosh, I don't even understand what,
-like, it's asking me to do.'' Like, or how to make it do what it needed
-to do, which was to like store in numbers in arrays, change them, and
-add `em together and stuff. (Interview 1 of 5, February 10, 2012)
-
-I asked Rebecca if she felt she had a better grasp by the time she
-turned in the project.
-
-Rebecca: No. Cuz, there was a friend on my floor, who---very smart at
-programming, he's had like years of experience. And he gave me a lot of
-pointers and tips doing it, so I felt like I was just using him a lot
-instead of actually learning it, which was a downfall for me.
-
-Interviewer: Huh. So, how, I mean. I feel like sometimes you could---you
-might ask somebody for help, and then they show you something and then
-you might be like oh, you've learned it now /yes/, but it sounds like
-you're saying /no/
-
-Rebecca: Just, I don't, like, I don't blame him at all because he helped
-me, but, it was kind of more of a, he showed me how to do it, I was like
-\{mimes exaggerated typing, in an almost singsong voice\} *OK, sure,
-I'll type that code*, but I don't really, I never really, I don't know I
-never made the connection. (Interview 1 of 5, February 10, 2012).
-
-\paragraph{Rebecca's ``terrible'' array experience has epistemological
-components}\label{rebeccas-terrible-array-experience-has-epistemological-components}
-
-In our second interview a week later when we discussed debugging,
-Rebecca returned to that experience in her Basic Programming class:
-
-Rebecca: My array project, my---the second project we did last semester,
-that I had a big error. I had to delete like half the code I worked on.
-Like, just cuz, like I had said last week, I didn't do very well on
-arrays. So.
-
-Interviewer: Oh. Was it cuz of one thing, or was it like a repeated---
-
-Rebecca: It was kind like a general, like, what I was trying to do
-didn't work with C.
-
-Interviewer: Do you remember what it was you were trying to do?
-
-Rebecca: I don't know exactly, but uh, I just remember, like, whatever I
-was tryin to do, cuz I had my friend help me, he's like ``you can't do
-this this way,'' like, it just, C doesn't recognize whatever I was
-trying to do.
-
-Interviewer: Huh. So that was like, it sounds like it's almost kind of a
-case where your, your plain English of what should happen /yeah/ can't
-actually translate---it's not even that you're not sure how it's---
-
-Rebecca: It's like, it was impossible.
-
-Interviewer: It was never---
-
-Rebecca: It was never allowed.
-
-Interviewer: There is no word for that /yeah/ in this language.
-
-Rebecca: I \{laughs\}, that was back then, so. That was always fun.
-
-(Interview 2 of 5, February 17, 2012)
-
-Epistemologically, two patterns stand out in Rebecca's recollections.
-First, as Rebecca remembers it, one of her core problems was that her
-visions for how her program should accomplish things were ``impossible''
-and ``never allowed'' in C. Second, in order to get past that obstacle,
-she ultimately typed code a friend told her to without feeling like she
-understood it. As this analysis unfolds, I think these two points give
-us a crucial starting point to understand Rebecca's practice. The
-experience underscored a gulf between ideas Rebecca wanted to invoke and
-the syntax she needed to realize them in C. And, her way of dealing with
-that gulf was to make an end-run around it by taking and using code she
-didn't fully understand.\footnote{I say this without any intended
-  judgment (or pejorative connotations) about what Rebecca did. Given
-  the assessment structure of the course, where 90\% of a student's
-  grade came entirely from whether the program worked to spec, she may
-  have felt that getting a good grade by using code she didn't
-  understand was preferable to getting a bad grade with code that didn't
-  work.}
-
-It's worth noting that whether her friend's code worked to
-spec\footnote{Here, I'm using ``worked to spec'' as a shorthand for
-  ``worked in such a way that it passed all of the input/output tests
-  the instructor would run to determine 90\% of a student's grade.''} is
-immaterial to the two epistemological issues I raise. First, if the code
-she copied worked, Rebecca would get a good grade, but she wouldn't
-understand how her project functioned. And, if it didn't work, she would
-get a bad grade \emph{and} she wouldn't understand how her project
-functioned. Ultimately, no matter how the code performed, Rebecca would
-likely be in a tough spot if she ever had to maintain, revisit, or
-refactor the code she copied. Second, no matter how her copied-code
-project functioned, it represented \emph{someone else's} articulation of
-how a procedure should go. That is, the copied code was an artifact of
-someone else bridging the gulf between high-level idea and articulation
-in C. Rebecca still found difficulty traversing that gulf.
-
-\paragraph{Rebecca's array project experience had a lasting effect on
-her confidence in
-programming}\label{rebeccas-array-project-experience-had-a-lasting-effect-on-her-confidence-in-programming}
-
-As that interview drew to a close, I wrapped it up with what had become
-standard protocol: asking whether there was anything else she wanted to
-talk about.
-
-Rebecca: I think I'm good, I---just
-
-Interviewer: Anything else on your mind?
-
-Rebecca: Sorry I'm not the greatest programmer \{laughs\}
-
-Interviewer: No, hey hey hey. That's \{shakes head\}, first of all, I
-mean, um, how come you're apologizing?
-
-Rebecca: I dunno. It's a study, and I'm probably not the best test
-subject. (Interview 2 of 5, February 17, 2012)
-
-Rebecca's spontaneous apology spawned a discussion about why she feels
-diffident in programming. For her, that diffidence traced its way back
-Basic Programming. Feeling lost and stuck on that array project was more
-than just a tough experience; it was a turning point:
-
-Rebecca: When I got lost I think is when I started losing that
-confidence thing. And then I never really got it back, so.
-
-Interviewer: When was it you started to get lost?
-
-Rebecca: Halfway through last semester, like with the arrays in the
-second project. So\ldots{}. I guess it was like, demoralizing, kind of,
-and, like, like I don't want to sound cocky or arrogant or anything, but
-in high school I did very well in academics. And, so, I never really,
-like anything that I had to work for, I worked for for a short period of
-time, got it, and like I'd understand it, and if I didn't, like it was
-really easy for someone usually to explain it to me. And, I felt like I
-didn't really get that after I lost everything on that second project, I
-was like, I wasn't able to build myself back up out of that, which
-probably stems to why, not as confident in the class. (Interview 2 of 5,
-February 17, 2012)
-
-As I interpret how this array project experience may be playing a part
-in Rebecca's later approaches to programming, I stress that my analysis
-is about Rebecca's perceptions and recollections.\footnote{Due to
-  deliberate study limitations I don't have access to the actual code
-  from that project. To be as conservative as possible in our data
-  collection and respect the privacy of participants, we asked
-  participants to grant us access only to folders that would contain
-  their Intermediate Programming projects. Outside code/data, including
-  code from prior semesters, was excluded from the scope of our code
-  snapshot data collection.} It's possible that her ideas for how her
-project should work wouldn't have been implementable in any language,
-let alone in C. And, while I think that possibility is unlikely, I have
-to concede that I only know what Rebecca recalled; I wasn't present for
-the experiences she's describing. Nevertheless, the question for my
-research is not ``can we know what really happened in that experience?''
-Rather, my question is, ``how might Rebecca's \emph{image} of that
-experience --- with its specific epistemological facets --- be playing a
-part in her approach to programming now?''
-
-Part of the answer to that question may be in her quote above. She talks
-about losing her confidence and ``never really {[}getting{]} it back.''
-Later, she says ``I felt like I didn't really get that after I lost
-everything on that second project.'' Because of the peculiarities of
-speech, I had a difficult time parsing Rebecca's statement about losing
-everything on that second project. It wasn't clear to me whether there
-was a pause between ``get that'' and after, but the meaning of ``that,''
-and in turn the meaning of the sentences, depends on the antecedent of
-``that.'' So what she said could have two consequentially different
-meanings:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  ``I feel like I didn't really get \textbf{that}. After I lost
-  everything I lost everything on that second project, I was like, I
-  wasn't able to build myself back up out of that.'' In this version,
-  ``that'' may mean the topic---arrays---or the project as a whole. I
-  might then interpret Rebecca's speech as saying this project was a
-  turning point because of the particular topic. Her being ``unable to
-  build myself back up out of that'' reflects a kind of debt that might
-  have piled up when later assessments depended cumulatively on that
-  early topic. Her failure to grasp arrays was thus cumulatively
-  punishing because arrays formed a core part of many projects that
-  followed.
-\item
-  ``I feel like I didn't really get \textbf{that} after I lost
-  everything on the second project. I was like, I wasn't able to build
-  myself back up out of that.'' In this version, ``that'' may mean her
-  ready insight into topics or her facility in understanding when
-  friends explain them to her. The distinction I draw in this second
-  interpretation is ``that'' refers to a kind of continuity in her
-  identity---someone for whom school always came easy. The array project
-  disturbed that continuity because it was an instance in which things
-  no longer came easy to her. And, the experience may have been so
-  troubling that she was never able to re-establish being able to easily
-  grasp things in programming. The result, then, is that she couldn't
-  build herself---in the sense of her continuous identity as one who
-  quickly grasps school topics in general and programming topics in
-  particular---back up after the array project.
-\end{enumerate}
-
-Distinguishing between these possible meanings of Rebecca's speech might
-seem pedantic. But, I think the bifurcation is important for theorizing
-about learning and instruction in computing. If Rebecca's difficulty was
-about the topic of arrays, a plausible narrative is that the difficulty
-of that concept---and the assessments that continued to depend on
-it---caused a run of bad grades, and the bad grades shook Rebecca's
-confidence. But, if Rebecca's difficulty comes from troubling her sense
-that she has a ready understanding of programming, the attack on her
-confidence is more complex. It's not just that she doesn't get arrays,
-it's that she feels she doesn't readily understand other new code she's
-seeing in the course---even code that doesn't strongly rely on arrays.
-In this view, the confidence Rebecca speaks of isn't just a product of
-getting good grades; it's part of a feedback loop in which grades,
-confidence, and a feeling of ready insight into programming are all
-components.
-
-That potential coupling of grades, confidence, and feelings about ready
-insight into code matters because it further pushes the discussion
-toward epistemological considerations. It is why, for example, I was
-paying such careful attention to Rebecca's recollection of copying down
-code her classmate told her. Instead of helping her confidence in
-programming by potentially boosting her grade, the copying may have
-diminished her confidence because she couldn't understand why someone
-else's code worked.
-
-\paragraph{Epistemological issues continued to interact with Rebecca's
-confidence in Intermediate
-Programming}\label{epistemological-issues-continued-to-interact-with-rebeccas-confidence-in-intermediate-programming}
-
-Rebecca started our fourth interview with a warning to me that she might
-``rant'' about programming.
-
-Interviewer: What would you rant about? Go ahead.
-
-Rebecca: Ohh god. Um, I am not doing well \{laughs\} in the class right
-now.
-
-Interviewer: OK
-
-Rebecca: And it's not good, so. Yeah, like, we just had our last project
-due, which I did a lot better on than the first project
-
-Interviewer: Uh-huh
-
-Rebecca: Which is making me really happy. Um, but, everything about
-programming just upsets me, because there's the two guys, like I had
-mentioned before who, uh, help me
-
-Interviewer: Mmmhmm
-
-Rebecca: Uh, one of them, he---like I worked every day from the day we
-got this project on this project, and my project still didn't turn out
-perfect. He wrote his twice in one day because he lost the whole thing.
-And, his was perfect. And, I dunno---it just irks me *so* bad that that,
-it can be that much of a disparity between people and programming.
-(Interview 4 of 5, April 6, 2012)
-
-In her telling, Rebecca wasn't the only person upset by the relative
-ease with which this student programmed.
-
-Rebecca: we didn't say it to him, but some of us were like, it's, we
-just found it like, so, like aggravating. Like, cuz there's me and
-another girl who, on my floor who, she's uh, never had programming
-experience either---
-
-Interviewer: Uh-huh
-
-Rebecca: And we're just like it gets so aggravating that it comes easy,
-like to him, and that he does it well, but, that's just---he ju---he is
-very good at school. He's very smart. So. (Interview 4 of 5, April 6,
-2012)
-
-Note here that Rebecca is connecting the idea that programming came easy
-to this student to the notion that he was ``very good at school.'' In
-this moment, the statement seems to suggest Rebecca has a \emph{fixed
-mindset} \{Dweck and colleagues\} about programming expertise. While it
-would be naïve to say Rebecca believed --- globally --- that programming
-came easily to those who were ``good at school,'' it's nonetheless true
-she saw a discrepancy between how easy programming was for her and her
-friend and how easy it was for this ``very smart'' classmate.
-
-As we kept talking about this discrepancy, Rebecca offered further
-evidence that struggling with programming troubled her academic
-identity. In high school, she ``didn't have to study a lot'' because ``a
-lot of things came relatively easy'' (Interview 4 of 5, April 6, 2012).
-But, her experiences in Intermediate Programming pushed her into a new
-position. As she said,
-
-now I'm on the other end and I'm realizing how aggravating it
-is\ldots{}. And, I don'---I guess it's just a shock, cuz I do not like
-not being good at things \{laughs\}. So...'' (Interview 4 of 5, April 6,
-2012).
-
-Again, Rebecca's comments stress the frustration of feeling like
-programming comes easily to some people but not to her.\footnote{Or, at
-  least, at that moment, in that semester, it did not seem to be coming
-  easily to her.} And, they are charged with emotion. She speaks of
-being ``on the other end,'' connoting an idea of a structure whose poles
-have people who ``get'' programming on one end and those who don't on
-the other. It's ``aggravating'' to be on that end, and a ``shock'' to be
-there for the first time.
-
-I asked Rebecca to reflect on whether she could ever remember seeing
-classmates in the position she was now, where something seemed to come
-easily to Rebecca but not to them.
-
-Rebecca: Um. I mean, there was a---there's a friend of mine, who, from
-my high school, who, I, I mean just always got math. And, uh, she was
-struggling in math a lot.
-
-Interviewer: Mmmhmm
-
-Rebecca: But her attitude was, uh, ``I don't need math. I'm gonna be an
-English major.''
-
-Interviewer: OK
-
-Rebecca: So, she didn't even care about it
-
-Interviewer: Oh, OK
-
-Rebecca: Whereas, like, I was like, I'm an electrical engineer major, I
-actually need this. So. Like that's the only time I think I've ever felt
-like someone's had that feeling that I probably did.
-
-Interviewer: Right.
-
-Rebecca: And, I mean I'm sure there are people that I---uh, that I
-didn't know, or didn't realize they had that feeling. But, it's just,
-interesting to be on the other end. \{6 second pause\}
-
-Rebecca: Umm. Just. Ugh. I've like, debated withdrawing from the class.
-And, like, I've talked to three different advisors or whatever, and
-they're like ``don't do it. It's not---you can still bring your---you
-can bring your grade up. And even if you don't you have freshman
-forgiveness,\footnote{In the interview, Rebecca explained that Freshman
-  Forgiveness was a policy extended to students during their first 24
-  credit-hours at the university. If a student gets a grade she's
-  unhappy with, she can retake that course and her second final grade
-  will replace the first.} so.''
-
-Rebecca, unlike her friend, couldn't afford \emph{not} to care about
-Intermediate Programming; it was a required course for her electrical
-engineering major. But, the combination of feelings she had---consisting
-partly of a feeling that programming did not come easily to her---pushed
-her to think about withdrawing from the course anyway. And, Rebecca was
-not in the habit of withdrawing from classes. Intermediate Programming
-was the first and only class where she was experiencing this level of
-difficulty (Interview 4 of 5, April 6, 2012).
-
-It would be tempting to think Rebecca struggled in Intermediate
-Programming because she didn't work hard enough. But, by her own
-assertions she \emph{did} put a considerable amount of effort on
-projects. On Project 2, for instance, she said ``I worked every day from
-the day we got this project on this project, and my project still didn't
-turn out perfect'' (Interview 4 of 5, April 6, 2012). Figure 5 below
-shows Rebecca's activity over time from March 5, the day Project 2 was
-handed out, to March 28, the day it was due:
-
-\includegraphics[width=5.73333in,height=3.98889in]{media/image10.emf}
-
-\protect\hypertarget{ux5fToc252445971}{}{}Figure ‑ -- Rebecca's
-compilation activity over time for Project 2. Each compile a student
-initiates creates a commit, so long as there has been a change to the
-underlying code. The height of each bar maps to the number of commits
-recorded that day. The red line charts \emph{cumulative} commits over
-time. The red line is steepest in periods of frequent activity and
-shallowest in periods with little or no compile activity.
-
-From the data we can tell Rebecca was compiling Project 2 code almost
-every day between March 19 and March 26. Moreover, roughly two thirds of
-compilations for the entire project happened in the last 3 days leading
-up to the deadline. So, at the very least our data shows she was
-actively compiling her code for 8 out the last 10 days.
-
-Rebecca registered no commits between March 12 and March 19. There were
-also other days for which no commits were recorded. One possible cause
-for those zero values is our data collection system was faulty. A second
-possible explanation is that Rebecca was not working. But, a third
-explanation is that Rebecca was working but not compiling. This third
-explanation actually seems most plausible. It fits the pattern of
-Rebecca's work on project 1, where she wrote dozens of lines of code
-before ever compiling (Interview 2 of 5, February 17, 2012). It also
-accounts for Rebecca's lack of compilations despite her claim that she
-worked on Project 2 every day.
-
-So, we can be fairly certain Rebecca was spending time on
-projects.\footnote{The specific nature of \emph{what} Rebecca's work
-  looked like on Project 2 is the subject of the first study of this
-  dissertation, so I don't expand upon it substantially in this study.}
-Another possibility for why Rebecca was struggling might be that she
-wasn't studying. To address that conjecture, we have to step back to
-when I asked Rebecca whether other classes --- classes other than
-Intermediate Programming --- had ever pushed her close to withdrawing
-from them. She said no.
-
-Rebecca: the thing is, I *don't* know how to like, study for
-programming. Cuz like, I look at code and stuff, and I try, and I work
-every day on it, but like, even when---I dunno, it just, does not come
-naturally.
-
-Interviewer: So, what is it like when you study it? What do you usually
-try to do?
-
-Rebecca: Uhh, a lot of times, like, I'll read through the notes we took
-in class, like, cuz, what we're doing now---we have a project, uh, we're
-doing linked lists, and like, malloc'ing data, which is, uh, *you*
-storing it in, as you go. And, like, I get the theory and everything
-behind it, and, like it makes sense what it does, but, I just like try
-and look at his code examples and stuff, and try and replicate it, but I
-don't know what's happening right now, but it's not doing it correctly.
-I just, like, my *biggest* problem, I realized, in programming, is I
-think I have the logic, my logic just doesn't transfer to code. Or I
-don't know *how* to transfer it correctly.
-
-Interviewer: So, um. when you're spending time studying, um, do you
-spend most of it on your computer trying to actually program or---
-
-Rebecca: A lot of it is on my computer and a lot of it is also trying to
-just look at his code and notes in class. Uh, that's probably about
-split, 50/50. Uh, because, sometimes I'll just end up staring at my
-computer screen, like, ``what do I even type in to try?'' So, I try and
-go look at his examples, so.
-
-(Interview 4 of 5, April 6, 2012)
-
-There are two striking features of this exchange.\footnote{There are
-  actually three striking features. The one I left out above is that
-  Rebecca describes malloc'ing as \emph{you} storing in data. I think a
-  great deal of Rebecca's difficulty with malloc specifically hinges on
-  what she means by ``you'' storing it in.} The first is that Rebecca
-felt she didn't know how to study for programming. The second striking
-feature of this exchange is how Rebecca elaborated on what it might mean
-for programming to come naturally to her.
-
-Rebecca feels programming doesn't come naturally to her because she
-routinely finds herself unable to transfer her ``logic'' to code. As a
-follow-up to Rebecca's answer, I asked what ``logic'' meant to her in
-this case.
-
-Rebecca: Like, like how to get the---like, if I need to get a certain
-data, like I have a process of how I want to get it, I ju---and I'm
-like, ``Oh, you would just do this,'' but can the *computer* do that?
-Can I---and it's my---me being able to tell the computer to do that is
-where I get lost. (Interview 4 of 5, April 6, 2012)
-
-The tension Rebecca is describing is one I see as strongly
-epistemological. First, it's epistemological in the sense that it
-reflects a struggle to articulate knowledge of \emph{how to do
-something}. Papert (1980) and Abelson \& Sussman (1996) would describe
-such a struggle as a pertaining to \emph{procedural epistemology}: the
-struggle concerns a particular kind of knowledge work in which the goal
-is not to describe what is, but rather how to. The second sense in which
-it's an epistemological struggle is closer to how Hammer, Elby, and
-colleagues use the word epistemological: what does students' activity
-reveal about how they orient toward knowledge and knowing in a
-discipline (Hammer et al., 2005; Hammer \& Elby, 2002, 2003; Scherr \&
-Hammer, 2009). In this second sense, we can observe that Rebecca defines
-her difficulty in terms of the creation of proper code. We can also pose
-the question ``if Rebecca struggles to transfer logic to code, what
-directs her activity as she \emph{tries} to transfer logic to code?''
-
-Later in the interview, she compared the iTunes project --- on which she
-was stuck --- to Project 2.\footnote{Project 2 did not require students
-  to use malloc() to dynamically allocate memory. Project 3 did.} In so
-doing, she offered another telling idea about the relationship between
-her logic and ``syntax'':
-
-Rebecca: Like, cuz, I guess I feel like I did a lot better on the last
-project because it was so similar to stuff that I'm used to /Mmmhmm/ um,
-whereas all this is brand new and I still don't have any of the syntax
-down yet. /Mmmhmm/ Like, if-statements and while-loops, they make sense
-to me, whereas I'm still like, trying to grasp at this stuff.
-
-By ``this stuff,'' the context of the conversation suggests Rebecca was
-referring to dynamic memory allocation and malloc, which we had been
-discussing.
-
-It seems sensible to think if-statements and while-loops might ``make
-sense'' to Rebecca because their syntactical form aligns strongly with
-their computational operation.\footnote{To borrow terminology from
-  Sherin (2001) and rebut Saussure (1986) at the same time, the symbol
-  template for an if-statement is strongly connected to its conceptual
-  schema because we so often use everyday language in a manner
-  concordant with an if-statement's branched control flow. In this
-  instance, the relationship between signifier and signified is the
-  \emph{opposite} of arbitrary. The similarity between the syntax for an
-  if-statement and an understanding of what that statement does is by
-  design, because C itself was a designed language.} By that, I mean in
-principle the designers of C could have chosen \emph{any} symbol to
-represent branched control flow, but they chose a word --- if --- that
-invites connections to its meaning in everyday speech. If-statements in
-C split control flow based on a Boolean expression's value; while-loops
-allow continued iteration until a Boolean expression turns up false.
-Both structures have everyday analogues with similar conceptual
-properties. A parent might say, ``if you're staying out past 9, let me
-know,'' or, ``while we're gone, can you make sure Sadie doesn't eat
-anything from the garbage?'' If a child will be home by 8, there is no
-need to let that parent know. Similarly, it's reasonable to expect that
-once the parents return, the responsibility of keeping Sadie out of the
-garbage no longer falls solely to the child.\footnote{Some parents might
-  object that family members should \emph{always} be on the lookout for
-  Sadie rooting through the trash, that these parents were just
-  emphasizing the point. And, reasonable children might in general try
-  to heed that. But, what makes computers computers is their lack of
-  qualms about abandoning any and all responsibility to keep Sadie from
-  the trash once the parents return.} The point of this observation is
-that Rebecca may feel certain kinds of programming structures ``make
-sense'' not simply because they were covered in introductory
-programming, but because something about their syntactical form offers
-an affordance for thinking about how they work.
-
-\subsubsection{Rebecca has productive capacities for making progress on
-software design
-work}\label{rebecca-has-productive-capacities-for-making-progress-on-software-design-work}
-
-In our fourth of five interviews, Rebecca was having trouble making
-progress on Project 3, which the instructor described as an ``online
-music server.''\footnote{Students in my study colloquially referred to
-  this project as the ``iTunes project,'' so ``the iTunes project'' and
-  ``Project 3'' all refer to this project.} ``Like, I am so lost, like I
-don't even know where to start,'' she said. We discussed why she was
-stuck. Then, with my help, we began to work through what Rebecca
-identified as a core cause of her confusion. In brief, the assignment
-contained a diagram showing the data scheme students were required to
-use to store data about the music (Figure 8 below). Specifically, the
-scheme required an array of pointers --- each entry of which pointed to
-an array of characters --- as a mechanism for storing data about song
-titles for the music server. Rebecca's work to resolve her confusion
-offers evidence for her productive capacities in designing programs. In
-this section, I argue:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Rebecca made sense of the array-of-pointers design\footnote{By
-    ``design scheme,'' here, I mean the instructor-provided diagram
-    indicating how students were to structure their data for Project 3.}
-  using talk and gesture. She was able to explain both why an array of
-  pointers \emph{should} exist in C as well as suggest candidate syntax
-  for what might create such an array.
-\item
-  When I proposed we pretend her candidate syntax worked, Rebecca
-  articulated through gesture, talk, and written pseudo-code how part of
-  her design would incorporate that syntax.
-\item
-  After the interview, Rebecca resumed work on the project and
-  incorporated a version of the design she developed in the interview
-  into her code.
-\end{enumerate}
-
-\paragraph{Rebecca made sense of the array-of-pointers design using talk
-and
-gesture}\label{rebecca-made-sense-of-the-array-of-pointers-design-using-talk-and-gesture}
-
-The figure below shows the instructor-provided diagram for how students
-should structure their data.
-
-\includegraphics[width=5.36667in,height=4.10000in]{media/image11.emf}
-
-\protect\hypertarget{ux5fToc252445972}{}{}Figure ‑ -- The instructor
-required students to use this data arrangement for storing information
-in the music server. The scheme uses an array of pointers to represent
-the track names of an album.
-
-Rebecca would have seen this diagram before because the project
-assignment was given out 11 days prior to our interview. But, the data I
-present is the first time I had any access to how she was thinking about
-it.
-
-At the time of the interview Rebecca felt stuck on handling issues with
-\emph{music albums} --- the top-left and top-right structures depicted
-on the diagram. The assignment required that ``all of the data
-structures described {[}in this diagram{]}'', including music albums and
-all of the data they referenced, ``must be allocated dynamically''
-(Instructor-provided Project 3 description). So, Rebecca and I started
-discussing what parts of that scheme she understood. Table 3 below
-presents what she said and what she did while beginning to explain her
-understanding.\footnote{}
-
-\protect\hypertarget{ux5fToc252445961}{}{}Table - Rebecca describes an
-array of structures using gestures (Interview 4 of 5, April 6, 2012)
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-What she says & What she does\tabularnewline
-\midrule
-\endhead
-Like, the---the thing that makes sense to me that I, uh, got right now,
-is that, to malloc the array, to \{bounds a wide space with her hands\}
-make the array a certain length, do that. &
-\includegraphics[width=2.03333in,height=2.72222in]{media/image12.png}\tabularnewline
-And then I know that each of the \{chops out three invisible boxes in
-the air\} I guess nu---characters---I---uh, I'll call `em characters. &
-\includegraphics[width=1.37778in,height=3.66667in]{media/image13.emf}\tabularnewline
-Like, each of the spots in the array holds a structure and each of the
-structures holds three things. & (gestures obscured by the computer
-)\tabularnewline
-\bottomrule
-\end{longtable}
-
-Rebecca's gestures create a kind of virtual object that looks much like
-the music albums array in the diagram. As she chops out spaces for each
-of the ``characters,'' those spaces are in series and fit --- more or
-less --- within the larger area she bounded with her when she said ``to
-malloc the array.'' The spatial subdivision is key because it offers
-supporting evidence that in some way, Rebecca might have been thinking
-about the album structures themselves as being \emph{contained} in or by
-the array. And, that spatial containment metaphor resurfaced later when
-we discussed the array of pointers, which I'll get to in a moment.
-
-As Rebecca explained her understanding to me, I took to writing down
-what I understood her to be saying on paper:
-
-Interviewer: OK. So you have, so here's what you just said. You said
-\{writes\} ``malloc the array'' um, ``each element has a structure,''
-and then ``each structure has'' you said---
-
-Rebecca: Has three parts
-
-Interviewer: OK. Has three parts. OK.
-
-Rebecca: Yeah. So, and the first part I can do `cuz it's an integer
-\{laughs\}. The first part is just the number of tracks on the album.
-So---
-
-Interviewer: Oh, oh. So, of the three parts {[}{[}I'm sorry
-
-Rebecca: Yes{]}{]}
-
-Interviewer: I {[}{[}\{untillegible\}
-
-Rebecca: Oh yeah, sorry{]}{]}
-
-Interviewer: So you're saying, um, and then---and then this is number of
-tracks, which is an integer, right? /yeah/ OK (Interview 4 of 5, April
-6, 2012)
-
-Figure 9 below shows what I wrote as we worked:
-
-\includegraphics[width=5.62222in,height=1.48889in]{media/image14.emf}
-
-\protect\hypertarget{ux5fToc252445973}{}{}Figure ‑ -- I write down what
-I understand as Rebecca explains the overall data structure of albums to
-me
-
-It wasn't the first time --- either in my series of interviews with
-Rebecca or in that interview in particular --- that I tried to write
-down what she said. But, I highlight it here because it became part of
-my activity as we worked together in this episode. Rebecca would
-articulate something, often through talk and gesture, and I would try to
-write down and offer back to her what I understood her to be explaining.
-
-As we continued, we landed on trying to understand the ``tracks''
-portion of the album data structure. Storing tracks was where Rebecca
-felt she was getting ``lost'':
-
-\protect\hypertarget{ux5fToc252445962}{}{}Table -- Rebecca says gets
-lost on the array of pointers pointing to track names (Interview 4 of 5,
-April 6, 2012)
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-What she says & What she does\tabularnewline
-\midrule
-\endhead
-Rebecca: And, the next one is where I get lost, on the tracks part.
-Because it's a pointer \{2 sec pause\} to \{hands make a vertical
-cylinder in front of her; note that the array is represented on the
-assignment as a narrow, vertical rectangle\}--- &
-\includegraphics[width=3.40000in,height=2.62222in]{media/image15.emf}\tabularnewline
-it looks like another array that's
-\textbar{}\textbar{}pointing\textbar{}\textbar{} \textbar{}\{left hand
-crosses from her left to right, at chest height, index finger extended
-in the direction of motion; this is the same direction (from Rebecca's
-perspective) in which the pointer array on paper points to each track
-name\}\textbar{} to the names. &
-\includegraphics[width=1.98889in,height=3.38889in]{media/image16.emf}\tabularnewline
-\bottomrule
-\end{longtable}
-
-Rebecca's gestures for the array of pointers again spatially mimic the
-depiction of the array on the page: a vertical cylinder. The reason I
-note the orientation of her array gestures is that for practical
-purposes arrays as represented by bits in C have no orientation in space
-that the programmer could know, much less care about. An array cannot be
-parallel to the ground, nor can it be oriented toward the
-ceiling.\footnote{The computational hardware representing the array does
-  have an orientation in space, but that's clearly not what is at issue
-  here. The whole point of the array is to be an abstraction away from
-  the soldered transistors, capacitive plates, flipped switches, vacuum
-  tubes, tinker toys, or other physical means that store the state of
-  the array.} But, the assignment \emph{depicted} the album array as
-horizontal, and so too did Rebecca's gestures for it. The assignment
-\emph{depicted} the pointer array as vertical, and so too did Rebecca's
-gestures for it. In other words, the observation that Rebecca's gestures
-align with the assignments pictures might suggest that she is doing more
-than thinking about arrays in the abstract. Rather, some elements of the
-representational features of the page have made their way into her
-activity and, arguably, her cognition.\footnote{One could argue that
-  gesture is possibly just undirected flailing, and that the correlation
-  between Rebecca's gestures and the page is weak at best --- the sort
-  of thing you'd expect to happen some percentage of the time anyway
-  under the assumption that the pictures on the page have nothing to do
-  with how she's thinking about it. I concede that it's \emph{possible}
-  the correlation isn't structurally meaningful. But, I think the
-  correlation is meaningful given how many other times her gestures
-  align with canonical written representations of computational
-  structures and processes.}
-
-Rebecca continued on to try thinking through why an array of pointers
-was part of the design.
-
-Rebecca: But, I've always been confused as to why you---I guess, I
-always---I was always like, when we learned pointers, I was like ``why
-do you need pointers when you could just call it the name? Why do you
-need two names for it?'' But, I think, what---at least what I'm seeing
-here maybe is like, this is just another array /Mmmhmm/ because you
-can't put more than one character in each element of the array, so that
-is just an array of pointers that point to strings. And the strings are
-the names.
-
-Rebecca: So, that would make sense. So you would---the tracks would
-point to the array, and so you could do the pointer of that \{1.5 sec
-pause\} The pointer to the array, an array---the array of like 1 would
-be a pointer to track 1's name.
-
-Interviewer: Mmmhmm
-
-Rebecca: So. OK, so---see that part makes sense now. I just---no idea
-how I would ever access that. Or, store it, I guess, is the better word.
-(Interview 4 of 5, April 6, 2012)
-
-What ``confused'' Rebecca has an understandable origin. Most of the work
-she would have been doing until that point would have used named
-variables. For example:
-
-int age;
-
-float weight;
-
-char *name;
-
-If age, weight, and name are already-named pieces of data, why would
-something ever need to refer to them indirectly? It would be as if we
-had to call me ``The second-born son of the second-born daughter of
-Henry'' for some reason. I already have a name. Why do I need another
-name that uses the names and relationships of my forebears to refer to
-me when historically my given name has perfectly fine for all
-referential purposes?
-
-Rebecca's way of finding sense in pointers was to observe that you can't
-put names --- each of which is defined as an array of characters ---
-into an array because ``you can't put more than one character in an
-array.'' To verify I was understanding her idea correctly, I tried to
-restate it:
-
-Interviewer: OK, so you were saying the reason you---when you're tryin
-to convince yourself like---you originally you were thinking ``why don't
-you just put all the names {[}{[}here
-
-Rebecca: Yeah{]}{]}
-
-Interviewer: And you were {[}{[}saying the reason is
-
-Rebecca: saying, no{]}{]}
-
-Interviewer: because==
-
-Rebecca: ==That makes sense now. Yeah, you can't put more than one
-character in an element. And, it's a whole name of a song, so you would
-need to point---that element would just \{gesture obscured by my
-laptop\} be a pointer to a name. OK. (Interview 4 of 5, April 6, 2012)
-
-The latching and overlapping turns of talk above offer strong evidence
-that Rebecca's explanation aligned with my interpretation. And again,
-notions of \emph{containment} come into play. Rebecca says, ``you can't
-\emph{put} more than one character \emph{in} an element'' in part
-because you're talking about ``a \emph{whole name} of a song'' (my
-emphasis added here). If Rebecca has a capacity for thinking about the
-canonical concept of type-restrictions on arrays, it's manifesting here
-as a part-physical metaphor, evoked through both talk (the quote above)
-and gesture (Table 4), in the service of making sense of a design
-decision.
-
-As she continued, Rebecca discerned that because the input data
-contained the number of tracks on each album, she could use that number
-to define how long the pointer array needed to be. But, something was
-still troubling her.
-
-Rebecca: So. \{4 sec pause\} That would \{6 sec pause\}
-
-Interviewer: What're you thinkin?
-
-Rebecca: Just that, um, I'm just tryna think, cuz like that makes sense,
-but I don't know how to get to that. Cuz, how do you make a pointer
-point to a random array that you just made? Uh, I guess \{4 sec pause\}
-
-Rebecca: Uh, p---the pointer syntax I need---I would need to go back and
-look at them because now you have an array, each array---my issue---cuz
-like now that makes sense. Like, it's just how do I get to that point of
-making that work?
-
-Interviewer: {[}{[}So
-
-Rebecca: And this is what{]}{]} I mean by my logic /OK/ I get the logic
-and the theory behind it but I don't know how to actually put it into C.
-
-Interviewer: OK, so in other words, what makes
-\textbar{}\textbar{}sense\textbar{}\textbar{} \textbar{}\{air
-quotes\}\textbar{} to you know /Mmmhmm/ that I guess didn't before is
-that *that* \{gestures to the instructor's structural diagram on
-screen\} is a---that's an understandable arrangement /yes/ for the data
-
-Rebecca: Yes.
-
-Interviewer: But then, how do I {[}{[}make it
-
-Rebecca: make it{]}{]} Yeah \{nods\}. (Interview 4 of 5, April 6, 2012)
-
-When Rebecca said ``I don't know how to get to that,'' I take ``how to
-get to that'' to mean \emph{how to write the code for that procedure}.
-We had just agreed that design-wise, she could use a number provided in
-the input data to define the length of the pointer of array. But, she
-was stuck because of her difficulty translating the ``logic'' we
-developed into code.
-
-From the outside, what Rebecca and I had done together was something I
-think crucially important to the design of a program. We worked to
-understand how someone else chose to structure their data.\footnote{That
-  the ``someone else'' was in this case the instructor is important for
-  larger power dynamic considerations.} We tried to decide for ourselves
-whether and how those structures made sense. Along the way we created
-ephemeral objects of that sense-making in the form of talk and gesture
-as well as the durable artifacts of my inscriptions. But, none of that
-work resulted in C code.
-
-For Rebecca, the lack of certainty about \emph{how to make} our ideas in
-C code was a marked concern. And, again, that concern makes sense. We
-had just convinced ourselves that we could understand \emph{why} an
-array-of-pointers. We even figured out \emph{how} we could use the input
-data to provide the length of the arrays we would need to make. But we
-had not laid out how to declare the array \emph{and} preserve a
-reference to it (``how do you make a pointer point to a random array
-that you just made?). Without that piece --- a piece Rebecca was unsure
-of and felt she would need to look up --- our solution clearly wouldn't
-work.
-
-Absent that piece, Rebecca tried thinking of possible ways to fill in
-the missing syntax.
-
-Rebecca: Cuz, I'm like, like I'm trying to think right now of things
-that I would try /Mmmhmm/ but I don't know what I would try first
-because, all I'm thinking right now is that I know tracks is going to be
-a pointer. /Mmmhmm/ So you make that a pointer.
-
-Interviewer: OK.
-
-Rebecca: But I'm not sure where it would point to, because you don't
-have that array made yet. So I \{3 sec pause\} so you'd have to, I guess
-you could make the array, and then make it point to the array, but \{3
-sec pause\} Cuz you have to make an array of pointers.
-
-Interviewer: Mmmhmm
-
-Rebecca: And I don't know how to do that. Because I've made arrays of
-characters and integers and stuff before /OK/ but never pointers, so.
-(Interview 4 of 5, April 6, 2012)
-
-Rebecca was seeing a temporal problem: how could she reference an array
-that didn't exist yet?\footnote{As a crude analogy, imagine trying to
-  write your own will. If you think forward in time, you might later
-  have children to whom you'd like to bequeath your assets. But how can
-  you enumerate those children in the will now, before you've had them?
-  You need your will to refer to something that does not exist at the
-  time you write the will. Extrapolating forward and rather ludicrously,
-  we could imagine you are always capable of producing children. So, up
-  until the day of your death, there is always the seeming risk that
-  your will might need to refer to a beneficiary who does not yet exist.
-  Recognizing this subtle problem as Rebecca did is an act I consider to
-  be sense-making, even though she as-yet still did not have a solution.}
-
-Rebecca started exploring ways to solve that problem of getting tracks
-to point to the array of pointers.
-
-Rebecca: But I'm not sure where it (tracks) would point to, because you
-don't have that array made yet. So I \{3 sec pause\} so you'd have to, I
-guess you could make the array, and then make it point to the array, but
-\{3 sec pause\} Cuz you have to make an array of pointers.
-
-Interviewer: Mmmhmm
-
-Rebecca: And I don't know how to do that. Because I've made arrays of
-characters and integers and stuff before /OK/ but never pointers, so.
-(Interview 4 of 5, April 6, 2012)
-
-Rebecca landed on a potential solution: make the array first, then make
-tracks point to the array. But, with that solution came another problem:
-how do you make an array of pointers?
-
-\paragraph{Rebecca could propose candidate syntax and build pseudocode
-around
-it.}\label{rebecca-could-propose-candidate-syntax-and-build-pseudocode-around-it.}
-
-Rebecca came up with an idea to create an array of pointers, but she
-wasn't sure if it would work.
-
-Rebecca: if I were to d---when I declared the array, if I were to---I
-dunno if you even can, like, whenever we---we do ints, and then it would
-be like star-p /Mmmhmm/ and then star-p would be the pointer /ok/ I
-don't know if you can do like int star-p
-\textbar{}\textbar{}bracket-bracket\textbar{}\textbar{}
-\textbar{}\{makes square brackets with hands\}\textbar{} and that would
-be an array or not.
-
-Interviewer: {[}{[}OK
-
-Rebecca: But I don't know{]}{]} if you can do that.
-
-Notice what's happening here. Rebecca is taking patterns that she knows
-work:
-
-int age; // creates an integer variable called age
-
-int ages{[}35{]}; // creates an array called ages,
-
-// 35 integers long
-
-int *p; // creates a pointer-to-an-integer;
-
-// said pointer is called p
-
-and considering a pattern that might work:
-
-int *p{[}35{]}; // Rebecca is not exactly sure what this will do
-
-Notice further that in this exchange, Rebecca wondered ``if you can do
-that.'' Determining the referent of \emph{that} was, again,
-consequential. One possibility was Rebecca meant ``int *p{[}35{]}''
-might not be the proper syntax for creating an array of 35
-pointers-to-ints. If so, Rebecca was struggling with her aforementioned
-problem of not being able to translate her logic to code. Another
-possibility was Rebecca meant ``I don't know if there exists \emph{any}
-syntax in C to create an array of pointers.'' So, I followed up on this
-latter interpretation. We both agreed it \emph{should} be possible to
-create an array of pointers, it was indeed just a question of whether
-that syntax accomplished it (Interview 4 of 5, April 6, 2012). So, that
-left the former interpretation: it looked like Rebecca was unsure how to
-``transfer her logic to code.''
-
-We both agreed creating an array of pointers should be possible. So, I
-asked Rebecca to suppose her candidate syntax worked. She readily
-agreed.
-
-Interviewer: OK. So then, um, what if we just pretended for a minute,
-that that, like==
-
-Rebecca: ==That that works==
-
-Interviewer: ==That that worked==
-
-Rebecca: ==OK==
-
-Interviewer: ==OK. So then, um. \{begins writing\} So then you might
-write like to make an array of \{stops writing\} actually what would we
-call this? This is==
-
-Rebecca: ==array of pointers, I guess==
-
-Interviewer: ==OK. \{under breath\}
-\textbar{}\textbar{}pointers\textbar{}\textbar{} \textbar{}\{writes
-``pointers''\}\textbar{}. It'd be, well. So. The---the one we had was
-like \textbar{}\textbar{}int, star p, um, it would be some
-number\textbar{}\textbar{} \textbar{}\{writing\}\textbar{}, um, and I
-guess that'd be it in order to just declare /Yes/ that right? OK.
-
-In this exchange note every turn boundary is latched. Once I bid for
-pretending, Rebecca accedes. When I'm unsure what to call the array,
-Rebecca finishes my sentence. Rebecca rapidly approves the final product
-before I even finish my last sentence. Figure 10 below is the syntax I
-wrote in the exchange.
-
-\includegraphics[width=5.12500in,height=1.09792in]{media/image17.emf}
-
-If that syntax created the array-of-pointers, the next challenge for
-storing track names was how to access elements in array-of-pointers.
-Figure 11 presents the pseudo-code Rebecca ultimately wrote to scan in
-and store track names.
-
-\includegraphics[width=5.36667in,height=1.30000in]{media/image18.emf}
-
-\protect\hypertarget{ux5fToc252445974}{}{}Figure ‑ -- Rebecca's
-pseudo-code for scanning in and storing track names
-
-The next section will show the rich talk and gestures that accompanied
-Rebecca's generation of this pseudo-code. I argue activity in the form
-of gesture and talk constituted and sustained Rebecca's in-the-moment
-approach to programming.
-
-\paragraph{Rebecca's talk and gestures constituted and sustained her
-in-the-moment approach to
-programming}\label{rebeccas-talk-and-gestures-constituted-and-sustained-her-in-the-moment-approach-to-programming}
-
-Throughout this section I detail the speech and gestures that accompany
-Rebecca's pseudocode production. Along both modalities I saw evidence of
-a marked shift in Rebecca's orientation. It was a shift from being
-hunched, silent, thinking over the keyboard to being animated, open, and
-gestural toward me. It was also a shift from seeming diffident,
-uncertain, and hedging in speech to seeming confident, assured, and able
-to dispatch my questions. Taken in their totality across the episode,
-these subtleties constitute evidence of an approach that strongly
-differs from those Rebecca had when discussing her struggles in the
-course (section 3.5.1).
-
-First, Rebecca suggested she could write the pseudo-code as we talked.
-
-\protect\hypertarget{ux5fToc252445963}{}{}Table -- Rebecca starts
-talking out pseudo-code and asks for the pen to write it
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-What she said & What she did\tabularnewline
-\midrule
-\endhead
-Uh, well, what I was think---like if I wanted to like,
-\textbar{}\textbar{}save it\textbar{}\textbar{} \textbar{}\{pinches
-right hand, thumb and index finger a few inches apart\}\textbar{} or
-whatever, /Yeah/ &
-\includegraphics[width=2.68889in,height=3.23333in]{media/image19.png}\tabularnewline
-I could make a while loop, \textbar{}\textbar{}scan in
-the\textbar{}\textbar{} \textbar{}\{drags hand across the
-table\}\textbar{} ---scan in the data from the album, /OK/ \{interviewer
-begins writing\} uh, which is, {[}{[}I can write it if you want &
-\includegraphics[width=2.66667in,height=3.23333in]{media/image20.png}\tabularnewline
-\bottomrule
-\end{longtable}
-
-Rebecca then took the pen and started describing her thinking while she
-wrote.
-
-\protect\hypertarget{ux5fToc252445964}{}{}Table -- Rebecca's
-verbal/gestural overview of looping through the input track titles
-(Interview 4 of 5, April 6, 2012)
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule
-Panel \# & What she said & What she did\tabularnewline
-\midrule
-\endhead
-1 & And then my thinking at least, is you should be able to, um, say
-that
-
-``star p of i'' /mmhmm/ equals, uh, the title, and then you just do i++,
-so then it'll \textbar{}\textbar{}move to the next
-one\textbar{}\textbar{} \textbar{}\{makes looping gesture with left
-hand\}\textbar{} /OK/ &
-\includegraphics[width=2.03333in,height=2.33333in]{media/image21.png}\tabularnewline
-2 & and you just keep \textbar{}\textbar{}saving each of the
-pointers\textbar{}\textbar{} \textbar{}\{left hand makes horizontal
-chops in the air, like rungs down a ladder\}\textbar{} &
-\includegraphics[width=3.40000in,height=1.92222in]{media/image22.emf}\tabularnewline
-3 & in the array \textbar{}\textbar{}to a title\textbar{}\textbar{}
-\textbar{}\{right hand makes pinching motion, left-hand points to
-it\}\textbar{} &
-\includegraphics[width=2.18889in,height=2.24444in]{media/image23.png}\tabularnewline
-4 & \textbar{}\textbar{}And\textbar{}\textbar{} \textbar{}\{right hand
-makes a large loop counterclockwise (from Rebecca's
-perspective)\}\textbar{} &
-\includegraphics[width=3.13333in,height=1.54444in]{media/image24.emf}\tabularnewline
-5 & \textbar{}\textbar{}you just increment by 1\textbar{}\textbar{}
-\textbar{}\{right hand makes a cycloid to her right, in a plane parallel
-to the back wall, like counting dots on a line\}\textbar{} until you
-reach the end of file. &
-\includegraphics[width=1.06667in,height=3.62222in]{media/image25.emf}\tabularnewline
-\bottomrule
-\end{longtable}
-
-``Like, that would make sense to me,'' Rebecca said. I asked whether the
-while loop would get a fresh line of text when it runs again.
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule
-Panel \# & What she said & What she did\tabularnewline
-\midrule
-\endhead
-1 & Yeah, because{]}{]} uh, \textbar{}\textbar{}what the while loop
-does\textbar{}\textbar{} \textbar{}\{pinches right thumb and
-forefinger\}\textbar{} &
-\includegraphics[width=1.74444in,height=1.95556in]{media/image26.png}\tabularnewline
-2 & \textbar{}\textbar{}is it reads in the line\textbar{}\textbar{}
-\textbar{}\{right hand, palm down, chest height, slides out to the
-right\}\textbar{} &
-\includegraphics[width=1.76667in,height=2.83333in]{media/image27.emf}\tabularnewline
-3 & and then \textbar{}\textbar{}once it reaches\textbar{}\textbar{}
-\textbar{}\{right hand rises, lowers in a chop\}\textbar{} &
-\includegraphics[width=1.78889in,height=2.11111in]{media/image28.png}\tabularnewline
-4 & the end of \textbar{}\textbar{}character it'll\textbar{}\textbar{}
-\textbar{}\{right hand rises, swoops down in a crescent, index finger
-extended, looping back up\}\textbar{} &
-\includegraphics[width=2.74444in,height=1.36667in]{media/image29.emf}\tabularnewline
-5 & go back down---it'll \textbar{}\textbar{}do the while
-loop\textbar{}\textbar{} \textbar{}\{right hand moves back down toward
-table\}\textbar{} \textbar{}\textbar{}and then\textbar{}\textbar{}
-\textbar{}\{right hand moves back up above shoulder-height,
-palm-down\}\textbar{} &
-\includegraphics[width=2.63333in,height=1.55556in]{media/image30.emf}\tabularnewline
-6 & \textbar{}\textbar{}go back\textbar{}\textbar{} \textbar{}\{right
-hand creates a loop down and back up\}\textbar{} down to the next
-thing---line, &
-\includegraphics[width=3.38889in,height=1.31111in]{media/image31.emf}\tabularnewline
-7 & it'll \textbar{}\textbar{}read in the line\textbar{}\textbar{}
-\textbar{}\{right hand scans rightward at shoulder height\}\textbar{} &
-\includegraphics[width=1.61111in,height=2.47778in]{media/image32.emf}\tabularnewline
-8 & \textbar{}\textbar{}until it\textbar{}\textbar{} \textbar{}\{pinches
-right thumb and forefinger, swipes right hand back to the left, palm
-down\}\textbar{} reaches the end of file &
-\includegraphics[width=3.38889in,height=1.45556in]{media/image33.emf}\tabularnewline
-\bottomrule
-\end{longtable}
-
-In panel 4, notice that Rebecca's gesture of swooping back actually
-\emph{precedes} her saying ``go back down.'' In other words, her body
-articulates the idea of cycling back before she actually speaks it. A
-similar argument could be made for panel 8, where Rebecca's gesture for
-a process hitting the bottom (hand swooping and hitting an inflection
-point) precedes her saying ``reaches the end of file''
-
-I tried to sum up my understanding of Rebecca's description:
-
-Interviewer: So as you step through this loop /mmhmm/ \{points to i++\}
-i keeps going up by one==
-
-Rebecca: ==Yes. (Interview 4 of 5, April 6, 2012)
-
-As I thought about that, Rebecca continued explaining and gesturing:
-
-\protect\hypertarget{ux5fToc252445965}{}{}Table -- Rebecca concludes her
-visual and gestural explanation for scanning in track titles (Interview
-4 of 5, April 6, 2012)
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule
-Panel \# & What she said & What she did\tabularnewline
-\midrule
-\endhead
-1 & And the \textbar{}\textbar{}lines keep going
-down\textbar{}\textbar{} \textbar{}\{left hand horizontally chops the
-air, creating ``ladder rungs''\}\textbar{} & (Gesture partially obscured
-by computer )\tabularnewline
-2 & so, \textbar{}\textbar{}that way,\textbar{}\textbar{}
-\textbar{}\{left hand raises up to her head, palm down and parallel to
-the table, and it is replaced by right hand, index finger extended and
-pointing to her left\}\textbar{} &
-\includegraphics[width=2.10000in,height=2.60000in]{media/image34.emf}\tabularnewline
-3 & \textbar{}\textbar{}the first line\textbar{}\textbar{}
-\textbar{}\{right hand scans across to her right\}\textbar{} &
-\includegraphics[width=1.57778in,height=3.50000in]{media/image35.emf}\tabularnewline
-4 & is going to \textbar{}\textbar{}be\textbar{}\textbar{}
-\textbar{}\{right hand rotates to become pinched thumb and forefinger,
-palm out\}\textbar{} , \textbar{}\textbar{}uh the
-p---\textbar{}\textbar{} \textbar{}\{right-hand wiggles\}\textbar{} &
-\includegraphics[width=3.25556in,height=1.62222in]{media/image36.emf}\tabularnewline
-5 & element zero /OK/ uh \textbar{}\textbar{}the
-second\textbar{}\textbar{} \textbar{}\{hand swoops slightly up, then
-curves down and locks in in a position below the prior one\}\textbar{}
-one'll be element one &
-\includegraphics[width=3.25556in,height=1.55556in]{media/image37.emf}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\subsection{Conclusion}\label{conclusion-1}
-
-In pulling together concluding ideas, I revisit the points I established
-at the end of section 3.1.
-
-\subsubsection{Students' early-stage design activity reveals patterns
-outside the explanatory scope of (mis)conceptions
-accounts}\label{students-early-stage-design-activity-reveals-patterns-outside-the-explanatory-scope-of-misconceptions-accounts}
-
-Lionel's strategy of working at a whiteboard, keeping himself at a
-top-level of planning, and copying pseudo-code into a computer isn't
-explained by appealing to ``concepts'' in computer science, which in
-many research accounts are simply mappings of content in computer
-science curricula. I demonstrate this gap in Appendix 4, where the
-``conceptual features'' of Lionel's code reveal little (if anything)
-about the resources involved in Lionel's design process.
-
-Rebecca, in a similar fashion, has lots of productive knowledge for
-thinking about arrays (section 3.5.2), but at the time of the interview
-none of her code reflected that knowledge. Moreover, Rebecca was able to
-sense-make (Danielak et al., in press) about an array-of-pointers
-design, where her sense-making was again a complex activity that would
-be poorly accounted for in CSEd frameworks that appeal to
-misconceptions. In Rebecca's case, the situation is particularly
-paradoxical if we consider conceptual knowledge about a topic to naively
-be something students have or don't have.
-
-Suppose Rebecca had the requisite conceptual knowledge for thinking
-about an array of pointers. It's a puzzle, then, to explain solely via
-conceptual knowledge why she was so frustrated with programming and felt
-stuck when she began Interview 4. Why hadn't the knowledge she
-possessed---relevant to solving the problem at hand---manifested already
-some way? Why was she stuck?
-
-The alternative is to assume that as of Interview 4 she didn't have the
-conceptual knowledge for thinking about the problem at hand. But, that
-doesn't make sense either. If Rebecca didn't know how to think about an
-array of pointers, why was she able to do it so well during the
-interview? Both possibilities---that either Rebecca had or did not have
-conceptual knowledge about an array of pointers---lead to fairly
-non-sensical conclusions under naïve have/don't have assumptions of
-conceptual knowledge. Yet, again, the majority of conceptual knowledge
-research in CSEd today is silent on the issue of degree when it comes to
-conceptual knowledge: students either have a mental model that matches
-canonical function or they don't.
-
-\subsubsection{Rebecca had epistemological resources to support
-expert-like practices, but she framed those practices
-differently}\label{rebecca-had-epistemological-resources-to-support-expert-like-practices-but-she-framed-those-practices-differently}
-
-Consider these summative statements made by Lionel and Rebecca on how
-they use pseudo-code in their work.
-
-\protect\hypertarget{ux5fToc252445966}{}{}Table -- Comparing Lionel's
-and Rebecca's views toward pseudo-code
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Lionel & Rebecca\tabularnewline
-\midrule
-\endhead
-``on my computer I'd, you know I'd write out the pseudocode, I'd
-ac---I'd literally write out the pseudocode, even though obviously it
-wouldn't compile and actually work.'' (Interview 1 of 1, October 17,
-2011) & ``So, I can't just type in ``if the white piece reaches in,''
-I---the, the C language, I guess, putting it in those, their terms,
-their terminology into programming language.'' (Interview 2 of 5,
-February 17, 2012)\tabularnewline
-\bottomrule
-\end{longtable}
-
-As illustrative examples, they capture a fundamental difference in
-practice between Lionel and Rebecca. Lionel worked in pseudo-code on his
-computer, even going so far as to copy it into his source-code files
-despite the fact that it ``obviously wouldn't compile and actually
-work.'' Rebecca also wrote pseudo-code, and she even wrote it into her
-source code files. But, her local sense of the activity of writing
-pseudo-code was different. It came to represent a discontinuity with the
-final code she was trying to write because there was nonetheless a gap
-between writing in English and writing C in ``their terms.''
-
-These diverging stances take the same practice---in this case,
-pseudo-coding---and situate it within a different kind of
-epistemological coherence. For Lionel, the coherence associated with
-pseudo-coding is productive and optimistic. Copying non-working
-pseudo-code into the computer is, in a colloquial sense, all part of the
-plan. It's a legitimate step toward creating a final working program
-and, crucially, it reflects his own understanding of what's supposed to
-happen in his code. For Rebecca, by contrast, writing pseudo-code is a
-fallback. Pseudo-code is what she writes ``if I know what I want to put
-underneath of it,'' by which she means if she's not sure how to flesh
-out a loop or other control structure. Rather than being part of a
-planful coherence, the practice of pseudo-coding for Rebecca gets
-triggered as part of a stopgap coherence; a contingent measure at times
-associated with a kind of diffidence.
-
-Crucially, Rebecca has intellectual resources for pseudo-coding.
-Moreover, her practice of writing pseudo-code to flesh out a loop is the
-plausible beginning of ``method stubbing,'' the process by which one
-might defer implementing the guts of a function and instead create a
-simple stopgap. For example, the function below will always print 28
-Fahrenheit, which is fairly useless as far as temperature-getting
-functions go:
-
-getOutsideTemperature \textless{}- function(sensor) \{
-
-print(``28F'')
-
-\}
-
-But, the advantage of having \emph{something} in the function body, no
-matter how trivial, is enormous. Now, because
-\textbf{getOutsideTemperature} is defined and takes arguments, other
-functions can safely call it. It also offers a diagnostic output---by
-printing ``\textbf{28F}'' out---that we can rely on if we ever need to
-make sure the function was called. At the time of creating
-\textbf{getOutsideTemperature}, a programmer may have no idea how the
-temperature will be obtained but will nonetheless need to be able to
-write code that depends on it getting the temperature. Stubbing is, in
-such cases, a highly productive thing for a programmer to be able to do.
-But, stubbing is as much a strict practice as it is an epistemological
-move, because it represents a decision that an entity-to-be-known can
-be, for the moment, underspecified so it can be incorporated into a
-system before its behavior is defined.
-
-Rebecca is already capable of simple stubbing. It's part of an
-epistemological coherence that's triggered when she doesn't know how to
-flesh out a control structure. So, from a constructivist standpoint
-(Smith et al., 1993), she has knowledge of something that can be further
-refined into software engineering knowledge. Indeed, much of the
-surrounding data from Rebecca suggests that she would have been served
-well if an instructional intervention had taught her some simple ways to
-stub out functions so she could call them even if she hadn't defined
-their bodies yet. To my knowledge, she was never exposed to stubbing
-formally, which is unfortunate considering how much it might have led to
-improving her confidence in programming.
-
-Lastly, consider Rebecca's episode in reasoning about an array of
-pointers (section 3.5.2). As I show, Rebecca clearly had resources for
-reasoning about the sensibility of an array-of-pointers design and for
-predicting array-of-pointers syntax by extrapolating from patterns she
-had seen before. And, when I explicitly suggested that we suppose a
-candidate syntax works, Rebecca was able to fluidly articulate the
-procedure she would build around that syntax (section 3.5.2.3). Rebecca
-knew things, but her knowledge was deployed in such a highly contextual
-way as to be sensitive to what kind of knowledge-activity she thought
-she was supposed to be doing. It's thus sensible to model that set of
-phenomena from an epistemological standpoint, rather than a strictly
-conceptual one. Rebecca had certain kinds of knowledge, but part of my
-in-interview intervention was to establish a frame where \emph{supposing
-syntax worked} was a valid part of programming. Once we mutually
-negotiated that frame there was a strong microcoherence of Rebecca
-stably explaining her code. But, crucially, my intervention wasn't about
-introducing conceptual content but rather establishing a kind of
-knowledge game---supposing---that one could play as part of programming
-design.
-
-\subsubsection{Students displayed a diversity of approaches to
-programming in the
-moment}\label{students-displayed-a-diversity-of-approaches-to-programming-in-the-moment}
-
-The diversity of programming approaches I saw students take is far
-greater than just the space spanned by Lionel and Rebecca. Within and
-across my data, I continually observed students saying and doing things
-that not only distinguished them from each other, but constituted
-phenomena I haven't seen described in prior CSEd research. I saw
-students employ clever tests for debugging, create their own makeshift
-debugger, and discuss at length whether certain programming constructs
-represented natural ways of thinking. For illustrative purposes, I'll
-discuss three such examples. While I don't analyze them at length, I
-include them here as proof of principle of the kinds of phenomena we're
-not currently capturing that nonetheless have a strong effect on
-students' programming approaches. Moreover, these ``in the wild''
-(Hutchins, 1995a) practices are further evidence of the kinds of
-knowledge that could be further refined into student expertise (Smith et
-al., 1993). So, not only do our research accounts miss these kinds
-phenomena, by missing them they preclude research and practice from
-building off them.
-
-\paragraph{Isaac used a thoughtful debugging strategy that code
-snapshots alone could never
-capture}\label{isaac-used-a-thoughtful-debugging-strategy-that-code-snapshots-alone-could-never-capture}
-
-When Isaac was debugging an error in his checkers game code, he made use
-of a clever test to discover he had reversed array index subscripts. He
-had become uncertain about which of the two subscripts corresponded to
-what he thought of as the x-coordinate on the board and which controlled
-the y-coordinate. So, he took a piece whose x-y position he knew and
-instructed his program to print it as a ``7'' on the board. Then, he
-told it to increment the first index of that known piece by 1, leave the
-second index alone and print the resulting piece on the board as ``8''.
-Finally, he told his program to increment the second index of the known
-piece by 1, leave the first piece alone, and print the resulting piece
-on the board as ``9''. By inspecting the visual output, Isaac could tell
-which subscript controlled which position because he knew where 8 and 9
-were relative to the unchanged 7. When he reflected on his work in the
-interview, he explained that what was tricky was realizing that the
-coordinate system for the checkers board didn't work the way standard
-coordinate systems in math worked; in the checkers program the indices
-were reversed.
-
-Because Isaac had been set up as a code-snapshot participant, we have a
-snapshot record of his work for that portion of the interview. But, the
-only snapshot-visible changes are the reversal of the indices and the
-accompanying test cases. The narrative surrounding Isaac's particular
-debugging text, including his reflection that coordinate systems in
-programming are ``trickier'' than those in math, is entirely invisible
-to the snapshot record. If I hadn't interviewed him I never would have
-captured it.
-
-\paragraph{Dana created her own debugging
-environment}\label{dana-created-her-own-debugging-environment}
-
-Dana was working on her checkers program and needed to debug some errant
-behavior. In class and in discussion students had been exposed to the
-GNU Debugger (GDB), so in the interview I asked Dana if she'd like to
-use it. She said she didn't feel comfortable using GDB. Instead, using
-my MacBook, Dana decided to create three separate instances of a
-terminal. In one instance, she had her checkers program source code
-open. In the second, she had the source code for a set of test inputs
-she was developing to try to pinpoint the error. The third window was
-dedicated to compiling and running the code. Since the program was
-designed to print an ASCII (text) representation of the board at each
-turn, the dedicated program window was Dana's visual output for which
-pieces were where at any state in the game.
-
-The important finding from this episode is how Dana deliberately
-structured her environment to support her activity. At the lowest level,
-she could have tried to do everything in a single terminal window. But,
-if she did that she'd have to constantly switch contexts. Because
-students in the course used a terminal version of emacs, she had no easy
-way to simultaneously view her code while compiling it. GDB would have
-been in many respects the right tool for that problem: it would let her
-set breakpoints, step through iterations of code, and inspect variable
-values as she did so. But Dana chose not to use GDB.
-
-Instead, Dana exploited the fact that Terminal.app on Mac OS X 10.7 can
-have multiple instances of itself open at the same time on the same
-screen. With three terminals running, she could persistently have her
-source code visible and executing at the same time. With one glance
-pattern she could move from the board's visual output to the test to the
-program source and back again. Collectively, the field of displays
-enabled a kind of programming that was substantially different---in
-terms of its representational affordances---from using GDB or a single
-terminal instance. Moreover, a snapshot-history-only research approach
-would have never had access to how or why Dana (re)configured her
-development environment.
-
-\paragraph{Toby said recursion was the ``hardest programming way to
-think''}\label{toby-said-recursion-was-the-hardest-programming-way-to-think}
-
-When I gave Toby a code sample task in an interview, he immediately
-commented on the fact that it used recursion. Recursion was, as he said,
-``the hardest programming way to think.'' At one point his exact
-phrasing, quoting the film \emph{Ice Age}, was that ``recursion is bad
-juju.'' The task in question was adapted from an example in Abelson and
-Sussman (1996) that approximates the square root of a number using an
-elegant recursive approach.
-
-Toby's first explanation for why recursion was bad was that it's a
-completely unnatural way to think. He cited examples from cooking by
-saying, in effect, ``Nobody ever stirs by saying stir once, and if it's
-not stirred, stir once. They say stir 100 times.'' He cited simple
-operations---including incrementing a number---where using recursion to
-push and pop a stack seemed needlessly complex. One of his most damning
-indictments of recursion was that his class introduced it as a solution
-to generate numbers in the Fibbonacci sequence. Of course recursion
-works for examples that are mathematically recursive, he reasoned, but
-outside of those obviously contrived examples it was bad joo joo when
-compared to understandable, sensible iteration.
-
-The fuller account of Toby's resistance to recursion---and slight change
-of heart after an instructional moment---is beyond the scope of what I
-can present here. Suffice it to say that Toby's feelings toward
-recursion---rooted in a sense of it being an unnatural way to
-think---strongly directed his approach to programming. The approach was
-so strong, in fact, that during our first interview he wasn't fully able
-to produce working code examples of why recursion was absurd despite his
-vehement conviction that it was. In other words, he maintained that
-recursion was bad and inefficient even though every time he tried to
-demonstrate its inferiority he made mistakes in his code and his
-examples didn't work.
-
-\subsubsection{Dynamic epistemological models can offer a lens for
-reforming assessment and
-instruction.}\label{dynamic-epistemological-models-can-offer-a-lens-for-reforming-assessment-and-instruction.}
-
-Historically, a recognition of in-pieces cognitive dynamics in science
-education (diSessa, 1993; Hammer, 1994) has led to more direct research
-on how those frameworks inform instruction (Hammer \& Elby, 2003;
-Hammer, 1996; Louca et al., 2004). What I describe here is research
-still at the formative end of generating models to explain cognition.
-And, because the focus of the research was explicitly on learning and
-not instruction, I advise prudence in trying to draw certain kinds of
-specific recommendations from the work I present.
-
-Those restrictions being said, we can think carefully about what this
-work means for instruction. For practical purposes, let's refine
-terminology so we can talk more specifically and precisely about
-components of teaching. Below are my working definitions for teaching
-components:
-
-\begin{itemize}
-\item
-  How should instructors change the way they act in the classroom
-  (\emph{instruction})?
-\item
-  How should this research inform the set of intended learning outcomes
-  instructors prepare (\emph{curriculum})?
-\item
-  How should we change the way we measure or otherwise go about
-  ascertaining what students know (\emph{assessment})?
-\end{itemize}
-
-The clearest implications of my work are in instruction and assessment.
-
-From an instruction perspective, it may help teachers to know that
-students can enact different epistemological stances as part of their
-programming practice. As stated, that finding is in principle not new
-(cf., Gaspar \& Langevin, 2007). What is new, I think, is the
-characterization that these stances evidence epistemological resources
-that can be tapped for learning (cf., Hammer \& Elby, 2003). That
-sense-making about program design choices or computational concepts is a
-worthwhile activity, for example, is a message instructors can send to
-students that resonates with larger findings in engineering education
-research (Danielak et al., in press). Moreover, instructors can be more
-sensitive to what kinds of knowledge-activity students think they're
-doing, particularly when making epistemological moves like
-\emph{supposing} might help students work through design problems.
-
-My findings also speak to assessment. In the course I studied, there
-were few (if any) assessment instruments that would have revealed the
-kinds of knowledge I document Lionel and Rebecca having in this study.
-At no point in my observational data of the course were students asked
-to explain a design choice or reason about competing solutions to handle
-a problem. At no point were they encouraged to deeply sense-make about
-the conceptual content of the course, particularly the traditionally
-tricky topics of pointers and pointer arithmetic. At no point in my data
-did the instructor model the kind of sense-making Rebecca did in her
-interviews, a fact which she noted at one point as distinguishing the
-course from Basic Programming, its first-semester counterpart. Because
-the course never explicitly asked students to sense-make, reason about
-design choices, or explain designs, it never used information about what
-knowledge students had about those things to inform its instruction. It
-couldn't. Moreover, because the course never explicitly asked students
-to sense-make, reason about design choices, or explain designs, I was
-left to conclude such things were not prioritized learning outcomes of
-the course. To be clear, all courses establish a kind of focus by
-deciding what won't be covered, and I don't fault the instructor at all
-for running a course where design knowledge wasn't an enacted learning
-outcome. But, I'm also left to wonder: why wasn't it? Given that even
-beginning students can think about software design, and given that
-engineering as a discipline fundamentally involves design, shouldn't it
-have been?
-
-\section{Conclusion}\label{conclusion-2}
-
-If there is one overarching finding from this dissertation, it's that
-students clearly have resources for thinking about designing programs
-and a diversity of approaches to programming in the moment. Below, I
-explain the kind of diversity I saw in programming approaches. Then, I
-conclude with a discussion of what my research might mean for
-assessment.
-
-\subsection{We should think carefully about what students' programming
-design knowledge means for
-assessment}\label{we-should-think-carefully-about-what-students-programming-design-knowledge-means-for-assessment}
-
-My research helps us document and model the kinds of knowledge students
-have. In so doing, it points out kinds of information that the course's
-assessments were apt to miss:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  How students frame or otherwise approach the task of programming in
-  the moment
-\item
-  The role of different kinds of prior experience in stabilizing (or
-  potentially destabilizing) certain kinds of frames
-\item
-  In-the-moment practices---including talking out solutions, sketching
-  out debugging strategies, and writing out pseudo-code---that display
-  students' competence and sense-making
-\item
-  The code history that traces how designs evolve, including how
-  students start projects and what parts of their designs become
-  dead-ends
-\end{enumerate}
-
-Together, those four points are relevant for formative assessment (Black
-\& Wiliam, 1998) in introductory programming. (1) and (2) point us
-toward what students think they're doing when they're programming.
-Analogous work from science education tells us that students' sense of
-the kind of knowledge activity they're enacting matters for learning and
-assessment (Russ et al., 2008). For example, knowing early on that
-students are blindly copying code or randomly trying syntax gives
-instructors a chance to intervene. But, my research suggests
-intervention can't just be about stopping a bad behavior: knowing that
-copy-paste behavior is happening is different from knowing \emph{why}
-it's happening.
-
-Formative assessment has to be about diagnosing causes, not just
-identifying symptoms. Otherwise, we run the risk of ignoring or even
-harming students' productive knowledge. Take the example of copy-paste
-behavior, as documented by Gaspar and Langevin (Gaspar \& Langevin,
-2007). One reason we may see copy-paste behavior, as Study 1
-demonstrates, is that students might see situations as new instances of
-already-solved problems. In professional practice, seeing old solved
-problems in new situations can be productive. Such insights can, for
-example, direct engineers to use a pre-built library of functions
-instead of building their own. But, another related reason for
-copy-paste may be efficiency. For Rebecca, copying and pasting code was
-very fast; it required only a few quick keystrokes. In the short run,
-Rebecca's choice to copy the code was a faster, less demanding, more
-trustworthy route to go on than was abstracting the code to a function.
-When we consider students trying to see common problems in new scenarios
-and solve such problems efficiently, ``copy-paste'' becomes a symptom
-rather than a root cause. Epistemological frameworks, then, can inform
-assessments by offering explanations that aim at the root cause of
-certain behaviors.
-
-Points 3 and 4 complement knowledge analysis by drawing focus to
-artifacts, practices, and history. As data from Lionel shows (Study 2),
-a crucial part of his design process involves artifacts and activities
-that were only distally knowable to the assessments he got in class.
-Lionel's instructor had no direct access to:
-
-\begin{itemize}
-\item
-  Lionel's whiteboard
-\item
-  the hours Lionel may have spent working out designs in chalk
-\item
-  how Lionel might have talked out design features
-\item
-  how Lionel's initial ``pseudo-code'' evolved into his final design.
-\end{itemize}
-
-In Rebecca's case, commented-out code in her final project submission
-were perhaps the only clues at all that she tried abstracting some
-flight day-checking procedures into functions (Study 1). Without
-Rebecca's history, the instructor would have had no reliable way of
-knowing:
-
-\begin{itemize}
-\item
-  Rebecca began her flights database project by copying code from an
-  earlier project
-\item
-  Rebecca's original solution for day-checking evolved from a seven-fold
-  conditional structure, one for each day
-\item
-  Rebecca may have tried abstracting day-checking procedures, even
-  creating a function called check\_days.
-\end{itemize}
-
-To sum up: traditional assessments can tell us what students finally
-produce but not how they produce it, when they produce it, why they
-produce it, or what the production process was. In the summative
-assessment structure of the course, final student products were the only
-products submitted to the instructor. My research shows that what's
-happening in the interstices---before typing, between compiles, away
-from the computer---can be captured and, in principle, analyzed and
-acted upon by instructors.
-
-\subsection{What We Might Change About Classroom
-Practice}\label{what-we-might-change-about-classroom-practice}
-
-What follows is my speculation about how we might specifically change
-classroom practices in light of my research findings. I move from
-recommendations I think are most strongly supported by my data to
-recommendations that align with my findings but are more expansive, and
-thus less strongly supported. As my recommendations broaden, I try to
-offer not only an instructional recommendation but a concomitant
-question for research.
-
-\subsubsection{Instructors could look beyond content to understand
-student
-difficulties}\label{instructors-could-look-beyond-content-to-understand-student-difficulties}
-
-I think first and foremost, instructors have to have a willingness to
-recognize that the difficulties they think they're seeing in students
-are difficulties they may be viewing through a lens of content. For
-example, after seeing a student struggle an instructor might say ``the
-student doesn't know assignment statements.'' But, those difficulties
-almost certainly have a deeper explanation. I say ``almost certainly,''
-where what I mean is that there are a number of patterns we've
-identified as common novice errors, but most research stops before
-asking why that's a common novice error. With rare exceptions (Fleury,
-1991, 2000), the computing education community doesn't encourage asking
-the question \emph{why might a student be doing this?} and \emph{what
-might this tell me about the way a student is making sense of this?} By
-contrast, my research suggests those orientations are ripe, low-hanging
-fruit for instructors to take on. I think one of the first things as
-instructor could ask is, \emph{why would a student be thinking that this
-is the appropriate thing to be doing?} And that holds true whether the
-thing in question is writing the statement this way, or interpreting the
-code this way, or enacting this kind of programming activity this way.
-
-Moreover, we know from research on metacognition (Schoenfeld, 1987,
-1992) that thinking beyond content opens the palette of kinds of
-interventions an instructor can make. Schoenfeld (1987), for example,
-came to encourage metacognition in his classroom by doggedly asking
-student groups what they were doing, why they were doing it, and how
-they hoped it would lead them to a solution. Students ultimately came to
-internalize such strategies and spent far less time floundering along
-solution paths that weren't ultimately productive. Might the same idea
-be true in introductory programming courses? To find out, we could begin
-researching in earnest how often and in what ways instructors model
-metacognition for their students. Currently, I would argue, we don't
-know whether and how instructors do so.
-
-\subsubsection{Instructors could use code history to inform
-interventions}\label{instructors-could-use-code-history-to-inform-interventions}
-
-When, for example, a student comes to office hours with a problem on a
-project, an instructor needs to quickly come to grips with the state of
-the students' project, the logic of their design, where they're stuck,
-why they're stuck, and what might best help them. That's no small task.
-But, given my analysis I have strong reason to believe that having code
-history available to instructor could change both the nature of coming
-to terms with a student's project and the conversation with a student
-that results. If an instructor can see the evolution of a student's
-code, at the very least they could see where the student started, how
-the code was growing, and where the student was working most recently.
-Even if the instructor had never before seen the code (or its history)
-until that moment in office hours, having both available changes the
-kind of view the instructor can get of the code and the specifics of the
-intervention that might result. In Rebecca's case, for example, it might
-have been an opportunity to explore why her check\_days abstracted
-function was failing on her flights database project.
-
-Having code-snapshot capabilities also changes the kind of research we
-can do. First and foremost, a result of this dissertation was to create
-a freely-available, lightweight, open-source framework for capturing and
-visualizing students' code histories.\footnote{https://github.com/briandk/gitvisualizations}
-So, the most basic kind of study would involve deploying that framework
-from the instructional side of a course and exploring what happens, as,
-for example, Hurd (2013) has done. One could ask questions of how
-instructors build code snapshotting into their course, how such
-information could or did inform assessment, and how such information
-could or did change the nature of instructional interactions with
-students. Was it for the better? If so, how do we know? If not, how do
-we know?
-
-\subsubsection{Instructors could establish a norm of asking why a design
-choice makes
-sense}\label{instructors-could-establish-a-norm-of-asking-why-a-design-choice-makes-sense}
-
-The most far-reaching implication of the research I present here is that
-instructors should establish a norm in their classes where anyone, at
-any point, for any piece of code, is allowed to ask \emph{why does this
-make sense as a design choice? Why is that an obvious choice to make?
-How does that work?} Consider a parallel example from mathematics
-education. In Don Saari's calculus class at UC Irvine, he ``invokes the
-principle of what he calls ``WGAD''---``Who gives a damn?'' (Bain, 2004,
-pp. 38--39). Bain explains:
-
-At the beginning of his courses, he tells his students that they are
-free to ask him the question on any day during the course, at any moment
-in class. He will stop and explain to his students why the material
-under consideration at that moment---however abstruse and minuscule a
-piece of the big picture it may be---is important, and how it relates to
-the larger questions and issues of the course. (Bain, 2004, p. 39)
-
-What I'm suggesting is even broader than that, because it's not just a
-question students can ask of professors; it's a question \emph{anyone}
-can ask of \emph{anyone.} I see it as the programming and design
-extension of a sociomathematical norm (Yackel \& Cobb, 1996), and I
-think it could lead to collaborative sense-making. That is, I'm asking
-for an accepted cultural practice that is also itself a design practice,
-whereby an instructor can help create a safe, stable space for a
-community to be reflective and critical about design choices. I also
-think it's an idea that leads to others, such as letting, if not
-requiring, students review one another's code. When students are forced
-to reckon with someone else's code and understand their design
-decisions, they're also forced to justify their own decisions. And
-establishing ``why that design choice?'' or ``how does that make
-sense?'' as a norm provides reciprocal opportunities for students, not
-just instructors, to improve the code of others.
-
-The research questions that come out of such an idea would include:
-
-\begin{itemize}
-\item
-  How can an instructor satisfactorily establish norms about design in a
-  classroom?
-\item
-  What does it look like when students collaboratively sense-make about
-  a program's design choices?
-\item
-  What does it look like when students are asked to reflect on their own
-  design choices?
-\item
-  How might introducing a code review component into a course change the
-  way students approach design? How might it improve students'
-  conceptual understanding? How might it improve the quality of the code
-  students produce?
-\end{itemize}
-
-\subsection{Final Remarks}\label{final-remarks}
-
-How students design programs matters for learning and instruction in
-engineering. It matters because finished code reflects what students
-know about design, whether or not instructors capture such information.
-It matters because students have resources for learning about and
-engaging in design; whether or not curriculum, instruction, and
-assessment choose to tap into those resources. It matters because design
-should be an intellectual thread that runs through all engineering
-courses. That thread shouldn't stop when we introduce students to
-programming.
-
-\section{Appendix 1 -- Transcript
-conventions}\label{appendix-1-transcript-conventions}
-
-\begin{itemize}
-\item
-  Turns are not numbered, but they are blank-line-delimited
-\item
-  Short interjected speech that does not interrupt a speaker's turn is
-  bounded /by slashes/
-\item
-  Matching double brackets show the {[}{[}onset and termination{]}{]} of
-  overlapping talk across turns.
-\item
-  *emphasized speech* is bounded by asterisks
-\item
-  Parenthetical clarifications by the analyst appear in parentheses. (He
-  considered but rejected square brackets.)
-\item
-  matching double equals signs mark turn boundaries== ==with minimal or
-  no audible silence (also known as latching turns)
-\item
-  Where gestures don't overlap speech they are in-lined by curly braces
-  when they happen \{folds arms, having made his point\}. All gestures
-  enacted \emph{by} a speaker appear within that speaker's turn unless
-  otherwise noted \{smiles, self-satisfied at having made this important
-  clarification\} \{audience scoffs\}.
-\item
-  When gestures happen \emph{during} speech, the speech is presented
-  first and bounded by double pipes. Gestures that happen simultaneously
-  with such speech are bounded by curly braces, nested within double
-  pipes, and immediately follow the speech they overlap. For example:
-\end{itemize}
-
-Throwing chainsaws \textbar{}\textbar{}*up*\textbar{}\textbar{}
-\textbar{}\{throws chainsaw\}\textbar{} is easy. Just be careful when
-they come \textbar{}\textbar{}*down*\textbar{}\textbar{}
-\textbar{}\{catches chainsaw for a punctuated finish\}\textbar{}.
-
-\section{Appendix 2 -- Visual conventions for
-gestures}\label{appendix-2-visual-conventions-for-gestures}
-
-A challenge of this dissertation was trying to use still-frames to
-convey motion and animation. Where I could, I annotated pictures with
-arrows to show trajectories of movement:
-
-\includegraphics[width=5.62222in,height=0.87778in]{media/image38.emf}
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-\textbf{Description} & \textbf{Example}\tabularnewline
-\midrule
-\endhead
-\textbf{Blue} arrows project forward to show action that will occur but
-hasn't yet in the frame you're looking at. The origin is where a hand
-\emph{is}; the terminus is where it \emph{will be} shortly. &
-\includegraphics[width=1.67778in,height=1.58889in]{media/image39.png}\tabularnewline
-\textbf{Pink} arrows show action that already occurred before the frame
-you're looking at. The origin is where a hand \emph{was}; the terminus
-is where it \emph{is}. &
-\includegraphics[width=1.65556in,height=2.37778in]{media/image20.png}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\section{\texorpdfstring{Appendix 3 -- Transcript of Rebecca's
-pseudo-code episode \emph{without} gesture
-codes}{Appendix 3 -- Transcript of Rebecca's pseudo-code episode without gesture codes}}\label{appendix-3-transcript-of-rebeccas-pseudo-code-episode-without-gesture-codes}
-
-Interviewer: OK. So then, um, what if we just pretended for a minute,
-that that, like==
-
-Rebecca: ==That that works=
-
-Interviewer: ==That that worked==
-
-Rebecca: ==OK==
-
-Interviewer: ==OK. So then, um. So then you might write like to make an
-array of \{stops writing\} actually what would we call this? This is==
-
-Rebecca: ==array of pointers, I guess==
-
-Interviewer: ==OK. Pointers. It'd be, well. So. The---the one we had was
-like int, star p, um, it would be some number, um, and I guess that'd be
-it in order to just declare /Yes/ that right? OK. And then you're
-saying, how would you like access an element of it?
-
-Rebecca: Uh, well, what I was think---like if I wanted to like, save it
-or whatever, /Yeah/ I could make a while loop, scan in the---scan in the
-data from the album, /OK/ uh, which is, I can write it {[}if you want
-
-Interviewer: Sure, go ahead{]}{]}
-
-Rebecca: Um, so, it was==
-
-Interviewer: {[}{[}Oh, sorry, just have to{]}
-
-Rebecca: {[}Oh{]}
-
-Interviewer: It's /Oh, OK/ one of those annoying ticks, but it works
-best cause of the camera if, ah, /Ah, OK/ if that's pointing away from
-your hand
-
-Rebecca: {[}OK{]}
-
-Interviewer: {[}That's fine{]}
-
-Rebecca: Um, there was album, and then, so there was, uh, percent d, and
-percent s, and then give those names, just, I'll just call it number,
-and title. /Mmmhmm/ And so, at least, my---until it does not equal EOF,
-and then my thinking at least, is you should be able to, um, say that
-
-Rebecca: ``star p of i'' /mmhmm/ equals, uh, the title, and then you
-just do i++, so then it'll move to the next one /OK/ and you just keep
-saving each of the pointers in the array to a title /Mmmhmm/ And you
-just increment by 1 until you reach the end of file.
-
-Interviewer: OK==
-
-Rebecca: ==Like *that* would make sense to me.
-
-Interviewer: So then, like, what would go in p of one would be the first
-title we read in==
-
-Rebecca: ==Yes,
-
-Interviewer: uh, would, when the while loop runs again, does it get a
-fresh line {[}{[}line from that
-
-Rebecca: Yeah, because{]}{]} uh, what the while loop does is it reads in
-the line and then once it reaches the end of character it'll go back
-down---it'll do the while loop and then go back down to the next
-thing---line, it'll read in the line until it reaches the end of file
-
-Interviewer: So as you step through this loop /mmhmm/ i keeps going up
-by one==
-
-Rebecca: ==Yes.
-
-Interviewer: Uh
-
-Rebecca: And the lines keep going down so, that way, the first line is
-going to be, uh the p---element zero /OK/ uh the second one'll be
-element one (Interview 4 of 5, April 6, 2012)
-
-\section{Appendix 4 -- Conceptual knowledge frameworks in computing
-don't tell us much about how Lionel structured an in-interview
-program}\label{appendix-4-conceptual-knowledge-frameworks-in-computing-dont-tell-us-much-about-how-lionel-structured-an-in-interview-program}
-
-In this section, I explore Lionel's work in solving an in-interview
-programming task. First, I reproduce the prompt Lionel was given. Next,
-I show the final source code of his solution to the problem, written in
-C. In the analysis that follows, I move outward to explore the context
-of that code's production.
-
-My choice to start with the code first is deliberate. I'm trying to
-mimic what the instructor of a typical programming course might see: the
-final submitted form of a student's code. This approach is admittedly a
-bit backwards, since the data I show after I present Lionel's submission
-is all about what happened \emph{before} the code reached its final
-form. But, I think this approach useful because it invites us to explore
-an artifact first, then raise questions about the conditions of its
-production. In that sense, my presentation has the anthropological feel
-of understanding a found object, which I think is a faithful way of
-looking at what many instructors (not to mention researchers, and of
-course active software developers) face in their day-to-day work.
-
-\subsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188555}{}{\protect\hypertarget{ux5fToc252445950}{}{}}Analyzing
-Lionel's Solution From a Conceptual Knowledge
-View}{Analyzing Lionel's Solution From a Conceptual Knowledge View}}\label{analyzing-lionels-solution-from-a-conceptual-knowledge-view}
-
-To infer conceptual knowledge, I'll use Elliott Tew's (2010) conceptual
-categories for procedural programming in first-semester computer science
-(CS1) courses. My reasons for doing so are:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Elliott Tew (2010) sought concepts that were language-agnostic. That
-  is, the concepts Elliott Tew (2010) identified are not idiomatic to
-  just one language. Rather, they are general ideas implemented in
-  nearly every major procedural language taught to undergraduates (Java,
-  Python, C-based languages, Scheme).
-\item
-  Elliott Tew's (2010, pp. 23--27) selection process was extensive. She
-  began with curricular documents, moved to canonical texts, then
-  ultimately conducted a thematic analysis on emerging categories to
-  produce her final list of concepts.
-\item
-  To date, Elliott Tew's (2010) identification of CS1 concepts is the
-  only work I know of that has been carried forward to create,
-  administer, and validate concept inventories in CS1 (Elliott Tew \&
-  Guzdial, 2010, 2011)
-\end{enumerate}
-
-Figure 2 below presents an overview of the conceptual features I
-identified in Lionel's code. Even in a short, relatively simple program
-Lionel exhibits four of the ten concept categories Elliott Tew (2010)
-identifies. Moreover, he uses each of them ``correctly,'' so to speak,
-in that not only does his program compile, it also properly finds the
-range of the numbers it's given. It's also worth noting that Lionel's
-code displays consistent, helpful use of whitespace (indentation) as
-well as informative comments. While these attributes aren't conceptual,
-per se, they are the sorts of features an instructor might consider in
-assigning a grade to this program.
-
-\includegraphics[width=5.60000in,height=5.67778in]{media/image40.jpeg}
-
-\protect\hypertarget{ux5fToc252445975}{}{}Figure ‑ -- Highlighting the
-conceptual features of Lionel's code given Elliott Tew's (2010) outline
-of CS 1 concepts
-
-\section{Appendix 5 -- Neverly-Asked Questions
-(NAQs)}\label{appendix-5-neverly-asked-questions-naqs}
-
-\subsubsection{Neverly Asked Questions (NAQs) about my conceptual
-analysis of Lionel's
-code}\label{neverly-asked-questions-naqs-about-my-conceptual-analysis-of-lionels-code}
-
-\emph{Didn't Elliott Tew (2010) define those concepts as a way of
-creating a concept inventory?}
-
-Yes. In her work, I believe each concept was represented by at least one
-multiple-choice question (MCQ) with carefully-chosen distractors. In
-each MCQ a student would see a snippet of pseudo-code and be asked to
-choose from a list of responses. The structure was explicitly modeled
-after Force Concept Inventory (FCI) work in physics education.
-
-\emph{So, if her intent (and research) was about making a concept
-\textbf{inventory}, aren't you misusing her ideas of concepts?}
-
-What makes you say that?
-
-\emph{Well, first off, she used categories of concept to create concrete
-instances of code, which she tests students on}\ldots{}
-
-That's my understanding, yep.
-
-\emph{But you used a concrete instance of code --- Lionel's code --- and
-inferred the existence of concepts in it. So, that's not the same thing
-as what Elliott Tew (2010) did.}
-
-I agree I'm not doing what she did, but I don't think that observation
-necessarily undermines the usefulness of what I \emph{am} doing. If the
-concepts she identifies really are general concepts, then it would seem
-silly to think we couldn't find concrete instances of code that
-exemplify them. I think I might rephrase your objection: ``how can I
-trust that you properly identified concepts in Lionel's code?'' My
-answer to that is, again, yes, I have no coding scheme from Elliott Tew
-to apply. But, the code Lionel wrote is there and my candidates for
-assigning concepts are there: my results are open to inspection and
-challenge.
-
-\emph{OK. Fine. But what ``conceptual knowledge analysis'' is there to
-speak of? You just made a slide and identified which parts of his code
-might correspond to which concepts.}
-
-What I did was assign concepts to canonically-correct --- that is,
-syntactically valid parts of Lionel's code that also work properly to
-make the program produce the right results --- sections of Lionel's
-code. I think your objection is about my inferential basis: ``why does
-the presence of those code chunks --- and my tagging certain chunks as
-aligning with Elliott Tew's (2010) concepts --- imply that Lionel has
-those concepts in his head?''
-
-\emph{I accept your amendment. So, what basis do you have for saying
-``this concept is in Lionel's mind?''}
-
-I feel like I'm on the same playing field as the concept inventory
-folks, though in some ways I'm setting a modified standard for validity.
-Elliott Tew (2010) says this about validity:
-
-validity is the evidence that assures us that questions about a
-particular concept are indeed measuring that concept. For instance, a
-question about arrays should require a student to have knowledge about
-arrays, but should not require knowledge about another concept, such as
-recursion. In addition, it is important that the question cannot be
-answered correctly without knowledge of arrays. (Elliott Tew, 2010, p.
-10)
-
-I don't know that I share her assumptions about validity. But, what
-specifically might be at issue here is ``does the presence of a for-loop
-in code imply the student has knowledge of definite loops?'' My answer
-is, ``if not that, then what?''
-
-\emph{But couldn't I argue that the for-loop could be there because he
-copied and pasted it from the internet? That he wouldn't have had to
-know anything about definite loops to do that?}
-
-Yes, you could. My rejoinder to that is in section 3.3.3.
-
-\subsubsection{Neverly-asked questions about Lionel's verbal pseudo-code
-description}\label{neverly-asked-questions-about-lionels-verbal-pseudo-code-description}
-
-Why do you so pedantically analyze such a short, small snippet of
-conversation?
-
-Because by Lionel's own assertion what was said constitutes his ``main
-concept'' (line 12) for the program. That is, by his own account he has
-just expressed the entire top-level design for the program. But, he's
-done it all all without writing a single additional line of code.
-
-So?
-
-If anything, this episode suggests that Lionel can and does have ways of
-expressing a process or procedure that don't rely on writing code. More
-importantly, I think, is that he views it as a legitimate activity to
-express ideas at a level \emph{above} the particularities of syntax and
-implementation.
-
-What do you mean ``views it as a legitimate activity?''
-
-I think here I'm appealing to my interpretation of Goffman's (1974)
-\emph{frames}: a participant's understanding of ``what is it'' that's
-going on in a social situation. Lionel takes up my bid for him to
-explain what he's thinking, but he does so first by explaining his
-``general concept'' for the program.
-
-Right. But, I mean, he's just doing what you \textbf{asked} him to do. I
-don't understand how any of this relates to Goffman or frames. That just
-seems like a needlessly complicated way to explain he did what you
-asked.
-
-OK, let's step back for a second. Suppose you've just started going to
-group therapy. One of the things group therapy environments often stress
-is that you should try to talk about how \emph{you feel}, which can be a
-hard transition for people. Group therapy participants may be used to
-saying ``you always do {[}X{]}'' when in a fight with someone. And, if
-prompted by a group leader to discuss how they feel about troublesome
-instance, a participant may adopt that kind of language. ``My partner
-always yells at me if I'm out late,'' for example.
-
-Group therapy can encourage inward reflection in a way that reshapes how
-participants talk about themselves and their feelings. So, after some
-time in group therapy our hypothetical participant may change the way
-they orient toward questions. If asked ``how are you feeling about the
-fight you and your partner had?'' our hypothetical participant may now
-say ``I feel like I don't have the freedom to see the people I want to
-see.'' The focus in speech shifts from the \emph{other}---in this case a
-partner---to the \emph{self}. Moreover, the substance of the speech is
-about feelings, rather than behavior.
-
-I think these reshapings of speech are neither trivial nor incidental.
-If we had transcripts of our hypothetical person's early sessions in
-group therapy and compared them to those of later sessions, I think we
-would not be surprised to see a marked change. When asked the same
-question---in this case ``how are you feeling about {[}X{]},'' the
-structure and character of our participant's response changes after
-therapy.
-
-My point here is that an everyday sense of how participants' speech
-patterns change in therapy primes our intuition for thinking more
-broadly about social phenomena. A cynical view of our participant would
-still have to acknowledge that, observably, the structure of their
-response to the question ``how does that make you feel?'' changes after
-therapy. I think a sociolinguistic stance would be more generous: our
-participant is \emph{orienting} differently toward the question.
-Furthermore, I take the following as evidence of such an orientation:
-
-\begin{itemize}
-\item
-  A change in sentence focus from the other to the self
-\item
-  Introspection into the source of a feeling
-\item
-  Acknowledgment of personal responsibility for feeling those feelings
-\end{itemize}
-
-I think we as analysts can acknowledge that after therapy, our
-participant frames the event of being asked, ``how does that make you
-feel?'' differently.
-
-OK. So, you're saying that if we look at the change in how someone
-answers the question ``how does that make you feel?'' we can find
-evidence of the superstructure---a frame---that directs their sense of
-appropriate ways to answer that question?
-
-Yes.
-
-So how does this get back to your claims about Lionel?
-
-I think it's tempting to take this position that, ``it's
-inevitable/totally expected that he would just verbalize the top-level
-description of his procedure! I mean, what else could he possibly do?''
-
-I mean, yeah. Kind of.
-
-Right, but I think actually that assumption isn't in line with a classic
-frame analysis approach. Even if the assumption is right, frame analysis
-changes the explanation from ``it's inevitable'' to ``hmm, I guess
-almost everyone I've ever seen brings or enacts the same structure of
-expectations.''
-
-So, my argument is, if you ask someone to describe to you how they're
-thinking about their program, they're gonna do that. What else
-\textbf{would} they do?
-
-Right. And I don't think we disagree that they'll describe their
-thinking. I think where we disagree is that everyone will describe in
-the same way. I'm saying \emph{if} they describe their program to you,
-\emph{that} they're able to express it, what they choose to verbalize,
-what they choose to leave out, how their gestures accompany the
-explanations, and their ultimate criteria for having satisfied the task
-of describing for themselves are all driven underneath (or from above?)
-by structures of expectations. And even if \emph{one thousand} out of
-one thousand people do the same kind of thing Lionel does, that doesn't
-undermine the idea that people have and bring structures of expectations
-to a task. What I think it does is show that one thousand out of one
-thousand people share the same sets of expectations about what it is
-that's going on when they're asked to verbalize their thinking on a
-programming protocol.
-
-So, what's the point of this whole section on Lionel's verbal
-description, then?
-
-I think the point is that when I ask Lionel to verbalize his thinking,
-he outlines through talk what the ``main concept'' of his program is. I
-focus on \emph{that} he decides to describe it out in words, \emph{how}
-he describes it, \emph{what} he leaves out\emph{, how} he repairs, and
-how all of that reflects structures of expectations about the activity
-and what he's being asked to do. Ultimately, the argument is that
-Lionel's choice to take me up on that bid and how he does so
-reflects---to me, the analyst---that he views \emph{verbal description
-of a high-level program} as a legitimate activity and part of his
-approach to programming.
-
-\section{References}\label{references}
-
-Abelson, H., \& Sussman, G. J. (1996). \emph{Structure and
-Interpretation of Computer Programs} (2nd ed.). Cambridge, Mass: MIT
-Press.
-
-Adelson, B., \& Soloway, E. (1985). The Role of Domain Experience in
-Software Design. \emph{IEEE Transactions on Software Engineering},
-\emph{SE-11}(11), 1351 -- 1360. doi:10.1109/TSE.1985.231883
-
-Archer, L. B. (1979). Whatever became of design methodology.
-\emph{Design Studies}, \emph{1}(1), 17--18.
-
-Bain, K. (2004). \emph{What the best college teachers do}. Cambridge,
-Mass: Harvard University Press.
-
-Baker, A., \& van der Hoek, A. (2010). Ideas, subjects, and cycles as
-lenses for understanding the software design process. \emph{Design
-Studies}, \emph{31}(6), 590--613. doi:10.1016/j.destud.2010.09.008
-
-Ball, L. J., Onarheim, B., \& Christensen, B. T. (2010). Design
-requirements, epistemic uncertainty and~solution development strategies
-in software design. \emph{Design Studies}, \emph{31}(6), 567--589.
-doi:10.1016/j.destud.2010.09.003
-
-Bayman, P., \& Mayer, R. E. (1983). A diagnosis of beginning
-programmers' misconceptions of BASIC programming statements.
-\emph{Communications of the ACM}, \emph{26}(9), 677--679.
-doi:http://doi.acm.org/10.1145/358172.358408
-
-Black, P., \& Wiliam, D. (1998). Inside the Black Box: Raising Standards
-Through Classroom Assessment. \emph{Phi Delta Kappan}, \emph{80}(2),
-139--44.
-
-Boaler, J. (1998). Open and closed mathematics: Student experiences and
-understandings. \emph{Journal for Research in Mathematics Education},
-\emph{29}(1), 41--62.
-
-Boaler, J. (2000). Mathematics from another world: Traditional
-communities and the alienation of learners. \emph{The Journal of
-Mathematical Behavior}, \emph{18}(4), 379--397.
-doi:10.1016/S0732-3123(00)00026-2
-
-Boaler, J. (2002). The development of disciplinary relationships:
-Knowledge, practice and identity in mathematics classrooms. \emph{For
-the Learning of Mathematics}, \emph{22}(1), 42--47.
-
-Boaler, J., \& Greeno, J. G. (2000). Identity, agency, and knowing in
-mathematical worlds. In J. Boaler (Ed.), \emph{Multiple perspectives on
-mathematics teaching and learning} (pp. 171--200). Westport, CT: Ablex
-Pub.
-
-Bonar, J., \& Soloway, E. (1983). Uncovering principles of novice
-programming. \emph{Proceedings of the 10th ACM SIGACT-SIGPLAN Symposium
-on Principles of Programming Languages}, 10--13.
-doi:10.1145/567067.567069
-
-Bonar, J., \& Soloway, E. (1985). Preprogramming Knowledge: A Major
-Source of Misconceptions in Novice Programmers. \emph{Human-Computer
-Interaction}, \emph{1}(2), 133.
-
-Bucciarelli, L. L. (1994). \emph{Designing engineers}. Cambridge, Mass:
-MIT Press.
-
-Clancy, M. (2004). Misconceptions and Attitudes that Interfere with
-Learning to Program. In S. Fincher \& M. Petre (Eds.), \emph{Computer
-Science Education Research} (pp. 85--100). London, UK: RoutledgeFalmer.
-
-Danielak, B. A., Gupta, A., \& Elby, A. (in press). The Marginalized
-Identities of Sense-Makers: Reframing Engineering Student Retention.
-\emph{Journal of Engineering Education}.
-
-Danielsiek, H., Paul, W., \& Vahrenhold, J. (2012). Detecting and
-Understanding Students' Misconceptions Related to Algorithms and Data
-Structures. In \emph{Proceedings of the 43rd ACM Technical Symposium on
-Computer Science Education} (pp. 21--26). New York, NY, USA: ACM.
-doi:10.1145/2157136.2157148
-
-diSessa, A. A. (1986). Models of Computation. In D. A. Norman \& S. W.
-Draper (Eds.), \emph{User centered system design: new perspectives on
-human-computer interaction} (pp. 201--218). Hillsdale, N.J: L. Erlbaum
-Associates.
-
-diSessa, A. A. (1993). Toward an Epistemology of Physics.
-\emph{Cognition and Instruction}, \emph{10}(2/3), 105--225.
-
-diSessa, A. A. (2002). Why ``Conceptual Ecology'' is a good idea. In M.
-Limón \& L. Mason (Eds.), \emph{Reconsidering conceptual change: issues
-in theory and practice} (pp. 29--60). Dordrecht ; Boston: Kluwer
-Academic Publishers.
-
-diSessa, A. A., \& Sherin, B. L. (1998). What changes in conceptual
-change? \emph{International Journal of Science Education},
-\emph{20}(10), 1155--1191. doi:10.1080/0950069980201002
-
-Duckworth, E. R. (2006). \emph{``The having of wonderful ideas'' and
-other essays on teaching and learning} (3rd ed.). New York: Teachers
-College Press.
-
-Elby, A., \& Hammer, D. (2010). Epistemological resources and framing: A
-cognitive framework for helping teachers interpret and respond to their
-students' epistemologies. In L. D. Bendixen \& F. C. Feucht (Eds.),
-\emph{Personal epistemology in the classroom: theory, research, and
-implications for practice} (pp. 409--434). Cambridge, UK ; New York:
-Cambridge University Press.
-
-Elliott Tew, A. (2010). \emph{Assessing fundamental introductory
-computing concept knowledge in a language independent manner} (Ph.D.).
-Georgia Institute of Technology, Ann Arbor. Retrieved from ProQuest
-Dissertations \& Theses Full Text. (873212789)
-
-Elliott Tew, A., \& Guzdial, M. (2010). Developing a validated
-assessment of fundamental CS1 concepts. \emph{Proceedings of the 41st
-ACM Technical Symposium on Computer Science Education}, 97--101.
-doi:10.1145/1734263.1734297
-
-Elliott Tew, A., \& Guzdial, M. (2011). The FCS1: a language independent
-assessment of CS1 knowledge. In \emph{Proceedings of the 42nd ACM
-technical symposium on Computer science education} (pp. 111--116). New
-York, NY, USA: ACM. doi:10.1145/1953163.1953200
-
-Erickson, F. (1986). Qualitative methods in research on teaching. In M.
-C. Wittrock (Ed.), \emph{Handbook of research on teaching} (3rd ed., pp.
-119--161). New York : London: Macmillan ; Collier Macmillan.
-
-Eynde, P., \& Hannula, M. (2006). The Case Study of Frank.
-\emph{Educational Studies in Mathematics}, \emph{63}(2), 123--129.
-doi:10.1007/s10649-006-9030-8
-
-Fleury, A. E. (1991). Parameter passing: the rules the students
-construct. In \emph{Proceedings of the twenty-second SIGCSE technical
-symposium on Computer science education} (pp. 283--286). New York, NY,
-USA: ACM. doi:10.1145/107004.107066
-
-Fleury, A. E. (1993). Student Beliefs about Pascal Programming.
-\emph{Journal of Educational Computing Research}, \emph{9}(3), 355--371.
-doi:10.2190/VECR-P8T6-GB10-MXJ5
-
-Fleury, A. E. (2000). Programming in Java: student-constructed rules.
-\emph{SIGCSE Bull.}, \emph{32}(1), 197--201. doi:10.1145/331795.331854
-
-Gainsburg, J. (2006). The mathematical modeling of structural engineers.
-\emph{Mathematical Thinking \& Learning}, \emph{8}(1), 3--36.
-doi:10.1207/s15327833mtl0801\_2
-
-Gal-Ezer, J., \& Zur, E. (2004). The efficiency of
-algorithms-\/-misconceptions. \emph{Computers \& Education},
-\emph{42}(3), 215--226.
-
-Gaspar, A., \& Langevin, S. (2007). Restoring ``coding with intention''
-in introductory programming courses. \emph{Proceedings of the 8th ACM
-SIGITE Conference on Information Technology Education}, 91--98.
-doi:10.1145/1324302.1324323
-
-Ginsburg, H. P. (1997). \emph{Entering the child's mind: the clinical
-interview in psychological research and practice}. Cambridge ; New York:
-Cambridge University Press.
-
-Ginsburg, H. P., \& Opper, S. (1988). \emph{Piaget's Theory of
-Intellectual Development} (3rd ed.). Englewood Cliffs, N.J:
-Prentice-Hall. Retrieved from http://lccn.loc.gov/87017353
-
-Goffman, E. (1974). \emph{Frame Analysis: An Essay on the Organization
-of Experience}. New York: Harper \& Row.
-
-Goldin-Meadow, S. (2003). \emph{Hearing gesture: how our hands help us
-think}. Cambridge, Mass: Belknap Press of Harvard University Press.
-
-Goodwin, C. (2000). Action and embodiment within situated human
-interaction. \emph{Journal of Pragmatics}, \emph{32}(10), 1489--1522.
-doi:10.1016/S0378-2166(99)00096-X
-
-Gupta, A., Hammer, D., \& Redish, E. F. (2010). The Case for Dynamic
-Models of Learners' Ontologies in Physics. \emph{Journal of the Learning
-Sciences}, \emph{19}(3), 285. doi:10.1080/10508406.2010.491751
-
-Hall, R. (1999). Following mathematical practices in design-oriented
-work. In C. Hoyles, C. Morgan, \& G. Woodhouse (Eds.), \emph{Rethinking
-the Mathematics Curriculum} (pp. 29--47). London: Falmer Press.
-
-Hall, R., \& Nemirovsky, R. (2012). Introduction to the Special Issue:
-Modalities of Body Engagement in Mathematical Activity and Learning.
-\emph{Journal of the Learning Sciences}, \emph{21}(2), 207--215.
-doi:10.1080/10508406.2011.611447
-
-Hall, R., \& Stevens, R. (1995). Making space: A comparison of
-mathematical work in school and professional design practices. In S. L.
-Star (Ed.), \emph{The cultures of computing} (pp. 118--145). Oxford, UK:
-Blackwell Publisher.
-
-Hall, R., Stevens, R., \& Torralba, T. (2002). Disrupting
-representational infrastructure in conversations across disciplines.
-\emph{Mind, Culture \& Activity}, \emph{9}(3), 179--210.
-
-Hall, R., Wright, K., \& Wieckert, K. (2007). Interactive and Historical
-Processes of Distributing Statistical Concepts Through Work
-Organization. \emph{Mind, Culture \& Activity}, \emph{14}(1/2),
-103--127. doi:10.1080/10749030701307770
-
-Hammer, D. (1989). Two approaches to learning physics. \emph{The Physics
-Teacher}, \emph{27}(9), 664--670. doi:10.1119/1.2342910
-
-Hammer, D. (1994). Epistemological beliefs in introductory physics.
-\emph{Cognition and Instruction}, \emph{12}(2), 151--183.
-doi:10.2307/3233679
-
-Hammer, D. (1996). Misconceptions or P-Prims: How May Alternative
-Perspectives of Cognitive Structure Influence Instructional Perceptions
-and Intentions? \emph{Journal of the Learning Sciences}, \emph{5}(2),
-97--127. doi:10.1207/s15327809jls0502\_1
-
-Hammer, D., \& Elby, A. (2002). On the form of a personal epistemology.
-In B. K. Hofer \& P. R. Pintrich (Eds.), \emph{Personal epistemology:
-The psychology of beliefs about knowledge and knowing} (pp. 169--190).
-Mahwah, N.J: L. Erlbaum Associates.
-
-Hammer, D., \& Elby, A. (2003). Tapping epistemological resources for
-learning physics. \emph{The Journal of the Learning Sciences},
-\emph{12}(1), 53--90. doi:10.2307/1466634
-
-Hammer, D., Elby, A., Scherr, R. E., \& Redish, E. F. (2005). Resources,
-framing, and transfer. In J. P. Mestre (Ed.), \emph{Transfer of learning
-from a modern multidisciplinary perspective}. Greenwich, CT: IAP.
-
-Hannula, M., Evans, J., Philippou, G., \& Zan, R. (2004). Affect in
-Mathematics Education--Exploring Theoretical Frameworks. Research Forum.
-\emph{International Group for the Psychology of Mathematics Education},
-30.
-
-Henderson, K. (1999). \emph{On line and on paper: visual
-representations, visual culture, and computer graphics in design
-engineering}. Cambridge, Mass: MIT Press.
-
-Herman, G. L., Kaczmarczyk, L., Loui, M. C., \& Zilles, C. (2008). Proof
-by incomplete enumeration and other logical misconceptions.
-\emph{Proceeding of the Fourth International Workshop on Computing
-Education Research}, 59--70. doi:10.1145/1404520.1404527
-
-Hofer, B. K., \& Pintrich, P. R. (1997). The development of
-epistemological theories: Beliefs about knowledge and knowing and their
-relation to learning. \emph{Review of Educational Research},
-\emph{67}(1), 88--140. doi:10.3102/00346543067001088
-
-Holland, S., Griffiths, R., \& Woodman, M. (1997). Avoiding object
-misconceptions. In \emph{Proceedings of the twenty-eighth SIGCSE
-technical symposium on Computer science education} (pp. 131--134). New
-York, NY, USA: ACM. doi:10.1145/268084.268132
-
-Hurd, A. (2013, March). \emph{Assessment in an Introduction to
-Programming Course}. Presented at the 16th Annual Course Technology
-Conference, San Diego, CA, USA. Retrieved from
-http://www.slideshare.net/CengageLearning/andrew-hurd-assessment-in-an-intro-to-programming-course
-
-Hutchins, E. (1995a). \emph{Cognition in the Wild}. Cambridge, Mass: MIT
-Press.
-
-Hutchins, E. (1995b). How a cockpit remembers its speeds.
-\emph{Cognitive Science}, \emph{19}(3), 265--288.
-doi:10.1016/0364-0213(95)90020-9
-
-Izsák, A. (2004). Students' Coordination of Knowledge When Learning to
-Model Physical Situations. \emph{Cognition \& Instruction},
-\emph{22}(1), 81--128.
-
-Jackson, M. (2010). Representing structure in a software system design.
-\emph{Design Studies}, \emph{31}(6), 545--566.
-doi:10.1016/j.destud.2010.09.002
-
-Jadud, M. C. (2006). Methods and tools for exploring novice compilation
-behaviour. In \emph{Proceedings of the 2006 international workshop on
-Computing education research - ICER '06} (p. 73). Canterbury, United
-Kingdom. doi:10.1145/1151588.1151600
-
-Jordan, B., \& Henderson, A. (1995). Interaction analysis: Foundations
-and practice. \emph{Journal of the Learning Sciences}, \emph{4}(1), 39.
-doi:10.1207/s15327809jls0401\_2
-
-Joy, M., Sinclair, J., Sun, S., Sitthiworachart, J., \& López-González,
-J. (2009). Categorising computer science education research.
-\emph{Education and Information Technologies}, \emph{14}(2), 105--126.
-doi:10.1007/s10639-008-9078-4
-
-Kaczmarczyk, L. C., Petrick, E. R., East, J. P., \& Herman, G. L.
-(2010). Identifying student misconceptions of programming.
-\emph{Proceedings of the 41st ACM Technical Symposium on Computer
-Science Education}, 107--111. doi:10.1145/1734263.1734299
-
-Kaiser, D. (2005). \emph{Drawing theories apart: the dispersion of
-Feynman diagrams in postwar physics}. Chicago: University of Chicago
-Press.
-
-Keller, E. F. (1983). \emph{A feeling for the organism: the life and
-work of Barbara McClintock}. San Francisco: W.H. Freeman.
-
-Kolikant, Y. B.-D., \& Mussai, M. (2008). ``So my program doesn't run!''
-Definition, origins, and practical expressions of students'
-(mis)conceptions of correctness. \emph{Computer Science Education},
-\emph{18}(2), 135.
-
-Kölling, M., \& Utting, I. (2012). Building an open, large-scale
-research data repository of initial programming student behaviour. In
-\emph{Proceedings of the 43rd ACM technical symposium on Computer
-Science Education} (pp. 323--324). New York, NY, USA: ACM.
-doi:10.1145/2157136.2157234
-
-Latour, B. (1987). \emph{Science in action: how to follow scientists and
-engineers through society}. Cambridge, Mass: Harvard University Press.
-
-Latour, B. (1990). Drawing things together. In M. Lynch \& S. Woolgar
-(Eds.), \emph{Representation in Scientific Practice} (1st MIT Press ed.,
-pp. 19--68). Cambridge, Mass: MIT Press.
-
-Lehrer, R., Schauble, L., Carpenter, S., \& Penner, D. (2000). The
-interrrelated development of inscriptions and conceptual understanding.
-In P. Cobb, E. Yackel, \& K. McClain (Eds.), \emph{Symbolizing and
-Communicating in Mathematics Classrooms: Perspectives on Discourse,
-Tools, and Instructional Design} (pp. 325--360). Mahwah, N.J: Lawrence
-Erlbaum Associates.
-
-Lising, L., \& Elby, A. (2005). The impact of epistemology on learning:
-A case study from introductory physics. \emph{American Journal of
-Physics}, \emph{73}(4), 372. doi:10.1119/1.1848115
-
-Louca, L., Elby, A., Hammer, D., \& Kagey, T. (2004). Epistemological
-Resources: Applying a New Epistemological Framework to Science
-Instruction. \emph{Educational Psychologist}, \emph{39}(1), 57--68.
-
-Malmi, L., Sheard, J., Simon, Bednarik, R., Helminen, J., Korhonen, A.,
-\ldots{} Taherkhani, A. (2010). Characterizing research in computing
-education: a preliminary analysis of the literature. In
-\emph{Proceedings of the Sixth international workshop on Computing
-education research} (pp. 3--12). New York, NY, USA: ACM.
-doi:10.1145/1839594.1839597
-
-Martin, R. C. (2009). \emph{Clean code: a handbook of agile software
-craftsmanship}. Upper Saddle River, NJ: Prentice Hall.
-
-Mayer, R. E. (1981). The Psychology of How Novices Learn Computer
-Programming. \emph{ACM Computing Surveys (CSUR)}, \emph{13}, 121--141.
-doi:10.1145/356835.356841
-
-Minsky, M. L. (1986). \emph{The Society of Mind}. New York: Simon and
-Schuster.
-
-Nasir, N. S., \& Cooks, J. (2009). Becoming a Hurdler: How Learning
-Settings Afford Identities. \emph{Anthropology \& Education Quarterly},
-\emph{40}(1), 41--61. doi:10.1111/j.1548-1492.2009.01027.x
-
-Nasir, N. S., \& Hand, V. (2008). From the court to the classroom:
-Opportunities for engagement, learning, and identity in basketball and
-classroom mathematics. \emph{Journal of the Learning Sciences},
-\emph{17}(2), 143--179. doi:10.1080/10508400801986108
-
-Nasir, N. S., \& Hand, V. M. (2006). Exploring Sociocultural
-Perspectives on Race, Culture, and Learning. \emph{Review of Educational
-Research}, \emph{76}(4), 449--475.
-
-Nemirovsky, R., Rasmussen, C., Sweeney, G., \& Wawro, M. (2012). When
-the Classroom Floor Becomes the Complex Plane: Addition and
-Multiplication as Ways of Bodily Navigation. \emph{Journal of the
-Learning Sciences}, \emph{21}(2), 287--323.
-doi:10.1080/10508406.2011.611445
-
-Ochs, E., Gonzales, P., \& Jacoby, S. (1996). ``When I come down I'm in
-the domain state'': grammar and graphic representation in the
-interpretive activity of physicists. In \emph{Interaction and Grammar}
-(pp. 328--369). Cambridge: Cambridge University Press.
-
-Papert, S. (1980). \emph{Mindstorms: Children, Computers, and Powerful
-Ideas}. New York: Basic Books. Retrieved from
-http://lccn.loc.gov/79005200
-
-Parsons, J., \& Saunders, C. (2004). Cognitive heuristics in software
-engineering: Applying and extending anchoring and adjustment to artifact
-reuse. \emph{IEEE Transactions on Software Engineering}, \emph{30}(12),
-873 -- 888. doi:10.1109/TSE.2004.94
-
-Patitsas, E., Craig, M., \& Easterbrook, S. (2013). On the Countably
-Many Misconceptions About \#Hashtables (Abstract Only). In
-\emph{Proceeding of the 44th ACM Technical Symposium on Computer Science
-Education} (pp. 739--739). New York, NY, USA: ACM.
-doi:10.1145/2445196.2445443
-
-Paul, W., \& Vahrenhold, J. (2013). Hunting High and Low: Instruments to
-Detect Misconceptions Related to Algorithms and Data Structures. In
-\emph{Proceeding of the 44th ACM Technical Symposium on Computer Science
-Education} (pp. 29--34). New York, NY, USA: ACM.
-doi:10.1145/2445196.2445212
-
-Pea, R. D. (1986). Language-independent conceptual`` bugs'' in novice
-programming. \emph{Journal of Educational Computing Research},
-\emph{2}(1), 25--36.
-
-Pea, R. D., Soloway, E., \& Spohrer, J. C. (1987). The Buggy Path to the
-Development of Programming Expertise. \emph{Focus on Learning Problems
-in Mathematics}, \emph{9}(1), 5--30.
-
-Petre, M., van der Hoek, A., \& Baker, A. (2010). Editorial.
-\emph{Design Studies}, \emph{31}(6), 533--544.
-doi:10.1016/j.destud.2010.09.001
-
-Rodrigo, M. M. T., \& Baker, R. S. J. d. (2009). Coarse-grained
-detection of student frustration in an introductory programming course.
-In \emph{Proceedings of the fifth international workshop on Computing
-education research workshop} (pp. 75--80). New York, NY, USA: ACM.
-doi:10.1145/1584322.1584332
-
-Rodrigo, M. M. T., Tabanao, E., Lahoz, M. B. ., \& Jadud, M. C. (2009).
-Analyzing Online Protocols to Characterize Novice Java Programmers.
-\emph{Philippine Journal of Science}, \emph{138}(2), 177--190.
-
-Rooksby, J. (2010). ``Just try to do it at the whiteboard'':
-Researcher-participant interaction and issues of generalisation. In
-\emph{Proceedings of the Studying Professional Software Design (SPSD)
-Conference}. San Diego, CA, USA.
-
-Rooksby, J., \& Ikeya, N. (2012). Collaboration in Formative Design:
-Working Together at a Whiteboard. \emph{IEEE Software}, \emph{29}(1), 56
---60. doi:10.1109/MS.2011.123
-
-Rosenberg, S., Hammer, D., \& Phelan, J. (2006). Multiple
-Epistemological Coherences in an Eighth-Grade Discussion of the Rock
-Cycle. \emph{Journal of the Learning Sciences}, \emph{15}(2), 261--292.
-doi:10.1207/s15327809jls1502\_4
-
-Russ, R. S., Coffey, J. E., Hammer, D., \& Hutchison, P. (2008). Making
-classroom assessment more accountable to scientific reasoning: A case
-for attending to mechanistic thinking. \emph{Science Education},
-\emph{93}(5), 875--891. doi:10.1002/sce.20320
-
-Saussure, F. de. (1986). \emph{Course in general linguistics}. LaSalle,
-Ill: Open Court.
-
-Scherr, R. E., \& Hammer, D. (2009). Student Behavior and
-Epistemological Framing: Examples from Collaborative Active-Learning
-Activities in Physics. \emph{Cognition \& Instruction}, \emph{27}(2),
-147--174. doi:10.1080/07370000902797379
-
-Schoenfeld, A. H. (1987). What's all the fuss about metacognition? In
-\emph{Cognitive Science and Mathematics Education} (pp. 189--215).
-Lawrence Erlbaum Associates.
-
-Schoenfeld, A. H. (1988). When good teaching leads to bad results: The
-disasters of ``well-taught'' mathematics courses. \emph{Educational
-Psychologist}, \emph{23}(2), 145--166. doi:10.1207/s15326985ep2302\_5
-
-Schoenfeld, A. H. (1991). On mathematics as sense-making: An informal
-attack on the unfortunate divorce of formal and informal mathematics. In
-\emph{Informal reasoning and education} (pp. 311--343).
-
-Schoenfeld, A. H. (1992). Learning to Think Mathematically: Problem
-Solving, Metacognition, and Sense-Making in Mathematics. In D. Grouws
-(Ed.), \emph{Handbook for research on mathematics teaching and learning}
-(pp. 334--370). New York: MacMillan.
-
-Schwartz, D. L., Chase, C. C., \& Bransford, J. D. (2012). Resisting
-Overzealous Transfer: Coordinating Previously Successful Routines With
-Needs for New Learning. \emph{Educational Psychologist}, \emph{47}(3),
-204--214. doi:10.1080/00461520.2012.696317
-
-Seppälä, O., Malmi, L., \& Korhonen, A. (2006). Observations on student
-misconceptions---A case study of the Build -- Heap Algorithm.
-\emph{Computer Science Education}, \emph{16}(3), 241--255.
-doi:10.1080/08993400600913523
-
-Sherin, B. L. (2001). How students understand physics equations.
-\emph{Cognition and Instruction}, \emph{19}(4), 479--541.
-
-Smith, J. P., diSessa, A. A., \& Roschelle, J. (1993). Misconceptions
-Reconceived: A Constructivist Analysis of Knowledge in Transition.
-\emph{The Journal of the Learning Sciences}, \emph{3}(2), 115--163.
-
-Soloway, E. (1986). Learning to program = learning to construct
-mechanisms and explanations. \emph{Communications of the ACM},
-\emph{29}, 850--858. doi:10.1145/6592.6594
-
-Soloway, E., Bonar, J., \& Ehrlich, K. (1983). Cognitive Strategies and
-Looping Constructs: An Empirical Study. \emph{Commun. ACM},
-\emph{26}(11), 853--860. doi:10.1145/182.358436
-
-Soloway, E., \& Spohrer, J. C. (Eds.). (1989). \emph{Studying the Novice
-Programmer}. Hillsdale, N.J: L. Erlbaum Associates.
-
-Spacco, J., Hovemeyer, D., Pugh, W., Hollingsworth, J., Padua-Perez, N.,
-\& Emad, F. (2006). Experiences with Marmoset: Designing and Using an
-Advanced Submission and Testing System for Programming Courses. In
-\emph{ITiCSE '06: Proceedings of the 11th annual conference on
-Innovation and technology in computer science education}. ACM Press.
-
-Spacco, J., Pugh, W., Ayewah, N., \& Hovemeyer, D. (2006). The Marmoset
-project: an automated snapshot, submission, and testing system. In
-\emph{Companion to the 21st ACM SIGPLAN symposium on Object-oriented
-programming systems, languages, and applications} (pp. 669--670). New
-York, NY, USA: ACM. doi:10.1145/1176617.1176665
-
-Spacco, J., Strecker, J., Hovemeyer, D., \& Pugh, W. (2005). Software
-Repository Mining with Marmoset: An Automated Programming Project
-Snapshot and Testing System. In \emph{Proceedings of the Mining Software
-Repositories Workshop (MSR 2005)}. St. Louis, Missouri, USA.
-
-Spohrer, J. C., \& Soloway, E. (1986). Alternatives to construct-based
-programming misconceptions. \emph{ACM SIGCHI Bulletin}, \emph{17},
-183--191. doi:10.1145/22339.22369
-
-Stevens, R., \& Hall, R. (1998). Disciplined perception: Learning to see
-in technoscience. In \emph{Talking Mathematics in School: Studies of
-Teaching and Learning} (pp. 107--150). Cambridge, U.K: Cambridge
-University Press.
-
-Stieff, M. (2007). Mental rotation and diagrammatic reasoning in
-science. \emph{Learning and Instruction}, \emph{17}(2), 219--234.
-doi:10.1016/j.learninstruc.2007.01.012
-
-Tabanao, E. S., Rodrigo, M. M. T., \& Jadud, M. C. (2011). Predicting
-at-risk novice Java programmers through the analysis of online
-protocols. In \emph{Proceedings of the seventh international workshop on
-Computing education research} (pp. 85--92). New York, NY, USA: ACM.
-doi:10.1145/2016911.2016930
-
-Tannen, D. (Ed.). (1993). \emph{Framing in Discourse}. New York: Oxford
-University Press.
-
-Trakhtenbrot, M. (2013). Students Misconceptions in Analysis of
-Algorithmic and Computational Complexity of Problems. In
-\emph{Proceedings of the 18th ACM Conference on Innovation and
-Technology in Computer Science Education} (pp. 353--354). New York, NY,
-USA: ACM. doi:10.1145/2462476.2465604
-
-Traweek, S. (1988). \emph{Beamtimes and lifetimes: the world of high
-energy physicists}. Cambridge, Mass: Harvard University Press.
-
-Turkle, S., \& Papert, S. (1991). Epistemological Pluralism and the
-Reevaluation of the Concrete. In I. Harel, S. Papert, \& Massachusetts
-Institute of Technology (Eds.), \emph{Constructionism: research reports
-and essays, 1985-1990}. Norwood, N.J: Ablex Pub. Corp.
-
-Valentine, D. W. (2004). CS educational research: a meta-analysis of
-SIGCSE technical symposium proceedings. \emph{SIGCSE Bull.},
-\emph{36}(1), 255--259. doi:10.1145/1028174.971391
-
-Van de Sande, C. C., \& Greeno, J. G. (2012). Achieving Alignment of
-Perspectival Framings in Problem-Solving Discourse. \emph{Journal of the
-Learning Sciences}, \emph{21}(1), 1--44.
-doi:10.1080/10508406.2011.639000
-
-VanLehn, K. (1990). \emph{Mind Bugs: The Origins of Procedural
-Misconceptions}. Cambridge, Mass: MIT Press.
-
-Wagner, J. F. (2006). Transfer in Pieces. \emph{Cognition \&
-Instruction}, \emph{24}(1), 1--71. doi:10.1207/s1532690xci2401\_1
-
-Wortham, S. (2006). \emph{Learning Identity: The Joint Emergence of
-Social Identification and Academic Learning}. Cambridge: Cambridge
-University Press.
-
-Yackel, E., \& Cobb, P. (1996). Sociomathematical Norms, Argumentation,
-and Autonomy in Mathematics. \emph{Journal for Research in Mathematics
-Education}, \emph{27}(4), 458--477. doi:10.2307/749877
-
-Yin, R. K. (2009). \emph{Case study research: design and methods} (4th
-ed.). Los Angeles, Calif: Sage Publications.
+% \begin{center}
+%
+%
+% \begin{supertabular}[]{@{}ll@{}}
+% \toprule
+% Statement & Syntax\tabularnewline
+% \midrule
+% \endhead
+% 1 & \texttt{x\ =\ x\ +\ d}\tabularnewline
+% 2 & \texttt{x\ =\ x\ +\ update}\tabularnewline
+% 3 & \texttt{x\_position\ =\ x\_position\ +\ update}\tabularnewline
+% 4 & \texttt{x\_position\ =\ x\_position\ +\ update(t)}\tabularnewline
+% 5 & \texttt{x\_position\ =\ x\_position\ +\ update(...)}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+% \end{center}
+\begin{center}
+  \begin{tabular}{ | l | l |}
+    1 & \texttt{x\ =\ x\ +\ d} \\
+    2 & \texttt{x\ =\ x\ +\ update} \\
+    3 & \texttt{x\_position\ =\ x\_position\ +\ update} \\
+    4 & \texttt{x\_position\ =\ x\_position\ +\ update(t)} \\
+    5 & \texttt{x\_position\ =\ x\_position\ +\ update(...)} \\
+  \end{tabular}
+\end{center}
+
+% I stress that the third column is about \emph{how I might think about}
+% each programming statement\emph{.} By no means am I making normative
+% claims about how one \emph{ought} to think about it, or whether the
+% statement actually does what its names might suggest it does. Rather,
+% ``How I might think about it'' reflects the kind of local meaning or
+% interpretation I might attach to such a statement when I work with it,
+% given my understanding of its role and context.
+%
+% Consider specifically statements 1 and 4. Statement 1 could very well be
+% an incrementer in some kind of iterative code. If I had to debug code
+% that employs statement 1, what I might do is exploit my knowledge of
+% what d is (does it hold a constant value? Is it 1?) and try inserting
+% intermediate print statements into the iterative code. Doing so, I can
+% inspect textually how values change with each iteration. In statement
+% 4's case, it could very well be that the position-update code animates
+% images on a screen. If there's a problem with that code, one of the
+% easiest ways I might notice is that the acceleration seems off in the
+% graphics. Consequently, my debugging might call upon knowledge I have
+% from kinematics. I might try inserting code that draws a dot at the
+% object's on-screen position with each iteration, leaving a trail I can
+% visually inspect. I could then look at the path of the object's
+% trajectory and the spacing patterns between successive dots as a
+% first-pass test of whether my code achieves the motion I want.
+%
+% To be clear, it's not just that graphics might improve my efficiency in
+% debugging a statement. Rather, applying an interpretive frame that
+% treats an assignment statement as a kinematic position update lets me
+% \emph{use conceptual knowledge from physics} to diagnose and fix
+% problems in my code. If I view the assignment statement as saying
+% something about the motion of an object (cf., Hammer, 1994, p. 165), a
+% field of knowledge and concomitant techniques from physics becomes
+% available to me to think with. But, if I focus only on the
+% semantic-level equivalence of statements 1 through 5, there would be no
+% obvious reason for me to access what I know about physics in order to
+% reason about the code.
+%
+% My second example concerns statements that are syntactically-similar,
+% rather than semantically-similar. I proposed that the statements in
+% Table 2 all came from the same language. But, another phenomenon comes
+% into play when different languages use the same symbology for
+% conceptually different operations. Table 3 shows examples of what look
+% like semantically-equivalent operations, but in fact are not.
+%
+% \protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
+% syntactically-similar statements with very different semantics
+%
+% \begin{supertabular}[]{@{}llll@{}}
+% \toprule
+% \textbf{Statement } & Programming Syntax & Language & How I think about
+% it\tabularnewline
+% \midrule
+% \endhead
+% \textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
+% \textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
+% the string w\tabularnewline
+% \textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
+% layer of points onto a plot p\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% The catch here is that the plus operator takes on different roles in
+% different languages because of how those languages define its use.
+% Statement 1 increments a number in C; Statement 2 appends the letters
+% ``ly'' to a string; Statement 3 adds a layer of points to a statistical
+% graphics plot. These different kinds of operations become even more
+% apparent and consequential when, for example, such statements are
+% repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
+%   operator can be written as a directional arrow. The symbol
+%   \textless{}- (``less-than, hyphen'') indicates the value on the right
+%   side of the symbol is being assigned to the variable on the left side
+%   of the symbol. As used, the symbol itself is semantically equivalent
+%   to having written an equals sign (``=``).} I'm using multiple
+% reassignment statements to compose a statistical graphics plot. The
+% layered creation of a plot invites a very different kind of conceptual
+% interpretation than, say, repeatedly accumulating numbers into a running
+% sum:
+%
+% p \textless{}- InitializeGgplot\_1w()
+%
+% p \textless{}- p + GrandMeanLine(owp)
+%
+% p \textless{}- p + GrandMeanPoint(owp)
+%
+% p \textless{}- p + ScaleX\_1w(owp)
+%
+% p \textless{}- p + ScaleY\_1w(owp)
+%
+% p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)
+%
+% One obvious reason for thinking about this code with a different
+% interpretive frame is that numeric addition is commutative; composing a
+% plot is not necessarily commutative.\footnote{To convince yourself that
+%   plot composition isn't commutative, imagine a scale function that
+%   squares up the aspect ratio of a plot and another scale function that
+%   sets the aspect ratio to 1.5:1. Applying the square function last
+%   produces a square plot; applying it first produces a rectangular plot.
+%   String concatenation is also not necessarily commutative. ``cat'' +
+%   ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
+%   ``dogcat.''} So, despite the syntactic similarity, reassignments that
+% compose a plot using reassignment (as above) does not obey the same
+% rules as reassignments for a running total. But, even within this code
+% block, statements that look alike perform operations of a different
+% nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
+% visual layer to a plot, others modify features of the plot (e.g, ``+
+% Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
+% the numeric range of the data).
+%
+% Given these motivating examples, it seems sensible to think there's
+% utility in a programmer having different conceptual metaphors available
+% to think about and work with code. Example 1 shows that choosing to
+% apply knowledge from physics to a piece of code can change the cognitive
+% nature of debugging. There, debugging a position-update statement
+% becomes, in part, reasoning kinematically about the properties of motion
+% trails.\footnote{For comparison, consider the argument that ringing an
+%   aircraft speedometer with physical markers changes the nature of
+%   cognition when pilots work to land a plane (Hutchins, 1995b).} Example
+% 2 shows that across languages, programmers might have to deploy
+% different conceptual metaphors to reason about statements in a
+% locally-consistent way. Knowing that plots in ggplot2 can be composed
+% layer-by-layer with reassignment is crucial if you're trying to write or
+% understand code that creates statistical graphics. But, I would argue
+% that thinking about ◻ = ◻ + ◻ as ``compose new layer onto plot'' can and
+% does appeal to different kinds of knowledge when compared to thinking
+% about ◻ = ◻ + ◻ as ``include this addend in the sum,'' which itself can
+% and does appeal to different kinds of concepts when compared to thinking
+% about ◻ = ◻ + ◻ as ``increment the counter.''
+%
+% Stepping back, I can build the following argument
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   Programming can be helped by applying conceptual models to code,
+%   particularly when relevant domain-knowledge structures can
+%   advantageously transform a problem (example 1)
+% \item
+%   But, conceptual models don't work all the time for all statements.
+%   Because languages are designed differently, the same syntax can
+%   actually correspond to very different operations in code (example 2).
+%   And, that's true both within and across languages.
+% \item
+%   Consequently, it makes less sense to treat conceptual models as right
+%   or wrong, and more sense to treat them as differentially advantageous
+%   for thinking about what a piece of code does. (Thinking a ``+''
+%   implies numerical addition isn't \emph{globally wrong} in JavaScript,
+%   but it won't explain why 1 + ``1'' yields ``11'' as a result.)
+% \item
+%   It seems plausible that successful programmers, when reasoning about
+%   or writing code, are able to dynamically access or deploy conceptual
+%   models that are advantageous given the context (language, syntax,
+%   surrounding code). Certainly such a supposition is in line with work
+%   suggesting students have resources for thinking conceptually and
+%   dynamically about how mathematics models real-world situations (Izsák,
+%   2004; Sherin, 2001)
+% \item
+%   To model how programmers think with conceptual models, a suitable
+%   framework should be able to account for the dynamic, context-sensitive
+%   deployment of conceptual knowledge.
+% \item
+%   To model how programmers develop expertise, a suitable framework
+%   should be able to describe higher-order phenomena. Such phenomena
+%   include explaining how programmers come to have conceptual models or
+%   generate new ones, why they decide to deploy them, and how programmers
+%   consider which conceptual model (i.e., which way to think about code)
+%   is appropriate.
+% \end{enumerate}
+%
+% Taken together, these assertions propose criteria for how we might
+% strive to model cognition in programming. Our modeling frameworks should
+% be context-dependent, dynamic, and capable of explaining where
+% conceptual models come from. They should also be able to account for
+% phenomena that are not themselves conceptual, including what directs the
+% use of certain kinds of conceptual knowledge. In the learning sciences,
+% such models already exist and have proven useful and productive for
+% thinking about thinking.
+%
+% \subsubsection{Manifold models of cognition explain context-dependence
+% and the growth of
+% expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}
+%
+% In 1993, a pair of articles in the learning sciences staked a strong
+% claim for viewing knowledge as a network of pieces, isolated enough to
+% be locally triggered but trainable enough to fire in larger concerted
+% patterns (diSessa, 1993; Smith, diSessa, \& Roschelle, 1993). Informed
+% in part by agent-based accounts of cognition (Minsky, 1986) and complex
+% systems models (diSessa, 2002), the central tenets of an ``in-pieces''
+% approach hold that knowledge is emergent from interacting primitives,
+% rather than unitary and monolithic. An example from Smith, diSessa, and
+% Roschelle helps illustrate the point.
+%
+% The authors show that we might think of a rubber band as a different
+% conceptual entity depending on context. In several different
+% situations---wrapped around a newspaper, pulled taut as a string, spun
+% to store energy in a toy plane propeller---we intuitively think about
+% the rubber band's physical behavior differently: one as a negligible
+% part of the newspaper's point mass, another as a transverse pendulum
+% (likely) obeying Hooke's Law, and the third as a torsional spring. Those
+% differences in intuitive thinking reflect the contextual dependency of
+% what we know about the physical world:
+%
+% In each of the rubberband examples, various pieces of intuitive physical
+% knowledge describe the mechanism at work: the rubber band binds the
+% newspaper, grips the jar lid, and acts a source of springiness for the
+% bobbing object. Although a mapping cannot be made from the rubberband to
+% scientific entities, it is quite easy to map these qualitatively
+% distinct physical processes to scientific entities and laws. For
+% example, instances of binding almost always map to a practically rigid
+% body. Likewise, gripping maps to friction forces, and springiness maps
+% to Hooke's law. This suggests that applicability can depend directly on
+% our intuitive knowledge---knowledge that exists prior to any formal
+% scientific training (Smith et al., 1993, p. 144).
+%
+% The in-pieces approach to modeling cognition has been used, among other
+% things, to explain how experts reason about fractions and decimals
+% (Smith et al., 1993), how students reason about forces in physics
+% (diSessa \& Sherin, 1998; diSessa, 1993; Hammer, 1996; Sherin, 2001),
+% how students construct and evaluate algebraic representations of
+% physical situations (Izsák, 2004), and how knowledge transfers across
+% contexts (Hammer et al., 2005; Wagner, 2006). Because its starting
+% assumption is that knowledge is fragmented, knowledge-in-pieces can
+% account for wide variations of how people---particularly novices---use
+% knowledge on a moment-to-moment basis. In other words, because it
+% assumes knowledge is local, it can still explain the kinds of
+% globally-inconsistent ways people might reason about physical situations
+% (diSessa, 1993). As a framework, an in-pieces approach ultimately argues
+% that models of concept replacement and good/bad criteria for knowledge
+% should be supplanted by a learning model of alignment/refinement of
+% prior knowledge and the consideration of knowledge as
+% productive/unproductive.
+%
+% Hammer and colleagues have worked to extend the in-pieces approach to
+% explain how students' epistemological activity---how they orient toward
+% knowledge and knowing in a context (Hammer et al., 2005; Hammer \& Elby,
+% 2002, 2003). Specifically, those authors use two core theoretical
+% constructs to explain students' stances toward knowledge and knowing:
+%
+% \begin{itemize}
+% \item
+%   \emph{Epistemological Resources} are the epistemological equivalent of
+%   diSessa's phenomenological primitives (p-prims). Resources, the
+%   authors propose, are the atomic units involved in how people cognize
+%   about the source of knowledge, the nature of knowledge, and
+%   epistemological activities (Hammer \& Elby, 2003; Louca, Elby, Hammer,
+%   \& Kagey, 2004).
+% \item
+%   \emph{Epistemological Frames} are the emergent result of subsets of
+%   resources acting in concert. Drawing from both Goffman's (1974)
+%   sociological notion of frame as structures of expectations and
+%   subsequent work on framing in discourse (Tannen, 1993),
+%   epistemological frames are a participant's local answer to the
+%   question ``what is it {[}specifically, what knowledge activity{]}
+%   that's going on here'' (Goffman, 1974, p. 8).
+% \end{itemize}
+%
+% Resources can frames can interact in activity settings to produce
+% larger-scale patterns called ``epistemological coherences'' (Rosenberg,
+% Hammer, \& Phelan, 2006) where evidence from data suggests that a
+% network of discrete cognitive units can nonetheless give rise to stable
+% cognition.
+%
+% A useful conceptual metaphor for resources and framing is to think about
+% a lecture hall with different sets of lights: a spotlight in the back to
+% highlight a lecturer, chalkboard lights, house lights, and a projector.
+% Each light has an individual brightness, but the lights are only
+% controllable at the per-bulb level. In that sense, they are atomic. But,
+% light patterns interact with one another to create a field of lighting
+% for the room. Thus, lights are a bit like resources, and different light
+% configurations can correspond to different, sociologically-stable uses
+% of the room.
+%
+% \protect\hypertarget{ux5fToc252445958}{}{}Table -- Using room lighting
+% configurations to think about epistemological framing
+%
+% \begin{supertabular}[]{@{}lllll@{}}
+% \toprule
+% \textbf{What is it that's going on?} & Houselights & Spotlight &
+% Chalkboard & Projector\tabularnewline
+% \midrule
+% \endhead
+% \textbf{On-stage monologue} & Low & ON & Low & OFF\tabularnewline
+% \textbf{Presenting slides} & Low & ON & Low & ON\tabularnewline
+% \textbf{Working through equations} & Low & ON & High &
+% OFF\tabularnewline
+% \textbf{Students discussing with each other} & High & OFF & Low &
+% OFF\tabularnewline
+% \textbf{Watching a movie} & Low & OFF & Low & ON\tabularnewline
+% \textbf{Cleaning the room after a movie} & High & OFF & High &
+% OFF\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% To convince yourself of the sociological stability of these
+% configurations, imagine you were watching a movie when suddenly the
+% house and chalkboard lights came up to full intensity. It would jar you
+% out of the experience. You might wonder whether something was wrong: is
+% there an emergency? Should you evacuate?
+%
+% The metaphor isn't perfect. Ontologically, for example, resource and
+% framing theory propose these constructs exist not in the world but in
+% the minds of individuals. But, the metaphor is quite useful for
+% understanding how individual elements---in this case lights---can work
+% in concert to create and sustain a stable frame. And, crucially for a
+% cognitive system, the metaphor lets us account for variation in
+% activity. The same lecture hall can become a window onto one person's
+% thoughts (monologue), a site for instruction (working through equations
+% at the board), a shared space for collaboration (student discussion), or
+% a place in need of repair (cleaning up) simply by varying the
+% intensities of lighting banks in particular ways.
+%
+% An example helps ground this in-pieces approach to epistemology. In
+% Russ, Coffey, Hammer, and Hutchison (2008), the authors describe the
+% situation where an elementary student reasons about why an empty juice
+% box collapses when you suck on the straw. One student gives what the
+% authors deem to be an excellent mechanistic account of why the juice box
+% collapses. But, as the teacher seems to steer the discussion toward
+% vocabulary---in this case, ``pressure''---the student clearly pulls back
+% from her mechanistic reasoning, seems much more diffident, and claims
+% that pressure is hard to explain. That example highlights the disconnect
+% between doing science as knowing vocabulary and doing science as
+% reasoning mechanistically. Moreover, it strikingly highlights that a
+% student who by all accounts produced an excellent explanation of how
+% pressure works was left nonetheless with the impression that pressure
+% was hard to explain.\footnote{In fairness to the teacher, I'm trying to
+%   focus on the result of the interaction and far less on the intent the
+%   teacher might have had. The student still left with a sense that
+%   science might be about vocabulary, even if that's not the view the
+%   teacher would espouse or was trying to enact in the moment.}
+%
+% \subsection{Methods and Theoretical
+% Commitments}\label{methods-and-theoretical-commitments}
+%
+% \subsubsection{Student population, course background, and selection
+% }\label{student-population-course-background-and-selection}
+%
+% This paper focuses on Electrical Engineering (EE) students from Flagship
+% State, a large public research institution in the mid-Atlantic. The
+% students I studied were taking Intermediate Programming, the second
+% semester of an introductory programming course taught in C. The course
+% was exclusively for EE majors and taught by EE faculty, and its
+% enrollees were typically first-year or second-year EE majors. Its model
+% was two 75-minute lectures and a 1-hour teaching assistant-led
+% discussion section for students each week. Course topics included:
+%
+% \begin{itemize}
+% \item
+%   Strings
+% \item
+%   Pointers
+% \item
+%   Dynamic Memory Allocation
+% \item
+%   Testing
+% \item
+%   Debugging
+% \item
+%   Hash tables
+% \item
+%   Trees
+% \item
+%   Linked Lists
+% \item
+%   Abstract Data Types
+% \item
+%   Functional Decomposition
+% \end{itemize}
+%
+% Students in the course had varying degrees of experience with
+% programming. Partly, that variation is because students with Advanced
+% Placement (AP) Computer Science credit could place out of Basic
+% Programming, the first semester C course. So, some students in
+% Intermediate Programming came from technical magnet schools where they
+% may have already had one or more years of programming in multiple
+% languages, while other students may have been first-time programmers
+% with only one semester of programming experience: Basic Programming.
+%
+% I studied the same course, Intermediate Programming, during the fall
+% 2011 semester and the spring 2012 semester. During fall 2011 I
+% ethnographically observed over 50\% of the course lectures, and during
+% spring 2012 I sat in on several discussion sections. At the beginning of
+% each semester I solicited student participants for semi-structured
+% clinical interviews. Of the students who responded to the solicitation,
+% I scheduled interviews opportunistically with students given my
+% resources as the sole researcher on the project. In total, I worked with
+% a cohort of 6 students in fall 2011 and 4 students in spring 2012.
+%
+% My specific interest was in how students design computer programs. By
+% that, I mean I wanted to know not just how students program, but whether
+% and how they structured programs, how they did (or did not) try to
+% incorporate modularity into their programs, and how the final form of a
+% program reflected what they had learned about how to manage complexity
+% through software. In the sections that follow I outline the palette of
+% methods I used to pursue those questions.\footnote{The language in
+%   sections 3.3.2 and 3.3.3 is taken with slight modification, from
+%   ``Studying students' early-stage software design practices,'' a paper
+%   I authored, along with William Doane, and submitted to the 2014
+%   International Conference of the Learning Sciences (ICLS).}
+%
+% \subsubsection{Studying design as a complex phenomenon of disciplinary
+% practice}\label{studying-design-as-a-complex-phenomenon-of-disciplinary-practice}
+%
+% Practicing professionals make complex use of talk, gesture, and
+% representational artifacts in physics (Gupta, Hammer, \& Redish, 2010;
+% Kaiser, 2005; Ochs, Gonzales, \& Jacoby, 1996); chemistry (Stieff,
+% 2007); field biology (Hall, Stevens, \& Torralba, 2002); civil
+% engineering (Stevens \& Hall, 1998); structural engineering (Gainsburg,
+% 2006); mechanical and electrical engineering (Bucciarelli, 1994;
+% Henderson, 1999); and architectural design (Hall et al., 2002; Hall,
+% 1999). Given that:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   Science and engineering education research has made progress by
+%   looking for continuities in how novice learners develop disciplinary
+%   practices (Gupta et al., 2010; Hall \& Stevens, 1995; Smith et al.,
+%   1993; Stevens \& Hall, 1998), and
+% \item
+%   Emerging research on software engineering reveals that early-stage
+%   software design involves complex inscriptional, discursively, and
+%   epistemic practices,
+% \end{enumerate}
+%
+% it seems striking that there is no contemporary body of research,
+% comparable to studies of expert practice, that looks at students'
+% software design practices. In other words, we have every reason to
+% believe that expertise in software design involves complex practices,
+% but little (if any) research on what productive capacities students have
+% for them. Finding those productive design capacities requires a shift
+% from questions such as \emph{how can we assess and mitigate students'
+% difficulty in programming?} toward questions such as \emph{how do
+% students learn and display evidence of design thinking in programming?}
+%
+% Rephrasing research questions asked of professional software engineers
+% (Petre et al., 2010, p. 533) and instead treating \emph{students} as
+% designers, I ask:
+%
+% \begin{itemize}
+% \item
+%   What do students actually do during early stage software design work?
+% \item
+%   What does students' exploratory design thinking look like?
+% \item
+%   How do students communicate?
+% \item
+%   What sorts of drawings do students create when they design software?
+% \item
+%   What kinds of strategies do students apply in exploring the vast space
+%   of possible software designs?
+% \end{itemize}
+%
+% \subsubsection{Deploying methods to capture the complexity of
+% early-stage design
+% work}\label{deploying-methods-to-capture-the-complexity-of-early-stage-design-work}
+%
+% The methods below form the core of my developing program to study
+% students as software designers. None of the methods below are new; all
+% have been used in prior educational research. What \emph{is} new, I
+% believe, is the opportunity to combine them all under the umbrella of
+% understanding what happens when students design software. For each of
+% the four students in my spring 2012 cohort I captured multiple streams
+% of data.
+%
+% \begin{itemize}
+% \item
+%   I collected their \textbf{code history}. By this, I mean I preserved
+%   frequent snapshots in time of what students' code looks like. Research
+%   has already shown code snapshotting to be a useful method for
+%   understanding large-scale patterns of student error (Jadud, 2006;
+%   Rodrigo et al., 2009; Spacco, Hovemeyer, et al., 2006). And, the
+%   resolution of snapshots is extremely fine: Spacco et al. are able to
+%   capture the contents of a file each and every time a student saves it
+%   to disk. But, that research takes an aggregate view: it identifies
+%   large-scale error patterns at the expense of detailed naturalistic
+%   understandings of why students make those errors. Moreover, it's
+%   primarily used to identify what \emph{mistakes} students make, which
+%   is distinctly different from a research orientation that considers the
+%   negative \emph{and} positive consequences of students' design choices.
+%   A currently untapped advantage of collecting code history data, then,
+%   is that while we historically use it to aggregate \emph{across
+%   programmers} it nevertheless also gives us fine-grained individual or
+%   team-based histories of how designs evolve.
+% \item
+%   I conducted \textbf{clinical interviews} with them. Clinical
+%   interviews have proven historically useful in understanding the
+%   substance of students' knowledge and the nature of conceptual change
+%   (diSessa \& Sherin, 1998; Duckworth, 2006; Ginsburg \& Opper, 1988;
+%   Ginsburg, 1997). Crucially for Computer Science Education (CSEd)
+%   research, clinical interviews can tell us about students'
+%   epistemologies---how they view knowledge and knowing in a discipline
+%   (Hofer \& Pintrich, 1997)---which can affect how they approach and
+%   adopt that discipline's practices (Hammer, 1989, 1994; Lising \& Elby,
+%   2005). Moreover, because my interviews were videotaped and analyzed
+%   from perspectives of interaction analysis (Goodwin, 2000; Jordan \&
+%   Henderson, 1995), they offer rich evidence of the substance of
+%   students' design practices
+% \item
+%   I analyzed their in-interview \textbf{inscriptions} when they designed
+%   --- what they wrote, how they wrote it, and how those writings got
+%   used. Evidence from both science studies (Hall et al., 2002;
+%   Henderson, 1999; Hutchins, 1995a; Kaiser, 2005; Latour, 1990; Ochs et
+%   al., 1996) and educational research (Hall \& Stevens, 1995; Lehrer,
+%   Schauble, Carpenter, \& Penner, 2000; Stevens \& Hall, 1998)
+%   highlights the centrality of inscriptions to disciplinary practice in
+%   science and mathematics. Ethnomethdological data from studies of
+%   professional software engineers supports the same result:
+%   inscriptional practice is central to how professional engineers design
+%   software (Rooksby \& Ikeya, 2012; Rooksby, 2010). And, since part of
+%   the inscriptional environment when designing software is the computer
+%   itself, I captured and analyzed what happened on-screen as students
+%   design programs.
+% \item
+%   I paid close attention to students' in-interview \textbf{gestures}.
+%   Perspectives of gestural analysis hold that gestures can not only
+%   support or extend thinking, they can also communicate entirely
+%   different information than what's being said (Goldin-Meadow, 2003).
+%   Moreover, perspectives from cognitive anthropology and embodied
+%   cognition studies argue that bodily motion \emph{is itself} cognition
+%   (Hall \& Nemirovsky, 2012; Hutchins, 1995a; Nemirovsky, Rasmussen,
+%   Sweeney, \& Wawro, 2012). For example, when students describe a part
+%   of code by moving their hand across an imaginary row of items and
+%   tapping each item, their bodies convey information we can interpret
+%   about how they understand iteration.
+% \end{itemize}
+%
+% Figure 3-1 presents a visual overview of some of these methods and modes
+% of analysis. In particular, the first (top-left) panel depicts how we
+% deploy these data collection methods during a clinical interview:
+%
+% \begin{itemize}
+% \item
+%   a voice recorder captures speech (often a redundancy in case another
+%   recording device fails)
+% \item
+%   a LiveScribe Pulse pen digitizes written inscriptions, allowing us to
+%   play back what was written in time
+% \item
+%   a videocamera records data for knowledge analysis, interaction
+%   analysis, and gestural analysis,
+% \item
+%   an in-interview computer tracks code history and its screen-capturing
+%   software records real-time activity.
+% \end{itemize}
+%
+% \includegraphics[width=6.25556in,height=4.20000in]{media/image4.jpeg}
+%
+% \protect\hypertarget{ux5fToc252445967}{}{}Figure ‑ -- An comic-based
+% overview of my methods for capturing students' design practices
+%
+% \subsubsection{Developing a revelatory case of students' early-stage
+% design
+% work}\label{developing-a-revelatory-case-of-students-early-stage-design-work}
+%
+% In the empirical work that follows I explore what productive knowledge
+% and resources students have for structuring programs. I begin with
+% Lionel, a male student in Intermediate Programming, fall 2011. First, I
+% propose the idea that Lionel connects ``designerly'' (Archer, 1979)
+% stances across contexts. I analyze his retrospective account of
+% modifying a bike and compare it to his retrospective accounts of how he
+% programs in Intermediate Programming. Then I detail Lionel's multi-stage
+% cascading approach to structuring a program solution in an interview.
+% The argument throughout is that from bike design to code design Lionel
+% evinces knowledge and stances that productively help him solve problems.
+%
+% Next I transition to Rebecca, first explaining the epistemological
+% difficulty she felt when coding, then showing how she circumvented that
+% block by structuring a program piece using gesture and talk. I argue
+% that what Rebecca and Lionel have in common are resources for handling
+% design obstacles as they emerge. Where the two students differ is in
+% their in-the-moment judgments about what sorts of knowledge and
+% activities are appropriate to work through a design problem.
+%
+% I argue that taken together, Rebecca and Lionel constitute what Yin
+% (2009) calls a ``revelatory case,'' by which I mean my investigation
+% arises from ``an opportunity to observe and analyze a phenomenon
+% previously inaccessible to social science inquiry.'' In my case, I argue
+% the phenomenon I'm observing is what engineering students do in the
+% early stages of design work on a program. Its inaccessibility is
+% demonstrated by the almost complete paucity of studies that draw on
+% real-time (e.g., video records) of how students begin work on a complex
+% program. Contrary to being ``representative'' or ``typical,'' my
+% revelatory case does not aim to generalize. Rather, its value is in
+% phenomenological richness. As Yin (2009) writes, and I think others
+% including Erickson (1986) would agree, such case studies are worth doing
+% ``because the descriptive information alone will be revelatory.''
+%
+% \subsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188553}{}{\protect\hypertarget{ux5fToc252445926}{}{}}Lionel's
+% approach to design and to
+% programming}{Lionel's approach to design and to programming}}\label{lionels-approach-to-design-and-to-programming}
+%
+% \subsubsection{While modifying a bike Lionel decomposed tasks, explained
+% solutions to himself, and created intermediate design
+% representations}\label{while-modifying-a-bike-lionel-decomposed-tasks-explained-solutions-to-himself-and-created-intermediate-design-representations}
+%
+% Prior to entering my study, Lionel had already earned a Bachelor's
+% degree in Business Information Technology at a different university. We
+% began the interview by discussing why he re-matriculated to earn an
+% electrical engineering degree.
+%
+% Interviewer: How did you, um end up coming out of your business program
+% and your business experience and deciding that electrical engineering
+% was what you wanted?
+%
+% Lionel: Um, well, since I was Business Information Technology, basically
+% all my job was doing was writing SQL code and, you know, short
+% statements like that. I also worked at a help desk for---I guess I
+% worked at Department of Homeland Security. And, pretty much besides the
+% SQL statements I worked with this one program and whenever somebody had
+% a problem with it they'd come to me. And basically that---that's what my
+% whole degree was---was going down that track and then that's really not
+% what I wanted to do: just sitting at a desk, you know people coming to
+% me being like ``oh, you gotta do this for me.'' So, I was like, alright.
+% Or, I have to fix a problem for them, pretty much. And that's really
+% what I didn't want to do. I wanted to do something more hands-on where I
+% could, you know, I'd be given a project where I'd do something on my
+% own, create something on my own. And, um, also my Dad and my uncle were
+% both electrical engineers. And my brother became a computer scientist.
+% Cu---so I've always been around, you know, that type of environment,
+% where my dad's always building something and, you know, I always see it
+% and I'm like, you know, that's very interesting. I'd like to do
+% something like that too. So it's really just uh, a mixture of bad
+% experience with my previous job and being around company that likes to
+% do that sort of stuff. (Interview 1 of 1, October 17, 2011)
+%
+% Lionel apprenticed building things with his father first, taking on
+% small, highly-directed roles in projects.
+%
+% Lionel: Be-before I started my degree, you know, I really didn't know
+% how to do any of this stuff. So, I would just kinda help my dad out a
+% little bit. He'd be like ``Oh, here,'' you know, ``to, to do this you
+% have to do exactly this.'' You know, ``solder this here,'' or, you know,
+% ``drill this,'' blah blah blah. He'd tell me exactly what to do. So, I
+% just kind of followed along. (Interview 1 of 1, October 17, 2011)
+%
+% Presumably in those projects Lionel was at a periphery and direction
+% came from his father. More recently, though, Lionel had struck out on
+% projects of his own. As Lionel said, ``now that uh, my dad kinda showed
+% me how to do certain typical types of things, I've tried to become
+% \textbar{}\textbar{}more adventurous\textbar{}\textbar{} \textbar{}\{air
+% quotes\}\textbar{} /Sure/ and do things on my own'' (Interview 1 of 1,
+% October 17, 2011).
+%
+% One of Lionel's recent adventurous projects was adding a motor to a
+% beach cruiser bicycle. The gist of the project was simple: using a
+% low-priced motor he found online, Lionel realized he could mount the
+% motor to an inexpensive bicycle and make a motorized bike. With one
+% exception, Lionel was solely responsible for the project from start to
+% finish.\footnote{Lionel told me that the only part his father helped him
+%   with was showing him how to use a Dremel power tool to smooth, sand,
+%   and cut edges (Interview 1 of 1, October 17, 2011).} He bought the
+% parts, he mounted the motor, and he maintained it as a working vehicle.
+% But, in practice the project was not straightforward. In the next
+% subsections I show at length how Lionel addressed the emergent
+% challenges of modifying\footnote{Often called ``modding'', for short.}
+% his bike with a motor. My analysis focuses on three features of Lionel's
+% story that I argue carry over to his programming approach:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   When faced with ambiguity, Lionel worked to decompose large tasks into
+%   smaller subtasks he could understand.
+% \item
+%   Throughout the task, Lionel was metacognitive about the state of his
+%   design. His constant refrain was ``\emph{how} am I gonna make this
+%   work?''
+% \item
+%   Before committing work and materials to a solution, Lionel created
+%   intermediate designs whose suitability he could evaluate.
+% \end{enumerate}
+%
+% I argue in later sections that these three features of Lionel's
+% approach---decomposing subtasks, explaining procedures to himself, and
+% explicitly planning prior to committing time and resources to a final
+% product---have strong resonances with the approach Lionel took to while
+% programming.
+%
+% \paragraph{Lionel had to decompose Step 1, which simply said ``Install
+% the
+% Engine''}\label{lionel-had-to-decompose-step-1-which-simply-said-install-the-engine}
+%
+% Lionel bought an inexpensive motor online with the plan of attaching it
+% to a cheap bike. As he discovered, though, frugality came with a hidden
+% cost: poor instructions. Attaching the motor to the bike was not at all
+% a straightforward process.
+%
+% Lionel: it was funny. When I, when I got the motor /Mmmhmm/ Um, oh,
+% it---like I said it was like cheaply made and whatnot and it came with
+% instructions, but um, no exaggeration at all, step 1 said ``install the
+% engine''. And that was it. \{Laughs\}
+%
+% Interviewer: Huh
+%
+% Lionel: And then step 2 was, um, I guess it was somethin like ``put the
+% gas tank on.'' It really didn't tell me anything. It was kinda like
+% ``OK, those are the obvious steps, but *how* do I do that? So I kinda
+% had to figure stuff out on my own. /So/ yup
+%
+% Interviewer: Did it---how did it end up going? Was it a lot of trial and
+% error? Like, did they even have pictures of how it was supposed to look
+% when it /There, there/ was installed?
+%
+% Lionel: There are maybe two or three pictures, but they were black and
+% white pictures printed out on like a, I guess like a crappy printer. So
+% they were kinda hard to see. Um, so I guess my first step was to go look
+% online at better quality pictures. Unfortunately nobody really---I was
+% hopin to find you know, like a YouTube video: this is how I did it.
+%
+% Interviewer: Mmmhmm.
+%
+% Lionel: I couldn't really find somethin like that, so I just looked at
+% pictures. (Interview 1 of 1, October 17, 2011).
+%
+% Lionel recognized that having an attached motor is an obvious goal
+% state, but that goal state was poorly-specified by ``crappy pictures''
+% that were hard to interpret. Moreover, the states in between start and
+% goal were undocumented and not obvious. Lionel's subsequent search for
+% materials that showed a step-by-step progression came up empty. Faced
+% with a poorly-specified goal-state and no intermediate guidance,
+% Lionel's response was to force himself to think carefully about how he
+% would proceed.
+%
+% \paragraph{In making modifications Lionel frequently asked himself ``how
+% am I gonna make this
+% work?''}\label{in-making-modifications-lionel-frequently-asked-himself-how-am-i-gonna-make-this-work}
+%
+% Faced with the challenge of interpolating between ``crappy pictures,''
+% Lionel thought hard about what to do.
+%
+% Lionel: I spent, I guess the, I guess the first five hours I just spent
+% kinda staring at it like, ``hmm, *what* am I gonna do?'' Like, ``*how*
+% am I gonna make this work?'' (Interview 1 of 1, October 17, 2011)
+%
+% Ultimately, he was struggling with more than just poor instructions.
+% Even if he did figure out how the motor was supposed to be mounted, he
+% still faced another problem: the motor was designed for a mountain bike.
+%
+% Lionel: I had to make a few modifications on my own. Because I guess
+% it's built for uh, a mountain bike. And the tubes on mountain bikes are,
+% y'know like *that* thin \{makes circle with right thumb and index
+% finger\}, but on my beach cruiser there's one tube \{widens thumb and
+% forefinger\} that's pretty thick where I had to mount the engine.
+%
+% Lionel: And since the engine mount was built for a motorized bike it was
+% really small. So I kinda had to like devise my own plan on *how* to,
+% y'know, add a bigger mount.
+%
+% Interviewer: Mmmhmm
+%
+% Lionel: So, that was, that was my first problem. That was, y'know the
+% first step was to put the motor on I'm thinkin, well, how am I gonna put
+% the motor on if, y'know the, the specs don't even align. So, I guess
+% spent y'know five hours just thinkin', like, ``what am I gonna do?''
+%
+% Lionel: Finally I came up with a plan that seems to be working. So. And
+% then, I just kinda went from there. I'd put it together, look at the
+% picture and say ``OK, this piece looks like that piece, and y'know I
+% guess it looks like it attaches that way, so I'm gonna try that.'' And
+% just put it together and, y'know if it fit, it fit. If it didn't, then
+% I'd, y'know I'd be like ``OK, maybe that was wrong. Lemme, lemme work on
+% somethin' else on it.'' So. (Interview 1 of 1, October 17, 2011)
+%
+% Asking himself \emph{how} to do something helped bring sub-problems into
+% focus. Within a larger mod task (viz., adding a motor to a bike), Lionel
+% in effect entered a nested task: modding a mountain bike engine mount to
+% fit a beach cruiser. Modding the mount itself was entirely undocumented,
+% so Lionel's solution in part was to look for affordances in the physical
+% structure of parts to see what he could attach. Crucially, the
+% environment afforded Lionel rapid, iterative feedback. It wasn't hard to
+% tell when a design choice worked because ``if it fit, it fit.''
+%
+% \paragraph{Lionel used intermediate representations to consider
+% alternative
+% designs}\label{lionel-used-intermediate-representations-to-consider-alternative-designs}
+%
+% Lionel grew increasingly frustrated as he saw how difficult it would be
+% to mount the motor. Crucially, part of his tactic around that
+% frustration was to resolve to keep trying and thinking of possible
+% solutions.
+%
+% Lionel: So, I guess I had to give myself, like a few hours just to cool
+% down and then actually think ``OK, well maybe, you know I've already
+% bought it, so, I'm gonna continue on with it. So I had to think, you
+% know, OK, what's my next step. *How* could I make this work? You know,
+% I---I thought of a few different ideas. You know, I'd think of one idea
+% and think ``OK, would this work? It might, but let me think of another
+% idea.'' So I'd come up with another idea and say, ``would this work?''
+% You know, so once I got a few ideas I would just kind of, in my head
+% picture it. Say, ``OK, this is how it would mount, or this is how it
+% would mount. Which one would be safer? Or will one of these not even
+% work?''
+%
+% Interviewer: Right.
+%
+% Lionel: So, I guess after, you know, awhile of deciding which one was
+% better, you know, which one would be easier and safe at the same time
+% /mmhmm/ so then, that's when I made my decision, and just kinda, went
+% for it to see if it would work. And luckily it did \{laughs\}.
+% (Interview 1 of 1, October 17, 2011)
+%
+% Before committing to a design, Lionel forced himself to evaluate
+% alternatives first. Naively, he could have gone with the first idea that
+% came to him. Instead, even if he thought an idea might work he pushed
+% himself to consider others first. Then, with a collection of possible
+% designs he evaluated what he saw as their suitability in his head. He
+% applied criteria as he evaluated, trying to balance the ease with which
+% a mounting scheme might work with a concern for the safety of the
+% user---in this case, himself. Again, at least in Lionel's telling, those
+% steps happened \emph{before} committing to a final implementation. In
+% other words, Lionel generated and evaluated multiple configurations with
+% an eye toward both the builder of and user of those configurations
+% before settling on a path to take.
+%
+% In sum, I note three key features of Lionel's approach to the design
+% challenges of modifying his bike. First, he repeatedly asked himself
+% ``how am I gonna do this,'' forcing himself to think in terms of a plan
+% or procedure to fill in the gaps between the instructional pictures.
+% Second, he proposed ideas, tried to evaluate their feasibility, and
+% continued pressing for alternate solutions (``would this work? It might,
+% but let me think of another solution''). Third, so much of his planning
+% took place before he ever mounted the motor. By his account, he fiddled
+% with pieces, envisioned multiple ways the engine could mount, and
+% evaluated their feasibility before ever taking concrete steps to mount
+% the motor.
+%
+% \subsubsection{Lionel's approach to programming resonates strongly with
+% the design stance he held when working on his
+% bike}\label{lionels-approach-to-programming-resonates-strongly-with-the-design-stance-he-held-when-working-on-his-bike}
+%
+% \paragraph{Lionel thinks about steps at a
+% chalkboard}\label{lionel-thinks-about-steps-at-a-chalkboard}
+%
+% Later in the interview I asked Lionel to compare his experiences in
+% Basic Programming with those he was having in Intermediate Programming.
+% His response revealed a great deal about how he starts projects in
+% Intermediate Programming: he starts at a chalkboard.
+%
+% Lionel: Well, {[}Basic Programming{]} I already learned the basics of
+% the class so I know the basics of where to start my code. And then,
+% obviously the projects in this class {[}Intermediate Programming{]} are
+% harder than last class, but I have the basics of, ``OK, well, I know how
+% a program runs. So how can I translate that into my program, for, for
+% {[}Intermediate Programming{]}?'' And, yeah, so I guess I'll sit down,
+% like, yeah, and I'll have like a chalkboard and I'll write things out,
+% and say OK well first, you know, \{sweeps hands{]} not even think about
+% the code. Just, in \{air quotes\} English words, so you know, my
+% project's gonna do this. You know, first I need to make this
+% calculation, then I'm gonna print this. And then I'm gonna make this
+% calculation. Et cetera. And I'll sit down and write that out, in the
+% order that it needs to be done. And then from there I'll go back and
+% think, ``OK, what's the code, to, you know, make this calculation or
+% print that?''
+%
+% Interviewer: So, when you start a project, you don't start at a
+% computer, it sounds like you kinda start at your chalkboard.
+%
+% Lionel: Right. Exactly. Yeah, cuz I've tried to start a project at a
+% computer, and you know I'll write a---I'll start writing a few lines of
+% code, but then I just really zone in on what I'm doing and kinda lose
+% sight of the whole picture.
+%
+% Lionel: So then I'll write this code for, you know, this one little
+% section of the project, and then after I've done that I'll look back and
+% say ``Oh, yeah, I need to write the whole project,'' but then this
+% little code doesn't actually fit into my project, so it's kinda, you
+% know I guess I get tunnel vision when I start at a computer. (Interview
+% 1 of 1, October 17, 2011)
+%
+% For Lionel, the chalkboard is a pivotal object in his design activity.
+% It holds the top-down ``English words'' description of his program. In
+% so doing, it serves as the earliest durable representation of his
+% program's intended hierarchy and relationships. That function is
+% important, because without it Lionel is apt to ``lose sight of the whole
+% picture'' of what his code is supposed to do. Moreover, as the next
+% section reveals, the board continues to help Lionel orient his activity
+% as he digs into the specifics of how his program will work.
+%
+% \paragraph{Lionel copies pseudo-code into the computer even though it
+% won't
+% compile}\label{lionel-copies-pseudo-code-into-the-computer-even-though-it-wont-compile}
+%
+% As Lionel described the role of the chalkboard, I initially thought
+% there was a distinct separation between the chalkboard and the computer
+% as activity objects. The chalkboard, I assumed, was where Lionel
+% sketched out design plans while the computer was where he wrote code.
+% When I probed Lionel based on my hunch, I discovered Lionel's activity
+% wasn't cleanly separable into design on the board / code on the
+% computer.
+%
+% Interviewer: Did you, when you start thinking about the code, do you end
+% up usually filling it in on your board, or will you actually go straight
+% over to the computer and do it?
+%
+% Lionel: Um, on my board, I'll do, I guess like
+% \textbar{}\textbar{}pseudo\textbar{}\textbar{} \textbar{}\{air
+% quotes\}\textbar{} code /mmhmm/ uh, you know, first on the board it
+% would only say ``this function calls this function, this function calls
+% this function.''
+%
+% Interviewer: Mmmhmm
+%
+% Lionel: So then on the board I'd write down, ``alright, well this
+% function'' um, I dunno, for instance I'll, I'll say you know, I'm workin
+% with a function that checks to make sure that it's on the---that I'm
+% moving *on* the board between spaces 1 and 24.\footnote{At that point in
+%   the semester, Lionel's class was working through a Backgammon project,
+%   so the spaces and moves he's referring to pertain to programming a
+%   basic Backgammon game.} So then, you know I'll write down like,
+% pseudocode, write ``OK, well I'm gonna start. I'm gonna have a for-loop
+% that will go from 1 to 24, and then here I'll, you know I'll have an
+% if() statement that says if it's between you know, 1 and 24 it's valid,
+% if not it's invalid.'' So, that'd be my pseudocode, and then when I
+% actually---I'd, I'd try to write that out for the entire program, or as
+% much as I could.
+%
+% Interviewer: Filling out the tree, you mean.
+%
+% Lionel: Right, fillin out the tree. And then I'd go to my computer, so
+% that at least I already had a baseline of what to work with. And then,
+% on my computer I'd, you know I'd write out the pseudocode, I'd ac---I'd
+% literally write out the pseudocode, even though obviously it wouldn't
+% compile and actually work.
+%
+% Interviewer: Mmmhmm
+%
+% Lionel: So, then, from the pseudocode, as, as I read down on my computer
+% what the pseudocode said, I'd actually try to write the actual code that
+% would work. (Interview 1 of 1, October 17, 2011)
+%
+% Lionel blurred my initial chalkboard / computer distinction in two ways.
+% First, his overall ``English words'' plan of functions calling functions
+% evolved on the chalkboard to include computational idioms, including
+% if-statements and for-loops. But, then, according to Lionel, he would
+% copy that pseudo-code as-was into the computer, knowing full well it
+% wouldn't compile. These observations---that Lionel writes computational
+% pseudo-code on the board, then copies that same code into the
+% computer---may seem trivial, but I contend they aren't.
+%
+% Because of the choices Lionel makes, neither chalkboard nor computer has
+% a distinct, privileged kind of syntax. Rather, ``English words'' and
+% pseudo-code can peacefully co-exist on the chalkboard, and non-working
+% pseudo-code gets deliberately typed into the computer before it's
+% expanded out into ``actual code.'' Put another way, Lionel is capable of
+% distinguishing between ``English,'' ``pseudo-code,'' and ``actual
+% code,'' but all three forms of expression are a part of his design
+% process. It's perfectly acceptable for him to have something other than
+% ``actual code'' exist in intermediate representations of his design.
+%
+% \subsubsection{Gestural, inscriptional, and verbal in-interview evidence
+% reveals Lionel's resources for structuring a
+% program}\label{gestural-inscriptional-and-verbal-in-interview-evidence-reveals-lionels-resources-for-structuring-a-program}
+%
+% To probe Lionel's retrospective account of how he programs, I gave him
+% an in-interview task to work on. In this section, I argue Lionel's
+% development process on that task moved rather linearly through what I
+% call a ``representational cascade.''\footnote{As I'll explain later, I
+%   think it would be a mistake to assume that all development activity
+%   proceeds linearly and top-down in the manner Lionel's does. Not only
+%   do I think Lionel's clean linearity is rare, I also think it's not
+%   normative. So, my point here isn't to stress that Lionel moved
+%   linearly, monotonically through the cascade. Rather, my point is to
+%   show that he moved between different representational levels \emph{at
+%   all}.} My schematic for that cascade --- along with the features and
+% affordances in each layer of the cascade --- is depicted in Figure 3.
+% What I show here is that actually, the final code Lionel produces is in
+% fact one representation of a procedure that actually existed in
+% different forms and modalities before it became the final code in Figure
+% 3.
+%
+% In the analysis that follows, I explore two of those modalities---verbal
+% description and written pseudo-code---and explain why each is crucial to
+% understanding the final produced code. Ultimately, I argue there is a
+% need to explain \emph{how} Lionel creates and moves fluidly between
+% these markedly different cascade layers. Specifically, I attempt to
+% answer the question of \emph{Why does he see the upper layers (verbal
+% code, hand-written pseudocode, and handwritten source code) as
+% legitimate activity?} My candidate answer is that Lionel's approach to
+% programming is in part stabilized by epistemological resources that
+% allow him to treat the upper layers of the cascade as valid, productive
+% knowledge expressions in programming. Such an answer is strongly in line
+% with the findings from sections 3.4.1, where we see Lionel deliberately
+% making use of other intermediate design representations.
+%
+% \includegraphics[width=5.74444in,height=4.31111in]{media/image5.jpeg}
+%
+% \protect\hypertarget{ux5fToc252445968}{}{}Figure ‑ -- An overview of
+% Lionel's representational cascade and the features and affordances of
+% each representational layer.
+%
+% \paragraph{The range-finding prompt and Lionel's final source
+% code}\label{the-range-finding-prompt-and-lionels-final-source-code}
+%
+% About 35 minutes into our interview, I gave Lionel the programming task
+% reproduced below:
+%
+% Suppose you want to write a program that finds the range of a set of
+% numbers. So, the program would take 10 numbers as input, then it would
+% compute the range as the difference between the largest number and the
+% smallest number.
+%
+% How would you write such a program? (Interview prompt, October 10, 2011)
+%
+% The prompt was deliberately vague about certain constraints. It did not
+% prescribe \emph{how} the program should get input (Should the programmer
+% expect the user might type it in, or feed in a formatted file?). It also
+% did not prescribe how the program should provide the result (should the
+% programmer print the result on screen? Store it to a file? Return the
+% result to be used by another function?) The reasons for vagueness were
+% in part because I wanted to see what assumptions students might make
+% during their design process. Some solutions, I reasoned, might decouple
+% the procedures of getting input, computing range, and returning output.
+% Others might simply hard-code a procedure for getting keyboard-based
+% input. I wanted to prescribe no specific approach up front if I didn't
+% have to. I have reconstructed his original source code from
+% screen-capture footage of his in-interview programming:
+%
+% \includegraphics[width=5.74444in,height=7.11111in]{media/image6.emf}
+%
+% \protect\hypertarget{ux5fToc252445969}{}{}Figure ‑ -- Lionel's final
+% source code for the range-finding prompt
+%
+% Let's explore some top-level features of Lionel's code. First, Lionel
+% chose to have a user enter text directly using the keyboard (lines
+% 9--12), and his program displays the range result as printed text (lines
+% 28--29). His strategy for finding the maximum and minimum is to
+% implement a procedure that iterates through the array of numbers and
+% compares the current number to the globally stored max and
+% min.\footnote{I think of this as a ``king of the court'' procedure. The
+%   current value stays on as an extreme if and only if it beats the
+%   current ``king'' for that extreme. Otherwise, the challenger is
+%   replaced and the contest resumes with a new challenger until there are
+%   no more challengers.} If the current number (array{[}i{]}) ``beats''
+% the global min (by being lower than min), its value replaces that of the
+% global min. Similarly, if the current number exceeds the global max, its
+% value replaces that of max. Before the procedure starts, Lionel assigns
+% the first number of the array as the value for both extrema.
+%
+% \paragraph{Lionel's verbal description of the
+% program}\label{lionels-verbal-description-of-the-program}
+%
+% Lionel chose to start the task on pencil and paper. After scribbling a
+% line about scanning input into an array (which I discuss in a later
+% section), Lionel stopped and began explaining out loud to me.
+%
+% Lionel: so I'm thinkin, my general concept is I'm going to parse through
+% this array, check each number /Mmmhmm/ and I'll have two variables, min
+% and max, and, first I'll set both min and max to the first variable in
+% array, er, yeah first number in array /Mmmhmm/ and, I'll parse through
+% the rest of the array, checking to see if,
+%
+% Lionel: Umm, you know if---I'll s---I'll start with max, for example.
+% /Mmmhmm/ And I'll check to see if the number in that section of the
+% array is bigger than max. /uh-huh/ And if it is, I'll set max to that
+% new number. /OK/ And then I'll do the same thing for minimum. And then
+% at the very end, I'll have my minimum and max and then I'll just
+% subtract `em.
+%
+% Interviewer: OK
+%
+% Lionel: So that's my main concept. (Interview, October 17, 2011)
+%
+% This verbal description is in some ways bound by the least stringent
+% rules of expression. If we're hard-nosed about syntax, Lionel's speech
+% stops and starts (line 5 into line 6), which leans on the unspoken
+% understanding that ideas in talk are under negotiation. Stopping
+% mid-sentence and leaving that sentence unfinished is a valid and
+% acceptable move in conversational space. Moreover, there is a strong
+% mapping between Lionel's temporal language and the intended
+% flow-of-control for the program. ``First I'll set,'' ``then I'll do,''
+% ``and then at the very end'' all describe when instructions will be run.
+%
+% Finally, Lionel's instructions begin at the high level, and in some
+% cases implementation details are completely left out. For example,
+% Lionel does not specify precisely how he'll ``parse through the array.''
+% At this stage of the cascade, it seems sufficient for him to explain
+% \emph{that} he plans to parse through, but now how. What Lionel
+% \emph{does} do is gesture while describing ``parsing''. This gesturing,
+% I argue, creates a concrete representation of iteration and
+% incrementation --- the gestural marker of a loop.
+%
+% \includegraphics[width=5.74444in,height=3.23333in]{media/image7.png}
+%
+% \protect\hypertarget{ux5fToc252445970}{}{}Figure ‑ -- Lionel makes a
+% cycloid/helix gesture as he says ``parse through this array''
+%
+% Contrast how Lionel specifies ``parsing'' through the array with how he
+% describes ``checking'' its contents. First, Lionel explains that he'll
+% ``check each number'' (line 2), decides that he has underspecified how
+% ``checking'' happens (line 5--6), and finally makes ``checking''
+% relatively more explicit (lines 7--8). Once checking for the maximum has
+% been made explicit (lines 7--8), Lionel says ``and then I'll do the same
+% thing for minimum'' (line 9). Line 9 thus becomes the verbal equivalent
+% of a repeated, parameterized procedure. It is, in a sense, the
+% talk-space seed of a function. Conversationally, both participants
+% understand that Lionel saying, ``then I'll do the same thing for
+% minimum'' (line 9) refers to a procedure defined elsewhere in the
+% conversation, not a new set of instructions.
+%
+% \paragraph{Lionel's pseudo-code}\label{lionels-pseudo-code}
+%
+% Immediately following the verbal description above, Lionel begins
+% creating what he calls his ``pseudo-code,'' narrating it as he writes.
+% There are four key features in Lionel's talk (below) that I discuss in
+% my analysis. Below is an overview of each feature and what importance I
+% think it has:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   Lionel iterates over his initial verbal description by fleshing out
+%   some implementation specifics --- structures that further specify
+%   \emph{how} something in a program is to be done.\footnote{It's a
+%     limitation of my writing style that in this document I use
+%     ``iteration'' both to describe aspects of Lionel's design process
+%     and as a formal term of art in computer science. In both cases,
+%     though, the conceptual meaning is the same: the incremental,
+%     rule-based set of transitions to the state of an object.} This
+%   iteration is important because it suggests that his subgoals --- and
+%   criteria for satisficing them --- change from one micro-coherence to
+%   the next.
+% \item
+%   Reconciling what Lionel \emph{narrates} with what Lionel \emph{writes}
+%   reveals peculiar mismatches. The mismatches suggests a productive
+%   capacity: that Lionel can \emph{background} certain details within a
+%   micro-coherence by omitting them from the written code while still
+%   maintaining an understanding of how his program will work. Put another
+%   way: the inscriptional object in a microcoherence captures what Lionel
+%   thinks is important within that microcoherence, but the inscription
+%   does not necessarily represent the totality of his conceptual
+%   understanding.
+% \item
+%   At one point in his code, Lionel writes ``same for min,'' which we can
+%   reasonably infer means he intends to repeat a procedure for
+%   calculating the minimum. That line encapsulates the precursors of
+%   functional abstraction: Lionel not only recognizes procedures that
+%   share the same structure, he can background the specifics (point 2) of
+%   code he repeats.
+% \end{enumerate}
+%
+% Lionel: So it'll be scanf, um, and I'll have my for-loop, and I'll just
+% have my int-i, and I'll go from zero uh, zero to, four actually cuz
+% it'll be zero, one, two, three, four, yeah, zero to four. /Mmmhmm/
+%
+% Lionel: Umm, and I'll have, oh, up top here, to set the, right---right
+% after the scanf /Mmmhmm/, I'll just say min and max equal the
+% first---mm, hiccup, sorry
+%
+% Interviewer: It's OK.
+%
+% Lionel: \{laughs\} the first number in the array. So I'll say min and
+% max equal array of zero. And that's just setting them so that I have
+% something to compare to, I guess. /uh-huh/ I think that will work. Yeah.
+%
+% Lionel: So, so then, as I parse through my array of five variables, I'll
+% say ``if min,'' er no, I'll say ``if array of i, you know of, of that,
+% if, uh, the number in the array that I'm looking at is greater than max,
+% then max equals that number, array of i'' /Mmmhmm/ and then I'll do the
+% same thing, same for min. And then at the very bottom, I'll have, uh,
+% range equals max minus min.
+%
+% Lionel: So that's my pseudo-code, just so that *I* understand what's
+% goin' on in my head.
+%
+% \includegraphics[width=5.75000in,height=3.00208in]{media/image8.emf}
+%
+% First, Lionel's talk iterates over his initial verbal description by
+% fleshing out \emph{implementation details} --- programming structures
+% that further specify \emph{how} something is to be done. Compare here
+% how Lionel instantiates ideas of iteration in Table 1 below.
+%
+% \protect\hypertarget{ux5fToc252445959}{}{}Table -- Comparing Lionel's
+% words and actions across different stages of ``programming'' over the
+% same putative section of his program. I have bolded some
+% language-specific syntax.
+%
+% \begin{supertabular}[]{@{}lll@{}}
+% \toprule
+% \textbf{Design Coherence} & What he says & What he does\tabularnewline
+% \midrule
+% \endhead
+% \textbf{Verbal Description} & ``I'm going to parse through this array,
+% check each number'' &
+% \includegraphics[width=2.22222in,height=1.48889in]{media/image9.jpeg}\tabularnewline
+% \textbf{Narrated pseudocode} & ``I'll have my \textbf{for-loop}, and
+% I'll just have my \textbf{int-i}, and I'll go from zero uh, zero to,
+% four actually cuz it'll be zero, one, two, three, four, yeah, zero to
+% four. &
+% \includegraphics[width=2.23333in,height=0.35556in]{media/image8.emf}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% In his narrated pseudo-code, he introduces the C-specific syntax of
+% ``for-loop'' and ``int-i,'' which handle iteration and indexing,
+% respectively. Moreover, he counts in order to specify the index bounds
+% of the loop (zero to four).
+%
+% Second, it's actually somewhat complicated to address the question ``how
+% do Lionel's inscriptions align with what he says?'' He \emph{says}
+% ``I'll have a for-loop'', but the visual evidence shows that he hasn't
+% bounded his for-loop body in curly-braces.\footnote{As a language, C
+%   typically uses curly braces to bound instructions that span multiple
+%   lines. So, a typical for-loop would be written like this, where the
+%   loop preamble is bounded by parenthesis and the loop body is bounded
+%   by curly braces (bolded):
+%
+%   for(int i=0; i \textless{} maximum\_value; i++) \textbf{\{}
+%
+%   // Do something
+%
+%   \textbf{\}}} If he had done so, his for-loop preamble would end with
+% an opening curly brace; it doesn't (Table 1 graphic above). Also, for
+% matching, his for-loop should have concluded with a closing curly brace;
+% it doesn't. Nevertheless, his verbal evidence --- particularly him
+% saying ``int-i'' --- suggests that he understands details that variables
+% need to be declared. Together, these mismatches suggest that features
+% (viz., denoting iteration and setting its bounds) are important enough
+% to be included in this representation. Other specifics, including proper
+% declaration of variables, are things Lionel understands but does not
+% insist on writing.\footnote{One might argue that Lionel is simply being
+%   sloppy and forgetting to include details, even though he said them.
+%   The point I'd make here is that even if he is being sloppy, I would
+%   think it unfair to say, for example, that Lionel has conceptual issues
+%   with declaring variables. His talk (and many later iterations of his
+%   design) suggests he understands that declarations are necessary for
+%   working with variables in C. Moreover, even if he is being sloppy on
+%   the particular point of variable declarations, the overall character
+%   of this pseudo-code is still starkly different from that of the full,
+%   working source code solution he'll ultimately produce. So, even if we
+%   propose that Lionel has some unintended tolerance --- such as
+%   sometimes overlooking missing variable declarations --- for assessing
+%   whether his pseudo-code satisfices, there is still strong evidence
+%   that his criteria for evaluating his work in pseudo-code stage are
+%   markedly different from the criteria he'll apply to later iterations
+%   of design.
+%
+%   no one would confuse the pseudo-code skeleton here for Lionel's
+%   fleshed-out, fully-working code later. So, even if Lionel \emph{meant}
+%   to put them in and forgot, he still judged this pseudocode as
+%   satisficing his subgoal} Together, and combined with later evidence of
+% Lionel's successful working program, these observations suggest that
+% ignoring the details is actually productive for Lionel.
+%
+% Third, Lionel's talk reveals productive capacities for dealing with
+% design work. For example, he describes the guts of his program through
+% brief perspective-taking (lines 14 -- 18 above). As a reminder, Lionel's
+% procedure is what I refer to as a ``king of the court'' approach: if the
+% number currently-being-inspected exceeds the reigning champ, the champ
+% is dethroned and replaced by the number currently-being-inspected. But,
+% the way Lionel articulates that process actually uses multiple senses of
+% ``I,'' where it's at times actually not clear whether ``I'' means ``I
+% the programmer'' or ``I the program.''\footnote{To keep your head from
+%   exploding, I remind you that the lower-case version of ``i'' in
+%   Lionel's talk actually refers to the subscript for indexing an array.}
+% For example, the I in ``I'll say `if the array of i'' could very well be
+% the programmer, where by ``I'll say'' he means ``I the programmer will
+% write a conditional test.'' But, that phrase is immediately followed by
+% ``if, uh, the number in the array that I'm looking at is greater than
+% max'', where I believe we may be seeing Lionel taking the perspective of
+% the program itself.\footnote{My point here is that technically, Lionel
+%   can never inspect an array element himself; he can only write code
+%   that does that. You might disagree, in which case we get to have a
+%   really fun discussion over beer about how we think programming is a
+%   case of ``distributing cognition'' (Hall, Wright, \& Wieckert, 2007).}
+% If anything, the seemingly untroubled, interchangeable use of ``I'' here
+% might suggest that to do the work of programming one can occupy a hybrid
+% space of both being the program and being the programmer at the same
+% time.
+%
+% Lionel concludes this procedure with the invocation ``same for min.''
+% Picking up the thread of hybridity, this phrase could mean, ``I the
+% programmer will write a similar procedure for min'' or ``I the program
+% will perform similar operations to determine min.'' That Lionel leaves
+% his pseudocode there (``so that's my pseudocode'', line 20) suggests
+% that whatever perspective it might be written from, this designed
+% artifact satisfices his pseudo-code production subgoal. And, that
+% satisfaction is remarkable because if ``same for min'' is an appropriate
+% place-holder, it suggests Lionel has ways of seeing that procedure as
+% \emph{similar enough} in structure to the procedure already laid out as
+% to not require further specification. This is an important point: if a
+% precursor to functional abstraction is recognizing repeated code, Lionel
+% has actually recognized repeated code \emph{before ever writing it}. As
+% we'll see later, he ends up writing this procedure out explicitly, which
+% means that whatever answer we might come up with as to why Lionel
+% doesn't abstract this to a function \emph{can't} be that he doesn't see
+% the repeated computation; to the contrary, repeating core parts of the
+% max computation was built into his design from the start.
+%
+% \subsubsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188557}{}{\protect\hypertarget{ux5fToc252445930}{}{}}We
+% can trace continuities in Lionel across different kinds of design
+% activities}{We can trace continuities in Lionel across different kinds of design activities}}\label{we-can-trace-continuities-in-lionel-across-different-kinds-of-design-activities}
+%
+% We can trace patterns in Lionel's design activity by asking questions
+% about the intermediate design artifacts he creates. Specifically, we can
+% pay attention to \emph{what} representations he creates, \emph{where}
+% they live, and \emph{how} he uses them. Table 3 outlines those
+% relationships for the three contexts discussed: Lionel's bike modding
+% project, Lionel's retrospective account of his programming practices,
+% and Lionel's in-interview work on the range-finding task. Table 3 helps
+% support three key points:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   There is a specific consistency to Lionel's design activity across
+%   contexts. All activities---given the limits of Lionel's
+%   self-reports---\emph{involved the generation and use of intermediate
+%   design representations}. Those representations may at times have lived
+%   in Lionel's head, on paper, or in electronic format.
+% \item
+%   A thorough account of Lionel's in-interview programming activity
+%   cannot and should not ignore those intermediate representations and
+%   how he used them.
+% \item
+%   Because intermediate representations are a feature of all three design
+%   contexts, it seems likely that what stabilizes their creation and use
+%   is tied to something deeper or more global than just conceptual
+%   knowledge in a domain such as computer programming.
+% \end{enumerate}
+%
+% \protect\hypertarget{ux5fToc252445960}{}{}Table -- Comparing Lionel's
+% design activity across contexts
+%
+% \begin{supertabular}[]{@{}llll@{}}
+% \toprule
+% \textbf{Context / Question} & Modding the bike & Retrospective
+% programming account & In-interview task\tabularnewline
+% \midrule
+% \endhead
+% \textbf{\emph{What} intermediate representations did he create?} &
+% Possible configurations for mounting the engine & Relationships of
+% functions to each other; pseudo-code for how functions would work & A
+% verbal description; narrated pseudo-code; hand-written pseudo-code;
+% hand-written source code\tabularnewline
+% \textbf{\emph{Where} do the intermediate representations live?} & In
+% Lionel's head & On a chalkboard and, later, in the text editor & In
+% Lionel's talk and gestures; on paper; in the text editor\tabularnewline
+% \textbf{\emph{How} did he use those intermediate representations?} & To
+% compare mounting configurations for ease and safety & To ``see the whole
+% picture'' of a project & To ``understand what's goin on in my head''
+% when structuring his code\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% \subsection{Rebecca's approach to
+% programming}\label{rebeccas-approach-to-programming}
+%
+% In this section, I analyze data from a different student --- Rebecca. I
+% offer this data as an example of what developing design expertise might
+% look like. If Lionel's approach to programming uses high-level work
+% (e.g., his pseudo-code) to cascade down the specifics of syntax, it
+% seems reasonable to ask whether other students use pseudo-code in the
+% same way. And, if they do not, it seems worth asking how uses of
+% pseudo-code differ and what that difference might tell us about
+% students' approaches to programming.
+%
+% In Rebecca's case, the data I'll present begins with her stuck on the
+% third of four projects in the course: the ``iTunes'' project.\footnote{In
+%   this project, students were given a series of text files ---
+%   representing the user's music collection --- that contained formatted
+%   albums and song titles. The task was to create a program to read in
+%   the music information, create a database representation of it, and
+%   allow the user to transact with that database. Typical transactions
+%   might be asking for a listing of all the albums in a user's library or
+%   creating a playlist by allowing the user to choose specific songs.} My
+% analysis will expand on the following points:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   To understand Rebecca's struggle on the iTunes project and her
+%   approach to programming during that struggle, we need to understand
+%   some of her prior experiences in programming. In our fourth interview
+%   Rebecca recalls a pivotal project experience from her Basic
+%   Programming course. At the cost of hours of work, half her codebase,
+%   and a good grade, Rebecca discovered that her sense of how her program
+%   should work did not match what was expressible in C. I analyze
+%   Rebecca's recollection because it sits at the intersection of
+%   self-efficacy and disciplinary practice. Rebecca lost confidence and
+%   felt ``demoralized'' because her attempts at expressing a procedure
+%   did not match the constructs and underlying workings of the C
+%   language. Moreover, the fallout from that experience persisted into
+%   Intermediate Programming. By our fourth interview she had considered
+%   dropping Intermediate Programming entirely because of her low grades.
+%   ``My *biggest* problem'' in programming, she said, ``is I think I have
+%   the logic, my logic just doesn't transfer to code. Or I don't know
+%   *how* to transfer it correctly'' (Interview 4 of 5, April 6, 2012).
+% \item
+%   Rebecca has the capacity to make sense of program designs. As she
+%   tries to understand why the instructor would prescribe an array of
+%   pointers to store track titles in the music server, Rebecca makes
+%   progress by reasoning about what she knows about pointers and arrays.
+%   Like Lionel, Rebecca can use gestures, talk, and inscriptions (viz.,
+%   pseudocode) to express how she might intend for her program to work.
+%   Rebecca also has a grasp of the conceptual issues --- from a
+%   computational and programming perspective --- relevant to the part of
+%   her design she's struggling with. But, Because Rebecca's approach to
+%   programming so strongly orients toward producing valid, working
+%   syntax, it may be limiting Rebecca from seeing other ways in which she
+%   can make progress on her design. In other words, Rebecca \emph{can} do
+%   things like write pseudo-code to plan, but she doesn't tend toward
+%   those behaviors when she's stuck. I'll offer an explanation based on
+%   epistemological dynamics to explain why that's the case.
+% \end{enumerate}
+%
+% In what follows, I first present point 1 to give readers a sense of how
+% Rebecca's prior experiences may inform her approach to programming. I
+% then outline the particular part of the iTunes project where she feels
+% ``stuck,'' discussing point 2. Before I jump into that analysis, I
+% \emph{strongly} urge readers who are not familiar with pointers or
+% dynamic memory allocation to refresh themselves on the subject.
+%
+% \subsubsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188558}{}{\protect\hypertarget{ux5fToc252445932}{}{}}Rebecca's
+% struggles on the iTunes project make sense in light of some prior
+% programming experiences she
+% had}{Rebecca's struggles on the iTunes project make sense in light of some prior programming experiences she had}}\label{rebeccas-struggles-on-the-itunes-project-make-sense-in-light-of-some-prior-programming-experiences-she-had}
+%
+% At the time of this study Rebecca was a first-year electrical
+% engineering major who had gone to high school in rural Maryland. Her
+% university courses in Basic and Intermediate Programming were the first
+% experiences she had doing any programming at all, let alone in the C
+% language. Though she felt academically at ease in high school, a tough
+% project in her first-semester Basic Programming course shook her
+% confidence in programming. In the data presentation and analysis that
+% follows, there are four sub-points I'd like to make about Rebecca's
+% development in programming.
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   Prior to the semester I worked with her, Rebecca had difficult
+%   experience trying to understand arrays on a Basic Programming course
+%   project.
+% \item
+%   That difficult experience had two pieces to it that I believe relate
+%   to personal epistemology:
+%
+%   \begin{enumerate}
+%   \def\labelenumii{\alph{enumii}.}
+%   \item
+%     Rebecca came to feel that her idea for a solution wasn't something
+%     expressible in C.
+%   \item
+%     When she got help from a friend who was ``very smart in
+%     programming,'' she felt like she typed the code he told her to into
+%     her project without really coming to understand it.
+%   \end{enumerate}
+% \item
+%   Reflecting on it in our first interview, Rebecca pinpoints the
+%   experience of that array project as demoralizing and having a lasting
+%   effect on her confidence in programming.
+% \item
+%   Epistemological issues continue to factor into Rebecca's confidence in
+%   programming. By our fourth interview, Rebecca was getting low grades
+%   in Intermediate Programming. She had strongly considered dropping the
+%   course. Trying to articulate her ``biggest problem,'' Rebecca
+%   explained that she felt like she understood the theory behind the code
+%   examples in class but had difficulty transferring her logic into code.
+% \end{enumerate}
+%
+% \paragraph{Rebecca had a difficult experience on a Basic Programming
+% arrays
+% project}\label{rebecca-had-a-difficult-experience-on-a-basic-programming-arrays-project}
+%
+% In our first interview, I asked Rebecca about how her experiences in
+% Basic Programming compared to those she was having in Intermediate
+% Programming. She explained that in that first semester, she ``kinda got
+% lost on one part'' --- arrays --- ``that was key to the rest of the
+% year.''
+%
+% Rebecca: {[}I{]}n {[}Beginning Programming{]} we had three projects, and
+% the first one, uh, we did---I did discuss it with other students, and we
+% all worked together kind of. But that one I felt much more like I had a
+% grasp on everything I was doing. Whereas the second project, I would
+% stare at the code and be like ``ohmygosh, I don't even understand what,
+% like, it's asking me to do.'' Like, or how to make it do what it needed
+% to do, which was to like store in numbers in arrays, change them, and
+% add `em together and stuff. (Interview 1 of 5, February 10, 2012)
+%
+% I asked Rebecca if she felt she had a better grasp by the time she
+% turned in the project.
+%
+% Rebecca: No. Cuz, there was a friend on my floor, who---very smart at
+% programming, he's had like years of experience. And he gave me a lot of
+% pointers and tips doing it, so I felt like I was just using him a lot
+% instead of actually learning it, which was a downfall for me.
+%
+% Interviewer: Huh. So, how, I mean. I feel like sometimes you could---you
+% might ask somebody for help, and then they show you something and then
+% you might be like oh, you've learned it now /yes/, but it sounds like
+% you're saying /no/
+%
+% Rebecca: Just, I don't, like, I don't blame him at all because he helped
+% me, but, it was kind of more of a, he showed me how to do it, I was like
+% \{mimes exaggerated typing, in an almost singsong voice\} *OK, sure,
+% I'll type that code*, but I don't really, I never really, I don't know I
+% never made the connection. (Interview 1 of 5, February 10, 2012).
+%
+% \paragraph{Rebecca's ``terrible'' array experience has epistemological
+% components}\label{rebeccas-terrible-array-experience-has-epistemological-components}
+%
+% In our second interview a week later when we discussed debugging,
+% Rebecca returned to that experience in her Basic Programming class:
+%
+% Rebecca: My array project, my---the second project we did last semester,
+% that I had a big error. I had to delete like half the code I worked on.
+% Like, just cuz, like I had said last week, I didn't do very well on
+% arrays. So.
+%
+% Interviewer: Oh. Was it cuz of one thing, or was it like a repeated---
+%
+% Rebecca: It was kind like a general, like, what I was trying to do
+% didn't work with C.
+%
+% Interviewer: Do you remember what it was you were trying to do?
+%
+% Rebecca: I don't know exactly, but uh, I just remember, like, whatever I
+% was tryin to do, cuz I had my friend help me, he's like ``you can't do
+% this this way,'' like, it just, C doesn't recognize whatever I was
+% trying to do.
+%
+% Interviewer: Huh. So that was like, it sounds like it's almost kind of a
+% case where your, your plain English of what should happen /yeah/ can't
+% actually translate---it's not even that you're not sure how it's---
+%
+% Rebecca: It's like, it was impossible.
+%
+% Interviewer: It was never---
+%
+% Rebecca: It was never allowed.
+%
+% Interviewer: There is no word for that /yeah/ in this language.
+%
+% Rebecca: I \{laughs\}, that was back then, so. That was always fun.
+%
+% (Interview 2 of 5, February 17, 2012)
+%
+% Epistemologically, two patterns stand out in Rebecca's recollections.
+% First, as Rebecca remembers it, one of her core problems was that her
+% visions for how her program should accomplish things were ``impossible''
+% and ``never allowed'' in C. Second, in order to get past that obstacle,
+% she ultimately typed code a friend told her to without feeling like she
+% understood it. As this analysis unfolds, I think these two points give
+% us a crucial starting point to understand Rebecca's practice. The
+% experience underscored a gulf between ideas Rebecca wanted to invoke and
+% the syntax she needed to realize them in C. And, her way of dealing with
+% that gulf was to make an end-run around it by taking and using code she
+% didn't fully understand.\footnote{I say this without any intended
+%   judgment (or pejorative connotations) about what Rebecca did. Given
+%   the assessment structure of the course, where 90\% of a student's
+%   grade came entirely from whether the program worked to spec, she may
+%   have felt that getting a good grade by using code she didn't
+%   understand was preferable to getting a bad grade with code that didn't
+%   work.}
+%
+% It's worth noting that whether her friend's code worked to
+% spec\footnote{Here, I'm using ``worked to spec'' as a shorthand for
+%   ``worked in such a way that it passed all of the input/output tests
+%   the instructor would run to determine 90\% of a student's grade.''} is
+% immaterial to the two epistemological issues I raise. First, if the code
+% she copied worked, Rebecca would get a good grade, but she wouldn't
+% understand how her project functioned. And, if it didn't work, she would
+% get a bad grade \emph{and} she wouldn't understand how her project
+% functioned. Ultimately, no matter how the code performed, Rebecca would
+% likely be in a tough spot if she ever had to maintain, revisit, or
+% refactor the code she copied. Second, no matter how her copied-code
+% project functioned, it represented \emph{someone else's} articulation of
+% how a procedure should go. That is, the copied code was an artifact of
+% someone else bridging the gulf between high-level idea and articulation
+% in C. Rebecca still found difficulty traversing that gulf.
+%
+% \paragraph{Rebecca's array project experience had a lasting effect on
+% her confidence in
+% programming}\label{rebeccas-array-project-experience-had-a-lasting-effect-on-her-confidence-in-programming}
+%
+% As that interview drew to a close, I wrapped it up with what had become
+% standard protocol: asking whether there was anything else she wanted to
+% talk about.
+%
+% Rebecca: I think I'm good, I---just
+%
+% Interviewer: Anything else on your mind?
+%
+% Rebecca: Sorry I'm not the greatest programmer \{laughs\}
+%
+% Interviewer: No, hey hey hey. That's \{shakes head\}, first of all, I
+% mean, um, how come you're apologizing?
+%
+% Rebecca: I dunno. It's a study, and I'm probably not the best test
+% subject. (Interview 2 of 5, February 17, 2012)
+%
+% Rebecca's spontaneous apology spawned a discussion about why she feels
+% diffident in programming. For her, that diffidence traced its way back
+% Basic Programming. Feeling lost and stuck on that array project was more
+% than just a tough experience; it was a turning point:
+%
+% Rebecca: When I got lost I think is when I started losing that
+% confidence thing. And then I never really got it back, so.
+%
+% Interviewer: When was it you started to get lost?
+%
+% Rebecca: Halfway through last semester, like with the arrays in the
+% second project. So\ldots{}. I guess it was like, demoralizing, kind of,
+% and, like, like I don't want to sound cocky or arrogant or anything, but
+% in high school I did very well in academics. And, so, I never really,
+% like anything that I had to work for, I worked for for a short period of
+% time, got it, and like I'd understand it, and if I didn't, like it was
+% really easy for someone usually to explain it to me. And, I felt like I
+% didn't really get that after I lost everything on that second project, I
+% was like, I wasn't able to build myself back up out of that, which
+% probably stems to why, not as confident in the class. (Interview 2 of 5,
+% February 17, 2012)
+%
+% As I interpret how this array project experience may be playing a part
+% in Rebecca's later approaches to programming, I stress that my analysis
+% is about Rebecca's perceptions and recollections.\footnote{Due to
+%   deliberate study limitations I don't have access to the actual code
+%   from that project. To be as conservative as possible in our data
+%   collection and respect the privacy of participants, we asked
+%   participants to grant us access only to folders that would contain
+%   their Intermediate Programming projects. Outside code/data, including
+%   code from prior semesters, was excluded from the scope of our code
+%   snapshot data collection.} It's possible that her ideas for how her
+% project should work wouldn't have been implementable in any language,
+% let alone in C. And, while I think that possibility is unlikely, I have
+% to concede that I only know what Rebecca recalled; I wasn't present for
+% the experiences she's describing. Nevertheless, the question for my
+% research is not ``can we know what really happened in that experience?''
+% Rather, my question is, ``how might Rebecca's \emph{image} of that
+% experience --- with its specific epistemological facets --- be playing a
+% part in her approach to programming now?''
+%
+% Part of the answer to that question may be in her quote above. She talks
+% about losing her confidence and ``never really {[}getting{]} it back.''
+% Later, she says ``I felt like I didn't really get that after I lost
+% everything on that second project.'' Because of the peculiarities of
+% speech, I had a difficult time parsing Rebecca's statement about losing
+% everything on that second project. It wasn't clear to me whether there
+% was a pause between ``get that'' and after, but the meaning of ``that,''
+% and in turn the meaning of the sentences, depends on the antecedent of
+% ``that.'' So what she said could have two consequentially different
+% meanings:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   ``I feel like I didn't really get \textbf{that}. After I lost
+%   everything I lost everything on that second project, I was like, I
+%   wasn't able to build myself back up out of that.'' In this version,
+%   ``that'' may mean the topic---arrays---or the project as a whole. I
+%   might then interpret Rebecca's speech as saying this project was a
+%   turning point because of the particular topic. Her being ``unable to
+%   build myself back up out of that'' reflects a kind of debt that might
+%   have piled up when later assessments depended cumulatively on that
+%   early topic. Her failure to grasp arrays was thus cumulatively
+%   punishing because arrays formed a core part of many projects that
+%   followed.
+% \item
+%   ``I feel like I didn't really get \textbf{that} after I lost
+%   everything on the second project. I was like, I wasn't able to build
+%   myself back up out of that.'' In this version, ``that'' may mean her
+%   ready insight into topics or her facility in understanding when
+%   friends explain them to her. The distinction I draw in this second
+%   interpretation is ``that'' refers to a kind of continuity in her
+%   identity---someone for whom school always came easy. The array project
+%   disturbed that continuity because it was an instance in which things
+%   no longer came easy to her. And, the experience may have been so
+%   troubling that she was never able to re-establish being able to easily
+%   grasp things in programming. The result, then, is that she couldn't
+%   build herself---in the sense of her continuous identity as one who
+%   quickly grasps school topics in general and programming topics in
+%   particular---back up after the array project.
+% \end{enumerate}
+%
+% Distinguishing between these possible meanings of Rebecca's speech might
+% seem pedantic. But, I think the bifurcation is important for theorizing
+% about learning and instruction in computing. If Rebecca's difficulty was
+% about the topic of arrays, a plausible narrative is that the difficulty
+% of that concept---and the assessments that continued to depend on
+% it---caused a run of bad grades, and the bad grades shook Rebecca's
+% confidence. But, if Rebecca's difficulty comes from troubling her sense
+% that she has a ready understanding of programming, the attack on her
+% confidence is more complex. It's not just that she doesn't get arrays,
+% it's that she feels she doesn't readily understand other new code she's
+% seeing in the course---even code that doesn't strongly rely on arrays.
+% In this view, the confidence Rebecca speaks of isn't just a product of
+% getting good grades; it's part of a feedback loop in which grades,
+% confidence, and a feeling of ready insight into programming are all
+% components.
+%
+% That potential coupling of grades, confidence, and feelings about ready
+% insight into code matters because it further pushes the discussion
+% toward epistemological considerations. It is why, for example, I was
+% paying such careful attention to Rebecca's recollection of copying down
+% code her classmate told her. Instead of helping her confidence in
+% programming by potentially boosting her grade, the copying may have
+% diminished her confidence because she couldn't understand why someone
+% else's code worked.
+%
+% \paragraph{Epistemological issues continued to interact with Rebecca's
+% confidence in Intermediate
+% Programming}\label{epistemological-issues-continued-to-interact-with-rebeccas-confidence-in-intermediate-programming}
+%
+% Rebecca started our fourth interview with a warning to me that she might
+% ``rant'' about programming.
+%
+% Interviewer: What would you rant about? Go ahead.
+%
+% Rebecca: Ohh god. Um, I am not doing well \{laughs\} in the class right
+% now.
+%
+% Interviewer: OK
+%
+% Rebecca: And it's not good, so. Yeah, like, we just had our last project
+% due, which I did a lot better on than the first project
+%
+% Interviewer: Uh-huh
+%
+% Rebecca: Which is making me really happy. Um, but, everything about
+% programming just upsets me, because there's the two guys, like I had
+% mentioned before who, uh, help me
+%
+% Interviewer: Mmmhmm
+%
+% Rebecca: Uh, one of them, he---like I worked every day from the day we
+% got this project on this project, and my project still didn't turn out
+% perfect. He wrote his twice in one day because he lost the whole thing.
+% And, his was perfect. And, I dunno---it just irks me *so* bad that that,
+% it can be that much of a disparity between people and programming.
+% (Interview 4 of 5, April 6, 2012)
+%
+% In her telling, Rebecca wasn't the only person upset by the relative
+% ease with which this student programmed.
+%
+% Rebecca: we didn't say it to him, but some of us were like, it's, we
+% just found it like, so, like aggravating. Like, cuz there's me and
+% another girl who, on my floor who, she's uh, never had programming
+% experience either---
+%
+% Interviewer: Uh-huh
+%
+% Rebecca: And we're just like it gets so aggravating that it comes easy,
+% like to him, and that he does it well, but, that's just---he ju---he is
+% very good at school. He's very smart. So. (Interview 4 of 5, April 6,
+% 2012)
+%
+% Note here that Rebecca is connecting the idea that programming came easy
+% to this student to the notion that he was ``very good at school.'' In
+% this moment, the statement seems to suggest Rebecca has a \emph{fixed
+% mindset} \{Dweck and colleagues\} about programming expertise. While it
+% would be naïve to say Rebecca believed --- globally --- that programming
+% came easily to those who were ``good at school,'' it's nonetheless true
+% she saw a discrepancy between how easy programming was for her and her
+% friend and how easy it was for this ``very smart'' classmate.
+%
+% As we kept talking about this discrepancy, Rebecca offered further
+% evidence that struggling with programming troubled her academic
+% identity. In high school, she ``didn't have to study a lot'' because ``a
+% lot of things came relatively easy'' (Interview 4 of 5, April 6, 2012).
+% But, her experiences in Intermediate Programming pushed her into a new
+% position. As she said,
+%
+% now I'm on the other end and I'm realizing how aggravating it
+% is\ldots{}. And, I don'---I guess it's just a shock, cuz I do not like
+% not being good at things \{laughs\}. So...'' (Interview 4 of 5, April 6,
+% 2012).
+%
+% Again, Rebecca's comments stress the frustration of feeling like
+% programming comes easily to some people but not to her.\footnote{Or, at
+%   least, at that moment, in that semester, it did not seem to be coming
+%   easily to her.} And, they are charged with emotion. She speaks of
+% being ``on the other end,'' connoting an idea of a structure whose poles
+% have people who ``get'' programming on one end and those who don't on
+% the other. It's ``aggravating'' to be on that end, and a ``shock'' to be
+% there for the first time.
+%
+% I asked Rebecca to reflect on whether she could ever remember seeing
+% classmates in the position she was now, where something seemed to come
+% easily to Rebecca but not to them.
+%
+% Rebecca: Um. I mean, there was a---there's a friend of mine, who, from
+% my high school, who, I, I mean just always got math. And, uh, she was
+% struggling in math a lot.
+%
+% Interviewer: Mmmhmm
+%
+% Rebecca: But her attitude was, uh, ``I don't need math. I'm gonna be an
+% English major.''
+%
+% Interviewer: OK
+%
+% Rebecca: So, she didn't even care about it
+%
+% Interviewer: Oh, OK
+%
+% Rebecca: Whereas, like, I was like, I'm an electrical engineer major, I
+% actually need this. So. Like that's the only time I think I've ever felt
+% like someone's had that feeling that I probably did.
+%
+% Interviewer: Right.
+%
+% Rebecca: And, I mean I'm sure there are people that I---uh, that I
+% didn't know, or didn't realize they had that feeling. But, it's just,
+% interesting to be on the other end. \{6 second pause\}
+%
+% Rebecca: Umm. Just. Ugh. I've like, debated withdrawing from the class.
+% And, like, I've talked to three different advisors or whatever, and
+% they're like ``don't do it. It's not---you can still bring your---you
+% can bring your grade up. And even if you don't you have freshman
+% forgiveness,\footnote{In the interview, Rebecca explained that Freshman
+%   Forgiveness was a policy extended to students during their first 24
+%   credit-hours at the university. If a student gets a grade she's
+%   unhappy with, she can retake that course and her second final grade
+%   will replace the first.} so.''
+%
+% Rebecca, unlike her friend, couldn't afford \emph{not} to care about
+% Intermediate Programming; it was a required course for her electrical
+% engineering major. But, the combination of feelings she had---consisting
+% partly of a feeling that programming did not come easily to her---pushed
+% her to think about withdrawing from the course anyway. And, Rebecca was
+% not in the habit of withdrawing from classes. Intermediate Programming
+% was the first and only class where she was experiencing this level of
+% difficulty (Interview 4 of 5, April 6, 2012).
+%
+% It would be tempting to think Rebecca struggled in Intermediate
+% Programming because she didn't work hard enough. But, by her own
+% assertions she \emph{did} put a considerable amount of effort on
+% projects. On Project 2, for instance, she said ``I worked every day from
+% the day we got this project on this project, and my project still didn't
+% turn out perfect'' (Interview 4 of 5, April 6, 2012). Figure 5 below
+% shows Rebecca's activity over time from March 5, the day Project 2 was
+% handed out, to March 28, the day it was due:
+%
+% \includegraphics[width=5.73333in,height=3.98889in]{media/image10.emf}
+%
+% \protect\hypertarget{ux5fToc252445971}{}{}Figure ‑ -- Rebecca's
+% compilation activity over time for Project 2. Each compile a student
+% initiates creates a commit, so long as there has been a change to the
+% underlying code. The height of each bar maps to the number of commits
+% recorded that day. The red line charts \emph{cumulative} commits over
+% time. The red line is steepest in periods of frequent activity and
+% shallowest in periods with little or no compile activity.
+%
+% From the data we can tell Rebecca was compiling Project 2 code almost
+% every day between March 19 and March 26. Moreover, roughly two thirds of
+% compilations for the entire project happened in the last 3 days leading
+% up to the deadline. So, at the very least our data shows she was
+% actively compiling her code for 8 out the last 10 days.
+%
+% Rebecca registered no commits between March 12 and March 19. There were
+% also other days for which no commits were recorded. One possible cause
+% for those zero values is our data collection system was faulty. A second
+% possible explanation is that Rebecca was not working. But, a third
+% explanation is that Rebecca was working but not compiling. This third
+% explanation actually seems most plausible. It fits the pattern of
+% Rebecca's work on project 1, where she wrote dozens of lines of code
+% before ever compiling (Interview 2 of 5, February 17, 2012). It also
+% accounts for Rebecca's lack of compilations despite her claim that she
+% worked on Project 2 every day.
+%
+% So, we can be fairly certain Rebecca was spending time on
+% projects.\footnote{The specific nature of \emph{what} Rebecca's work
+%   looked like on Project 2 is the subject of the first study of this
+%   dissertation, so I don't expand upon it substantially in this study.}
+% Another possibility for why Rebecca was struggling might be that she
+% wasn't studying. To address that conjecture, we have to step back to
+% when I asked Rebecca whether other classes --- classes other than
+% Intermediate Programming --- had ever pushed her close to withdrawing
+% from them. She said no.
+%
+% Rebecca: the thing is, I *don't* know how to like, study for
+% programming. Cuz like, I look at code and stuff, and I try, and I work
+% every day on it, but like, even when---I dunno, it just, does not come
+% naturally.
+%
+% Interviewer: So, what is it like when you study it? What do you usually
+% try to do?
+%
+% Rebecca: Uhh, a lot of times, like, I'll read through the notes we took
+% in class, like, cuz, what we're doing now---we have a project, uh, we're
+% doing linked lists, and like, malloc'ing data, which is, uh, *you*
+% storing it in, as you go. And, like, I get the theory and everything
+% behind it, and, like it makes sense what it does, but, I just like try
+% and look at his code examples and stuff, and try and replicate it, but I
+% don't know what's happening right now, but it's not doing it correctly.
+% I just, like, my *biggest* problem, I realized, in programming, is I
+% think I have the logic, my logic just doesn't transfer to code. Or I
+% don't know *how* to transfer it correctly.
+%
+% Interviewer: So, um. when you're spending time studying, um, do you
+% spend most of it on your computer trying to actually program or---
+%
+% Rebecca: A lot of it is on my computer and a lot of it is also trying to
+% just look at his code and notes in class. Uh, that's probably about
+% split, 50/50. Uh, because, sometimes I'll just end up staring at my
+% computer screen, like, ``what do I even type in to try?'' So, I try and
+% go look at his examples, so.
+%
+% (Interview 4 of 5, April 6, 2012)
+%
+% There are two striking features of this exchange.\footnote{There are
+%   actually three striking features. The one I left out above is that
+%   Rebecca describes malloc'ing as \emph{you} storing in data. I think a
+%   great deal of Rebecca's difficulty with malloc specifically hinges on
+%   what she means by ``you'' storing it in.} The first is that Rebecca
+% felt she didn't know how to study for programming. The second striking
+% feature of this exchange is how Rebecca elaborated on what it might mean
+% for programming to come naturally to her.
+%
+% Rebecca feels programming doesn't come naturally to her because she
+% routinely finds herself unable to transfer her ``logic'' to code. As a
+% follow-up to Rebecca's answer, I asked what ``logic'' meant to her in
+% this case.
+%
+% Rebecca: Like, like how to get the---like, if I need to get a certain
+% data, like I have a process of how I want to get it, I ju---and I'm
+% like, ``Oh, you would just do this,'' but can the *computer* do that?
+% Can I---and it's my---me being able to tell the computer to do that is
+% where I get lost. (Interview 4 of 5, April 6, 2012)
+%
+% The tension Rebecca is describing is one I see as strongly
+% epistemological. First, it's epistemological in the sense that it
+% reflects a struggle to articulate knowledge of \emph{how to do
+% something}. Papert (1980) and Abelson \& Sussman (1996) would describe
+% such a struggle as a pertaining to \emph{procedural epistemology}: the
+% struggle concerns a particular kind of knowledge work in which the goal
+% is not to describe what is, but rather how to. The second sense in which
+% it's an epistemological struggle is closer to how Hammer, Elby, and
+% colleagues use the word epistemological: what does students' activity
+% reveal about how they orient toward knowledge and knowing in a
+% discipline (Hammer et al., 2005; Hammer \& Elby, 2002, 2003; Scherr \&
+% Hammer, 2009). In this second sense, we can observe that Rebecca defines
+% her difficulty in terms of the creation of proper code. We can also pose
+% the question ``if Rebecca struggles to transfer logic to code, what
+% directs her activity as she \emph{tries} to transfer logic to code?''
+%
+% Later in the interview, she compared the iTunes project --- on which she
+% was stuck --- to Project 2.\footnote{Project 2 did not require students
+%   to use malloc() to dynamically allocate memory. Project 3 did.} In so
+% doing, she offered another telling idea about the relationship between
+% her logic and ``syntax'':
+%
+% Rebecca: Like, cuz, I guess I feel like I did a lot better on the last
+% project because it was so similar to stuff that I'm used to /Mmmhmm/ um,
+% whereas all this is brand new and I still don't have any of the syntax
+% down yet. /Mmmhmm/ Like, if-statements and while-loops, they make sense
+% to me, whereas I'm still like, trying to grasp at this stuff.
+%
+% By ``this stuff,'' the context of the conversation suggests Rebecca was
+% referring to dynamic memory allocation and malloc, which we had been
+% discussing.
+%
+% It seems sensible to think if-statements and while-loops might ``make
+% sense'' to Rebecca because their syntactical form aligns strongly with
+% their computational operation.\footnote{To borrow terminology from
+%   Sherin (2001) and rebut Saussure (1986) at the same time, the symbol
+%   template for an if-statement is strongly connected to its conceptual
+%   schema because we so often use everyday language in a manner
+%   concordant with an if-statement's branched control flow. In this
+%   instance, the relationship between signifier and signified is the
+%   \emph{opposite} of arbitrary. The similarity between the syntax for an
+%   if-statement and an understanding of what that statement does is by
+%   design, because C itself was a designed language.} By that, I mean in
+% principle the designers of C could have chosen \emph{any} symbol to
+% represent branched control flow, but they chose a word --- if --- that
+% invites connections to its meaning in everyday speech. If-statements in
+% C split control flow based on a Boolean expression's value; while-loops
+% allow continued iteration until a Boolean expression turns up false.
+% Both structures have everyday analogues with similar conceptual
+% properties. A parent might say, ``if you're staying out past 9, let me
+% know,'' or, ``while we're gone, can you make sure Sadie doesn't eat
+% anything from the garbage?'' If a child will be home by 8, there is no
+% need to let that parent know. Similarly, it's reasonable to expect that
+% once the parents return, the responsibility of keeping Sadie out of the
+% garbage no longer falls solely to the child.\footnote{Some parents might
+%   object that family members should \emph{always} be on the lookout for
+%   Sadie rooting through the trash, that these parents were just
+%   emphasizing the point. And, reasonable children might in general try
+%   to heed that. But, what makes computers computers is their lack of
+%   qualms about abandoning any and all responsibility to keep Sadie from
+%   the trash once the parents return.} The point of this observation is
+% that Rebecca may feel certain kinds of programming structures ``make
+% sense'' not simply because they were covered in introductory
+% programming, but because something about their syntactical form offers
+% an affordance for thinking about how they work.
+%
+% \subsubsection{Rebecca has productive capacities for making progress on
+% software design
+% work}\label{rebecca-has-productive-capacities-for-making-progress-on-software-design-work}
+%
+% In our fourth of five interviews, Rebecca was having trouble making
+% progress on Project 3, which the instructor described as an ``online
+% music server.''\footnote{Students in my study colloquially referred to
+%   this project as the ``iTunes project,'' so ``the iTunes project'' and
+%   ``Project 3'' all refer to this project.} ``Like, I am so lost, like I
+% don't even know where to start,'' she said. We discussed why she was
+% stuck. Then, with my help, we began to work through what Rebecca
+% identified as a core cause of her confusion. In brief, the assignment
+% contained a diagram showing the data scheme students were required to
+% use to store data about the music (Figure 8 below). Specifically, the
+% scheme required an array of pointers --- each entry of which pointed to
+% an array of characters --- as a mechanism for storing data about song
+% titles for the music server. Rebecca's work to resolve her confusion
+% offers evidence for her productive capacities in designing programs. In
+% this section, I argue:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   Rebecca made sense of the array-of-pointers design\footnote{By
+%     ``design scheme,'' here, I mean the instructor-provided diagram
+%     indicating how students were to structure their data for Project 3.}
+%   using talk and gesture. She was able to explain both why an array of
+%   pointers \emph{should} exist in C as well as suggest candidate syntax
+%   for what might create such an array.
+% \item
+%   When I proposed we pretend her candidate syntax worked, Rebecca
+%   articulated through gesture, talk, and written pseudo-code how part of
+%   her design would incorporate that syntax.
+% \item
+%   After the interview, Rebecca resumed work on the project and
+%   incorporated a version of the design she developed in the interview
+%   into her code.
+% \end{enumerate}
+%
+% \paragraph{Rebecca made sense of the array-of-pointers design using talk
+% and
+% gesture}\label{rebecca-made-sense-of-the-array-of-pointers-design-using-talk-and-gesture}
+%
+% The figure below shows the instructor-provided diagram for how students
+% should structure their data.
+%
+% \includegraphics[width=5.36667in,height=4.10000in]{media/image11.emf}
+%
+% \protect\hypertarget{ux5fToc252445972}{}{}Figure ‑ -- The instructor
+% required students to use this data arrangement for storing information
+% in the music server. The scheme uses an array of pointers to represent
+% the track names of an album.
+%
+% Rebecca would have seen this diagram before because the project
+% assignment was given out 11 days prior to our interview. But, the data I
+% present is the first time I had any access to how she was thinking about
+% it.
+%
+% At the time of the interview Rebecca felt stuck on handling issues with
+% \emph{music albums} --- the top-left and top-right structures depicted
+% on the diagram. The assignment required that ``all of the data
+% structures described {[}in this diagram{]}'', including music albums and
+% all of the data they referenced, ``must be allocated dynamically''
+% (Instructor-provided Project 3 description). So, Rebecca and I started
+% discussing what parts of that scheme she understood. Table 3 below
+% presents what she said and what she did while beginning to explain her
+% understanding.\footnote{}
+%
+% \protect\hypertarget{ux5fToc252445961}{}{}Table - Rebecca describes an
+% array of structures using gestures (Interview 4 of 5, April 6, 2012)
+%
+% \begin{supertabular}[]{@{}ll@{}}
+% \toprule
+% What she says & What she does\tabularnewline
+% \midrule
+% \endhead
+% Like, the---the thing that makes sense to me that I, uh, got right now,
+% is that, to malloc the array, to \{bounds a wide space with her hands\}
+% make the array a certain length, do that. &
+% \includegraphics[width=2.03333in,height=2.72222in]{media/image12.png}\tabularnewline
+% And then I know that each of the \{chops out three invisible boxes in
+% the air\} I guess nu---characters---I---uh, I'll call `em characters. &
+% \includegraphics[width=1.37778in,height=3.66667in]{media/image13.emf}\tabularnewline
+% Like, each of the spots in the array holds a structure and each of the
+% structures holds three things. & (gestures obscured by the computer
+% )\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% Rebecca's gestures create a kind of virtual object that looks much like
+% the music albums array in the diagram. As she chops out spaces for each
+% of the ``characters,'' those spaces are in series and fit --- more or
+% less --- within the larger area she bounded with her when she said ``to
+% malloc the array.'' The spatial subdivision is key because it offers
+% supporting evidence that in some way, Rebecca might have been thinking
+% about the album structures themselves as being \emph{contained} in or by
+% the array. And, that spatial containment metaphor resurfaced later when
+% we discussed the array of pointers, which I'll get to in a moment.
+%
+% As Rebecca explained her understanding to me, I took to writing down
+% what I understood her to be saying on paper:
+%
+% Interviewer: OK. So you have, so here's what you just said. You said
+% \{writes\} ``malloc the array'' um, ``each element has a structure,''
+% and then ``each structure has'' you said---
+%
+% Rebecca: Has three parts
+%
+% Interviewer: OK. Has three parts. OK.
+%
+% Rebecca: Yeah. So, and the first part I can do `cuz it's an integer
+% \{laughs\}. The first part is just the number of tracks on the album.
+% So---
+%
+% Interviewer: Oh, oh. So, of the three parts {[}{[}I'm sorry
+%
+% Rebecca: Yes{]}{]}
+%
+% Interviewer: I {[}{[}\{untillegible\}
+%
+% Rebecca: Oh yeah, sorry{]}{]}
+%
+% Interviewer: So you're saying, um, and then---and then this is number of
+% tracks, which is an integer, right? /yeah/ OK (Interview 4 of 5, April
+% 6, 2012)
+%
+% Figure 9 below shows what I wrote as we worked:
+%
+% \includegraphics[width=5.62222in,height=1.48889in]{media/image14.emf}
+%
+% \protect\hypertarget{ux5fToc252445973}{}{}Figure ‑ -- I write down what
+% I understand as Rebecca explains the overall data structure of albums to
+% me
+%
+% It wasn't the first time --- either in my series of interviews with
+% Rebecca or in that interview in particular --- that I tried to write
+% down what she said. But, I highlight it here because it became part of
+% my activity as we worked together in this episode. Rebecca would
+% articulate something, often through talk and gesture, and I would try to
+% write down and offer back to her what I understood her to be explaining.
+%
+% As we continued, we landed on trying to understand the ``tracks''
+% portion of the album data structure. Storing tracks was where Rebecca
+% felt she was getting ``lost'':
+%
+% \protect\hypertarget{ux5fToc252445962}{}{}Table -- Rebecca says gets
+% lost on the array of pointers pointing to track names (Interview 4 of 5,
+% April 6, 2012)
+%
+% \begin{supertabular}[]{@{}ll@{}}
+% \toprule
+% What she says & What she does\tabularnewline
+% \midrule
+% \endhead
+% Rebecca: And, the next one is where I get lost, on the tracks part.
+% Because it's a pointer \{2 sec pause\} to \{hands make a vertical
+% cylinder in front of her; note that the array is represented on the
+% assignment as a narrow, vertical rectangle\}--- &
+% \includegraphics[width=3.40000in,height=2.62222in]{media/image15.emf}\tabularnewline
+% it looks like another array that's
+% \textbar{}\textbar{}pointing\textbar{}\textbar{} \textbar{}\{left hand
+% crosses from her left to right, at chest height, index finger extended
+% in the direction of motion; this is the same direction (from Rebecca's
+% perspective) in which the pointer array on paper points to each track
+% name\}\textbar{} to the names. &
+% \includegraphics[width=1.98889in,height=3.38889in]{media/image16.emf}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% Rebecca's gestures for the array of pointers again spatially mimic the
+% depiction of the array on the page: a vertical cylinder. The reason I
+% note the orientation of her array gestures is that for practical
+% purposes arrays as represented by bits in C have no orientation in space
+% that the programmer could know, much less care about. An array cannot be
+% parallel to the ground, nor can it be oriented toward the
+% ceiling.\footnote{The computational hardware representing the array does
+%   have an orientation in space, but that's clearly not what is at issue
+%   here. The whole point of the array is to be an abstraction away from
+%   the soldered transistors, capacitive plates, flipped switches, vacuum
+%   tubes, tinker toys, or other physical means that store the state of
+%   the array.} But, the assignment \emph{depicted} the album array as
+% horizontal, and so too did Rebecca's gestures for it. The assignment
+% \emph{depicted} the pointer array as vertical, and so too did Rebecca's
+% gestures for it. In other words, the observation that Rebecca's gestures
+% align with the assignments pictures might suggest that she is doing more
+% than thinking about arrays in the abstract. Rather, some elements of the
+% representational features of the page have made their way into her
+% activity and, arguably, her cognition.\footnote{One could argue that
+%   gesture is possibly just undirected flailing, and that the correlation
+%   between Rebecca's gestures and the page is weak at best --- the sort
+%   of thing you'd expect to happen some percentage of the time anyway
+%   under the assumption that the pictures on the page have nothing to do
+%   with how she's thinking about it. I concede that it's \emph{possible}
+%   the correlation isn't structurally meaningful. But, I think the
+%   correlation is meaningful given how many other times her gestures
+%   align with canonical written representations of computational
+%   structures and processes.}
+%
+% Rebecca continued on to try thinking through why an array of pointers
+% was part of the design.
+%
+% Rebecca: But, I've always been confused as to why you---I guess, I
+% always---I was always like, when we learned pointers, I was like ``why
+% do you need pointers when you could just call it the name? Why do you
+% need two names for it?'' But, I think, what---at least what I'm seeing
+% here maybe is like, this is just another array /Mmmhmm/ because you
+% can't put more than one character in each element of the array, so that
+% is just an array of pointers that point to strings. And the strings are
+% the names.
+%
+% Rebecca: So, that would make sense. So you would---the tracks would
+% point to the array, and so you could do the pointer of that \{1.5 sec
+% pause\} The pointer to the array, an array---the array of like 1 would
+% be a pointer to track 1's name.
+%
+% Interviewer: Mmmhmm
+%
+% Rebecca: So. OK, so---see that part makes sense now. I just---no idea
+% how I would ever access that. Or, store it, I guess, is the better word.
+% (Interview 4 of 5, April 6, 2012)
+%
+% What ``confused'' Rebecca has an understandable origin. Most of the work
+% she would have been doing until that point would have used named
+% variables. For example:
+%
+% int age;
+%
+% float weight;
+%
+% char *name;
+%
+% If age, weight, and name are already-named pieces of data, why would
+% something ever need to refer to them indirectly? It would be as if we
+% had to call me ``The second-born son of the second-born daughter of
+% Henry'' for some reason. I already have a name. Why do I need another
+% name that uses the names and relationships of my forebears to refer to
+% me when historically my given name has perfectly fine for all
+% referential purposes?
+%
+% Rebecca's way of finding sense in pointers was to observe that you can't
+% put names --- each of which is defined as an array of characters ---
+% into an array because ``you can't put more than one character in an
+% array.'' To verify I was understanding her idea correctly, I tried to
+% restate it:
+%
+% Interviewer: OK, so you were saying the reason you---when you're tryin
+% to convince yourself like---you originally you were thinking ``why don't
+% you just put all the names {[}{[}here
+%
+% Rebecca: Yeah{]}{]}
+%
+% Interviewer: And you were {[}{[}saying the reason is
+%
+% Rebecca: saying, no{]}{]}
+%
+% Interviewer: because==
+%
+% Rebecca: ==That makes sense now. Yeah, you can't put more than one
+% character in an element. And, it's a whole name of a song, so you would
+% need to point---that element would just \{gesture obscured by my
+% laptop\} be a pointer to a name. OK. (Interview 4 of 5, April 6, 2012)
+%
+% The latching and overlapping turns of talk above offer strong evidence
+% that Rebecca's explanation aligned with my interpretation. And again,
+% notions of \emph{containment} come into play. Rebecca says, ``you can't
+% \emph{put} more than one character \emph{in} an element'' in part
+% because you're talking about ``a \emph{whole name} of a song'' (my
+% emphasis added here). If Rebecca has a capacity for thinking about the
+% canonical concept of type-restrictions on arrays, it's manifesting here
+% as a part-physical metaphor, evoked through both talk (the quote above)
+% and gesture (Table 4), in the service of making sense of a design
+% decision.
+%
+% As she continued, Rebecca discerned that because the input data
+% contained the number of tracks on each album, she could use that number
+% to define how long the pointer array needed to be. But, something was
+% still troubling her.
+%
+% Rebecca: So. \{4 sec pause\} That would \{6 sec pause\}
+%
+% Interviewer: What're you thinkin?
+%
+% Rebecca: Just that, um, I'm just tryna think, cuz like that makes sense,
+% but I don't know how to get to that. Cuz, how do you make a pointer
+% point to a random array that you just made? Uh, I guess \{4 sec pause\}
+%
+% Rebecca: Uh, p---the pointer syntax I need---I would need to go back and
+% look at them because now you have an array, each array---my issue---cuz
+% like now that makes sense. Like, it's just how do I get to that point of
+% making that work?
+%
+% Interviewer: {[}{[}So
+%
+% Rebecca: And this is what{]}{]} I mean by my logic /OK/ I get the logic
+% and the theory behind it but I don't know how to actually put it into C.
+%
+% Interviewer: OK, so in other words, what makes
+% \textbar{}\textbar{}sense\textbar{}\textbar{} \textbar{}\{air
+% quotes\}\textbar{} to you know /Mmmhmm/ that I guess didn't before is
+% that *that* \{gestures to the instructor's structural diagram on
+% screen\} is a---that's an understandable arrangement /yes/ for the data
+%
+% Rebecca: Yes.
+%
+% Interviewer: But then, how do I {[}{[}make it
+%
+% Rebecca: make it{]}{]} Yeah \{nods\}. (Interview 4 of 5, April 6, 2012)
+%
+% When Rebecca said ``I don't know how to get to that,'' I take ``how to
+% get to that'' to mean \emph{how to write the code for that procedure}.
+% We had just agreed that design-wise, she could use a number provided in
+% the input data to define the length of the pointer of array. But, she
+% was stuck because of her difficulty translating the ``logic'' we
+% developed into code.
+%
+% From the outside, what Rebecca and I had done together was something I
+% think crucially important to the design of a program. We worked to
+% understand how someone else chose to structure their data.\footnote{That
+%   the ``someone else'' was in this case the instructor is important for
+%   larger power dynamic considerations.} We tried to decide for ourselves
+% whether and how those structures made sense. Along the way we created
+% ephemeral objects of that sense-making in the form of talk and gesture
+% as well as the durable artifacts of my inscriptions. But, none of that
+% work resulted in C code.
+%
+% For Rebecca, the lack of certainty about \emph{how to make} our ideas in
+% C code was a marked concern. And, again, that concern makes sense. We
+% had just convinced ourselves that we could understand \emph{why} an
+% array-of-pointers. We even figured out \emph{how} we could use the input
+% data to provide the length of the arrays we would need to make. But we
+% had not laid out how to declare the array \emph{and} preserve a
+% reference to it (``how do you make a pointer point to a random array
+% that you just made?). Without that piece --- a piece Rebecca was unsure
+% of and felt she would need to look up --- our solution clearly wouldn't
+% work.
+%
+% Absent that piece, Rebecca tried thinking of possible ways to fill in
+% the missing syntax.
+%
+% Rebecca: Cuz, I'm like, like I'm trying to think right now of things
+% that I would try /Mmmhmm/ but I don't know what I would try first
+% because, all I'm thinking right now is that I know tracks is going to be
+% a pointer. /Mmmhmm/ So you make that a pointer.
+%
+% Interviewer: OK.
+%
+% Rebecca: But I'm not sure where it would point to, because you don't
+% have that array made yet. So I \{3 sec pause\} so you'd have to, I guess
+% you could make the array, and then make it point to the array, but \{3
+% sec pause\} Cuz you have to make an array of pointers.
+%
+% Interviewer: Mmmhmm
+%
+% Rebecca: And I don't know how to do that. Because I've made arrays of
+% characters and integers and stuff before /OK/ but never pointers, so.
+% (Interview 4 of 5, April 6, 2012)
+%
+% Rebecca was seeing a temporal problem: how could she reference an array
+% that didn't exist yet?\footnote{As a crude analogy, imagine trying to
+%   write your own will. If you think forward in time, you might later
+%   have children to whom you'd like to bequeath your assets. But how can
+%   you enumerate those children in the will now, before you've had them?
+%   You need your will to refer to something that does not exist at the
+%   time you write the will. Extrapolating forward and rather ludicrously,
+%   we could imagine you are always capable of producing children. So, up
+%   until the day of your death, there is always the seeming risk that
+%   your will might need to refer to a beneficiary who does not yet exist.
+%   Recognizing this subtle problem as Rebecca did is an act I consider to
+%   be sense-making, even though she as-yet still did not have a solution.}
+%
+% Rebecca started exploring ways to solve that problem of getting tracks
+% to point to the array of pointers.
+%
+% Rebecca: But I'm not sure where it (tracks) would point to, because you
+% don't have that array made yet. So I \{3 sec pause\} so you'd have to, I
+% guess you could make the array, and then make it point to the array, but
+% \{3 sec pause\} Cuz you have to make an array of pointers.
+%
+% Interviewer: Mmmhmm
+%
+% Rebecca: And I don't know how to do that. Because I've made arrays of
+% characters and integers and stuff before /OK/ but never pointers, so.
+% (Interview 4 of 5, April 6, 2012)
+%
+% Rebecca landed on a potential solution: make the array first, then make
+% tracks point to the array. But, with that solution came another problem:
+% how do you make an array of pointers?
+%
+% \paragraph{Rebecca could propose candidate syntax and build pseudocode
+% around
+% it.}\label{rebecca-could-propose-candidate-syntax-and-build-pseudocode-around-it.}
+%
+% Rebecca came up with an idea to create an array of pointers, but she
+% wasn't sure if it would work.
+%
+% Rebecca: if I were to d---when I declared the array, if I were to---I
+% dunno if you even can, like, whenever we---we do ints, and then it would
+% be like star-p /Mmmhmm/ and then star-p would be the pointer /ok/ I
+% don't know if you can do like int star-p
+% \textbar{}\textbar{}bracket-bracket\textbar{}\textbar{}
+% \textbar{}\{makes square brackets with hands\}\textbar{} and that would
+% be an array or not.
+%
+% Interviewer: {[}{[}OK
+%
+% Rebecca: But I don't know{]}{]} if you can do that.
+%
+% Notice what's happening here. Rebecca is taking patterns that she knows
+% work:
+%
+% int age; // creates an integer variable called age
+%
+% int ages{[}35{]}; // creates an array called ages,
+%
+% // 35 integers long
+%
+% int *p; // creates a pointer-to-an-integer;
+%
+% // said pointer is called p
+%
+% and considering a pattern that might work:
+%
+% int *p{[}35{]}; // Rebecca is not exactly sure what this will do
+%
+% Notice further that in this exchange, Rebecca wondered ``if you can do
+% that.'' Determining the referent of \emph{that} was, again,
+% consequential. One possibility was Rebecca meant ``int *p{[}35{]}''
+% might not be the proper syntax for creating an array of 35
+% pointers-to-ints. If so, Rebecca was struggling with her aforementioned
+% problem of not being able to translate her logic to code. Another
+% possibility was Rebecca meant ``I don't know if there exists \emph{any}
+% syntax in C to create an array of pointers.'' So, I followed up on this
+% latter interpretation. We both agreed it \emph{should} be possible to
+% create an array of pointers, it was indeed just a question of whether
+% that syntax accomplished it (Interview 4 of 5, April 6, 2012). So, that
+% left the former interpretation: it looked like Rebecca was unsure how to
+% ``transfer her logic to code.''
+%
+% We both agreed creating an array of pointers should be possible. So, I
+% asked Rebecca to suppose her candidate syntax worked. She readily
+% agreed.
+%
+% Interviewer: OK. So then, um, what if we just pretended for a minute,
+% that that, like==
+%
+% Rebecca: ==That that works==
+%
+% Interviewer: ==That that worked==
+%
+% Rebecca: ==OK==
+%
+% Interviewer: ==OK. So then, um. \{begins writing\} So then you might
+% write like to make an array of \{stops writing\} actually what would we
+% call this? This is==
+%
+% Rebecca: ==array of pointers, I guess==
+%
+% Interviewer: ==OK. \{under breath\}
+% \textbar{}\textbar{}pointers\textbar{}\textbar{} \textbar{}\{writes
+% ``pointers''\}\textbar{}. It'd be, well. So. The---the one we had was
+% like \textbar{}\textbar{}int, star p, um, it would be some
+% number\textbar{}\textbar{} \textbar{}\{writing\}\textbar{}, um, and I
+% guess that'd be it in order to just declare /Yes/ that right? OK.
+%
+% In this exchange note every turn boundary is latched. Once I bid for
+% pretending, Rebecca accedes. When I'm unsure what to call the array,
+% Rebecca finishes my sentence. Rebecca rapidly approves the final product
+% before I even finish my last sentence. Figure 10 below is the syntax I
+% wrote in the exchange.
+%
+% \includegraphics[width=5.12500in,height=1.09792in]{media/image17.emf}
+%
+% If that syntax created the array-of-pointers, the next challenge for
+% storing track names was how to access elements in array-of-pointers.
+% Figure 11 presents the pseudo-code Rebecca ultimately wrote to scan in
+% and store track names.
+%
+% \includegraphics[width=5.36667in,height=1.30000in]{media/image18.emf}
+%
+% \protect\hypertarget{ux5fToc252445974}{}{}Figure ‑ -- Rebecca's
+% pseudo-code for scanning in and storing track names
+%
+% The next section will show the rich talk and gestures that accompanied
+% Rebecca's generation of this pseudo-code. I argue activity in the form
+% of gesture and talk constituted and sustained Rebecca's in-the-moment
+% approach to programming.
+%
+% \paragraph{Rebecca's talk and gestures constituted and sustained her
+% in-the-moment approach to
+% programming}\label{rebeccas-talk-and-gestures-constituted-and-sustained-her-in-the-moment-approach-to-programming}
+%
+% Throughout this section I detail the speech and gestures that accompany
+% Rebecca's pseudocode production. Along both modalities I saw evidence of
+% a marked shift in Rebecca's orientation. It was a shift from being
+% hunched, silent, thinking over the keyboard to being animated, open, and
+% gestural toward me. It was also a shift from seeming diffident,
+% uncertain, and hedging in speech to seeming confident, assured, and able
+% to dispatch my questions. Taken in their totality across the episode,
+% these subtleties constitute evidence of an approach that strongly
+% differs from those Rebecca had when discussing her struggles in the
+% course (section 3.5.1).
+%
+% First, Rebecca suggested she could write the pseudo-code as we talked.
+%
+% \protect\hypertarget{ux5fToc252445963}{}{}Table -- Rebecca starts
+% talking out pseudo-code and asks for the pen to write it
+%
+% \begin{supertabular}[]{@{}ll@{}}
+% \toprule
+% What she said & What she did\tabularnewline
+% \midrule
+% \endhead
+% Uh, well, what I was think---like if I wanted to like,
+% \textbar{}\textbar{}save it\textbar{}\textbar{} \textbar{}\{pinches
+% right hand, thumb and index finger a few inches apart\}\textbar{} or
+% whatever, /Yeah/ &
+% \includegraphics[width=2.68889in,height=3.23333in]{media/image19.png}\tabularnewline
+% I could make a while loop, \textbar{}\textbar{}scan in
+% the\textbar{}\textbar{} \textbar{}\{drags hand across the
+% table\}\textbar{} ---scan in the data from the album, /OK/ \{interviewer
+% begins writing\} uh, which is, {[}{[}I can write it if you want &
+% \includegraphics[width=2.66667in,height=3.23333in]{media/image20.png}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% Rebecca then took the pen and started describing her thinking while she
+% wrote.
+%
+% \protect\hypertarget{ux5fToc252445964}{}{}Table -- Rebecca's
+% verbal/gestural overview of looping through the input track titles
+% (Interview 4 of 5, April 6, 2012)
+%
+% \begin{supertabular}[]{@{}lll@{}}
+% \toprule
+% Panel \# & What she said & What she did\tabularnewline
+% \midrule
+% \endhead
+% 1 & And then my thinking at least, is you should be able to, um, say
+% that
+%
+% ``star p of i'' /mmhmm/ equals, uh, the title, and then you just do i++,
+% so then it'll \textbar{}\textbar{}move to the next
+% one\textbar{}\textbar{} \textbar{}\{makes looping gesture with left
+% hand\}\textbar{} /OK/ &
+% \includegraphics[width=2.03333in,height=2.33333in]{media/image21.png}\tabularnewline
+% 2 & and you just keep \textbar{}\textbar{}saving each of the
+% pointers\textbar{}\textbar{} \textbar{}\{left hand makes horizontal
+% chops in the air, like rungs down a ladder\}\textbar{} &
+% \includegraphics[width=3.40000in,height=1.92222in]{media/image22.emf}\tabularnewline
+% 3 & in the array \textbar{}\textbar{}to a title\textbar{}\textbar{}
+% \textbar{}\{right hand makes pinching motion, left-hand points to
+% it\}\textbar{} &
+% \includegraphics[width=2.18889in,height=2.24444in]{media/image23.png}\tabularnewline
+% 4 & \textbar{}\textbar{}And\textbar{}\textbar{} \textbar{}\{right hand
+% makes a large loop counterclockwise (from Rebecca's
+% perspective)\}\textbar{} &
+% \includegraphics[width=3.13333in,height=1.54444in]{media/image24.emf}\tabularnewline
+% 5 & \textbar{}\textbar{}you just increment by 1\textbar{}\textbar{}
+% \textbar{}\{right hand makes a cycloid to her right, in a plane parallel
+% to the back wall, like counting dots on a line\}\textbar{} until you
+% reach the end of file. &
+% \includegraphics[width=1.06667in,height=3.62222in]{media/image25.emf}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% ``Like, that would make sense to me,'' Rebecca said. I asked whether the
+% while loop would get a fresh line of text when it runs again.
+%
+% \begin{supertabular}[]{@{}lll@{}}
+% \toprule
+% Panel \# & What she said & What she did\tabularnewline
+% \midrule
+% \endhead
+% 1 & Yeah, because{]}{]} uh, \textbar{}\textbar{}what the while loop
+% does\textbar{}\textbar{} \textbar{}\{pinches right thumb and
+% forefinger\}\textbar{} &
+% \includegraphics[width=1.74444in,height=1.95556in]{media/image26.png}\tabularnewline
+% 2 & \textbar{}\textbar{}is it reads in the line\textbar{}\textbar{}
+% \textbar{}\{right hand, palm down, chest height, slides out to the
+% right\}\textbar{} &
+% \includegraphics[width=1.76667in,height=2.83333in]{media/image27.emf}\tabularnewline
+% 3 & and then \textbar{}\textbar{}once it reaches\textbar{}\textbar{}
+% \textbar{}\{right hand rises, lowers in a chop\}\textbar{} &
+% \includegraphics[width=1.78889in,height=2.11111in]{media/image28.png}\tabularnewline
+% 4 & the end of \textbar{}\textbar{}character it'll\textbar{}\textbar{}
+% \textbar{}\{right hand rises, swoops down in a crescent, index finger
+% extended, looping back up\}\textbar{} &
+% \includegraphics[width=2.74444in,height=1.36667in]{media/image29.emf}\tabularnewline
+% 5 & go back down---it'll \textbar{}\textbar{}do the while
+% loop\textbar{}\textbar{} \textbar{}\{right hand moves back down toward
+% table\}\textbar{} \textbar{}\textbar{}and then\textbar{}\textbar{}
+% \textbar{}\{right hand moves back up above shoulder-height,
+% palm-down\}\textbar{} &
+% \includegraphics[width=2.63333in,height=1.55556in]{media/image30.emf}\tabularnewline
+% 6 & \textbar{}\textbar{}go back\textbar{}\textbar{} \textbar{}\{right
+% hand creates a loop down and back up\}\textbar{} down to the next
+% thing---line, &
+% \includegraphics[width=3.38889in,height=1.31111in]{media/image31.emf}\tabularnewline
+% 7 & it'll \textbar{}\textbar{}read in the line\textbar{}\textbar{}
+% \textbar{}\{right hand scans rightward at shoulder height\}\textbar{} &
+% \includegraphics[width=1.61111in,height=2.47778in]{media/image32.emf}\tabularnewline
+% 8 & \textbar{}\textbar{}until it\textbar{}\textbar{} \textbar{}\{pinches
+% right thumb and forefinger, swipes right hand back to the left, palm
+% down\}\textbar{} reaches the end of file &
+% \includegraphics[width=3.38889in,height=1.45556in]{media/image33.emf}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% In panel 4, notice that Rebecca's gesture of swooping back actually
+% \emph{precedes} her saying ``go back down.'' In other words, her body
+% articulates the idea of cycling back before she actually speaks it. A
+% similar argument could be made for panel 8, where Rebecca's gesture for
+% a process hitting the bottom (hand swooping and hitting an inflection
+% point) precedes her saying ``reaches the end of file''
+%
+% I tried to sum up my understanding of Rebecca's description:
+%
+% Interviewer: So as you step through this loop /mmhmm/ \{points to i++\}
+% i keeps going up by one==
+%
+% Rebecca: ==Yes. (Interview 4 of 5, April 6, 2012)
+%
+% As I thought about that, Rebecca continued explaining and gesturing:
+%
+% \protect\hypertarget{ux5fToc252445965}{}{}Table -- Rebecca concludes her
+% visual and gestural explanation for scanning in track titles (Interview
+% 4 of 5, April 6, 2012)
+%
+% \begin{supertabular}[]{@{}lll@{}}
+% \toprule
+% Panel \# & What she said & What she did\tabularnewline
+% \midrule
+% \endhead
+% 1 & And the \textbar{}\textbar{}lines keep going
+% down\textbar{}\textbar{} \textbar{}\{left hand horizontally chops the
+% air, creating ``ladder rungs''\}\textbar{} & (Gesture partially obscured
+% by computer )\tabularnewline
+% 2 & so, \textbar{}\textbar{}that way,\textbar{}\textbar{}
+% \textbar{}\{left hand raises up to her head, palm down and parallel to
+% the table, and it is replaced by right hand, index finger extended and
+% pointing to her left\}\textbar{} &
+% \includegraphics[width=2.10000in,height=2.60000in]{media/image34.emf}\tabularnewline
+% 3 & \textbar{}\textbar{}the first line\textbar{}\textbar{}
+% \textbar{}\{right hand scans across to her right\}\textbar{} &
+% \includegraphics[width=1.57778in,height=3.50000in]{media/image35.emf}\tabularnewline
+% 4 & is going to \textbar{}\textbar{}be\textbar{}\textbar{}
+% \textbar{}\{right hand rotates to become pinched thumb and forefinger,
+% palm out\}\textbar{} , \textbar{}\textbar{}uh the
+% p---\textbar{}\textbar{} \textbar{}\{right-hand wiggles\}\textbar{} &
+% \includegraphics[width=3.25556in,height=1.62222in]{media/image36.emf}\tabularnewline
+% 5 & element zero /OK/ uh \textbar{}\textbar{}the
+% second\textbar{}\textbar{} \textbar{}\{hand swoops slightly up, then
+% curves down and locks in in a position below the prior one\}\textbar{}
+% one'll be element one &
+% \includegraphics[width=3.25556in,height=1.55556in]{media/image37.emf}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% \subsection{Conclusion}\label{conclusion-1}
+%
+% In pulling together concluding ideas, I revisit the points I established
+% at the end of section 3.1.
+%
+% \subsubsection{Students' early-stage design activity reveals patterns
+% outside the explanatory scope of (mis)conceptions
+% accounts}\label{students-early-stage-design-activity-reveals-patterns-outside-the-explanatory-scope-of-misconceptions-accounts}
+%
+% Lionel's strategy of working at a whiteboard, keeping himself at a
+% top-level of planning, and copying pseudo-code into a computer isn't
+% explained by appealing to ``concepts'' in computer science, which in
+% many research accounts are simply mappings of content in computer
+% science curricula. I demonstrate this gap in Appendix 4, where the
+% ``conceptual features'' of Lionel's code reveal little (if anything)
+% about the resources involved in Lionel's design process.
+%
+% Rebecca, in a similar fashion, has lots of productive knowledge for
+% thinking about arrays (section 3.5.2), but at the time of the interview
+% none of her code reflected that knowledge. Moreover, Rebecca was able to
+% sense-make (Danielak et al., in press) about an array-of-pointers
+% design, where her sense-making was again a complex activity that would
+% be poorly accounted for in CSEd frameworks that appeal to
+% misconceptions. In Rebecca's case, the situation is particularly
+% paradoxical if we consider conceptual knowledge about a topic to naively
+% be something students have or don't have.
+%
+% Suppose Rebecca had the requisite conceptual knowledge for thinking
+% about an array of pointers. It's a puzzle, then, to explain solely via
+% conceptual knowledge why she was so frustrated with programming and felt
+% stuck when she began Interview 4. Why hadn't the knowledge she
+% possessed---relevant to solving the problem at hand---manifested already
+% some way? Why was she stuck?
+%
+% The alternative is to assume that as of Interview 4 she didn't have the
+% conceptual knowledge for thinking about the problem at hand. But, that
+% doesn't make sense either. If Rebecca didn't know how to think about an
+% array of pointers, why was she able to do it so well during the
+% interview? Both possibilities---that either Rebecca had or did not have
+% conceptual knowledge about an array of pointers---lead to fairly
+% non-sensical conclusions under naïve have/don't have assumptions of
+% conceptual knowledge. Yet, again, the majority of conceptual knowledge
+% research in CSEd today is silent on the issue of degree when it comes to
+% conceptual knowledge: students either have a mental model that matches
+% canonical function or they don't.
+%
+% \subsubsection{Rebecca had epistemological resources to support
+% expert-like practices, but she framed those practices
+% differently}\label{rebecca-had-epistemological-resources-to-support-expert-like-practices-but-she-framed-those-practices-differently}
+%
+% Consider these summative statements made by Lionel and Rebecca on how
+% they use pseudo-code in their work.
+%
+% \protect\hypertarget{ux5fToc252445966}{}{}Table -- Comparing Lionel's
+% and Rebecca's views toward pseudo-code
+%
+% \begin{supertabular}[]{@{}ll@{}}
+% \toprule
+% Lionel & Rebecca\tabularnewline
+% \midrule
+% \endhead
+% ``on my computer I'd, you know I'd write out the pseudocode, I'd
+% ac---I'd literally write out the pseudocode, even though obviously it
+% wouldn't compile and actually work.'' (Interview 1 of 1, October 17,
+% 2011) & ``So, I can't just type in ``if the white piece reaches in,''
+% I---the, the C language, I guess, putting it in those, their terms,
+% their terminology into programming language.'' (Interview 2 of 5,
+% February 17, 2012)\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% As illustrative examples, they capture a fundamental difference in
+% practice between Lionel and Rebecca. Lionel worked in pseudo-code on his
+% computer, even going so far as to copy it into his source-code files
+% despite the fact that it ``obviously wouldn't compile and actually
+% work.'' Rebecca also wrote pseudo-code, and she even wrote it into her
+% source code files. But, her local sense of the activity of writing
+% pseudo-code was different. It came to represent a discontinuity with the
+% final code she was trying to write because there was nonetheless a gap
+% between writing in English and writing C in ``their terms.''
+%
+% These diverging stances take the same practice---in this case,
+% pseudo-coding---and situate it within a different kind of
+% epistemological coherence. For Lionel, the coherence associated with
+% pseudo-coding is productive and optimistic. Copying non-working
+% pseudo-code into the computer is, in a colloquial sense, all part of the
+% plan. It's a legitimate step toward creating a final working program
+% and, crucially, it reflects his own understanding of what's supposed to
+% happen in his code. For Rebecca, by contrast, writing pseudo-code is a
+% fallback. Pseudo-code is what she writes ``if I know what I want to put
+% underneath of it,'' by which she means if she's not sure how to flesh
+% out a loop or other control structure. Rather than being part of a
+% planful coherence, the practice of pseudo-coding for Rebecca gets
+% triggered as part of a stopgap coherence; a contingent measure at times
+% associated with a kind of diffidence.
+%
+% Crucially, Rebecca has intellectual resources for pseudo-coding.
+% Moreover, her practice of writing pseudo-code to flesh out a loop is the
+% plausible beginning of ``method stubbing,'' the process by which one
+% might defer implementing the guts of a function and instead create a
+% simple stopgap. For example, the function below will always print 28
+% Fahrenheit, which is fairly useless as far as temperature-getting
+% functions go:
+%
+% getOutsideTemperature \textless{}- function(sensor) \{
+%
+% print(``28F'')
+%
+% \}
+%
+% But, the advantage of having \emph{something} in the function body, no
+% matter how trivial, is enormous. Now, because
+% \textbf{getOutsideTemperature} is defined and takes arguments, other
+% functions can safely call it. It also offers a diagnostic output---by
+% printing ``\textbf{28F}'' out---that we can rely on if we ever need to
+% make sure the function was called. At the time of creating
+% \textbf{getOutsideTemperature}, a programmer may have no idea how the
+% temperature will be obtained but will nonetheless need to be able to
+% write code that depends on it getting the temperature. Stubbing is, in
+% such cases, a highly productive thing for a programmer to be able to do.
+% But, stubbing is as much a strict practice as it is an epistemological
+% move, because it represents a decision that an entity-to-be-known can
+% be, for the moment, underspecified so it can be incorporated into a
+% system before its behavior is defined.
+%
+% Rebecca is already capable of simple stubbing. It's part of an
+% epistemological coherence that's triggered when she doesn't know how to
+% flesh out a control structure. So, from a constructivist standpoint
+% (Smith et al., 1993), she has knowledge of something that can be further
+% refined into software engineering knowledge. Indeed, much of the
+% surrounding data from Rebecca suggests that she would have been served
+% well if an instructional intervention had taught her some simple ways to
+% stub out functions so she could call them even if she hadn't defined
+% their bodies yet. To my knowledge, she was never exposed to stubbing
+% formally, which is unfortunate considering how much it might have led to
+% improving her confidence in programming.
+%
+% Lastly, consider Rebecca's episode in reasoning about an array of
+% pointers (section 3.5.2). As I show, Rebecca clearly had resources for
+% reasoning about the sensibility of an array-of-pointers design and for
+% predicting array-of-pointers syntax by extrapolating from patterns she
+% had seen before. And, when I explicitly suggested that we suppose a
+% candidate syntax works, Rebecca was able to fluidly articulate the
+% procedure she would build around that syntax (section 3.5.2.3). Rebecca
+% knew things, but her knowledge was deployed in such a highly contextual
+% way as to be sensitive to what kind of knowledge-activity she thought
+% she was supposed to be doing. It's thus sensible to model that set of
+% phenomena from an epistemological standpoint, rather than a strictly
+% conceptual one. Rebecca had certain kinds of knowledge, but part of my
+% in-interview intervention was to establish a frame where \emph{supposing
+% syntax worked} was a valid part of programming. Once we mutually
+% negotiated that frame there was a strong microcoherence of Rebecca
+% stably explaining her code. But, crucially, my intervention wasn't about
+% introducing conceptual content but rather establishing a kind of
+% knowledge game---supposing---that one could play as part of programming
+% design.
+%
+% \subsubsection{Students displayed a diversity of approaches to
+% programming in the
+% moment}\label{students-displayed-a-diversity-of-approaches-to-programming-in-the-moment}
+%
+% The diversity of programming approaches I saw students take is far
+% greater than just the space spanned by Lionel and Rebecca. Within and
+% across my data, I continually observed students saying and doing things
+% that not only distinguished them from each other, but constituted
+% phenomena I haven't seen described in prior CSEd research. I saw
+% students employ clever tests for debugging, create their own makeshift
+% debugger, and discuss at length whether certain programming constructs
+% represented natural ways of thinking. For illustrative purposes, I'll
+% discuss three such examples. While I don't analyze them at length, I
+% include them here as proof of principle of the kinds of phenomena we're
+% not currently capturing that nonetheless have a strong effect on
+% students' programming approaches. Moreover, these ``in the wild''
+% (Hutchins, 1995a) practices are further evidence of the kinds of
+% knowledge that could be further refined into student expertise (Smith et
+% al., 1993). So, not only do our research accounts miss these kinds
+% phenomena, by missing them they preclude research and practice from
+% building off them.
+%
+% \paragraph{Isaac used a thoughtful debugging strategy that code
+% snapshots alone could never
+% capture}\label{isaac-used-a-thoughtful-debugging-strategy-that-code-snapshots-alone-could-never-capture}
+%
+% When Isaac was debugging an error in his checkers game code, he made use
+% of a clever test to discover he had reversed array index subscripts. He
+% had become uncertain about which of the two subscripts corresponded to
+% what he thought of as the x-coordinate on the board and which controlled
+% the y-coordinate. So, he took a piece whose x-y position he knew and
+% instructed his program to print it as a ``7'' on the board. Then, he
+% told it to increment the first index of that known piece by 1, leave the
+% second index alone and print the resulting piece on the board as ``8''.
+% Finally, he told his program to increment the second index of the known
+% piece by 1, leave the first piece alone, and print the resulting piece
+% on the board as ``9''. By inspecting the visual output, Isaac could tell
+% which subscript controlled which position because he knew where 8 and 9
+% were relative to the unchanged 7. When he reflected on his work in the
+% interview, he explained that what was tricky was realizing that the
+% coordinate system for the checkers board didn't work the way standard
+% coordinate systems in math worked; in the checkers program the indices
+% were reversed.
+%
+% Because Isaac had been set up as a code-snapshot participant, we have a
+% snapshot record of his work for that portion of the interview. But, the
+% only snapshot-visible changes are the reversal of the indices and the
+% accompanying test cases. The narrative surrounding Isaac's particular
+% debugging text, including his reflection that coordinate systems in
+% programming are ``trickier'' than those in math, is entirely invisible
+% to the snapshot record. If I hadn't interviewed him I never would have
+% captured it.
+%
+% \paragraph{Dana created her own debugging
+% environment}\label{dana-created-her-own-debugging-environment}
+%
+% Dana was working on her checkers program and needed to debug some errant
+% behavior. In class and in discussion students had been exposed to the
+% GNU Debugger (GDB), so in the interview I asked Dana if she'd like to
+% use it. She said she didn't feel comfortable using GDB. Instead, using
+% my MacBook, Dana decided to create three separate instances of a
+% terminal. In one instance, she had her checkers program source code
+% open. In the second, she had the source code for a set of test inputs
+% she was developing to try to pinpoint the error. The third window was
+% dedicated to compiling and running the code. Since the program was
+% designed to print an ASCII (text) representation of the board at each
+% turn, the dedicated program window was Dana's visual output for which
+% pieces were where at any state in the game.
+%
+% The important finding from this episode is how Dana deliberately
+% structured her environment to support her activity. At the lowest level,
+% she could have tried to do everything in a single terminal window. But,
+% if she did that she'd have to constantly switch contexts. Because
+% students in the course used a terminal version of emacs, she had no easy
+% way to simultaneously view her code while compiling it. GDB would have
+% been in many respects the right tool for that problem: it would let her
+% set breakpoints, step through iterations of code, and inspect variable
+% values as she did so. But Dana chose not to use GDB.
+%
+% Instead, Dana exploited the fact that Terminal.app on Mac OS X 10.7 can
+% have multiple instances of itself open at the same time on the same
+% screen. With three terminals running, she could persistently have her
+% source code visible and executing at the same time. With one glance
+% pattern she could move from the board's visual output to the test to the
+% program source and back again. Collectively, the field of displays
+% enabled a kind of programming that was substantially different---in
+% terms of its representational affordances---from using GDB or a single
+% terminal instance. Moreover, a snapshot-history-only research approach
+% would have never had access to how or why Dana (re)configured her
+% development environment.
+%
+% \paragraph{Toby said recursion was the ``hardest programming way to
+% think''}\label{toby-said-recursion-was-the-hardest-programming-way-to-think}
+%
+% When I gave Toby a code sample task in an interview, he immediately
+% commented on the fact that it used recursion. Recursion was, as he said,
+% ``the hardest programming way to think.'' At one point his exact
+% phrasing, quoting the film \emph{Ice Age}, was that ``recursion is bad
+% juju.'' The task in question was adapted from an example in Abelson and
+% Sussman (1996) that approximates the square root of a number using an
+% elegant recursive approach.
+%
+% Toby's first explanation for why recursion was bad was that it's a
+% completely unnatural way to think. He cited examples from cooking by
+% saying, in effect, ``Nobody ever stirs by saying stir once, and if it's
+% not stirred, stir once. They say stir 100 times.'' He cited simple
+% operations---including incrementing a number---where using recursion to
+% push and pop a stack seemed needlessly complex. One of his most damning
+% indictments of recursion was that his class introduced it as a solution
+% to generate numbers in the Fibbonacci sequence. Of course recursion
+% works for examples that are mathematically recursive, he reasoned, but
+% outside of those obviously contrived examples it was bad joo joo when
+% compared to understandable, sensible iteration.
+%
+% The fuller account of Toby's resistance to recursion---and slight change
+% of heart after an instructional moment---is beyond the scope of what I
+% can present here. Suffice it to say that Toby's feelings toward
+% recursion---rooted in a sense of it being an unnatural way to
+% think---strongly directed his approach to programming. The approach was
+% so strong, in fact, that during our first interview he wasn't fully able
+% to produce working code examples of why recursion was absurd despite his
+% vehement conviction that it was. In other words, he maintained that
+% recursion was bad and inefficient even though every time he tried to
+% demonstrate its inferiority he made mistakes in his code and his
+% examples didn't work.
+%
+% \subsubsection{Dynamic epistemological models can offer a lens for
+% reforming assessment and
+% instruction.}\label{dynamic-epistemological-models-can-offer-a-lens-for-reforming-assessment-and-instruction.}
+%
+% Historically, a recognition of in-pieces cognitive dynamics in science
+% education (diSessa, 1993; Hammer, 1994) has led to more direct research
+% on how those frameworks inform instruction (Hammer \& Elby, 2003;
+% Hammer, 1996; Louca et al., 2004). What I describe here is research
+% still at the formative end of generating models to explain cognition.
+% And, because the focus of the research was explicitly on learning and
+% not instruction, I advise prudence in trying to draw certain kinds of
+% specific recommendations from the work I present.
+%
+% Those restrictions being said, we can think carefully about what this
+% work means for instruction. For practical purposes, let's refine
+% terminology so we can talk more specifically and precisely about
+% components of teaching. Below are my working definitions for teaching
+% components:
+%
+% \begin{itemize}
+% \item
+%   How should instructors change the way they act in the classroom
+%   (\emph{instruction})?
+% \item
+%   How should this research inform the set of intended learning outcomes
+%   instructors prepare (\emph{curriculum})?
+% \item
+%   How should we change the way we measure or otherwise go about
+%   ascertaining what students know (\emph{assessment})?
+% \end{itemize}
+%
+% The clearest implications of my work are in instruction and assessment.
+%
+% From an instruction perspective, it may help teachers to know that
+% students can enact different epistemological stances as part of their
+% programming practice. As stated, that finding is in principle not new
+% (cf., Gaspar \& Langevin, 2007). What is new, I think, is the
+% characterization that these stances evidence epistemological resources
+% that can be tapped for learning (cf., Hammer \& Elby, 2003). That
+% sense-making about program design choices or computational concepts is a
+% worthwhile activity, for example, is a message instructors can send to
+% students that resonates with larger findings in engineering education
+% research (Danielak et al., in press). Moreover, instructors can be more
+% sensitive to what kinds of knowledge-activity students think they're
+% doing, particularly when making epistemological moves like
+% \emph{supposing} might help students work through design problems.
+%
+% My findings also speak to assessment. In the course I studied, there
+% were few (if any) assessment instruments that would have revealed the
+% kinds of knowledge I document Lionel and Rebecca having in this study.
+% At no point in my observational data of the course were students asked
+% to explain a design choice or reason about competing solutions to handle
+% a problem. At no point were they encouraged to deeply sense-make about
+% the conceptual content of the course, particularly the traditionally
+% tricky topics of pointers and pointer arithmetic. At no point in my data
+% did the instructor model the kind of sense-making Rebecca did in her
+% interviews, a fact which she noted at one point as distinguishing the
+% course from Basic Programming, its first-semester counterpart. Because
+% the course never explicitly asked students to sense-make, reason about
+% design choices, or explain designs, it never used information about what
+% knowledge students had about those things to inform its instruction. It
+% couldn't. Moreover, because the course never explicitly asked students
+% to sense-make, reason about design choices, or explain designs, I was
+% left to conclude such things were not prioritized learning outcomes of
+% the course. To be clear, all courses establish a kind of focus by
+% deciding what won't be covered, and I don't fault the instructor at all
+% for running a course where design knowledge wasn't an enacted learning
+% outcome. But, I'm also left to wonder: why wasn't it? Given that even
+% beginning students can think about software design, and given that
+% engineering as a discipline fundamentally involves design, shouldn't it
+% have been?
+%
+% \section{Conclusion}\label{conclusion-2}
+%
+% If there is one overarching finding from this dissertation, it's that
+% students clearly have resources for thinking about designing programs
+% and a diversity of approaches to programming in the moment. Below, I
+% explain the kind of diversity I saw in programming approaches. Then, I
+% conclude with a discussion of what my research might mean for
+% assessment.
+%
+% \subsection{We should think carefully about what students' programming
+% design knowledge means for
+% assessment}\label{we-should-think-carefully-about-what-students-programming-design-knowledge-means-for-assessment}
+%
+% My research helps us document and model the kinds of knowledge students
+% have. In so doing, it points out kinds of information that the course's
+% assessments were apt to miss:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   How students frame or otherwise approach the task of programming in
+%   the moment
+% \item
+%   The role of different kinds of prior experience in stabilizing (or
+%   potentially destabilizing) certain kinds of frames
+% \item
+%   In-the-moment practices---including talking out solutions, sketching
+%   out debugging strategies, and writing out pseudo-code---that display
+%   students' competence and sense-making
+% \item
+%   The code history that traces how designs evolve, including how
+%   students start projects and what parts of their designs become
+%   dead-ends
+% \end{enumerate}
+%
+% Together, those four points are relevant for formative assessment (Black
+% \& Wiliam, 1998) in introductory programming. (1) and (2) point us
+% toward what students think they're doing when they're programming.
+% Analogous work from science education tells us that students' sense of
+% the kind of knowledge activity they're enacting matters for learning and
+% assessment (Russ et al., 2008). For example, knowing early on that
+% students are blindly copying code or randomly trying syntax gives
+% instructors a chance to intervene. But, my research suggests
+% intervention can't just be about stopping a bad behavior: knowing that
+% copy-paste behavior is happening is different from knowing \emph{why}
+% it's happening.
+%
+% Formative assessment has to be about diagnosing causes, not just
+% identifying symptoms. Otherwise, we run the risk of ignoring or even
+% harming students' productive knowledge. Take the example of copy-paste
+% behavior, as documented by Gaspar and Langevin (Gaspar \& Langevin,
+% 2007). One reason we may see copy-paste behavior, as Study 1
+% demonstrates, is that students might see situations as new instances of
+% already-solved problems. In professional practice, seeing old solved
+% problems in new situations can be productive. Such insights can, for
+% example, direct engineers to use a pre-built library of functions
+% instead of building their own. But, another related reason for
+% copy-paste may be efficiency. For Rebecca, copying and pasting code was
+% very fast; it required only a few quick keystrokes. In the short run,
+% Rebecca's choice to copy the code was a faster, less demanding, more
+% trustworthy route to go on than was abstracting the code to a function.
+% When we consider students trying to see common problems in new scenarios
+% and solve such problems efficiently, ``copy-paste'' becomes a symptom
+% rather than a root cause. Epistemological frameworks, then, can inform
+% assessments by offering explanations that aim at the root cause of
+% certain behaviors.
+%
+% Points 3 and 4 complement knowledge analysis by drawing focus to
+% artifacts, practices, and history. As data from Lionel shows (Study 2),
+% a crucial part of his design process involves artifacts and activities
+% that were only distally knowable to the assessments he got in class.
+% Lionel's instructor had no direct access to:
+%
+% \begin{itemize}
+% \item
+%   Lionel's whiteboard
+% \item
+%   the hours Lionel may have spent working out designs in chalk
+% \item
+%   how Lionel might have talked out design features
+% \item
+%   how Lionel's initial ``pseudo-code'' evolved into his final design.
+% \end{itemize}
+%
+% In Rebecca's case, commented-out code in her final project submission
+% were perhaps the only clues at all that she tried abstracting some
+% flight day-checking procedures into functions (Study 1). Without
+% Rebecca's history, the instructor would have had no reliable way of
+% knowing:
+%
+% \begin{itemize}
+% \item
+%   Rebecca began her flights database project by copying code from an
+%   earlier project
+% \item
+%   Rebecca's original solution for day-checking evolved from a seven-fold
+%   conditional structure, one for each day
+% \item
+%   Rebecca may have tried abstracting day-checking procedures, even
+%   creating a function called check\_days.
+% \end{itemize}
+%
+% To sum up: traditional assessments can tell us what students finally
+% produce but not how they produce it, when they produce it, why they
+% produce it, or what the production process was. In the summative
+% assessment structure of the course, final student products were the only
+% products submitted to the instructor. My research shows that what's
+% happening in the interstices---before typing, between compiles, away
+% from the computer---can be captured and, in principle, analyzed and
+% acted upon by instructors.
+%
+% \subsection{What We Might Change About Classroom
+% Practice}\label{what-we-might-change-about-classroom-practice}
+%
+% What follows is my speculation about how we might specifically change
+% classroom practices in light of my research findings. I move from
+% recommendations I think are most strongly supported by my data to
+% recommendations that align with my findings but are more expansive, and
+% thus less strongly supported. As my recommendations broaden, I try to
+% offer not only an instructional recommendation but a concomitant
+% question for research.
+%
+% \subsubsection{Instructors could look beyond content to understand
+% student
+% difficulties}\label{instructors-could-look-beyond-content-to-understand-student-difficulties}
+%
+% I think first and foremost, instructors have to have a willingness to
+% recognize that the difficulties they think they're seeing in students
+% are difficulties they may be viewing through a lens of content. For
+% example, after seeing a student struggle an instructor might say ``the
+% student doesn't know assignment statements.'' But, those difficulties
+% almost certainly have a deeper explanation. I say ``almost certainly,''
+% where what I mean is that there are a number of patterns we've
+% identified as common novice errors, but most research stops before
+% asking why that's a common novice error. With rare exceptions (Fleury,
+% 1991, 2000), the computing education community doesn't encourage asking
+% the question \emph{why might a student be doing this?} and \emph{what
+% might this tell me about the way a student is making sense of this?} By
+% contrast, my research suggests those orientations are ripe, low-hanging
+% fruit for instructors to take on. I think one of the first things as
+% instructor could ask is, \emph{why would a student be thinking that this
+% is the appropriate thing to be doing?} And that holds true whether the
+% thing in question is writing the statement this way, or interpreting the
+% code this way, or enacting this kind of programming activity this way.
+%
+% Moreover, we know from research on metacognition (Schoenfeld, 1987,
+% 1992) that thinking beyond content opens the palette of kinds of
+% interventions an instructor can make. Schoenfeld (1987), for example,
+% came to encourage metacognition in his classroom by doggedly asking
+% student groups what they were doing, why they were doing it, and how
+% they hoped it would lead them to a solution. Students ultimately came to
+% internalize such strategies and spent far less time floundering along
+% solution paths that weren't ultimately productive. Might the same idea
+% be true in introductory programming courses? To find out, we could begin
+% researching in earnest how often and in what ways instructors model
+% metacognition for their students. Currently, I would argue, we don't
+% know whether and how instructors do so.
+%
+% \subsubsection{Instructors could use code history to inform
+% interventions}\label{instructors-could-use-code-history-to-inform-interventions}
+%
+% When, for example, a student comes to office hours with a problem on a
+% project, an instructor needs to quickly come to grips with the state of
+% the students' project, the logic of their design, where they're stuck,
+% why they're stuck, and what might best help them. That's no small task.
+% But, given my analysis I have strong reason to believe that having code
+% history available to instructor could change both the nature of coming
+% to terms with a student's project and the conversation with a student
+% that results. If an instructor can see the evolution of a student's
+% code, at the very least they could see where the student started, how
+% the code was growing, and where the student was working most recently.
+% Even if the instructor had never before seen the code (or its history)
+% until that moment in office hours, having both available changes the
+% kind of view the instructor can get of the code and the specifics of the
+% intervention that might result. In Rebecca's case, for example, it might
+% have been an opportunity to explore why her check\_days abstracted
+% function was failing on her flights database project.
+%
+% Having code-snapshot capabilities also changes the kind of research we
+% can do. First and foremost, a result of this dissertation was to create
+% a freely-available, lightweight, open-source framework for capturing and
+% visualizing students' code histories.\footnote{https://github.com/briandk/gitvisualizations}
+% So, the most basic kind of study would involve deploying that framework
+% from the instructional side of a course and exploring what happens, as,
+% for example, Hurd (2013) has done. One could ask questions of how
+% instructors build code snapshotting into their course, how such
+% information could or did inform assessment, and how such information
+% could or did change the nature of instructional interactions with
+% students. Was it for the better? If so, how do we know? If not, how do
+% we know?
+%
+% \subsubsection{Instructors could establish a norm of asking why a design
+% choice makes
+% sense}\label{instructors-could-establish-a-norm-of-asking-why-a-design-choice-makes-sense}
+%
+% The most far-reaching implication of the research I present here is that
+% instructors should establish a norm in their classes where anyone, at
+% any point, for any piece of code, is allowed to ask \emph{why does this
+% make sense as a design choice? Why is that an obvious choice to make?
+% How does that work?} Consider a parallel example from mathematics
+% education. In Don Saari's calculus class at UC Irvine, he ``invokes the
+% principle of what he calls ``WGAD''---``Who gives a damn?'' (Bain, 2004,
+% pp. 38--39). Bain explains:
+%
+% At the beginning of his courses, he tells his students that they are
+% free to ask him the question on any day during the course, at any moment
+% in class. He will stop and explain to his students why the material
+% under consideration at that moment---however abstruse and minuscule a
+% piece of the big picture it may be---is important, and how it relates to
+% the larger questions and issues of the course. (Bain, 2004, p. 39)
+%
+% What I'm suggesting is even broader than that, because it's not just a
+% question students can ask of professors; it's a question \emph{anyone}
+% can ask of \emph{anyone.} I see it as the programming and design
+% extension of a sociomathematical norm (Yackel \& Cobb, 1996), and I
+% think it could lead to collaborative sense-making. That is, I'm asking
+% for an accepted cultural practice that is also itself a design practice,
+% whereby an instructor can help create a safe, stable space for a
+% community to be reflective and critical about design choices. I also
+% think it's an idea that leads to others, such as letting, if not
+% requiring, students review one another's code. When students are forced
+% to reckon with someone else's code and understand their design
+% decisions, they're also forced to justify their own decisions. And
+% establishing ``why that design choice?'' or ``how does that make
+% sense?'' as a norm provides reciprocal opportunities for students, not
+% just instructors, to improve the code of others.
+%
+% The research questions that come out of such an idea would include:
+%
+% \begin{itemize}
+% \item
+%   How can an instructor satisfactorily establish norms about design in a
+%   classroom?
+% \item
+%   What does it look like when students collaboratively sense-make about
+%   a program's design choices?
+% \item
+%   What does it look like when students are asked to reflect on their own
+%   design choices?
+% \item
+%   How might introducing a code review component into a course change the
+%   way students approach design? How might it improve students'
+%   conceptual understanding? How might it improve the quality of the code
+%   students produce?
+% \end{itemize}
+%
+% \subsection{Final Remarks}\label{final-remarks}
+%
+% How students design programs matters for learning and instruction in
+% engineering. It matters because finished code reflects what students
+% know about design, whether or not instructors capture such information.
+% It matters because students have resources for learning about and
+% engaging in design; whether or not curriculum, instruction, and
+% assessment choose to tap into those resources. It matters because design
+% should be an intellectual thread that runs through all engineering
+% courses. That thread shouldn't stop when we introduce students to
+% programming.
+%
+% \section{Appendix 1 -- Transcript
+% conventions}\label{appendix-1-transcript-conventions}
+%
+% \begin{itemize}
+% \item
+%   Turns are not numbered, but they are blank-line-delimited
+% \item
+%   Short interjected speech that does not interrupt a speaker's turn is
+%   bounded /by slashes/
+% \item
+%   Matching double brackets show the {[}{[}onset and termination{]}{]} of
+%   overlapping talk across turns.
+% \item
+%   *emphasized speech* is bounded by asterisks
+% \item
+%   Parenthetical clarifications by the analyst appear in parentheses. (He
+%   considered but rejected square brackets.)
+% \item
+%   matching double equals signs mark turn boundaries== ==with minimal or
+%   no audible silence (also known as latching turns)
+% \item
+%   Where gestures don't overlap speech they are in-lined by curly braces
+%   when they happen \{folds arms, having made his point\}. All gestures
+%   enacted \emph{by} a speaker appear within that speaker's turn unless
+%   otherwise noted \{smiles, self-satisfied at having made this important
+%   clarification\} \{audience scoffs\}.
+% \item
+%   When gestures happen \emph{during} speech, the speech is presented
+%   first and bounded by double pipes. Gestures that happen simultaneously
+%   with such speech are bounded by curly braces, nested within double
+%   pipes, and immediately follow the speech they overlap. For example:
+% \end{itemize}
+%
+% Throwing chainsaws \textbar{}\textbar{}*up*\textbar{}\textbar{}
+% \textbar{}\{throws chainsaw\}\textbar{} is easy. Just be careful when
+% they come \textbar{}\textbar{}*down*\textbar{}\textbar{}
+% \textbar{}\{catches chainsaw for a punctuated finish\}\textbar{}.
+%
+% \section{Appendix 2 -- Visual conventions for
+% gestures}\label{appendix-2-visual-conventions-for-gestures}
+%
+% A challenge of this dissertation was trying to use still-frames to
+% convey motion and animation. Where I could, I annotated pictures with
+% arrows to show trajectories of movement:
+%
+% \includegraphics[width=5.62222in,height=0.87778in]{media/image38.emf}
+%
+% \begin{supertabular}[]{@{}ll@{}}
+% \toprule
+% \textbf{Description} & \textbf{Example}\tabularnewline
+% \midrule
+% \endhead
+% \textbf{Blue} arrows project forward to show action that will occur but
+% hasn't yet in the frame you're looking at. The origin is where a hand
+% \emph{is}; the terminus is where it \emph{will be} shortly. &
+% \includegraphics[width=1.67778in,height=1.58889in]{media/image39.png}\tabularnewline
+% \textbf{Pink} arrows show action that already occurred before the frame
+% you're looking at. The origin is where a hand \emph{was}; the terminus
+% is where it \emph{is}. &
+% \includegraphics[width=1.65556in,height=2.37778in]{media/image20.png}\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% \section{\texorpdfstring{Appendix 3 -- Transcript of Rebecca's
+% pseudo-code episode \emph{without} gesture
+% codes}{Appendix 3 -- Transcript of Rebecca's pseudo-code episode without gesture codes}}\label{appendix-3-transcript-of-rebeccas-pseudo-code-episode-without-gesture-codes}
+%
+% Interviewer: OK. So then, um, what if we just pretended for a minute,
+% that that, like==
+%
+% Rebecca: ==That that works=
+%
+% Interviewer: ==That that worked==
+%
+% Rebecca: ==OK==
+%
+% Interviewer: ==OK. So then, um. So then you might write like to make an
+% array of \{stops writing\} actually what would we call this? This is==
+%
+% Rebecca: ==array of pointers, I guess==
+%
+% Interviewer: ==OK. Pointers. It'd be, well. So. The---the one we had was
+% like int, star p, um, it would be some number, um, and I guess that'd be
+% it in order to just declare /Yes/ that right? OK. And then you're
+% saying, how would you like access an element of it?
+%
+% Rebecca: Uh, well, what I was think---like if I wanted to like, save it
+% or whatever, /Yeah/ I could make a while loop, scan in the---scan in the
+% data from the album, /OK/ uh, which is, I can write it {[}if you want
+%
+% Interviewer: Sure, go ahead{]}{]}
+%
+% Rebecca: Um, so, it was==
+%
+% Interviewer: {[}{[}Oh, sorry, just have to{]}
+%
+% Rebecca: {[}Oh{]}
+%
+% Interviewer: It's /Oh, OK/ one of those annoying ticks, but it works
+% best cause of the camera if, ah, /Ah, OK/ if that's pointing away from
+% your hand
+%
+% Rebecca: {[}OK{]}
+%
+% Interviewer: {[}That's fine{]}
+%
+% Rebecca: Um, there was album, and then, so there was, uh, percent d, and
+% percent s, and then give those names, just, I'll just call it number,
+% and title. /Mmmhmm/ And so, at least, my---until it does not equal EOF,
+% and then my thinking at least, is you should be able to, um, say that
+%
+% Rebecca: ``star p of i'' /mmhmm/ equals, uh, the title, and then you
+% just do i++, so then it'll move to the next one /OK/ and you just keep
+% saving each of the pointers in the array to a title /Mmmhmm/ And you
+% just increment by 1 until you reach the end of file.
+%
+% Interviewer: OK==
+%
+% Rebecca: ==Like *that* would make sense to me.
+%
+% Interviewer: So then, like, what would go in p of one would be the first
+% title we read in==
+%
+% Rebecca: ==Yes,
+%
+% Interviewer: uh, would, when the while loop runs again, does it get a
+% fresh line {[}{[}line from that
+%
+% Rebecca: Yeah, because{]}{]} uh, what the while loop does is it reads in
+% the line and then once it reaches the end of character it'll go back
+% down---it'll do the while loop and then go back down to the next
+% thing---line, it'll read in the line until it reaches the end of file
+%
+% Interviewer: So as you step through this loop /mmhmm/ i keeps going up
+% by one==
+%
+% Rebecca: ==Yes.
+%
+% Interviewer: Uh
+%
+% Rebecca: And the lines keep going down so, that way, the first line is
+% going to be, uh the p---element zero /OK/ uh the second one'll be
+% element one (Interview 4 of 5, April 6, 2012)
+%
+% \section{Appendix 4 -- Conceptual knowledge frameworks in computing
+% don't tell us much about how Lionel structured an in-interview
+% program}\label{appendix-4-conceptual-knowledge-frameworks-in-computing-dont-tell-us-much-about-how-lionel-structured-an-in-interview-program}
+%
+% In this section, I explore Lionel's work in solving an in-interview
+% programming task. First, I reproduce the prompt Lionel was given. Next,
+% I show the final source code of his solution to the problem, written in
+% C. In the analysis that follows, I move outward to explore the context
+% of that code's production.
+%
+% My choice to start with the code first is deliberate. I'm trying to
+% mimic what the instructor of a typical programming course might see: the
+% final submitted form of a student's code. This approach is admittedly a
+% bit backwards, since the data I show after I present Lionel's submission
+% is all about what happened \emph{before} the code reached its final
+% form. But, I think this approach useful because it invites us to explore
+% an artifact first, then raise questions about the conditions of its
+% production. In that sense, my presentation has the anthropological feel
+% of understanding a found object, which I think is a faithful way of
+% looking at what many instructors (not to mention researchers, and of
+% course active software developers) face in their day-to-day work.
+%
+% \subsection{\texorpdfstring{\protect\hypertarget{ux5fToc247188555}{}{\protect\hypertarget{ux5fToc252445950}{}{}}Analyzing
+% Lionel's Solution From a Conceptual Knowledge
+% View}{Analyzing Lionel's Solution From a Conceptual Knowledge View}}\label{analyzing-lionels-solution-from-a-conceptual-knowledge-view}
+%
+% To infer conceptual knowledge, I'll use Elliott Tew's (2010) conceptual
+% categories for procedural programming in first-semester computer science
+% (CS1) courses. My reasons for doing so are:
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   Elliott Tew (2010) sought concepts that were language-agnostic. That
+%   is, the concepts Elliott Tew (2010) identified are not idiomatic to
+%   just one language. Rather, they are general ideas implemented in
+%   nearly every major procedural language taught to undergraduates (Java,
+%   Python, C-based languages, Scheme).
+% \item
+%   Elliott Tew's (2010, pp. 23--27) selection process was extensive. She
+%   began with curricular documents, moved to canonical texts, then
+%   ultimately conducted a thematic analysis on emerging categories to
+%   produce her final list of concepts.
+% \item
+%   To date, Elliott Tew's (2010) identification of CS1 concepts is the
+%   only work I know of that has been carried forward to create,
+%   administer, and validate concept inventories in CS1 (Elliott Tew \&
+%   Guzdial, 2010, 2011)
+% \end{enumerate}
+%
+% Figure 2 below presents an overview of the conceptual features I
+% identified in Lionel's code. Even in a short, relatively simple program
+% Lionel exhibits four of the ten concept categories Elliott Tew (2010)
+% identifies. Moreover, he uses each of them ``correctly,'' so to speak,
+% in that not only does his program compile, it also properly finds the
+% range of the numbers it's given. It's also worth noting that Lionel's
+% code displays consistent, helpful use of whitespace (indentation) as
+% well as informative comments. While these attributes aren't conceptual,
+% per se, they are the sorts of features an instructor might consider in
+% assigning a grade to this program.
+%
+% \includegraphics[width=5.60000in,height=5.67778in]{media/image40.jpeg}
+%
+% \protect\hypertarget{ux5fToc252445975}{}{}Figure ‑ -- Highlighting the
+% conceptual features of Lionel's code given Elliott Tew's (2010) outline
+% of CS 1 concepts
+%
+% \section{Appendix 5 -- Neverly-Asked Questions
+% (NAQs)}\label{appendix-5-neverly-asked-questions-naqs}
+%
+% \subsubsection{Neverly Asked Questions (NAQs) about my conceptual
+% analysis of Lionel's
+% code}\label{neverly-asked-questions-naqs-about-my-conceptual-analysis-of-lionels-code}
+%
+% \emph{Didn't Elliott Tew (2010) define those concepts as a way of
+% creating a concept inventory?}
+%
+% Yes. In her work, I believe each concept was represented by at least one
+% multiple-choice question (MCQ) with carefully-chosen distractors. In
+% each MCQ a student would see a snippet of pseudo-code and be asked to
+% choose from a list of responses. The structure was explicitly modeled
+% after Force Concept Inventory (FCI) work in physics education.
+%
+% \emph{So, if her intent (and research) was about making a concept
+% \textbf{inventory}, aren't you misusing her ideas of concepts?}
+%
+% What makes you say that?
+%
+% \emph{Well, first off, she used categories of concept to create concrete
+% instances of code, which she tests students on}\ldots{}
+%
+% That's my understanding, yep.
+%
+% \emph{But you used a concrete instance of code --- Lionel's code --- and
+% inferred the existence of concepts in it. So, that's not the same thing
+% as what Elliott Tew (2010) did.}
+%
+% I agree I'm not doing what she did, but I don't think that observation
+% necessarily undermines the usefulness of what I \emph{am} doing. If the
+% concepts she identifies really are general concepts, then it would seem
+% silly to think we couldn't find concrete instances of code that
+% exemplify them. I think I might rephrase your objection: ``how can I
+% trust that you properly identified concepts in Lionel's code?'' My
+% answer to that is, again, yes, I have no coding scheme from Elliott Tew
+% to apply. But, the code Lionel wrote is there and my candidates for
+% assigning concepts are there: my results are open to inspection and
+% challenge.
+%
+% \emph{OK. Fine. But what ``conceptual knowledge analysis'' is there to
+% speak of? You just made a slide and identified which parts of his code
+% might correspond to which concepts.}
+%
+% What I did was assign concepts to canonically-correct --- that is,
+% syntactically valid parts of Lionel's code that also work properly to
+% make the program produce the right results --- sections of Lionel's
+% code. I think your objection is about my inferential basis: ``why does
+% the presence of those code chunks --- and my tagging certain chunks as
+% aligning with Elliott Tew's (2010) concepts --- imply that Lionel has
+% those concepts in his head?''
+%
+% \emph{I accept your amendment. So, what basis do you have for saying
+% ``this concept is in Lionel's mind?''}
+%
+% I feel like I'm on the same playing field as the concept inventory
+% folks, though in some ways I'm setting a modified standard for validity.
+% Elliott Tew (2010) says this about validity:
+%
+% validity is the evidence that assures us that questions about a
+% particular concept are indeed measuring that concept. For instance, a
+% question about arrays should require a student to have knowledge about
+% arrays, but should not require knowledge about another concept, such as
+% recursion. In addition, it is important that the question cannot be
+% answered correctly without knowledge of arrays. (Elliott Tew, 2010, p.
+% 10)
+%
+% I don't know that I share her assumptions about validity. But, what
+% specifically might be at issue here is ``does the presence of a for-loop
+% in code imply the student has knowledge of definite loops?'' My answer
+% is, ``if not that, then what?''
+%
+% \emph{But couldn't I argue that the for-loop could be there because he
+% copied and pasted it from the internet? That he wouldn't have had to
+% know anything about definite loops to do that?}
+%
+% Yes, you could. My rejoinder to that is in section 3.3.3.
+%
+% \subsubsection{Neverly-asked questions about Lionel's verbal pseudo-code
+% description}\label{neverly-asked-questions-about-lionels-verbal-pseudo-code-description}
+%
+% Why do you so pedantically analyze such a short, small snippet of
+% conversation?
+%
+% Because by Lionel's own assertion what was said constitutes his ``main
+% concept'' (line 12) for the program. That is, by his own account he has
+% just expressed the entire top-level design for the program. But, he's
+% done it all all without writing a single additional line of code.
+%
+% So?
+%
+% If anything, this episode suggests that Lionel can and does have ways of
+% expressing a process or procedure that don't rely on writing code. More
+% importantly, I think, is that he views it as a legitimate activity to
+% express ideas at a level \emph{above} the particularities of syntax and
+% implementation.
+%
+% What do you mean ``views it as a legitimate activity?''
+%
+% I think here I'm appealing to my interpretation of Goffman's (1974)
+% \emph{frames}: a participant's understanding of ``what is it'' that's
+% going on in a social situation. Lionel takes up my bid for him to
+% explain what he's thinking, but he does so first by explaining his
+% ``general concept'' for the program.
+%
+% Right. But, I mean, he's just doing what you \textbf{asked} him to do. I
+% don't understand how any of this relates to Goffman or frames. That just
+% seems like a needlessly complicated way to explain he did what you
+% asked.
+%
+% OK, let's step back for a second. Suppose you've just started going to
+% group therapy. One of the things group therapy environments often stress
+% is that you should try to talk about how \emph{you feel}, which can be a
+% hard transition for people. Group therapy participants may be used to
+% saying ``you always do {[}X{]}'' when in a fight with someone. And, if
+% prompted by a group leader to discuss how they feel about troublesome
+% instance, a participant may adopt that kind of language. ``My partner
+% always yells at me if I'm out late,'' for example.
+%
+% Group therapy can encourage inward reflection in a way that reshapes how
+% participants talk about themselves and their feelings. So, after some
+% time in group therapy our hypothetical participant may change the way
+% they orient toward questions. If asked ``how are you feeling about the
+% fight you and your partner had?'' our hypothetical participant may now
+% say ``I feel like I don't have the freedom to see the people I want to
+% see.'' The focus in speech shifts from the \emph{other}---in this case a
+% partner---to the \emph{self}. Moreover, the substance of the speech is
+% about feelings, rather than behavior.
+%
+% I think these reshapings of speech are neither trivial nor incidental.
+% If we had transcripts of our hypothetical person's early sessions in
+% group therapy and compared them to those of later sessions, I think we
+% would not be surprised to see a marked change. When asked the same
+% question---in this case ``how are you feeling about {[}X{]},'' the
+% structure and character of our participant's response changes after
+% therapy.
+%
+% My point here is that an everyday sense of how participants' speech
+% patterns change in therapy primes our intuition for thinking more
+% broadly about social phenomena. A cynical view of our participant would
+% still have to acknowledge that, observably, the structure of their
+% response to the question ``how does that make you feel?'' changes after
+% therapy. I think a sociolinguistic stance would be more generous: our
+% participant is \emph{orienting} differently toward the question.
+% Furthermore, I take the following as evidence of such an orientation:
+%
+% \begin{itemize}
+% \item
+%   A change in sentence focus from the other to the self
+% \item
+%   Introspection into the source of a feeling
+% \item
+%   Acknowledgment of personal responsibility for feeling those feelings
+% \end{itemize}
+%
+% I think we as analysts can acknowledge that after therapy, our
+% participant frames the event of being asked, ``how does that make you
+% feel?'' differently.
+%
+% OK. So, you're saying that if we look at the change in how someone
+% answers the question ``how does that make you feel?'' we can find
+% evidence of the superstructure---a frame---that directs their sense of
+% appropriate ways to answer that question?
+%
+% Yes.
+%
+% So how does this get back to your claims about Lionel?
+%
+% I think it's tempting to take this position that, ``it's
+% inevitable/totally expected that he would just verbalize the top-level
+% description of his procedure! I mean, what else could he possibly do?''
+%
+% I mean, yeah. Kind of.
+%
+% Right, but I think actually that assumption isn't in line with a classic
+% frame analysis approach. Even if the assumption is right, frame analysis
+% changes the explanation from ``it's inevitable'' to ``hmm, I guess
+% almost everyone I've ever seen brings or enacts the same structure of
+% expectations.''
+%
+% So, my argument is, if you ask someone to describe to you how they're
+% thinking about their program, they're gonna do that. What else
+% \textbf{would} they do?
+%
+% Right. And I don't think we disagree that they'll describe their
+% thinking. I think where we disagree is that everyone will describe in
+% the same way. I'm saying \emph{if} they describe their program to you,
+% \emph{that} they're able to express it, what they choose to verbalize,
+% what they choose to leave out, how their gestures accompany the
+% explanations, and their ultimate criteria for having satisfied the task
+% of describing for themselves are all driven underneath (or from above?)
+% by structures of expectations. And even if \emph{one thousand} out of
+% one thousand people do the same kind of thing Lionel does, that doesn't
+% undermine the idea that people have and bring structures of expectations
+% to a task. What I think it does is show that one thousand out of one
+% thousand people share the same sets of expectations about what it is
+% that's going on when they're asked to verbalize their thinking on a
+% programming protocol.
+%
+% So, what's the point of this whole section on Lionel's verbal
+% description, then?
+%
+% I think the point is that when I ask Lionel to verbalize his thinking,
+% he outlines through talk what the ``main concept'' of his program is. I
+% focus on \emph{that} he decides to describe it out in words, \emph{how}
+% he describes it, \emph{what} he leaves out\emph{, how} he repairs, and
+% how all of that reflects structures of expectations about the activity
+% and what he's being asked to do. Ultimately, the argument is that
+% Lionel's choice to take me up on that bid and how he does so
+% reflects---to me, the analyst---that he views \emph{verbal description
+% of a high-level program} as a legitimate activity and part of his
+% approach to programming.
+%
+% \section{References}\label{references}
+%
+% Abelson, H., \& Sussman, G. J. (1996). \emph{Structure and
+% Interpretation of Computer Programs} (2nd ed.). Cambridge, Mass: MIT
+% Press.
+%
+% Adelson, B., \& Soloway, E. (1985). The Role of Domain Experience in
+% Software Design. \emph{IEEE Transactions on Software Engineering},
+% \emph{SE-11}(11), 1351 -- 1360. doi:10.1109/TSE.1985.231883
+%
+% Archer, L. B. (1979). Whatever became of design methodology.
+% \emph{Design Studies}, \emph{1}(1), 17--18.
+%
+% Bain, K. (2004). \emph{What the best college teachers do}. Cambridge,
+% Mass: Harvard University Press.
+%
+% Baker, A., \& van der Hoek, A. (2010). Ideas, subjects, and cycles as
+% lenses for understanding the software design process. \emph{Design
+% Studies}, \emph{31}(6), 590--613. doi:10.1016/j.destud.2010.09.008
+%
+% Ball, L. J., Onarheim, B., \& Christensen, B. T. (2010). Design
+% requirements, epistemic uncertainty and~solution development strategies
+% in software design. \emph{Design Studies}, \emph{31}(6), 567--589.
+% doi:10.1016/j.destud.2010.09.003
+%
+% Bayman, P., \& Mayer, R. E. (1983). A diagnosis of beginning
+% programmers' misconceptions of BASIC programming statements.
+% \emph{Communications of the ACM}, \emph{26}(9), 677--679.
+% doi:http://doi.acm.org/10.1145/358172.358408
+%
+% Black, P., \& Wiliam, D. (1998). Inside the Black Box: Raising Standards
+% Through Classroom Assessment. \emph{Phi Delta Kappan}, \emph{80}(2),
+% 139--44.
+%
+% Boaler, J. (1998). Open and closed mathematics: Student experiences and
+% understandings. \emph{Journal for Research in Mathematics Education},
+% \emph{29}(1), 41--62.
+%
+% Boaler, J. (2000). Mathematics from another world: Traditional
+% communities and the alienation of learners. \emph{The Journal of
+% Mathematical Behavior}, \emph{18}(4), 379--397.
+% doi:10.1016/S0732-3123(00)00026-2
+%
+% Boaler, J. (2002). The development of disciplinary relationships:
+% Knowledge, practice and identity in mathematics classrooms. \emph{For
+% the Learning of Mathematics}, \emph{22}(1), 42--47.
+%
+% Boaler, J., \& Greeno, J. G. (2000). Identity, agency, and knowing in
+% mathematical worlds. In J. Boaler (Ed.), \emph{Multiple perspectives on
+% mathematics teaching and learning} (pp. 171--200). Westport, CT: Ablex
+% Pub.
+%
+% Bonar, J., \& Soloway, E. (1983). Uncovering principles of novice
+% programming. \emph{Proceedings of the 10th ACM SIGACT-SIGPLAN Symposium
+% on Principles of Programming Languages}, 10--13.
+% doi:10.1145/567067.567069
+%
+% Bonar, J., \& Soloway, E. (1985). Preprogramming Knowledge: A Major
+% Source of Misconceptions in Novice Programmers. \emph{Human-Computer
+% Interaction}, \emph{1}(2), 133.
+%
+% Bucciarelli, L. L. (1994). \emph{Designing engineers}. Cambridge, Mass:
+% MIT Press.
+%
+% Clancy, M. (2004). Misconceptions and Attitudes that Interfere with
+% Learning to Program. In S. Fincher \& M. Petre (Eds.), \emph{Computer
+% Science Education Research} (pp. 85--100). London, UK: RoutledgeFalmer.
+%
+% Danielak, B. A., Gupta, A., \& Elby, A. (in press). The Marginalized
+% Identities of Sense-Makers: Reframing Engineering Student Retention.
+% \emph{Journal of Engineering Education}.
+%
+% Danielsiek, H., Paul, W., \& Vahrenhold, J. (2012). Detecting and
+% Understanding Students' Misconceptions Related to Algorithms and Data
+% Structures. In \emph{Proceedings of the 43rd ACM Technical Symposium on
+% Computer Science Education} (pp. 21--26). New York, NY, USA: ACM.
+% doi:10.1145/2157136.2157148
+%
+% diSessa, A. A. (1986). Models of Computation. In D. A. Norman \& S. W.
+% Draper (Eds.), \emph{User centered system design: new perspectives on
+% human-computer interaction} (pp. 201--218). Hillsdale, N.J: L. Erlbaum
+% Associates.
+%
+% diSessa, A. A. (1993). Toward an Epistemology of Physics.
+% \emph{Cognition and Instruction}, \emph{10}(2/3), 105--225.
+%
+% diSessa, A. A. (2002). Why ``Conceptual Ecology'' is a good idea. In M.
+% Limón \& L. Mason (Eds.), \emph{Reconsidering conceptual change: issues
+% in theory and practice} (pp. 29--60). Dordrecht ; Boston: Kluwer
+% Academic Publishers.
+%
+% diSessa, A. A., \& Sherin, B. L. (1998). What changes in conceptual
+% change? \emph{International Journal of Science Education},
+% \emph{20}(10), 1155--1191. doi:10.1080/0950069980201002
+%
+% Duckworth, E. R. (2006). \emph{``The having of wonderful ideas'' and
+% other essays on teaching and learning} (3rd ed.). New York: Teachers
+% College Press.
+%
+% Elby, A., \& Hammer, D. (2010). Epistemological resources and framing: A
+% cognitive framework for helping teachers interpret and respond to their
+% students' epistemologies. In L. D. Bendixen \& F. C. Feucht (Eds.),
+% \emph{Personal epistemology in the classroom: theory, research, and
+% implications for practice} (pp. 409--434). Cambridge, UK ; New York:
+% Cambridge University Press.
+%
+% Elliott Tew, A. (2010). \emph{Assessing fundamental introductory
+% computing concept knowledge in a language independent manner} (Ph.D.).
+% Georgia Institute of Technology, Ann Arbor. Retrieved from ProQuest
+% Dissertations \& Theses Full Text. (873212789)
+%
+% Elliott Tew, A., \& Guzdial, M. (2010). Developing a validated
+% assessment of fundamental CS1 concepts. \emph{Proceedings of the 41st
+% ACM Technical Symposium on Computer Science Education}, 97--101.
+% doi:10.1145/1734263.1734297
+%
+% Elliott Tew, A., \& Guzdial, M. (2011). The FCS1: a language independent
+% assessment of CS1 knowledge. In \emph{Proceedings of the 42nd ACM
+% technical symposium on Computer science education} (pp. 111--116). New
+% York, NY, USA: ACM. doi:10.1145/1953163.1953200
+%
+% Erickson, F. (1986). Qualitative methods in research on teaching. In M.
+% C. Wittrock (Ed.), \emph{Handbook of research on teaching} (3rd ed., pp.
+% 119--161). New York : London: Macmillan ; Collier Macmillan.
+%
+% Eynde, P., \& Hannula, M. (2006). The Case Study of Frank.
+% \emph{Educational Studies in Mathematics}, \emph{63}(2), 123--129.
+% doi:10.1007/s10649-006-9030-8
+%
+% Fleury, A. E. (1991). Parameter passing: the rules the students
+% construct. In \emph{Proceedings of the twenty-second SIGCSE technical
+% symposium on Computer science education} (pp. 283--286). New York, NY,
+% USA: ACM. doi:10.1145/107004.107066
+%
+% Fleury, A. E. (1993). Student Beliefs about Pascal Programming.
+% \emph{Journal of Educational Computing Research}, \emph{9}(3), 355--371.
+% doi:10.2190/VECR-P8T6-GB10-MXJ5
+%
+% Fleury, A. E. (2000). Programming in Java: student-constructed rules.
+% \emph{SIGCSE Bull.}, \emph{32}(1), 197--201. doi:10.1145/331795.331854
+%
+% Gainsburg, J. (2006). The mathematical modeling of structural engineers.
+% \emph{Mathematical Thinking \& Learning}, \emph{8}(1), 3--36.
+% doi:10.1207/s15327833mtl0801\_2
+%
+% Gal-Ezer, J., \& Zur, E. (2004). The efficiency of
+% algorithms-\/-misconceptions. \emph{Computers \& Education},
+% \emph{42}(3), 215--226.
+%
+% Gaspar, A., \& Langevin, S. (2007). Restoring ``coding with intention''
+% in introductory programming courses. \emph{Proceedings of the 8th ACM
+% SIGITE Conference on Information Technology Education}, 91--98.
+% doi:10.1145/1324302.1324323
+%
+% Ginsburg, H. P. (1997). \emph{Entering the child's mind: the clinical
+% interview in psychological research and practice}. Cambridge ; New York:
+% Cambridge University Press.
+%
+% Ginsburg, H. P., \& Opper, S. (1988). \emph{Piaget's Theory of
+% Intellectual Development} (3rd ed.). Englewood Cliffs, N.J:
+% Prentice-Hall. Retrieved from http://lccn.loc.gov/87017353
+%
+% Goffman, E. (1974). \emph{Frame Analysis: An Essay on the Organization
+% of Experience}. New York: Harper \& Row.
+%
+% Goldin-Meadow, S. (2003). \emph{Hearing gesture: how our hands help us
+% think}. Cambridge, Mass: Belknap Press of Harvard University Press.
+%
+% Goodwin, C. (2000). Action and embodiment within situated human
+% interaction. \emph{Journal of Pragmatics}, \emph{32}(10), 1489--1522.
+% doi:10.1016/S0378-2166(99)00096-X
+%
+% Gupta, A., Hammer, D., \& Redish, E. F. (2010). The Case for Dynamic
+% Models of Learners' Ontologies in Physics. \emph{Journal of the Learning
+% Sciences}, \emph{19}(3), 285. doi:10.1080/10508406.2010.491751
+%
+% Hall, R. (1999). Following mathematical practices in design-oriented
+% work. In C. Hoyles, C. Morgan, \& G. Woodhouse (Eds.), \emph{Rethinking
+% the Mathematics Curriculum} (pp. 29--47). London: Falmer Press.
+%
+% Hall, R., \& Nemirovsky, R. (2012). Introduction to the Special Issue:
+% Modalities of Body Engagement in Mathematical Activity and Learning.
+% \emph{Journal of the Learning Sciences}, \emph{21}(2), 207--215.
+% doi:10.1080/10508406.2011.611447
+%
+% Hall, R., \& Stevens, R. (1995). Making space: A comparison of
+% mathematical work in school and professional design practices. In S. L.
+% Star (Ed.), \emph{The cultures of computing} (pp. 118--145). Oxford, UK:
+% Blackwell Publisher.
+%
+% Hall, R., Stevens, R., \& Torralba, T. (2002). Disrupting
+% representational infrastructure in conversations across disciplines.
+% \emph{Mind, Culture \& Activity}, \emph{9}(3), 179--210.
+%
+% Hall, R., Wright, K., \& Wieckert, K. (2007). Interactive and Historical
+% Processes of Distributing Statistical Concepts Through Work
+% Organization. \emph{Mind, Culture \& Activity}, \emph{14}(1/2),
+% 103--127. doi:10.1080/10749030701307770
+%
+% Hammer, D. (1989). Two approaches to learning physics. \emph{The Physics
+% Teacher}, \emph{27}(9), 664--670. doi:10.1119/1.2342910
+%
+% Hammer, D. (1994). Epistemological beliefs in introductory physics.
+% \emph{Cognition and Instruction}, \emph{12}(2), 151--183.
+% doi:10.2307/3233679
+%
+% Hammer, D. (1996). Misconceptions or P-Prims: How May Alternative
+% Perspectives of Cognitive Structure Influence Instructional Perceptions
+% and Intentions? \emph{Journal of the Learning Sciences}, \emph{5}(2),
+% 97--127. doi:10.1207/s15327809jls0502\_1
+%
+% Hammer, D., \& Elby, A. (2002). On the form of a personal epistemology.
+% In B. K. Hofer \& P. R. Pintrich (Eds.), \emph{Personal epistemology:
+% The psychology of beliefs about knowledge and knowing} (pp. 169--190).
+% Mahwah, N.J: L. Erlbaum Associates.
+%
+% Hammer, D., \& Elby, A. (2003). Tapping epistemological resources for
+% learning physics. \emph{The Journal of the Learning Sciences},
+% \emph{12}(1), 53--90. doi:10.2307/1466634
+%
+% Hammer, D., Elby, A., Scherr, R. E., \& Redish, E. F. (2005). Resources,
+% framing, and transfer. In J. P. Mestre (Ed.), \emph{Transfer of learning
+% from a modern multidisciplinary perspective}. Greenwich, CT: IAP.
+%
+% Hannula, M., Evans, J., Philippou, G., \& Zan, R. (2004). Affect in
+% Mathematics Education--Exploring Theoretical Frameworks. Research Forum.
+% \emph{International Group for the Psychology of Mathematics Education},
+% 30.
+%
+% Henderson, K. (1999). \emph{On line and on paper: visual
+% representations, visual culture, and computer graphics in design
+% engineering}. Cambridge, Mass: MIT Press.
+%
+% Herman, G. L., Kaczmarczyk, L., Loui, M. C., \& Zilles, C. (2008). Proof
+% by incomplete enumeration and other logical misconceptions.
+% \emph{Proceeding of the Fourth International Workshop on Computing
+% Education Research}, 59--70. doi:10.1145/1404520.1404527
+%
+% Hofer, B. K., \& Pintrich, P. R. (1997). The development of
+% epistemological theories: Beliefs about knowledge and knowing and their
+% relation to learning. \emph{Review of Educational Research},
+% \emph{67}(1), 88--140. doi:10.3102/00346543067001088
+%
+% Holland, S., Griffiths, R., \& Woodman, M. (1997). Avoiding object
+% misconceptions. In \emph{Proceedings of the twenty-eighth SIGCSE
+% technical symposium on Computer science education} (pp. 131--134). New
+% York, NY, USA: ACM. doi:10.1145/268084.268132
+%
+% Hurd, A. (2013, March). \emph{Assessment in an Introduction to
+% Programming Course}. Presented at the 16th Annual Course Technology
+% Conference, San Diego, CA, USA. Retrieved from
+% http://www.slideshare.net/CengageLearning/andrew-hurd-assessment-in-an-intro-to-programming-course
+%
+% Hutchins, E. (1995a). \emph{Cognition in the Wild}. Cambridge, Mass: MIT
+% Press.
+%
+% Hutchins, E. (1995b). How a cockpit remembers its speeds.
+% \emph{Cognitive Science}, \emph{19}(3), 265--288.
+% doi:10.1016/0364-0213(95)90020-9
+%
+% Izsák, A. (2004). Students' Coordination of Knowledge When Learning to
+% Model Physical Situations. \emph{Cognition \& Instruction},
+% \emph{22}(1), 81--128.
+%
+% Jackson, M. (2010). Representing structure in a software system design.
+% \emph{Design Studies}, \emph{31}(6), 545--566.
+% doi:10.1016/j.destud.2010.09.002
+%
+% Jadud, M. C. (2006). Methods and tools for exploring novice compilation
+% behaviour. In \emph{Proceedings of the 2006 international workshop on
+% Computing education research - ICER '06} (p. 73). Canterbury, United
+% Kingdom. doi:10.1145/1151588.1151600
+%
+% Jordan, B., \& Henderson, A. (1995). Interaction analysis: Foundations
+% and practice. \emph{Journal of the Learning Sciences}, \emph{4}(1), 39.
+% doi:10.1207/s15327809jls0401\_2
+%
+% Joy, M., Sinclair, J., Sun, S., Sitthiworachart, J., \& López-González,
+% J. (2009). Categorising computer science education research.
+% \emph{Education and Information Technologies}, \emph{14}(2), 105--126.
+% doi:10.1007/s10639-008-9078-4
+%
+% Kaczmarczyk, L. C., Petrick, E. R., East, J. P., \& Herman, G. L.
+% (2010). Identifying student misconceptions of programming.
+% \emph{Proceedings of the 41st ACM Technical Symposium on Computer
+% Science Education}, 107--111. doi:10.1145/1734263.1734299
+%
+% Kaiser, D. (2005). \emph{Drawing theories apart: the dispersion of
+% Feynman diagrams in postwar physics}. Chicago: University of Chicago
+% Press.
+%
+% Keller, E. F. (1983). \emph{A feeling for the organism: the life and
+% work of Barbara McClintock}. San Francisco: W.H. Freeman.
+%
+% Kolikant, Y. B.-D., \& Mussai, M. (2008). ``So my program doesn't run!''
+% Definition, origins, and practical expressions of students'
+% (mis)conceptions of correctness. \emph{Computer Science Education},
+% \emph{18}(2), 135.
+%
+% Kölling, M., \& Utting, I. (2012). Building an open, large-scale
+% research data repository of initial programming student behaviour. In
+% \emph{Proceedings of the 43rd ACM technical symposium on Computer
+% Science Education} (pp. 323--324). New York, NY, USA: ACM.
+% doi:10.1145/2157136.2157234
+%
+% Latour, B. (1987). \emph{Science in action: how to follow scientists and
+% engineers through society}. Cambridge, Mass: Harvard University Press.
+%
+% Latour, B. (1990). Drawing things together. In M. Lynch \& S. Woolgar
+% (Eds.), \emph{Representation in Scientific Practice} (1st MIT Press ed.,
+% pp. 19--68). Cambridge, Mass: MIT Press.
+%
+% Lehrer, R., Schauble, L., Carpenter, S., \& Penner, D. (2000). The
+% interrrelated development of inscriptions and conceptual understanding.
+% In P. Cobb, E. Yackel, \& K. McClain (Eds.), \emph{Symbolizing and
+% Communicating in Mathematics Classrooms: Perspectives on Discourse,
+% Tools, and Instructional Design} (pp. 325--360). Mahwah, N.J: Lawrence
+% Erlbaum Associates.
+%
+% Lising, L., \& Elby, A. (2005). The impact of epistemology on learning:
+% A case study from introductory physics. \emph{American Journal of
+% Physics}, \emph{73}(4), 372. doi:10.1119/1.1848115
+%
+% Louca, L., Elby, A., Hammer, D., \& Kagey, T. (2004). Epistemological
+% Resources: Applying a New Epistemological Framework to Science
+% Instruction. \emph{Educational Psychologist}, \emph{39}(1), 57--68.
+%
+% Malmi, L., Sheard, J., Simon, Bednarik, R., Helminen, J., Korhonen, A.,
+% \ldots{} Taherkhani, A. (2010). Characterizing research in computing
+% education: a preliminary analysis of the literature. In
+% \emph{Proceedings of the Sixth international workshop on Computing
+% education research} (pp. 3--12). New York, NY, USA: ACM.
+% doi:10.1145/1839594.1839597
+%
+% Martin, R. C. (2009). \emph{Clean code: a handbook of agile software
+% craftsmanship}. Upper Saddle River, NJ: Prentice Hall.
+%
+% Mayer, R. E. (1981). The Psychology of How Novices Learn Computer
+% Programming. \emph{ACM Computing Surveys (CSUR)}, \emph{13}, 121--141.
+% doi:10.1145/356835.356841
+%
+% Minsky, M. L. (1986). \emph{The Society of Mind}. New York: Simon and
+% Schuster.
+%
+% Nasir, N. S., \& Cooks, J. (2009). Becoming a Hurdler: How Learning
+% Settings Afford Identities. \emph{Anthropology \& Education Quarterly},
+% \emph{40}(1), 41--61. doi:10.1111/j.1548-1492.2009.01027.x
+%
+% Nasir, N. S., \& Hand, V. (2008). From the court to the classroom:
+% Opportunities for engagement, learning, and identity in basketball and
+% classroom mathematics. \emph{Journal of the Learning Sciences},
+% \emph{17}(2), 143--179. doi:10.1080/10508400801986108
+%
+% Nasir, N. S., \& Hand, V. M. (2006). Exploring Sociocultural
+% Perspectives on Race, Culture, and Learning. \emph{Review of Educational
+% Research}, \emph{76}(4), 449--475.
+%
+% Nemirovsky, R., Rasmussen, C., Sweeney, G., \& Wawro, M. (2012). When
+% the Classroom Floor Becomes the Complex Plane: Addition and
+% Multiplication as Ways of Bodily Navigation. \emph{Journal of the
+% Learning Sciences}, \emph{21}(2), 287--323.
+% doi:10.1080/10508406.2011.611445
+%
+% Ochs, E., Gonzales, P., \& Jacoby, S. (1996). ``When I come down I'm in
+% the domain state'': grammar and graphic representation in the
+% interpretive activity of physicists. In \emph{Interaction and Grammar}
+% (pp. 328--369). Cambridge: Cambridge University Press.
+%
+% Papert, S. (1980). \emph{Mindstorms: Children, Computers, and Powerful
+% Ideas}. New York: Basic Books. Retrieved from
+% http://lccn.loc.gov/79005200
+%
+% Parsons, J., \& Saunders, C. (2004). Cognitive heuristics in software
+% engineering: Applying and extending anchoring and adjustment to artifact
+% reuse. \emph{IEEE Transactions on Software Engineering}, \emph{30}(12),
+% 873 -- 888. doi:10.1109/TSE.2004.94
+%
+% Patitsas, E., Craig, M., \& Easterbrook, S. (2013). On the Countably
+% Many Misconceptions About \#Hashtables (Abstract Only). In
+% \emph{Proceeding of the 44th ACM Technical Symposium on Computer Science
+% Education} (pp. 739--739). New York, NY, USA: ACM.
+% doi:10.1145/2445196.2445443
+%
+% Paul, W., \& Vahrenhold, J. (2013). Hunting High and Low: Instruments to
+% Detect Misconceptions Related to Algorithms and Data Structures. In
+% \emph{Proceeding of the 44th ACM Technical Symposium on Computer Science
+% Education} (pp. 29--34). New York, NY, USA: ACM.
+% doi:10.1145/2445196.2445212
+%
+% Pea, R. D. (1986). Language-independent conceptual`` bugs'' in novice
+% programming. \emph{Journal of Educational Computing Research},
+% \emph{2}(1), 25--36.
+%
+% Pea, R. D., Soloway, E., \& Spohrer, J. C. (1987). The Buggy Path to the
+% Development of Programming Expertise. \emph{Focus on Learning Problems
+% in Mathematics}, \emph{9}(1), 5--30.
+%
+% Petre, M., van der Hoek, A., \& Baker, A. (2010). Editorial.
+% \emph{Design Studies}, \emph{31}(6), 533--544.
+% doi:10.1016/j.destud.2010.09.001
+%
+% Rodrigo, M. M. T., \& Baker, R. S. J. d. (2009). Coarse-grained
+% detection of student frustration in an introductory programming course.
+% In \emph{Proceedings of the fifth international workshop on Computing
+% education research workshop} (pp. 75--80). New York, NY, USA: ACM.
+% doi:10.1145/1584322.1584332
+%
+% Rodrigo, M. M. T., Tabanao, E., Lahoz, M. B. ., \& Jadud, M. C. (2009).
+% Analyzing Online Protocols to Characterize Novice Java Programmers.
+% \emph{Philippine Journal of Science}, \emph{138}(2), 177--190.
+%
+% Rooksby, J. (2010). ``Just try to do it at the whiteboard'':
+% Researcher-participant interaction and issues of generalisation. In
+% \emph{Proceedings of the Studying Professional Software Design (SPSD)
+% Conference}. San Diego, CA, USA.
+%
+% Rooksby, J., \& Ikeya, N. (2012). Collaboration in Formative Design:
+% Working Together at a Whiteboard. \emph{IEEE Software}, \emph{29}(1), 56
+% --60. doi:10.1109/MS.2011.123
+%
+% Rosenberg, S., Hammer, D., \& Phelan, J. (2006). Multiple
+% Epistemological Coherences in an Eighth-Grade Discussion of the Rock
+% Cycle. \emph{Journal of the Learning Sciences}, \emph{15}(2), 261--292.
+% doi:10.1207/s15327809jls1502\_4
+%
+% Russ, R. S., Coffey, J. E., Hammer, D., \& Hutchison, P. (2008). Making
+% classroom assessment more accountable to scientific reasoning: A case
+% for attending to mechanistic thinking. \emph{Science Education},
+% \emph{93}(5), 875--891. doi:10.1002/sce.20320
+%
+% Saussure, F. de. (1986). \emph{Course in general linguistics}. LaSalle,
+% Ill: Open Court.
+%
+% Scherr, R. E., \& Hammer, D. (2009). Student Behavior and
+% Epistemological Framing: Examples from Collaborative Active-Learning
+% Activities in Physics. \emph{Cognition \& Instruction}, \emph{27}(2),
+% 147--174. doi:10.1080/07370000902797379
+%
+% Schoenfeld, A. H. (1987). What's all the fuss about metacognition? In
+% \emph{Cognitive Science and Mathematics Education} (pp. 189--215).
+% Lawrence Erlbaum Associates.
+%
+% Schoenfeld, A. H. (1988). When good teaching leads to bad results: The
+% disasters of ``well-taught'' mathematics courses. \emph{Educational
+% Psychologist}, \emph{23}(2), 145--166. doi:10.1207/s15326985ep2302\_5
+%
+% Schoenfeld, A. H. (1991). On mathematics as sense-making: An informal
+% attack on the unfortunate divorce of formal and informal mathematics. In
+% \emph{Informal reasoning and education} (pp. 311--343).
+%
+% Schoenfeld, A. H. (1992). Learning to Think Mathematically: Problem
+% Solving, Metacognition, and Sense-Making in Mathematics. In D. Grouws
+% (Ed.), \emph{Handbook for research on mathematics teaching and learning}
+% (pp. 334--370). New York: MacMillan.
+%
+% Schwartz, D. L., Chase, C. C., \& Bransford, J. D. (2012). Resisting
+% Overzealous Transfer: Coordinating Previously Successful Routines With
+% Needs for New Learning. \emph{Educational Psychologist}, \emph{47}(3),
+% 204--214. doi:10.1080/00461520.2012.696317
+%
+% Seppälä, O., Malmi, L., \& Korhonen, A. (2006). Observations on student
+% misconceptions---A case study of the Build -- Heap Algorithm.
+% \emph{Computer Science Education}, \emph{16}(3), 241--255.
+% doi:10.1080/08993400600913523
+%
+% Sherin, B. L. (2001). How students understand physics equations.
+% \emph{Cognition and Instruction}, \emph{19}(4), 479--541.
+%
+% Smith, J. P., diSessa, A. A., \& Roschelle, J. (1993). Misconceptions
+% Reconceived: A Constructivist Analysis of Knowledge in Transition.
+% \emph{The Journal of the Learning Sciences}, \emph{3}(2), 115--163.
+%
+% Soloway, E. (1986). Learning to program = learning to construct
+% mechanisms and explanations. \emph{Communications of the ACM},
+% \emph{29}, 850--858. doi:10.1145/6592.6594
+%
+% Soloway, E., Bonar, J., \& Ehrlich, K. (1983). Cognitive Strategies and
+% Looping Constructs: An Empirical Study. \emph{Commun. ACM},
+% \emph{26}(11), 853--860. doi:10.1145/182.358436
+%
+% Soloway, E., \& Spohrer, J. C. (Eds.). (1989). \emph{Studying the Novice
+% Programmer}. Hillsdale, N.J: L. Erlbaum Associates.
+%
+% Spacco, J., Hovemeyer, D., Pugh, W., Hollingsworth, J., Padua-Perez, N.,
+% \& Emad, F. (2006). Experiences with Marmoset: Designing and Using an
+% Advanced Submission and Testing System for Programming Courses. In
+% \emph{ITiCSE '06: Proceedings of the 11th annual conference on
+% Innovation and technology in computer science education}. ACM Press.
+%
+% Spacco, J., Pugh, W., Ayewah, N., \& Hovemeyer, D. (2006). The Marmoset
+% project: an automated snapshot, submission, and testing system. In
+% \emph{Companion to the 21st ACM SIGPLAN symposium on Object-oriented
+% programming systems, languages, and applications} (pp. 669--670). New
+% York, NY, USA: ACM. doi:10.1145/1176617.1176665
+%
+% Spacco, J., Strecker, J., Hovemeyer, D., \& Pugh, W. (2005). Software
+% Repository Mining with Marmoset: An Automated Programming Project
+% Snapshot and Testing System. In \emph{Proceedings of the Mining Software
+% Repositories Workshop (MSR 2005)}. St. Louis, Missouri, USA.
+%
+% Spohrer, J. C., \& Soloway, E. (1986). Alternatives to construct-based
+% programming misconceptions. \emph{ACM SIGCHI Bulletin}, \emph{17},
+% 183--191. doi:10.1145/22339.22369
+%
+% Stevens, R., \& Hall, R. (1998). Disciplined perception: Learning to see
+% in technoscience. In \emph{Talking Mathematics in School: Studies of
+% Teaching and Learning} (pp. 107--150). Cambridge, U.K: Cambridge
+% University Press.
+%
+% Stieff, M. (2007). Mental rotation and diagrammatic reasoning in
+% science. \emph{Learning and Instruction}, \emph{17}(2), 219--234.
+% doi:10.1016/j.learninstruc.2007.01.012
+%
+% Tabanao, E. S., Rodrigo, M. M. T., \& Jadud, M. C. (2011). Predicting
+% at-risk novice Java programmers through the analysis of online
+% protocols. In \emph{Proceedings of the seventh international workshop on
+% Computing education research} (pp. 85--92). New York, NY, USA: ACM.
+% doi:10.1145/2016911.2016930
+%
+% Tannen, D. (Ed.). (1993). \emph{Framing in Discourse}. New York: Oxford
+% University Press.
+%
+% Trakhtenbrot, M. (2013). Students Misconceptions in Analysis of
+% Algorithmic and Computational Complexity of Problems. In
+% \emph{Proceedings of the 18th ACM Conference on Innovation and
+% Technology in Computer Science Education} (pp. 353--354). New York, NY,
+% USA: ACM. doi:10.1145/2462476.2465604
+%
+% Traweek, S. (1988). \emph{Beamtimes and lifetimes: the world of high
+% energy physicists}. Cambridge, Mass: Harvard University Press.
+%
+% Turkle, S., \& Papert, S. (1991). Epistemological Pluralism and the
+% Reevaluation of the Concrete. In I. Harel, S. Papert, \& Massachusetts
+% Institute of Technology (Eds.), \emph{Constructionism: research reports
+% and essays, 1985-1990}. Norwood, N.J: Ablex Pub. Corp.
+%
+% Valentine, D. W. (2004). CS educational research: a meta-analysis of
+% SIGCSE technical symposium proceedings. \emph{SIGCSE Bull.},
+% \emph{36}(1), 255--259. doi:10.1145/1028174.971391
+%
+% Van de Sande, C. C., \& Greeno, J. G. (2012). Achieving Alignment of
+% Perspectival Framings in Problem-Solving Discourse. \emph{Journal of the
+% Learning Sciences}, \emph{21}(1), 1--44.
+% doi:10.1080/10508406.2011.639000
+%
+% VanLehn, K. (1990). \emph{Mind Bugs: The Origins of Procedural
+% Misconceptions}. Cambridge, Mass: MIT Press.
+%
+% Wagner, J. F. (2006). Transfer in Pieces. \emph{Cognition \&
+% Instruction}, \emph{24}(1), 1--71. doi:10.1207/s1532690xci2401\_1
+%
+% Wortham, S. (2006). \emph{Learning Identity: The Joint Emergence of
+% Social Identification and Academic Learning}. Cambridge: Cambridge
+% University Press.
+%
+% Yackel, E., \& Cobb, P. (1996). Sociomathematical Norms, Argumentation,
+% and Autonomy in Mathematics. \emph{Journal for Research in Mathematics
+% Education}, \emph{27}(4), 458--477. doi:10.2307/749877
+%
+% Yin, R. K. (2009). \emph{Case study research: design and methods} (4th
+% ed.). Los Angeles, Calif: Sage Publications.

### 2016-04-22 manuscript-body.tex
Copy parts of exported dissertation into manuscript body
But, I forgot to apply `--no-wrap` to the pandoc export, so now I have
to go fix that.

@@
-\section{Introduction}\label{introduction}
+\subsection{Introduction}\label{introduction-2}
+
+Although imaginative and carefully designed, the Harvard course taught
+that there is only one right way to approach the computer, a way that
+emphasizes control through structure and planning. There are many
+virtues to this computational approach (it certainly makes sense when
+dividing the labor on a large programming project), but Lisa and Robin
+have intellectual styles at war with it. Lisa says she has ``turned
+herself into a different kind of person'' in order to perform, and Robin
+says she has learned to ``fake it.'' Although both women are able to get
+good grades in their programming course, they represent casualties of
+this war. Both deny who they are in order to succeed. (Turkle \& Papert,
+1991, p. 165)
+
+Over 20 years ago Sherry Turkle and Seymour Papert called for
+epistemological pluralism in computing education. Their argument --- in
+my view ---had multiple parts:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Programming a computer involved a shift in how one thought about the
+  nature knowledge from a propositional \emph{what is} to an imperative
+  \emph{how to} (Abelson \& Sussman, 1996; Papert, 1980).
+\item
+  Historically, there was a tendency to treat computers and the act of
+  programming them as an extension of the same Western logico-deductive
+  knowledge tradition that fueled science.
+\item
+  Recent scholarship, often at the intersection of post-structuralism
+  and social studies of science, had been troubling the idea that
+  knowledge-work in science proceeded solely by way of logical
+  deductions (Keller, 1983; Latour, 1987; Traweek, 1988)
+\item
+  Data from professionals suggested programming, too, didn't proceed
+  solely by way of the structured-planning approach emergent from
+  Western logico-deductive knowledge traditions. There were many ways of
+  knowing and constructing knowledge in programming.
+\item
+  A viewpoint of \emph{epistemological pluralism} --- embracing multiple
+  kinds and ways of knowing --- was informing the field of science
+  studies but not, to a large degree, how education researchers thought
+  about or taught computing.
+\item
+  The lack of support for diverse ways of knowing in computing
+  classrooms demonstrably hurt students.
+\end{enumerate}
+
+To Turkle and Papert (1991), the result of this culturally-rooted
+epistemological tension was a war. More specifically, it was a war where
+students of computing were the casualties, and the aggressor was a
+distributed failure to recognize, embrace, and support diverse ways of
+knowing. While I think their strong-form characterization of a ``war''
+no longer applies, their consideration of epistemological issues is
+still relevant two decades later. An understanding that computing does
+in fact involve diverse ways of knowing should continue to inform
+research and discussions on learning
+
+This paper tries to model what directs and sustains students'
+in-the-moment activity when they program. Its focus is on early-stage
+program design; analyzing how what students say, do, write, and gesture
+\emph{even before they type code} can help us improve theories of
+cognition and activity in learning programming. Using a ``revelatory
+case study'' (Yin, 2009) of two students in an introductory programming
+course, I argue the following:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Students' early-stage design activity reveals patterns outside the
+  explanatory scope of misconception-based accounts of cognition.
+\item
+  For a subset of phenomena, we can recast students' productive
+  capacities and their difficulties in terms of epistemological stances.
+\item
+  Such recastings are analytically powerful. Beyond Rebecca and Lionel,
+  they could potentially explain the diversity of practices I saw other
+  students take up.
+\item
+  As evidenced by work in science and math education, dynamic
+  epistemological models can offer a lens for reforming assessment and
+  instruction.
+\end{enumerate}
+
+\subsection{Literature Review}\label{literature-review}
+
+I begin by exploring the historical legacy of Turkle and Papert's (1991)
+remarks on epistemological pluralism in computing. I'll then describe
+what I see as a well-intentioned obstacle: the research-based
+preoccupation with students' misconceptions in computing education.
+These two stra.
+
+\subsubsection{Computing culture should support a diversity of ways of
+knowing}\label{computing-culture-should-support-a-diversity-of-ways-of-knowing}
+
+Turkle \& Papert (1991) describe a tension between individuals and a
+kind of cultural collective. In this tension, individual students and
+their personally-identified ways of knowing were in conflict with the
+larger culture of computing and its manifested ways of knowing. Lisa,
+for instance, was a Harvard student who identified as a poet:
+
+Lisa experiences language as transparent, she knows where all the
+elements are at every point in the development of her ideas. She wants
+her relationship to computer language to be similarly transparent. When
+she builds large programs she prefers to write her own smaller
+``building block'' procedures even though she could use prepackaged ones
+from a program library; she resents the latter's opacity. (Turkle \&
+Papert, 1991, p. 164)
+
+Turkle and Papert connect Lisa's programming practices to Lisa's
+personal view of knowledge. She writes building block procedures because
+of a continuity between herself as knower-of-her-own-poetry and that of
+knower-of-her-own-programs. Across both contexts, Lisa \emph{resents
+opacity} because it is an obstacle to her knowing. But, that resentment
+slows her progress on projects in the class.
+
+In part because of her commitments to transparency and the difficulty
+they created, Lisa changed her approach to programming. She ultimately
+had to, in her words, ``be a different kind of person with the machine''
+(Turkle \& Papert, 1991, p. 164). Turkle and Papert explain:
+
+She had been told that the ``right way'' to do things was to control a
+program through planning and black­boxing, the technique that lets you
+exploit opacity to plan something large without knowing in advance how
+the details will be managed. Lisa recognized the value of these
+techniques --- for someone else. She struggled against using them as the
+starting points for her learning. Lisa ended up abandoning the fight,
+doing things ``their way,'' and accepting the inevitable alienation from
+her work. (Turkle \& Papert, 1991, p. 164)
+
+The phrase ``inevitable alienation'' is particularly damning. One might
+accuse Turkle and Papert (1991) of hyperbole, but research from other
+disciplines suggests they're not overstating the case. Identity
+alienation for epistemological reasons --- or some variation of them ---
+is playing out in other subjects as well.
+
+Connections and tensions between identities, learning, and personal ways
+of knowing are not unique to computing education. Wortham (2006), for
+example, detailed instances of social identification and academic
+identity as \emph{jointly emerging} in a combined high school English
+and History class. Nasir and colleagues explored how different learning
+environments afford access to identity, domain practices, and
+self-expression (Nasir \& Cooks, 2009; Nasir \& Hand, 2008). To single
+out one study, Nasir \& Hand (2008) followed players ``from the
+{[}basketball{]} court to the classroom.'' They found the court afforded
+players a sense of role on the team, chances to be expressive through
+play, and ``access to the domain {[}of basketball{]} as a whole'' (Nasir
+\& Hand, 2006, p. 147). Comparatively, the mathematics classroom offered
+little (if any) sense that students were on a team, a much narrower role
+for them as having either right or wrong answers, and no strong sense of
+self-expression or creativity through activity. These markedly different
+contexts led not only to differences in kinds of activity but, Nasir and
+Hand (2006) argue, divergent \emph{practice-linked identities} for the
+players. In a sense, players had a different identity on the court than
+they did in the mathematics classroom because who they were was strongly
+coupled to a sense of their relationship with a practice.
+
+Boaler's work on learning and identity in mathematics classrooms (Boaler
+\& Greeno, 2000; Boaler, 1998, 2000, 2002) offers resonant findings. In
+particular, Boaler (2000) strongly echoes the student voices in Turkle
+and Papert (1991). Where Turkle and Papert's aggressor is the dominant
+computing culture, Boaler describes a beast with similar effects in the
+form of ``school mathematics'' --- the \emph{monotonous},
+\emph{meaningless} (Boaler, 2002, pp. 383--384) imposition of routinized
+mathematical procedures in ways that seem far removed from the real
+world. Of school mathematics, Boaler writes:
+
+School mathematics, for many of them, was \emph{of another world} and to
+fully engage in that world, students needed to suspend their knowledge
+of the real world, suppress their desire to interact with others, and
+strive to reproduce standard procedures that held little meaning for
+them. (Boaler, 2002, p. 392 emphasis in original)
+
+Schoenfeld (1988, 1991) had already argued school mathematics could and
+did depart substantially from the core of mathematical thinking. Boaler
+established that such school mathematics could alienate learners
+(Boaler, 1998, 2000) and influence their feelings about pursuing
+mathematics after secondary school (Boaler \& Greeno, 2000).
+
+Most recently, Danielak, Gupta, and Elby (in press) extended those
+findings in engineering education. In their 3 ½ year case study of an
+electrical engineering undergraduate named ``Michael,'' those authors
+argued Michael's practice of sense-making in engineering coursework was
+strongly coupled to his identity. Michael's passion for deeply
+understanding concepts, much like Lisa's (Turkle \& Papert, 1991) desire
+to understand every element of her program, did not mesh neatly with the
+larger culture of his courses. But, because Michael's practice of
+sense-making entwined with his identity, forces that pushed against
+sense-making alienated him. That tension was particularly pointed when
+Michael described his father's resistance to Michael's way of thinking:
+
+{[}My dad said{]} ``you're just an undergraduate. Nobody expects
+undergraduates to understand how anything works. That's why you go to
+graduate school.'' I was like ``look, you know. I'm gonna be an unhappy
+person if I have to....I have life goals other than to just get a good
+grade on the exam. Other things are important to me.'' (Danielak et al.,
+in press)
+
+And, much like Turkle and Papert's (1991) Lisa, Michael learned to
+curtail parts of his identity to get by in class:
+
+I think the reason {[}pursuing deep learning{]} actually hasn't affected
+my GPA is because I view learning as a hobby. So, as with any hobby, you
+shouldn't let it interfere your GPA. But it \emph{is} one of my hobbies,
+and I \emph{do} enjoy learning, I just---up to the point where I get my
+grades done \{raises eyebrow\}. (Danielak et al., in press)
+
+Ultimately, research suggests the identity-epistemology tensions Turkle
+and Papert (1991) described in computing resonates with those found in
+related disciplines. And, what such accounts --- particularly those of
+Boaler and others (Boaler \& Greeno, 2000; Boaler, 1998, 2000, 2002;
+Danielak et al., in press) --- have in common is a culturally-sanctioned
+``right way'' (Turkle \& Papert, 1991, p. 164) of thinking or knowing
+that alienates some students. Avowedly such ways of knowing have use and
+value. Turkle \& Papert themselves acknowledge ``there are many
+virtues'' to black-boxing and top-down design in programming (Turkle \&
+Papert, 1991, p. 165). But,
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  When a preponderance of results from science studies suggests a
+  plurality of ways of knowing among practicing scientists and
+  programmers, and
+\item
+  When research from education shows that the inflexible imposition of
+  one way of knowing above all else is alienating students, then
+\item
+  It's worth reconsidering whether a narrow sense of what counts as
+  knowing might still be hindering improvements in computing education.
+\end{enumerate}
+
+In the next section, I turn to misconceptions research in computing
+education. In keeping with observations 1--3,
+
+\subsubsection{Misconceptions research in computing education tends to
+ignore students' productive
+knowledge}\label{misconceptions-research-in-computing-education-tends-to-ignore-students-productive-knowledge}
+
+I argue research
+approaches that inflexibly privilege canonical knowledge do so at the
+expense of other productive knowledge and ways of knowing that students
+have.
+
+In the past three decades, educational research has had a marked focus
+on students' misconceptions in programming. It's a focus with a sensible
+origin. Students get things wrong in programming --- often
+systematically so --- and in ways that seem resistant to instruction.
+The cause of those errors is theorized to be something cognitive,
+whether it's a ``bug'' (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
+``misconception'' (Bayman \& Mayer, 1983; Bonar \& Soloway, 1985;
+Clancy, 2004; Gal-Ezer \& Zur, 2004; Herman et al., 2008; Kaczmarczyk et
+al., 2010), a ``belief'' (Fleury, 1993) or a ``student-constructed
+rule'' (Fleury, 1991, 2000). Instruction should try to identify,
+address, and correct these misconceptions (Clancy, 2004) because they
+can represent barriers to learning.
+
+How that line of thinking and research becomes problematic is two-fold.
+First, when taken in total the alleged brokenness of student knowledge
+begins eclipsing all else in studying the cognition of learning to
+program. In other words, most cognitively-focused educational research
+in computer science treats students as having varied degrees of
+deficiency with respect to canonical knowledge. Below is an unordered,
+partial sampling of topics about which researchers have documented
+students' misconceptions. Note across the list the variation in both the
+grain sizes of students' misconceptions and the programming languages in
+which they manifest:
+
+\begin{itemize}
+\item
+  Objects in object-oriented programming (Holland, Griffiths, \&
+  Woodman, 1997)
+\item
+  Algorithms and data structures (Danielsiek, Paul, \& Vahrenhold, 2012;
+  Paul \& Vahrenhold, 2013)
+\item
+  Programming statements in BASIC (Bayman \& Mayer, 1983)
+\item
+  Programming in Java (Fleury, 2000)
+\item
+  Programming in Pascal (Fleury, 1993)
+\item
+  Parameter-passing (Fleury, 1991)
+\item
+  Arrays in Java (Kaczmarczyk et al., 2010)
+\item
+  Objects in Java (Kaczmarczyk et al., 2010)
+\item
+  Algorithms and computational complexity (Trakhtenbrot, 2013)
+\item
+  Boolean logic (Herman et al., 2008)
+\item
+  The efficiency of algorithms (Gal-Ezer \& Zur, 2004)
+\item
+  The Build-Heap algorithm (Seppälä, Malmi, \& Korhonen, 2006)
+\item
+  Hashtables (Patitsas, Craig, \& Easterbrook, 2013)
+\item
+  The correctness of programs (Kolikant \& Mussai, 2008)
+\end{itemize}
+
+Clancy (2004) provides a comprehensive overview of this line of
+research, though in the past decade it has only grown. Indeed, roughly
+half the articles above were published in the ten years since Clancy's
+overview.
+
+Identifying and removing barriers to student learning seems like a good
+thing. So, it should follow that cataloging student misconceptions and
+developing remedies for them should also be a good thing. But, the
+logical implication isn't that clean. In some cases students display
+productive, useful knowledge that's either ignored or outright
+criticized by researchers. Aligning students toward canonical knowledge
+makes sense, but doing so at the expense of---or in direct contradiction
+to---useful ways of knowing seems undesirable at best. Next, I expand on
+two examples from misconceptions research in programming. Specifically,
+I show how and why I think a misconceptions focus in programming casts
+aside students' useful intuitions and understandings.
+
+The first example comes from Kaczmarczyk et al. (2010). Part of that
+study involved giving students snippets of Java code and asking students
+to diagram (or pseudo-code) how the information would be stored in
+memory. Below I have reproduced the code for Problem 2 (Kaczmarczyk et
+al., 2010, p. 110):
+
+Cheese{[}{]} cheeses = new Cheese{[}4{]};
+
+Meat{[}{]} meats = new Meat{[}2{]};
+
+Turkey turkey;
+
+Ham ham;
+
+RoastBeef roastBeef;
+
+boolean lettuce = true;
+
+boolean tomato = true;
+
+SauceType sauceType = new SauceType();
+
+int numMeat;
+
+int numCheese;
+
+In diagramming this information, a student in the study makes a mistake:
+
+Student3 makes incorrect assumptions about connections between variables
+to the extent that the student makes a mistake concerning the types of
+the variables. As a result, the student places Objects of different
+types in an array whose type matches none of them: ``And so because
+there's two arrays, cheese and meats, uh, all those turkey and ham and
+roast beef are gonna be sorted into the meats array.'' (Kaczmarczyk et
+al., 2010, p. 110)
+
+The researchers are correct in the sense that \textbf{turkey} and
+\textbf{ham} and \textbf{roastBeef} will \emph{not} be sorted into the
+\textbf{meats} array. First, there is no code here that places
+\textbf{turkey} and \textbf{ham} in the array; there is only code that
+declares them as variables. Moreover, as written, an attempt to place
+\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} into the array
+would fail. Because of type restrictions in Java, only objects of class
+Meat (or objects that inherit from class Meat) can go in the array.
+\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} are, perhaps
+confusingly, references to object instances of classes \textbf{Turkey}
+and \textbf{Ham} and \textbf{RoastBeef}, so in the current snippet they
+cannot enter an array of type \textbf{Meat} because (1) they don't yet
+exist as objects and (2) even if they did exist, their types don't match
+the array's type. The authors call this misconception \emph{semantics to
+semantics}, which occurs ``when the student inappropriately
+assume{[}s{]} details about the relationship and operation of code
+samples, although such information was neither given nor implied''
+(Kaczmarczyk et al., 2010, p. 110).
+
+Again, the researchers are right that the student is failing to describe
+the code in a way consistent with canon.\footnote{In this case, canon is
+  the specifications and operations of the Java language and its
+  compilers.} But, in their non-canonical thinking Student3 evidences
+potentially productive insights about design. Precisely \emph{because}
+there is no code stating that \textbf{turkey} and \textbf{roastBeef} and
+ham are sorted into the array, the student is \emph{inferring} that to
+be true. And, while that behavior is not what's happening, it
+\emph{would} be sensible to design a program where specific instances of
+classes \textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} could go
+into an array of type \textbf{Meat}. To do so, a designer could define
+\textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} as inheriting
+from \textbf{Meat}.
+
+Student3 has an idea about a relationship between entities where that
+relationship is not specified in the code. Kaczmarczyk et al.
+(Kaczmarczyk et al., 2010) focus only on the downside of the idea: the
+student fails to display a proper understanding of how arrays work in
+Java. Moreover, the student might be prone to similar mistakes of
+inferring information that does not actually exist in code. But, there
+is also an upside of this idea. Because Student3 is thinking about
+real-world propositions like turkey and ham being kinds of meat, they
+might be prepared to appreciate and discuss an object-oriented way to
+put turkey in a \textbf{Meat} array. But, that possibility is
+speculative conjecture. We can't know for certain whether Student3 could
+be tipped into a productive object-oriented design activity around the
+meats example because that question was not a focus of the research.
+
+My second example of research that criticizes students' non-canonical
+understandings comes from Bonar and Soloway's (1983) study of Pascal
+programmers, part of which is discussed in Pea (1986). A student in
+Bonar \& Soloway's study was asked to ``Write a program which reads in
+ten integers and prints the average of those integers'' (Bonar \&
+Soloway, 1983, p. 12). In pseudo-code, she wrote:
+
+Repeat
+
+(1) Read a number (Num)
+
+(1a) Count := Count + 1
+
+(2) Add the number to Sum
+
+(2a) Sum := Sum + Num
+
+(3) until Count :=10
+
+(4) Average := Sum div Num
+
+(5) writeln (`average = `,Average)
+
+The interviewer then asked whether (1a) and (2a) were ``the same kinds
+of statements.'' That interchange is reproduced here:
+
+Interviewer: Steps 1a and 2a: are those the same kinds of statements?
+
+Subject: How's that, are they the same \emph{kind}. Ahhh, ummm, not
+exactly, because with this {[}1a{]} you are adding - you initialize it
+at zero and you're adding one to it {[}points to the right side of 1a{]}
+which is just a constant kind of thing.
+
+Interviewer: Yes
+
+Subject: {[}points to 2a{]} Sum, initialized, to, uhh Sum to Sum plus
+Num, ahh - thats {[}points to left side of 2a{]} storing two values in
+one, two variables {[}points to Sum and Num on the right side of 2a{]}.
+That's {[}now points to 1a{]} a counter, that's what keeps the whole
+loop under control. Whereas, this thing {[}points to 2a{]} was probably
+the most interesting thing\ldots{}about Pascal when I hit it. That you
+could have the same, you sorta have the same thing here {[}points to
+1a{]}, it was interesting that you cold have, you could save space by
+having the Sum re-storing information on the left with two different
+things there {[}points to right side of 2a{]}, so I didn't need to have
+two. No, they're different to me.
+
+Interviewer: So - in summary, how do you think of 1a?
+
+Subject: I think of this {[}point to 1a{]} as just a constant, something
+that keeps the loop under control. And this {[}points to 2a{]} has
+something to do with something that you are gonna, that stores more
+kinds of information that you are going to take out of the loop with
+you. (Bonar \& Soloway, 1983, p. 12)
+
+Pea's (1986) interpretation? ``Here, again, we see the student believing
+that the programming language knows more about her intentions than it
+possibly can'' (p. 32).
+
+As before, this student has an idea about relationships in code. Pea
+(1986) see the downside of her idea: believing PASCAL can understand
+shades of programmer intent when, in fact, it cannot. And again, that
+downside is real. It could cause trouble for this programmer later on if
+she expects PASCAL to interpret her intent and it cannot.
+
+In defense of the student, the question --- as asked --- is vague. Are
+those statements the same \emph{to whom and in what way?} Pea (1986)
+treats the data as though she meant ``the same to PASCAL.'' Indeed,
+maybe she did, in which case his interpretation has traction. But,
+another interpretation is that she meant to herself, or to someone else
+reading the code. Those statements might not be the same \emph{to her}
+because she treats (1a) as having a function of controlling iteration
+while (2a)'s job is to combine two numbers into a new sum.
+
+These two purposes, which for the sake of description I'll call
+\emph{keeping control} and \emph{totaling up} are, in a sense,
+different. The PASCAL compiler (and runtime) does not differentiate
+them, but humans can. And, humans may well \emph{want} to differentiate
+them. diSessa (1986) describes exactly this kind of differentiation as a
+consequence of separating the structural understanding of a programming
+language from a functional understanding of a language. As an example,
+he discusses the structure/function difference with respect to
+variables:
+
+The structural aspects of a variable in a computer language are given
+primarily by the rules for setting their values and for getting access
+to their values. These rules apply in all contexts. In contrast, a
+variable's functions might vary. Sometimes they might be described as
+``a flag'' or more generally, as ``a communications device.'' At other
+times a variable might function as ``a counter,'' ``data,'' or
+``input.'' (diSessa, 1986, p. 202)
+
+The student in Bonar \& Soloway's (1983) study did not show evidence of
+understanding the structural similarities between (1a) and (2a) in her
+pseudo-code. And, those authors as well as Pea (1986) justly insist that
+similarity is important for students to understand. From a conceptual
+standpoint, seeing the structural similarity constitutes a part of
+``knowing'' PASCAL. But, even if knowing PASCAL were not the goal,
+seeing the similarity helps one to take the perspective of a computing
+agent that has no means for discerning programmer intent. Such
+perspective-taking may help students avoid mistakes that arise from
+over-assuming what a computer ``understands.''
+
+The student did show evidence of understanding a functional difference
+between (1a) and (2a), but Pea (1986) does not remark on that kind of
+understanding at all.\footnote{Also glossed over is, to me, another
+  important difference: a programmer might not know in advance which
+  numbers are being passed in to the sum statement. So, in advance the
+  programmer can say nothing about how the value of \textbf{Sum} will
+  change as the loop iterates. In contrast, the programmer knows exactly
+  how the value of Count will change with each loop iteration.} Again, I
+claim this oversight is part of a subtle but observable trend in
+programming misconceptions literature. While, or perhaps because
+research has been so preoccupied warring with students' problematic
+knowledge, it has sometimes failed to recover the productive knowledge
+(or resources for building it) students have. In this example, the
+student already has a grasp that syntactically similar statements could
+serve different conceptual purposes. Hypothetically, the student might
+use that information in design by calling up the ``◻=︎ + 1'' syntactical
+template when the situation seems to demand \emph{keeping control},
+while calling up ``◻= ︎ + number'' when \emph{totaling up} is the goal.
+And, the idea that structurally identical symbol templates can serve
+different functional and conceptual purposes fits precisely in line with
+Sherin's (2001) theory of symbolic forms. For example, the symbolic
+forms \emph{parts-of-a-whole} and \emph{base+change} have different
+conceptual schemata. Parts-of-a-whole refers to the contributions of
+component entities while base±change describes a kind of accumulation.
+Specifically, the terms in base±change ``play different roles'' (Sherin,
+2001, p. 534) But, the two distinct conceptual schemata share what I
+would argue is the same concrete symbol template of how to write an
+expression: ◻=︎ + ◻.
+
+The problem, for learning to program, comes in needing to fluidly
+interpret and write code in languages that may demand incommensurable,
+or at least distinct, conceptual schemata. As I show later in Table 4,
+three current programming languages make remarkably different use of the
+plus sign (+) as an operator. Crucially, some of the entailing ways to
+make sense of how + works in those languages don't exist in Sherin's
+(2001) catalog of conceptual schemata. In other words, I would argue
+there are conceptual ways a programmer may need to think about
+interpreting or writing a ◻=︎ +◻ symbol template that Sherin doesn't
+enumerate.\footnote{One of the most obvious is the movement from seeing
+  ``◻=︎ +◻'' as a statement of equality to seeing it as the storage of a
+  sum to a variable.} To follow that implication, the canonical body of
+knowledge about which programmers must reason is itself fractured,
+because different languages design their operations around different
+symbolic and conceptual metaphors.
+
+To return to misconceptions, what drives research on students'
+misconceptions is largely a need to get students to program computers
+and reason about computation in ways that are canonically correct. And,
+those are assuredly worthwhile goals. But, as I've argued, we can
+already identify cases where a narrow misconceptions focus is silent
+about or dismissive of students' useful intuitions. We can also identify
+the further problems whereby unilateral emphasis on one language's canon
+opposes the vocabulary of symbolic forms (and diverse conceptual
+schemata) expert programmers ultimately need (Sherin, 2001) Taken in
+total, that silence, dismissiveness, and narrow view of refining
+knowledge perpetuates a deficit-focused discourse about student's
+knowledge in computing.\footnote{Worse still, in my view, is that there
+  is a pattern of researchers failing to discuss or acknowledge
+  students' productive knowledge \emph{even when their own data supports
+  it}.} Such a perspective also fails to address what aspects of
+students reasoning might get broken by attempts to ``fix'' such
+``misconceptions''.
+
+It's important to note that misconceptions research in computing didn't
+always treat students' non-canonical knowledge this way. I begin the
+next section by backtracing to some of the earliest work on students'
+cognitive ``bugs''. There, we find researchers talking more explicitly
+about what's useful in students' non-canonical knowledge---a stance
+largely absent in modern computing misconceptions research.
+
+\subsubsection{Not all cognitive programming bugs imply a problem with
+the
+student}\label{not-all-cognitive-programming-bugs-imply-a-problem-with-the-student}
+
+What's curious about Pea's (1986) comments on Bonar \& Soloway (1983) is
+that Pea drew different conclusions from the data than Bonar and Soloway
+did. Pea (1986) emphasized that when the student thought two
+semantically-equivalent assignment statements in PASCAL were different,
+it was a problem of egocentrism: ``students assume that there is
+\emph{more} of their meaning for what they want to accomplish in the
+program than is actually present in the code they have written'' (p.
+30). In other words, Pea treated the data as a fairly clear example of a
+class of bugs. Specifically, he saw a bug class in which students simply
+assumed the interpreter or runtime could infer the code author's shades
+of intent.
+
+Bonar and Soloway (1983), by contrast, were less quick to make
+inferences either about the nature of the bug or about the intervention
+it entailed. Rather than jump to definitive conclusions, they were
+circumspect:
+
+It is not clear exactly how to react to the bugs we have uncovered in
+novice understanding of programming. In some cases it may be appropriate
+to design new languages or constructs. Often, better instruction would
+take care of the problem. The intent of our studies is to better
+understand the source of the mismatches and misconceptions that cause
+novice bugs. Only once a bug is uncovered and understood are we ready to
+create a remedy for that bug. (Bonar \& Soloway, 1983, p. 12)
+
+That the authors would even consider developing new constructs or
+languages is a noteworthy distinction. Rather than assume in toto that
+students with ``bugs'' had wrong knowledge, the authors instead suppose
+cognitive bugs have plausible origins worth designing around. Moreover,
+they treat students' divergence from canon as \emph{an opportunity for
+research to learn from students}. Precisely because students saw
+functional differences (cf., diSessa, 1986) in semantically-equivalent
+PASCAL statements, Bonar and Soloway (1983) reflect that perhaps
+programming languages should be more expressive:
+
+We find it quite interesting that novices seem to understand the role or
+strategy of statements more clearly than the standard semantics. Such
+roles discussed here include ``counter variable,'' ``running total
+variable,'' ``running total loop,'' and ``first, then rest loop''. (See
+Soloway et al {[}1982b{]} for a detailed discussion of novice looping
+strategies.) Much work in programming languages is concerned with
+allowing a programmer to more accurately express his or her intentions
+in the program. Perhaps we can learn something from novices here - our
+programming systems should support recording the roles the programmer
+intends for various statements and variables. (Bonar \& Soloway, 1983,
+p. 12)
+
+Again, what's noteworthy here is that rather than treating students'
+non-canonical views as a burden for instruction, Bonar and Soloway
+instead see them as an opportunity for programming language designers to
+make languages better.
+
+That insight---that instruction and design can meet novices where they
+are---carries through to their final remarks about studying and
+analyzing novice programming knowledge:
+
+The experience and understanding of a novice are available for analysis.
+In particular, our results suggest that the knowledge people bring from
+natural language has a key effect on their early programming efforts.
+Our work suggests that we need serious study of the knowledge novices
+bring to a computing system. For most computerized tasks there is some
+model that a novice will use in his or her first attempts. We need to
+understand when it is appropriate to appeal to this model, and, when
+necessary, how to move a novice to some more appropriate model. (Bonar
+\& Soloway, 1983, p. 13)
+
+I assume Bonar and Soloway structured their final line deliberately. If
+so, their phrasing has three consequences:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Appealing to novice's existing models gets precedence. That is,
+  understanding how to leverage novices' existing knowledge comes first
+  in research.
+\item
+  Changing the models novices have comes next, and ``when necessary.''
+\item
+  They speak of ``how to move a novice to some more appropriate model,''
+  which does not necessarily entail the removal or destruction of
+  novice's existing models.
+\end{enumerate}
+
+Taken together, these points convey a sense of how Bonar and Soloway
+view learning and instruction in computing. Instruction explicitly
+includes appeals to prior models and knowledge students might already
+have. Learning, meanwhile, involves the movement \emph{when necessary}
+to more appropriate models of computation. As I explain in the next
+section, such a view exactly aligns with a particular branch of
+constructivism, where cognition is viewed as the complex activation of
+manifold resources for thinking and knowing.
+
+\subsubsection{Examples motivate the need for contextual-sensitivity in
+modeling programming
+cognition}\label{examples-motivate-the-need-for-contextual-sensitivity-in-modeling-programming-cognition}
+
+Let's begin with two motivating examples. My aim with these examples is
+to show how a practicing programmer might employ specific, distinct
+conceptual models to reason locally about a piece of code. First,
+consider variants of the PASCAL statements from Bonar and Soloway
+(1983), where an assignment statement worked to increment a value and
+store the result back to that value. In Table 2 below, I imagine five
+different ways of writing a programming statement. Each of the five ways
+is semantically equivalent to the others (or near enough for explanatory
+purposes). And, in each example I add a layer of specificity to the
+syntax. I also propose a corresponding interpretation of how I might
+apply interpret and think about a given statement.\footnote{I assume for
+  simplicity's sake that all variables and functions in my code samples
+  evaluate to floating-point numbers that can be combined under
+  addition. I also assume, crucially, that all statements are in the
+  same hypothetical language and that the variable names connote nothing
+  to the computational interpreter.}
+
+\protect\hypertarget{ux5fToc252445956}{}{}Table -- Five different
+interpretations of semantically equivalent programming statements in the
+same language
+
+\begin{center}
+  \begin{tabular}{ | l | l |}
+    1 & \texttt{x\ =\ x\ +\ d} \\
+    2 & \texttt{x\ =\ x\ +\ update} \\
+    3 & \texttt{x\_position\ =\ x\_position\ +\ update} \\
+    4 & \texttt{x\_position\ =\ x\_position\ +\ update(t)} \\
+    5 & \texttt{x\_position\ =\ x\_position\ +\ update(...)} \\
+  \end{tabular}
+\end{center}
+
+% I stress that the third column is about \emph{how I might think about}
+% each programming statement\emph{.} By no means am I making normative
+% claims about how one \emph{ought} to think about it, or whether the
+% statement actually does what its names might suggest it does. Rather,
+% ``How I might think about it'' reflects the kind of local meaning or
+% interpretation I might attach to such a statement when I work with it,
+% given my understanding of its role and context.
+%
+% Consider specifically statements 1 and 4. Statement 1 could very well be
+% an incrementer in some kind of iterative code. If I had to debug code
+% that employs statement 1, what I might do is exploit my knowledge of
+% what d is (does it hold a constant value? Is it 1?) and try inserting
+% intermediate print statements into the iterative code. Doing so, I can
+% inspect textually how values change with each iteration. In statement
+% 4's case, it could very well be that the position-update code animates
+% images on a screen. If there's a problem with that code, one of the
+% easiest ways I might notice is that the acceleration seems off in the
+% graphics. Consequently, my debugging might call upon knowledge I have
+% from kinematics. I might try inserting code that draws a dot at the
+% object's on-screen position with each iteration, leaving a trail I can
+% visually inspect. I could then look at the path of the object's
+% trajectory and the spacing patterns between successive dots as a
+% first-pass test of whether my code achieves the motion I want.
+%
+% To be clear, it's not just that graphics might improve my efficiency in
+% debugging a statement. Rather, applying an interpretive frame that
+% treats an assignment statement as a kinematic position update lets me
+% \emph{use conceptual knowledge from physics} to diagnose and fix
+% problems in my code. If I view the assignment statement as saying
+% something about the motion of an object (cf., Hammer, 1994, p. 165), a
+% field of knowledge and concomitant techniques from physics becomes
+% available to me to think with. But, if I focus only on the
+% semantic-level equivalence of statements 1 through 5, there would be no
+% obvious reason for me to access what I know about physics in order to
+% reason about the code.
+%
+% My second example concerns statements that are syntactically-similar,
+% rather than semantically-similar. I proposed that the statements in
+% Table 2 all came from the same language. But, another phenomenon comes
+% into play when different languages use the same symbology for
+% conceptually different operations. Table 3 shows examples of what look
+% like semantically-equivalent operations, but in fact are not.
+%
+% \protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
+% syntactically-similar statements with very different semantics
+%
+% \begin{supertabular}[]{@{}llll@{}}
+% \toprule
+% \textbf{Statement } & Programming Syntax & Language & How I think about
+% it\tabularnewline
+% \midrule
+% \endhead
+% \textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
+% \textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
+% the string w\tabularnewline
+% \textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
+% layer of points onto a plot p\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% The catch here is that the plus operator takes on different roles in
+% different languages because of how those languages define its use.
+% Statement 1 increments a number in C; Statement 2 appends the letters
+% ``ly'' to a string; Statement 3 adds a layer of points to a statistical
+% graphics plot. These different kinds of operations become even more
+% apparent and consequential when, for example, such statements are
+% repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
+%   operator can be written as a directional arrow. The symbol
+%   \textless{}- (``less-than, hyphen'') indicates the value on the right
+%   side of the symbol is being assigned to the variable on the left side
+%   of the symbol. As used, the symbol itself is semantically equivalent
+%   to having written an equals sign (``=``).} I'm using multiple
+% reassignment statements to compose a statistical graphics plot. The
+% layered creation of a plot invites a very different kind of conceptual
+% interpretation than, say, repeatedly accumulating numbers into a running
+% sum:
+%
+% p \textless{}- InitializeGgplot\_1w()
+%
+% p \textless{}- p + GrandMeanLine(owp)
+%
+% p \textless{}- p + GrandMeanPoint(owp)
+%
+% p \textless{}- p + ScaleX\_1w(owp)
+%
+% p \textless{}- p + ScaleY\_1w(owp)
+%
+% p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)
+%
+% One obvious reason for thinking about this code with a different
+% interpretive frame is that numeric addition is commutative; composing a
+% plot is not necessarily commutative.\footnote{To convince yourself that
+%   plot composition isn't commutative, imagine a scale function that
+%   squares up the aspect ratio of a plot and another scale function that
+%   sets the aspect ratio to 1.5:1. Applying the square function last
+%   produces a square plot; applying it first produces a rectangular plot.
+%   String concatenation is also not necessarily commutative. ``cat'' +
+%   ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
+%   ``dogcat.''} So, despite the syntactic similarity, reassignments that
+% compose a plot using reassignment (as above) does not obey the same
+% rules as reassignments for a running total. But, even within this code
+% block, statements that look alike perform operations of a different
+% nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
+% visual layer to a plot, others modify features of the plot (e.g, ``+
+% Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
+% the numeric range of the data).
+%
+% Given these motivating examples, it seems sensible to think there's
+% utility in a programmer having different conceptual metaphors available
+% to think about and work with code. Example 1 shows that choosing to
+% apply knowledge from physics to a piece of code can change the cognitive
+% nature of debugging. There, debugging a position-update statement
+% becomes, in part, reasoning kinematically about the properties of motion
+% trails.\footnote{For comparison, consider the argument that ringing an
+%   aircraft speedometer with physical markers changes the nature of
+%   cognition when pilots work to land a plane (Hutchins, 1995b).} Example
+% 2 shows that across languages, programmers might have to deploy
+% different conceptual metaphors to reason about statements in a
+% locally-consistent way. Knowing that plots in ggplot2 can be composed
+% layer-by-layer with reassignment is crucial if you're trying to write or
+% understand code that creates statistical graphics. But, I would argue
+% that thinking about ◻ = ◻ + ◻ as ``compose new layer onto plot'' can and
+% does appeal to different kinds of knowledge when compared to thinking
+% about ◻ = ◻ + ◻ as ``include this addend in the sum,'' which itself can
+% and does appeal to different kinds of concepts when compared to thinking
+% about ◻ = ◻ + ◻ as ``increment the counter.''
+%
+% Stepping back, I can build the following argument
+%
+% \begin{enumerate}
+% \def\labelenumi{\arabic{enumi}.}
+% \item
+%   Programming can be helped by applying conceptual models to code,
+%   particularly when relevant domain-knowledge structures can
+%   advantageously transform a problem (example 1)
+% \item
+%   But, conceptual models don't work all the time for all statements.
+%   Because languages are designed differently, the same syntax can
+%   actually correspond to very different operations in code (example 2).
+%   And, that's true both within and across languages.
+% \item
+%   Consequently, it makes less sense to treat conceptual models as right
+%   or wrong, and more sense to treat them as differentially advantageous
+%   for thinking about what a piece of code does. (Thinking a ``+''
+%   implies numerical addition isn't \emph{globally wrong} in JavaScript,
+%   but it won't explain why 1 + ``1'' yields ``11'' as a result.)
+% \item
+%   It seems plausible that successful programmers, when reasoning about
+%   or writing code, are able to dynamically access or deploy conceptual
+%   models that are advantageous given the context (language, syntax,
+%   surrounding code). Certainly such a supposition is in line with work
+%   suggesting students have resources for thinking conceptually and
+%   dynamically about how mathematics models real-world situations (Izsák,
+%   2004; Sherin, 2001)
+% \item
+%   To model how programmers think with conceptual models, a suitable
+%   framework should be able to account for the dynamic, context-sensitive
+%   deployment of conceptual knowledge.
+% \item
+%   To model how programmers develop expertise, a suitable framework
+%   should be able to describe higher-order phenomena. Such phenomena
+%   include explaining how programmers come to have conceptual models or
+%   generate new ones, why they decide to deploy them, and how programmers
+%   consider which conceptual model (i.e., which way to think about code)
+%   is appropriate.
+% \end{enumerate}
+%
+% Taken together, these assertions propose criteria for how we might
+% strive to model cognition in programming. Our modeling frameworks should
+% be context-dependent, dynamic, and capable of explaining where
+% conceptual models come from. They should also be able to account for
+% phenomena that are not themselves conceptual, including what directs the
+% use of certain kinds of conceptual knowledge. In the learning sciences,
+% such models already exist and have proven useful and productive for
+% thinking about thinking.
+%
+% \subsubsection{Manifold models of cognition explain context-dependence
+% and the growth of
+% expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}
+%
+% In 1993, a pair of articles in the learning sciences staked a strong
+% claim for viewing knowledge as a network of pieces, isolated enough to
+% be locally triggered but trainable enough to fire in larger concerted
+% patterns (diSessa, 1993; Smith, diSessa, \& Roschelle, 1993). Informed
+% in part by agent-based accounts of cognition (Minsky, 1986) and complex
+% systems models (diSessa, 2002), the central tenets of an ``in-pieces''
+% approach hold that knowledge is emergent from interacting primitives,
+% rather than unitary and monolithic. An example from Smith, diSessa, and
+% Roschelle helps illustrate the point.
+%
+% The authors show that we might think of a rubber band as a different
+% conceptual entity depending on context. In several different
+% situations---wrapped around a newspaper, pulled taut as a string, spun
+% to store energy in a toy plane propeller---we intuitively think about
+% the rubber band's physical behavior differently: one as a negligible
+% part of the newspaper's point mass, another as a transverse pendulum
+% (likely) obeying Hooke's Law, and the third as a torsional spring. Those
+% differences in intuitive thinking reflect the contextual dependency of
+% what we know about the physical world:
+%
+% In each of the rubberband examples, various pieces of intuitive physical
+% knowledge describe the mechanism at work: the rubber band binds the
+% newspaper, grips the jar lid, and acts a source of springiness for the
+% bobbing object. Although a mapping cannot be made from the rubberband to
+% scientific entities, it is quite easy to map these qualitatively
+% distinct physical processes to scientific entities and laws. For
+% example, instances of binding almost always map to a practically rigid
+% body. Likewise, gripping maps to friction forces, and springiness maps
+% to Hooke's law. This suggests that applicability can depend directly on
+% our intuitive knowledge---knowledge that exists prior to any formal
+% scientific training (Smith et al., 1993, p. 144).
+%
+% The in-pieces approach to modeling cognition has been used, among other
+% things, to explain how experts reason about fractions and decimals
+% (Smith et al., 1993), how students reason about forces in physics
+% (diSessa \& Sherin, 1998; diSessa, 1993; Hammer, 1996; Sherin, 2001),
+% how students construct and evaluate algebraic representations of
+% physical situations (Izsák, 2004), and how knowledge transfers across
+% contexts (Hammer et al., 2005; Wagner, 2006). Because its starting
+% assumption is that knowledge is fragmented, knowledge-in-pieces can
+% account for wide variations of how people---particularly novices---use
+% knowledge on a moment-to-moment basis. In other words, because it
+% assumes knowledge is local, it can still explain the kinds of
+% globally-inconsistent ways people might reason about physical situations
+% (diSessa, 1993). As a framework, an in-pieces approach ultimately argues
+% that models of concept replacement and good/bad criteria for knowledge
+% should be supplanted by a learning model of alignment/refinement of
+% prior knowledge and the consideration of knowledge as
+% productive/unproductive.
+%
+% Hammer and colleagues have worked to extend the in-pieces approach to
+% explain how students' epistemological activity---how they orient toward
+% knowledge and knowing in a context (Hammer et al., 2005; Hammer \& Elby,
+% 2002, 2003). Specifically, those authors use two core theoretical
+% constructs to explain students' stances toward knowledge and knowing:
+%
+% \begin{itemize}
+% \item
+%   \emph{Epistemological Resources} are the epistemological equivalent of
+%   diSessa's phenomenological primitives (p-prims). Resources, the
+%   authors propose, are the atomic units involved in how people cognize
+%   about the source of knowledge, the nature of knowledge, and
+%   epistemological activities (Hammer \& Elby, 2003; Louca, Elby, Hammer,
+%   \& Kagey, 2004).
+% \item
+%   \emph{Epistemological Frames} are the emergent result of subsets of
+%   resources acting in concert. Drawing from both Goffman's (1974)
+%   sociological notion of frame as structures of expectations and
+%   subsequent work on framing in discourse (Tannen, 1993),
+%   epistemological frames are a participant's local answer to the
+%   question ``what is it {[}specifically, what knowledge activity{]}
+%   that's going on here'' (Goffman, 1974, p. 8).
+% \end{itemize}
+%
+% Resources can frames can interact in activity settings to produce
+% larger-scale patterns called ``epistemological coherences'' (Rosenberg,
+% Hammer, \& Phelan, 2006) where evidence from data suggests that a
+% network of discrete cognitive units can nonetheless give rise to stable
+% cognition.
+%
+% A useful conceptual metaphor for resources and framing is to think about
+% a lecture hall with different sets of lights: a spotlight in the back to
+% highlight a lecturer, chalkboard lights, house lights, and a projector.
+% Each light has an individual brightness, but the lights are only
+% controllable at the per-bulb level. In that sense, they are atomic. But,
+% light patterns interact with one another to create a field of lighting
+% for the room. Thus, lights are a bit like resources, and different light
+% configurations can correspond to different, sociologically-stable uses
+% of the room.
+%
+% \protect\hypertarget{ux5fToc252445958}{}{}Table -- Using room lighting
+% configurations to think about epistemological framing
+%
+% \begin{supertabular}[]{@{}lllll@{}}
+% \toprule
+% \textbf{What is it that's going on?} & Houselights & Spotlight &
+% Chalkboard & Projector\tabularnewline
+% \midrule
+% \endhead
+% \textbf{On-stage monologue} & Low & ON & Low & OFF\tabularnewline
+% \textbf{Presenting slides} & Low & ON & Low & ON\tabularnewline
+% \textbf{Working through equations} & Low & ON & High &
+% OFF\tabularnewline
+% \textbf{Students discussing with each other} & High & OFF & Low &
+% OFF\tabularnewline
+% \textbf{Watching a movie} & Low & OFF & Low & ON\tabularnewline
+% \textbf{Cleaning the room after a movie} & High & OFF & High &
+% OFF\tabularnewline
+% \bottomrule
+% \end{supertabular}
+%
+% To convince yourself of the sociological stability of these
+% configurations, imagine you were watching a movie when suddenly the
+% house and chalkboard lights came up to full intensity. It would jar you
+% out of the experience. You might wonder whether something was wrong: is
+% there an emergency? Should you evacuate?
+%
+% The metaphor isn't perfect. Ontologically, for example, resource and
+% framing theory propose these constructs exist not in the world but in
+% the minds of individuals. But, the metaphor is quite useful for
+% understanding how individual elements---in this case lights---can work
+% in concert to create and sustain a stable frame. And, crucially for a
+% cognitive system, the metaphor lets us account for variation in
+% activity. The same lecture hall can become a window onto one person's
+% thoughts (monologue), a site for instruction (working through equations
+% at the board), a shared space for collaboration (student discussion), or
+% a place in need of repair (cleaning up) simply by varying the
+% intensities of lighting banks in particular ways.
+%
+% An example helps ground this in-pieces approach to epistemology. In
+% Russ, Coffey, Hammer, and Hutchison (2008), the authors describe the
+% situation where an elementary student reasons about why an empty juice
+% box collapses when you suck on the straw. One student gives what the
+% authors deem to be an excellent mechanistic account of why the juice box
+% collapses. But, as the teacher seems to steer the discussion toward
+% vocabulary---in this case, ``pressure''---the student clearly pulls back
+% from her mechanistic reasoning, seems much more diffident, and claims
+% that pressure is hard to explain. That example highlights the disconnect
+% between doing science as knowing vocabulary and doing science as
+% reasoning mechanistically. Moreover, it strikingly highlights that a
+% student who by all accounts produced an excellent explanation of how
+% pressure works was left nonetheless with the impression that pressure
+% was hard to explain.\footnote{In fairness to the teacher, I'm trying to
+%   focus on the result of the interaction and far less on the intent the
+%   teacher might have had. The student still left with a sense that
+%   science might be about vocabulary, even if that's not the view the
+%   teacher would espouse or was trying to enact in the moment.}
+%
 
@@
 
-Blinded for review. \cite{hammer_tapping_2003}
-
-\cite{hammer_two_1989}
+Blinded for review.
 

### 2016-04-22 manuscript-body.tex
Get manuscript to at least compile

@@
-\subsection{Introduction}\label{introduction-2}
-
-Although imaginative and carefully designed, the Harvard course taught
-that there is only one right way to approach the computer, a way that
-emphasizes control through structure and planning. There are many
-virtues to this computational approach (it certainly makes sense when
-dividing the labor on a large programming project), but Lisa and Robin
-have intellectual styles at war with it. Lisa says she has ``turned
-herself into a different kind of person'' in order to perform, and Robin
-says she has learned to ``fake it.'' Although both women are able to get
-good grades in their programming course, they represent casualties of
-this war. Both deny who they are in order to succeed. (Turkle \& Papert,
-1991, p. 165)
-
-Over 20 years ago Sherry Turkle and Seymour Papert called for
-epistemological pluralism in computing education. Their argument --- in
-my view ---had multiple parts:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Programming a computer involved a shift in how one thought about the
-  nature knowledge from a propositional \emph{what is} to an imperative
-  \emph{how to} (Abelson \& Sussman, 1996; Papert, 1980).
-\item
-  Historically, there was a tendency to treat computers and the act of
-  programming them as an extension of the same Western logico-deductive
-  knowledge tradition that fueled science.
-\item
-  Recent scholarship, often at the intersection of post-structuralism
-  and social studies of science, had been troubling the idea that
-  knowledge-work in science proceeded solely by way of logical
-  deductions (Keller, 1983; Latour, 1987; Traweek, 1988)
-\item
-  Data from professionals suggested programming, too, didn't proceed
-  solely by way of the structured-planning approach emergent from
-  Western logico-deductive knowledge traditions. There were many ways of
-  knowing and constructing knowledge in programming.
-\item
-  A viewpoint of \emph{epistemological pluralism} --- embracing multiple
-  kinds and ways of knowing --- was informing the field of science
-  studies but not, to a large degree, how education researchers thought
-  about or taught computing.
-\item
-  The lack of support for diverse ways of knowing in computing
-  classrooms demonstrably hurt students.
-\end{enumerate}
-
-To Turkle and Papert (1991), the result of this culturally-rooted
-epistemological tension was a war. More specifically, it was a war where
-students of computing were the casualties, and the aggressor was a
-distributed failure to recognize, embrace, and support diverse ways of
-knowing. While I think their strong-form characterization of a ``war''
-no longer applies, their consideration of epistemological issues is
-still relevant two decades later. An understanding that computing does
-in fact involve diverse ways of knowing should continue to inform
-research and discussions on learning
-
-This paper tries to model what directs and sustains students'
-in-the-moment activity when they program. Its focus is on early-stage
-program design; analyzing how what students say, do, write, and gesture
-\emph{even before they type code} can help us improve theories of
-cognition and activity in learning programming. Using a ``revelatory
-case study'' (Yin, 2009) of two students in an introductory programming
-course, I argue the following:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Students' early-stage design activity reveals patterns outside the
-  explanatory scope of misconception-based accounts of cognition.
-\item
-  For a subset of phenomena, we can recast students' productive
-  capacities and their difficulties in terms of epistemological stances.
-\item
-  Such recastings are analytically powerful. Beyond Rebecca and Lionel,
-  they could potentially explain the diversity of practices I saw other
-  students take up.
-\item
-  As evidenced by work in science and math education, dynamic
-  epistemological models can offer a lens for reforming assessment and
-  instruction.
-\end{enumerate}
-
-\subsection{Literature Review}\label{literature-review}
-
-I begin by exploring the historical legacy of Turkle and Papert's (1991)
-remarks on epistemological pluralism in computing. I'll then describe
-what I see as a well-intentioned obstacle: the research-based
-preoccupation with students' misconceptions in computing education.
-These two stra.
-
-\subsubsection{Computing culture should support a diversity of ways of
-knowing}\label{computing-culture-should-support-a-diversity-of-ways-of-knowing}
-
-Turkle \& Papert (1991) describe a tension between individuals and a
-kind of cultural collective. In this tension, individual students and
-their personally-identified ways of knowing were in conflict with the
-larger culture of computing and its manifested ways of knowing. Lisa,
-for instance, was a Harvard student who identified as a poet:
-
-Lisa experiences language as transparent, she knows where all the
-elements are at every point in the development of her ideas. She wants
-her relationship to computer language to be similarly transparent. When
-she builds large programs she prefers to write her own smaller
-``building block'' procedures even though she could use prepackaged ones
-from a program library; she resents the latter's opacity. (Turkle \&
-Papert, 1991, p. 164)
-
-Turkle and Papert connect Lisa's programming practices to Lisa's
-personal view of knowledge. She writes building block procedures because
-of a continuity between herself as knower-of-her-own-poetry and that of
-knower-of-her-own-programs. Across both contexts, Lisa \emph{resents
-opacity} because it is an obstacle to her knowing. But, that resentment
-slows her progress on projects in the class.
-
-In part because of her commitments to transparency and the difficulty
-they created, Lisa changed her approach to programming. She ultimately
-had to, in her words, ``be a different kind of person with the machine''
-(Turkle \& Papert, 1991, p. 164). Turkle and Papert explain:
-
-She had been told that the ``right way'' to do things was to control a
-program through planning and black­boxing, the technique that lets you
-exploit opacity to plan something large without knowing in advance how
-the details will be managed. Lisa recognized the value of these
-techniques --- for someone else. She struggled against using them as the
-starting points for her learning. Lisa ended up abandoning the fight,
-doing things ``their way,'' and accepting the inevitable alienation from
-her work. (Turkle \& Papert, 1991, p. 164)
-
-The phrase ``inevitable alienation'' is particularly damning. One might
-accuse Turkle and Papert (1991) of hyperbole, but research from other
-disciplines suggests they're not overstating the case. Identity
-alienation for epistemological reasons --- or some variation of them ---
-is playing out in other subjects as well.
-
-Connections and tensions between identities, learning, and personal ways
-of knowing are not unique to computing education. Wortham (2006), for
-example, detailed instances of social identification and academic
-identity as \emph{jointly emerging} in a combined high school English
-and History class. Nasir and colleagues explored how different learning
-environments afford access to identity, domain practices, and
-self-expression (Nasir \& Cooks, 2009; Nasir \& Hand, 2008). To single
-out one study, Nasir \& Hand (2008) followed players ``from the
-{[}basketball{]} court to the classroom.'' They found the court afforded
-players a sense of role on the team, chances to be expressive through
-play, and ``access to the domain {[}of basketball{]} as a whole'' (Nasir
-\& Hand, 2006, p. 147). Comparatively, the mathematics classroom offered
-little (if any) sense that students were on a team, a much narrower role
-for them as having either right or wrong answers, and no strong sense of
-self-expression or creativity through activity. These markedly different
-contexts led not only to differences in kinds of activity but, Nasir and
-Hand (2006) argue, divergent \emph{practice-linked identities} for the
-players. In a sense, players had a different identity on the court than
-they did in the mathematics classroom because who they were was strongly
-coupled to a sense of their relationship with a practice.
-
-Boaler's work on learning and identity in mathematics classrooms (Boaler
-\& Greeno, 2000; Boaler, 1998, 2000, 2002) offers resonant findings. In
-particular, Boaler (2000) strongly echoes the student voices in Turkle
-and Papert (1991). Where Turkle and Papert's aggressor is the dominant
-computing culture, Boaler describes a beast with similar effects in the
-form of ``school mathematics'' --- the \emph{monotonous},
-\emph{meaningless} (Boaler, 2002, pp. 383--384) imposition of routinized
-mathematical procedures in ways that seem far removed from the real
-world. Of school mathematics, Boaler writes:
-
-School mathematics, for many of them, was \emph{of another world} and to
-fully engage in that world, students needed to suspend their knowledge
-of the real world, suppress their desire to interact with others, and
-strive to reproduce standard procedures that held little meaning for
-them. (Boaler, 2002, p. 392 emphasis in original)
-
-Schoenfeld (1988, 1991) had already argued school mathematics could and
-did depart substantially from the core of mathematical thinking. Boaler
-established that such school mathematics could alienate learners
-(Boaler, 1998, 2000) and influence their feelings about pursuing
-mathematics after secondary school (Boaler \& Greeno, 2000).
-
-Most recently, Danielak, Gupta, and Elby (in press) extended those
-findings in engineering education. In their 3 ½ year case study of an
-electrical engineering undergraduate named ``Michael,'' those authors
-argued Michael's practice of sense-making in engineering coursework was
-strongly coupled to his identity. Michael's passion for deeply
-understanding concepts, much like Lisa's (Turkle \& Papert, 1991) desire
-to understand every element of her program, did not mesh neatly with the
-larger culture of his courses. But, because Michael's practice of
-sense-making entwined with his identity, forces that pushed against
-sense-making alienated him. That tension was particularly pointed when
-Michael described his father's resistance to Michael's way of thinking:
-
-{[}My dad said{]} ``you're just an undergraduate. Nobody expects
-undergraduates to understand how anything works. That's why you go to
-graduate school.'' I was like ``look, you know. I'm gonna be an unhappy
-person if I have to....I have life goals other than to just get a good
-grade on the exam. Other things are important to me.'' (Danielak et al.,
-in press)
-
-And, much like Turkle and Papert's (1991) Lisa, Michael learned to
-curtail parts of his identity to get by in class:
-
-I think the reason {[}pursuing deep learning{]} actually hasn't affected
-my GPA is because I view learning as a hobby. So, as with any hobby, you
-shouldn't let it interfere your GPA. But it \emph{is} one of my hobbies,
-and I \emph{do} enjoy learning, I just---up to the point where I get my
-grades done \{raises eyebrow\}. (Danielak et al., in press)
-
-Ultimately, research suggests the identity-epistemology tensions Turkle
-and Papert (1991) described in computing resonates with those found in
-related disciplines. And, what such accounts --- particularly those of
-Boaler and others (Boaler \& Greeno, 2000; Boaler, 1998, 2000, 2002;
-Danielak et al., in press) --- have in common is a culturally-sanctioned
-``right way'' (Turkle \& Papert, 1991, p. 164) of thinking or knowing
-that alienates some students. Avowedly such ways of knowing have use and
-value. Turkle \& Papert themselves acknowledge ``there are many
-virtues'' to black-boxing and top-down design in programming (Turkle \&
-Papert, 1991, p. 165). But,
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  When a preponderance of results from science studies suggests a
-  plurality of ways of knowing among practicing scientists and
-  programmers, and
-\item
-  When research from education shows that the inflexible imposition of
-  one way of knowing above all else is alienating students, then
-\item
-  It's worth reconsidering whether a narrow sense of what counts as
-  knowing might still be hindering improvements in computing education.
-\end{enumerate}
-
-In the next section, I turn to misconceptions research in computing
-education. In keeping with observations 1--3,
-
 \subsubsection{Misconceptions research in computing education tends to
@@
 The cause of those errors is theorized to be something cognitive,
-whether it's a ``bug'' (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
+whether it's a ``bug'' \cite{pea_buggy_1987} (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
 ``misconception'' (Bayman \& Mayer, 1983; Bonar \& Soloway, 1985;
@@
 \begin{itemize}
+\tightlist
 \item
@@
 %
-
-\subsection{Acknowledgments}\label{acknowledgments}
-
-Blinded for review.
-
-\clearpage
-
-\bibliographystyle{acm}
-\bibliography{bibliography}

### 2016-04-22 manuscript-body.tex
Promot subsubsections to subsections

@@
-\subsubsection{Misconceptions research in computing education tends to
+\subsection{Misconceptions research in computing education tends to
 ignore students' productive
@@
 
-\subsubsection{Not all cognitive programming bugs imply a problem with
+\subsection{Not all cognitive programming bugs imply a problem with
 the
@@
 
-\subsubsection{Examples motivate the need for contextual-sensitivity in
+\subsection{Examples motivate the need for contextual-sensitivity in
 modeling programming
@@
 %
-% \subsubsection{Manifold models of cognition explain context-dependence
+% \subsection{Manifold models of cognition explain context-dependence
 % and the growth of

### 2016-04-22 manuscript-body.tex
Convert references to natbib in first paragraph

@@
 The cause of those errors is theorized to be something cognitive,
-whether it's a ``bug'' \cite{pea_buggy_1987} (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
-``misconception'' (Bayman \& Mayer, 1983; Bonar \& Soloway, 1985;
-Clancy, 2004; Gal-Ezer \& Zur, 2004; Herman et al., 2008; Kaczmarczyk et
-al., 2010), a ``belief'' (Fleury, 1993) or a ``student-constructed
-rule'' (Fleury, 1991, 2000). Instruction should try to identify,
-address, and correct these misconceptions (Clancy, 2004) because they
-can represent barriers to learning.
+whether it's a ``bug'' \cite{pea_buggy_1987,pea_languageindependent_1986,vanlehn_mind_1990}, a
+``misconception'' \cite{bayman_diagnosis_1983,bonar_preprogramming_1985,clancy_misconceptions_2004,gal-ezer_efficiency_2004,herman_proof_2008,kaczmarczyk_identifying_2010}, a ``belief'' \cite{fleury_student_1993}, or a ``student-constructed
+rule'' \cite{fleury_parameter_1991,fleury_programming_2000} in programming. Instruction, Clancy argues, should try to identify,
+address, and correct these misconceptions because they
+can represent barriers to learning \cite{clancy_misconceptions_2004}.
 
@@
 serve different conceptual purposes. Hypothetically, the student might
-use that information in design by calling up the ``◻=︎ + 1'' syntactical
+use that information in design by calling up the $\Box=︎ \Box + 1$ syntactical
 template when the situation seems to demand \emph{keeping control},
-while calling up ``◻= ︎ + number'' when \emph{totaling up} is the goal.
+while calling up $\Box= ︎\Box + \mathrm{number}$ when \emph{totaling up} is the goal.
 And, the idea that structurally identical symbol templates can serve
@@
 would argue is the same concrete symbol template of how to write an
-expression: ◻=︎ + ◻.
+expression: $\Box=︎ \Box + \Box$.
 
@@
 there are conceptual ways a programmer may need to think about
-interpreting or writing a ◻=︎ +◻ symbol template that Sherin doesn't
+interpreting or writing a $\Box=︎ \Box +\Box$ symbol template that Sherin doesn't
 enumerate.\footnote{One of the most obvious is the movement from seeing
-  ``◻=︎ +◻'' as a statement of equality to seeing it as the storage of a
+  $\Box=︎ \Box +\Box$ as a statement of equality to seeing it as the storage of a
   sum to a variable.} To follow that implication, the canonical body of
@@
 
-% I stress that the third column is about \emph{how I might think about}
-% each programming statement\emph{.} By no means am I making normative
-% claims about how one \emph{ought} to think about it, or whether the
-% statement actually does what its names might suggest it does. Rather,
-% ``How I might think about it'' reflects the kind of local meaning or
-% interpretation I might attach to such a statement when I work with it,
-% given my understanding of its role and context.
-%
-% Consider specifically statements 1 and 4. Statement 1 could very well be
-% an incrementer in some kind of iterative code. If I had to debug code
-% that employs statement 1, what I might do is exploit my knowledge of
-% what d is (does it hold a constant value? Is it 1?) and try inserting
-% intermediate print statements into the iterative code. Doing so, I can
-% inspect textually how values change with each iteration. In statement
-% 4's case, it could very well be that the position-update code animates
-% images on a screen. If there's a problem with that code, one of the
-% easiest ways I might notice is that the acceleration seems off in the
-% graphics. Consequently, my debugging might call upon knowledge I have
-% from kinematics. I might try inserting code that draws a dot at the
-% object's on-screen position with each iteration, leaving a trail I can
-% visually inspect. I could then look at the path of the object's
-% trajectory and the spacing patterns between successive dots as a
-% first-pass test of whether my code achieves the motion I want.
-%
-% To be clear, it's not just that graphics might improve my efficiency in
-% debugging a statement. Rather, applying an interpretive frame that
-% treats an assignment statement as a kinematic position update lets me
-% \emph{use conceptual knowledge from physics} to diagnose and fix
-% problems in my code. If I view the assignment statement as saying
-% something about the motion of an object (cf., Hammer, 1994, p. 165), a
-% field of knowledge and concomitant techniques from physics becomes
-% available to me to think with. But, if I focus only on the
-% semantic-level equivalence of statements 1 through 5, there would be no
-% obvious reason for me to access what I know about physics in order to
-% reason about the code.
-%
-% My second example concerns statements that are syntactically-similar,
-% rather than semantically-similar. I proposed that the statements in
-% Table 2 all came from the same language. But, another phenomenon comes
-% into play when different languages use the same symbology for
-% conceptually different operations. Table 3 shows examples of what look
-% like semantically-equivalent operations, but in fact are not.
-%
-% \protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
-% syntactically-similar statements with very different semantics
-%
+I stress that the third column is about \emph{how I might think about}
+each programming statement\emph{.} By no means am I making normative
+claims about how one \emph{ought} to think about it, or whether the
+statement actually does what its names might suggest it does. Rather,
+``How I might think about it'' reflects the kind of local meaning or
+interpretation I might attach to such a statement when I work with it,
+given my understanding of its role and context.
+
+Consider specifically statements 1 and 4. Statement 1 could very well be
+an incrementer in some kind of iterative code. If I had to debug code
+that employs statement 1, what I might do is exploit my knowledge of
+what d is (does it hold a constant value? Is it 1?) and try inserting
+intermediate print statements into the iterative code. Doing so, I can
+inspect textually how values change with each iteration. In statement
+4's case, it could very well be that the position-update code animates
+images on a screen. If there's a problem with that code, one of the
+easiest ways I might notice is that the acceleration seems off in the
+graphics. Consequently, my debugging might call upon knowledge I have
+from kinematics. I might try inserting code that draws a dot at the
+object's on-screen position with each iteration, leaving a trail I can
+visually inspect. I could then look at the path of the object's
+trajectory and the spacing patterns between successive dots as a
+first-pass test of whether my code achieves the motion I want.
+
+To be clear, it's not just that graphics might improve my efficiency in
+debugging a statement. Rather, applying an interpretive frame that
+treats an assignment statement as a kinematic position update lets me
+\emph{use conceptual knowledge from physics} to diagnose and fix
+problems in my code. If I view the assignment statement as saying
+something about the motion of an object (cf., Hammer, 1994, p. 165), a
+field of knowledge and concomitant techniques from physics becomes
+available to me to think with. But, if I focus only on the
+semantic-level equivalence of statements 1 through 5, there would be no
+obvious reason for me to access what I know about physics in order to
+reason about the code.
+
+My second example concerns statements that are syntactically-similar,
+rather than semantically-similar. I proposed that the statements in
+Table 2 all came from the same language. But, another phenomenon comes
+into play when different languages use the same symbology for
+conceptually different operations. Table 3 shows examples of what look
+like semantically-equivalent operations, but in fact are not.
+
+\protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
+syntactically-similar statements with very different semantics
+
 % \begin{supertabular}[]{@{}llll@{}}
@@
 % \end{supertabular}
-%
-% The catch here is that the plus operator takes on different roles in
-% different languages because of how those languages define its use.
-% Statement 1 increments a number in C; Statement 2 appends the letters
-% ``ly'' to a string; Statement 3 adds a layer of points to a statistical
-% graphics plot. These different kinds of operations become even more
-% apparent and consequential when, for example, such statements are
-% repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
-%   operator can be written as a directional arrow. The symbol
-%   \textless{}- (``less-than, hyphen'') indicates the value on the right
-%   side of the symbol is being assigned to the variable on the left side
-%   of the symbol. As used, the symbol itself is semantically equivalent
-%   to having written an equals sign (``=``).} I'm using multiple
-% reassignment statements to compose a statistical graphics plot. The
-% layered creation of a plot invites a very different kind of conceptual
-% interpretation than, say, repeatedly accumulating numbers into a running
-% sum:
-%
-% p \textless{}- InitializeGgplot\_1w()
-%
-% p \textless{}- p + GrandMeanLine(owp)
-%
-% p \textless{}- p + GrandMeanPoint(owp)
-%
-% p \textless{}- p + ScaleX\_1w(owp)
-%
-% p \textless{}- p + ScaleY\_1w(owp)
-%
-% p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)
-%
-% One obvious reason for thinking about this code with a different
-% interpretive frame is that numeric addition is commutative; composing a
-% plot is not necessarily commutative.\footnote{To convince yourself that
-%   plot composition isn't commutative, imagine a scale function that
-%   squares up the aspect ratio of a plot and another scale function that
-%   sets the aspect ratio to 1.5:1. Applying the square function last
-%   produces a square plot; applying it first produces a rectangular plot.
-%   String concatenation is also not necessarily commutative. ``cat'' +
-%   ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
-%   ``dogcat.''} So, despite the syntactic similarity, reassignments that
-% compose a plot using reassignment (as above) does not obey the same
-% rules as reassignments for a running total. But, even within this code
-% block, statements that look alike perform operations of a different
-% nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
-% visual layer to a plot, others modify features of the plot (e.g, ``+
-% Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
-% the numeric range of the data).
-%
-% Given these motivating examples, it seems sensible to think there's
-% utility in a programmer having different conceptual metaphors available
-% to think about and work with code. Example 1 shows that choosing to
-% apply knowledge from physics to a piece of code can change the cognitive
-% nature of debugging. There, debugging a position-update statement
-% becomes, in part, reasoning kinematically about the properties of motion
-% trails.\footnote{For comparison, consider the argument that ringing an
-%   aircraft speedometer with physical markers changes the nature of
-%   cognition when pilots work to land a plane (Hutchins, 1995b).} Example
-% 2 shows that across languages, programmers might have to deploy
-% different conceptual metaphors to reason about statements in a
-% locally-consistent way. Knowing that plots in ggplot2 can be composed
-% layer-by-layer with reassignment is crucial if you're trying to write or
-% understand code that creates statistical graphics. But, I would argue
-% that thinking about ◻ = ◻ + ◻ as ``compose new layer onto plot'' can and
-% does appeal to different kinds of knowledge when compared to thinking
-% about ◻ = ◻ + ◻ as ``include this addend in the sum,'' which itself can
-% and does appeal to different kinds of concepts when compared to thinking
-% about ◻ = ◻ + ◻ as ``increment the counter.''
-%
-% Stepping back, I can build the following argument
-%
-% \begin{enumerate}
-% \def\labelenumi{\arabic{enumi}.}
-% \item
-%   Programming can be helped by applying conceptual models to code,
-%   particularly when relevant domain-knowledge structures can
-%   advantageously transform a problem (example 1)
-% \item
-%   But, conceptual models don't work all the time for all statements.
-%   Because languages are designed differently, the same syntax can
-%   actually correspond to very different operations in code (example 2).
-%   And, that's true both within and across languages.
-% \item
-%   Consequently, it makes less sense to treat conceptual models as right
-%   or wrong, and more sense to treat them as differentially advantageous
-%   for thinking about what a piece of code does. (Thinking a ``+''
-%   implies numerical addition isn't \emph{globally wrong} in JavaScript,
-%   but it won't explain why 1 + ``1'' yields ``11'' as a result.)
-% \item
-%   It seems plausible that successful programmers, when reasoning about
-%   or writing code, are able to dynamically access or deploy conceptual
-%   models that are advantageous given the context (language, syntax,
-%   surrounding code). Certainly such a supposition is in line with work
-%   suggesting students have resources for thinking conceptually and
-%   dynamically about how mathematics models real-world situations (Izsák,
-%   2004; Sherin, 2001)
-% \item
-%   To model how programmers think with conceptual models, a suitable
-%   framework should be able to account for the dynamic, context-sensitive
-%   deployment of conceptual knowledge.
-% \item
-%   To model how programmers develop expertise, a suitable framework
-%   should be able to describe higher-order phenomena. Such phenomena
-%   include explaining how programmers come to have conceptual models or
-%   generate new ones, why they decide to deploy them, and how programmers
-%   consider which conceptual model (i.e., which way to think about code)
-%   is appropriate.
-% \end{enumerate}
-%
-% Taken together, these assertions propose criteria for how we might
-% strive to model cognition in programming. Our modeling frameworks should
-% be context-dependent, dynamic, and capable of explaining where
-% conceptual models come from. They should also be able to account for
-% phenomena that are not themselves conceptual, including what directs the
-% use of certain kinds of conceptual knowledge. In the learning sciences,
-% such models already exist and have proven useful and productive for
-% thinking about thinking.
-%
-% \subsection{Manifold models of cognition explain context-dependence
-% and the growth of
-% expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}
-%
-% In 1993, a pair of articles in the learning sciences staked a strong
-% claim for viewing knowledge as a network of pieces, isolated enough to
-% be locally triggered but trainable enough to fire in larger concerted
-% patterns (diSessa, 1993; Smith, diSessa, \& Roschelle, 1993). Informed
-% in part by agent-based accounts of cognition (Minsky, 1986) and complex
-% systems models (diSessa, 2002), the central tenets of an ``in-pieces''
-% approach hold that knowledge is emergent from interacting primitives,
-% rather than unitary and monolithic. An example from Smith, diSessa, and
-% Roschelle helps illustrate the point.
-%
-% The authors show that we might think of a rubber band as a different
-% conceptual entity depending on context. In several different
-% situations---wrapped around a newspaper, pulled taut as a string, spun
-% to store energy in a toy plane propeller---we intuitively think about
-% the rubber band's physical behavior differently: one as a negligible
-% part of the newspaper's point mass, another as a transverse pendulum
-% (likely) obeying Hooke's Law, and the third as a torsional spring. Those
-% differences in intuitive thinking reflect the contextual dependency of
-% what we know about the physical world:
-%
-% In each of the rubberband examples, various pieces of intuitive physical
-% knowledge describe the mechanism at work: the rubber band binds the
-% newspaper, grips the jar lid, and acts a source of springiness for the
-% bobbing object. Although a mapping cannot be made from the rubberband to
-% scientific entities, it is quite easy to map these qualitatively
-% distinct physical processes to scientific entities and laws. For
-% example, instances of binding almost always map to a practically rigid
-% body. Likewise, gripping maps to friction forces, and springiness maps
-% to Hooke's law. This suggests that applicability can depend directly on
-% our intuitive knowledge---knowledge that exists prior to any formal
-% scientific training (Smith et al., 1993, p. 144).
-%
-% The in-pieces approach to modeling cognition has been used, among other
-% things, to explain how experts reason about fractions and decimals
-% (Smith et al., 1993), how students reason about forces in physics
-% (diSessa \& Sherin, 1998; diSessa, 1993; Hammer, 1996; Sherin, 2001),
-% how students construct and evaluate algebraic representations of
-% physical situations (Izsák, 2004), and how knowledge transfers across
-% contexts (Hammer et al., 2005; Wagner, 2006). Because its starting
-% assumption is that knowledge is fragmented, knowledge-in-pieces can
-% account for wide variations of how people---particularly novices---use
-% knowledge on a moment-to-moment basis. In other words, because it
-% assumes knowledge is local, it can still explain the kinds of
-% globally-inconsistent ways people might reason about physical situations
-% (diSessa, 1993). As a framework, an in-pieces approach ultimately argues
-% that models of concept replacement and good/bad criteria for knowledge
-% should be supplanted by a learning model of alignment/refinement of
-% prior knowledge and the consideration of knowledge as
-% productive/unproductive.
-%
-% Hammer and colleagues have worked to extend the in-pieces approach to
-% explain how students' epistemological activity---how they orient toward
-% knowledge and knowing in a context (Hammer et al., 2005; Hammer \& Elby,
-% 2002, 2003). Specifically, those authors use two core theoretical
-% constructs to explain students' stances toward knowledge and knowing:
-%
-% \begin{itemize}
-% \item
-%   \emph{Epistemological Resources} are the epistemological equivalent of
-%   diSessa's phenomenological primitives (p-prims). Resources, the
-%   authors propose, are the atomic units involved in how people cognize
-%   about the source of knowledge, the nature of knowledge, and
-%   epistemological activities (Hammer \& Elby, 2003; Louca, Elby, Hammer,
-%   \& Kagey, 2004).
-% \item
-%   \emph{Epistemological Frames} are the emergent result of subsets of
-%   resources acting in concert. Drawing from both Goffman's (1974)
-%   sociological notion of frame as structures of expectations and
-%   subsequent work on framing in discourse (Tannen, 1993),
-%   epistemological frames are a participant's local answer to the
-%   question ``what is it {[}specifically, what knowledge activity{]}
-%   that's going on here'' (Goffman, 1974, p. 8).
-% \end{itemize}
-%
-% Resources can frames can interact in activity settings to produce
-% larger-scale patterns called ``epistemological coherences'' (Rosenberg,
-% Hammer, \& Phelan, 2006) where evidence from data suggests that a
-% network of discrete cognitive units can nonetheless give rise to stable
-% cognition.
-%
-% A useful conceptual metaphor for resources and framing is to think about
-% a lecture hall with different sets of lights: a spotlight in the back to
-% highlight a lecturer, chalkboard lights, house lights, and a projector.
-% Each light has an individual brightness, but the lights are only
-% controllable at the per-bulb level. In that sense, they are atomic. But,
-% light patterns interact with one another to create a field of lighting
-% for the room. Thus, lights are a bit like resources, and different light
-% configurations can correspond to different, sociologically-stable uses
-% of the room.
-%
-% \protect\hypertarget{ux5fToc252445958}{}{}Table -- Using room lighting
-% configurations to think about epistemological framing
-%
-% \begin{supertabular}[]{@{}lllll@{}}
-% \toprule
-% \textbf{What is it that's going on?} & Houselights & Spotlight &
-% Chalkboard & Projector\tabularnewline
-% \midrule
-% \endhead
-% \textbf{On-stage monologue} & Low & ON & Low & OFF\tabularnewline
-% \textbf{Presenting slides} & Low & ON & Low & ON\tabularnewline
-% \textbf{Working through equations} & Low & ON & High &
-% OFF\tabularnewline
-% \textbf{Students discussing with each other} & High & OFF & Low &
-% OFF\tabularnewline
-% \textbf{Watching a movie} & Low & OFF & Low & ON\tabularnewline
-% \textbf{Cleaning the room after a movie} & High & OFF & High &
-% OFF\tabularnewline
-% \bottomrule
-% \end{supertabular}
-%
-% To convince yourself of the sociological stability of these
-% configurations, imagine you were watching a movie when suddenly the
-% house and chalkboard lights came up to full intensity. It would jar you
-% out of the experience. You might wonder whether something was wrong: is
-% there an emergency? Should you evacuate?
-%
-% The metaphor isn't perfect. Ontologically, for example, resource and
-% framing theory propose these constructs exist not in the world but in
-% the minds of individuals. But, the metaphor is quite useful for
-% understanding how individual elements---in this case lights---can work
-% in concert to create and sustain a stable frame. And, crucially for a
-% cognitive system, the metaphor lets us account for variation in
-% activity. The same lecture hall can become a window onto one person's
-% thoughts (monologue), a site for instruction (working through equations
-% at the board), a shared space for collaboration (student discussion), or
-% a place in need of repair (cleaning up) simply by varying the
-% intensities of lighting banks in particular ways.
-%
-% An example helps ground this in-pieces approach to epistemology. In
-% Russ, Coffey, Hammer, and Hutchison (2008), the authors describe the
-% situation where an elementary student reasons about why an empty juice
-% box collapses when you suck on the straw. One student gives what the
-% authors deem to be an excellent mechanistic account of why the juice box
-% collapses. But, as the teacher seems to steer the discussion toward
-% vocabulary---in this case, ``pressure''---the student clearly pulls back
-% from her mechanistic reasoning, seems much more diffident, and claims
-% that pressure is hard to explain. That example highlights the disconnect
-% between doing science as knowing vocabulary and doing science as
-% reasoning mechanistically. Moreover, it strikingly highlights that a
-% student who by all accounts produced an excellent explanation of how
-% pressure works was left nonetheless with the impression that pressure
-% was hard to explain.\footnote{In fairness to the teacher, I'm trying to
-%   focus on the result of the interaction and far less on the intent the
-%   teacher might have had. The student still left with a sense that
-%   science might be about vocabulary, even if that's not the view the
-%   teacher would espouse or was trying to enact in the moment.}
-%
+
+The catch here is that the plus operator takes on different roles in
+different languages because of how those languages define its use.
+Statement 1 increments a number in C; Statement 2 appends the letters
+``ly'' to a string; Statement 3 adds a layer of points to a statistical
+graphics plot. These different kinds of operations become even more
+apparent and consequential when, for example, such statements are
+repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
+  operator can be written as a directional arrow. The symbol
+  \textless{}- (``less-than, hyphen'') indicates the value on the right
+  side of the symbol is being assigned to the variable on the left side
+  of the symbol. As used, the symbol itself is semantically equivalent
+  to having written an equals sign (``=``).} I'm using multiple
+reassignment statements to compose a statistical graphics plot. The
+layered creation of a plot invites a very different kind of conceptual
+interpretation than, say, repeatedly accumulating numbers into a running
+sum:
+
+p \textless{}- InitializeGgplot\_1w()
+
+p \textless{}- p + GrandMeanLine(owp)
+
+p \textless{}- p + GrandMeanPoint(owp)
+
+p \textless{}- p + ScaleX\_1w(owp)
+
+p \textless{}- p + ScaleY\_1w(owp)
+
+p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)
+
+One obvious reason for thinking about this code with a different
+interpretive frame is that numeric addition is commutative; composing a
+plot is not necessarily commutative.\footnote{To convince yourself that
+  plot composition isn't commutative, imagine a scale function that
+  squares up the aspect ratio of a plot and another scale function that
+  sets the aspect ratio to 1.5:1. Applying the square function last
+  produces a square plot; applying it first produces a rectangular plot.
+  String concatenation is also not necessarily commutative. ``cat'' +
+  ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
+  ``dogcat.''} So, despite the syntactic similarity, reassignments that
+compose a plot using reassignment (as above) does not obey the same
+rules as reassignments for a running total. But, even within this code
+block, statements that look alike perform operations of a different
+nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
+visual layer to a plot, others modify features of the plot (e.g, ``+
+Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
+the numeric range of the data).
+
+Given these motivating examples, it seems sensible to think there's
+utility in a programmer having different conceptual metaphors available
+to think about and work with code. Example 1 shows that choosing to
+apply knowledge from physics to a piece of code can change the cognitive
+nature of debugging. There, debugging a position-update statement
+becomes, in part, reasoning kinematically about the properties of motion
+trails.\footnote{For comparison, consider the argument that ringing an
+  aircraft speedometer with physical markers changes the nature of
+  cognition when pilots work to land a plane (Hutchins, 1995b).} Example
+2 shows that across languages, programmers might have to deploy
+different conceptual metaphors to reason about statements in a
+locally-consistent way. Knowing that plots in ggplot2 can be composed
+layer-by-layer with reassignment is crucial if you're trying to write or
+understand code that creates statistical graphics. But, I would argue
+that thinking about $\Box = \Box + \Box$ as ``compose new layer onto plot'' can and
+does appeal to different kinds of knowledge when compared to thinking
+about $\Box = \Box + \Box$ as ``include this addend in the sum,'' which itself can
+and does appeal to different kinds of concepts when compared to thinking
+about $\Box = \Box + \Box$ as ``increment the counter.''
+
+Stepping back, I can build the following argument
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Programming can be helped by applying conceptual models to code,
+  particularly when relevant domain-knowledge structures can
+  advantageously transform a problem (example 1)
+\item
+  But, conceptual models don't work all the time for all statements.
+  Because languages are designed differently, the same syntax can
+  actually correspond to very different operations in code (example 2).
+  And, that's true both within and across languages.
+\item
+  Consequently, it makes less sense to treat conceptual models as right
+  or wrong, and more sense to treat them as differentially advantageous
+  for thinking about what a piece of code does. (Thinking a ``+''
+  implies numerical addition isn't \emph{globally wrong} in JavaScript,
+  but it won't explain why 1 + ``1'' yields ``11'' as a result.)
+\item
+  It seems plausible that successful programmers, when reasoning about
+  or writing code, are able to dynamically access or deploy conceptual
+  models that are advantageous given the context (language, syntax,
+  surrounding code). Certainly such a supposition is in line with work
+  suggesting students have resources for thinking conceptually and
+  dynamically about how mathematics models real-world situations (Izsák,
+  2004; Sherin, 2001)
+\item
+  To model how programmers think with conceptual models, a suitable
+  framework should be able to account for the dynamic, context-sensitive
+  deployment of conceptual knowledge.
+\item
+  To model how programmers develop expertise, a suitable framework
+  should be able to describe higher-order phenomena. Such phenomena
+  include explaining how programmers come to have conceptual models or
+  generate new ones, why they decide to deploy them, and how programmers
+  consider which conceptual model (i.e., which way to think about code)
+  is appropriate.
+\end{enumerate}
+
+Taken together, these assertions propose criteria for how we might
+strive to model cognition in programming. Our modeling frameworks should
+be context-dependent, dynamic, and capable of explaining where
+conceptual models come from. They should also be able to account for
+phenomena that are not themselves conceptual, including what directs the
+use of certain kinds of conceptual knowledge. In the learning sciences,
+such models already exist and have proven useful and productive for
+thinking about thinking.
+
+\subsection{Manifold models of cognition explain context-dependence
+and the growth of
+expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}
+
+In 1993, a pair of articles in the learning sciences staked a strong
+claim for viewing knowledge as a network of pieces, isolated enough to
+be locally triggered but trainable enough to fire in larger concerted
+patterns (diSessa, 1993; Smith, diSessa, \& Roschelle, 1993). Informed
+in part by agent-based accounts of cognition (Minsky, 1986) and complex
+systems models (diSessa, 2002), the central tenets of an ``in-pieces''
+approach hold that knowledge is emergent from interacting primitives,
+rather than unitary and monolithic. An example from Smith, diSessa, and
+Roschelle helps illustrate the point.
+
+The authors show that we might think of a rubber band as a different
+conceptual entity depending on context. In several different
+situations---wrapped around a newspaper, pulled taut as a string, spun
+to store energy in a toy plane propeller---we intuitively think about
+the rubber band's physical behavior differently: one as a negligible
+part of the newspaper's point mass, another as a transverse pendulum
+(likely) obeying Hooke's Law, and the third as a torsional spring. Those
+differences in intuitive thinking reflect the contextual dependency of
+what we know about the physical world:
+
+In each of the rubberband examples, various pieces of intuitive physical
+knowledge describe the mechanism at work: the rubber band binds the
+newspaper, grips the jar lid, and acts a source of springiness for the
+bobbing object. Although a mapping cannot be made from the rubberband to
+scientific entities, it is quite easy to map these qualitatively
+distinct physical processes to scientific entities and laws. For
+example, instances of binding almost always map to a practically rigid
+body. Likewise, gripping maps to friction forces, and springiness maps
+to Hooke's law. This suggests that applicability can depend directly on
+our intuitive knowledge---knowledge that exists prior to any formal
+scientific training (Smith et al., 1993, p. 144).
+
+The in-pieces approach to modeling cognition has been used, among other
+things, to explain how experts reason about fractions and decimals
+(Smith et al., 1993), how students reason about forces in physics
+(diSessa \& Sherin, 1998; diSessa, 1993; Hammer, 1996; Sherin, 2001),
+how students construct and evaluate algebraic representations of
+physical situations (Izsák, 2004), and how knowledge transfers across
+contexts (Hammer et al., 2005; Wagner, 2006). Because its starting
+assumption is that knowledge is fragmented, knowledge-in-pieces can
+account for wide variations of how people---particularly novices---use
+knowledge on a moment-to-moment basis. In other words, because it
+assumes knowledge is local, it can still explain the kinds of
+globally-inconsistent ways people might reason about physical situations
+(diSessa, 1993). As a framework, an in-pieces approach ultimately argues
+that models of concept replacement and good/bad criteria for knowledge
+should be supplanted by a learning model of alignment/refinement of
+prior knowledge and the consideration of knowledge as
+productive/unproductive.
+
+Hammer and colleagues have worked to extend the in-pieces approach to
+explain how students' epistemological activity---how they orient toward
+knowledge and knowing in a context (Hammer et al., 2005; Hammer \& Elby,
+2002, 2003). Specifically, those authors use two core theoretical
+constructs to explain students' stances toward knowledge and knowing:
+
+\begin{itemize}
+\item
+  \emph{Epistemological Resources} are the epistemological equivalent of
+  diSessa's phenomenological primitives (p-prims). Resources, the
+  authors propose, are the atomic units involved in how people cognize
+  about the source of knowledge, the nature of knowledge, and
+  epistemological activities (Hammer \& Elby, 2003; Louca, Elby, Hammer,
+  \& Kagey, 2004).
+\item
+  \emph{Epistemological Frames} are the emergent result of subsets of
+  resources acting in concert. Drawing from both Goffman's (1974)
+  sociological notion of frame as structures of expectations and
+  subsequent work on framing in discourse (Tannen, 1993),
+  epistemological frames are a participant's local answer to the
+  question ``what is it {[}specifically, what knowledge activity{]}
+  that's going on here'' (Goffman, 1974, p. 8).
+\end{itemize}
+
+Resources can frames can interact in activity settings to produce
+larger-scale patterns called ``epistemological coherences'' (Rosenberg,
+Hammer, \& Phelan, 2006) where evidence from data suggests that a
+network of discrete cognitive units can nonetheless give rise to stable
+cognition.
+
+An example helps ground this in-pieces approach to epistemology. In
+Russ, Coffey, Hammer, and Hutchison (2008), the authors describe the
+situation where an elementary student reasons about why an empty juice
+box collapses when you suck on the straw. One student gives what the
+authors deem to be an excellent mechanistic account of why the juice box
+collapses. But, as the teacher seems to steer the discussion toward
+vocabulary---in this case, ``pressure''---the student clearly pulls back
+from her mechanistic reasoning, seems much more diffident, and claims
+that pressure is hard to explain. That example highlights the disconnect
+between doing science as knowing vocabulary and doing science as
+reasoning mechanistically. Moreover, it strikingly highlights that a
+student who by all accounts produced an excellent explanation of how
+pressure works was left nonetheless with the impression that pressure
+was hard to explain.\footnote{In fairness to the teacher, I'm trying to
+  focus on the result of the interaction and far less on the intent the
+  teacher might have had. The student still left with a sense that
+  science might be about vocabulary, even if that's not the view the
+  teacher would espouse or was trying to enact in the moment.}
+
+\subsection{Acknowledgments}\label{acknowledgments}
+
+Blinded for review.
+
+\clearpage
+
+\bibliographystyle{acm}
+\bibliography{bibliography}

### 2016-04-22 manuscript-body.tex
Fix references in itemized list of misconceptions

@@
 \item
-  Objects in object-oriented programming (Holland, Griffiths, \&
-  Woodman, 1997)
+  Objects in object-oriented programming \cite{holland_avoiding_1997}
 \item
-  Algorithms and data structures (Danielsiek, Paul, \& Vahrenhold, 2012;
-  Paul \& Vahrenhold, 2013)
+  Algorithms and data structures \cite{danielsiek_detecting_2012,paul_hunting_2013}
 \item
-  Programming statements in BASIC (Bayman \& Mayer, 1983)
+  Programming statements in BASIC \cite{bayman_diagnosis_1983}
 \item
-  Programming in Java (Fleury, 2000)
+  Programming in Java \cite{fleury_programming_2000}
 \item
-  Programming in Pascal (Fleury, 1993)
+  Programming in Pascal \cite{fleury_student_1993}
 \item
-  Parameter-passing (Fleury, 1991)
+  Parameter-passing \cite{fleury_parameter_1991}
 \item
-  Arrays in Java (Kaczmarczyk et al., 2010)
+  Arrays in Java \cite{kaczmarczyk_identifying_2010}
 \item
-  Objects in Java (Kaczmarczyk et al., 2010)
+  Objects in Java \cite{kaczmarczyk_identifying_2010}
 \item
-  Algorithms and computational complexity (Trakhtenbrot, 2013)
+  Algorithms and computational complexity \cite{trakhtenbrot_students_2013}
 \item
-  Boolean logic (Herman et al., 2008)
+  Boolean logic \cite{herman_proof_2008}
 \item
-  The efficiency of algorithms (Gal-Ezer \& Zur, 2004)
+  The efficiency of algorithms \cite{gal-ezer_efficiency_2004}
 \item
-  The Build-Heap algorithm (Seppälä, Malmi, \& Korhonen, 2006)
+  The Build-Heap algorithm \cite{seppala_observations_2006}
 \item
-  Hashtables (Patitsas, Craig, \& Easterbrook, 2013)
+  Hashtables \cite{patitsas_countably_2013}
 \item
-  The correctness of programs (Kolikant \& Mussai, 2008)
+  The correctness of programs \cite{kolikant_my_2008}
+\item
+  Polynomial and mapping reduction in the Theory and Computation of Complexity \cite{gal-ezer_identification_2016}
 \end{itemize}

### 2016-04-22 manuscript-body.tex
Add sandwiches example from Kaczmarczyk et al

@@
 
-Clancy (2004) provides a comprehensive overview of this line of
+Clancy \cite{clancy_misconceptions_2004} provides a comprehensive overview of this line of
 research, though in the past decade it has only grown. Indeed, roughly
@@
 
-The first example comes from Kaczmarczyk et al. (2010). Part of that
+The first example comes from \cite{kaczmarczyk_identifying_2010}. Part of that
 study involved giving students snippets of Java code and asking students
 to diagram (or pseudo-code) how the information would be stored in
-memory. Below I have reproduced the code for Problem 2 (Kaczmarczyk et
-al., 2010, p. 110):
+memory. Below I have reproduced the code for Problem 2:
 
-Cheese{[}{]} cheeses = new Cheese{[}4{]};
-
-Meat{[}{]} meats = new Meat{[}2{]};
-
-Turkey turkey;
-
-Ham ham;
-
-RoastBeef roastBeef;
-
-boolean lettuce = true;
-
-boolean tomato = true;
-
-SauceType sauceType = new SauceType();
-
-int numMeat;
-
-int numCheese;
+\hypertarget{sandiwches-code}{\label{sandwiches-code}}
+\begin{Shaded}
+\begin{Highlighting}[numbers=left,,]
+\NormalTok{Cheese[] cheeses = }\KeywordTok{new} \NormalTok{Cheese[}\DecValTok{4}\NormalTok{];}
+\NormalTok{Meat[] meats = }\KeywordTok{new} \NormalTok{Meat[}\DecValTok{2}\NormalTok{];}
+\NormalTok{Turkey turkey;}
+\NormalTok{Ham ham;}
+\NormalTok{RoastBeef roastBeef;}
+\DataTypeTok{boolean} \NormalTok{lettuce = }\KeywordTok{true}\NormalTok{;}
+\DataTypeTok{boolean} \NormalTok{tomato = }\KeywordTok{true}\NormalTok{;}
+\NormalTok{SauceType sauceType = }\KeywordTok{new} \FunctionTok{SauceType}\NormalTok{();}
+\DataTypeTok{int} \NormalTok{numMeat;}
+\DataTypeTok{int} \NormalTok{numCheese;}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
+\begin{quote}
   Student3 makes incorrect assumptions about connections between variables
@@
   there's two arrays, cheese and meats, uh, all those turkey and ham and
-roast beef are gonna be sorted into the meats array.'' (Kaczmarczyk et
-al., 2010, p. 110)
+  roast beef are gonna be sorted into the meats array.'' \cite{kaczmarczyk_identifying_2010}
+\end{quote}
 
-The researchers are correct in the sense that \textbf{turkey} and
-\textbf{ham} and \textbf{roastBeef} will \emph{not} be sorted into the
-\textbf{meats} array. First, there is no code here that places
-\textbf{turkey} and \textbf{ham} in the array; there is only code that
+The researchers are correct in the sense that \texttt{turkey} and
+\texttt{ham} and \texttt{roastBeef} will \emph{not} be sorted into the
+\texttt{meats} array. First, there is no code here that places
+\texttt{turkey} and \texttt{ham} in the array; there is only code that
 declares them as variables. Moreover, as written, an attempt to place
-\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} into the array
+\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} into the array
 would fail. Because of type restrictions in Java, only objects of class
 Meat (or objects that inherit from class Meat) can go in the array.
-\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} are, perhaps
-confusingly, references to object instances of classes \textbf{Turkey}
-and \textbf{Ham} and \textbf{RoastBeef}, so in the current snippet they
-cannot enter an array of type \textbf{Meat} because (1) they don't yet
+\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} are, perhaps
+confusingly, references to object instances of classes \texttt{Turkey}
+and \texttt{Ham} and \texttt{RoastBeef}, so in the current snippet they
+cannot enter an array of type \texttt{Meat} because (1) they don't yet
 exist as objects and (2) even if they did exist, their types don't match
@@
 samples, although such information was neither given nor implied''
-(Kaczmarczyk et al., 2010, p. 110).
+\cite{kaczmarczyk_identifying_2010}.
 

### 2016-04-22 manuscript-body.tex
Argue that a misconceptions focus is narrow

@@
 Again, the researchers are right that the student is failing to describe
-the code in a way consistent with canon.\footnote{In this case, canon is
-  the specifications and operations of the Java language and its
-  compilers.} But, in their non-canonical thinking Student3 evidences
-potentially productive insights about design. Precisely \emph{because}
+the code in a way consistent with canon. But, in their non-canonical thinking Student3 evidences
+potentially ``\emph{productive}'' \cite{hammer_form_2002,hammer_tapping_2003} insights about design. Precisely \emph{because}
 there is no code stating that \textbf{turkey} and \textbf{roastBeef} and
@@
 
-Student3 has an idea about a relationship between entities where that
-relationship is not specified in the code. Kaczmarczyk et al.
-(Kaczmarczyk et al., 2010) focus only on the downside of the idea: the
+Put another way, it's true Student3 has an idea about a relationship between entities that
+is not specified in the code. The study authors focus only on the \emph{downside} of the idea: the
 student fails to display a proper understanding of how arrays work in
-Java. Moreover, the student might be prone to similar mistakes of
-inferring information that does not actually exist in code. But, there
+Java. But, there
 is also an upside of this idea. Because Student3 is thinking about
@@
 My second example of research that criticizes students' non-canonical
-understandings comes from Bonar and Soloway's (1983) study of Pascal
-programmers, part of which is discussed in Pea (1986). A student in
-Bonar \& Soloway's study was asked to ``Write a program which reads in
+understandings comes from Bonar and Soloway's 1983 study of Pascal
+programmers \cite{bonar_uncovering_1983}, data from which is also analyzed and discussed in \cite{pea_languageindependent_1986}. A student in
+\cite{bonar_uncovering_1983} was asked to ``Write a program which reads in
 ten integers and prints the average of those integers'' (Bonar \&
@@
 
+\begin{verbatim}
 Repeat
-
 (1) Read a number (Num)
-
     (1a) Count := Count + 1
-
 (2) Add the number to Sum
-
     (2a) Sum := Sum + Num
-
 (3) until Count :=10
-
 (4) Average := Sum div Num
-
-(5) writeln (`average = `,Average)
+(5) writeln ('average = ',Average)
+\end{verbatim}
 
@@
 
+\begin{quote}
   Interviewer: Steps 1a and 2a: are those the same kinds of statements?
@@
   kinds of information that you are going to take out of the loop with
-you. (Bonar \& Soloway, 1983, p. 12)
+  you. \cite{bonar_uncovering_1983}
+\end{quote}
 
-Pea's (1986) interpretation? ``Here, again, we see the student believing
+Pea and Kurland's interpretation? ``Here, again, we see the student believing
 that the programming language knows more about her intentions than it
-possibly can'' (p. 32).
+possibly can'' \cite{pea_cognitive_1983}.
 
-As before, this student has an idea about relationships in code. Pea
-(1986) see the downside of her idea: believing PASCAL can understand
+As in Example 1, this student has an idea about relationships in code. Pea and Kurland
+\cite{pea_cognitive_1983} see the \emph{downside} of her idea: believing PASCAL can understand
 shades of programmer intent when, in fact, it cannot. And again, that
@@
 In defense of the student, the question --- as asked --- is vague. Are
-those statements the same \emph{to whom and in what way?} Pea (1986)
-treats the data as though she meant ``the same to PASCAL.'' Indeed,
+those statements the same \emph{to whom} and \emph{in what way?} Pea and Kurland
+treat the data as though she meant ``the same to PASCAL.'' Indeed,
 maybe she did, in which case his interpretation has traction. But,
-another interpretation is that she meant to herself, or to someone else
+another interpretation is that she meant to herself or to someone else
 reading the code. Those statements might not be the same \emph{to her}
@@
 These two purposes, which for the sake of description I'll call
-\emph{keeping control} and \emph{totaling up} are, in a sense,
+\emph{keeping control} (1a) and \emph{totaling up} (2a) are, in a sense,
 different. The PASCAL compiler (and runtime) does not differentiate
@@
 
+\begin{quote}
   The structural aspects of a variable in a computer language are given
@@
   times a variable might function as ``a counter,'' ``data,'' or
-``input.'' (diSessa, 1986, p. 202)
+  ``input.'' \cite{disessa_models_1986}
+\end{quote}
 
-The student in Bonar \& Soloway's (1983) study did not show evidence of
+The student in \cite{bonar_uncovering_1983} did not show evidence of
 understanding the structural similarities between (1a) and (2a) in her
-pseudo-code. And, those authors as well as Pea (1986) justly insist that
+pseudo-code. And, those authors as well as others \cite{pea_cognitive_1983} justly insist that
 similarity is important for students to understand. From a conceptual
@@
 
-The student did show evidence of understanding a functional difference
-between (1a) and (2a), but Pea (1986) does not remark on that kind of
-understanding at all.\footnote{Also glossed over is, to me, another
+The student did, however, show evidence of understanding a \emph{functional} \cite{disessa_models_1986} difference
+between (1a) and (2a), but Pea and Kurland do not remark on that kind of
+understanding at all. Both \cite{pea_cognitive_1983,bonar_uncovering_1983} also gloss over another
 important difference: a programmer might not know in advance which
@@
 change as the loop iterates. In contrast, the programmer knows exactly
-  how the value of Count will change with each loop iteration.} Again, I
+how the value of Count will change with each loop iteration. Again, we
 claim this oversight is part of a subtle but observable trend in
-programming misconceptions literature. While, or perhaps because
+programming misconceptions literature.
+
+While, or perhaps because
 research has been so preoccupied warring with students' problematic
 knowledge, it has sometimes failed to recover the productive knowledge
-(or resources for building it) students have. In this example, the
+(or resources for building it) students have. In Example 2, the
 student already has a grasp that syntactically similar statements could
-serve different conceptual purposes. Hypothetically, the student might
-use that information in design by calling up the $\Box=︎ \Box + 1$ syntactical
-template when the situation seems to demand \emph{keeping control},
-while calling up $\Box= ︎\Box + \mathrm{number}$ when \emph{totaling up} is the goal.
-And, the idea that structurally identical symbol templates can serve
-different functional and conceptual purposes fits precisely in line with
-Sherin's (2001) theory of symbolic forms. For example, the symbolic
+serve different conceptual purposes. And, that understanding could, in turn, shape how they program. The student might
+invoke the $\Box=︎ \Box + 1$ syntactical
+template (Or what Sherin would call ``symbol template'' \cite{sherin_how_2001}) when the situation seems to demand \emph{keeping control},
+while invoking $\Box= ︎\Box + \mathrm{number}$ when \emph{totaling up} is the goal.
+And, the idea that structurally similar symbol templates can serve
+different functional and conceptual purposes fits precisely in line with both diSessa's distinction of structural/functional understanding of a programming language \cite{disessa_models_1986} and
+Sherin's theory of symbolic forms \cite{sherin_how_2001}.
+
+An example drives home the connection between \cite{disessa_models_1986} and \cite{sherin_how_2001} as it applies to programming. The symbolic
 forms \emph{parts-of-a-whole} and \emph{base+change} have different
 conceptual schemata. Parts-of-a-whole refers to the contributions of
-component entities while base±change describes a kind of accumulation.
-Specifically, the terms in base±change ``play different roles'' (Sherin,
-2001, p. 534) But, the two distinct conceptual schemata share what I
-would argue is the same concrete symbol template of how to write an
-expression: $\Box=︎ \Box + \Box$.
+component entities while base±change describes a kind of accumulation \cite{sherin_how_2001}.
+Specifically, the terms in base±change ``contribute to a whole but play different roles.
+One is a base value; the other is a change to that base.'' But, the two distinct conceptual schemata share what I
+would argue is the same symbolic structure: $\Box=︎ \Box + \Box$ (parts of a whole) and $\Box = \Box \pm \Delta$ (base + change) \cite{sherin_how_2001}.
 
@@
 (2001) catalog of conceptual schemata. In other words, I would argue
-there are conceptual ways a programmer may need to think about
-interpreting or writing a $\Box=︎ \Box +\Box$ symbol template that Sherin doesn't
-enumerate.\footnote{One of the most obvious is the movement from seeing
-  $\Box=︎ \Box +\Box$ as a statement of equality to seeing it as the storage of a
-  sum to a variable.} To follow that implication, the canonical body of
+there are conceptual ways a \emph{programmer} may need to think about
+interpreting or writing a $\Box=︎ \Box +\Box$ symbol template that even Sherin \cite{sherin_how_2001} doesn't
+enumerate.One of the most obvious, for example, is the conceptual movement from seeing
+  $\Box=︎ \Box +\Box$ as a statement of equality (typical in a math class) to seeing it as the assignment of a
+  sum to a variable (typical in programming). To follow that implication, the canonical body of
 knowledge about which programmers must reason is itself fractured,
@@
 
-To return to misconceptions, what drives research on students'
-misconceptions is largely a need to get students to program computers
+To return to misconceptions, what seems to drive research on students'
+misconceptions is computing education largely a need to get students to program computers
 and reason about computation in ways that are canonically correct. And,
-those are assuredly worthwhile goals. But, as I've argued, we can
+that's a worthwhile goals. But, as I've argued, we can
 already identify cases where a narrow misconceptions focus is silent
-about or dismissive of students' useful intuitions. We can also identify
-the further problems whereby unilateral emphasis on one language's canon
+about (at best) or dismissive of (at worst) students' useful intuitions. We identified
+the further problems when a unilateral emphasis on one language's canon
 opposes the vocabulary of symbolic forms (and diverse conceptual
-schemata) expert programmers ultimately need (Sherin, 2001) Taken in
+schemata) expert programmers ultimately need. \emph{Taken in
 total, that silence, dismissiveness, and narrow view of refining
 knowledge perpetuates a deficit-focused discourse about student's
-knowledge in computing.\footnote{Worse still, in my view, is that there
-  is a pattern of researchers failing to discuss or acknowledge
-  students' productive knowledge \emph{even when their own data supports
-  it}.} Such a perspective also fails to address what aspects of
+knowledge in computing.} A knowledge-deficiency perspective also fails to address what aspects of
 students reasoning might get broken by attempts to ``fix'' such
@@
 It's important to note that misconceptions research in computing didn't
-always treat students' non-canonical knowledge this way. I begin the
+always treat students' non-canonical knowledge as a problem. I begin the
 next section by backtracing to some of the earliest work on students'

### 2016-04-22 manuscript-body.tex
Continue comparing Pea & Kurland with Bonar & Soloway

@@
 
-What's curious about Pea's (1986) comments on Bonar \& Soloway (1983) is
-that Pea drew different conclusions from the data than Bonar and Soloway
-did. Pea (1986) emphasized that when the student thought two
+What's curious about Pea and Kurland's \cite{pea_languageindependent_1986} comments on Bonar and Solowa \cite{bonar_uncovering_1983} is
+that \emph{Pea \cite{pea_languageindependent_1986} drew different conclusions from the data than did Bonar and Soloway
+}. Pea emphasized that when the student thought two
 semantically-equivalent assignment statements in PASCAL were different,
@@
 \emph{more} of their meaning for what they want to accomplish in the
-program than is actually present in the code they have written'' (p.
-30). In other words, Pea treated the data as a fairly clear example of a
-class of bugs. Specifically, he saw a bug class in which students simply
-assumed the interpreter or runtime could infer the code author's shades
+program than is actually present in the code they have written'' \cite{pea_languageindependent_1986}. In other words, Pea treated the data as a fairly clear example of a
+class of cognitive bugs. Specifically, he saw a bug class in which students simply
+assumed the interpreter or runtime could infer programmer's shades
 of intent.
@@
 Bonar and Soloway (1983), by contrast, were less quick to make
-inferences either about the nature of the bug or about the intervention
+inferences either about the nature of the bug or the intervention
 it entailed. Rather than jump to definitive conclusions, they were
@@
 
+\begin{quote}
   It is not clear exactly how to react to the bugs we have uncovered in
@@
   novice bugs. Only once a bug is uncovered and understood are we ready to
-create a remedy for that bug. (Bonar \& Soloway, 1983, p. 12)
+  create a remedy for that bug. \cite{bonar_uncovering_1983}
+\end{quote}
 
-That the authors would even consider developing new constructs or
-languages is a noteworthy distinction. Rather than assume in toto that
+That Bonar and Soloway would even consider developing new constructs or
+languages is a noteworthy distinction. Rather than assume wholesale that
 students with ``bugs'' had wrong knowledge, the authors instead suppose
@@
 
+\begin{quote}
   We find it quite interesting that novices seem to understand the role or
@@
   programming systems should support recording the roles the programmer
-intends for various statements and variables. (Bonar \& Soloway, 1983,
-p. 12)
+  intends for various statements and variables. \cite{bonar_uncovering_1983}
+\end{quote}
 
-Again, what's noteworthy here is that rather than treating students'
+Again, what's noteworthy is that rather than treating students'
 non-canonical views as a burden for instruction, Bonar and Soloway
@@
 
+\begin{quote}
   The experience and understanding of a novice are available for analysis.
@@
   \& Soloway, 1983, p. 13)
+\end{quote}
 
-I assume Bonar and Soloway structured their final line deliberately. If
+We assume Bonar and Soloway structured the final line of that quote deliberately. If
 so, their phrasing has three consequences:
@@
 \def\labelenumi{\arabic{enumi}.}
+\tightlist
 \item

### 2016-04-22 manuscript-body.tex
Rewrite section and subsection headings

@@
+\section{Introduction}\label{Introduction}
+
 \subsection{Misconceptions research in computing education tends to
@@
 
+\section{Core Arguments}\label{core-arguments}
+
 \subsection{Not all cognitive programming bugs imply a problem with
@@
 
-What's curious about Pea and Kurland's \cite{pea_languageindependent_1986} comments on Bonar and Solowa \cite{bonar_uncovering_1983} is
-that \emph{Pea \cite{pea_languageindependent_1986} drew different conclusions from the data than did Bonar and Soloway
-}. Pea emphasized that when the student thought two
+What's curious about Pea and Kurland's \cite{pea_languageindependent_1986} comments on Bonar and Soloway \cite{bonar_uncovering_1983} is
+that Pea's conclusions \cite{pea_languageindependent_1986} about Bonar and Soloway's data differed greatly from Bonar and Soloway's conclusions. Pea emphasized that when the student thought two
 semantically-equivalent assignment statements in PASCAL were different,
@@
 \item
-  Changing the models novices have comes next, and ``when necessary.''
+  Changing the models novices have comes next, and explicitly ``when necessary.''
 \item
   They speak of ``how to move a novice to some more appropriate model,''
-  which does not necessarily entail the removal or destruction of
-  novice's existing models.
+  which does not necessarily entail rejecting a student's existing model, treating it as a misconception to be corrected, or otherwise ignoring what productive elements are present in a given novice model.
 \end{enumerate}
 
-Taken together, these points convey a sense of how Bonar and Soloway
-view learning and instruction in computing. Instruction explicitly
+We believe that taken together, these points convey a sense of how Bonar and Soloway \cite{bonar_uncovering_1983}
+viewed learning and instruction in computing. Instruction explicitly
 includes appeals to prior models and knowledge students might already
 have. Learning, meanwhile, involves the movement \emph{when necessary}
-to more appropriate models of computation. As I explain in the next
-section, such a view exactly aligns with a particular branch of
+to more appropriate models of computation. As we explain in the next
+section, such a view of understanding and refining student ideas---in contrast to diagnosing the deficiencies in and replacement of student ideas---exactly aligns with a particular branch of
 constructivism, where cognition is viewed as the complex activation of

### 2016-04-22 manuscript-body.tex
Replace "I" with "we" for joint authorship

@@
 
-I argue research
+we argue research
 approaches that inflexibly privilege canonical knowledge do so at the
@@
 makes sense, but doing so at the expense of---or in direct contradiction
-to---useful ways of knowing seems undesirable at best. Next, I expand on
+to---useful ways of knowing seems undesirable at best. Next, we expand on
 two examples from misconceptions research in programming. Specifically,
-I show how and why I think a misconceptions focus in programming casts
+we show how and why we think a misconceptions focus in programming casts
 aside students' useful intuitions and understandings.
@@
 to diagram (or pseudo-code) how the information would be stored in
-memory. Below I have reproduced the code for Problem 2:
+memory. Below we have reproduced the code for Problem 2:
 
@@
 interpret and write code in languages that may demand incommensurable,
-or at least distinct, conceptual schemata. As I show later in Table 4,
+or at least distinct, conceptual schemata. As we show later in Table 4,
 three current programming languages make remarkably different use of the
@@
 make sense of how + works in those languages don't exist in Sherin's
-(2001) catalog of conceptual schemata. In other words, I would argue
+(2001) catalog of conceptual schemata. In other words, we would argue
 there are conceptual ways a \emph{programmer} may need to think about
@@
 It's important to note that misconceptions research in computing didn't
-always treat students' non-canonical knowledge as a problem. I begin the
+always treat students' non-canonical knowledge as a problem. we begin the
 next section by backtracing to some of the earliest work on students'
@@
 (1983), where an assignment statement worked to increment a value and
-store the result back to that value. In Table 2 below, I imagine five
+store the result back to that value. In Table 2 below, we imagine five
 different ways of writing a programming statement. Each of the five ways
 is semantically equivalent to the others (or near enough for explanatory
-purposes). And, in each example I add a layer of specificity to the
-syntax. I also propose a corresponding interpretation of how I might
-apply interpret and think about a given statement.\footnote{I assume for
+purposes). And, in each example we add a layer of specificity to the
+syntax. we also propose a corresponding interpretation of how we might
+apply interpret and think about a given statement.\footnote{we assume for
   simplicity's sake that all variables and functions in my code samples
   evaluate to floating-point numbers that can be combined under
-  addition. I also assume, crucially, that all statements are in the
+  addition. we also assume, crucially, that all statements are in the
   same hypothetical language and that the variable names connote nothing
@@
 
-I stress that the third column is about \emph{how I might think about}
-each programming statement\emph{.} By no means am I making normative
+we stress that the third column is about \emph{how we might think about}
+each programming statement\emph{.} By no means am we making normative
 claims about how one \emph{ought} to think about it, or whether the
 statement actually does what its names might suggest it does. Rather,
-``How I might think about it'' reflects the kind of local meaning or
-interpretation I might attach to such a statement when I work with it,
+``How we might think about it'' reflects the kind of local meaning or
+interpretation we might attach to such a statement when we work with it,
 given my understanding of its role and context.
@@
 Consider specifically statements 1 and 4. Statement 1 could very well be
-an incrementer in some kind of iterative code. If I had to debug code
-that employs statement 1, what I might do is exploit my knowledge of
+an incrementer in some kind of iterative code. If we had to debug code
+that employs statement 1, what we might do is exploit my knowledge of
 what d is (does it hold a constant value? Is it 1?) and try inserting
-intermediate print statements into the iterative code. Doing so, I can
+intermediate print statements into the iterative code. Doing so, we can
 inspect textually how values change with each iteration. In statement
@@
 images on a screen. If there's a problem with that code, one of the
-easiest ways I might notice is that the acceleration seems off in the
-graphics. Consequently, my debugging might call upon knowledge I have
-from kinematics. I might try inserting code that draws a dot at the
-object's on-screen position with each iteration, leaving a trail I can
-visually inspect. I could then look at the path of the object's
+easiest ways we might notice is that the acceleration seems off in the
+graphics. Consequently, my debugging might call upon knowledge we have
+from kinematics. we might try inserting code that draws a dot at the
+object's on-screen position with each iteration, leaving a trail we can
+visually inspect. we could then look at the path of the object's
 trajectory and the spacing patterns between successive dots as a
-first-pass test of whether my code achieves the motion I want.
+first-pass test of whether my code achieves the motion we want.
 
@@
 \emph{use conceptual knowledge from physics} to diagnose and fix
-problems in my code. If I view the assignment statement as saying
+problems in my code. If we view the assignment statement as saying
 something about the motion of an object (cf., Hammer, 1994, p. 165), a
 field of knowledge and concomitant techniques from physics becomes
-available to me to think with. But, if I focus only on the
+available to me to think with. But, if we focus only on the
 semantic-level equivalence of statements 1 through 5, there would be no
-obvious reason for me to access what I know about physics in order to
+obvious reason for me to access what we know about physics in order to
 reason about the code.
@@
 My second example concerns statements that are syntactically-similar,
-rather than semantically-similar. I proposed that the statements in
+rather than semantically-similar. we proposed that the statements in
 Table 2 all came from the same language. But, another phenomenon comes
@@
 % \toprule
-% \textbf{Statement } & Programming Syntax & Language & How I think about
+% \textbf{Statement } & Programming Syntax & Language & How we think about
 % it\tabularnewline
@@
 layer-by-layer with reassignment is crucial if you're trying to write or
-understand code that creates statistical graphics. But, I would argue
+understand code that creates statistical graphics. But, we would argue
 that thinking about $\Box = \Box + \Box$ as ``compose new layer onto plot'' can and
@@
 
-Stepping back, I can build the following argument
+Stepping back, we can build the following argument
 

### 2016-04-22 manuscript-body.tex
Drive home points about structural and syntactic similarity
Add example statements in javascript

@@
 research to learn from students}. Precisely because students saw
-functional differences (cf., diSessa, 1986) in semantically-equivalent
-PASCAL statements, Bonar and Soloway (1983) reflect that perhaps
+functional differences in semantically-equivalent
+PASCAL statements, Bonar and Soloway \cite{bonar_uncovering_1983} reflect that perhaps
 programming languages should be more expressive:
@@
 
-\subsection{Examples motivate the need for contextual-sensitivity in
-modeling programming
-cognition}\label{examples-motivate-the-need-for-contextual-sensitivity-in-modeling-programming-cognition}
+\subsection{Examples motivate the need for contextual-sensitivity in modeling programming cognition}\label{examples-motivate-the-need-for-contextual-sensitivity-in-modeling-programming-cognition}
 
-Let's begin with two motivating examples. My aim with these examples is
+Let's begin with two motivating examples. Our aim with these examples is
 to show how a practicing programmer might employ specific, distinct
@@
 (1983), where an assignment statement worked to increment a value and
-store the result back to that value. In Table 2 below, we imagine five
-different ways of writing a programming statement. Each of the five ways
-is semantically equivalent to the others (or near enough for explanatory
-purposes). And, in each example we add a layer of specificity to the
-syntax. we also propose a corresponding interpretation of how we might
-apply interpret and think about a given statement.\footnote{we assume for
-  simplicity's sake that all variables and functions in my code samples
-  evaluate to floating-point numbers that can be combined under
-  addition. we also assume, crucially, that all statements are in the
-  same hypothetical language and that the variable names connote nothing
-  to the computational interpreter.}
+store the result back to that value. Below are JavaScript statements that, as programmers, we have actually written in code as part of our careers. Each statement
+adheres to the structural similarity in Bonar and Soloway's PASCAL example \cite{bonar_uncovering_1983} discussed earlier.
+But, the comment above each statement reflects the \emph{functional understandings} \cite{disessa_models_1986} we made use of to write, interpret, and debug the code we were working on.
 
-\protect\hypertarget{ux5fToc252445956}{}{}Table -- Five different
-interpretations of semantically equivalent programming statements in the
-same language
+\begin{Shaded}
+\begin{Highlighting}[numbers=left,,]
+\CommentTok{// Increment a counter}
+\NormalTok{x }\OperatorTok{=} \NormalTok{x }\OperatorTok{+} \DecValTok{1}
 
-\begin{center}
-  \begin{tabular}{ | l | l |}
-    1 & \texttt{x\ =\ x\ +\ d} \\
-    2 & \texttt{x\ =\ x\ +\ update} \\
-    3 & \texttt{x\_position\ =\ x\_position\ +\ update} \\
-    4 & \texttt{x\_position\ =\ x\_position\ +\ update(t)} \\
-    5 & \texttt{x\_position\ =\ x\_position\ +\ update(...)} \\
-  \end{tabular}
-\end{center}
+\CommentTok{// Advance a simulation forward}
+\NormalTok{t }\OperatorTok{=} \NormalTok{t }\OperatorTok{+} \NormalTok{dt}
 
-we stress that the third column is about \emph{how we might think about}
-each programming statement\emph{.} By no means am we making normative
+\CommentTok{// Move an object under uniform velocity}
+\NormalTok{x_position }\OperatorTok{=} \NormalTok{x_position }\OperatorTok{+} \NormalTok{displacement}
+
+\CommentTok{// Accelerate an object along the x-dimension}
+\NormalTok{x_position }\OperatorTok{=} \NormalTok{x_position }\OperatorTok{+} \AttributeTok{displacement}\NormalTok{(t)}
+
+\CommentTok{// Jitter a plot point using a stochastic function}
+\NormalTok{y_position }\OperatorTok{=} \NormalTok{y_position }\OperatorTok{+} \AttributeTok{random_noise}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+We stress the comments reflect \emph{how one might think about}
+each programming statement. By no means are we making normative
 claims about how one \emph{ought} to think about it, or whether the
-statement actually does what its names might suggest it does. Rather,
+statement actually does what its variable names might suggest it does. Rather,
 ``How we might think about it'' reflects the kind of local meaning or
 interpretation we might attach to such a statement when we work with it,
-given my understanding of its role and context.
+given our understanding of its role and context.
 
-Consider specifically statements 1 and 4. Statement 1 could very well be
-an incrementer in some kind of iterative code. If we had to debug code
-that employs statement 1, what we might do is exploit my knowledge of
-what d is (does it hold a constant value? Is it 1?) and try inserting
-intermediate print statements into the iterative code. Doing so, we can
-inspect textually how values change with each iteration. In statement
-4's case, it could very well be that the position-update code animates
+Consider specifically lines 2 and 11. Line 2 could very well be
+an incrementer in some kind of iterative code. If we had to debug loop code
+that employs line 2, what we might do is try inserting
+intermediate print statements to see the value of x.
+We might also call on our knowledge that it's being incremented by 1 to check whether the difference between any two successive printouts is 1; if it isn't, we immediately know something's wrong.
+In line
+11's case, it could very well be that line 11's function is animating
 images on a screen. If there's a problem with that code, one of the
 easiest ways we might notice is that the acceleration seems off in the
-graphics. Consequently, my debugging might call upon knowledge we have
-from kinematics. we might try inserting code that draws a dot at the
-object's on-screen position with each iteration, leaving a trail we can
-visually inspect. we could then look at the path of the object's
+animation. Consequently, debugging might call upon knowledge we have
+from kinematics. We might try inserting code that draws a dot at the
+object's on-screen position with each iteration, leaving a trail of dots we can
+visually inspect. We could then look at the path of the object's
 trajectory and the spacing patterns between successive dots as a
-first-pass test of whether my code achieves the motion we want.
+first-pass test of whether the code achieves the motion we want.
+(If, for example, we're attempting a cubic easing animation, but the dots are all evenly spaced, then we might look first at the \texttt{displacement(t)}) function and what it's returning.
 
-To be clear, it's not just that graphics might improve my efficiency in
-debugging a statement. Rather, applying an interpretive frame that
-treats an assignment statement as a kinematic position update lets me
+To be clear, it's not just that graphics might improve our efficiency in
+debugging a statement like line 11. Rather, applying an interpretive frame that
+treats an assignment statement as a kinematic position update lets us
 \emph{use conceptual knowledge from physics} to diagnose and fix
-problems in my code. If we view the assignment statement as saying
-something about the motion of an object (cf., Hammer, 1994, p. 165), a
+problems in code. If we view the assignment statement as proposing
+something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves[blinded]---a
 field of knowledge and concomitant techniques from physics becomes
@@
 
-My second example concerns statements that are syntactically-similar,
-rather than semantically-similar. we proposed that the statements in
-Table 2 all came from the same language. But, another phenomenon comes
-into play when different languages use the same symbology for
-conceptually different operations. Table 3 shows examples of what look
+Our second example concerns statements that use the same symbol template for fundamentally different kinds of computation.
+In our JavaScript example statements, all code came from the same language.
+But, another phenomenon comes
+into play \emph{across} different languages when use the same symbology (written patterns) to stand for
+conceptually different operations. The statements below are examples of what look
 like semantically-equivalent operations, but in fact are not.
 
-\protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
-syntactically-similar statements with very different semantics
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{// C - Increments the value of i by 1}
+\NormalTok{i = i + }\DecValTok{1}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{// JavaScript - Appends "ly" to word.toString()}
+\NormalTok{word }\OperatorTok{=} \NormalTok{word }\OperatorTok{+} \StringTok{"ly"}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# R/ggplot2 - Composes a layer of points onto a plot}
+\NormalTok{p =}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
 
@@
 Statement 1 increments a number in C; Statement 2 appends the letters
-``ly'' to a string; Statement 3 adds a layer of points to a statistical
+``ly'' to a string; Statement 3 composes a layer of points onto a statistical
 graphics plot. These different kinds of operations become even more
 apparent and consequential when, for example, such statements are
-repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
-  operator can be written as a directional arrow. The symbol
-  \textless{}- (``less-than, hyphen'') indicates the value on the right
-  side of the symbol is being assigned to the variable on the left side
-  of the symbol. As used, the symbol itself is semantically equivalent
-  to having written an equals sign (``=``).} I'm using multiple
-reassignment statements to compose a statistical graphics plot. The
+repeated in the same language within the same file. In the R/ggplot2 code below, the author uses multiple
+reassignment statements to compose a statistical graphics plot.
+
+\begin{Shaded}
+\begin{Highlighting}[numbers=left,,]
+\NormalTok{p <-}\StringTok{ }\KeywordTok{InitializeGgplot_1w}\NormalTok{()}
+\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanLine}\NormalTok{(owp)}
+\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanPoint}\NormalTok{(owp)}
+\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleX_1w}\NormalTok{(owp)}
+\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleY_1w}\NormalTok{(owp)}
+\end{Highlighting}
+\end{Shaded}
+
+Despite the syntactic similarity, The
 layered creation of a plot invites a very different kind of conceptual
 interpretation than, say, repeatedly accumulating numbers into a running
-sum:
-
-p \textless{}- InitializeGgplot\_1w()
-
-p \textless{}- p + GrandMeanLine(owp)
-
-p \textless{}- p + GrandMeanPoint(owp)
-
-p \textless{}- p + ScaleX\_1w(owp)
-
-p \textless{}- p + ScaleY\_1w(owp)
-
-p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)
-
-One obvious reason for thinking about this code with a different
-interpretive frame is that numeric addition is commutative; composing a
-plot is not necessarily commutative.\footnote{To convince yourself that
-  plot composition isn't commutative, imagine a scale function that
-  squares up the aspect ratio of a plot and another scale function that
-  sets the aspect ratio to 1.5:1. Applying the square function last
-  produces a square plot; applying it first produces a rectangular plot.
-  String concatenation is also not necessarily commutative. ``cat'' +
-  ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
-  ``dogcat.''} So, despite the syntactic similarity, reassignments that
-compose a plot using reassignment (as above) does not obey the same
-rules as reassignments for a running total. But, even within this code
+sum. One obvious reason for thinking about this code with a different
+interpretive frame than incrementation is that numeric addition is commutative; composing a
+plot is not necessarily commutative. So, despite the syntactic similarity, reassignments that
+compose a plot using reassignment (as above) do not obey the same
+rules as reassignments for a running total. But, it turns out the statements above don't obey the same rules of string concatenation either. Within this code
 block, statements that look alike perform operations of a different
-nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
-visual layer to a plot, others modify features of the plot (e.g, ``+
-Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
-the numeric range of the data).
+nature. While some expressions (lines 2 and 3) compose
+visual layers onto a plot, others (lines 4 and 5) modify features of the plot, like the x and y scales.
 
@@
   models that are advantageous given the context (language, syntax,
-  surrounding code). Certainly such a supposition is in line with work
-  suggesting students have resources for thinking conceptually and
-  dynamically about how mathematics models real-world situations (Izsák,
-  2004; Sherin, 2001)
+  surrounding code).
+Certainly prior research demonstrates novices and experts both have resources for creating, evaluating, and adapting their own conceptual metaphors to suit the context of the problem \cite{hammer_static_2011,izsak_we_2003,izsak_students_2004}
 \item

### 2016-04-22 example-statements.js

@@ -1,0 +1,14 @@
+// Increment a counter
+x = x + 1
+
+// Advance a simulation forward
+t = t + dt
+
+// Move an object under uniform velocity
+x_position = x_position + displacement
+
+// Accelerate an object along the x-dimension
+x_position = x_position + displacement(t)
+
+// Jitter a plot point using a stochastic function
+y_position = y_position + random_noise()

### 2016-04-22 manuscript-body.tex
Add citations for knowledge-in-pieces and framing

@@
 be locally triggered but trainable enough to fire in larger concerted
-patterns (diSessa, 1993; Smith, diSessa, \& Roschelle, 1993). Informed
-in part by agent-based accounts of cognition (Minsky, 1986) and complex
-systems models (diSessa, 2002), the central tenets of an ``in-pieces''
+patterns \cite{disessa_epistemology_1993,smith_misconceptions_1993}. Informed
+in part by agent-based accounts of cognition \cite{minsky_society_1986} and complex
+systems models \cite{disessa_why_2002}, the central tenets of an ``in-pieces''
 approach hold that knowledge is emergent from interacting primitives,
@@
 
+\begin{quote}
   In each of the rubberband examples, various pieces of intuitive physical
@@
   our intuitive knowledge---knowledge that exists prior to any formal
-scientific training (Smith et al., 1993, p. 144).
+  scientific training \cite{smith_misconceptions_1993}.
+\end{quote}
 
 The in-pieces approach to modeling cognition has been used, among other
-things, to explain how experts reason about fractions and decimals
-(Smith et al., 1993), how students reason about forces in physics
-(diSessa \& Sherin, 1998; diSessa, 1993; Hammer, 1996; Sherin, 2001),
+things, to explain how experts reason about fractions and decimals \cite{smith_misconceptions_1993}, how students reason about forces in physics \cite{disessa_what_1998,disessa_epistemology_1993,hammer_misconceptions_1996,sherin_how_2001},
 how students construct and evaluate algebraic representations of
-physical situations (Izsák, 2004), and how knowledge transfers across
-contexts (Hammer et al., 2005; Wagner, 2006). Because its starting
+physical situations \cite{izsak_we_2003,izsak_students_2004}, and how knowledge transfers across
+contexts \cite{hammer_resources_2005,wagner_transfer_2006}. Because its starting
 assumption is that knowledge is fragmented, knowledge-in-pieces can
@@
 globally-inconsistent ways people might reason about physical situations
-(diSessa, 1993). As a framework, an in-pieces approach ultimately argues
+\cite{disessa_epistemology_1993}. As a framework, an in-pieces approach ultimately argues
 that models of concept replacement and good/bad criteria for knowledge
@@
 explain how students' epistemological activity---how they orient toward
-knowledge and knowing in a context (Hammer et al., 2005; Hammer \& Elby,
-2002, 2003). Specifically, those authors use two core theoretical
+knowledge and knowing in a context \cite{hammer_resources_2005,hammer_form_2002,elby_substance_2001}. Specifically, those authors use two core theoretical
 constructs to explain students' stances toward knowledge and knowing:
@@
 \item
-  \emph{Epistemological Resources} are the epistemological equivalent of
-  diSessa's phenomenological primitives (p-prims). Resources, the
+  \emph{Epistemological Resources} \cite{hammer_resources_2005} are the epistemological equivalent of
+  diSessa's phenomenological primitives (p-prims) \cite{disessa_epistemology_1993}. Resources, the
   authors propose, are the atomic units involved in how people cognize
   about the source of knowledge, the nature of knowledge, and
-  epistemological activities (Hammer \& Elby, 2003; Louca, Elby, Hammer,
-  \& Kagey, 2004).
+  epistemological activities \cite{hammer_tapping_2003,louca_epistemological_2004}.
 \item
   \emph{Epistemological Frames} are the emergent result of subsets of
-  resources acting in concert. Drawing from both Goffman's (1974)
-  sociological notion of frame as structures of expectations and
-  subsequent work on framing in discourse (Tannen, 1993),
+  resources acting in concert. Drawing from both Goffman's
+  sociological notion of frame as structures of expectations \cite{goffman_frame_1974} and
+  subsequent work on framing in discourse \cite{tannen_framing_1993},
   epistemological frames are a participant's local answer to the
   question ``what is it {[}specifically, what knowledge activity{]}
-  that's going on here'' (Goffman, 1974, p. 8).
+  that's going on here'' \cite{goffman_frame_1974}.
 \end{itemize}
@@
 Resources can frames can interact in activity settings to produce
-larger-scale patterns called ``epistemological coherences'' (Rosenberg,
-Hammer, \& Phelan, 2006) where evidence from data suggests that a
+larger-scale patterns called ``epistemological coherences'' \cite{rosenberg_multiple_2006} where evidence from data suggests that a
 network of discrete cognitive units can nonetheless give rise to stable
@@
 
-An example helps ground this in-pieces approach to epistemology. In
-Russ, Coffey, Hammer, and Hutchison (2008), the authors describe the
+An example helps ground this in-pieces approach to epistemology. Russ, Coffey, Hammer, and Hutchison, describe the
 situation where an elementary student reasons about why an empty juice
-box collapses when you suck on the straw. One student gives what the
+box collapses when you suck on the straw \cite{russ_making_2008}. One student gives what the
 authors deem to be an excellent mechanistic account of why the juice box

### 2016-04-23 manuscript-body.tex
Explain how kinematic debugging distributes cognition
and consequently transforms the activity

@@
 to show how a practicing programmer might employ specific, distinct
-conceptual models to reason locally about a piece of code. First,
+conceptual models to reason locally about a piece of code.
+
+\subsubsection{Example 1: Thinking With Kinematics Can Transform a Debugging Problem}
+
+First,
 consider variants of the PASCAL statements from Bonar and Soloway
@@
 
-Consider specifically lines 2 and 11. Line 2 could very well be
+Consider specifically what would happen if we found ourselves needing to debug lines 2 and 11. Line 2 could very well be
 an incrementer in some kind of iterative code. If we had to debug loop code
@@
 We might also call on our knowledge that it's being incremented by 1 to check whether the difference between any two successive printouts is 1; if it isn't, we immediately know something's wrong.
+But to diagnose a potential problem in line 11, we might take an altogether different strategy.
+
 In line
 11's case, it could very well be that line 11's function is animating
-images on a screen. If there's a problem with that code, one of the
-easiest ways we might notice is that the acceleration seems off in the
-animation. Consequently, debugging might call upon knowledge we have
-from kinematics. We might try inserting code that draws a dot at the
-object's on-screen position with each iteration, leaving a trail of dots we can
+images on a screen. Suppose that's the case. If there's a problem with that code, one of the
+easiest ways we might notice is that the \emph{acceleration} seems off in the
+animation. Consequently, in debugging we might call upon knowledge we have
+from kinematics to conceptually model what our code is doing. We might try inserting code that draws a dot at the
+animated object's on-screen position with each iteration, leaving a trail of dots we can
 visually inspect. We could then look at the path of the object's
@@
 first-pass test of whether the code achieves the motion we want.
-(If, for example, we're attempting a cubic easing animation, but the dots are all evenly spaced, then we might look first at the \texttt{displacement(t)}) function and what it's returning.
+(If, for example, our code is supposed to achieve a cubic easing animation, but the dots in the trail are all evenly spaced, basic kinematics tells us our object isn't accelerating at all.
+And, if there's no acceleration, we can pursue a strong hunch that the \texttt{displacement(t)}) function is returning the same constant value with each iternation.
 
 To be clear, it's not just that graphics might improve our efficiency in
-debugging a statement like line 11. Rather, applying an interpretive frame that
+debugging a statement like line 11. Rather, choosing to
+use conceptual knowledge from physics to understand a piece of code transforms debugging a position-update statement
+into a different activity: reasoning kinematically about the properties of motion
+trails.
+For a comparative perspective, consider Hutchins's argument that for aircraft pilots, ringing an aircraft speedometer with physical markers transforms the cognition involved in landing a plane \cite{hutchins_how_1995}:
+
+\begin{quote}
+  Without a speed bug, on final approach the PF [Pilot Flying] must remember the approach speed, read the airspeed indicator scale to find the remembered value of the approach speed on the airspeed indicator scale, and compare the position of the ASI [Airspeed Indicator] needle on the scale with the position of the approach speed on the scale.
+With the salmon bug set, the pilot no longer needs to read the airspeed indicator scale.
+He or she simply looks to see whether or not the indicator needle is lined up with the salmon bug.
+Thus, a memory and scale reading task is transformed into a judgment of spatial adjacency.
+\cite{hutchins_how_1995}
+\end{quote}
+
+Our physics debugging example and Hutchins's speed bug example \cite{hutchins_how_1995} share two core traits:
+
+\begin{enumerate}
+  \tightlist
+  \item Cognition gets \emph{distributed} \cite{hutchins_distributed_2000,hutchins_cognition_1995} across time and space.
+  \item That distribution doesn't simply enhance, but rather \emph{transforms} the nature of the activity, allowing cognitive agents to bring to bear resources that the prior or original context didn't afford.
+\end{enumerate}
+
+In Hutchins's
+
+the cognitive act of debugging in this instance. Applying an interpretive frame that
 treats an assignment statement as a kinematic position update lets us
 \emph{use conceptual knowledge from physics} to diagnose and fix
-problems in code. If we view the assignment statement as proposing
+problems in code.   There,  For comparison, consider
+
+ If we view the assignment statement as proposing
 something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves[blinded]---a
@@
 reason about the code.
+
+\subsection{Example 2: The Same Symbol Template Within The Same Language Can Still Mean Different Things}
 
@@
 \begin{Highlighting}[numbers=left,,]
-\NormalTok{p <-}\StringTok{ }\KeywordTok{InitializeGgplot_1w}\NormalTok{()}
-\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanLine}\NormalTok{(owp)}
-\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanPoint}\NormalTok{(owp)}
-\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleX_1w}\NormalTok{(owp)}
-\NormalTok{p <-}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleY_1w}\NormalTok{(owp)}
+\NormalTok{p =}\StringTok{ }\KeywordTok{InitializeGgplot_1w}\NormalTok{()}
+\NormalTok{p =}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanLine}\NormalTok{(owp)}
+\NormalTok{p =}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{GrandMeanPoint}\NormalTok{(owp)}
+\NormalTok{p =}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleX_1w}\NormalTok{(owp)}
+\NormalTok{p =}\StringTok{ }\NormalTok{p +}\StringTok{ }\KeywordTok{ScaleY_1w}\NormalTok{(owp)}
 \end{Highlighting}
@@
 utility in a programmer having different conceptual metaphors available
-to think about and work with code. Example 1 shows that choosing to
-apply knowledge from physics to a piece of code can change the cognitive
-nature of debugging. There, debugging a position-update statement
-becomes, in part, reasoning kinematically about the properties of motion
-trails.\footnote{For comparison, consider the argument that ringing an
-  aircraft speedometer with physical markers changes the nature of
-  cognition when pilots work to land a plane (Hutchins, 1995b).} Example
+to think about and work with code.
+
+We believe that working from conceptual metaphors---in our example, attacking a debugging problem by exploiting animation and drawing from physics knowledge---achieves the same kind of transformation Hutchins describes.
+
+Example
 2 shows that across languages, programmers might have to deploy

### 2016-04-23 manuscript-body.tex
Add introduction and flesh out juice box pressure example

@@
+\section{Abstract}\label{Abstract}
+
 \section{Introduction}\label{Introduction}
+
+In 1996, learning scientist David Hammer argued ``Not all thoughts students express need to be understood as directly reflecting stable, stored knowledge structures.
+What the misconceptions perspective treats as a stored construct may alternatively be treated as an act of construction.'' \cite{hammer_misconceptions_1996}.
+In the 20 years since, researchers---particularly in science education---have pushed cognitive models beyond the static assumptions of classical misconception accounts of knowledge.
+But, a striking preponderance of computing education research models cognition via misconceptions.
+Such research assumes students have entrenched, wrong ideas about computation.
+And, while misconceptions research in computing education has advanced our understanding of teaching and learning, its narrow also begets problems growing too big to ignore:
+
+\begin{enumerate}
+  \tightlist
+  \item When we use misconceptions to treat student ideas as right or wrong, we establish and perpetuate a deficit model of student understanding.
+  \item The more classical misconceptions-basedresearch we publish as a community, the stronger that deficit narrative becomes in our discourse.
+\end{enumerate}
+
+In this paper, we aim to start remedying those problems.
+We first reanalyze historical data using a knowledge-in-pieces perspective to show how we might recover accounts of students' productive knowledge.
+We then argue through example that expertise in programming can involve the fluid and active construction of conceptual metaphors to deal with idiosyncrasies within and across programming languages.
+These examples help broaden our sense of applicable cognitive theory by showing how distributed cognition can be a powerful framework for analyzing activity.
+Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as knowledge and knowing in computing learning environments.
 
@@
 To be clear, it's not just that graphics might improve our efficiency in
-debugging a statement like line 11. Rather, choosing to
-use conceptual knowledge from physics to understand a piece of code transforms debugging a position-update statement
-into a different activity: reasoning kinematically about the properties of motion
-trails.
-For a comparative perspective, consider Hutchins's argument that for aircraft pilots, ringing an aircraft speedometer with physical markers transforms the cognition involved in landing a plane \cite{hutchins_how_1995}:
+debugging a statement like line 11. Rather, applying an interpretive frame that
+treats an assignment statement as a kinematic position update lets us
+\emph{use conceptual knowledge from physics} to diagnose and fix
+problems in code.
+For a comparative perspective, consider Hutchins's argument that for aircraft pilots, ringing an aircraft speedometer with physical markers transforms the cognition involved in landing a plane:
 
@@
 
-In Hutchins's
+In Hutchins's \cite{hutchins_how_1995} example, pilots distribute cognition across time (1) by making approach speed calculations mid-flight, well before they're needed.
+Pilots then distribute those calculations across space (2) by ringing the speedometer with bugs to represent distinct speeds.
+And, it is this act of distribution that lets pilots use spatial-adjacency perception (visually making sure the needle is on the bugs) during approach instead of having to calculate, compute, and compare speed numbers during approach.
 
-the cognitive act of debugging in this instance. Applying an interpretive frame that
-treats an assignment statement as a kinematic position update lets us
-\emph{use conceptual knowledge from physics} to diagnose and fix
-problems in code.   There,  For comparison, consider
+In our kinematics debugging example, cognition gets distributed across time and space (1) by creating a visual trace of the animated object's position (a persistent trail of dots).
+And, it is this act of dot-tracing that lets a programmer invokve spatial-adjacency perception and conceptual kinematics knowledge (how positions uniformly sampled in time can reflect constant velocity vs.
+acceleration) that wouldn't seem applicable if the reassignment statement were stripped of its context.
+
+the cognitive act of debugging in this instance.    There,  For comparison, consider
 
@@
 
-\subsection{Example 2: The Same Symbol Template Within The Same Language Can Still Mean Different Things}
+\subsubsection{Example 2: The Plus Sign Means Lots of Things}
 
@@
 
-% \begin{supertabular}[]{@{}llll@{}}
-% \toprule
-% \textbf{Statement } & Programming Syntax & Language & How we think about
-% it\tabularnewline
-% \midrule
-% \endhead
-% \textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
-% \textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
-% the string w\tabularnewline
-% \textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
-% layer of points onto a plot p\tabularnewline
-% \bottomrule
-% \end{supertabular}
-
 The catch here is that the plus operator takes on different roles in
@@
 apparent and consequential when, for example, such statements are
-repeated in the same language within the same file. In the R/ggplot2 code below, the author uses multiple
+repeated in the same language within the same file, as we'll see in Example 3.
+
+\subsubsection{Example 3: The Same Symbol Template \emph{Within} The Same Language Can Still Mean Different Things}
+
+In the R/ggplot2 code below, the author uses multiple
 reassignment statements to compose a statistical graphics plot.
@@
   actually correspond to very different operations in code (example 2).
-  And, that's true both within and across languages.
+  And, that's true both within and across languages (example 3).
 \item
@@
 
-\subsection{Manifold models of cognition explain context-dependence
+\subsection{Manifold models of cognition can explain context-dependence
 and the growth of
@@
 conceptual entity depending on context. In several different
-situations---wrapped around a newspaper, pulled taut as a string, spun
+situations---wrapped around a newspaper, suspending a bob weight, pulled taut as a string, spun
 to store energy in a toy plane propeller---we intuitively think about
 the rubber band's physical behavior differently: one as a negligible
-part of the newspaper's point mass, another as a transverse pendulum
-(likely) obeying Hooke's Law, and the third as a torsional spring. Those
+part of the newspaper's point mass, two different kinds of harmonic pendulums, and finally a torsional spring. Those
 differences in intuitive thinking reflect the contextual dependency of
@@
 authors deem to be an excellent mechanistic account of why the juice box
-collapses. But, as the teacher seems to steer the discussion toward
-vocabulary---in this case, ``pressure''---the student clearly pulls back
+collapses:
+
+\begin{quote}
+  In explaining the phenomenon, Erin focuses on the role played by the air located inside the juice box.
+She describes the air inside as actively pushing out on all sides of the box holding them out and flat.
+When that air is removed from the box (by sucking it out through the straw), there is no longer anything pushing from the inside to hold the box out, so the sides cave in.
+We call this description of the phenomenon the “inside-pusher” model because of its focus on what happens inside the box.
+\cite{russ_making_2008}
+\end{quote}
+
+But, as the teacher seems to steer the discussion toward canonically correct
+vocabulary---in this case, ``pressure''---and Erin clearly pulls back
 from her mechanistic reasoning, seems much more diffident, and claims
@@
 pressure works was left nonetheless with the impression that pressure
-was hard to explain.\footnote{In fairness to the teacher, I'm trying to
-  focus on the result of the interaction and far less on the intent the
-  teacher might have had. The student still left with a sense that
-  science might be about vocabulary, even if that's not the view the
-  teacher would espouse or was trying to enact in the moment.}
+was hard to explain.
+In other words, in that moment what counted as knowing was using the word pressure.
+Erin's mechanistic explanation---beautiful and complex as it was---didn't use that magic word, and in a possible effort to push the class toward canonical behavior, the teacher shut down Erin's reasoning.
 

### 2016-04-23 manuscript-body.tex
Augment introduction
turn bold type into typewriter type for meats and sandwich example

@@
 
-In 1996, learning scientist David Hammer argued ``Not all thoughts students express need to be understood as directly reflecting stable, stored knowledge structures.
+Science and math education started challenging misconception models of mind in 1993 \cite{disessa_epistemology_1993,smith_misconceptions_1993}.
+By 1996, learning scientists were boldly repudiating the assumptions of misconceptions: ``Not all thoughts students express need to be understood as directly reflecting stable, stored knowledge structures.
 What the misconceptions perspective treats as a stored construct may alternatively be treated as an act of construction.'' \cite{hammer_misconceptions_1996}.
-In the 20 years since, researchers---particularly in science education---have pushed cognitive models beyond the static assumptions of classical misconception accounts of knowledge.
-But, a striking preponderance of computing education research models cognition via misconceptions.
+In the more than 20 years since those papers were published, researchers---particularly in science education---have pushed cognitive models beyond the static assumptions of classical misconception accounts of knowledge.
+But, a preponderance of research in computing education (CSEd) has modeled and continues to model cognition through misconceptions.
 Such research assumes students have entrenched, wrong ideas about computation.
-And, while misconceptions research in computing education has advanced our understanding of teaching and learning, its narrow also begets problems growing too big to ignore:
+And, while misconceptions research in computing education has advanced our understanding of teaching and learning, its narrow focus also begets problems that are growing too big to ignore:
 
@@
 
-In this paper, we aim to start remedying those problems.
-We first reanalyze historical data using a knowledge-in-pieces perspective to show how we might recover accounts of students' productive knowledge.
+This paper aims squarely at those problems by expanding a sense of what's possible in modeling cognition within computing education research. We argue research
+approaches that inflexibly privilege canonical knowledge do so at the
+expense of other productive knowledge and ways of knowing that students
+have.
+First, we reanalyze prior data to show how we might recast misconceptions as the potential seeds of productive knowledge.
 We then argue through example that expertise in programming can involve the fluid and active construction of conceptual metaphors to deal with idiosyncrasies within and across programming languages.
-These examples help broaden our sense of applicable cognitive theory by showing how distributed cognition can be a powerful framework for analyzing activity.
+These examples help broaden our sense of applicable cognitive theory by showing how distributed cognition and symbolic forms can be a powerful framework for analysis in CSEd.
 Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as knowledge and knowing in computing learning environments.
@@
 knowledge}\label{misconceptions-research-in-computing-education-tends-to-ignore-students-productive-knowledge}
-
-we argue research
-approaches that inflexibly privilege canonical knowledge do so at the
-expense of other productive knowledge and ways of knowing that students
-have.
 
@@
 potentially ``\emph{productive}'' \cite{hammer_form_2002,hammer_tapping_2003} insights about design. Precisely \emph{because}
-there is no code stating that \textbf{turkey} and \textbf{roastBeef} and
+there is no code stating that \texttt{turkey} and \texttt{roastBeef} and
 ham are sorted into the array, the student is \emph{inferring} that to
@@
 \emph{would} be sensible to design a program where specific instances of
-classes \textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} could go
-into an array of type \textbf{Meat}. To do so, a designer could define
-\textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} as inheriting
-from \textbf{Meat}.
+classes \texttt{Turkey} and \texttt{Ham} and \texttt{RoastBeef} could go
+into an array of type \texttt{Meat}. To do so, a designer could define
+\texttt{Turkey} and \texttt{Ham} and \texttt{RoastBeef} as inheriting
+from \texttt{Meat}.
 

### 2016-04-23 manuscript-body.tex
Fix typos; add Stefik text (need cite); add introduction

@@
 \section{Abstract}\label{Abstract}
+
+This paper aims to expand our sense of what's possible in modeling cognition within computing education research. We argue that research
+approaches that privilege canonical knowledge do so at the
+expense of other productive knowledge and ways of knowing that students
+have.
+We explore applicable cognitive theory by showing how distributed cognition and symbolic forms can be a powerful framework for analysis in CSEd.
+Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as \texttt{knowledge} and \texttt{knowing} in computing learning environments.
 
@@
   \item When we use misconceptions to treat student ideas as right or wrong, we establish and perpetuate a deficit model of student understanding.
-  \item The more classical misconceptions-basedresearch we publish as a community, the stronger that deficit narrative becomes in our discourse.
+  \item The more classical misconceptions-based research we publish as a community, the stronger that deficit narrative becomes in our discourse.
 \end{enumerate}
@@
 These examples help broaden our sense of applicable cognitive theory by showing how distributed cognition and symbolic forms can be a powerful framework for analysis in CSEd.
-Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as knowledge and knowing in computing learning environments.
+Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as \texttt{knowledge} and \texttt{knowing} in computing learning environments.
 
@@
 \texttt{meats} array. First, there is no code here that places
-\texttt{turkey} and \texttt{ham} in the array; there is only code that
+\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} in the array; 
+there is only code that
 declares them as variables. Moreover, as written, an attempt to place
@@
 would fail. Because of type restrictions in Java, only objects of class
-Meat (or objects that inherit from class Meat) can go in the array.
+\texttt{Meat} (or objects that inherit from class \texttt{Meat}) can go in the array.
 \texttt{turkey} and \texttt{ham} and \texttt{roastBeef} are, perhaps
@@
 there is no code stating that \texttt{turkey} and \texttt{roastBeef} and
-ham are sorted into the array, the student is \emph{inferring} that to
+\texttt{ham} are sorted into the array, the student is \emph{inferring} that to
 be true. And, while that behavior is not what's happening, it
@@
 
-My second example of research that criticizes students' non-canonical
+Our second example of research that criticizes students' non-canonical
 understandings comes from Bonar and Soloway's 1983 study of Pascal
@@
 
-These two purposes, which for the sake of description I'll call
+These two purposes, which for the sake of description we'll call
 \emph{keeping control} (1a) and \emph{totaling up} (2a) are, in a sense,
@@
 conceptual schemata. Parts-of-a-whole refers to the contributions of
-component entities while base±change describes a kind of accumulation \cite{sherin_how_2001}.
-Specifically, the terms in base±change ``contribute to a whole but play different roles.
-One is a base value; the other is a change to that base.'' But, the two distinct conceptual schemata share what I
+component entities while \emph{base+change} describes a kind of accumulation \cite{sherin_how_2001}.
+Specifically, the terms in \emph{base+change} ``contribute to a whole but play different roles.
+One is a base value; the other is a change to that base.'' But, the two distinct conceptual schemata share what we
 would argue is the same symbolic structure: $\Box=︎ \Box + \Box$ (parts of a whole) and $\Box = \Box \pm \Delta$ (base + change) \cite{sherin_how_2001}.
@@
 To return to misconceptions, what seems to drive research on students'
-misconceptions is computing education largely a need to get students to program computers
+misconceptions in computing education is largely a need to get students to program computers
 and reason about computation in ways that are canonically correct. And,
-that's a worthwhile goals. But, as I've argued, we can
+that's a worthwhile goal. But, as we've argued, we can
 already identify cases where a narrow misconceptions focus is silent
@@
 It's important to note that misconceptions research in computing didn't
-always treat students' non-canonical knowledge as a problem. we begin the
+always treat students' non-canonical knowledge as a problem. We begin the
 next section by backtracing to some of the earliest work on students'
@@
 instead see them as an opportunity for programming language designers to
-make languages better.
+make languages better. Indeed, an empirical approach to language design has recently been articulated in Stefik and Siebert's work regarding language syntax's effect on novices. \cite{}
 
 That insight---that instruction and design can meet novices where they
-are---carries through to their final remarks about studying and
+are---carries through to Bonar and Soloway's final remarks about studying and
 analyzing novice programming knowledge:
@@
 something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves[blinded]---a
-field of knowledge and concomitant techniques from physics becomes
+field of knowledge and concomitant techniques from physics become
 available to me to think with. But, if we focus only on the
@@
 But, another phenomenon comes
-into play \emph{across} different languages when use the same symbology (written patterns) to stand for
+into play \emph{across} different languages when they use the same symbology (written patterns) to stand for
 conceptually different operations. The statements below are examples of what look
@@
 interpretive frame than incrementation is that numeric addition is commutative; composing a
-plot is not necessarily commutative. So, despite the syntactic similarity, reassignments that
+plot is not necessarily commutative. So, despite the syntactic similarity, statements that
 compose a plot using reassignment (as above) do not obey the same
@@
 
-Resources can frames can interact in activity settings to produce
+Resources and frames can interact in activity settings to produce
 larger-scale patterns called ``epistemological coherences'' \cite{rosenberg_multiple_2006} where evidence from data suggests that a

### 2016-04-23 manuscript-body.tex
add conclusion; change title of paper

@@
 instead see them as an opportunity for programming language designers to
-make languages better. Indeed, an empirical approach to language design has recently been articulated in Stefik and Siebert's work regarding language syntax's effect on novices. \cite{}
+make languages better.
+Indeed, an empirical approach to language design has recently been articulated in Stefik and Siebert's work regarding language syntax's effect on novices.
+\cite{stefik_intuitive_2011}
 
@@
 
+In conclusion, we hold that Erin's reasoning---and that of other sense-makers---should count as valid ways of \texttt{knowing} in computing education and computing education research and should be attended to in computing learning environments.
+Modeling cognition only or primarily as a set of privileged, canonical knowledge is done at the expense of other productive knowledge and ways of knowing that students have.
+
 \subsection{Acknowledgments}\label{acknowledgments}

### 2016-04-23 manuscript-body.tex
Enhance Wil's conclusion

--- 2016-04-23 manuscript-body.tex
@@
 
+\section{Conclusion}
+
+In this article, we have argued for expanding our cognitive models beyond misconceptions in computing education research. We
+
+\begin{enumerate}
+  \tightlist
+  \item Reanalyzed historical data through the perspective of productive student knowledge, building evidence in classic computing education research that we can recover novice's productive knowledge without simply labeling it as non-canonical or wrong
+  \item Explored several examples that reveal both the adaptive nature of computing expertise and the power of non-misconception frameworks---namely distributed cognition and knowledge-in-pieces/symbolic forms.
+  \item Stressed that models to what counts as knowledge and knowing have both historical success in other fields and obvious implications for computing education
+\end{enumerate}
+
+Ultimately, the modeling of cognition only or primarily as a set of privileged, canonical knowledge and how students do or don't have it is eclipsing the other productive knowledge and ways of knowing that students do have.
+While useful, misconceptions can only take us so far in our efforts to understand and improve learning in computing education.
+
 \subsection{Acknowledgments}\label{acknowledgments}

### 2016-04-23 manuscript-body.tex
fix curly quotes; fix a couple of typos

@@
 
-\hypertarget{sandiwches-code}{\label{sandwiches-code}}
+\hypertarget{sandwiches-code}{\label{sandwiches-code}}
 \begin{Shaded}
@@
 In our kinematics debugging example, cognition gets distributed across time and space (1) by creating a visual trace of the animated object's position (a persistent trail of dots).
-And, it is this act of dot-tracing that lets a programmer invokve spatial-adjacency perception and conceptual kinematics knowledge (how positions uniformly sampled in time can reflect constant velocity vs.
+And, it is this act of dot-tracing that lets a programmer invoke spatial-adjacency perception and conceptual kinematics knowledge (how positions uniformly sampled in time can reflect constant velocity vs.
 acceleration) that wouldn't seem applicable if the reassignment statement were stripped of its context.
@@
  If we view the assignment statement as proposing
-something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves[blinded]---a
+something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves [blinded]---a
 field of knowledge and concomitant techniques from physics become
@@
 When that air is removed from the box (by sucking it out through the straw), there is no longer anything pushing from the inside to hold the box out, so the sides cave in.
-We call this description of the phenomenon the “inside-pusher” model because of its focus on what happens inside the box.
+We call this description of the phenomenon the "inside-pusher" model because of its focus on what happens inside the box.
 \cite{russ_making_2008}
